### 2020-04-28 issues.Rmd
Setting up the repo and listing issues

---
title: "Replicating Blair and Sambanis: Forecasting Civil Wars - JCR 2020"
output: github_document
---

### Introduction

While reading Blair and Sambanis (2020), a few issues stuck out. We downloaded the replication files authors provide from https://journals.sagepub.com/doi/suppl/10.1177/0022002720918923/suppl_file/sj-zip-1-jcr-10.1177_0022002720918923.zip. The original folder name of is `sj-zip-1-jcr-10.1177_0022002720918923/replication-3`. For simplicity, we extracted this `replication-3` and saved to the `Replication Files` folder of this repo.

Before listing the primary issues we identified, we should note that we are partial to their approach. Theory is important. It is helpful for forecasting problems, especially with regards to identifying which variables to include in a model. This is particularly true when the goal is to assess how ...

This document lists the primary issues we identified.

1. Improperly tuning of models
  - RM: In a twitter DM, I asked Blair to clarify -- "[I] was wondering how you all came to your tuning procedure for the RF model hyper-parameters?" He responded "mostly trial and error, honestly. ... Trial and error was on early data. Forecasts were for much later data"
2. Using a regression framework in a classification problem
3. Their train/validate/test approach
4. Rounding of AUC/ROC scores
5. Lack of yearly test forecasts in favor of a single 5-year test forecast
6. Table 4! 
    - There are only 20 onsets from 2001 to 2015 But there are 15 "Persistence" cases (top of table 4) and 16 "Change" cases (bottom of table 4)
    - Why only top 30?
7. Treating the output of a RF regression model as Pr() -- (RM: I guess this is akin to a Linear Probability Model, but I'm not sure if this is possible in a RF framework
8. The lack of procedures to account for rare events in an RF model
9.

### 2020-04-28 Replication files/replication-3/+master.R

@@ -1,0 +1,507 @@
+# Setup ------------------------------------------------------------
+
+         set.seed(92382)
+
+    # Clear environment
+    
+         rm(list = ls())
+    
+    # Set working directory
+    
+          setwd("YOUR DIRECTORY")
+    
+    # Load packages
+    
+          library(foreign)
+          library(separationplot)
+          library(dyn)
+          library(gdata)
+          library(ROCR)
+          library(readstata13)
+          library(verification)		
+          library(WriteXLS)
+          library(randomForest)
+          library(scales)
+          library(pROC)
+          library(extrafont)
+          library(fontcm)
+          library(corpcor)
+          library(ggplot2)
+          library(fontcm)
+          library(PRROC)
+          library(Bolstad2)
+          library(separationplot)
+
+    # Specify length of ROC curve
+    
+          test_observations <- 10000
+
+
+
+# Define 1-month models ------------------------------------------------------------
+
+    # Load data
+
+          data <- read.dta13("data/1mo_data.dta")
+
+    # Define training and testing sets for base specification
+          
+          train_period = mean(data$period[which(data$month==12 & data$year==2007)])
+          end_period = mean(data$period[which(data$month==12 & data$year==2015)])
+          train <- data[data$period<=train_period,]
+          test <- data[data$period>train_period & data$period<=end_period,]
+ 
+    # Define training and testing sets for robustness checks using alternate start dates
+          
+          train_period_robust_traintest1 = mean(data$period[which(data$month==12 & data$year==2008)])
+          end_period = mean(data$period[which(data$month==12 & data$year==2015)])
+          train_robust_traintest1 <- data[data$period<=train_period_robust_traintest1,]
+          test_robust_traintest1 <- data[data$period>train_period_robust_traintest1 & data$period<=end_period,]
+          
+          train_period_robust_traintest2 = mean(data$period[which(data$month==12 & data$year==2009)])
+          end_period = mean(data$period[which(data$month==12 & data$year==2015)])
+          train_robust_traintest2 <- data[data$period<=train_period_robust_traintest2,]
[…]

### 2020-04-28 Replication files/replication-3/code/1mo_define_models.R

@@ -1,0 +1,2582 @@
+
+
+  # Define model labels
+		
+    			escalation <- c(
+    				"gov_opp_low_level",
+    				"gov_reb_low_level",
+    				"opp_gov_low_level",
+    				"reb_gov_low_level",
+    				"gov_opp_nonviol_repression",
+    				"gov_reb_nonviol_repression",
+    				"gov_opp_accommodations",
+    				"gov_reb_accommodations",
+    				"reb_gov_demands",
+    				"opp_gov_demands"
+    				)
+		
+    			quad <- c(
+      			"gov_opp_vercf","gov_reb_vercf",
+    				"gov_opp_matcf","gov_reb_matcf",
+    				"opp_gov_vercf","reb_gov_vercf",
+    				"opp_gov_matcf","reb_gov_matcf",
+    				"gov_opp_vercp","gov_reb_vercp",
+    				"gov_opp_matcp","gov_reb_matcp",
+    				"opp_gov_vercp","reb_gov_vercp",
+    				"opp_gov_matcp","reb_gov_matcp"
+    				)
+
+        	goldstein <- c(
+      			"gov_opp_gold","gov_reb_gold",
+    				"opp_gov_gold","reb_gov_gold"
+            )
+
+      		all_CAMEO <- c(
+      			"gov_opp_cameo_010",
+    				"gov_opp_cameo_011",
+    				"gov_opp_cameo_012",
+    				"gov_opp_cameo_013",
+    				"gov_opp_cameo_014",
+    				"gov_opp_cameo_015",
+    				"gov_opp_cameo_016",
+    				"gov_opp_cameo_017",
+    				"gov_opp_cameo_018",
+    				"gov_opp_cameo_019",
+    				"gov_opp_cameo_020",
+    				"gov_opp_cameo_021",
+    				"gov_opp_cameo_0211",
+    				"gov_opp_cameo_0212",
+    				"gov_opp_cameo_0213",
+    				"gov_opp_cameo_0214",
+    				"gov_opp_cameo_022",
+    				"gov_opp_cameo_023",
+    				"gov_opp_cameo_0231",
+    				"gov_opp_cameo_0232",
+    				"gov_opp_cameo_0233",
+    				"gov_opp_cameo_0234",
+    				"gov_opp_cameo_024",
+    				"gov_opp_cameo_0241",
+    				"gov_opp_cameo_0242",
[…]

### 2020-04-28 Replication files/replication-3/code/1mo_make_ROC.R

@@ -1,0 +1,49 @@
+
+      loadfonts()
+
+    # Open plot window
+
+      quartz()
+
+      plot.new()
+
+    # Save PDF
+      
+  		pdf("figures/figure1_left.pdf", family = "CM Roman", width = 5.5, height = 5)
+      
+    # Plot ROC curves   
+      
+      plot(roc_escalation_inc_civil_ns,
+		    ylab = "True positive rate",
+		    xlab = "False positive rate",
+		    #col = c("brown"),
+		    lty = 1
+        )
+      
+      lines.roc(roc_quad_inc_civil_ns,
+        lty = 2
+        )
+    
+      lines.roc(roc_goldstein_inc_civil_ns,
+        lty = 3
+        )
+
+      lines.roc(roc_all_CAMEO_inc_civil_ns,
+        lty = 4
+        )
+      
+      lines.roc(roc_avg_inc_civil_ns,
+        lty = 5
+        )
+
+    # Add legend
+      
+  		legend("bottomright",
+  			c("Escalation", "Quad", "Goldstein", "CAMEO", "Average"),
+  			  lty = c(1, 2, 3, 4, 5) 
+  		      )
+
+      dev.off()
+      
+      
+      

### 2020-04-28 Replication files/replication-3/code/1mo_make_importance.R

@@ -1,0 +1,1282 @@
+
+  # Create left panel of Figure A11
+
+      importance <- RF_ranking_escalation_inc_civil_ns
+      
+      rownames(importance) <-  c("Gov -> opp low-level violence", 
+                                    "Gov -> reb low-level violence", 
+                                    "Opp -> gov low-level violence", 
+                                    "Reb -> gov low-level violence", 
+                                    "Gov -> opp non-violent repression", 
+                                    "Gov -> reb non-violent repression", 
+                                    "Gov -> opp accommodations", 
+                                    "Gov -> reb accommodations", 
+                                    "Reb -> gov demands",
+                                    "Opp -> gov demands")
+      
+      importance.sorted <- importance[order(importance$Importance.score),]
+
+      quartz()
+      
+      plot.new()
+      
+      pdf("figures/figureA11_left.pdf", family = "CM Roman", width = 5.5, height = 5)
+      
+      dotchart(importance.sorted$Importance.score,labels=row.names(importance.sorted), xlab="Decrease in MSE", cex=.7)
+      
+      dev.off()
+
+
+      
+      
+  # Create left panel of Figure A12
+      
+      importance <- RF_ranking_quad_inc_civil_ns
+      
+      rownames(importance) <-  c("Gov -> opp verbal conflict", 
+                                 "Gov -> reb verbal conflict",
+                                 "Gov -> opp material conflict",
+                                 "Gov -> reb material conflict",
+                                 "Opp -> gov verbal conflict",
+                                 "Reb -> gov verbal conflict",
+                                 "Opp -> gov material conflict",
+                                 "Reb -> gov material conflict",
+                                 "Gov -> opp verbal cooperation",
+                                 "Gov -> reb verbal cooperation",
+                                 "Gov -> opp material cooperation",
+                                 "Gov -> reb material cooperation",
+                                 "Opp -> gov verbal cooperation",
+                                 "Reb -> gov verbal cooperation",
+                                 "Opp -> gov material cooperation",
+                                 "Reb -> gov material cooperation"
+      )
+      
+      importance.sorted <- importance[order(importance$Importance.score),]
+
+      quartz()
+      
+      plot.new()
+      
[…]

### 2020-04-28 Replication files/replication-3/code/1mo_run_escalation.R

@@ -1,0 +1,219 @@
+
+			# Run random forests models			
+				         
+
+    				train_model <- randomForest(escalation_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+      			train_model_robust_maxnodes <- randomForest(escalation_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+        		train_model_robust_sampsize <- randomForest(escalation_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 500, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+      			train_model_robust_ntree <- randomForest(escalation_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+        		train_model_robust_traintest1 <- randomForest(escalation_train_formula_civil_ns_robust_traintest1,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_train_frame_civil_ns_robust_traintest1, forest = TRUE   
+    												  )			         
+
[…]

### 2020-04-28 Replication files/replication-3/code/1mo_run_escalation_simple.R

@@ -1,0 +1,30 @@
+
+			# Run random forests models			
+				         
+    				train_model <- randomForest(escalation_simple_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_simple_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+
+			# Generate out-of-sample predictions
+				
+    				prediction <- as.numeric(predict(train_model, newdata = data.matrix(escalation_simple_test_frame_civil_ns), type="response"))      								
+       				prediction[is.na(escalation_simple_test_frame_civil_ns$test_DV_civil_ns)] <- NA
+
+ 
+
+			# Calculate AUC
+			
+    				performance <- prediction(prediction, test_DV_civil_ns)
+              roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+    				    AUC_obs <- data.frame(as.numeric(roc$auc))
+
+
+			# Compile AUCs
+              
+              AUCs_escalation_simple <- AUC_obs
+

### 2020-04-28 Replication files/replication-3/code/1mo_run_escalation_weighted_PITF.R

@@ -1,0 +1,22 @@
+
+      # Weight predicted probabilities from escalation model
+
+            weight <- as.numeric(train$pred_prob_plus1)
+            length(weight) <- length(prediction_escalation_inc_civil_ns)
+            
+            prediction = prediction_escalation_inc_civil_ns*weight
+            prediction[is.na(escalation_test_frame_civil_ns$test_DV_civil_ns)] <- NA
+
+
+			# Calculate AUC
+			
+    				performance <- prediction(prediction, test_DV_civil_ns)
+              roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+    				    AUC_obs <- data.frame(as.numeric(roc$auc))
+
+
+			# Compile AUCs
+              
+              AUCs_escalation_weighted_PITF <- AUC_obs
+
+

### 2020-04-28 Replication files/replication-3/code/1mo_run_escalation_withlags.R

@@ -1,0 +1,30 @@
+
+			# Run random forests models			
+				         
+    				train_model <- randomForest(escalation_withlags_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_withlags_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+
+			# Generate out-of-sample predictions
+				
+    				prediction <- as.numeric(predict(train_model, newdata = data.matrix(escalation_withlags_test_frame_civil_ns), type="response"))      								
+       				prediction[is.na(escalation_withlags_test_frame_civil_ns$test_DV_civil_ns)] <- NA
+
+ 
+
+			# Calculate AUC
+			
+    				performance <- prediction(prediction, test_DV_civil_ns)
+              roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+    				    AUC_obs <- data.frame(as.numeric(roc$auc))
+
+
+			# Compile AUCs
+              
+              AUCs_escalation_withlags <- AUC_obs
+

### 2020-04-28 Replication files/replication-3/code/1mo_run_goldstein.R

@@ -1,0 +1,219 @@
+
+			# Run random forests models			
+				         
+
+    				train_model <- randomForest(goldstein_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = goldstein_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+      			train_model_robust_maxnodes <- randomForest(goldstein_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = goldstein_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+        		train_model_robust_sampsize <- randomForest(goldstein_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 500, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = goldstein_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+      			train_model_robust_ntree <- randomForest(goldstein_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = goldstein_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+        		train_model_robust_traintest1 <- randomForest(goldstein_train_formula_civil_ns_robust_traintest1,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = goldstein_train_frame_civil_ns_robust_traintest1, forest = TRUE   
+    												  )			         
+
[…]

### 2020-04-28 Replication files/replication-3/code/6mo_define_models_OOS.R

@@ -1,0 +1,33 @@
+
+    # Define model labels
+
+    			escalation_6mo <- c(
+    				"gov_opp_low_level",
+    				"gov_reb_low_level",
+    				"opp_gov_low_level",
+    				"reb_gov_low_level",
+    				"gov_opp_nonviol_repression",
+    				"gov_reb_nonviol_repression",
+    				"gov_opp_accommodations",
+    				"gov_reb_accommodations",
+    				"reb_gov_demands",
+    				"opp_gov_demands"
+    				)
+    
+    # Define predictors for base specification
+    			
+      			escalation_6mo_train_RHS <- train[escalation_6mo]
+      			escalation_6mo_test_RHS <- test[escalation_6mo]
+
+    # Append predictand for base specification
+      				
+      			escalation_6mo_train_frame_civil_ns = data.frame(train_DV_civil_ns, escalation_6mo_train_RHS)
+      			escalation_6mo_test_frame_civil_ns = data.frame(test_DV_civil_ns, escalation_6mo_test_RHS)
+
+    # Define model for base specification
+      			
+      			escalation_6mo_train_formula_civil_ns <- as.formula(paste("train_DV_civil_ns",paste(escalation_6mo, collapse=" + "), sep = "~", collapse=NULL))
+      			escalation_6mo_test_formula_civil_ns <- as.formula(paste("test_DV_civil_ns",paste(escalation_6mo, collapse=" + "), sep = "~", collapse=NULL))
+
+
+

### 2020-04-28 Replication files/replication-3/code/6mo_make_PR.R

@@ -1,0 +1,30 @@
+
+      loadfonts()
+
+    # Open plot window
+
+      quartz()
+
+      plot.new()
+      
+    # Save PDF
+      
+      pdf("figures/figure2_right.pdf", family = "Times", width = 5.5, height = 5)
+
+    # Plot PR curves
+
+      plot(precision_recall_escalation_6mo_inc_civil_ns,lty = 1)
+      plot(precision_recall_quad_6mo_inc_civil_ns,lty = 2,add=T)
+      plot(precision_recall_goldstein_6mo_inc_civil_ns,lty = 3,add=T)
+      plot(precision_recall_all_CAMEO_6mo_inc_civil_ns,lty = 4,add=T)
+      plot(precision_recall_avg_6mo_inc_civil_ns,lty = 5,add=T)
+
+    # Add legend
+      
+      legend("topright",
+             c("Escalation", "Quad", "Goldstein", "CAMEO", "Average"),
+             lty = c(1, 2, 3, 4, 5) 
+              )
+
+      dev.off()
+

### 2020-04-28 Replication files/replication-3/code/6mo_make_importance.R

@@ -1,0 +1,1282 @@
+
+  # Create right panel of Figure A11
+
+      importance <- RF_ranking_escalation_6mo_inc_civil_ns
+      
+      rownames(importance) <-  c("Gov -> opp low-level violence", 
+                                    "Gov -> reb low-level violence", 
+                                    "Opp -> gov low-level violence", 
+                                    "Reb -> gov low-level violence", 
+                                    "Gov -> opp non-violent repression", 
+                                    "Gov -> reb non-violent repression", 
+                                    "Gov -> opp accommodations", 
+                                    "Gov -> reb accommodations", 
+                                    "Reb -> gov demands",
+                                    "Opp -> gov demands")
+      
+      importance.sorted <- importance[order(importance$Importance.score),]
+
+      quartz()
+      
+      plot.new()
+      
+      pdf("figures/figureA11_right.pdf", family = "CM Roman", width = 5.5, height = 5)
+      
+      dotchart(importance.sorted$Importance.score,labels=row.names(importance.sorted), xlab="Decrease in MSE", cex=.7)
+      
+      dev.off()
+
+
+      
+      
+  # Create right panel of Figure A12
+      
+      importance <- RF_ranking_quad_6mo_inc_civil_ns
+      
+      rownames(importance) <-  c("Gov -> opp verbal conflict", 
+                                 "Gov -> reb verbal conflict",
+                                 "Gov -> opp material conflict",
+                                 "Gov -> reb material conflict",
+                                 "Opp -> gov verbal conflict",
+                                 "Reb -> gov verbal conflict",
+                                 "Opp -> gov material conflict",
+                                 "Reb -> gov material conflict",
+                                 "Gov -> opp verbal cooperation",
+                                 "Gov -> reb verbal cooperation",
+                                 "Gov -> opp material cooperation",
+                                 "Gov -> reb material cooperation",
+                                 "Opp -> gov verbal cooperation",
+                                 "Reb -> gov verbal cooperation",
+                                 "Opp -> gov material cooperation",
+                                 "Reb -> gov material cooperation"
+      )
+      
+      importance.sorted <- importance[order(importance$Importance.score),]
+
+      quartz()
+      
+      plot.new()
+      
[…]

### 2020-04-28 Replication files/replication-3/code/6mo_make_separation.R

@@ -1,0 +1,193 @@
+
+    # Load modified separation plot function from "separationplot" package
+      
+      separationplot_alt <-
+        function(pred, actual, type="line", line=T, lwd1=0.5, lwd2=1, heading="", xlab="", shuffle=T, width=8, height=2, col0="#FEF0D9", col1="#E34A33", flag=NULL, flagcol=1, file=NULL, newplot=T, locate=NULL, rectborder=NA, show.expected=F, zerosfirst=T, BW=F, family = "Times"){
+          
+          if (is.vector(pred)==F) stop("The pred argument needs to be a vector")
+          if (is.vector(actual)==F) stop("The actual argument needs to be a vector")
+          if (length(pred)!=length(actual)) stop("The pred and actual vectors are of different lengths.")
+          if (any(is.na(pred))) stop("Missing values in the pred vector.")
+          if (any(is.na(actual))) stop("Missing values in the actual vector.")
+          
+          resultsmatrix<-data.frame(pred, actual, flags=0)
+          rows<-nrow(resultsmatrix)
+          
+          if (!is.null(flag)) resultsmatrix$flags[flag]<-1
+          
+          if (shuffle==T){
+            set.seed(1)
+            resultsmatrix<-resultsmatrix[sample(1:rows,rows),]
+          } 
+          
+          resultsmatrix<-resultsmatrix[order(resultsmatrix$pred,resultsmatrix$actual),]
+          resultsmatrix<-cbind(resultsmatrix, position=1:rows)
+          
+          if (type=="bands"){width<-6; height<-2} 
+          
+          if (BW==T){col0="#FFFFFF"; col1="#636363"} 
+          
+          if (newplot==T){
+            if (is.null(file)) dev.new(width=width, height=height)
+            if (!is.null(file)) pdf(file=file, width=width, height=height)
+            par(mgp=c(3,0,0), lend=2, mar=c(3,2,2,2))	
+          } 
+          
+          if (type!="bands"){	 			
[…]

### 2020-04-28 Replication files/replication-3/code/6mo_run_avg.R

@@ -1,0 +1,156 @@
+
+      # Average predicted probabilities across models
+
+            prediction <- as.numeric((prediction_escalation_6mo_inc_civil_ns + 
+                            prediction_quad_6mo_inc_civil_ns + 
+                            prediction_goldstein_6mo_inc_civil_ns + 
+                            prediction_all_CAMEO_6mo_inc_civil_ns)
+                         /4)
+
+            prediction_robust_maxnodes <- as.numeric((prediction_robust_maxnodes_escalation_6mo_inc_civil_ns + 
+                                            prediction_robust_maxnodes_quad_6mo_inc_civil_ns + 
+                                            prediction_robust_maxnodes_goldstein_6mo_inc_civil_ns + 
+                                            prediction_robust_maxnodes_all_CAMEO_6mo_inc_civil_ns)
+                                         /4)
+
+            prediction_robust_sampsize <- as.numeric((prediction_robust_sampsize_escalation_6mo_inc_civil_ns + 
+                                            prediction_robust_sampsize_quad_6mo_inc_civil_ns + 
+                                            prediction_robust_sampsize_goldstein_6mo_inc_civil_ns + 
+                                            prediction_robust_sampsize_all_CAMEO_6mo_inc_civil_ns)
+                                         /4)
+
+            prediction_robust_ntree <- as.numeric((prediction_robust_ntree_escalation_6mo_inc_civil_ns + 
+                                         prediction_robust_ntree_quad_6mo_inc_civil_ns + 
+                                         prediction_robust_ntree_goldstein_6mo_inc_civil_ns + 
+                                         prediction_robust_ntree_all_CAMEO_6mo_inc_civil_ns)
+                                      /4)
+
+            prediction_robust_traintest1 <- as.numeric((prediction_robust_traintest1_escalation_6mo_inc_civil_ns + 
+                                              prediction_robust_traintest1_quad_6mo_inc_civil_ns + 
+                                              prediction_robust_traintest1_goldstein_6mo_inc_civil_ns + 
+                                              prediction_robust_traintest1_all_CAMEO_6mo_inc_civil_ns)
+                                           /4)
+
[…]

### 2020-04-28 Replication files/replication-3/code/6mo_run_escalation_simple.R

@@ -1,0 +1,30 @@
+
+			# Run random forests models			
+				         
+    				train_model <- randomForest(escalation_6mo_simple_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_6mo_simple_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+
+			# Generate out-of-sample predictions
+				
+    				prediction <- as.numeric(predict(train_model, newdata = data.matrix(escalation_6mo_simple_test_frame_civil_ns), type="response"))      								
+       				prediction[is.na(escalation_6mo_simple_test_frame_civil_ns$test_DV_civil_ns)] <- NA
+
+ 
+
+			# Calculate AUC
+			
+    				performance <- prediction(prediction, test_DV_civil_ns)
+              roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+    				    AUC_obs <- data.frame(as.numeric(roc$auc))
+
+
+			# Compile AUCs
+              
+              AUCs_escalation_6mo_simple <- AUC_obs
+

### 2020-04-28 Replication files/replication-3/code/6mo_run_escalation_withlags.R

@@ -1,0 +1,30 @@
+
+			# Run random forests models			
+				         
+    				train_model <- randomForest(escalation_6mo_withlags_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_6mo_withlags_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+
+			# Generate out-of-sample predictions
+				
+    				prediction <- as.numeric(predict(train_model, newdata = data.matrix(escalation_6mo_withlags_test_frame_civil_ns), type="response"))      								
+       				prediction[is.na(escalation_6mo_withlags_test_frame_civil_ns$test_DV_civil_ns)] <- NA
+
+ 
+
+			# Calculate AUC
+			
+    				performance <- prediction(prediction, test_DV_civil_ns)
+              roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+    				    AUC_obs <- data.frame(as.numeric(roc$auc))
+
+
+			# Compile AUCs
+              
+              AUCs_escalation_6mo_withlags <- AUC_obs
+

### 2020-04-28 Replication files/replication-3/code/6mo_run_quad.R

@@ -1,0 +1,218 @@
+
+			# Run random forests models			
+				         
+
+    				train_model <- randomForest(quad_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = quad_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+      			train_model_robust_maxnodes <- randomForest(quad_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = quad_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+        		train_model_robust_sampsize <- randomForest(quad_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 500, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = quad_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+      			train_model_robust_ntree <- randomForest(quad_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = quad_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+        		train_model_robust_traintest1 <- randomForest(quad_6mo_train_formula_civil_ns_robust_traintest1,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = quad_6mo_train_frame_civil_ns_robust_traintest1, forest = TRUE   
[…]

### 2020-04-28 issues.Rmd
some changes

@@
 The original folder name of is `sj-zip-1-jcr-10.1177_0022002720918923/replication-3`.
-For simplicity, we extracted this `replication-3` and saved to the `Replication Files` folder of this repo.
+For simplicity, we extracted this `replication-3` folder and saved to the `Replication Files` folder of this repo.
 
@@
 
-1. Improperly tuning of models
+- Improperly tuning of models
   - RM: In a twitter DM, I asked Blair to clarify -- "[I] was wondering how you all came to your tuning procedure for the RF model hyper-parameters?" He responded "mostly trial and error, honestly.
@@
 Forecasts were for much later data"
-2. Using a regression framework in a classification problem
-3. Their train/validate/test approach
-4. Rounding of AUC/ROC scores
-5. Lack of yearly test forecasts in favor of a single 5-year test forecast
-6. Table 4! 
+- Using a regression framework in a classification problem
+- Their train/test approach 
+  - 5-year forecasts for sep plots and AUC scores)
+- Rounding of AUC/ROC scores
+- Lack of yearly test forecasts in favor of a single 5-year test forecast
+- Table 4! 
   - There are only 20 onsets from 2001 to 2015 But there are 15 "Persistence" cases (top of table 4) and 16 "Change" cases (bottom of table 4)
-    - Why only top 30?
-7. Treating the output of a RF regression model as Pr() -- (RM: I guess this is akin to a Linear Probability Model, but I'm not sure if this is possible in a RF framework
-8. The lack of procedures to account for rare events in an RF model
-9.
+- Treating the output of a RF regression model as Pr() -- (RM: I guess this is akin to a Linear Probability Model, but I'm not sure if this is possible in a RF framework
+- The lack of procedures to account for rare events in an RF model
+
+To-do
+
+1. Full replication
+    - Can we get their results
+    - RM: maybe I can go through their code and bring it "up-to-date" i.e. make it tidy... 
+    - Figure out how they built table 4 (It looks like this is done in STATA)
+    - Check to see if it matter that they are using regression for a classification problem
+2. Address tuning 
+    - Does properly tuning the CAMEO model improve its performance?
+3. Address AUC/ROC smoothing... 
+    - How drastic is the change; does it change their conclusions? 
+4. Run a yearly test forecast -- as is the current standard in TSCS forecasts (Is it the current standard?) 
+5.

### 2020-04-29 Tables/Table 1.Rmd
first cut at narrative of issues with table 1/figure 1
I also copied all of their replication code to replication-3_BM so that we can key a clean version of their code and comment out things that we are not using right now

---
title: "Replication of Figure 1 and Table 1"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

library("kableExtra")
library("gridExtra")
library("tidyverse")
library("yardstick")
library("here")
```

### Introduction^[RM: As a way to maintain transparency and as a poor-mans version of preregistration, I'm pushing each stepto GitHub, where it will have a time-stamp and track all changes.]

One issue that stood out as we read through Blair and Sambanis' (2020) article was how smooth their Receiver Operating Characteristic curves where in Figure 1 on page 14. A Receiver Operating Characteristic curve displays the balance between the true positive rate and the false positive rate, across a range of acceptance thresholds -- the value in which a zero to one probability is assumed to be an "onset" (here: civil conflict). Given that the DV is civil conflict onset (1/0) and that this is a rare event, these curves should resemble a step-like function. Looking through the Blair/Sambanis replication code, we noticed that they we using a rarely used in practice smoothing function when producing their Receiver Operating Characteristic curves.

This is important because Blair and Sambanis use a common performance metric -- Area Under the Receiver Operating Characteristic Curves (AUC-ROC) -- as evidence that their escalation model out preforms all others. However, by smoothing the Receiver Operating Characteristic Curves, they are misrepresenting the true AUC-ROC scores of their various models. Not only does this make model-to-model comparison difficult, it also makes it impossible to assess whether any of their models are performing better than random chance (a AUC-ROC score greater than 0.5) or whether their AUC-ROC scores surpass the (soft) industry standard that suggests that a model is performing well (an AUC-ROC score greater than 0.8).

To assess whether their decision to implement a smoothing function on the Receiver Operating Characteristic Curves for their various models impacts their results (namely, that the escalation model is superior to all others), we replicate Figure 1. However, rather than a two panel figure (one ROC plot their one-month-ahead forecasts and one for their six-months-ahead forecasts), we now include for panels (one panel for the two forecast windows with the smoothed curve (as they produce) and one panel for the two forecast windows that are not smoothed -- the industry standard).

Using these smoothed and non-smoothed curves, we then replicate Table 1 (page 13), which reports the AUC-ROC scores for these four different Receiver Operating Characteristic Curves. If the non-smoothed AUC-ROC scores are drastically different from those of the smoothed AUC-ROC score (a difference of &ge; 0.05) or if the non-smoothed AUC-ROC scores are < 0.8, would suggest that their decision to smooth-out the curves was improper. Further, if the the non-smoothed AUC-ROC score of the Escalation model is smaller or the same as of the non-smoothed AUC-ROC scores other models (Quad, Goldstein, and CAMEO) would suggests that the evidence presented in the paper is not robust to standard AUC-ROC calculation procedures.

Note: we are only reproducing the "Base specification" scores found in Table 1, as these are the what Blair and Sambanis use to support their calims. In order to ensure that this replication is a close as possible to the authors' original work, I will be using the replication code they provided at https://journals.sagepub.com/doi/suppl/10.1177/0022002720918923. There is a lot going on in their code, so I will be commenting out all elements that are not necessary for this replication process. Futher, I will note any changes that I make this parsed code (i.e., removing the smoothing function). The raw replication files can be found at `GitHub/Blair-Sambanis-replication/Replication files/replication-3`, while all my edited code can be found at `GitHub/Blair-Sambanis-replication/Replication files/replication-3_BM`

```{r}
print(Sys.time())
```

### 2020-04-29 Replication files/replication-3_BM/+master.R

@@ -1,0 +1,508 @@
+# Setup ------------------------------------------------------------
+
+         set.seed(92382)
+
+    # Clear environment
+    
+         rm(list = ls())
+    
+    # Set working directory
+    
+         library(here) ## ADDED
+          
+         setwd(here::here("Replication files/replication-3_BM"))
+    
+    # Load packages
+    
+          library(foreign)
+          library(separationplot)
+          library(dyn)
+          library(gdata)
+          library(ROCR)
+          library(readstata13)
+          library(verification)		
+          library(WriteXLS)
+          library(randomForest)
+          library(scales)
+          library(pROC)
+          library(extrafont)
+          library(fontcm)
+          library(corpcor)
+          library(ggplot2)
+          library(fontcm)
+          library(PRROC)
+          library(Bolstad2)
+          library(separationplot)
+
+    # Specify length of ROC curve
+    
+          test_observations <- 10000
+
+# Define 1-month models ------------------------------------------------------------
+
+    # Load data
+
+          data <- read.dta13("data/1mo_data.dta")
+
+    # Define training and testing sets for base specification
+          
+          train_period = mean(data$period[which(data$month==12 & data$year==2007)])
+          end_period = mean(data$period[which(data$month==12 & data$year==2015)])
+          train <- data[data$period<=train_period,]
+          test <- data[data$period>train_period & data$period<=end_period,]
+      
+    #  NOT RUN # We are only running the base specification, for now... (RM: 4/29/2020)
+    # # Define training and testing sets for robustness checks using alternate start dates
+    #       
+    #       train_period_robust_traintest1 = mean(data$period[which(data$month==12 & data$year==2008)])
+    #       end_period = mean(data$period[which(data$month==12 & data$year==2015)])
+    #       train_robust_traintest1 <- data[data$period<=train_period_robust_traintest1,]
+    #       test_robust_traintest1 <- data[data$period>train_period_robust_traintest1 & data$period<=end_period,]
+    #       
+    #       train_period_robust_traintest2 = mean(data$period[which(data$month==12 & data$year==2009)])
[…]

### 2020-04-29 Replication files/replication-3_BM/code/1mo_define_models.R

@@ -1,0 +1,2586 @@
+
+
+  # Define model labels
+		
+    			escalation <- c(
+    				"gov_opp_low_level",
+    				"gov_reb_low_level",
+    				"opp_gov_low_level",
+    				"reb_gov_low_level",
+    				"gov_opp_nonviol_repression",
+    				"gov_reb_nonviol_repression",
+    				"gov_opp_accommodations",
+    				"gov_reb_accommodations",
+    				"reb_gov_demands",
+    				"opp_gov_demands"
+    				)
+		
+    			quad <- c(
+      			"gov_opp_vercf","gov_reb_vercf",
+    				"gov_opp_matcf","gov_reb_matcf",
+    				"opp_gov_vercf","reb_gov_vercf",
+    				"opp_gov_matcf","reb_gov_matcf",
+    				"gov_opp_vercp","gov_reb_vercp",
+    				"gov_opp_matcp","gov_reb_matcp",
+    				"opp_gov_vercp","reb_gov_vercp",
+    				"opp_gov_matcp","reb_gov_matcp"
+    				)
+
+        	goldstein <- c(
+      			"gov_opp_gold","gov_reb_gold",
+    				"opp_gov_gold","reb_gov_gold"
+            )
+
+      		all_CAMEO <- c(
+      			"gov_opp_cameo_010",
+    				"gov_opp_cameo_011",
+    				"gov_opp_cameo_012",
+    				"gov_opp_cameo_013",
+    				"gov_opp_cameo_014",
+    				"gov_opp_cameo_015",
+    				"gov_opp_cameo_016",
+    				"gov_opp_cameo_017",
+    				"gov_opp_cameo_018",
+    				"gov_opp_cameo_019",
+    				"gov_opp_cameo_020",
+    				"gov_opp_cameo_021",
+    				"gov_opp_cameo_0211",
+    				"gov_opp_cameo_0212",
+    				"gov_opp_cameo_0213",
+    				"gov_opp_cameo_0214",
+    				"gov_opp_cameo_022",
+    				"gov_opp_cameo_023",
+    				"gov_opp_cameo_0231",
+    				"gov_opp_cameo_0232",
+    				"gov_opp_cameo_0233",
+    				"gov_opp_cameo_0234",
+    				"gov_opp_cameo_024",
+    				"gov_opp_cameo_0241",
+    				"gov_opp_cameo_0242",
[…]

### 2020-04-29 Replication files/replication-3_BM/code/1mo_make_PR.R

@@ -1,0 +1,30 @@
+
+      loadfonts()
+
+    # Open plot window
+
+      quartz()
+
+      plot.new()
+      
+    # Save PDF
+      
+      pdf("figures/figure2_left.pdf", family = "Times", width = 5.5, height = 5)
+
+    # Plot PR curves
+
+      plot(precision_recall_escalation_inc_civil_ns,lty = 1)
+      plot(precision_recall_quad_inc_civil_ns,lty = 2,add=T)
+      plot(precision_recall_goldstein_inc_civil_ns,lty = 3,add=T)
+      plot(precision_recall_all_CAMEO_inc_civil_ns,lty = 4,add=T)
+      plot(precision_recall_avg_inc_civil_ns,lty = 5,add=T)
+
+    # Add legend
+      
+      legend("topright",
+             c("Escalation", "Quad", "Goldstein", "CAMEO", "Average"),
+             lty = c(1, 2, 3, 4, 5) 
+              )
+
+      dev.off()
+

### 2020-04-29 Replication files/replication-3_BM/code/1mo_run_PITF.R

@@ -1,0 +1,30 @@
+
+			# Run random forests models			
+				         
+    				train_model <- randomForest(PITF_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = PITF_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+
+			# Generate out-of-sample predictions
+				
+    				prediction <- as.numeric(predict(train_model, newdata = data.matrix(PITF_test_frame_civil_ns), type="response"))      								
+       				prediction[is.na(PITF_test_frame_civil_ns$test_DV_civil_ns)] <- NA
+
+ 
+
+			# Calculate AUC
+			
+    				performance <- prediction(prediction, test_DV_civil_ns)
+              roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+    				    AUC_obs <- data.frame(as.numeric(roc$auc))
+
+
+			# Compile AUCs
+              
+              AUCs_PITF <- AUC_obs
+

### 2020-04-29 Replication files/replication-3_BM/code/1mo_run_all_CAMEO.R

@@ -1,0 +1,219 @@
+
+			# Run random forests models			
+				         
+
+    				train_model <- randomForest(all_CAMEO_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = all_CAMEO_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+      			train_model_robust_maxnodes <- randomForest(all_CAMEO_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = all_CAMEO_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+        		train_model_robust_sampsize <- randomForest(all_CAMEO_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 500, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = all_CAMEO_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+      			train_model_robust_ntree <- randomForest(all_CAMEO_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = all_CAMEO_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+        		train_model_robust_traintest1 <- randomForest(all_CAMEO_train_formula_civil_ns_robust_traintest1,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = all_CAMEO_train_frame_civil_ns_robust_traintest1, forest = TRUE   
[…]

### 2020-04-29 Replication files/replication-3_BM/code/1mo_run_avg.R

@@ -1,0 +1,157 @@
+
+      # Average predicted probabilities across models
+
+            prediction <- as.numeric((prediction_escalation_inc_civil_ns + 
+                            prediction_quad_inc_civil_ns + 
+                            prediction_goldstein_inc_civil_ns + 
+                            prediction_all_CAMEO_inc_civil_ns)
+                         /4)
+
+            prediction_robust_maxnodes <- as.numeric((prediction_robust_maxnodes_escalation_inc_civil_ns + 
+                                            prediction_robust_maxnodes_quad_inc_civil_ns + 
+                                            prediction_robust_maxnodes_goldstein_inc_civil_ns + 
+                                            prediction_robust_maxnodes_all_CAMEO_inc_civil_ns)
+                                         /4)
+
+            prediction_robust_sampsize <- as.numeric((prediction_robust_sampsize_escalation_inc_civil_ns + 
+                                            prediction_robust_sampsize_quad_inc_civil_ns + 
+                                            prediction_robust_sampsize_goldstein_inc_civil_ns + 
+                                            prediction_robust_sampsize_all_CAMEO_inc_civil_ns)
+                                         /4)
+
+            prediction_robust_ntree <- as.numeric((prediction_robust_ntree_escalation_inc_civil_ns + 
+                                         prediction_robust_ntree_quad_inc_civil_ns + 
+                                         prediction_robust_ntree_goldstein_inc_civil_ns + 
+                                         prediction_robust_ntree_all_CAMEO_inc_civil_ns)
+                                      /4)
+
+            prediction_robust_traintest1 <- as.numeric((prediction_robust_traintest1_escalation_inc_civil_ns + 
+                                              prediction_robust_traintest1_quad_inc_civil_ns + 
+                                              prediction_robust_traintest1_goldstein_inc_civil_ns + 
+                                              prediction_robust_traintest1_all_CAMEO_inc_civil_ns)
+                                           /4)
+
+            prediction_robust_traintest2 <- as.numeric((prediction_robust_traintest2_escalation_inc_civil_ns + 
+                                              prediction_robust_traintest2_quad_inc_civil_ns + 
[…]

### 2020-04-29 Replication files/replication-3_BM/code/1mo_run_escalation_with_PITF.R

@@ -1,0 +1,30 @@
+
+			# Run random forests models			
+				         
+    				train_model <- randomForest(escalation_with_PITF_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_with_PITF_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+
+			# Generate out-of-sample predictions
+				
+    				prediction <- as.numeric(predict(train_model, newdata = data.matrix(escalation_with_PITF_test_frame_civil_ns), type="response"))      								
+       				prediction[is.na(escalation_with_PITF_test_frame_civil_ns$test_DV_civil_ns)] <- NA
+
+ 
+
+			# Calculate AUC
+			
+    				performance <- prediction(prediction, test_DV_civil_ns)
+              roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+    				    AUC_obs <- data.frame(as.numeric(roc$auc))
+
+
+			# Compile AUCs
+              
+              AUCs_escalation_with_PITF <- AUC_obs
+

### 2020-04-29 Replication files/replication-3_BM/code/1mo_run_quad.R

@@ -1,0 +1,219 @@
+
+			# Run random forests models			
+				         
+
+    				train_model <- randomForest(quad_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = quad_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+      			train_model_robust_maxnodes <- randomForest(quad_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = quad_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+        		train_model_robust_sampsize <- randomForest(quad_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 500, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = quad_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+      			train_model_robust_ntree <- randomForest(quad_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = quad_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+        		train_model_robust_traintest1 <- randomForest(quad_train_formula_civil_ns_robust_traintest1,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = quad_train_frame_civil_ns_robust_traintest1, forest = TRUE   
+    												  )			         
+
[…]

### 2020-04-29 Replication files/replication-3_BM/code/6mo_define_models.R

@@ -1,0 +1,2581 @@
+
+
+  # Define model labels
+		
+    			escalation_6mo <- c(
+    				"gov_opp_low_level",
+    				"gov_reb_low_level",
+    				"opp_gov_low_level",
+    				"reb_gov_low_level",
+    				"gov_opp_nonviol_repression",
+    				"gov_reb_nonviol_repression",
+    				"gov_opp_accommodations",
+    				"gov_reb_accommodations",
+    				"reb_gov_demands",
+    				"opp_gov_demands"
+    				)
+		
+    			quad_6mo <- c(
+      			"gov_opp_vercf","gov_reb_vercf",
+    				"gov_opp_matcf","gov_reb_matcf",
+    				"opp_gov_vercf","reb_gov_vercf",
+    				"opp_gov_matcf","reb_gov_matcf",
+    				"gov_opp_vercp","gov_reb_vercp",
+    				"gov_opp_matcp","gov_reb_matcp",
+    				"opp_gov_vercp","reb_gov_vercp",
+    				"opp_gov_matcp","reb_gov_matcp"
+    				)
+
+        	goldstein_6mo <- c(
+      			"gov_opp_gold","gov_reb_gold",
+    				"opp_gov_gold","reb_gov_gold"
+            )
+
+      		all_CAMEO_6mo <- c(
+      			"gov_opp_cameo_010",
+    				"gov_opp_cameo_011",
+    				"gov_opp_cameo_012",
+    				"gov_opp_cameo_013",
+    				"gov_opp_cameo_014",
+    				"gov_opp_cameo_015",
+    				"gov_opp_cameo_016",
+    				"gov_opp_cameo_017",
+    				"gov_opp_cameo_018",
+    				"gov_opp_cameo_019",
+    				"gov_opp_cameo_020",
+    				"gov_opp_cameo_021",
+    				"gov_opp_cameo_0211",
+    				"gov_opp_cameo_0212",
+    				"gov_opp_cameo_0213",
+    				"gov_opp_cameo_0214",
+    				"gov_opp_cameo_022",
+    				"gov_opp_cameo_023",
+    				"gov_opp_cameo_0231",
+    				"gov_opp_cameo_0232",
+    				"gov_opp_cameo_0233",
+    				"gov_opp_cameo_0234",
+    				"gov_opp_cameo_024",
+    				"gov_opp_cameo_0241",
[…]

### 2020-04-29 Replication files/replication-3_BM/code/6mo_make_ROC.R

@@ -1,0 +1,49 @@
+
+      loadfonts()
+
+    # Open plot window
+
+      quartz()
+
+      plot.new()
+
+    # Save PDF
+      
+  		pdf("figures/figure1_right.pdf", family = "CM Roman", width = 5.5, height = 5)
+      
+    # Plot ROC curves   
+      
+      plot(roc_escalation_6mo_inc_civil_ns,
+		    ylab = "True positive rate",
+		    xlab = "False positive rate",
+		    #col = c("brown"),
+		    lty = 1
+        )
+      
+      lines.roc(roc_quad_6mo_inc_civil_ns,
+        lty = 2
+        )
+    
+      lines.roc(roc_goldstein_6mo_inc_civil_ns,
+        lty = 3
+        )
+
+      lines.roc(roc_all_CAMEO_6mo_inc_civil_ns,
+        lty = 4
+        )
+      
+      lines.roc(roc_avg_6mo_inc_civil_ns,
+        lty = 5
+        )
+
+    # Add legend
+      
+  		legend("bottomright",
+  			c("Escalation", "Quad", "Goldstein", "CAMEO", "Average"),
+  			  lty = c(1, 2, 3, 4, 5) 
+  		      )
+
+      dev.off()
+      
+      
+      

### 2020-04-29 Replication files/replication-3_BM/code/6mo_run_PITF.R

@@ -1,0 +1,30 @@
+
+			# Run random forests models			
+				         
+    				train_model <- randomForest(PITF_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = PITF_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+
+			# Generate out-of-sample predictions
+				
+    				prediction <- as.numeric(predict(train_model, newdata = data.matrix(PITF_6mo_test_frame_civil_ns), type="response"))      								
+       				prediction[is.na(PITF_6mo_test_frame_civil_ns$test_DV_civil_ns)] <- NA
+
+ 
+
+			# Calculate AUC
+			
+    				performance <- prediction(prediction, test_DV_civil_ns)
+              roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+    				    AUC_obs <- data.frame(as.numeric(roc$auc))
+
+
+			# Compile AUCs
+              
+              AUCs_PITF_6mo <- AUC_obs
+

### 2020-04-29 Replication files/replication-3_BM/code/6mo_run_all_CAMEO.R

@@ -1,0 +1,218 @@
+
+			# Run random forests models			
+				         
+
+    				train_model <- randomForest(all_CAMEO_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = all_CAMEO_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+      			train_model_robust_maxnodes <- randomForest(all_CAMEO_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = all_CAMEO_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+        		train_model_robust_sampsize <- randomForest(all_CAMEO_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 500, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = all_CAMEO_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+      			train_model_robust_ntree <- randomForest(all_CAMEO_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = all_CAMEO_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+        		train_model_robust_traintest1 <- randomForest(all_CAMEO_6mo_train_formula_civil_ns_robust_traintest1,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
[…]

### 2020-04-29 Replication files/replication-3_BM/code/6mo_run_escalation.R

@@ -1,0 +1,218 @@
+
+			# Run random forests models			
+				         
+
+    				train_model <- randomForest(escalation_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+      			train_model_robust_maxnodes <- randomForest(escalation_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+        		train_model_robust_sampsize <- randomForest(escalation_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 500, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+      			train_model_robust_ntree <- randomForest(escalation_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+        		train_model_robust_traintest1 <- randomForest(escalation_6mo_train_formula_civil_ns_robust_traintest1,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
[…]

### 2020-04-29 Replication files/replication-3_BM/code/6mo_run_escalation_OOS.R

@@ -1,0 +1,37 @@
+##############################
+# RHS = escalation_6mo 
+# DV = incidence_civil_ns_plus1
+# model = randomForest
+##############################
+
+			# Run random forests models			
+				         
+      				train_model <- randomForest(escalation_6mo_train_formula_civil_ns,
+      												  type = regression, importance = TRUE, 
+      												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+      												  sampsize = 100, 
+      												  replace = FALSE, do.trace = FALSE , 
+      												  data = escalation_6mo_train_frame_civil_ns, forest = TRUE   
+      												  )			         
+    				
+			# Generate out-of-sample predictions
+				
+      				prediction <- as.numeric(predict(train_model, newdata = data.matrix(escalation_6mo_test_frame_civil_ns), type="response"))      								
+
+      # Export predicted probabilities
+
+              country <- as.vector(test$country_iso3)
+              year <- as.vector(test$year)
+              period <- as.vector(test$period)
+              incidence_civil_ns <- as.vector(test$incidence_civil_ns)
+              incidence_civil_ns_plus1 <- as.vector(test$incidence_civil_ns_plus1)
+      
+              pred_escalation_6mo_inc_civil_ns <- data.frame(cbind(country, year, period,
+                                                      incidence_civil_ns, incidence_civil_ns_plus1,
+                                                      prediction))
+    
+              colnames(pred_escalation_6mo_inc_civil_ns) <- c("country", "year", "period",
+                                                         "incidence_civil_ns", "incidence_civil_ns_plus1",
+                                                        "prediction")
+    
+              

### 2020-04-29 Replication files/replication-3_BM/code/6mo_run_escalation_split_PITF.R

@@ -1,0 +1,44 @@
+  		
+        	# Run random forests model for high-risk split	
+				         
+        			train_model_highriskPITF <- randomForest(escalation_6mo_highriskPITF_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 10000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_6mo_highriskPITF_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+      		# Run random forests model for low-risk split
+
+        			train_model_lowriskPITF <- randomForest(escalation_6mo_lowriskPITF_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 10000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = escalation_6mo_lowriskPITF_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+      		# Combine random forests models for high-risk and low-risk splits
+
+           		 	train_model <- combine(train_model_highriskPITF, train_model_lowriskPITF)
+    
+
+		    	# Generate out-of-sample predictions
+				
+    				    prediction <- as.numeric(predict(train_model, newdata = data.matrix(escalation_6mo_splitPITF_test_frame_civil_ns), type="response"))      								
+       			    	prediction[is.na(escalation_6mo_splitPITF_test_frame_civil_ns$test_DV_civil_ns)] <- NA
+
+ 
+
+			# Calculate AUC
+			
+    				performance <- prediction(prediction, test_DV_civil_ns)
+              			roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+    				    	AUC_obs <- data.frame(as.numeric(roc$auc))
+
+
+			# Compile AUCs
+              
+             		AUCs_escalation_6mo_split_PITF <- AUC_obs
+

### 2020-04-29 Replication files/replication-3_BM/code/6mo_run_escalation_weighted_PITF.R

@@ -1,0 +1,22 @@
+
+      # Weight predicted probabilities from escalation_6mo model
+
+            weight <- as.numeric(train$pred_prob_plus1)
+            length(weight) <- length(prediction_escalation_6mo_inc_civil_ns)
+            
+            prediction = prediction_escalation_6mo_inc_civil_ns*weight
+            prediction[is.na(escalation_6mo_test_frame_civil_ns$test_DV_civil_ns)] <- NA
+
+
+			# Calculate AUC
+			
+    				performance <- prediction(prediction, test_DV_civil_ns)
+              roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+    				    AUC_obs <- data.frame(as.numeric(roc$auc))
+
+
+			# Compile AUCs
+              
+              AUCs_escalation_6mo_weighted_PITF <- AUC_obs
+
+

### 2020-04-29 Replication files/replication-3_BM/code/6mo_run_goldstein.R

@@ -1,0 +1,218 @@
+
+			# Run random forests models			
+				         
+
+    				train_model <- randomForest(goldstein_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = goldstein_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+      			train_model_robust_maxnodes <- randomForest(goldstein_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = goldstein_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+    
+        		train_model_robust_sampsize <- randomForest(goldstein_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 500, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = goldstein_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+      			train_model_robust_ntree <- randomForest(goldstein_6mo_train_formula_civil_ns,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
+    												  data = goldstein_6mo_train_frame_civil_ns, forest = TRUE   
+    												  )			         
+
+        		train_model_robust_traintest1 <- randomForest(goldstein_6mo_train_formula_civil_ns_robust_traintest1,
+    												  type = regression, importance = TRUE, 
+    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
+    												  sampsize = 100, 
+    												  replace = FALSE, do.trace = FALSE , 
[…]

### 2020-04-29 issues.Rmd
update issues doc

@@
 
+1. Using AUC for smoothed ROC curves. Does the interpretation of Table 1, and superiority of the escalation model, depend on the choice to use smoothed ROC curves?
+
+2.
+In Table 4, are the forecasts evaluated against incidence or onset?
+The data from 2001 to 2015 have 20 civil war onsets in total, yet Table 4 appears to show 15 and 16 positive events for 2016 only.
+If this is indeed incidence, but the original models and forecasts were for onset, then this is a incorrect assessment of the forecasts' accuracy.
+
+- Table 4! 
+  - There are only 20 onsets from 2001 to 2015 But there are 15 "Persistence" cases (top of table 4) and 16 "Change" cases (bottom of table 4)
+
+3.
+Unorthodox RF hyperparameter choices that may be driving the OOS test prediction results.
+There are two issues here.
+One is that the RF models are so unorthodox.
+I'm not sure that this in itself is a problem.
+But the second potential problem is that the hyperparameter values may not have been tuned in a way that works the best for the non-escalation model, and especially the CAMEO model with 1,100+ features.
+
 - Improperly tuning of models
@@
 - Using a regression framework in a classification problem
+- Rounding of AUC/ROC scores
+    + AB: smoothing or rounding? I think rounding is ok. 
+- Treating the output of a RF regression model as Pr() -- (RM: I guess this is akin to a Linear Probability Model, but I'm not sure if this is possible in a RF framework
+    + It's hacky but it works. The resulting forecasts are within the 0-1 interval.
+
+The table below shows the randomForest package default hyperparameter values for a binary classification problem like the one at hand and compares them to the B&S base settings.
+Some hyperparameters have heuristics that determine the default value based on characteristics of the input data; we note these in the second column and what the realized default setting would be for the basic Escalation model (first row in Table 1 in the paper) in the 3rd column.
+The last column has the settings B&S use.
+
+|Hyperparameter|Default heuristic|Default values (Escalation)|B&S value|
+|--------------|-----------------|--------------------------|---------|
+|type          |                 | classification           | regression|
+|ntree         |                 | 500                      | 100,000 |
+|mtry          |`floor(sqrt(ncol(x)))`|3|3|
+|replace       |                 | true | false |
+|sampsize | `nrow(x)` if replace, else `ceiling(.632*nrow(x))` | 11,869 | 100 |
+|nodesize | 1 for classification | 1 | 1 |
+|maxnodes | | null | 5 |
+
+It is worth noting that commonly the ntree, mtry, and nodesize parameters are the main parameters tuned; of these three in the B&S specification only 1, the ntree parameter, deviates from the default settings.
+
+In any case, there is a stark contrast in the default RF settings and the way B&S use the RF models.
+The default approach is to train a relatively small number (ntree; 500) of classification trees, but where each tree is fairly big in that it is trained on data that has the same number of rows as the training data, albeit sampled with replacement (replace is true; sampsize is 11,869), and allowed to grow fairly deep (this is governed by nodesize (1), which is the minimum size a terminal node must have).
+In contrast, B&S grow very extensive forests with a large number of trees (100,000 compared to 500), but each tree is very small and shallow; only a 100 rows are sampled from the training data for each tree, and the trees are constrained to at most 5 terminal nodes (maxnodes).
+
+This approach only works due to the other unorthodox choice, which is to use regression, not classificaton, trees.
+Trying to use classification trees with the other parameter settings in fact does not work at all because it is almost guaranteed that a sample of 100 from the 11,869 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample.
+As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:
+
+```
+Warning message:
+In randomForest.default(y = as.integer(train_df$incidence_civil_ns_plus1 ==  :
+  The response has five or fewer unique values.  Are you sure you want to do regression?
+```
+
+While unorthodox, the approach does work.
+It produces predictions that are within the 0 to 1 interval.
+The concern is that the settings work well only for the escalation model, and only for the particular test set at hand.
+Specifically we wonder if the CAMEO model with 1,100+ features, compared to 10 or less for the other specifications, would perform better with a more explicit tuning procedure.
+
+The table below shows cross-validated training and OOS test AUC-ROC for the original 1 month escalation model and a modified version which uses more conventional random forest parameter settings (classification, 5000 trees, and default settings otherwise).
+The first and second columns are average and SD AUC-ROC obtained through repeated cross-validation on the training data set.
+Because there are only 9 positive cases in the whole training data, and thus to ensure that any given split will include at least 1 positive case in each data partition, we use 2-fold CV, i.e.
+splitting the original training data into equally-sized new training and validation sets.
+This is repeated 21 times for a total of 42 OOS performance samples for each model.
+The last column shows the OOS performance on the original test set.
+The first value in this column corresponds to the base escalation model results reported Table 1 in the B&S paper when *not* using smoothed ROC curves.
+
+```{r}
+tbl <- structure(list(Model = c("Escalation, 1mo", "Modified Escalation, 1mo"
+), Avg_CV_ROC_AUC = c(0.698772142673124, 0.643982732304452), 
+    SD_CV_ROC_AUC = c(0.136834250535024, 0.104102374345325), 
+    Test_ROC_AUC = c(0.786872878159185, 0.583391239320482)), row.names = c(NA, 
+-2L), class = c("tbl_df", "tbl", "data.frame"))
+
+knitr::kable(tbl, digits = 2)
+```
+
+Two thigs stand out:
+
+- The escalation model has better performance in the test data than in the training CV samples. This is consistent with hyperparameter values that were (over-)fit to the test data. 
+- Although the original escalation model outperforms the modified model in the training CV resamples (p < 0.05), the difference in means is much less pronounced than in the test data.
+
+4. Design choices
+
+Rick, I think the points below are more subjective choices.
+The first two points above are IMO potentially objectively incorrect technical errors that could undermine the original B&S findings.
+The 3rd point about the RF models is somewhere between objective error and subjective choice.
+I'm not sure.
+But the points below I would say are subjective choices that could alter their findings and that I think are justifiable criticisms, but I think it'll be easier for someone to push back on this if they wanted to.
+
 - Their train/test approach 
   - 5-year forecasts for sep plots and AUC scores)
-- Rounding of AUC/ROC scores
 - Lack of yearly test forecasts in favor of a single 5-year test forecast
-- Table 4! 
-  - There are only 20 onsets from 2001 to 2015 But there are 15 "Persistence" cases (top of table 4) and 16 "Change" cases (bottom of table 4)
-- Treating the output of a RF regression model as Pr() -- (RM: I guess this is akin to a Linear Probability Model, but I'm not sure if this is possible in a RF framework
 - The lack of procedures to account for rare events in an RF model
@@
     - RM: maybe I can go through their code and bring it "up-to-date" i.e. make it tidy... 
+        + AB: that sounds like a lot of work
     - Figure out how they built table 4 (It looks like this is done in STATA)
     - Check to see if it matter that they are using regression for a classification problem
+        + AB: i tried switching the models as they have them just from regression to classification and it breaks because the data samples end up being all 0's. 
 2. Address tuning 

### 2020-04-29 tuning-experiments/scratch.R

@@ -49,7 +49,7 @@ table(complete.cases(train_df[, escalation]))
 table(complete.cases(train_df[, all_cameo]))
 
 set.seed(5234)
-folds <- vfold_cv(train_df, v = 2, repeats = 7*1)
+folds <- vfold_cv(train_df, v = 2, repeats = 7*3)
 map_dbl(folds$splits, function(x) {dat = testing(x); sum(dat[[dv]]=="1")})
 
 res <- foreach(i=1:nrow(folds)) %dopar% {
@@ -97,44 +97,44 @@ res <- foreach(i=1:nrow(folds)) %dopar% {
            AUC = roc_auc(test_preds_2, truth, preds)[[".estimate"]]
     )))
 
-  # CAMEO models
-  mdl_3 <- suppressWarnings({
-    randomForest(y = as.integer(train_i$incidence_civil_ns_plus1=="1"),
-                 x = train_i[, all_cameo],
-                 type = "regression",
-                 ntree = 100000,
-                 maxnodes = 5,
-                 sampsize = 100,
-                 replace = FALSE,
-                 do.trace = FALSE)
-  })
-
-  test_preds_3 <- tibble(
-    preds = as.vector(predict(mdl_3, newdata = test_i[, all_cameo],
-                              type = "response")),
-    truth = test_i[, dv])
-  res_i <- c(res_i, list(
-    tibble(i = i,
-           model = "3",
-           AUC = roc_auc(test_preds_3, truth, preds)[[".estimate"]]
-    )))
-
-  mdl_4 <- randomForest(y = train_i$incidence_civil_ns_plus1,
-                                   x = train_i[, all_cameo],
-                                   type = "classification",
-                                   ntree = 5000,
-                                   mtry  = 34,
-                                   replace = TRUE,
-                                   do.trace = FALSE)
-
-  test_preds_4 <- tibble(
-    preds = as.vector(predict(mdl_4, newdata = test_i[, all_cameo], type = "prob")[, "1"]),
-    truth = test_i[, dv])
-  res_i <- c(res_i, list(
-    tibble(i = i,
-           model = "4",
-           AUC = roc_auc(test_preds_4, truth, preds)[[".estimate"]]
-    )))
+  # # CAMEO models
+  # mdl_3 <- suppressWarnings({
+  #   randomForest(y = as.integer(train_i$incidence_civil_ns_plus1=="1"),
[…]

### 2020-04-29 Tables/Table 1.Rmd
removed r time stamp
Created new figure 1 and table 1
shows smoothed and not smoothed ROC plots and AUC-ROC scores for the base specification

@@
 
-### Introduction^[RM: As a way to maintain transparency and as a poor-mans version of preregistration, I'm pushing each stepto GitHub, where it will have a time-stamp and track all changes.]
+### Introduction^[RM: As a way to maintain transparency and as a poor-mans version of preregistration, I'm pushing each step to GitHub, where it will have a time-stamp and track all changes.]
 
 One issue that stood out as we read through Blair and Sambanis' (2020) article was how smooth their Receiver Operating Characteristic curves where in Figure 1 on page 14.
-A Receiver Operating Characteristic curve displays the balance between the true positive rate and the false positive rate, across a range of acceptance thresholds -- the value in which a zero to one probability is assumed to be an "onset" (here: civil conflict).
+A Receiver Operating Characteristic curve displays the balance between the true positive rate and the false positive rate, across a range of discrimination thresholds -- the value in which a zero to one probability is assumed to be an "onset" (here: civil conflict).
 Given that the DV is civil conflict onset (1/0) and that this is a rare event, these curves should resemble a step-like function.
@@
 
-Note: we are only reproducing the "Base specification" scores found in Table 1, as these are the what Blair and Sambanis use to support their calims.
+Note: we are only reproducing the "Base specification" scores found in Table 1, as these are the what Blair and Sambanis use to support their claims.
 In order to ensure that this replication is a close as possible to the authors' original work, I will be using the replication code they provided at https://journals.sagepub.com/doi/suppl/10.1177/0022002720918923.
 There is a lot going on in their code, so I will be commenting out all elements that are not necessary for this replication process.
-Futher, I will note any changes that I make this parsed code (i.e., removing the smoothing function).
+Further, I will note any changes that I make this parsed code (i.e., removing the smoothing function).
 The raw replication files can be found at `GitHub/Blair-Sambanis-replication/Replication files/replication-3`, while all my edited code can be found at `GitHub/Blair-Sambanis-replication/Replication files/replication-3_BM`
-
-```{r}
-print(Sys.time())
-```

### 2020-04-29 issues.Rmd
Created new figure 1 and table 1
shows smoothed and not smoothed ROC plots and AUC-ROC scores for the base specification

--- 2020-04-28 issues.Rmd
@@
 For simplicity, we extracted this `replication-3` folder and saved to the `Replication Files` folder of this repo.
+To ensure that we have a clean copy of the original replication scripts, I copied all files to `replication-3_BM` I'm (RM) currently using the sripts in this folder to recreate figure 1 and table 1 with and without smoothing -- see below
 
@@
   - 5-year forecasts for sep plots and AUC scores)
-- Rounding of AUC/ROC scores
+- Smoothing of AUC/ROC scores
 - Lack of yearly test forecasts in favor of a single 5-year test forecast

### 2020-04-29 Replication files/replication-3_BM/code/1mo_define_models.R

@@ -1203,7 +1203,7 @@
     
       		# We are only running the base specification, for now... (RM: 4/29/2020)
       		
-    for(model in c("escalation", "goldstein", "all_CAMEO")){
+    for(model in c("escalation", "quad", "goldstein", "all_CAMEO")){
       
     if(model == "escalation"){
       variables <- c("gov_opp_low_level",

### 2020-04-29 Replication files/replication-3_BM/code/1mo_make_ROC.R

@@ -1,15 +1,19 @@
 
-      loadfonts()
+      # loadfonts()
 
     # Open plot window
 
-      quartz()
+      # quartz()
 
-      plot.new()
-
-    # Save PDF
+## RM: ADDED 4/29/2020       
+      # plot.new()
       
-  		pdf("figures/figure1_left.pdf", family = "CM Roman", width = 5.5, height = 5)
+      # Save PDF
+## RM: ADDED 4/29/2020   
+      png("figures/figure1_top.png", family = "Times", width = 11, height = 5, units = "in", res = 1050)
+      
+      par(mfrow = c(1, 2))
+      # pdf("figures/figure1_left.pdf", family = "CM Roman", width = 5.5, height = 5)
       
     # Plot ROC curves   
       
@@ -17,6 +21,7 @@
 		    ylab = "True positive rate",
 		    xlab = "False positive rate",
 		    #col = c("brown"),
+		    main = "ROC - Smoothed (1-mon)",
 		    lty = 1
         )
       
@@ -42,6 +47,31 @@
   			c("Escalation", "Quad", "Goldstein", "CAMEO", "Average"),
   			  lty = c(1, 2, 3, 4, 5) 
   		      )
+  		
+  		plot(roc_noSmooth_escalation_inc_civil_ns,
+  		     ylab = "True positive rate",
+  		     xlab = "False positive rate",
+  		     #col = c("brown"),
+  		     main = "ROC - Not Smoothed (1-mon)",
+  		     lty = 1
+  		)
+  		
+  		lines.roc(roc_noSmooth_quad_inc_civil_ns,
+  		          lty = 2
+  		)
+  		
+  		lines.roc(roc_noSmooth_goldstein_inc_civil_ns,
+  		          lty = 3
+  		)
+  		
+  		lines.roc(roc_noSmooth_all_CAMEO_inc_civil_ns,
+  		          lty = 4
+  		)
+  		
+  		lines.roc(roc_noSmooth_avg_inc_civil_ns,
+  		          lty = 5
+  		)
+  		
 
       dev.off()
       

### 2020-04-29 Replication files/replication-3_BM/code/1mo_run_all_CAMEO.R

@@ -10,69 +10,69 @@
     												  data = all_CAMEO_train_frame_civil_ns, forest = TRUE   
     												  )			         
     
-      			train_model_robust_maxnodes <- randomForest(all_CAMEO_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = all_CAMEO_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-    
-        		train_model_robust_sampsize <- randomForest(all_CAMEO_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 500, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = all_CAMEO_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-      			train_model_robust_ntree <- randomForest(all_CAMEO_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = all_CAMEO_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest1 <- randomForest(all_CAMEO_train_formula_civil_ns_robust_traintest1,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = all_CAMEO_train_frame_civil_ns_robust_traintest1, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest2 <- randomForest(all_CAMEO_train_formula_civil_ns_robust_traintest2,
-        		                  type = regression, importance = TRUE, 
-        		                  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-        		                  sampsize = 100, 
-        		                  replace = FALSE, do.trace = FALSE , 
[…]

### 2020-04-29 Replication files/replication-3_BM/code/1mo_run_avg.R

@@ -7,119 +7,123 @@
                             prediction_all_CAMEO_inc_civil_ns)
                          /4)
 
-            prediction_robust_maxnodes <- as.numeric((prediction_robust_maxnodes_escalation_inc_civil_ns + 
-                                            prediction_robust_maxnodes_quad_inc_civil_ns + 
-                                            prediction_robust_maxnodes_goldstein_inc_civil_ns + 
-                                            prediction_robust_maxnodes_all_CAMEO_inc_civil_ns)
-                                         /4)
-
-            prediction_robust_sampsize <- as.numeric((prediction_robust_sampsize_escalation_inc_civil_ns + 
-                                            prediction_robust_sampsize_quad_inc_civil_ns + 
-                                            prediction_robust_sampsize_goldstein_inc_civil_ns + 
-                                            prediction_robust_sampsize_all_CAMEO_inc_civil_ns)
-                                         /4)
-
-            prediction_robust_ntree <- as.numeric((prediction_robust_ntree_escalation_inc_civil_ns + 
-                                         prediction_robust_ntree_quad_inc_civil_ns + 
-                                         prediction_robust_ntree_goldstein_inc_civil_ns + 
-                                         prediction_robust_ntree_all_CAMEO_inc_civil_ns)
-                                      /4)
-
-            prediction_robust_traintest1 <- as.numeric((prediction_robust_traintest1_escalation_inc_civil_ns + 
-                                              prediction_robust_traintest1_quad_inc_civil_ns + 
-                                              prediction_robust_traintest1_goldstein_inc_civil_ns + 
-                                              prediction_robust_traintest1_all_CAMEO_inc_civil_ns)
-                                           /4)
-
-            prediction_robust_traintest2 <- as.numeric((prediction_robust_traintest2_escalation_inc_civil_ns + 
-                                              prediction_robust_traintest2_quad_inc_civil_ns + 
-                                              prediction_robust_traintest2_goldstein_inc_civil_ns + 
-                                              prediction_robust_traintest2_all_CAMEO_inc_civil_ns)
-                                           /4)
-
[…]

### 2020-04-29 Replication files/replication-3_BM/code/1mo_run_escalation.R

@@ -10,69 +10,69 @@
     												  data = escalation_train_frame_civil_ns, forest = TRUE   
     												  )			         
     
-      			train_model_robust_maxnodes <- randomForest(escalation_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = escalation_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-    
-        		train_model_robust_sampsize <- randomForest(escalation_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 500, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = escalation_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-      			train_model_robust_ntree <- randomForest(escalation_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = escalation_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest1 <- randomForest(escalation_train_formula_civil_ns_robust_traintest1,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = escalation_train_frame_civil_ns_robust_traintest1, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest2 <- randomForest(escalation_train_formula_civil_ns_robust_traintest2,
-        		                  type = regression, importance = TRUE, 
-        		                  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-        		                  sampsize = 100, 
-        		                  replace = FALSE, do.trace = FALSE , 
-        		                  data = escalation_train_frame_civil_ns_robust_traintest2, forest = TRUE   
[…]

### 2020-04-29 Replication files/replication-3_BM/code/1mo_run_goldstein.R

@@ -10,69 +10,69 @@
     												  data = goldstein_train_frame_civil_ns, forest = TRUE   
     												  )			         
     
-      			train_model_robust_maxnodes <- randomForest(goldstein_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = goldstein_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-    
-        		train_model_robust_sampsize <- randomForest(goldstein_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 500, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = goldstein_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-      			train_model_robust_ntree <- randomForest(goldstein_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = goldstein_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest1 <- randomForest(goldstein_train_formula_civil_ns_robust_traintest1,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = goldstein_train_frame_civil_ns_robust_traintest1, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest2 <- randomForest(goldstein_train_formula_civil_ns_robust_traintest2,
-        		                  type = regression, importance = TRUE, 
-        		                  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-        		                  sampsize = 100, 
-        		                  replace = FALSE, do.trace = FALSE , 
[…]

### 2020-04-29 Replication files/replication-3_BM/code/1mo_run_quad.R

@@ -10,69 +10,69 @@
     												  data = quad_train_frame_civil_ns, forest = TRUE   
     												  )			         
     
-      			train_model_robust_maxnodes <- randomForest(quad_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = quad_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-    
-        		train_model_robust_sampsize <- randomForest(quad_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 500, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = quad_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-      			train_model_robust_ntree <- randomForest(quad_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = quad_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest1 <- randomForest(quad_train_formula_civil_ns_robust_traintest1,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = quad_train_frame_civil_ns_robust_traintest1, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest2 <- randomForest(quad_train_formula_civil_ns_robust_traintest2,
-        		                  type = regression, importance = TRUE, 
-        		                  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-        		                  sampsize = 100, 
-        		                  replace = FALSE, do.trace = FALSE , 
-        		                  data = quad_train_frame_civil_ns_robust_traintest2, forest = TRUE   
[…]

### 2020-04-29 Replication files/replication-3_BM/code/6mo_define_models.R

@@ -1199,8 +1199,12 @@
   # Define model features
 
 
-  for(model in c("escalation_6mo", "escalation_6mo_with_PITF", "PITF_6mo", "escalation_6mo_splitPITF", "escalation_6mo_highriskPITF", "escalation_6mo_lowriskPITF", "escalation_6mo_withlags", "escalation_6mo_simple", "quad_6mo", "goldstein_6mo", "all_CAMEO_6mo")){
-    
+  # for(model in c("escalation_6mo", "escalation_6mo_with_PITF", "PITF_6mo", "escalation_6mo_splitPITF", "escalation_6mo_highriskPITF", "escalation_6mo_lowriskPITF", "escalation_6mo_withlags", "escalation_6mo_simple", "quad_6mo", "goldstein_6mo", "all_CAMEO_6mo")){
+  #   
+      		
+# We are only running the base specification, for now... (RM: 4/29/2020)
+    for(model in c("escalation_6mo", "quad_6mo", "goldstein_6mo", "all_CAMEO_6mo")){
+      
     if(model == "escalation_6mo"){
       variables <- c("gov_opp_low_level",
                      "gov_reb_low_level",
@@ -1214,113 +1218,113 @@
                      "opp_gov_demands")
     } 
  
-    if(model == "escalation_6mo_with_PITF"){
-      variables <- c("gov_opp_low_level",
-                     "gov_reb_low_level",
-                     "opp_gov_low_level",
-                     "reb_gov_low_level",
-                     "gov_opp_nonviol_repression",
-                     "gov_reb_nonviol_repression",
-                     "gov_opp_accommodations",
-                     "gov_reb_accommodations",
-                     "reb_gov_demands",
-                     "opp_gov_demands",
-                     "region_africa",
-                     "region_easia",
-                     "region_europe",
-                     "region_latam",
-                     "region_neast",
-                     "disp4cat_minus1",
-                     "faction_minus1",
-                     "bnkv100cat_minus1",
-                     "lncimr_minus1",
-                     "lntpopx_minus1")
-    }
-      
-      if(model == "PITF_6mo"){
-        variables <- c("region_africa",
-                       "region_easia",
-                       "region_europe",
-                       "region_latam",
-                       "region_neast",
-                       "disp4cat_minus1",
-                       "faction_minus1",
-                       "bnkv100cat_minus1",
-                       "lncimr_minus1",
-                       "lntpopx_minus1")
-      }
- 
-    if(model == "escalation_6mo_highriskPITF"){
[…]

### 2020-04-29 Replication files/replication-3_BM/code/6mo_make_ROC.R

@@ -1,49 +1,77 @@
 
-      loadfonts()
+# loadfonts()
 
-    # Open plot window
+# Open plot window
 
-      quartz()
+# quartz()
 
-      plot.new()
+## RM: ADDED 4/29/2020       
+# plot.new()
 
-    # Save PDF
+# Save PDF
+## RM: ADDED 4/29/2020   
+      png("figures/figure1_bottom.png", family = "times", width = 11, height = 5, units = "in", res = 1050)
       
-  		pdf("figures/figure1_right.pdf", family = "CM Roman", width = 5.5, height = 5)
+      par(mfrow = c(1, 2))
+      # pdf("figures/figure1_left.pdf", family = "CM Roman", width = 5.5, height = 5)
       
-    # Plot ROC curves   
+      # Plot ROC curves   
       
       plot(roc_escalation_6mo_inc_civil_ns,
-		    ylab = "True positive rate",
-		    xlab = "False positive rate",
-		    #col = c("brown"),
-		    lty = 1
-        )
+           ylab = "True positive rate",
+           xlab = "False positive rate",
+           #col = c("brown"),
+           main = "ROC - Smoothed (6-mons)",
+           lty = 1
+      )
       
       lines.roc(roc_quad_6mo_inc_civil_ns,
-        lty = 2
-        )
-    
+                lty = 2
+      )
+      
       lines.roc(roc_goldstein_6mo_inc_civil_ns,
-        lty = 3
-        )
-
+                lty = 3
+      )
+      
       lines.roc(roc_all_CAMEO_6mo_inc_civil_ns,
-        lty = 4
-        )
+                lty = 4
+      )
       
       lines.roc(roc_avg_6mo_inc_civil_ns,
-        lty = 5
-        )
-
-    # Add legend
+                lty = 5
+      )
       
-  		legend("bottomright",
-  			c("Escalation", "Quad", "Goldstein", "CAMEO", "Average"),
-  			  lty = c(1, 2, 3, 4, 5) 
-  		      )
-
+      # Add legend
+      
+      legend("bottomright",
+             c("Escalation", "Quad", "Goldstein", "CAMEO", "Average"),
+             lty = c(1, 2, 3, 4, 5) 
+      )
+      
+      plot(roc_noSmooth_escalation_6mo_inc_civil_ns,
+           ylab = "True positive rate",
+           xlab = "False positive rate",
+           #col = c("brown"),
[…]

### 2020-04-29 Replication files/replication-3_BM/code/6mo_run_all_CAMEO.R

@@ -10,69 +10,69 @@
     												  data = all_CAMEO_6mo_train_frame_civil_ns, forest = TRUE   
     												  )			         
     
-      			train_model_robust_maxnodes <- randomForest(all_CAMEO_6mo_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = all_CAMEO_6mo_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-    
-        		train_model_robust_sampsize <- randomForest(all_CAMEO_6mo_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 500, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = all_CAMEO_6mo_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-      			train_model_robust_ntree <- randomForest(all_CAMEO_6mo_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = all_CAMEO_6mo_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest1 <- randomForest(all_CAMEO_6mo_train_formula_civil_ns_robust_traintest1,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = all_CAMEO_6mo_train_frame_civil_ns_robust_traintest1, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest2 <- randomForest(all_CAMEO_6mo_train_formula_civil_ns_robust_traintest2,
-        		                  type = regression, importance = TRUE, 
-        		                  na.action = na.omit, ntree = 100000, maxnodes = 5, 
[…]

### 2020-04-29 Replication files/replication-3_BM/code/6mo_run_avg.R

@@ -7,54 +7,54 @@
                             prediction_all_CAMEO_6mo_inc_civil_ns)
                          /4)
 
-            prediction_robust_maxnodes <- as.numeric((prediction_robust_maxnodes_escalation_6mo_inc_civil_ns + 
-                                            prediction_robust_maxnodes_quad_6mo_inc_civil_ns + 
-                                            prediction_robust_maxnodes_goldstein_6mo_inc_civil_ns + 
-                                            prediction_robust_maxnodes_all_CAMEO_6mo_inc_civil_ns)
-                                         /4)
-
-            prediction_robust_sampsize <- as.numeric((prediction_robust_sampsize_escalation_6mo_inc_civil_ns + 
-                                            prediction_robust_sampsize_quad_6mo_inc_civil_ns + 
-                                            prediction_robust_sampsize_goldstein_6mo_inc_civil_ns + 
-                                            prediction_robust_sampsize_all_CAMEO_6mo_inc_civil_ns)
-                                         /4)
-
-            prediction_robust_ntree <- as.numeric((prediction_robust_ntree_escalation_6mo_inc_civil_ns + 
-                                         prediction_robust_ntree_quad_6mo_inc_civil_ns + 
-                                         prediction_robust_ntree_goldstein_6mo_inc_civil_ns + 
-                                         prediction_robust_ntree_all_CAMEO_6mo_inc_civil_ns)
-                                      /4)
-
-            prediction_robust_traintest1 <- as.numeric((prediction_robust_traintest1_escalation_6mo_inc_civil_ns + 
-                                              prediction_robust_traintest1_quad_6mo_inc_civil_ns + 
-                                              prediction_robust_traintest1_goldstein_6mo_inc_civil_ns + 
-                                              prediction_robust_traintest1_all_CAMEO_6mo_inc_civil_ns)
-                                           /4)
-
-            prediction_robust_traintest2 <- as.numeric((prediction_robust_traintest2_escalation_6mo_inc_civil_ns + 
-                                              prediction_robust_traintest2_quad_6mo_inc_civil_ns + 
-                                              prediction_robust_traintest2_goldstein_6mo_inc_civil_ns + 
[…]

### 2020-04-29 Replication files/replication-3_BM/code/6mo_run_escalation.R

@@ -10,69 +10,69 @@
     												  data = escalation_6mo_train_frame_civil_ns, forest = TRUE   
     												  )			         
     
-      			train_model_robust_maxnodes <- randomForest(escalation_6mo_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = escalation_6mo_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-    
-        		train_model_robust_sampsize <- randomForest(escalation_6mo_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 500, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = escalation_6mo_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-      			train_model_robust_ntree <- randomForest(escalation_6mo_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = escalation_6mo_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest1 <- randomForest(escalation_6mo_train_formula_civil_ns_robust_traintest1,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = escalation_6mo_train_frame_civil_ns_robust_traintest1, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest2 <- randomForest(escalation_6mo_train_formula_civil_ns_robust_traintest2,
-        		                  type = regression, importance = TRUE, 
-        		                  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-        		                  sampsize = 100, 
-        		                  replace = FALSE, do.trace = FALSE , 
[…]

### 2020-04-29 Replication files/replication-3_BM/code/6mo_run_goldstein.R

@@ -10,69 +10,69 @@
     												  data = goldstein_6mo_train_frame_civil_ns, forest = TRUE   
     												  )			         
     
-      			train_model_robust_maxnodes <- randomForest(goldstein_6mo_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = goldstein_6mo_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-    
-        		train_model_robust_sampsize <- randomForest(goldstein_6mo_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 500, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = goldstein_6mo_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-      			train_model_robust_ntree <- randomForest(goldstein_6mo_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = goldstein_6mo_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest1 <- randomForest(goldstein_6mo_train_formula_civil_ns_robust_traintest1,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = goldstein_6mo_train_frame_civil_ns_robust_traintest1, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest2 <- randomForest(goldstein_6mo_train_formula_civil_ns_robust_traintest2,
-        		                  type = regression, importance = TRUE, 
-        		                  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-        		                  sampsize = 100, 
[…]

### 2020-04-29 Replication files/replication-3_BM/code/6mo_run_quad.R

@@ -10,69 +10,69 @@
     												  data = quad_6mo_train_frame_civil_ns, forest = TRUE   
     												  )			         
     
-      			train_model_robust_maxnodes <- randomForest(quad_6mo_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 10, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = quad_6mo_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-    
-        		train_model_robust_sampsize <- randomForest(quad_6mo_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 500, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = quad_6mo_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-      			train_model_robust_ntree <- randomForest(quad_6mo_train_formula_civil_ns,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 1000000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = quad_6mo_train_frame_civil_ns, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest1 <- randomForest(quad_6mo_train_formula_civil_ns_robust_traintest1,
-    												  type = regression, importance = TRUE, 
-    												  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-    												  sampsize = 100, 
-    												  replace = FALSE, do.trace = FALSE , 
-    												  data = quad_6mo_train_frame_civil_ns_robust_traintest1, forest = TRUE   
-    												  )			         
-
-        		train_model_robust_traintest2 <- randomForest(quad_6mo_train_formula_civil_ns_robust_traintest2,
-        		                  type = regression, importance = TRUE, 
-        		                  na.action = na.omit, ntree = 100000, maxnodes = 5, 
-        		                  sampsize = 100, 
-        		                  replace = FALSE, do.trace = FALSE , 
[…]

### 2020-04-29 issues.Rmd

--- 2020-04-29 issues.Rmd
@@
   - 5-year forecasts for sep plots and AUC scores)
+- Smoothing of AUC/ROC scores
 - Lack of yearly test forecasts in favor of a single 5-year test forecast

### 2020-04-30 issues.Rmd
update issues note

@@
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
 
@@
 The concern is that the settings work well only for the escalation model, and only for the particular test set at hand.
-Specifically we wonder if the CAMEO model with 1,100+ features, compared to 10 or less for the other specifications, would perform better with a more explicit tuning procedure.
+Specifically, we wonder (1) if the hyperparameter settings only work well for the particular test set chosen, and (2) if the CAMEO model with 1,100+ features, compared to 10 or less for the other specifications, would perform better with a more explicit tuning procedure.
 
-The table below shows cross-validated training and OOS test AUC-ROC for the original 1 month escalation model and a modified version which uses more conventional random forest parameter settings (classification, 5000 trees, and default settings otherwise).
-The first and second columns are average and SD AUC-ROC obtained through repeated cross-validation on the training data set.
+## Are the RF hyperparameters overfit to the test set?
+
+Does the base specification's good performance in the test set generalize?
+
+One initial piece of evidence is already available from Table 1.
+In addition to the test set starting 2008, B&S also evaluate test sets starting in 2009, 2010, and 2011 (sets 1, 2, and 3 in Table 1).
+For the 1 month models, when we compare the base specification performance of each model to the performance in the alternate test sets, the AUC-ROC increases in 2 (of 15) cases, is the same in 0 cases, and decreases in 13 cases.
+For the 6 month models, it increases in 1 case, is the same in 3 cases, and decreases in 11 cases.
+Altering the test set thus generally shows reduced performance.
+The table below shows cross-validated training and OOS test AUC-ROC for the original 1 month escalation model, a RF model with default hyperparameter settings, and a tuned RF model with 10,000 trees and otherwise the default settings.
+The first set of results are out of sampel results from repeated cross-validations performed on the training data.
+We show the average AUC-ROC, it's lower and upper 95\% CI, obtained via bootstrapping, and the standard deviation of the distribution of resampled AUC-ROC values.
+
 Because there are only 9 positive cases in the whole training data, and thus to ensure that any given split will include at least 1 positive case in each data partition, we use 2-fold CV, i.e.
@@
 ```{r}
-tbl <- structure(list(Model = c("Escalation, 1mo", "Modified Escalation, 1mo"
-), Avg_CV_ROC_AUC = c(0.698772142673124, 0.643982732304452), 
-    SD_CV_ROC_AUC = c(0.136834250535024, 0.104102374345325), 
-    Test_ROC_AUC = c(0.786872878159185, 0.583391239320482)), row.names = c(NA, 
--2L), class = c("tbl_df", "tbl", "data.frame"))
+tbl <- structure(list(Model = c("Escalation, 1mo", "Modified Escalation, 1mo", 
+"Tuned Escalation, 1mo"), Avg_CV_ROC_AUC = c(0.677707841149067, 
+0.619719616721681, 0.671630392456263), ci_lower = c(0.634101138978023, 
+0.5835673668958, 0.634713167892701), ci_upper = c(0.722598486545989, 
+0.65584632967561, 0.706310478133188), SD_CV_ROC_AUC = c(0.147542574525528, 
+0.122337186539073, 0.117973980721395), Test_ROC_AUC = c(0.783352194140576, 
+0.586147564308735, 0.582927991423296)), row.names = c(NA, -3L
+), class = c("tbl_df", "tbl", "data.frame"))
 
@@
 
-Two thigs stand out:
+The results show that:
 
-- The escalation model has better performance in the test data than in the training CV samples. This is consistent with hyperparameter values that were (over-)fit to the test data. 
-- Although the original escalation model outperforms the modified model in the training CV resamples (p < 0.05), the difference in means is much less pronounced than in the test data.
+- Both of the alternative RF models are able to achieve roughly similar OOS performance in the training data split.
+*T*-tests comparing model 1 to the other models fail to reject the null hypothesis at a 95\% confidence level.
+(A *t*-test comparing model 1 and 2 average AUC-ROC just barely fails to reject the null hypothesis (*p* slightly above 0.05).)
+- Only the base escalation model is able to achive good test performance; the other two models, despite achieving similar training data performance, have significantly lower test performance.
+
+This suggests that the base RF specifications are (over-)fit to the test data.
 

### 2020-05-01 Replication files/replication-3_BM/tables/replicated-table-1.tex
built table 1

<table class="table" style="margin-left: auto; margin-right: auto;">
<caption></caption>
 <thead>
  <tr>
   <th style="text-align:left;"> Model </th>
   <th style="text-align:left;"> Horizon </th>
   <th style="text-align:left;"> ROC Smoothed </th>
   <th style="text-align:center;"> Escalation </th>
   <th style="text-align:center;"> Quad </th>
   <th style="text-align:center;"> Goldstein </th>
   <th style="text-align:center;"> CAMEO </th>
   <th style="text-align:center;"> Avg </th>
  </tr>
 </thead>
<tbody>
  <tr>
   <td style="text-align:left;vertical-align: middle !important;" rowspan="4"> Base Specification </td>
   <td style="text-align:left;vertical-align: middle !important;" rowspan="2"> 1 Month </td>
   <td style="text-align:left;font-weight: bold;"> Yes </td>
   <td style="text-align:center;"> 0.853 </td>
   <td style="text-align:center;"> 0.804 </td>
   <td style="text-align:center;"> 0.791 </td>
   <td style="text-align:center;"> 0.837 </td>
   <td style="text-align:center;"> 0.825 </td>
  </tr>
  <tr>

   <td style="text-align:left;font-weight: bold;"> No </td>
   <td style="text-align:center;"> 0.786 </td>
   <td style="text-align:center;"> 0.786 </td>
   <td style="text-align:center;"> 0.794 </td>
   <td style="text-align:center;"> 0.809 </td>
   <td style="text-align:center;"> 0.822 </td>
  </tr>
  <tr>

   <td style="text-align:left;vertical-align: middle !important;" rowspan="2"> 6 Months </td>
   <td style="text-align:left;font-weight: bold;"> Yes </td>
   <td style="text-align:center;"> 0.853 </td>
   <td style="text-align:center;"> 0.804 </td>
   <td style="text-align:center;"> 0.791 </td>
   <td style="text-align:center;"> 0.837 </td>
   <td style="text-align:center;"> 0.825 </td>
  </tr>
  <tr>

   <td style="text-align:left;font-weight: bold;"> No </td>
   <td style="text-align:center;"> 0.786 </td>
   <td style="text-align:center;"> 0.786 </td>
   <td style="text-align:center;"> 0.794 </td>
   <td style="text-align:center;"> 0.809 </td>
   <td style="text-align:center;"> 0.822 </td>
  </tr>
</tbody>
<tfoot>
<tr>
<td style = 'padding: 0; border:0;' colspan='100%'><sup></sup> To create Table 1, Blair and Sambanis use a non-standard &quot;smoothing&quot; function when creating their various Receiver Operating Characteristic Curves (ROC). These &quot;smoothed&quot; curves are then used to compute their model performance metric -- the Area Under the Receiver Operating Characteristic Curve (AUC-ROC). We recreate these AUC-ROC scores with and without the smoothing function (Blair and Sambanis' specification and the standard specification, respectively). This allows us to see if their findings are sensitive to the use of the smoothing function.</td>
</tr>
</tfoot>
</table>

### 2020-05-01 Replication files/replication-3_BM/code/1mo_run_avg.R

@@ -67,10 +67,12 @@ prediction <- as.numeric((prediction_escalation_inc_civil_ns +
 # Calculate AUC
 performance <- prediction(prediction, test_DV_civil_ns)
 roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+saveRDS(roc, "figures/roc-1mo-avg.rds")
 AUC_obs <- data.frame(as.numeric(roc$auc))
 
 ## RM: ADDED 4/29/2020
 roc_noSmooth <- roc(test_DV_civil_ns, prediction, smooth=FALSE, auc = TRUE)
+saveRDS(roc, "figures/non-smooth-roc-1mo-avg.rds")
 AUC_obs_noSmooth <- data.frame(as.numeric(roc_noSmooth$auc))
 #   			performance_robust_maxnodes <- prediction(prediction_robust_maxnodes, test_DV_civil_ns)
 #           roc_robust_maxnodes <- roc(test_DV_civil_ns, prediction_robust_maxnodes, smooth=TRUE, auc = TRUE)

### 2020-05-01 Replication files/replication-3_BM/code/1mo_run_escalation.R

@@ -122,10 +122,12 @@ prediction[is.na(escalation_test_frame_civil_ns$test_DV_civil_ns)] <- NA
 
 performance <- prediction(prediction, test_DV_civil_ns)
 roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+saveRDS(roc, "figures/roc-1mo-escalation.rds")
 AUC_obs <- data.frame(as.numeric(roc$auc))
 
 ## RM: ADDED 4/29/2020
 roc_noSmooth <- roc(test_DV_civil_ns, prediction, smooth=FALSE, auc = TRUE)
+saveRDS(roc, "figures/non-smooth-roc-1mo-escalation.rds")
 AUC_obs_noSmooth <- data.frame(as.numeric(roc_noSmooth$auc))
 #
 #       			performance_robust_maxnodes <- prediction(prediction_robust_maxnodes, test_DV_civil_ns)

### 2020-05-01 Replication files/replication-3_BM/code/1mo_run_goldstein.R

@@ -123,9 +123,11 @@ prediction[is.na(goldstein_test_frame_civil_ns$test_DV_civil_ns)] <- NA
 
 performance <- prediction(prediction, test_DV_civil_ns)
 roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+saveRDS(roc, "figures/roc-1mo-goldstein.rds")
 AUC_obs <- data.frame(as.numeric(roc$auc))
 ## RM: ADDED 4/29/2020
 roc_noSmooth <- roc(test_DV_civil_ns, prediction, smooth=FALSE, auc = TRUE)
+saveRDS(roc, "figures/non-smooth-roc-1mo-goldstein.rds")
 AUC_obs_noSmooth <- data.frame(as.numeric(roc_noSmooth$auc))
 
 # 				  performance_robust_maxnodes <- prediction(prediction_robust_maxnodes, test_DV_civil_ns)

### 2020-05-01 Replication files/replication-3_BM/code/6mo_run_all_CAMEO.R

@@ -123,10 +123,12 @@ prediction[is.na(all_CAMEO_6mo_test_frame_civil_ns$test_DV_civil_ns)] <- NA
 
 performance <- prediction(prediction, test_DV_civil_ns)
 roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+saveRDS(roc, "figures/roc-6mo-cameo.rds")
 AUC_obs <- data.frame(as.numeric(roc$auc))
 
 ## RM: ADDED 4/29/2020
 roc_noSmooth <- roc(test_DV_civil_ns, prediction, smooth=FALSE, auc = TRUE)
+saveRDS(roc, "figures/non-smooth-roc-6mo-cameo.rds")
 AUC_obs_noSmooth <- data.frame(as.numeric(roc_noSmooth$auc))
 
 #   			performance_robust_maxnodes <- prediction(prediction_robust_maxnodes, test_DV_civil_ns)
@@ -169,7 +171,20 @@ AUC_obs_noSmooth <- data.frame(as.numeric(roc_noSmooth$auc))
 #                        AUC_obs_robust_traintest3, AUC_obs_robust_civil_ns_alt1, AUC_obs_robust_civil_ns_alt2))
 
 ## RM: ADDED 4/29/2020
+# keep track of both smoothed and original AUC ROC; write a table with these
+# intermediate results so that Table 1 can be reconstructed later without having
+# to re-run all of the model scripts.
 AUCs <- as.matrix(c(AUC_obs, AUC_obs_noSmooth))
+
+tbl <- data.frame(
+  model = "base specification",
+  specification = "CAMEO",
+  horizon = "6 month",
+  smoothed = unname(AUC_obs),
+  original = unname(AUC_obs_noSmooth)
+)
+write.csv(tbl, "tables/auc-6mo-cameo.csv", row.names = FALSE)
+
 # Calculate PR
 
 precision_recall <- performance(performance, "prec", "rec")

### 2020-05-01 Replication files/replication-3_BM/code/6mo_run_quad.R

@@ -123,10 +123,12 @@ prediction[is.na(quad_6mo_test_frame_civil_ns$test_DV_civil_ns)] <- NA
 
 performance <- prediction(prediction, test_DV_civil_ns)
 roc <- roc(test_DV_civil_ns, prediction, smooth=TRUE, auc = TRUE)
+saveRDS(roc, "figures/roc-6mo-quad.rds")
 AUC_obs <- data.frame(as.numeric(roc$auc))
 
 ## RM: ADDED 4/29/2020
 roc_noSmooth <- roc(test_DV_civil_ns, prediction, smooth=FALSE, auc = TRUE)
+saveRDS(roc, "figures/non-smooth-roc-6mo-quad.rds")
 AUC_obs_noSmooth <- data.frame(as.numeric(roc_noSmooth$auc))
 
 #     				  performance_robust_maxnodes <- prediction(prediction_robust_maxnodes, test_DV_civil_ns)

### 2020-05-01 Replication files/replication-3_BM/master.R

@@ -348,26 +348,26 @@ source("code/6mo_define_models.R")
 # Run models
 
 source("code/6mo_run_escalation.R")
+write.dta(pred_escalation_6mo_inc_civil_ns,"predictions/6mo_predictions_escalation.dta", convert.factors="string")
 
 source("code/6mo_run_quad.R")
+write.dta(pred_quad_6mo_inc_civil_ns,"predictions/6mo_predictions_quad.dta", convert.factors="string")
 
 source("code/6mo_run_goldstein.R")
+write.dta(pred_goldstein_6mo_inc_civil_ns,"predictions/6mo_predictions_goldstein.dta", convert.factors="string")
 
 source("code/6mo_run_all_CAMEO.R")
+write.dta(pred_all_CAMEO_6mo_inc_civil_ns,"predictions/6mo_predictions_all_CAMEO.dta", convert.factors="string")
 
 source("code/6mo_run_avg.R")
+write.dta(pred_avg_6mo_inc_civil_ns,"predictions/6mo_predictions_avg.dta", convert.factors="string")
 
 # Save predictions
 
-write.dta(pred_escalation_6mo_inc_civil_ns,"predictions/6mo_predictions_escalation.dta", convert.factors="string")
 
-write.dta(pred_quad_6mo_inc_civil_ns,"predictions/6mo_predictions_quad.dta", convert.factors="string")
 
-write.dta(pred_goldstein_6mo_inc_civil_ns,"predictions/6mo_predictions_goldstein.dta", convert.factors="string")
 
-write.dta(pred_all_CAMEO_6mo_inc_civil_ns,"predictions/6mo_predictions_all_CAMEO.dta", convert.factors="string")
 
-write.dta(pred_avg_6mo_inc_civil_ns,"predictions/6mo_predictions_avg.dta", convert.factors="string")
 
 # Create bottom panel of Table 1
 

### 2020-05-01 Replication files/replication-3_BM/recreate-table1.R

@@ -1,20 +1,51 @@
 #
 #   Recreate table 1 with non-smoothed ROC
 #
-
+# install.packages("")
 library(readr)
+library(kableExtra)
+library(knitr)
 library(dplyr)
 library(here)
 
 setwd(here::here("Replication files/replication-3_BM"))
 
-auc <- lapply(dir("tables", pattern = "auc-", full.names = TRUE), read_csv,
-              col_types = cols(
-                model = col_character(),
-                specification = col_character(),
-                horizon = col_character(),
-                smoothed = col_double(),
-                original = col_double()
-              )) %>%
-  bind_rows() %>%
-  select(model, horizon, specification, smoothed, original)
+top <- read_csv("tables/table1_top.csv") %>% 
+  rename(`ROC Smoothed` = X1) %>% 
+  mutate(Horizon = "1 Month", 
+         model = "Base Specification", 
+         `ROC Smoothed` = case_when(`ROC Smoothed` == "Smoothed" ~ "Yes", 
+                                    `ROC Smoothed` == "Not Smoothed" ~ "No")) %>% 
+  select(model, Horizon, everything())
+
+bottom <- read_csv("tables/table1_bottom.csv") %>% 
+  rename(`ROC Smoothed` = X1) %>% 
+  mutate(Horizon = "6 Months", 
+         model = "Base Specification", 
+         `ROC Smoothed` = case_when(`ROC Smoothed` == "Smoothed" ~ "Yes", 
+                                    `ROC Smoothed` == "Not Smoothed" ~ "No")) %>% 
+  select(model, Horizon, everything())
+
+tab_dat <- bind_rows(top, bottom)
+names(tab_dat) <- c("Model", "Horizon", "ROC Smoothed", "Escalation", "Quad", "Goldstein", "CAMEO", "Avg")
+
+footnote_text <- 'To create Table 1, Blair and Sambanis use a non-standard "smoothing" function when creating their various Receiver Operating Characteristic Curves (ROC). These "smoothed" curves are then used to compute their model performance metric -- the Area Under the Receiver Operating Characteristic Curve (AUC-ROC). We recreate these AUC-ROC scores with and without the smoothing function (Blair and Sambanis\' specification and the standard specification, respectively). This allows us to see if their findings are sensitive to the use of the smoothing function.'
+
+knitr::kable(tab_dat, "html", booktabs = TRUE, linesep = "\\addlinespace", digits = c(3),
[…]

### 2020-05-01 issues.Rmd
change issues.Rmd to output PDF

@@
 ---
-title: "Replicating Blair and Sambanis: Forecasting Civil Wars - JCR 2020"
-output: github_document
+title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
+author: Richard K. Morgan and Andreas Beger
+date: "`r format(Sys.time(), '%d %B %Y')`"
+output: 
+  pdf_document:
+    keep_tex: false
+references:
+- id: blair2020forecasting
+  title: "Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning"
+  author:
+  - family: Blair
+    given: Robert A.
+  - family: Sambanis
+    given: Nicholas
+  container-title: Journal of Conflict Resolution
+  type: article-journal
+  issued:
+    year: 2020
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
+
+library(readr)
+library(dplyr)
 ```
 
-### Introduction
+We highlight some issues with the analysis in the recent Blair and Sambanis paper that undermine their main empirical finding, that a theoretically-informed model can forecast civil war onsets better than more inductive alternatives.
 
-While reading Blair and Sambanis (2020), a few issues stuck out.
-We downloaded the replication files authors provide from https://journals.sagepub.com/doi/suppl/10.1177/0022002720918923/suppl_file/sj-zip-1-jcr-10.1177_0022002720918923.zip.
-The original folder name of is `sj-zip-1-jcr-10.1177_0022002720918923/replication-3`.
-For simplicity, we extracted this `replication-3` folder and saved to the `Replication Files` folder of this repo.
-To ensure that we have a clean copy of the original replication scripts, I copied all files to `replication-3_BM` I'm (RM) currently using the sripts in this folder to recreate figure 1 and table 1 with and without smoothing -- see below
+- The key findings depend on the decision to use smoothed rather than original ROC curves when calculating area under the ROC curve (AUC-ROC) fit statistics.
+When using regular, not smoothed ROC curves, the findings reverse.
+- The 2016 forecasts for the risk of civil war onset appear to be evaluated using civil war *incidence* not *onset*. This distorts the assessment of how accurate they were. (TBD)
+- The way the random forest models are fitted by B&S is very unorthodox.^[This is in fact what first led us to examine the analysis in more detail.] We evaluate two questions: (1) does the key model's superior test forecast performance generalize or is it a result of inadvertant fitting of the model hyperparameters to test forecast performance, and (2) does the inferior performance of the alternative models, and especially the CAMEO model with 1,159 covariates, reflect a lack of proper hyperparameter tuning for these models?
+(TBD)
+
+# Review of the Blair and Sambanis paper
+
+@blair2020forecasting (B&S) argue that theory contributes to forecasting accuracy even when using non-parametric machine learning models and alternative specifications that are not informed by theory.
+They arrive at this conclusion by examining the problem of predicting civil war onset, and find that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory and a kitchen sink model with more than 1,100 covariates.
+
+Their data consist of country-month observations covering 2001 to 2015.
+The main outcome is civil war onset, measured using Sambanis' data.
+The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country.
+They key results are based on a comparison of the forecast accuracy of a random forest (RF) model with 4 covariate specifications as well as an ensemble model:
+
+- Escalation: a set of 10 theoretically-informed indicators.
+- Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
+- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government one one side and opposition or rebel actors on the other.
+These are directed, thus making for 4 total covariates.
+- CAMEO: counts for all CAMEO event codes, thus a total of 1,159 covariates. 
+- Average: unweighted average of the predictions from the 4 models above.
+
+To assess forecast accuracy, B&S:
+
+1. Split the training data into training (2001 - 2007) and test (2008 - 2015) sets.
+2. Estimate 5 random forest models with the specifications from above. 
+3. Create out-of-sample (OOS) predictions from each model for the test set. 
+4. Calculate AUC-ROC measures for each set of OOS predictions.
+
+This is done for both 1-month ahead and 6-month ahead forecasts.
+B&S also conduct a series of robustness tests that (1) vary hyperparameters of the RF model, (2) change the split year for the trainin/test data split, and (3) alternative codings of the dependent variable.
+
+The results for this analysis are summarized in Table 1 in the B&S paper, which we reproduce here:
+
+![](figures/bands-table1.png)
+
+The escalation model generally outperforms the alternative models/specifications, and on that basis B&S conclude that theory does indeed provide better forecast accuracy when compared to models not informed by theory.
 
@@
 
-This document lists the primary issues we identified.
+# Issues
 
-1. Using AUC for smoothed ROC curves. Does the interpretation of Table 1, and superiority of the escalation model, depend on the choice to use smoothed ROC curves?
+## Smoothed ROC curves
 
-2.
-In Table 4, are the forecasts evaluated against incidence or onset?
-The data from 2001 to 2015 have 20 civil war onsets in total, yet Table 4 appears to show 15 and 16 positive events for 2016 only.
-If this is indeed incidence, but the original models and forecasts were for onset, then this is a incorrect assessment of the forecasts' accuracy.
+```{r}
+auc <- read_csv("data/table1-redone.csv") %>%
+  mutate(specification = factor(specification, levels = c("escalation", "quad", "goldstein", "CAMEO", "avg"))) %>%
+  arrange(model, horizon, specification)
 
-- Table 4! 
-  - There are only 20 onsets from 2001 to 2015 But there are 15 "Persistence" cases (top of table 4) and 16 "Change" cases (bottom of table 4)
+auc %>%
+  knitr::kable(digits = 2)
+```
+
+Using AUC-ROC values from un-smoothed, original ROC curves reverses the original conclusions from B&S Table 1.
+In both the 1 month-ahead and 6-month ahead forecasts, the escalation model is outperformed by alternative models.
+
+## Assessment of the 2016 forecasts
+
+The main data set covering 2001 - 2015 contains 20 civil war onsets. But there are 15 "Persistence" cases (top of table 4) and 16 "Change" cases (bottom of table 4)
+
+## Random forest model tuning
 
@@
 5.
+
+# References

### 2020-05-04 data/replicated-table-1.tex

\begin{table}

\begin{threeparttable}
\caption{\label{tab:tab-1}}
\centering
\begin{tabular}[t]{ll>{\bfseries}lccccc}
\toprule
 & Horizon & ROC Smoothed & Escalation & Quad & Goldstein & CAMEO & Avg\\
\midrule
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{Base Specification}}\\
\hspace{1em} & 1 Month & Yes & 0.853 & 0.804 & 0.791 & 0.837 & 0.825\\
\cmidrule{3-8}
\hspace{1em}\hspace{1em} &  & No & 0.786 & 0.786 & 0.794 & 0.809 & 0.822\\
\cmidrule{2-8}
\hspace{1em} & 6 Months & Yes & 0.853 & 0.804 & 0.791 & 0.837 & 0.825\\
\cmidrule{3-8}
 &  & No & 0.786 & 0.786 & 0.794 & 0.809 & 0.822\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item [] To create Table 1, Blair and Sambanis use a non-standard "smoothing" function when creating their various Receiver Operating Characteristic Curves (ROC). These "smoothed" curves are then used to compute their model performance metric -- the Area Under the Receiver Operating Characteristic Curve (AUC-ROC). We recreate these AUC-ROC scores with and without the smoothing function (Blair and Sambanis' specification and the standard specification, respectively). This allows us to see if their findings are sensitive to the use of the smoothing function.
\end{tablenotes}
\end{threeparttable}
\end{table}

### 2020-05-04 issues.Rmd
Small changes to issues, remade the table 1 csv to include all of the 6 mon mods, and moved my table 1 replication
investigate table 4

@@
 library(dplyr)
+library(kableExtra)
 ```
@@
 This is particularly true when the goal is to assess how ...
+(RM: I want to add something here about identifying potential policy interventions that can help advert onset...)
 
@@
 
+<!-- \input{data/replicated-table-1.tex} -->
+
 ## Assessment of the 2016 forecasts
@@
 The main data set covering 2001 - 2015 contains 20 civil war onsets. But there are 15 "Persistence" cases (top of table 4) and 16 "Change" cases (bottom of table 4)
+
+<!-- 
+AB: going to put some detailed technical notes in here.
+
+The 2016 forecasts are created in R and contained in pred_escalation_6mo_inc_civil_ns.
+That data frame actually has predictions for 2008 to 2016.
+
+The data from which all this is created has the following values for the plus1 and non-lead DV:
+
+  incidence_civil_ns_plus1     n
+                     <dbl> <int>
+1                        0  4581
+2                        1    20
+3                       NA   483
+
+  incidence_civil_ns     n
+               <dbl> <int>
+1                  0  4562
+2                  1    20
+3                 NA   502
+
+Looking at the pattern of missing values, it is fairly clear that the incidence variables were originally incidence, but changed to onset such that the first value in a sequence of 1's remained a 1 while the rest were set to NA.
+Like:
+
+c(0, 1, 1, 1, 0) becomes (c(0, 1, NA, NA, 0))
+
+And it is also fairly clear that "incidence_civil_ns_plus1" is a 1-period lead indicator for the outcome that one can use as a prediction target insteaad of having to lag all inputs.
+
+Anyways, in R training, "incidence_civil_ns_plus1" is used as the target for the models. It has 20 positive values for data ranging from 2001 to 2016.
+
+The predictions, including 2016 forecasts, are written to disk and then Stata picks up. The crucial coding for Table 4 is in `6mo_make_confusion_matrix.do`.
+
+Two strange things happen there.
+
+Sub-issue 1: First, the forecasts are assessed with "incidence_civil_ns", not "incidence_civil_ns_plus1".
+Based on the DV names, I would imagine that "plus1" is a lead version, so the actual predictions for 2016-H1 are contained in the predicted value for "incidence_civil_ns_plus1" in 2015-H2, since the plus 1 puts the time period to 2016-H1.
+
+(It could be that the 2016-H1 data in "data/6mo_data_OOS.dta" are just a copy of 2015-H2, in which case ok. I haven't checked.)
+
+This is a relatively minor point.
+
+Sub-issue 2: The second issue, and this is key, is that missing values in "incidence_civil_ns" are recoded to 1 in `6mo_make_confusion_matrix.do` line 52.
+This is what does all the heavy lifting in Table 4.
+Otherwise there are no 1 values in the data:
+
+  incidence_civil_ns     n
+  <chr>              <int>
+1 0                    149
+2 NA                    15
+
+So it appears that the variable is essentially changed back from the onset coding in which ongonig conflicts are NA, back to incidence coding.
+
+-->
 

### 2020-05-04 issues.Rmd

--- 2020-05-01 issues.Rmd
@@
 library(dplyr)
+library(kableExtra)
 ```
 
-We highlight some issues with the analysis in the recent Blair and Sambanis paper that undermine their main empirical finding, that a theoretically-informed model can forecast civil war onsets better than more inductive alternatives.
+We highlight some issues with the analysis in the recent Blair and Sambanis paper that undermine their main empirical finding, that a theoretically-informed escalation model can forecast civil war onsets better than more inductive alternatives.
+Their paper has two key results:
 
-- The key findings depend on the decision to use smoothed rather than original ROC curves when calculating area under the ROC curve (AUC-ROC) fit statistics.
-When using regular, not smoothed ROC curves, the findings reverse.
-- The 2016 forecasts for the risk of civil war onset appear to be evaluated using civil war *incidence* not *onset*. This distorts the assessment of how accurate they were. (TBD)
+1. A theoretically-informed model outperforms alternative specifications in predicting civil war onset. 
+2. Structural variables do not add, and sometimes detract, from the forecast accuracy of the escalation model.
+
+They also evaluate forecasts for the first half of 2016 and conclude good accuracy in forecasting civil war.
+
+We find three issues in their empirical analyses:
+
+- The first key finding depends entirely on using smoothed ROC curves when calculating the AUC-ROC values in Table 1.
+When using the original ROC curves, the findings reverse.
+Alternative specifications consistently beat the escalation model.
+- Regarding the forecast assessment in Table 4, the forecatss for 2016-H1 are erraneously assessed using civil war incidence, not onset, even though the models from which they are gneerated are clearly fitted to onset.
+This vastly overstates the precision of the forecasts.
 - The way the random forest models are fitted by B&S is very unorthodox.^[This is in fact what first led us to examine the analysis in more detail.] We evaluate two questions: (1) does the key model's superior test forecast performance generalize or is it a result of inadvertant fitting of the model hyperparameters to test forecast performance, and (2) does the inferior performance of the alternative models, and especially the CAMEO model with 1,159 covariates, reflect a lack of proper hyperparameter tuning for these models?
-(TBD)
+(work still in progress)
 
@@
 
-```{r}
+```{r, message=FALSE}
 auc <- read_csv("data/table1-redone.csv") %>%
@@
 auc %>%
-  knitr::kable(digits = 2)
+  knitr::kable(digits = 2, booktabs = TRUE, caption = "Comparison of smoothed and original AUC-ROC values")
 ```
@@
 
+<!-- \input{data/replicated-table-1.tex} -->
+
 ## Assessment of the 2016 forecasts
 
+The predictions for the first half of 2016 are assessed in Table 4:
+
+![](figures/bands-table4.png)
+
+The table presents 15 or 16 positive cases for 2016-H1, depending on a minor coding decision. These numgers correspond to a positive rate of around 9.5% for the first half of 2016.
+
+By comparison, the 6-month version of the full data from 2001 to 2015 contains only 20 positive cases, with a positive rate of around 0.5%.
+
+The relevant variables in the data are "incidence_civil_ns" and "incidence_civli_ns_plus1", which appears to be a 1-period lead version of the DV that is used in the actual prediction models.
+The incidence DV contains both 0/1 and missing values.
+By examining the pattern of missing values, it seems clear that this was originally an incidence variable indicating whether a country was at civil war in a given year or not, and which was converted to an onset version so that onsets retain the value of 1 but continuing civil war years are coded as missing.
+
 The main data set covering 2001 - 2015 contains 20 civil war onsets. But there are 15 "Persistence" cases (top of table 4) and 16 "Change" cases (bottom of table 4)
+
+By examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
+In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable back to incidence.
+
+The actual data contains 0 new civil war onsets in 2016-H1.
+Using those data, the 2016-H1 forecasts would have retained a recall value of 1, but precision would have decreased from 13/30 = 0.76 to 1/30 = 0.03.
+
+The second part of the Table 4 script manually codes onset in two countries, Turkey and Burundi.
+With this coding, the correct recall (bottom of Table 4) is 1 (not 14/16 = 0.88) and the precision changes from 14/30 = 0.47 to 2/30 = 0.07.
+
+Another, minor issue or rather coding error, is related to using a lead version of the DV.
+With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 actually indicates the risk of civil war onset in 2016-H2.
+In the Table 4 script referenced above, the 2016-H1 predictions (for 2016-H2) are assessed using the raw DV, "incidence_civil_ns", not the lead version.
+Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1.
+In this case it probably doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.
 

### 2020-05-05 issues.Rmd
working on the issues note...

@@
 
-We highlight some issues with the analysis in the recent Blair and Sambanis paper that undermine their main empirical finding, that a theoretically-informed escalation model can forecast civil war onsets better than more inductive alternatives.
+We highlight some issues with the analysis in the recent @blair2020forecasting paper that undermine their main empirical finding -- that a theoretically-informed escalation model can forecast civil war onsets better than more inductive alternatives.
 Their paper has two key results:
@@
 1. A theoretically-informed model outperforms alternative specifications in predicting civil war onset. 
-2. Structural variables do not add, and sometimes detract, from the forecast accuracy of the escalation model.
+2. Structural variables do not add, and sometimes detract, from the forecast accuracy of the escalation model they advance.
 
-They also evaluate forecasts for the first half of 2016 and conclude good accuracy in forecasting civil war.
+These results are highlighted in Table 1 and Table 2 [@blair2020forecasting, pg.
+13 and 19 respectively].
+They also evaluate forecasts for the first half of 2016 and conclude good accuracy in forecasting civil war (Table 4 on pg.23).
 
-We find three issues in their empirical analyses:
+We find three main issues in their empirical analyses:
 
-- The first key finding depends entirely on using smoothed ROC curves when calculating the AUC-ROC values in Table 1.
-When using the original ROC curves, the findings reverse.
-Alternative specifications consistently beat the escalation model.
-- Regarding the forecast assessment in Table 4, the forecatss for 2016-H1 are erraneously assessed using civil war incidence, not onset, even though the models from which they are gneerated are clearly fitted to onset.
+- Their primary finding -- that the escalation model they advance consistently out-performs the alternation models -- depends entirely on the unorthodox use of smoothed ROC curves when calculating the AUC-ROC values in Table 1 and Table 2.
+When using the standard (not smoothed) ROC curves, this finding is reversed.
+That is, the escalation performs no better and often worse than the other models.
+- Regarding the forecast assessment in Table 4, the forecasts for 2016-H1 are erroneously assessed using civil war _incidence_, not _onset_, even though the models from which they are generated are clearly fitted to civil conflict onset.
 This vastly overstates the precision of the forecasts.
-- The way the random forest models are fitted by B&S is very unorthodox.^[This is in fact what first led us to examine the analysis in more detail.] We evaluate two questions: (1) does the key model's superior test forecast performance generalize or is it a result of inadvertant fitting of the model hyperparameters to test forecast performance, and (2) does the inferior performance of the alternative models, and especially the CAMEO model with 1,159 covariates, reflect a lack of proper hyperparameter tuning for these models?
-(work still in progress)
+- The way the random forest models are fitted is very unorthodox.^[This is in fact what first led us to examine the analysis in more detail.] The number of tree the model grow, the number of nodes in these trees, and the sample size used for each tree deviates drastically for traditional approaches (e.g., they set the number of trees to 100,000 where the default is 500).
+
+We demonstrate the first point through model replication: we use the replication files that Blair and Sambanis provide to calculate the AUC-ROC scores using smoothed ROC curves (the published results) and then calculate the AUC-ROC scores using the standard approach -- without the unorthodox smoother.^[In this draft, we focus our replication efforts on what Blair and Sambanis call their "Base specification".
+This set of models are what Blair and Sambanis use when discussing their main findings and the implications of their results.] We demonstrate the second point using the replicated results from the original model Blair and Sambanis use to create Table 4 and then compare these results to the observed outcomes in the 2016 data.
+
+For the final point, we evaluate two questions: (1) even if we allow for the use of smoothed ROC curves, does the key model's superior test forecast performance generalize or is it a result of inadvertent fitting of the model hyperparameters to test forecast performance, and (2) does the inferior performance of the alternative models, and especially the CAMEO model with 1,159 covariates, reflect a lack of proper hyperparameter tuning for these models?
+(work still in progress).
 
@@
 
-@blair2020forecasting (B&S) argue that theory contributes to forecasting accuracy even when using non-parametric machine learning models and alternative specifications that are not informed by theory.
-They arrive at this conclusion by examining the problem of predicting civil war onset, and find that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory and a kitchen sink model with more than 1,100 covariates.
+@blair2020forecasting argue that theory contributes to forecasting accuracy even when using non-parametric machine learning models (in this case, the random forest model) and alternative specifications that are not informed by theory.
+They arrive at this conclusion by examining the problem of predicting civil war onset, and find that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory and a kitchen sink model with more than 1,000 covariates.
 
 Their data consist of country-month observations covering 2001 to 2015.
-The main outcome is civil war onset, measured using Sambanis' data.
+The main outcome is civil war onset, measured using Sambanis' civil war dataset.
 The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country.
-They key results are based on a comparison of the forecast accuracy of a random forest (RF) model with 4 covariate specifications as well as an ensemble model:
+The key results are based on a comparison of the forecast accuracy of a random forest (RF) model with 4 covariate specifications as well as an ensemble model.
+These models are:
 
@@
 
-To assess forecast accuracy, B&S:
+To assess forecast accuracy, Blair and Sambanis conduct the following procedures:
 
 1. Split the training data into training (2001 - 2007) and test (2008 - 2015) sets.
-2. Estimate 5 random forest models with the specifications from above. 
-3. Create out-of-sample (OOS) predictions from each model for the test set. 
+2. Estimate 4 random forest models with the specifications from above as well as a unweighted average of the models. 
+3. Create out-of-sample (OOS) predictions from each model using the test set. 
 4. Calculate AUC-ROC measures for each set of OOS predictions.
@@
 This is done for both 1-month ahead and 6-month ahead forecasts.
-B&S also conduct a series of robustness tests that (1) vary hyperparameters of the RF model, (2) change the split year for the trainin/test data split, and (3) alternative codings of the dependent variable.
+Blair and Sambanis also conduct a series of robustness tests that (1) vary hyperparameters of the RF model, (2) change the split year for the training/test data split, and (3) alternative codings of the dependent variable.
 
-The results for this analysis are summarized in Table 1 in the B&S paper, which we reproduce here:
+The results for this analysis are summarized in Table 1 in @blair2020forecasting, which we reproduce here:
 
@@
 
-The escalation model generally outperforms the alternative models/specifications, and on that basis B&S conclude that theory does indeed provide better forecast accuracy when compared to models not informed by theory.
+The escalation model generally out-performs the alternative models/specifications, and on that basis Blair and Sambanis conclude that theory does indeed provide better forecast accuracy when compared to models not informed by theory.
 
-Before listing the primary issues we identified, we should note that we are partial to their approach.
-Theory is important.
-It is helpful for forecasting problems, especially with regards to identifying which variables to include in a model.
-This is particularly true when the goal is to assess how ...
-(RM: I want to add something here about identifying potential policy interventions that can help advert onset...)
+After establishing that the escalation model they advance out-performs all other covariate specifications, Blair and Sambanis move to show how the inclusion of structural risk variable (using the set of indicators identified by the Political Instability Task Force (PITF) )
 
 # Issues
+
+Below, we work through our issues we identified above, starting first with the unconventional use of a smoother when calculating the AUC-ROC scores. We then discusses
 
@@
 
-Using AUC-ROC values from un-smoothed, original ROC curves reverses the original conclusions from B&S Table 1.
+Using AUC-ROC values from un-smoothed, original ROC curves reverses the original conclusions from Blair and Sambanis Table 1.
 In both the 1 month-ahead and 6-month ahead forecasts, the escalation model is outperformed by alternative models.
 
-<!-- \input{data/replicated-table-1.tex} -->
+<!-- \input{Tables/replicated-table-1.tex} -->
 
@@
 
-The table presents 15 or 16 positive cases for 2016-H1, depending on a minor coding decision. These numgers correspond to a positive rate of around 9.5% for the first half of 2016.
+The table presents 15 or 16 positive cases for 2016-H1, depending on a minor coding decision. These numbers correspond to a positive rate of around 9.5% for the first half of 2016.
 
@@
 
-The relevant variables in the data are "incidence_civil_ns" and "incidence_civli_ns_plus1", which appears to be a 1-period lead version of the DV that is used in the actual prediction models.
+The relevant variables in the data are "incidence_civil_ns" and "incidence_civil_ns_plus1", which appears to be a 1-period lead version of the DV that is used in the actual prediction models.
 The incidence DV contains both 0/1 and missing values.
@@
 
-The table below shows the randomForest package default hyperparameter values for a binary classification problem like the one at hand and compares them to the B&S base settings.
+The table below shows the randomForest package default hyperparameter values for a binary classification problem like the one at hand and compares them to the Blair and Sambanis base settings.
 Some hyperparameters have heuristics that determine the default value based on characteristics of the input data; we note these in the second column and what the realized default setting would be for the basic Escalation model (first row in Table 1 in the paper) in the 3rd column.
-The last column has the settings B&S use.
+The last column has the settings Blair and Sambanis use.
 
@@
 
-It is worth noting that commonly the ntree, mtry, and nodesize parameters are the main parameters tuned; of these three in the B&S specification only 1, the ntree parameter, deviates from the default settings.
+It is worth noting that commonly the ntree, mtry, and nodesize parameters are the main parameters tuned; of these three in the Blair and Sambanis specification only 1, the ntree parameter, deviates from the default settings.
 
-In any case, there is a stark contrast in the default RF settings and the way B&S use the RF models.
+In any case, there is a stark contrast in the default RF settings and the way Blair and Sambanis use the RF models.
 The default approach is to train a relatively small number (ntree; 500) of classification trees, but where each tree is fairly big in that it is trained on data that has the same number of rows as the training data, albeit sampled with replacement (replace is true; sampsize is 11,869), and allowed to grow fairly deep (this is governed by nodesize (1), which is the minimum size a terminal node must have).
-In contrast, B&S grow very extensive forests with a large number of trees (100,000 compared to 500), but each tree is very small and shallow; only a 100 rows are sampled from the training data for each tree, and the trees are constrained to at most 5 terminal nodes (maxnodes).
+In contrast, Blair and Sambanis grow very extensive forests with a large number of trees (100,000 compared to 500), but each tree is very small and shallow; only a 100 rows are sampled from the training data for each tree, and the trees are constrained to at most 5 terminal nodes (maxnodes).
 
-This approach only works due to the other unorthodox choice, which is to use regression, not classificaton, trees.
+This approach only works due to the other unorthodox choice, which is to use regression, not classification, trees.
 Trying to use classification trees with the other parameter settings in fact does not work at all because it is almost guaranteed that a sample of 100 from the 11,869 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample.
@@
 One initial piece of evidence is already available from Table 1.
-In addition to the test set starting 2008, B&S also evaluate test sets starting in 2009, 2010, and 2011 (sets 1, 2, and 3 in Table 1).
+In addition to the test set starting 2008, Blair and Sambanis also evaluate test sets starting in 2009, 2010, and 2011 (sets 1, 2, and 3 in Table 1).
 For the 1 month models, when we compare the base specification performance of each model to the performance in the alternate test sets, the AUC-ROC increases in 2 (of 15) cases, is the same in 0 cases, and decreases in 13 cases.
@@
 The table below shows cross-validated training and OOS test AUC-ROC for the original 1 month escalation model, a RF model with default hyperparameter settings, and a tuned RF model with 10,000 trees and otherwise the default settings.
-The first set of results are out of sampel results from repeated cross-validations performed on the training data.
+The first set of results are out of sample results from repeated cross-validations performed on the training data.
 We show the average AUC-ROC, it's lower and upper 95\% CI, obtained via bootstrapping, and the standard deviation of the distribution of resampled AUC-ROC values.
@@
 The last column shows the OOS performance on the original test set.
-The first value in this column corresponds to the base escalation model results reported Table 1 in the B&S paper when *not* using smoothed ROC curves.
+The first value in this column corresponds to the base escalation model results reported Table 1 in the Blair and Sambanis paper when *not* using smoothed ROC curves.
 
@@
 (A *t*-test comparing model 1 and 2 average AUC-ROC just barely fails to reject the null hypothesis (*p* slightly above 0.05).)
-- Only the base escalation model is able to achive good test performance; the other two models, despite achieving similar training data performance, have significantly lower test performance.
+- Only the base escalation model is able to achieve good test performance; the other two models, despite achieving similar training data performance, have significantly lower test performance.
 
@@
 Rick, I think the points below are more subjective choices.
-The first two points above are IMO potentially objectively incorrect technical errors that could undermine the original B&S findings.
+The first two points above are IMO potentially objectively incorrect technical errors that could undermine the original Blair and Sambanis findings.
 The 3rd point about the RF models is somewhere between objective error and subjective choice.

### 2020-05-05 Tables/replicated-table-1.tex
Added stuff on table 1 replication

--- 2020-05-04 data/replicated-table-1.tex
@@
 
-\begin{threeparttable}
-\caption{\label{tab:tab-1}}
+\caption{\label{tab:tab-1}Replication of Table 1 in Blair and Sambanis (2020) with and without ROC smoothers}
 \centering
@@
 \multicolumn{8}{l}{\textbf{Base Specification}}\\
-\hspace{1em} & 1 Month & Yes & 0.853 & 0.804 & 0.791 & 0.837 & 0.825\\
+\addlinespace[0.2em]
+\hspace{1em} & 1 Month & Yes & 0.85 & 0.80 & 0.79 & 0.84 & 0.82\\
 \cmidrule{3-8}
-\hspace{1em}\hspace{1em} &  & No & 0.786 & 0.786 & 0.794 & 0.809 & 0.822\\
+\hspace{1em}\hspace{1em} &  & No & 0.79 & 0.79 & 0.79 & 0.81 & 0.82\\
 \cmidrule{2-8}
-\hspace{1em} & 6 Months & Yes & 0.853 & 0.804 & 0.791 & 0.837 & 0.825\\
+\hspace{1em} & 6 Months & Yes & 0.85 & 0.80 & 0.79 & 0.84 & 0.82\\
 \cmidrule{3-8}
- &  & No & 0.786 & 0.786 & 0.794 & 0.809 & 0.822\\
+ &  & No & 0.79 & 0.79 & 0.79 & 0.81 & 0.82\\
 \bottomrule
 \end{tabular}
-\begin{tablenotes}
-\small
-\item [] To create Table 1, Blair and Sambanis use a non-standard "smoothing" function when creating their various Receiver Operating Characteristic Curves (ROC).
-These "smoothed" curves are then used to compute their model performance metric -- the Area Under the Receiver Operating Characteristic Curve (AUC-ROC).
-We recreate these AUC-ROC scores with and without the smoothing function (Blair and Sambanis' specification and the standard specification, respectively).
-This allows us to see if their findings are sensitive to the use of the smoothing function.
-\end{tablenotes}
-\end{threeparttable}
 \end{table}

### 2020-05-05 issues.Rmd
Added stuff on table 1 replication

@@
 library(dplyr)
+library(tidyr)
 library(kableExtra)
@@
 
-The results for this analysis are summarized in Table 1 in @blair2020forecasting, which we reproduce here:
+The results for this analysis are summarized in Table 1 in @blair2020forecasting, which we copied in here:
 
@@
 
-After establishing that the escalation model they advance out-performs all other covariate specifications, Blair and Sambanis move to show how the inclusion of structural risk variable (using the set of indicators identified by the Political Instability Task Force (PITF) )
+After establishing that the escalation model they advance out-performs all other covariate specifications, Blair and Sambanis move to show how the inclusion of structural risk variables -- the set of structural variables identified by the Political Instability Task Force (PITF) that can contribute to an increased risk of civil conflict.
+The figure below is Table 2 from @blair2020forecasting, which presents these findings.
+
+![](figures/bands-table2.png)
 
@@
 
-Below, we work through our issues we identified above, starting first with the unconventional use of a smoother when calculating the AUC-ROC scores. We then discusses
+Below, we work through our issues we identified above, starting first with the unconventional use of a smoother when calculating the AUC-ROC scores.
+We reproduce Table 1 and Table 2 from @blair2020forecasting to show that their results are dependent on the unconventional use of a smoother.
+
+We then discuss... (RM: I'll fill this out as I work through the evidence.)
 
 ## Smoothed ROC curves
+
+One issue that stood out as we read @blair2020forecasting was how smooth their Receiver Operating Characteristic curves where in Figure 1 on page 14.
+A Receiver Operating Characteristic curve displays the balance between the true positive rate and the false positive rate, across a range of discrimination thresholds -- the value in which a zero to one probability is assumed to be an "onset".
+Given that the DV is civil conflict onset (1/0) and that this is a rare event, these curves should resemble a step-like function.
+Looking through the Blair/Sambanis replication code, we noticed that they use a smoothing function when producing their Receiver Operating Characteristic curves.
+
+This is important because Blair and Sambanis use a common performance metric -- Area Under the Receiver Operating Characteristic Curves (AUC-ROC) -- as evidence that their escalation model out preforms all others.
+However, by smoothing the Receiver Operating Characteristic Curves, they are misrepresenting the true AUC-ROC scores of their various models.
+To help make this point clear, we replicate Figure 1 below.
+However, rather than a two panel figure (one ROC plot their one-month-ahead forecasts and one for their six-months-ahead forecasts, as is presented in the article), we now include four panels: one panel for the two forecast windows with the smoothed curve, as they produce, and one panel for the two forecast windows that are not smoothed.
+
+\clearpage
+
+Using these smoothed and non-smoothed curves, we then replicate Table 1 (page 13), but now the table reports the AUC-ROC scores for each model specification with and without the use of a smoother.
+If the non-smoothed AUC-ROC scores are drastically different from those of the smoothed AUC-ROC score (a difference of $\ge 0.05$) or if the non-smoothed AUC-ROC scores are < 0.8, would suggest the findings presented in @blair2020forecasting are fragile to the use of a non-standard procedure.
 
@@
   mutate(specification = factor(specification, levels = c("escalation", "quad", "goldstein", "CAMEO", "avg"))) %>%
-  arrange(model, horizon, specification)
+  arrange(model, horizon, specification) %>% 
+  pivot_longer(-c(model, horizon, specification), names_to = "ROC Smoothed", values_to = "Value") %>% 
+  pivot_wider( names_from = "specification", values_from = "Value") %>% 
+  mutate(`ROC Smoothed` = ifelse(`ROC Smoothed` == "smoothed", "Yes", "No"))
 
 auc %>%
-  knitr::kable(digits = 2, booktabs = TRUE, caption = "Comparison of smoothed and original AUC-ROC values")
+  knitr::kable(digits = 2, booktabs = TRUE, caption = "Comparison of smoothed and not smoothed AUC-ROC values") %>% 
+  # column_spec(c(3), bold = TRUE) %>%
+  collapse_rows(1:2, row_group_label_position = 'stack')
+
 ```
 
-Using AUC-ROC values from un-smoothed, original ROC curves reverses the original conclusions from Blair and Sambanis Table 1.
-In both the 1 month-ahead and 6-month ahead forecasts, the escalation model is outperformed by alternative models.
-
-<!-- \input{Tables/replicated-table-1.tex} -->
+As the table above shows, the results presented in @blair2020forecasting are fragile to the use of a smoothing option when calculating the AUC-ROC performance metric.
+Further, even when we rely solely on the replication files provided by Blair and Sambanis, we cannot replicate their findings exactly.
+For example, in our replication, the AUC-ROC score for the 1-month CAMEO model with the use of the smoother is 0.84; however, in the paper, the reported score is 0.82.
+This is most likely a function of the version of `R` Blair and Sambanis used to conduct their analysis.
+The current replication project was done using `R version 4.0.0`, which updated their random number seeding procedure, so small differences should be expected.
 

### 2020-05-05 MB-replication-checks/recreate-table1.R

@@ -10,8 +10,6 @@ library(here)
 
 setwd(here::here("MB-replication-checks"))
 
-
-
 # Write table summaries for monitoring on git
 
 auc <- lapply(dir("tables", pattern = "auc-", full.names = TRUE), read_csv,
@@ -28,45 +26,36 @@ auc <- lapply(dir("tables", pattern = "auc-", full.names = TRUE), read_csv,
 write_csv(auc, "tables/table1-redone.csv")
 write_csv(auc, "../data/table1-redone.csv")
 
-# Write latex versions of table 1
+## OLD
 
-top <- read_csv("tables/table1_top.csv") %>%
-  rename(`ROC Smoothed` = X1) %>%
-  mutate(Horizon = "1 Month",
-         model = "Base Specification",
-         `ROC Smoothed` = case_when(`ROC Smoothed` == "Smoothed" ~ "Yes",
-                                    `ROC Smoothed` == "Not Smoothed" ~ "No")) %>%
-  select(model, Horizon, everything())
-
-bottom <- read_csv("tables/table1_bottom.csv") %>%
-  rename(`ROC Smoothed` = X1) %>%
-  mutate(Horizon = "6 Months",
-         model = "Base Specification",
-         `ROC Smoothed` = case_when(`ROC Smoothed` == "Smoothed" ~ "Yes",
-                                    `ROC Smoothed` == "Not Smoothed" ~ "No")) %>%
-  select(model, Horizon, everything())
-
-tab_dat <- bind_rows(top, bottom)
-names(tab_dat) <- c("Model", "Horizon", "ROC Smoothed", "Escalation", "Quad", "Goldstein", "CAMEO", "Avg")
-
-footnote_text <- 'To create Table 1, Blair and Sambanis use a non-standard "smoothing" function when creating their various Receiver Operating Characteristic Curves (ROC). These "smoothed" curves are then used to compute their model performance metric -- the Area Under the Receiver Operating Characteristic Curve (AUC-ROC). We recreate these AUC-ROC scores with and without the smoothing function (Blair and Sambanis\' specification and the standard specification, respectively). This allows us to see if their findings are sensitive to the use of the smoothing function.'
-
-knitr::kable(tab_dat, "latex", booktabs = TRUE, linesep = "\\addlinespace", digits = c(3),
-             label = "tab-1", caption = "", align = c(rep("l", 3), rep("c", 5))) %>%
[…]

### 2020-05-05 issues.Rmd
commented out our notes and made small changes.

@@
 
-Below, we work through our issues we identified above, starting first with the unconventional use of a smoother when calculating the AUC-ROC scores.
+<!-- Below, we work through our issues we identified above, starting first with the unconventional use of a smoother when calculating the AUC-ROC scores.
 We reproduce Table 1 and Table 2 from @blair2020forecasting to show that their results are dependent on the unconventional use of a smoother.
+-->
 
-We then discuss... (RM: I'll fill this out as I work through the evidence.)
+<!-- We then discuss... (RM: I'll fill this out as I work through the evidence.)  -->
 
@@
 
+![](figures/figure1-replicated.png)
+
 \clearpage
@@
   pivot_wider( names_from = "specification", values_from = "Value") %>% 
-  mutate(`ROC Smoothed` = ifelse(`ROC Smoothed` == "smoothed", "Yes", "No"))
+  mutate(`ROC Smoothed` = ifelse(`ROC Smoothed` == "smoothed", "Yes", "No"), 
+         model = "Base Specification") %>% 
+  rename(Escalation = escalation, 
+         Quad = quad, 
+         Goldstein = goldstein, 
+         Avg = avg)
 
 auc %>%
-  knitr::kable(digits = 2, booktabs = TRUE, caption = "Comparison of smoothed and not smoothed AUC-ROC values") %>% 
+  knitr::kable(digits = 2, booktabs = TRUE, caption = "Comparison of smoothed and not smoothed AUC-ROC values", align = c(rep("l", 2), rep("c", 6))) %>% 
   # column_spec(c(3), bold = TRUE) %>%
-  collapse_rows(1:2, row_group_label_position = 'stack')
+  kableExtra::collapse_rows(1:2, row_group_label_position = 'stack')
 
@@
 
-- Improperly tuning of models
-  - RM: In a twitter DM, I asked Blair to clarify -- "[I] was wondering how you all came to your tuning procedure for the RF model hyper-parameters?" He responded "mostly trial and error, honestly.
+<!-- - Improperly tuning of models -->
+<!--   - RM: In a twitter DM, I asked Blair to clarify -- "[I] was wondering how you all came to your tuning procedure for the RF model hyper-parameters?" He responded "mostly trial and error, honestly.
 ...
 Trial and error was on early data.
-Forecasts were for much later data"
-- Using a regression framework in a classification problem
-- Rounding of AUC/ROC scores
-    + AB: smoothing or rounding? I think rounding is ok. 
-- Treating the output of a RF regression model as Pr() -- (RM: I guess this is akin to a Linear Probability Model, but I'm not sure if this is possible in a RF framework
-    + It's hacky but it works. The resulting forecasts are within the 0-1 interval.
+Forecasts were for much later data" -->
+<!-- - Using a regression framework in a classification problem -->
+<!-- - smoothing of AUC/ROC scores -->
+<!--     + AB: smoothing or rounding? I think rounding is ok.  -->
+<!-- - Treating the output of a RF regression model as Pr() -- (RM: I guess this is akin to a Linear Probability Model, but I'm not sure if this is possible in a RF framework -->
+<!--     + It's hacky but it works. The resulting forecasts are within the 0-1 interval. -->
 
@@
 
-4. Design choices
+<!-- 4. Design choices -->
 
-Rick, I think the points below are more subjective choices.
+<!-- Rick, I think the points below are more subjective choices.
 The first two points above are IMO potentially objectively incorrect technical errors that could undermine the original Blair and Sambanis findings.
@@
 But the points below I would say are subjective choices that could alter their findings and that I think are justifiable criticisms, but I think it'll be easier for someone to push back on this if they wanted to.
+-->
 
-- Their train/test approach 
-  - 5-year forecasts for sep plots and AUC scores)
-- Smoothing of AUC/ROC scores
-- Lack of yearly test forecasts in favor of a single 5-year test forecast
-- The lack of procedures to account for rare events in an RF model
+<!-- - Their train/test approach  -->
+<!--   - 5-year forecasts for sep plots and AUC scores) -->
+<!-- - Smoothing of AUC/ROC scores -->
+<!-- - Lack of yearly test forecasts in favor of a single 5-year test forecast -->
+<!-- - The lack of procedures to account for rare events in an RF model -->
 
-To-do
+<!-- To-do -->
 
-1. Full replication
-    - Can we get their results
-    - RM: maybe I can go through their code and bring it "up-to-date" i.e. make it tidy... 
-        + AB: that sounds like a lot of work
-    - Figure out how they built table 4 (It looks like this is done in STATA)
-    - Check to see if it matter that they are using regression for a classification problem
-        + AB: i tried switching the models as they have them just from regression to classification and it breaks because the data samples end up being all 0's. 
-2. Address tuning 
-    - Does properly tuning the CAMEO model improve its performance?
-3. Address AUC/ROC smoothing... 
-    - How drastic is the change; does it change their conclusions? 
-4. Run a yearly test forecast -- as is the current standard in TSCS forecasts (Is it the current standard?) 
-5.
+<!-- 1. Full replication -->
+<!--     - Can we get their results -->
+<!--     - RM: maybe I can go through their code and bring it "up-to-date" i.e. make it tidy...  -->
+<!--         + AB: that sounds like a lot of work -->
+<!--     - Figure out how they built table 4 (It looks like this is done in STATA) -->
+<!--     - Check to see if it matter that they are using regression for a classification problem -->
+<!--         + AB: i tried switching the models as they have them just from regression to classification and it breaks because the data samples end up being all 0's.  -->
+<!-- 2. Address tuning  -->
+<!--     - Does properly tuning the CAMEO model improve its performance? -->
+<!-- 3. Address AUC/ROC smoothing...  -->
+<!--     - How drastic is the change; does it change their conclusions?  -->
+<!-- 4. Run a yearly test forecast -- as is the current standard in TSCS forecasts (Is it the current standard?)  -->
+<!-- 5.  -->
 

### 2020-05-13 issues.Rmd

@@
 
-<!-- Below, we work through our issues we identified above, starting first with the unconventional use of a smoother when calculating the AUC-ROC scores.
-We reproduce Table 1 and Table 2 from @blair2020forecasting to show that their results are dependent on the unconventional use of a smoother.
--->
-
-<!-- We then discuss... (RM: I'll fill this out as I work through the evidence.)  -->
-
 ## Smoothed ROC curves
@@
 Given that the DV is civil conflict onset (1/0) and that this is a rare event, these curves should resemble a step-like function.
-Looking through the Blair/Sambanis replication code, we noticed that they use a smoothing function when producing their Receiver Operating Characteristic curves.
+Looking through the Blair/Sambanis replication code, we noticed that they use a smoothing option when producing their Receiver Operating Characteristic curves.
 
@@
 To help make this point clear, we replicate Figure 1 below.
-However, rather than a two panel figure (one ROC plot their one-month-ahead forecasts and one for their six-months-ahead forecasts, as is presented in the article), we now include four panels: one panel for the two forecast windows with the smoothed curve, as they produce, and one panel for the two forecast windows that are not smoothed.
+Rather than a two panel figure (one ROC plot their one-month-ahead forecasts and one for their six-months-ahead forecasts, as is presented in the article), we now include four panels: one panel for the two forecast windows with the smoothed curve, as they produce, and one panel for the two forecast windows that are not smoothed.
 
@@
 
+### Comparison of the escalation specification to alternatives (Table 1)
+
 Using these smoothed and non-smoothed curves, we then replicate Table 1 (page 13), but now the table reports the AUC-ROC scores for each model specification with and without the use of a smoother.
@@
 ```{r, message=FALSE}
-auc <- read_csv("data/table1-redone.csv") %>%
+table1 <- read_csv("data/table1-redone.csv") %>%
   mutate(specification = factor(specification, levels = c("escalation", "quad", "goldstein", "CAMEO", "avg"))) %>%
@@
 
-auc %>%
+table1 %>%
   knitr::kable(digits = 2, booktabs = TRUE, caption = "Comparison of smoothed and not smoothed AUC-ROC values", align = c(rep("l", 2), rep("c", 6))) %>% 
@@
 The current replication project was done using `R version 4.0.0`, which updated their random number seeding procedure, so small differences should be expected.
+(RM: I mention this last part because later when we replicate the PITF table, the results are very far off, beyond what we would expect from random fluctuations)
+
+### Comparison of the escalation model to structural extensions (Table 2)
+
+Table 2 in the B&S paper (p. 19) compares the escalation model, based only on ICEWS-derived indicators, to alternatives that incorporate aspects of a structural model in different ways.
+
+![](figures/bands-table2.png)
+
+On page 20:
+
+> Overall, our results suggest that while measures of structural risk may improve predictive performance, the value they add is marginal and inconsistent.
+Of the approaches we test, the split-population analog is most promising, with AUCs of .84 and .83 over one-month and six-month windows, respectively.
+
+Below is a recreation of this table with AUC values for both smoothed and the original ROC curves.
+
+```{r, message=FALSE}
+escalation_only <- read_csv("data/table1-redone.csv") %>%
+  filter(specification=="escalation") %>%
+  select(-model) %>%
+  mutate(horizon = case_when(horizon=="6 month" ~ "6 months", TRUE ~ horizon))
+
+table2 <- read_csv("data/table2-redone.csv") %>%
+  bind_rows(escalation_only) %>%
+  mutate(
+    specification = factor(
+      specification, 
+      levels = c("escalation", "escalation with PITF", 
+                 "escalation with PITF weights", 
+                 "escalation with Split PITF", "PITF"),
+      labels = c("Escalation\nOnly", "With PITF Predictors", "Weighted by PITF", "PITF Split Population", "PITF Only"))
+  ) %>%
+  arrange(horizon, specification) %>% 
+  pivot_longer(-c(horizon, specification), names_to = "ROC Smoothed") %>% 
+  pivot_wider(names_from = "specification") %>% 
+  mutate(`ROC Smoothed` = ifelse(`ROC Smoothed` == "smoothed", "Yes", "No")) %>%
+  rename(Horizon = horizon)
+
+table2 %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Comparison of smoothed and not smoothed AUC-ROC values", 
+    align = c(rep("l", 2), rep("c", 6))
+  ) %>% 
+  kableExtra::column_spec(c(2, 5, 7), width = "1.5cm") %>%
+  kableExtra::column_spec(3, width = "1.8cm") %>%
+  kableExtra::column_spec(c(4, 6), width = "2.2cm") %>%
+  kableExtra::collapse_rows(1)
+
+```
+
+Note that we were unable to reproduce some of the AUC values in Table 2.
+Values that are notably different are 1-month "Weighted by PITF" and "PITF Split Population", and 6-month "PITF Split Population".
+For all three of these, the AUC-ROC values we obtained are much better than those in the original Table 2.
+The values for 6-month "PITF Only" and "With PITF Predictors" are slightly off but could be rounding errors.
+
+Despite these reproduction differences, the original interpretation mostly still stands.
+The various efforts to incorporate structural variables into the forecasts largely do not work or do not add much predictive value.
+In the 6-month forecasts, the value added by simply adding the annual structural predictors to the existing random forest model is more pronounced than before, while the PITF split population approach fares much worse.
+
+Although the substantive interpretation does not change much as a result of using ROC curve smoothing, the absolute AUC-ROC values between the smoothed and original versions do differe by quite a bit in several instances.
+The AUC-ROC values differ by 0.05 or more in 5 of the 10 cells in the original Table 2, and the direction of change is in all cases such that using smoothed ROC curves present more optimistic performance values.
 
@@
 
-The table presents 15 or 16 positive cases for 2016-H1, depending on a minor coding decision. These numbers correspond to a positive rate of around 9.5% for the first half of 2016.
-
-By comparison, the 6-month version of the full data from 2001 to 2015 contains only 20 positive cases, with a positive rate of around 0.5%.
+The table presents 15 or 16 positive cases for 2016-H1, depending on a minor coding decision.
+This is the difference between the top and bottom confusion matrices.
+From the confusion matrices we can infer that the test data had a reported positive rate of around 9.5% for the first half of 2016.
+In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a positive rate of around 0.5%.
+This suggests that the civil war onset forecasts are erranously assessed using civil war incidence, not onset.
 
@@
 
-The main data set covering 2001 - 2015 contains 20 civil war onsets. But there are 15 "Persistence" cases (top of table 4) and 16 "Change" cases (bottom of table 4)
-
 By examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
@@
 
-The actual data contains 0 new civil war onsets in 2016-H1.
-Using those data, the 2016-H1 forecasts would have retained a recall value of 1, but precision would have decreased from 13/30 = 0.76 to 1/30 = 0.03.
+The correct confusion matrices when using observed onset (or the lack of it) are shown below.
 
-The second part of the Table 4 script manually codes onset in two countries, Turkey and Burundi.
-With this coding, the correct recall (bottom of Table 4) is 1 (not 14/16 = 0.88) and the precision changes from 14/30 = 0.47 to 2/30 = 0.07.
+```{r, message=FALSE}
+tab4 <- read_csv("data/tab4-fixed.csv")
+
+tab4 %>%
+  knitr::kable(
+    caption = "2016 Confusion Matrices for Six-month Escalation Model.",
+    booktabs = TRUE
+  ) %>%
+  kableExtra::collapse_rows(1)
+```
+
+There are no civil war onsets in the data for 2016-H1.
+Thus the recall values is undefined, while the precision is 0/30 = 0, compared to recall and precision values of 13/15 = 0.87 and 13/30 = 0.43 before.
+The the alternative coding ("Assuming Change") at the bottom of Table 4, there are 2 civil war onsets.
+Recall is 1 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.
 
@@
 
-## Random forest model tuning
+## Discussion
 
-3.
-Unorthodox RF hyperparameter choices that may be driving the OOS test prediction results.
-There are two issues here.
-One is that the RF models are so unorthodox.
-I'm not sure that this in itself is a problem.
-But the second potential problem is that the hyperparameter values may not have been tuned in a way that works the best for the non-escalation model, and especially the CAMEO model with 1,100+ features.
+When using the standard method of calculating AUC-ROC values, i.e.
+without smoothing the ROC curve beforehand, one of the principal B&S results reverses.
+The theory-driven escalation model does not in fact outperform the models that use alternative covariate specifications.
+In both the 1-month and 6-month versions of the base specification, all four other alternatives achieve equal or better AUC-ROC values compared to the escalation model.
+This includes the 1,159 covariate kitchen-sink CAMEO model.
 
-<!-- - Improperly tuning of models -->
-<!--   - RM: In a twitter DM, I asked Blair to clarify -- "[I] was wondering how you all came to your tuning procedure for the RF model hyper-parameters?" He responded "mostly trial and error, honestly.
-...
-Trial and error was on early data.
-Forecasts were for much later data" -->
-<!-- - Using a regression framework in a classification problem -->
-<!-- - smoothing of AUC/ROC scores -->
-<!--     + AB: smoothing or rounding? I think rounding is ok.  -->
-<!-- - Treating the output of a RF regression model as Pr() -- (RM: I guess this is akin to a Linear Probability Model, but I'm not sure if this is possible in a RF framework -->
-<!--     + It's hacky but it works. The resulting forecasts are within the 0-1 interval. -->
+We find no general change in the results regarding whether adding structural factors to the escalation model's forecasts improves performance, although the gain from incorporating the PITF structural variables into the 6-month escalation model is much more pronounced.
+There is no reason to rule out the possibility that the escalation model alternatives would not also similarly gain from incorporating structural variables in their 6-month versions.
+
+Furthermore, it appears that the forecasts for the risk of civil war onset in the first half of 2016 were erroneously evaluated using *incidence*, not *onset*.
+When correcting this mistake, the forecast evaluation changes significantly.
+There were no onsets in 2016, resulting in a precision of 0 and undefined recall; while with the manual re-coding reflected in the bottom portion of Table 4, the forecasts capture both civil war onsets for a recall of 1 and precision of 0.07.
+
+Both of these issues are relatively objective problems.
+It seems relatively clear to us that using incidence to assess onset forecasts is simply incorrect.
+Using smoothed ROC curves to calculate AUC is non-standard to the extent of our knowledge, and the decision to use them is not discussed or justified in the paper.
+Maybe it is possible to defend this choice, but given that one of the principal findings rests on it, this should be done explicitly.
+
+There are also other, more subjective concerns that we did not raise.
+For example, the paper does not explain how the random forest model hyperparemeters were selected.
+Although they do evaluate several changes in the hyperparameters as part of the robustness tests, the overall choice of hyperparameter values is unusual.
+Of note is also that this unusual approach only works (with warnings issued by the software) because B&S use random forests based on regression, not decision, trees.
+The concern to rule out would be that the hyperparameters are (over-) fit on the test data and that the reported performance values do not generalize.[^On page 10, B&S write that they use classification trees; this is strictly speaking incorrect and the random forest is built using regression tress on the 0 or 1 outcome.]
+
+We do not disagree with the overall point that theory is important for recognizing which variables or indicators might be important for forecasting.
+But it does not appear that B&S analysis demonstrates this, and that the paper's conclusions to this effect rest on faulty results.
+
+<!--
 
@@
 
+-->
+
 <!-- 4. Design choices -->
@@
 
+-->
+
 # References

### 2020-05-21 issues.Rmd

@@
 
-We find three main issues in their empirical analyses:
+We find that:
 
-- Their primary finding -- that the escalation model they advance consistently out-performs the alternation models -- depends entirely on the unorthodox use of smoothed ROC curves when calculating the AUC-ROC values in Table 1 and Table 2.
-When using the standard (not smoothed) ROC curves, this finding is reversed.
-That is, the escalation performs no better and often worse than the other models.
-- Regarding the forecast assessment in Table 4, the forecasts for 2016-H1 are erroneously assessed using civil war _incidence_, not _onset_, even though the models from which they are generated are clearly fitted to civil conflict onset.
-This vastly overstates the precision of the forecasts.
-- The way the random forest models are fitted is very unorthodox.^[This is in fact what first led us to examine the analysis in more detail.] The number of tree the model grow, the number of nodes in these trees, and the sample size used for each tree deviates drastically for traditional approaches (e.g., they set the number of trees to 100,000 where the default is 500).
-
-We demonstrate the first point through model replication: we use the replication files that Blair and Sambanis provide to calculate the AUC-ROC scores using smoothed ROC curves (the published results) and then calculate the AUC-ROC scores using the standard approach -- without the unorthodox smoother.^[In this draft, we focus our replication efforts on what Blair and Sambanis call their "Base specification".
-This set of models are what Blair and Sambanis use when discussing their main findings and the implications of their results.] We demonstrate the second point using the replicated results from the original model Blair and Sambanis use to create Table 4 and then compare these results to the observed outcomes in the 2016 data.
-
-For the final point, we evaluate two questions: (1) even if we allow for the use of smoothed ROC curves, does the key model's superior test forecast performance generalize or is it a result of inadvertent fitting of the model hyperparameters to test forecast performance, and (2) does the inferior performance of the alternative models, and especially the CAMEO model with 1,159 covariates, reflect a lack of proper hyperparameter tuning for these models?
-(work still in progress).
+- the first key finding depends on using AUC values calculated for *smoothed*, not the original, ROC curves
+- the 2016 forecasts are incorrectly assessed using the occurrence (incidence), not onset, of civil war
 
@@
 
-The results for this analysis are summarized in Table 1 in @blair2020forecasting, which we copied in here:
+# Smoothed ROC curves
 
-![](figures/bands-table1.png)
+## Comparison of the escalation specification to alternatives (Table 1)
+
+The results for the core analysis are summarized in Table 1 in @blair2020forecasting, which we copied in Figure 1 below.
+
+![Copy of B&S Table 1](figures/bands-table1.png)
+
+The first row, each, for the one-month and 6-month forecasts are the core base specification. The additional rows reflect various robustness tests.
 
@@
 
-After establishing that the escalation model they advance out-performs all other covariate specifications, Blair and Sambanis move to show how the inclusion of structural risk variables -- the set of structural variables identified by the Political Instability Task Force (PITF) that can contribute to an increased risk of civil conflict.
-The figure below is Table 2 from @blair2020forecasting, which presents these findings.
+A small detail, which is mentioned in the text and visually obvious in the ROC curves they show in B&S Figure 1, is that these AUC-ROC values are computed using smoothed ROC curves.
+We used their replication code to recreate the ROC curves in their figure, which are shown in the left two panels in the figure below.
+The curves indeed appear to be very smooth.
+But in the underlying test data for which these curves are, there are only 11 positive events, out of 2,366 rows for the 6-month data and 13,748 rows for the 1-month data.
+The ROC curves in the two panels on the right are the original, non-smoothed versions, which correctly reflect the small number of positive events.^[The diagonal lines in the ROC curves are there because there are groups of cases which have identical predictions, causing jumps in both the true and false positive rates at the same time.]
 
-![](figures/bands-table2.png)
-
-# Issues
-
-## Smoothed ROC curves
-
-One issue that stood out as we read @blair2020forecasting was how smooth their Receiver Operating Characteristic curves where in Figure 1 on page 14.
-A Receiver Operating Characteristic curve displays the balance between the true positive rate and the false positive rate, across a range of discrimination thresholds -- the value in which a zero to one probability is assumed to be an "onset".
-Given that the DV is civil conflict onset (1/0) and that this is a rare event, these curves should resemble a step-like function.
-Looking through the Blair/Sambanis replication code, we noticed that they use a smoothing option when producing their Receiver Operating Characteristic curves.
-
-This is important because Blair and Sambanis use a common performance metric -- Area Under the Receiver Operating Characteristic Curves (AUC-ROC) -- as evidence that their escalation model out preforms all others.
-However, by smoothing the Receiver Operating Characteristic Curves, they are misrepresenting the true AUC-ROC scores of their various models.
-To help make this point clear, we replicate Figure 1 below.
-Rather than a two panel figure (one ROC plot their one-month-ahead forecasts and one for their six-months-ahead forecasts, as is presented in the article), we now include four panels: one panel for the two forecast windows with the smoothed curve, as they produce, and one panel for the two forecast windows that are not smoothed.
-
-![](figures/figure1-replicated.png)
+![Replication of B&S Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)
 
@@
 
-### Comparison of the escalation specification to alternatives (Table 1)
-
-Using these smoothed and non-smoothed curves, we then replicate Table 1 (page 13), but now the table reports the AUC-ROC scores for each model specification with and without the use of a smoother.
-If the non-smoothed AUC-ROC scores are drastically different from those of the smoothed AUC-ROC score (a difference of $\ge 0.05$) or if the non-smoothed AUC-ROC scores are < 0.8, would suggest the findings presented in @blair2020forecasting are fragile to the use of a non-standard procedure.
+Using these smoothed and non-smoothed curves, we then replicate B&S Table 1 (page 13), but now the table reports the AUC-ROC scores for each model specification with and without the use of a smoother.
+We focus on the core "Base specification" models only, not the additional robustness checks shown in the other rows in the table.
+The results of this are shown in our Table 1 above.
 
@@
 
-As the table above shows, the results presented in @blair2020forecasting are fragile to the use of a smoothing option when calculating the AUC-ROC performance metric.
-Further, even when we rely solely on the replication files provided by Blair and Sambanis, we cannot replicate their findings exactly.
-For example, in our replication, the AUC-ROC score for the 1-month CAMEO model with the use of the smoother is 0.84; however, in the paper, the reported score is 0.82.
+First, note that we were unable to exactly reproduce the AUC-ROC values for all models.
+For example, in B&S the 1-month CAMEO has a AUC-ROC of 0.82, while our reproduction has a AUC-ROC value (using the smoothed version) of 0.84.
 This is most likely a function of the version of `R` Blair and Sambanis used to conduct their analysis.
 The current replication project was done using `R version 4.0.0`, which updated their random number seeding procedure, so small differences should be expected.
-(RM: I mention this last part because later when we replicate the PITF table, the results are very far off, beyond what we would expect from random fluctuations)
+Despite these small differences, the way we would interpret the smoothed-version results does not change.
+In the 1-month version of the Base Specification, the Escalation model outperforms all alternatives; in the 6-month version it outperforms all but one alternative and has equal performance with the Goldstein model.
 
-### Comparison of the escalation model to structural extensions (Table 2)
+When we calculate the AUC-ROC values using the original, non-smoothed curves, the findings change substantially.
+Rows 2 and 4 in Table 1 show the non-smoothed AUC-ROC values for the 1-month and 6-month versions, respectively.
+In the 1-month version, the escalation model is outperformed by 2 alternatives and even with the other 2.
+In the 6-month version, all 4 alternatives outperform the escalation model.
+When using the original, non-smoothed ROC curves, the escalation model has the worst overall performance out of the 5 models considered.
+The first key finding in the paper, that the escalation model is superior to the alternative models, thus depends on using smoothed ROC curves when calculating the AUC-ROC statistic.
 
-Table 2 in the B&S paper (p. 19) compares the escalation model, based only on ICEWS-derived indicators, to alternatives that incorporate aspects of a structural model in different ways.
+\newpage
 
-![](figures/bands-table2.png)
+## Comparison of the escalation model to structural extensions (Table 2)
+
+After establishing that the escalation model they advance out-performs all other covariate specifications, Blair and Sambanis move to show how the inclusion of structural risk variables -- the set of structural variables identified by the Political Instability Task Force (PITF) that can contribute to an increased risk of civil conflict.
+A copy of B&S Table 2 is included in Figure 3.
+
+![Copy of B&S Table 2](figures/bands-table2.png)
 
@@
 
-Below is a recreation of this table with AUC values for both smoothed and the original ROC curves.
+Table 2 is our recreation of B&S Table 2, copied in Figure 3, with AUC values for both smoothed and the original ROC curves.
+
+\newpage
 
@@
     digits = 2, booktabs = TRUE, 
-    caption = "Comparison of smoothed and not smoothed AUC-ROC values", 
+    caption = "Comparison of smoothed and not smoothed AUC-ROC values for B\\&S Table 2", 
     align = c(rep("l", 2), rep("c", 6))
@@
 
-Note that we were unable to reproduce some of the AUC values in Table 2.
+Note that we were again unable to reproduce some of the AUC values in Table 2.
 Values that are notably different are 1-month "Weighted by PITF" and "PITF Split Population", and 6-month "PITF Split Population".
@@
 
-## Assessment of the 2016 forecasts
+# Assessment of the 2016 forecasts
 
-The predictions for the first half of 2016 are assessed in Table 4:
+The predictions for the first half of 2016 are assessed in B&S Table 4:
 
@@
 In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a positive rate of around 0.5%.
-This suggests that the civil war onset forecasts are erranously assessed using civil war incidence, not onset.
+Since the positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data, we suspected that the forecasts were erranously assessed using civil war incidence, not onset.
 
@@
 
-## Discussion
+# Discussion
 
@@
 We find no general change in the results regarding whether adding structural factors to the escalation model's forecasts improves performance, although the gain from incorporating the PITF structural variables into the 6-month escalation model is much more pronounced.
-There is no reason to rule out the possibility that the escalation model alternatives would not also similarly gain from incorporating structural variables in their 6-month versions.
+There is also no reason to rule out the possibility that the escalation model alternatives would not also similarly gain from incorporating structural variables in their 6-month versions.
 
@@
 Of note is also that this unusual approach only works (with warnings issued by the software) because B&S use random forests based on regression, not decision, trees.
-The concern to rule out would be that the hyperparameters are (over-) fit on the test data and that the reported performance values do not generalize.[^On page 10, B&S write that they use classification trees; this is strictly speaking incorrect and the random forest is built using regression tress on the 0 or 1 outcome.]
+The concern to rule out would be that the hyperparameters are (over-) fit on the test data and that the reported performance values do not generalize.^[On page 10, B&S write that they use classification trees; this is strictly speaking incorrect and the random forest is built using regression tress on the 0 or 1 outcome.]
 
 We do not disagree with the overall point that theory is important for recognizing which variables or indicators might be important for forecasting.
-But it does not appear that B&S analysis demonstrates this, and that the paper's conclusions to this effect rest on faulty results.
-
-<!--
-
-The table below shows the randomForest package default hyperparameter values for a binary classification problem like the one at hand and compares them to the Blair and Sambanis base settings.
-Some hyperparameters have heuristics that determine the default value based on characteristics of the input data; we note these in the second column and what the realized default setting would be for the basic Escalation model (first row in Table 1 in the paper) in the 3rd column.
-The last column has the settings Blair and Sambanis use.
-
-|Hyperparameter|Default heuristic|Default values (Escalation)|B&S value|
-|--------------|-----------------|--------------------------|---------|
-|type          |                 | classification           | regression|
-|ntree         |                 | 500                      | 100,000 |
-|mtry          |`floor(sqrt(ncol(x)))`|3|3|
-|replace       |                 | true | false |
-|sampsize | `nrow(x)` if replace, else `ceiling(.632*nrow(x))` | 11,869 | 100 |
-|nodesize | 1 for classification | 1 | 1 |
-|maxnodes | | null | 5 |
-
-It is worth noting that commonly the ntree, mtry, and nodesize parameters are the main parameters tuned; of these three in the Blair and Sambanis specification only 1, the ntree parameter, deviates from the default settings.
-
-In any case, there is a stark contrast in the default RF settings and the way Blair and Sambanis use the RF models.
-The default approach is to train a relatively small number (ntree; 500) of classification trees, but where each tree is fairly big in that it is trained on data that has the same number of rows as the training data, albeit sampled with replacement (replace is true; sampsize is 11,869), and allowed to grow fairly deep (this is governed by nodesize (1), which is the minimum size a terminal node must have).
-In contrast, Blair and Sambanis grow very extensive forests with a large number of trees (100,000 compared to 500), but each tree is very small and shallow; only a 100 rows are sampled from the training data for each tree, and the trees are constrained to at most 5 terminal nodes (maxnodes).
-
-This approach only works due to the other unorthodox choice, which is to use regression, not classification, trees.
-Trying to use classification trees with the other parameter settings in fact does not work at all because it is almost guaranteed that a sample of 100 from the 11,869 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample.
-As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:
-
-```
-Warning message:
-In randomForest.default(y = as.integer(train_df$incidence_civil_ns_plus1 ==  :
-  The response has five or fewer unique values.  Are you sure you want to do regression?
-```
-
-While unorthodox, the approach does work.
-It produces predictions that are within the 0 to 1 interval.
-The concern is that the settings work well only for the escalation model, and only for the particular test set at hand.
-Specifically, we wonder (1) if the hyperparameter settings only work well for the particular test set chosen, and (2) if the CAMEO model with 1,100+ features, compared to 10 or less for the other specifications, would perform better with a more explicit tuning procedure.
-
-## Are the RF hyperparameters overfit to the test set?
-
-Does the base specification's good performance in the test set generalize?
-
-One initial piece of evidence is already available from Table 1.
-In addition to the test set starting 2008, Blair and Sambanis also evaluate test sets starting in 2009, 2010, and 2011 (sets 1, 2, and 3 in Table 1).
-For the 1 month models, when we compare the base specification performance of each model to the performance in the alternate test sets, the AUC-ROC increases in 2 (of 15) cases, is the same in 0 cases, and decreases in 13 cases.
-For the 6 month models, it increases in 1 case, is the same in 3 cases, and decreases in 11 cases.
-Altering the test set thus generally shows reduced performance.
-The table below shows cross-validated training and OOS test AUC-ROC for the original 1 month escalation model, a RF model with default hyperparameter settings, and a tuned RF model with 10,000 trees and otherwise the default settings.
-The first set of results are out of sample results from repeated cross-validations performed on the training data.
-We show the average AUC-ROC, it's lower and upper 95\% CI, obtained via bootstrapping, and the standard deviation of the distribution of resampled AUC-ROC values.
-
-Because there are only 9 positive cases in the whole training data, and thus to ensure that any given split will include at least 1 positive case in each data partition, we use 2-fold CV, i.e.
-splitting the original training data into equally-sized new training and validation sets.
-This is repeated 21 times for a total of 42 OOS performance samples for each model.
-The last column shows the OOS performance on the original test set.
-The first value in this column corresponds to the base escalation model results reported Table 1 in the Blair and Sambanis paper when *not* using smoothed ROC curves.
-
-```{r}
-tbl <- structure(list(Model = c("Escalation, 1mo", "Modified Escalation, 1mo", 
-"Tuned Escalation, 1mo"), Avg_CV_ROC_AUC = c(0.677707841149067, 
-0.619719616721681, 0.671630392456263), ci_lower = c(0.634101138978023, 
-0.5835673668958, 0.634713167892701), ci_upper = c(0.722598486545989, 
-0.65584632967561, 0.706310478133188), SD_CV_ROC_AUC = c(0.147542574525528, 
-0.122337186539073, 0.117973980721395), Test_ROC_AUC = c(0.783352194140576, 
-0.586147564308735, 0.582927991423296)), row.names = c(NA, -3L
-), class = c("tbl_df", "tbl", "data.frame"))
-
-knitr::kable(tbl, digits = 2)
-```
-
-The results show that:
-
-- Both of the alternative RF models are able to achieve roughly similar OOS performance in the training data split.
-*T*-tests comparing model 1 to the other models fail to reject the null hypothesis at a 95\% confidence level.
-(A *t*-test comparing model 1 and 2 average AUC-ROC just barely fails to reject the null hypothesis (*p* slightly above 0.05).)
-- Only the base escalation model is able to achieve good test performance; the other two models, despite achieving similar training data performance, have significantly lower test performance.
-
-This suggests that the base RF specifications are (over-)fit to the test data.
-
--->
-
-<!-- 4. Design choices -->
-
-<!-- Rick, I think the points below are more subjective choices.
-The first two points above are IMO potentially objectively incorrect technical errors that could undermine the original Blair and Sambanis findings.
-The 3rd point about the RF models is somewhere between objective error and subjective choice.
-I'm not sure.
-But the points below I would say are subjective choices that could alter their findings and that I think are justifiable criticisms, but I think it'll be easier for someone to push back on this if they wanted to.
--->
-
-<!-- - Their train/test approach  -->
-<!--   - 5-year forecasts for sep plots and AUC scores) -->
-<!-- - Smoothing of AUC/ROC scores -->
-<!-- - Lack of yearly test forecasts in favor of a single 5-year test forecast -->
-<!-- - The lack of procedures to account for rare events in an RF model -->
-
-<!-- To-do -->
-
-<!-- 1. Full replication -->
-<!--     - Can we get their results -->
-<!--     - RM: maybe I can go through their code and bring it "up-to-date" i.e. make it tidy...  -->
-<!--         + AB: that sounds like a lot of work -->
-<!--     - Figure out how they built table 4 (It looks like this is done in STATA) -->
-<!--     - Check to see if it matter that they are using regression for a classification problem -->
-<!--         + AB: i tried switching the models as they have them just from regression to classification and it breaks because the data samples end up being all 0's.  -->
-<!-- 2. Address tuning  -->
-<!--     - Does properly tuning the CAMEO model improve its performance? -->
-<!-- 3. Address AUC/ROC smoothing...  -->
-<!--     - How drastic is the change; does it change their conclusions?  -->
-<!-- 4. Run a yearly test forecast -- as is the current standard in TSCS forecasts (Is it the current standard?)  -->
-<!-- 5.  -->
-
--->
+But it does not appear that B&S analysis demonstrates this, and that the paper's conclusions to this effect rest on faulty analysis.
 

### 2020-05-29 rep_nosmooth/README.Rmd
refactor replication code (not done)
The RF models now can be run in parallel. Still need to handle the special models, like the average model in Table 1.

---
title: "Modified replication"
output: 
  github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE)

library(dplyr)
library(readr)
```

The code in this folder replicates the results in Tables 1 and 2 of B&S, but does so with both the smoothed ROC curve AUC values B&S report, and regular, non-smoothed ROC curve AUC.

In order to be able to run the models needed to create tables 1 and 2 more quickly than with the original replication code, the code here has been refactored to enable running in parallel. Here is a summary of the major changes:

- `re-save-data.R` changes the format of the saved 1-month and 6-month data versions from Stata 13 .dta files to R .rds format. This can be read much quicker than the Stata format. 
- `setup-model-table.R` encodes all the various pieces of information needed to actually run the models reflected in tables 1 and 2. In the original replication files this is done in bits and pieces in `+master.R`, `code/[1|6]mo_define_models.R`, and each model runner file, e.g. `code/1mo_run_escalation.R`. The primary output are several files written to the `output/model-definitions` folder. More details on this below.
- `model-runner.R` runs the models. Adjust the number of WORKERS at the top of the script as needed based on the number of cores on your computer. The primary output of this script is `output/model-table-w-results.csv`
- `recreate-tables.R` recreates the original and non-smooth ROC versions of Tables 1 and 2.

### Model definitions

The original replication code has a large number of scripts that define the various models/methods to run. In order to allow running the RF models in parallel, this refactored code sets up a table and other lists that define all the information needed to run the various models, and then runs through the table (in parallel) to run each model.

The various models going into tables 1 and 2 vary over several dimensions. We started untangling this by first encoding every AUC-ROC value/cell in Tables 1 and 2. These can be uniquely identified by the forecast horizon (1 or 6-months), and then the table row and columns. There are a total of 100 AUC-ROC values reported, 90 in Table 1, and 10 in Table 2.

_(The "Escalation only" values in Table 2 actuall correspond to the models that produce the \["Base specification", "Escalation"\] values in Table 1, but it is easier to duplicate them since the underlying models in any case estimate pretty quickly.)_

One of the principal outputs of `setup-model-definitions.R` is this encoding, in `output/model-definitions/model-table.rds`:

```{r, echo = TRUE}
model_table <- read_rds("output/model-definitions/model-table.rds")
head(model_table)
```

The majority of cells correspond to underlying random forest (RF) models--78 of 100 entries. For these RF cells, the table columns correspond to the feature specification going into the model as predictors. This mapping is encoded in `output/model-definitions/feature-specs.json`.

Forecasting horizon aside, the rows encode three dimensions of info:

- the DV to use (`dv-specs.yaml`), specifically either the regular one or one of the two alternate codings
- the train/test split to use (`train-end-year.yaml`), one of 4 possible values
- the RF hyperparameter settings to use (`hp-settings.yaml`).

Together these bits define all that we need to run the correct random forest models for these 78 cells.

The other 22 cells mostly involve averaging predictions produced by the other models. For example, the "Average" column in Table 1 is for a model that averages the predictions from the other four models(/specifications). This doesn't require running any model per se; just averaging existing predictions. These are handled via special treatment in `model-runner.R`, outside the main parallel loop.

## Table 1

### Original, smoothed ROC curves

```{r}
table1_smooth <- read_csv("output/table1-smooth.csv")

table1_smooth %>%
  knitr::kable(digits = 2,
               caption = "Alternative Table 1 with non-smoothed AUC-ROC")
```

### With conventional ROC curves

```{r}
table1_nosmooth <- read_csv("output/table1-nosmooth.csv")
table1_nosmooth %>%
  knitr::kable(digits = 2,
               caption = "Alternative Table 1 with non-smoothed AUC-ROC")
```

### 2020-05-29 rep_nosmooth/code/1mo_define_models.R

@@ -1,2585 +1,2575 @@
 
 
-  # Define model labels
+# Define model labels
 
-    			escalation <- c(
-    				"gov_opp_low_level",
-    				"gov_reb_low_level",
-    				"opp_gov_low_level",
-    				"reb_gov_low_level",
-    				"gov_opp_nonviol_repression",
-    				"gov_reb_nonviol_repression",
-    				"gov_opp_accommodations",
-    				"gov_reb_accommodations",
-    				"reb_gov_demands",
-    				"opp_gov_demands"
-    				)
+escalation <- c(
+  "gov_opp_low_level",
+  "gov_reb_low_level",
+  "opp_gov_low_level",
+  "reb_gov_low_level",
+  "gov_opp_nonviol_repression",
+  "gov_reb_nonviol_repression",
+  "gov_opp_accommodations",
+  "gov_reb_accommodations",
+  "reb_gov_demands",
+  "opp_gov_demands"
+)
 
-    			quad <- c(
-      			"gov_opp_vercf","gov_reb_vercf",
-    				"gov_opp_matcf","gov_reb_matcf",
-    				"opp_gov_vercf","reb_gov_vercf",
-    				"opp_gov_matcf","reb_gov_matcf",
-    				"gov_opp_vercp","gov_reb_vercp",
-    				"gov_opp_matcp","gov_reb_matcp",
-    				"opp_gov_vercp","reb_gov_vercp",
-    				"opp_gov_matcp","reb_gov_matcp"
-    				)
+quad <- c(
+  "gov_opp_vercf","gov_reb_vercf",
+  "gov_opp_matcf","gov_reb_matcf",
+  "opp_gov_vercf","reb_gov_vercf",
+  "opp_gov_matcf","reb_gov_matcf",
+  "gov_opp_vercp","gov_reb_vercp",
+  "gov_opp_matcp","gov_reb_matcp",
+  "opp_gov_vercp","reb_gov_vercp",
+  "opp_gov_matcp","reb_gov_matcp"
+)
 
-        	goldstein <- c(
-      			"gov_opp_gold","gov_reb_gold",
-    				"opp_gov_gold","reb_gov_gold"
-            )
+goldstein <- c(
+  "gov_opp_gold","gov_reb_gold",
+  "opp_gov_gold","reb_gov_gold"
+)
 
-      		all_CAMEO <- c(
-      			"gov_opp_cameo_010",
-    				"gov_opp_cameo_011",
[…]

### 2020-05-29 rep_nosmooth/code/1mo_run_escalation.R

@@ -9,69 +9,69 @@ train_model <- randomForest(escalation_train_formula_civil_ns,
                             replace = FALSE, do.trace = FALSE ,
                             data = escalation_train_frame_civil_ns, forest = TRUE)
 
-#   			train_model_robust_maxnodes <- randomForest(escalation_train_formula_civil_ns,
-# 												  type = regression, importance = TRUE,
-# 												  na.action = na.omit, ntree = 100000, maxnodes = 10,
-# 												  sampsize = 100,
-# 												  replace = FALSE, do.trace = FALSE ,
-# 												  data = escalation_train_frame_civil_ns, forest = TRUE
-# 												  )
-#
-#     		train_model_robust_sampsize <- randomForest(escalation_train_formula_civil_ns,
-# 												  type = regression, importance = TRUE,
-# 												  na.action = na.omit, ntree = 100000, maxnodes = 5,
-# 												  sampsize = 500,
-# 												  replace = FALSE, do.trace = FALSE ,
-# 												  data = escalation_train_frame_civil_ns, forest = TRUE
-# 												  )
-#
-#   			train_model_robust_ntree <- randomForest(escalation_train_formula_civil_ns,
-# 												  type = regression, importance = TRUE,
-# 												  na.action = na.omit, ntree = 1000000, maxnodes = 5,
-# 												  sampsize = 100,
-# 												  replace = FALSE, do.trace = FALSE ,
-# 												  data = escalation_train_frame_civil_ns, forest = TRUE
-# 												  )
-#
-#     		train_model_robust_traintest1 <- randomForest(escalation_train_formula_civil_ns_robust_traintest1,
-# 												  type = regression, importance = TRUE,
-# 												  na.action = na.omit, ntree = 100000, maxnodes = 5,
-# 												  sampsize = 100,
-# 												  replace = FALSE, do.trace = FALSE ,
-# 												  data = escalation_train_frame_civil_ns_robust_traintest1, forest = TRUE
-# 												  )
-#
-#     		train_model_robust_traintest2 <- randomForest(escalation_train_formula_civil_ns_robust_traintest2,
-#     		                  type = regression, importance = TRUE,
-#     		                  na.action = na.omit, ntree = 100000, maxnodes = 5,
-#     		                  sampsize = 100,
-#     		                  replace = FALSE, do.trace = FALSE ,
-#     		                  data = escalation_train_frame_civil_ns_robust_traintest2, forest = TRUE
-#     		                  )
-#
[…]

### 2020-05-29 rep_nosmooth/master.R

@@ -4,17 +4,13 @@ set.seed(92382)
 
 # Clear environment
 
-# rm(list = ls())
+rm(list = ls())
 
 # Set working directory
-
-## RM: ADDED 4/29/2020
-library(here) ##
 
 setwd(here::here("rep_nosmooth"))
 
 # Load packages
-
 
 library(foreign)
 library(separationplot)
@@ -53,23 +49,23 @@ end_period = mean(data$period[which(data$month==12 & data$year==2015)])
 train <- data[data$period<=train_period,]
 test <- data[data$period>train_period & data$period<=end_period,]
 
-# # Define training and testing sets for robustness checks using alternate start dates
-#
-#       train_period_robust_traintest1 = mean(data$period[which(data$month==12 & data$year==2008)])
-#       end_period = mean(data$period[which(data$month==12 & data$year==2015)])
-#       train_robust_traintest1 <- data[data$period<=train_period_robust_traintest1,]
-#       test_robust_traintest1 <- data[data$period>train_period_robust_traintest1 & data$period<=end_period,]
-#
-#       train_period_robust_traintest2 = mean(data$period[which(data$month==12 & data$year==2009)])
-#       end_period = mean(data$period[which(data$month==12 & data$year==2015)])
-#       train_robust_traintest2 <- data[data$period<=train_period_robust_traintest2,]
-#       test_robust_traintest2 <- data[data$period>train_period_robust_traintest2 & data$period<=end_period,]
-#
-#       train_period_robust_traintest3 = mean(data$period[which(data$month==12 & data$year==2010)])
-#       end_period = mean(data$period[which(data$month==12 & data$year==2015)])
-#       train_robust_traintest3 <- data[data$period<=train_period_robust_traintest3,]
-#       test_robust_traintest3 <- data[data$period>train_period_robust_traintest3 & data$period<=end_period,]
-#
+# Define training and testing sets for robustness checks using alternate start dates
+
+      train_period_robust_traintest1 = mean(data$period[which(data$month==12 & data$year==2008)])
+      end_period = mean(data$period[which(data$month==12 & data$year==2015)])
+      train_robust_traintest1 <- data[data$period<=train_period_robust_traintest1,]
[…]

### 2020-05-29 rep_nosmooth/model-runner.R

@@ -1,0 +1,144 @@
+#
+#   Run all the models behind Tables 1 and 2
+#
+
+WORKERS <- 7
+
+setwd(here::here("rep_nosmooth"))
+
+library(dplyr)
+library(readr)
+library(randomForest)
+library(yaml)
+library(jsonlite)
+library(pROC)
+library(foreach)
+library(doFuture)
+
+dir.create("output/predictions", recursive = TRUE, showWarnings = FALSE)
+
+registerDoFuture()
+plan("multisession", workers = WORKERS)
+
+# Function to wrap the AUC-ROC calculation
+auc_roc_vec <- function(pred, truth, smooth) {
+  roc_obj <- pROC::roc(truth, pred, auc = TRUE, quiet = TRUE, smooth = smooth)
+  as.numeric(roc_obj$auc)
+}
+
+# Load the model definition data
+
+model_table    <- read_rds("output/model-definitions/model-table.rds")
+feature_specs  <- read_json("output/model-definitions/feature-specs.json",
+                            simplifyVector = TRUE)
+dv_specs       <- read_yaml("output/model-definitions/dv-specs.yaml")
+hp_settings    <- read_yaml("output/model-definitions/hp-settings.yaml")
+train_end_year <- read_yaml("output/model-definitions/train-end-year.yaml")
+
+# Load the data
+data_1mo <- read_rds("trafo-data/1mo_data.rds")
+data_6mo <- read_rds("trafo-data/6mo_data.rds")
+
+# Split out the non-RF models that need special treatment
+non_rf_models <- model_table %>%
+  filter(non_RF==TRUE)
+rf_model_table <- model_table %>%
+  filter(!non_RF)
+
+# Shuffle the model table so the workload per worker is on average more evenly
+# distributed. Otherwise the script will take longer to run because it's waiting
+# on the last worker that ended up with the hardest workload to finish.
+rf_model_table <- rf_model_table[sample(1:nrow(rf_model_table)), ]
+
+rf_model_table <- foreach(
+  i = 1:nrow(rf_model_table),
+  .inorder = FALSE
+) %dopar% {
+
+  # keep track of run time
+  t0 <- proc.time()
+
+  horizon_i <- rf_model_table[i, ][["horizon"]]
+  row_i     <- rf_model_table[i, ][["row"]]
+  col_i     <- rf_model_table[i, ][["column"]]
+
+  if (horizon_i=="1 month") {
[…]

### 2020-05-29 rep_nosmooth/re-save-data.R

@@ -1,0 +1,18 @@
+#
+#   Re-save the replication data as RDS; the .dta versions take very long to
+#   read.
+#
+
+library(readstata13)
+library(readr)
+library(here)
+
+setwd(here::here("rep_nosmooth"))
+
+dir.create("trafo-data", showWarnings = FALSE)
+
+data_1mo <- read.dta13("data/1mo_data.dta")
+write_rds(data_1mo, "trafo-data/1mo_data.rds")
+
+data_6mo <- read.dta13("data/6mo_data.dta")
+write_rds(data_6mo, "trafo-data/6mo_data.rds")

### 2020-05-29 rep_nosmooth/recreate-table1.R

@@ -1,61 +1,0 @@
-#
-#   Recreate table 1 with non-smoothed ROC
-#
-# install.packages("")
-library(readr)
-library(kableExtra)
-library(knitr)
-library(dplyr)
-library(here)
-
-setwd(here::here("MB-replication-checks"))
-
-# Write table summaries for monitoring on git
-
-auc <- lapply(dir("tables", pattern = "auc-", full.names = TRUE), read_csv,
-              col_types = cols(
-                model = col_character(),
-                specification = col_character(),
-                horizon = col_character(),
-                smoothed = col_double(),
-                original = col_double()
-              )) %>%
-  bind_rows() %>%
-  dplyr::select(model, horizon, specification, smoothed, original)
-
-write_csv(auc, "tables/table1-redone.csv")
-write_csv(auc, "../data/table1-redone.csv")
-
-## OLD
-
-# # Write latex versions of table 1
-#
-# top <- read_csv("tables/table1_top.csv") %>%
-#   rename(`ROC Smoothed` = X1) %>%
-#   mutate(Horizon = "1 Month",
-#          model = "Base Specification",
-#          `ROC Smoothed` = case_when(`ROC Smoothed` == "Smoothed" ~ "Yes",
-#                                     `ROC Smoothed` == "Not Smoothed" ~ "No")) %>%
-#   select(model, Horizon, everything())
-#
-# bottom <- read_csv("tables/table1_bottom.csv") %>%
-#   rename(`ROC Smoothed` = X1) %>%
-#   mutate(Horizon = "6 Months",
-#          model = "Base Specification",
-#          `ROC Smoothed` = case_when(`ROC Smoothed` == "Smoothed" ~ "Yes",
-#                                     `ROC Smoothed` == "Not Smoothed" ~ "No")) %>%
-#   select(model, Horizon, everything())
-#
-# tab_dat <- bind_rows(top, bottom)
-# names(tab_dat) <- c("Model", "Horizon", "ROC Smoothed", "Escalation", "Quad", "Goldstein", "CAMEO", "Avg")
-#
[…]

### 2020-05-29 rep_nosmooth/recreate-table2.R

@@ -1,70 +1,0 @@
-#
-#   Recreate table 2 with non-smoothed ROC
-#
-# install.packages("")
-library(readr)
-library(kableExtra)
-library(knitr)
-library(dplyr)
-library(here)
-
-setwd(here::here("MB-replication-checks"))
-getwd()
-# Write table summaries for monitoring on git
-
-auc <- lapply(dir("tables/PITF/", pattern = "AUCs-", full.names = TRUE), read_csv,
-              col_types = cols(
-                specification = col_character(),
-                horizon = col_character(),
-                smoothed = col_double(),
-                original = col_double())) %>%
-  bind_rows() %>%
-  dplyr::select(horizon, specification, smoothed, original)
-
-write_csv(auc, "tables/table2-redone.csv")
-write_csv(auc, "../data/table2-redone.csv")
-
-# Write latex versions of table
-
-
-
-top <- read_csv("tables/table1_top.csv") %>%
-  rename(`ROC Smoothed` = X1) %>%
-  mutate(Horizon = "1 Month",
-         model = "Base Specification",
-         `ROC Smoothed` = case_when(`ROC Smoothed` == "Smoothed" ~ "Yes",
-                                    `ROC Smoothed` == "Not Smoothed" ~ "No")) %>%
-  select(model, Horizon, everything())
-
-bottom <- read_csv("tables/table1_bottom.csv") %>%
-  rename(`ROC Smoothed` = X1) %>%
-  mutate(Horizon = "6 Months",
-         model = "Base Specification",
-         `ROC Smoothed` = case_when(`ROC Smoothed` == "Smoothed" ~ "Yes",
-                                    `ROC Smoothed` == "Not Smoothed" ~ "No")) %>%
-  select(model, Horizon, everything())
-
-tab_dat <- bind_rows(top, bottom)
-names(tab_dat) <- c("Model", "Horizon", "ROC Smoothed", "Escalation", "Quad", "Goldstein", "CAMEO", "Avg")
-
-footnote_text <- 'To create Table 1, Blair and Sambanis use a non-standard "smoothing" function when creating their various Receiver Operating Characteristic Curves (ROC). These "smoothed" curves are then used to compute their model performance metric -- the Area Under the Receiver Operating Characteristic Curve (AUC-ROC). We recreate these AUC-ROC scores with and without the smoothing function (Blair and Sambanis\' specification and the standard specification, respectively). This allows us to see if their findings are sensitive to the use of the smoothing function.'
-
[…]

### 2020-05-29 rep_nosmooth/recreate-tables.R

@@ -1,0 +1,27 @@
+
+library(tidyr)
+library(kableExtra)
+
+results <- read_rds("output/model-table-w-results.rds")
+
+table1_smooth <- results %>%
+  filter(table=="Table 1") %>%
+  select(horizon, row, column, auc_roc_smoothed) %>%
+  arrange(horizon, row) %>%
+  pivot_wider(names_from = "column", values_from = "auc_roc_smoothed") %>%
+  select(horizon, row, Escalation, Quad, Goldstein, CAMEO, Average) %>%
+  rename(Model = row)
+
+write_csv(table1_smooth, "output/table1-smooth.csv")
+
+table1_nosmooth <- results %>%
+  filter(table=="Table 1") %>%
+  select(horizon, row, column, auc_roc) %>%
+  arrange(horizon, row) %>%
+  pivot_wider(names_from = "column", values_from = "auc_roc") %>%
+  select(horizon, row, Escalation, Quad, Goldstein, CAMEO, Average) %>%
+  rename(Model = row)
+
+write_csv(table1_nosmooth, "output/table1-nosmooth.csv")
+
+

### 2020-05-29 rep_nosmooth/setup-model-table.R

@@ -65,7 +65,7 @@ model_table <- bind_rows(table1, table2) %>%
 # add a unique ID to the model table so that saving models is easier
 model_table <- model_table %>%
   ungroup() %>%
-  mutate(model_id = 1:n())
+  mutate(cell_id = 1:n())
 
 # mark models that are not implemented as random forests
 model_table <- model_table %>%
@@ -73,7 +73,7 @@ model_table <- model_table %>%
                                 "PITF Split Population"))
 
 model_table <- model_table %>%
-  select(model_id, everything())
+  select(cell_id, everything())
 
 # Model definitions / feature sets ----------------------------------------
 

### 2020-06-01 paper.Rmd
start revising paper; drop in new tables

--- 2020-05-21 issues.Rmd
@@
 title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
-author: Richard K. Morgan and Andreas Beger
+author: Richard K. Morgan and Andreas Beger and Michael D. Ward
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE)
 
@@
 
-![Copy of B&S Table 1](figures/bands-table1.png)
-
 The first row, each, for the one-month and 6-month forecasts are the core base specification. The additional rows reflect various robustness tests.
@@
 The results of this are shown in our Table 1 above.
-
-```{r, message=FALSE}
-table1 <- read_csv("data/table1-redone.csv") %>%
-  mutate(specification = factor(specification, levels = c("escalation", "quad", "goldstein", "CAMEO", "avg"))) %>%
-  arrange(model, horizon, specification) %>% 
-  pivot_longer(-c(model, horizon, specification), names_to = "ROC Smoothed", values_to = "Value") %>% 
-  pivot_wider( names_from = "specification", values_from = "Value") %>% 
-  mutate(`ROC Smoothed` = ifelse(`ROC Smoothed` == "smoothed", "Yes", "No"), 
-         model = "Base Specification") %>% 
-  rename(Escalation = escalation, 
-         Quad = quad, 
-         Goldstein = goldstein, 
-         Avg = avg)
-
-table1 %>%
-  knitr::kable(digits = 2, booktabs = TRUE, caption = "Comparison of smoothed and not smoothed AUC-ROC values", align = c(rep("l", 2), rep("c", 6))) %>% 
-  # column_spec(c(3), bold = TRUE) %>%
-  kableExtra::collapse_rows(1:2, row_group_label_position = 'stack')
-
-```
 
@@
 
-![Copy of B&S Table 2](figures/bands-table2.png)
-
 On page 20:
@@
 Table 2 is our recreation of B&S Table 2, copied in Figure 3, with AUC values for both smoothed and the original ROC curves.
-
-\newpage
-
-```{r, message=FALSE}
-escalation_only <- read_csv("data/table1-redone.csv") %>%
-  filter(specification=="escalation") %>%
-  select(-model) %>%
-  mutate(horizon = case_when(horizon=="6 month" ~ "6 months", TRUE ~ horizon))
-
-table2 <- read_csv("data/table2-redone.csv") %>%
-  bind_rows(escalation_only) %>%
-  mutate(
-    specification = factor(
-      specification, 
-      levels = c("escalation", "escalation with PITF", 
-                 "escalation with PITF weights", 
-                 "escalation with Split PITF", "PITF"),
-      labels = c("Escalation\nOnly", "With PITF Predictors", "Weighted by PITF", "PITF Split Population", "PITF Only"))
-  ) %>%
-  arrange(horizon, specification) %>% 
-  pivot_longer(-c(horizon, specification), names_to = "ROC Smoothed") %>% 
-  pivot_wider(names_from = "specification") %>% 
-  mutate(`ROC Smoothed` = ifelse(`ROC Smoothed` == "smoothed", "Yes", "No")) %>%
-  rename(Horizon = horizon)
-
-table2 %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Comparison of smoothed and not smoothed AUC-ROC values for B\\&S Table 2", 
-    align = c(rep("l", 2), rep("c", 6))
-  ) %>% 
-  kableExtra::column_spec(c(2, 5, 7), width = "1.5cm") %>%
-  kableExtra::column_spec(3, width = "1.8cm") %>%
-  kableExtra::column_spec(c(4, 6), width = "2.2cm") %>%
-  kableExtra::collapse_rows(1)
-
-```
 
@@
 
+```{r table1-smooth}
+table1_smooth <- read_csv("data/table1-smooth.csv")
+
+table1_smooth %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Test set AUC-ROC; replication of the original Table 1 with smoothed ROC curves.") %>% 
+  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
+```
+
+```{r table1-nosmooth}
+table1_nosmooth <- read_csv("data/table1-nosmooth.csv")
+
+table1_nosmooth %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Test set AUC-ROC; replication of the original Table 1 \\textit{without} smoothed ROC curves.") %>% 
+  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
+```
+
+```{r table2-smooth}
+table2_smooth <- read_csv("data/table2-smooth.csv")
+
+table2_smooth %>%
+  dplyr::select(-Model) %>%
+  rename(Horizon = horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Test set AUC-ROC for Models including PITF; original version based on smoothed ROC curves") 
+```
+
+```{r table2-nosmooth}
+table2_nosmooth <- read_csv("data/table2-nosmooth.csv")
+
+table2_nosmooth %>%
+  dplyr::select(-Model) %>%
+  rename(Horizon = horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Test set AUC-ROC for Models including PITF; version \\textit{without} smoothed ROC curves") 
+```
+
 # References

### 2020-06-03 paper.Rmd
adjust Table 2 results for N differences

@@
 library(kableExtra)
+library(stringr)
+
+# function to adjust the "cases" labels in Table 2
+fix_case_labels <- function(x) {
+  # escape the opening and closing parentheses
+  stringr::str_replace_all(x, "([\\(\\)])", "\\\\\\1")
+}
+
+# function to adjust the model/column labels in Table 2
 ```
+
+Issues in the B&S replication:
+
+- The AUC-ROC statistics are calculated using smoothed ROC curves
+- The "Weighted by PITF" model is incorrectly implemented
+- The "PITF Split Population" model is incorrectly implemented
+- The forecasts are scored using incidence, not onset
+
+Pre-registration should have had:
+
+- The escalation model specification, prior to looking at data!
+- The hyper-parameter tuning strategy.
+
+With rare events, also a lot of variance over RNG state.
 
@@
 
-```{r table2-smooth}
-table2_smooth <- read_csv("data/table2-smooth.csv")
-
-table2_smooth %>%
-  dplyr::select(-Model) %>%
-  rename(Horizon = horizon) %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Test set AUC-ROC for Models including PITF; original version based on smoothed ROC curves") 
-```
-
 ```{r table2-nosmooth}
@@
 table2_nosmooth %>%
-  dplyr::select(-Model) %>%
-  rename(Horizon = horizon) %>%
+  dplyr::select(-Model, -horizon) %>%
   knitr::kable(
     digits = 2, booktabs = TRUE, 
-    caption = "Test set AUC-ROC for Models including PITF; version \\textit{without} smoothed ROC curves") 
+    caption = "Replication of BS Table 2: Test set AUC-ROC for Escalation model with and without structural PITF contribution") %>%
+  column_spec(2:6, width = "2cm") %>%
+  pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1)) %>%
+  footnote("Differences from the original BS Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.", footnote_as_chunk = TRUE, threeparttable = TRUE)
+```
+
+```{r}
+table2_full <- read_csv("data/table2-for-appendix.csv")
+
+table2_full %>%
+  select(horizon, column, N_test, cases) %>%
+  pivot_wider(names_from = column, values_from = N_test) %>%
+  arrange(horizon, desc(cases)) %>%
+  rename(Cases = cases, Horizon = horizon) %>%
+  select(Cases, Horizon, everything()) %>%
+  arrange(desc(Cases), Horizon) %>%
+  knitr::kable("latex", booktabs = TRUE, digits = 0,
+               caption = "Number of valid test predictions for each cell in BS Table 2") %>%
+  column_spec(3:7, width = "2cm") %>%
+  collapse_rows(1:2, row_group_label_position = "stack")
+
+table2_full %>%
+  pivot_longer(auc_roc:auc_roc_smoothed, names_to = "Smoothed ROC",
+               values_to = "auc_roc") %>%
+  mutate(`Smoothed ROC` = case_when(
+    `Smoothed ROC`=="auc_roc" ~ "No",
+    TRUE ~ "Yes")
+  ) %>%
+  select(horizon, column, cases, `Smoothed ROC`, auc_roc) %>%
+  pivot_wider(names_from = column, values_from = auc_roc) %>%
+  rename(Cases = cases, Horizon = horizon) %>%
+  select(Cases, Horizon, `Smoothed ROC`, everything()) %>%
+  arrange(desc(Cases), Horizon, desc(`Smoothed ROC`)) %>%
+  knitr::kable("latex", booktabs = TRUE, digits = 2,
+               caption = "Replication of BS Table 2 with smoothed/original ROC and with original varying N cases or adjusting for common case set with constant N") %>%
+  column_spec(3, width = "1.5cm") %>%
+  column_spec(4:8, width = "2cm") %>%
+  collapse_rows(1:3, row_group_label_position = "stack")
+
 ```

### 2020-06-03 rep_nosmooth/model-runner.R

@@ -301,7 +301,7 @@ for (i in 1:nrow(non_rf_models)) {
       # The number of non-missing rows in the training data used to fit the mdl
       N_train = nrow(train_data),
       # The number of non-missing predictions
-      N_test  = sum(!is.na(preds$pred)),
+      N_test  = sum(complete.cases(preds)),
       time    = as.numeric((proc.time() - t0)["elapsed"])
     )
   results[[i]] <- res_i

### 2020-06-03 rep_nosmooth/recreate-tables.R

@@ -1,13 +1,27 @@
 
+library(here)
 library(readr)
 library(dplyr)
 library(tidyr)
 library(kableExtra)
+library(purrr)
 
-results <- read_rds("output/model-table-w-results-1234.rds")
+setwd(here::here("rep_nosmooth"))
+results <- read_rds("output/model-table-w-results.rds")
 
 
 # Table 1 -----------------------------------------------------------------
+
+# Note that all rows in Table 1 operate, correctly, on the same test
+# set N
+results %>%
+  filter(table=="Table 1") %>%
+  select(table, horizon, row, column, N_test) %>%
+  arrange(table, horizon, row) %>%
+  pivot_wider(names_from = "column", values_from = "N_test") %>%
+  select(table, horizon, row, Escalation, Quad, Goldstein, CAMEO, Average) %>%
+  rename(Model = row)
+
 
 table1_smooth <- results %>%
   filter(table=="Table 1") %>%
@@ -17,7 +31,7 @@ table1_smooth <- results %>%
   dplyr::select(horizon, row, Escalation, Quad, Goldstein, CAMEO, Average) %>%
   rename(Model = row)
 
-write_csv(table1_smooth, "output/table1-smooth.csv")
+write_csv(table1_smooth, "output/tables/table1-smooth.csv")
 table1_smooth %>%
   knitr::kable("markdown", digits = 2) %>%
   writeLines("output/table1-smooth.md")
@@ -30,7 +44,7 @@ table1_nosmooth <- results %>%
   dplyr::select(horizon, row, Escalation, Quad, Goldstein, CAMEO, Average) %>%
   rename(Model = row)
 
-write_csv(table1_nosmooth, "output/table1-nosmooth.csv")
+write_csv(table1_nosmooth, "output/tables/table1-nosmooth.csv")
 table1_nosmooth %>%
   knitr::kable("markdown", digits = 2) %>%
   writeLines("output/table1-nosmooth.md")
@@ -44,76 +58,103 @@ smooth_benefit <- results %>%
   dplyr::select(horizon, row, Escalation, Quad, Goldstein, CAMEO, Average) %>%
   rename(Model = row)
 
-write_csv(smooth_benefit, "output/table1-smooth-benefit.csv")
+write_csv(smooth_benefit, "output/tables/table1-smooth-benefit.csv")
 smooth_benefit %>%
   knitr::kable("markdown", digits = 2) %>%
[…]

### 2020-06-03 paper.Rmd
work on paper

--- 2020-06-01 paper.Rmd
@@
 library(kableExtra)
+library(stringr)
 ```
 
-We highlight some issues with the analysis in the recent @blair2020forecasting paper that undermine their main empirical finding -- that a theoretically-informed escalation model can forecast civil war onsets better than more inductive alternatives.
-Their paper has two key results:
+# Review of Blair and Sambanis 2020
 
-1. A theoretically-informed model outperforms alternative specifications in predicting civil war onset. 
-2. Structural variables do not add, and sometimes detract, from the forecast accuracy of the escalation model they advance.
-
-These results are highlighted in Table 1 and Table 2 [@blair2020forecasting, pg.
-13 and 19 respectively].
-They also evaluate forecasts for the first half of 2016 and conclude good accuracy in forecasting civil war (Table 4 on pg.23).
-
-We find that:
-
-- the first key finding depends on using AUC values calculated for *smoothed*, not the original, ROC curves
-- the 2016 forecasts are incorrectly assessed using the occurrence (incidence), not onset, of civil war
-
-# Review of the Blair and Sambanis paper
-
-@blair2020forecasting argue that theory contributes to forecasting accuracy even when using non-parametric machine learning models (in this case, the random forest model) and alternative specifications that are not informed by theory.
+@blair2020forecasting (BS for the sake of brevity hereafter) aim to examine whether theory adds value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
+They find that it does.
 They arrive at this conclusion by examining the problem of predicting civil war onset, and find that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory and a kitchen sink model with more than 1,000 covariates.
 
-Their data consist of country-month observations covering 2001 to 2015.
+BS specifically examine three questions:
+
+- (1) how does the theoretically-driven Escalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories, 
+- (2) does annual, structural information from the PITF instability forecasting model add to the Escalation model's monthly and 6-month predictions,
+- and (3) how accurate were predictions for the first half of 2016 derived from the Escalation model?
+
+To assess the first two questions BS use global datasets covering all major countries from 2001 to 2015.
+Two versions of the dataset are used, one at the country-month level, the other aggregated to 6-month half-years.
 The main outcome is civil war onset, measured using Sambanis' civil war dataset.
+
+Both the first and second questions above rely on comparing the Escalation model to various alternative models. The same procedure is used in both cases:
+
+1. Split the training data into training (2001 - 2007) and test (2008 - 2015) sets.
+2. Estimate the Escalation and other competing models. 
+3. Create out-of-sample (OOS) predictions from each model using the test set. 
+4. Calculate AUC-ROC measures for each set of OOS predictions.
+
+The corresponding results for each question are shown in BS Tables 1 and 2, which we will replicate further below.
+
+To examine the first question, BS compare the test set of the Escalation model to four alternative models.
 The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country.
-The key results are based on a comparison of the forecast accuracy of a random forest (RF) model with 4 covariate specifications as well as an ensemble model.
-These models are:
+The models are:
 
@@
 
-To assess forecast accuracy, Blair and Sambanis conduct the following procedures:
+The results in Table 1, aside from the core base specification results, include 8 additional robustness tests for both the 1-month and 6-month versions.
+These robustness checks vary either (1) random forecast hyperparameter values, or (2) the year used to split the train/test data, or (3) alternative codings of the civil war onset dependent variable.
 
-1. Split the training data into training (2001 - 2007) and test (2008 - 2015) sets.
-2. Estimate 4 random forest models with the specifications from above as well as a unweighted average of the models. 
-3. Create out-of-sample (OOS) predictions from each model using the test set. 
-4. Calculate AUC-ROC measures for each set of OOS predictions.
+The second question, whether structural variables add to the Escalation model, is assessed by comparing the original escalation model to four alternatives that incorporate annual, structural variables that are used in the PITF instability forecasting model:
 
-This is done for both 1-month ahead and 6-month ahead forecasts.
-Blair and Sambanis also conduct a series of robustness tests that (1) vary hyperparameters of the RF model, (2) change the split year for the training/test data split, and (3) alternative codings of the dependent variable.
+- Escalation Only: the original basic escalation model with only ICEWS predictors
+- With PITF Predictors: a random forest that as predictors has the escalation model indicators but also the PITF annual, structural variables
+- Weighted by PITF: escalation model predictions weighted using the PITF instability model predictions
+- PITF Split Population: the training data are split into high and low risk portions based on the PITF instability model predictions, two separate escalation random forests are trained on the splits, then re-combined into a single random forest that is used to create the test set predictions
+- PITF Only: a random forest model based only on the annual, structural PITF model predictors
 
-# Smoothed ROC curves
+The corresponding results are shown in BS Table 2.
 
-## Comparison of the escalation specification to alternatives (Table 1)
+Finally, BS used the Escalation model to create forcasts for the first half of 2016, and in their third and final analysis, they score the forecasts accuracy using civil war onset data later observed.
+This is summarized in BS Table 3.
 
-The results for the core analysis are summarized in Table 1 in @blair2020forecasting, which we copied in Figure 1 below.
+# Replication problems
 
-The first row, each, for the one-month and 6-month forecasts are the core base specification. The additional rows reflect various robustness tests.
+While attempting to replicate BS's results, we found several issues.
 
-The escalation model generally out-performs the alternative models/specifications, and on that basis Blair and Sambanis conclude that theory does indeed provide better forecast accuracy when compared to models not informed by theory.
+## Smoothed ROC curves
 
-A small detail, which is mentioned in the text and visually obvious in the ROC curves they show in B&S Figure 1, is that these AUC-ROC values are computed using smoothed ROC curves.
-We used their replication code to recreate the ROC curves in their figure, which are shown in the left two panels in the figure below.
-The curves indeed appear to be very smooth.
-But in the underlying test data for which these curves are, there are only 11 positive events, out of 2,366 rows for the 6-month data and 13,748 rows for the 1-month data.
-The ROC curves in the two panels on the right are the original, non-smoothed versions, which correctly reflect the small number of positive events.^[The diagonal lines in the ROC curves are there because there are groups of cases which have identical predictions, causing jumps in both the true and false positive rates at the same time.]
+The most consequential issue that we found is that all AUC-ROC values reported in BS Tables 1 and 2 are calculated using smoothed ROC curves, not the original, actual ROC curves.
+A reference to smoothing is made in a single sentence in the paper (p.
+12):
 
-![Replication of B&S Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)
+> Figure 1 displays the corresponding ROC curves, smoothed for ease of interpretation.
 
-\clearpage
+This implies that the ROC curves were only smoothed in the references Figure 1, but actually all AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation.
 
-Using these smoothed and non-smoothed curves, we then replicate B&S Table 1 (page 13), but now the table reports the AUC-ROC scores for each model specification with and without the use of a smoother.
-We focus on the core "Base specification" models only, not the additional robustness checks shown in the other rows in the table.
-The results of this are shown in our Table 1 above.
+Figure 1 shows our replication of both the smoothed ROC curves BS report, and the actual ROC curves on the right.
 
-First, note that we were unable to exactly reproduce the AUC-ROC values for all models.
-For example, in B&S the 1-month CAMEO has a AUC-ROC of 0.82, while our reproduction has a AUC-ROC value (using the smoothed version) of 0.84.
-This is most likely a function of the version of `R` Blair and Sambanis used to conduct their analysis.
-The current replication project was done using `R version 4.0.0`, which updated their random number seeding procedure, so small differences should be expected.
-Despite these small differences, the way we would interpret the smoothed-version results does not change.
-In the 1-month version of the Base Specification, the Escalation model outperforms all alternatives; in the 6-month version it outperforms all but one alternative and has equal performance with the Goldstein model.
+![Replication of BS Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)
 
-When we calculate the AUC-ROC values using the original, non-smoothed curves, the findings change substantially.
-Rows 2 and 4 in Table 1 show the non-smoothed AUC-ROC values for the 1-month and 6-month versions, respectively.
-In the 1-month version, the escalation model is outperformed by 2 alternatives and even with the other 2.
-In the 6-month version, all 4 alternatives outperform the escalation model.
-When using the original, non-smoothed ROC curves, the escalation model has the worst overall performance out of the 5 models considered.
-The first key finding in the paper, that the escalation model is superior to the alternative models, thus depends on using smoothed ROC curves when calculating the AUC-ROC statistic.
+ROC curves typically appear step-like in response to the distribution of positive and negative cases in the data.
+In this case there are also groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines.
+In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached.
+For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in BS Table 1, there are only 11 civil war onset cases in the test set.
+Thus the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.
+We can only speculate, but maybe the very "step"-like nature of the ROC curves plus the presence of unusual diagonal lines accounts for the decision to use smoothed ROC curves.
 
-\newpage
+As it turns out, this decision has a dramatic impact on the AUC-ROC calculations.
+As we will show in the results further below, the difference between the smoothed and original ROC AUC values is up to 0.12--a huge difference given that row-wise, the models being compared typically differ by only 0.05 or less--and differentially impacts the models that are being compared.
+In fact, BS's original results and interpretation are entirely conditional on the use of smoothed AUC-ROC.
 
-## Comparison of the escalation model to structural extensions (Table 2)
+To our knowledge, the norm is to calculate AUC-ROC values using original, not smoothed ROC curves.
+We are in fact not aware of other work that uses smoothed ROC curves for AUC calculations.
+It might be that there are theoretical reasons justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the BS results, it minimally would have warranted an explicit discussion in the paper.
+This is not the case, and it is only mentioned in the sentence we quote above.
 
-After establishing that the escalation model they advance out-performs all other covariate specifications, Blair and Sambanis move to show how the inclusion of structural risk variables -- the set of structural variables identified by the Political Instability Task Force (PITF) that can contribute to an increased risk of civil conflict.
-A copy of B&S Table 2 is included in Figure 3.
+The next three issues we encountered all concern information in Table 2.
 
-On page 20:
+## Incorrect "Weighted by PITF" implementation
 
-> Overall, our results suggest that while measures of structural risk may improve predictive performance, the value they add is marginal and inconsistent.
-Of the approaches we test, the split-population analog is most promising, with AUCs of .84 and .83 over one-month and six-month windows, respectively.
+The "Weighted by PITF" model is described as follows in BS, page 19:
 
-Table 2 is our recreation of B&S Table 2, copied in Figure 3, with AUC values for both smoothed and the original ROC curves.
+> The [Weighted by PITF model] uses PITF predicted probabilities to weight the results of the escalation model, ensuring that high-risk and low-risk countries that happen to take similar values on ICEWS-based predictors are nonetheless assigned different predicted probabilities in most months.
 
-Note that we were again unable to reproduce some of the AUC values in Table 2.
-Values that are notably different are 1-month "Weighted by PITF" and "PITF Split Population", and 6-month "PITF Split Population".
-For all three of these, the AUC-ROC values we obtained are much better than those in the original Table 2.
-The values for 6-month "PITF Only" and "With PITF Predictors" are slightly off but could be rounding errors.
+We infer that the intent is that the Escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
+The implementation actually weighs the test set predictions using the PITF model predictions for the training set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`).
+The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions.
+Similarly in the 6 month version of this file.] This appears to be a coding error.
 
-Despite these reproduction differences, the original interpretation mostly still stands.
-The various efforts to incorporate structural variables into the forecasts largely do not work or do not add much predictive value.
-In the 6-month forecasts, the value added by simply adding the annual structural predictors to the existing random forest model is more pronounced than before, while the PITF split population approach fares much worse.
+## Incorrect "PITF Split Population" implementation
 
-Although the substantive interpretation does not change much as a result of using ROC curve smoothing, the absolute AUC-ROC values between the smoothed and original versions do differe by quite a bit in several instances.
-The AUC-ROC values differ by 0.05 or more in 5 of the 10 cells in the original Table 2, and the direction of change is in all cases such that using smoothed ROC curves present more optimistic performance values.
+Similarly, the "PITF Split Population" model appears to be incorrectly implemented. BS describe it as, on page 20:
 
-# Assessment of the 2016 forecasts
+> The final approach is a random forest analog to split-population modeling.
+first compute the average PITF predicted probability for each country across all years in our training set.
+We define those that fall in the bottom quartile as "low risk" and the rest as "high risk." We then run our escalation model on the high-risk and low-risk subsets separately, combining the results into a single random forest (column 4).
 
-The predictions for the first half of 2016 are assessed in B&S Table 4:
+The intention clearly was to run two separate random forest models, one each on the low- and high-risk training data splits.
+The replication code does indeed run two separate random forecasts, but they are both run on the exact same training data, which consists of the full training data all other models are run on.
+The models are also identical otherwise, i.e.
+they use the same *x* variables and the same random forest hyper-parameter settings.
+The *only* difference in the models as they are implemented in the BS replication code is due to the non-deterministic nature of the random forest model itself.
+If we ran both with the same RNG seed, they would be identical in every respect, producing identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined.
+We have documented details at https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5.]
 
-![](figures/bands-table4.png)
+The implementation error aside, this split-population analog model is actually quite odd and does not actually replicate an analogue of the idea behind split-population modeling.
+Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision (regression^[See further below.
+Although the RF models are used for a binary decision problem, the actual implementation uses regression RFs for continuous outcomes.]) trees.
+Thus while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does indeed have 200,000 decision trees.
+Furthermore, the PITF model predictions do not impact the way the combined RF model predicts at all, not even through a binary low-/high-risk split.
+The split-population PITF RF model is practically speaking just another Escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree).
 
-The table presents 15 or 16 positive cases for 2016-H1, depending on a minor coding decision.
-This is the difference between the top and bottom confusion matrices.
-From the confusion matrices we can infer that the test data had a reported positive rate of around 9.5% for the first half of 2016.
-In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a positive rate of around 0.5%.
-Since the positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data, we suspected that the forecasts were erranously assessed using civil war incidence, not onset.
+## Inconsistent test set N for the models in Table 2
+
+Lastly, the AUC-ROC values reported in the original BS Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{table2-N}).
+ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available.
+Those sets differ across models (columns) for each row in Table 2.
+Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
+In other words, the results for different models in BS Table 2 are actually not comparable to the other, and any conclusions drawn from such comparison are potentially incorrect.
+
+We fix this issue by only using predictions for common joint subset of cases that all models have non-missing predictions for.
+The original BS Table 2 1-month have *N*=11,806--12,495 versus a common joint subset of 9,811, and for the 6-month row *N*=2,070--2,233 versus *N*=1,915 for the common joint subset.
+
+## Incorrect scoring of the 2016 forecasts
+
+BS show a confusiong matrix to score their 2016-H1 forecasts in Table 4.
+Although the forecasts are for the probability of civil war onset, in the replication code they are actually scored using the much more common incidence of civil war, i.e.
+including ongoing civil wars as "1"'s.
 
@@
 By examining the pattern of missing values, it seems clear that this was originally an incidence variable indicating whether a country was at civil war in a given year or not, and which was converted to an onset version so that onsets retain the value of 1 but continuing civil war years are coded as missing.
+This reflects common practice in how these are coded.
 
@@
 
-The correct confusion matrices when using observed onset (or the lack of it) are shown below.
+# Results of the updated replication
+
+## Basic interpretation of Table 1
+
+Table \ref{table1-smooth} is our replication of BS Table 1 with smoothed AUC-ROC.
+The results differ slightly from the original BS Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models.
+It is the case that BS set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel.
+In any case, the interpretation of results should not be sensitive to random variation, i.e.
+it should not depend on using a specific RNG seed.
+On the basis of these results, BS conclude that the Escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.
+
+```{r table1-smooth}
+table1_smooth <- read_csv("data/table1-smooth.csv")
+
+table1_smooth %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Replication of BS Table 1 with smoothed ROC curves; test set AUC-ROC for various models",
+    label = "table1-smooth") %>% 
+  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
+```
+
+Table \ref{table1-nosmooth} shows a version of BS Table 1 with the conventional non-smoothed ROC curves.
+The Average model outperforms the Escalation model in 17 out of 18 cases, and the CAMEO model outperforms in 16 of 18 cases, with one tie.
+The Goldstein model generally outperforms the Escalation model in the 6-month version.
+The Quad model appears to be roughly on par with the Escalation model.
+Thus the original BS conclusion that the Escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.
+
+```{r table1-nosmooth}
+table1_nosmooth <- read_csv("data/table1-nosmooth.csv")
+
+table1_nosmooth %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Replication of BS Table 1 \\textit{without} smoothed ROC curves; test set AUC-ROC for various models",
+    label = "table1-nosmooth") %>% 
+  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
+```
+
+## Do structural variables add to the Escalation model?
+
+Table \ref{table2-fixed} shows our replication of BS Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common joint subset of tests cases that all models have non-missing predictions for.
+Table \ref{table2-full} further below shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.
+
+```{r table2-nosmooth}
+table2_nosmooth <- read_csv("data/table2-nosmooth.csv")
+
+table2_nosmooth %>%
+  dplyr::select(-Model, -horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Replication of BS Table 2: Test set AUC-ROC for Escalation model with and without structural PITF contribution",
+    label = "table2-fixed") %>%
+  column_spec(2:6, width = "2cm") %>%
+  pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1)) %>%
+  footnote("Differences from the original BS Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.", footnote_as_chunk = TRUE, threeparttable = TRUE)
+```
+
+BS interpret the results as follows, on page 20:^[We list the "Overall, ..." interpretation out of order, last, for clarity.]
+
+> Of the approaches we test, the split-population analog is most promising
+
+This is no the case anymore. It outperforms in the 1-month version and underperforms the Escalation model in the 6-month version.
+
+> Adding PITF predictors improves the performance of the escalation model over six-month windows but diminishes it over one-month windows.
+
+Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model.
+
+> The weighted model performs very poorly regardless.
+
+It performs roughly on par with the Escalation Only model.
+
+One finding that remains is that the "PITF Only" model is outperformed by the "Escalation Only" model.
+As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is not surprising.
+
+> Overall, our results suggest that while measures of structural risk may improve predictive performance, the value they add is marginal and inconsistent.
+[...] Incorporating PITF thus significantly reduces or only slightly improves the performance of the escalation model, regardless of the approach we take.
+
+The most straightforward method of incorporating the annual, structural PITF variables--adding them to the predictors of the Escalation RF model--strictly outperforms the Escalation Only model.
+Note that the two other combination models considered are both non-standard, and that the "PITF Split Population" model does not in fact actually incorporate structural information at all.
+We thus conclude that adding structural variables actually clearly improves predictive performance.
+
+## The effect of using smoothed ROC curves
+
+```{r}
+table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
+table1_benefit %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Smoothing advantage for BS Table 1: the gain in AUC-ROC when calculated using smoothed ROC curves",
+    label = "table1-benefit") %>% 
+  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
+
+table2_benefit <- read_csv("data/table2-smooth-benefit.csv") 
+table2_benefit %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Smoothing advantage for BS Table 2: the gain in AUC-ROC when calculated using smoothed ROC curves",
+    label = "table2-benefit") %>% 
+  column_spec(2:6, width = "2cm") %>%
+  pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1))
+```
+
+What impact did the ROC smoothing have overall on the results reported in BS Tables 1 and 2?
+Tables \ref{table1-benefit} and \ref{table2-benefit} show the increase in AUC-ROC when using smoothed ROC curves, compared to the standard non-smoothed AUC-ROC.
+Positive values indicate that smoothing increased a model's apparent performance.
+The Escalation model is the only model that consistently had a benefit from smoothing.
+For all 8 other models, smoothing sometimes gave a benefit, sometimes not.
+
+A more dramatic difference stands out when we consider the overall average impact of smoothing across all flavors of a model reported in the columns of the tables.
+The Average, Goldstein, Quad, PITF Split Population, and Weighted by PITF models are slightly hurt by smoothing, but on the order of less than 0.01 in absolute magnitude.
+The "With PITF Predictors" model is hurt on the order of 0.02, i.e.
+it appears substantially worse with smoothing.
+The CAMEO and PITF Only models benefit slightly, on the order of 0.01 or less.
+The Escalation Only model on the other hand has an average boost of 0.058 to it's AUC-ROC from smoothing.
+Considering the spread of AUC-ROC values if we compare across rows in BS Tables 1 and 2, that boost is substantial.
+
+The use of smoothed ROC curves to calculate AUC-ROC values only clearly benefits the Escalation model, it does so consistently, and by a considerable margin.
+All 8 alternative models reported in Tables 1 and 2 do not on average gain when using smoothed ROC curves to calculate AUC.
+
+## How accurate were the 2016 forecasts?
+
+# Additional concerns
+
+## Random forest hyperparameters
+
+## AUC-ROC sensitivity
+
+RNG and variation
+
+## Pre-registration
+
+- The escalation model specification, prior to looking at data!
+- The hyper-parameter tuning strategy.
+
+# Conclusion
+
+We encountered several issues in the code underlying the BS analysis.
+
+The issues we encountered are not subjective modeling choices.
+
+When we fix these issue and perform an updated analysis, the conclusions BS draw all essentially overturn. In other words, BS findings are based on a faulty analysis, and invalid.
+
+Using the same analysis BS intend to use, we in fact find that:
+
+- the theory-driven Escalation model is outperformed both by the low-effort 1,160 predictor all CAMEO model and the Average ensemble model
+- structural variables substantially improve the Escalation model's performance when added to the pool of predictors the underlying random forest model draws on
+
+**AB: add a table comparing BS original claims and updated results.**
+
+\clearpage
+
+![](figures/bands-table4.png)
 
@@
 
-There are no civil war onsets in the data for 2016-H1.
-Thus the recall values is undefined, while the precision is 0/30 = 0, compared to recall and precision values of 13/15 = 0.87 and 13/30 = 0.43 before.
-The the alternative coding ("Assuming Change") at the bottom of Table 4, there are 2 civil war onsets.
-Recall is 1 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.
+```{r}
+table2_full <- read_csv("data/table2-for-appendix.csv")
 
-Another, minor issue or rather coding error, is related to using a lead version of the DV.
-With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 actually indicates the risk of civil war onset in 2016-H2.
-In the Table 4 script referenced above, the 2016-H1 predictions (for 2016-H2) are assessed using the raw DV, "incidence_civil_ns", not the lead version.
-Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1.
-In this case it probably doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.
+table2_full %>%
+  select(horizon, column, N_test, cases) %>%
+  pivot_wider(names_from = column, values_from = N_test) %>%
+  arrange(horizon, desc(cases)) %>%
+  rename(Cases = cases, Horizon = horizon) %>%
+  select(Cases, Horizon, everything()) %>%
+  arrange(desc(Cases), Horizon) %>%
+  knitr::kable(
+    "latex", booktabs = TRUE, digits = 0,
+    caption = "Number of valid test predictions for each cell in BS Table 2",
+    label = "table2-N") %>%
+  column_spec(3:7, width = "2cm") %>%
+  collapse_rows(1:2, row_group_label_position = "stack")
 
-# Discussion
+table2_full %>%
+  pivot_longer(auc_roc:auc_roc_smoothed, names_to = "Smoothed ROC",
+               values_to = "auc_roc") %>%
+  mutate(`Smoothed ROC` = case_when(
+    `Smoothed ROC`=="auc_roc" ~ "No",
+    TRUE ~ "Yes")
+  ) %>%
+  select(horizon, column, cases, `Smoothed ROC`, auc_roc) %>%
+  pivot_wider(names_from = column, values_from = auc_roc) %>%
+  rename(Cases = cases, Horizon = horizon) %>%
+  select(Cases, Horizon, `Smoothed ROC`, everything()) %>%
+  arrange(desc(Cases), Horizon, desc(`Smoothed ROC`)) %>%
+  knitr::kable(
+    "latex", booktabs = TRUE, digits = 2,
+               caption = "Replication of BS Table 2 with smoothed/original ROC and with original varying N cases or adjusting for common case set with constant N",
+    label = "table2-full") %>%
+  column_spec(3, width = "1.5cm") %>%
+  column_spec(4:8, width = "2cm") %>%
+  collapse_rows(1:3, row_group_label_position = "stack")
 
-When using the standard method of calculating AUC-ROC values, i.e.
-without smoothing the ROC curve beforehand, one of the principal B&S results reverses.
-The theory-driven escalation model does not in fact outperform the models that use alternative covariate specifications.
-In both the 1-month and 6-month versions of the base specification, all four other alternatives achieve equal or better AUC-ROC values compared to the escalation model.
-This includes the 1,159 covariate kitchen-sink CAMEO model.
-
-We find no general change in the results regarding whether adding structural factors to the escalation model's forecasts improves performance, although the gain from incorporating the PITF structural variables into the 6-month escalation model is much more pronounced.
-There is also no reason to rule out the possibility that the escalation model alternatives would not also similarly gain from incorporating structural variables in their 6-month versions.
-
-Furthermore, it appears that the forecasts for the risk of civil war onset in the first half of 2016 were erroneously evaluated using *incidence*, not *onset*.
-When correcting this mistake, the forecast evaluation changes significantly.
-There were no onsets in 2016, resulting in a precision of 0 and undefined recall; while with the manual re-coding reflected in the bottom portion of Table 4, the forecasts capture both civil war onsets for a recall of 1 and precision of 0.07.
-
-Both of these issues are relatively objective problems.
-It seems relatively clear to us that using incidence to assess onset forecasts is simply incorrect.
-Using smoothed ROC curves to calculate AUC is non-standard to the extent of our knowledge, and the decision to use them is not discussed or justified in the paper.
-Maybe it is possible to defend this choice, but given that one of the principal findings rests on it, this should be done explicitly.
-
-There are also other, more subjective concerns that we did not raise.
-For example, the paper does not explain how the random forest model hyperparemeters were selected.
-Although they do evaluate several changes in the hyperparameters as part of the robustness tests, the overall choice of hyperparameter values is unusual.
-Of note is also that this unusual approach only works (with warnings issued by the software) because B&S use random forests based on regression, not decision, trees.
-The concern to rule out would be that the hyperparameters are (over-) fit on the test data and that the reported performance values do not generalize.^[On page 10, B&S write that they use classification trees; this is strictly speaking incorrect and the random forest is built using regression tress on the 0 or 1 outcome.]
-
-We do not disagree with the overall point that theory is important for recognizing which variables or indicators might be important for forecasting.
-But it does not appear that B&S analysis demonstrates this, and that the paper's conclusions to this effect rest on faulty analysis.
-
-```{r table1-smooth}
-table1_smooth <- read_csv("data/table1-smooth.csv")
-
-table1_smooth %>%
-  select(-horizon) %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Test set AUC-ROC; replication of the original Table 1 with smoothed ROC curves.") %>% 
-  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
-```
-
-```{r table1-nosmooth}
-table1_nosmooth <- read_csv("data/table1-nosmooth.csv")
-
-table1_nosmooth %>%
-  select(-horizon) %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Test set AUC-ROC; replication of the original Table 1 \\textit{without} smoothed ROC curves.") %>% 
-  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
-```
-
-```{r table2-smooth}
-table2_smooth <- read_csv("data/table2-smooth.csv")
-
-table2_smooth %>%
-  dplyr::select(-Model) %>%
-  rename(Horizon = horizon) %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Test set AUC-ROC for Models including PITF; original version based on smoothed ROC curves") 
-```
-
-```{r table2-nosmooth}
-table2_nosmooth <- read_csv("data/table2-nosmooth.csv")
-
-table2_nosmooth %>%
-  dplyr::select(-Model) %>%
-  rename(Horizon = horizon) %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Test set AUC-ROC for Models including PITF; version \\textit{without} smoothed ROC curves") 
 ```

### 2020-06-03 rep_nosmooth/recreate-tables.R

@@ -7,7 +7,7 @@ library(kableExtra)
 library(purrr)
 
 setwd(here::here("rep_nosmooth"))
-results <- read_rds("output/model-table-w-results.rds")
+results <- read_rds("output/model-table-w-results-1234.rds")
 
 
 # Table 1 -----------------------------------------------------------------
@@ -49,7 +49,7 @@ table1_nosmooth %>%
   knitr::kable("markdown", digits = 2) %>%
   writeLines("output/table1-nosmooth.md")
 
-smooth_benefit <- results %>%
+table1_smooth_benefit <- results %>%
   filter(table=="Table 1") %>%
   mutate(diff = auc_roc_smoothed - auc_roc) %>%
   dplyr::select(horizon, row, column, diff) %>%
@@ -58,8 +58,8 @@ smooth_benefit <- results %>%
   dplyr::select(horizon, row, Escalation, Quad, Goldstein, CAMEO, Average) %>%
   rename(Model = row)
 
-write_csv(smooth_benefit, "output/tables/table1-smooth-benefit.csv")
-smooth_benefit %>%
+write_csv(table1_smooth_benefit, "output/tables/table1-smooth-benefit.csv")
+table1_smooth_benefit %>%
   knitr::kable("markdown", digits = 2) %>%
   writeLines("output/table1-smooth-benefit.md")
 
@@ -99,6 +99,10 @@ common_subset <- preds %>%
       select(year, period, country_iso3)
     out
   }))
+common_subset %>%
+  group_by(horizon, column, preds) %>%
+  summarize(n = map_int(preds, nrow)) %>%
+  select(-preds)
 # Since we need to calculate common subsets by horizon, split the data frame
 # by horizon, then use inner_join on the row ID tibbles in each preds column
 # to get the case subsets
@@ -187,3 +191,17 @@ full_table2 <- bind_rows(orig, adj) %>%
 write_csv(full_table2, "output/tables/table2-for-appendix.csv")
 
 
+# Average smooth benefit --------------------------------------------------
+#
+#   This is mentioned in the text
+#
+
+table1_smooth_benefit %>%
+  pivot_longer(Escalation:Average) %>%
+  group_by(name) %>%
+  summarize(Avg_benefit = mean(value))
+
+table2_smooth_benefit %>%
+  pivot_longer(-c(horizon, Model)) %>%
+  group_by(name) %>%
+  summarize(Avg_benefit = mean(value))

### 2020-06-03 paper.Rmd
add RF HP settings table back in

@@
 
+What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated.
+
+|Hyperparameter|Default heuristic|Default values (Escalation)|B&S value|
+|--------------|-----------------|--------------------------|---------|
+|type          |                 | classification           | regression|
+|ntree         |                 | 500                      | 100,000 or 1e6 |
+|mtry          |`floor(sqrt(ncol(x)))`|3|3|
+|replace       |                 | true | false |
+|sampsize | `nrow(x)` if replace, | 11,869 | 100 or 500 |
+||else `ceiling(.632*nrow(x))`|||
+|nodesize | 1 for classification | 1 | 1 |
+|maxnodes | | null | 5 or 10 |
+
+There is a stark contrast in the default RF settings and the way Blair and Sambanis use the RF models.
+The default approach is to train a relatively small number (ntree; 500) of classification trees, but where each tree is fairly big in that it is trained on data that has the same number of rows as the training data, albeit sampled with replacement (replace is true; sampsize is 11,869), and allowed to grow fairly deep (this is governed by nodesize (1), which is the minimum size a terminal node must have).
+In contrast, Blair and Sambanis grow very extensive forests with a large number of trees (100,000 compared to 500), but each tree is very small and shallow; only a 100 rows are sampled from the training data for each tree, and the trees are constrained to at most 5 terminal nodes (maxnodes).
+
+This approach only works due to the other unorthodox choice, which is to use regression, not classification, trees.
+Trying to use classification trees with the other parameter settings in fact does not work at all because it is almost guaranteed that a sample of 100 from the 11,869 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample.
+As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:
+
+```
+Warning message:
+In randomForest.default(y = as.integer(train_df$incidence_civil_ns_plus1 ==  :
+  The response has five or fewer unique values.  Are you sure you want to do regression?
+```
+
+The RF approach BS use is essentially similar to the probability forests proposed in Malley 2012. It is not clear if this is intentional or not, as the paper is not cited.
+
+We tried to create versions of the RF models that use the regular framework for classification forests, but were unable to create models that approach the BS RFs test set performance.
+
+How were these hyper-parameter values determined?
+
 ## AUC-ROC sensitivity

### 2020-06-03 paper.Rmd
sync minor changes
Update paper.Rmd

@@
 ---
-title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
+title: "Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
 author: Richard K. Morgan and Andreas Beger and Michael D. Ward
@@
   pdf_document:
-    keep_tex: false
+    keep_tex: true
 references:
@@
     year: 2020
+bibliography: ~/Git/whistle/master.bib
 ---
@@
 library(tidyr)
+library(tidyverse)
 library(kableExtra)
@@
 
-How were these hyper-parameter values determined?
+How were these hyper-parameter values determined? Do they work better for some models than others?
 
@@
 # Conclusion
+
+BS has a clear misunderstanding of the role that out-of-sample prediction can plan in analysis.
+On the one hand it can be used for simple forecasting while on the other it can be used to evaluate the performance of models, specifically to overfitting and bias.
+They further misrepresent cross-validation which is not about fishing for results, as claimed by BS.
+The current state-of-the-art uses prediction and cross-validation most frequently to provide supportive evidence that is independent of the estimation procedure and the in-sample data.
+It is incorrect to paint this procedure as atheoretical since many studies will have some explanation for how the model was constructed.
+These procedures may simply be used to provide evidence for a theoretical argument.
+Though that argument may be unconvincing and/or atheoretical.
+
+The BS approach that is advocated is to use theory to guide prediction.
+But theory is a quite ambiguous and undefined concept.
+It is not a procedure.
+What they actually do is to create a model with four right hand side variables that is supposed to capture a complicated repression-dissent dynamic.
+The dynamic is probably not linear, but their model is.
+There is a wide-ranging literature on this dynamic that they do not rely on to construct their model.
+As such their baseline comparison is hardly a standard bearer for strong theory.
+
+Further, they complete misunderstand the use of ICEWS event data in current research.
+They claim that most uses to date have focused on the quad categories, but this ignores a wide swath of literature (Steinert-Threlkeld ApsR 2017) and Metternich et al (AJPX 2013) that uses a sepcific action---such as protest---defined in the CAMEO ontology.
+In the AJPS article we hand coded, for example, every actor in Thailand and focused on an analysis of how those have been interacting.
 

### 2020-06-03 paper.Rmd
moar stuff
moar stuff

@@
     year: 2020
-bibliography: ~/Git/whistle/master.bib
+bibliography: /Users/mdw/Git/whistle/master.bib
 ---

### 2020-06-03 paper.tex
moar stuff
moar stuff

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020, ``Forecasting Civil Wars: Theory and Structure in an Age of `Big Data' and Machine Learning'', JCR},
            pdfauthor={Richard K. Morgan and Andreas Beger and Michael D. Ward},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020,
``Forecasting Civil Wars: Theory and Structure in an Age of `Big Data'
and Machine Learning'', \emph{JCR}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Richard K. Morgan and Andreas Beger and Michael D. Ward}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{03 June 2020}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\maketitle

\hypertarget{review-of-blair-and-sambanis-2020}{%
\section{Review of Blair and Sambanis
2020}\label{review-of-blair-and-sambanis-2020}}

Blair and Sambanis (2020) (BS for the sake of brevity hereafter) aim to
examine whether theory adds value to a forecasting model when compared
to non-parametric machine learning models, and specifically random
forests, whose specifications are not theory-informed. They find that it
does. They arrive at this conclusion by examining the problem of
predicting civil war onset, and find that a parsimonious model using a
small number of covariates derived from escalation theories of conflict
can forecast civil war onset better than alternative specifications
based on generic covariates not specifically informed by theory and a
kitchen sink model with more than 1,000 covariates.

BS specifically examine three questions:

\begin{itemize}
\item
  \begin{enumerate}
  \def\labelenumi{(\arabic{enumi})}
  \tightlist
  \item
    how does the theoretically-driven Escalation model compare in
    forecast performance to alternative models not informed specifically
    by civil war onset theories,
  \end{enumerate}
\item
  \begin{enumerate}
  \def\labelenumi{(\arabic{enumi})}
  \setcounter{enumi}{1}
  \tightlist
  \item
    does annual, structural information from the PITF instability
    forecasting model add to the Escalation model's monthly and 6-month
    predictions,
  \end{enumerate}
\item
  and (3) how accurate were predictions for the first half of 2016
  derived from the Escalation model?
\end{itemize}

To assess the first two questions BS use global datasets covering all
major countries from 2001 to 2015. Two versions of the dataset are used,
one at the country-month level, the other aggregated to 6-month
half-years. The main outcome is civil war onset, measured using
Sambanis' civil war dataset.

Both the first and second questions above rely on comparing the
Escalation model to various alternative models. The same procedure is
used in both cases:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Split the training data into training (2001 - 2007) and test (2008 -
  2015) sets.
\item
  Estimate the Escalation and other competing models.
\item
  Create out-of-sample (OOS) predictions from each model using the test
  set.
\item
  Calculate AUC-ROC measures for each set of OOS predictions.
\end{enumerate}

The corresponding results for each question are shown in BS Tables 1 and
2, which we will replicate further below.

To examine the first question, BS compare the test set of the Escalation
model to four alternative models. The independent variables for the
first set of analysis reported in Table 1 in the paper are all derived
from the ICEWS event data, using domestic events between actors within a
country. The models are:

\begin{itemize}
\tightlist
\item
  Escalation: a set of 10 theoretically-informed indicators.
\item
  Quad: ICEWS quad counts, i.e.~material conflict, material cooperation,
  verbal conflict, verbal cooperation.
\item
  Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from
  the ICEWS data for interactions between the government one one side
  and opposition or rebel actors on the other. These are directed, thus
  making for 4 total covariates.
\item
  CAMEO: counts for all CAMEO event codes, thus a total of 1,159
  covariates.
\item
  Average: unweighted average of the predictions from the 4 models
  above.
\end{itemize}

The results in Table 1, aside from the core base specification results,
include 8 additional robustness tests for both the 1-month and 6-month
versions. These robustness checks vary either (1) random forecast
hyperparameter values, or (2) the year used to split the train/test
data, or (3) alternative codings of the civil war onset dependent
variable.

The second question, whether structural variables add to the Escalation
model, is assessed by comparing the original escalation model to four
alternatives that incorporate annual, structural variables that are used
in the PITF instability forecasting model:

\begin{itemize}
\tightlist
\item
  Escalation Only: the original basic escalation model with only ICEWS
  predictors
\item
  With PITF Predictors: a random forest that as predictors has the
  escalation model indicators but also the PITF annual, structural
  variables
\item
  Weighted by PITF: escalation model predictions weighted using the PITF
  instability model predictions
\item
  PITF Split Population: the training data are split into high and low
  risk portions based on the PITF instability model predictions, two
  separate escalation random forests are trained on the splits, then
  re-combined into a single random forest that is used to create the
  test set predictions
\item
  PITF Only: a random forest model based only on the annual, structural
  PITF model predictors
\end{itemize}

The corresponding results are shown in BS Table 2.

Finally, BS used the Escalation model to create forcasts for the first
half of 2016, and in their third and final analysis, they score the
forecasts accuracy using civil war onset data later observed. This is
summarized in BS Table 3.

\hypertarget{replication-problems}{%
\section{Replication problems}\label{replication-problems}}

While attempting to replicate BS's results, we found several issues.

\hypertarget{smoothed-roc-curves}{%
\subsection{Smoothed ROC curves}\label{smoothed-roc-curves}}

The most consequential issue that we found is that all AUC-ROC values
reported in BS Tables 1 and 2 are calculated using smoothed ROC curves,
not the original, actual ROC curves. A reference to smoothing is made in
a single sentence in the paper (p.~12):

\begin{quote}
Figure 1 displays the corresponding ROC curves, smoothed for ease of
interpretation.
\end{quote}

This implies that the ROC curves were only smoothed in the references
Figure 1, but actually all AUC-ROC calculations throughout the
replication code use an option to smooth the ROC curves prior to AUC
calculation.

Figure 1 shows our replication of both the smoothed ROC curves BS
report, and the actual ROC curves on the right.

\begin{figure}
\centering
\includegraphics{figures/figure1-replicated.png}
\caption{Replication of BS Figure 1 with both smooth and non-smoothed
ROC curves.}
\end{figure}

ROC curves typically appear step-like in response to the distribution of
positive and negative cases in the data. In this case there are also
groups of cases with identical predicted probabilities, which accounts
for the unusual diagonal lines. In any case, with a sparse outcome like
civil war onsets, the true positive rate on the \emph{y}-axis only
changes when the prediction for a observed positive case is reached. For
these ROC curves, and for that matter in the basic train/test split used
for 12 of the 18 rows/models in BS Table 1, there are only 11 civil war
onset cases in the test set. Thus the ROC curves here are very
step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct
\emph{y} coordinates. We can only speculate, but maybe the very
``step''-like nature of the ROC curves plus the presence of unusual
diagonal lines accounts for the decision to use smoothed ROC curves.

As it turns out, this decision has a dramatic impact on the AUC-ROC
calculations. As we will show in the results further below, the
difference between the smoothed and original ROC AUC values is up to
0.12--a huge difference given that row-wise, the models being compared
typically differ by only 0.05 or less--and differentially impacts the
models that are being compared. In fact, BS's original results and
interpretation are entirely conditional on the use of smoothed AUC-ROC.

To our knowledge, the norm is to calculate AUC-ROC values using
original, not smoothed ROC curves. We are in fact not aware of other
work that uses smoothed ROC curves for AUC calculations. It might be
that there are theoretical reasons justifying the use of smoothed ROC
curves over the original ROC curves; but given that this decision
dramatically impacts the interpretation of the BS results, it minimally
would have warranted an explicit discussion in the paper. This is not
the case, and it is only mentioned in the sentence we quote above.

The next three issues we encountered all concern information in Table 2.

\hypertarget{incorrect-weighted-by-pitf-implementation}{%
\subsection{Incorrect ``Weighted by PITF''
implementation}\label{incorrect-weighted-by-pitf-implementation}}

The ``Weighted by PITF'' model is described as follows in BS, page 19:

\begin{quote}
The {[}Weighted by PITF model{]} uses PITF predicted probabilities to
weight the results of the escalation model, ensuring that high-risk and
low-risk countries that happen to take similar values on ICEWS-based
predictors are nonetheless assigned different predicted probabilities in
most months.
\end{quote}

We infer that the intent is that the Escalation model's predictions for
the test set are weighted by the PITF model predictions for the test
set. The implementation actually weighs the test set predictions using
the PITF model predictions for the training set.\footnote{See
  \texttt{1mo\_run\_escalation\_weighted\_PITF.R} line 4, where the PITF
  predictions are taken from the training data set
  (\texttt{train\$pred\_prob\_plus1}). The next line is a hack extending
  the shorter \texttt{weight} vector with missing values to avoid a R
  warning when it is multiplied with the longer vector of escalation
  model test set predictions. Similarly in the 6 month version of this
  file.} This appears to be a coding error.

\hypertarget{incorrect-pitf-split-population-implementation}{%
\subsection{Incorrect ``PITF Split Population''
implementation}\label{incorrect-pitf-split-population-implementation}}

Similarly, the ``PITF Split Population'' model appears to be incorrectly
implemented. BS describe it as, on page 20:

\begin{quote}
The final approach is a random forest analog to split-population
modeling. first compute the average PITF predicted probability for each
country across all years in our training set. We define those that fall
in the bottom quartile as ``low risk'' and the rest as ``high risk.'' We
then run our escalation model on the high-risk and low-risk subsets
separately, combining the results into a single random forest (column
4).
\end{quote}

The intention clearly was to run two separate random forest models, one
each on the low- and high-risk training data splits. The replication
code does indeed run two separate random forecasts, but they are both
run on the exact same training data, which consists of the full training
data all other models are run on. The models are also identical
otherwise, i.e.~they use the same \emph{x} variables and the same random
forest hyper-parameter settings. The \emph{only} difference in the
models as they are implemented in the BS replication code is due to the
non-deterministic nature of the random forest model itself. If we ran
both with the same RNG seed, they would be identical in every respect,
producing identical predictions.\footnote{Disentangling this coding
  error is not straightforward as it occurs over several R scripts and
  requires (or at least is easier to verify by) running partway through
  the actual replication until the objects holding the training data for
  the models are instantiated and can be examined. We have documented
  details at
  \url{https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5}.}

The implementation error aside, this split-population analog model is
actually quite odd and does not actually replicate an analogue of the
idea behind split-population modeling. Although the two RFs are trained
on separate data (in our updated, fixed replication), the process of
combining them actually just creates a new, larger RF using both
component model's underlying decision (regression\footnote{See further
  below. Although the RF models are used for a binary decision problem,
  the actual implementation uses regression RFs for continuous outcomes.})
trees. Thus while all RF models throughout (except for one of the
robustness checks) are trained with 100,000 decision trees
(\texttt{ntree}), the new RF model after combination does indeed have
200,000 decision trees. Furthermore, the PITF model predictions do not
impact the way the combined RF model predicts at all, not even through a
binary low-/high-risk split. The split-population PITF RF model is
practically speaking just another Escalation model trained with
N=200,000 instead of N=100,000 trees and an extra odd randomization step
added to the already existing RF randomization facilities (row and
column sampling for each decision tree).

\hypertarget{inconsistent-test-set-n-for-the-models-in-table-2}{%
\subsection{Inconsistent test set N for the models in Table
2}\label{inconsistent-test-set-n-for-the-models-in-table-2}}

Lastly, the AUC-ROC values reported in the original BS Table 2 are
calculated on the basis of slightly different numbers of underlying test
set cases (see Table \ref{table2-N}). ROC calculations for a set of
predictions can only be done on the set of cases for which both
non-missing predictions and non-missing outcomes are available. Those
sets differ across models (columns) for each row in Table 2. Thus a
difference in AUC-ROC values for two models could be due to the fact
that they were calculated on different sets of underlying cases, not
because the models are systemically performing at a different level. In
other words, the results for different models in BS Table 2 are actually
not comparable to the other, and any conclusions drawn from such
comparison are potentially incorrect.

We fix this issue by only using predictions for common joint subset of
cases that all models have non-missing predictions for. The original BS
Table 2 1-month have \emph{N}=11,806--12,495 versus a common joint
subset of 9,811, and for the 6-month row \emph{N}=2,070--2,233 versus
\emph{N}=1,915 for the common joint subset.

\hypertarget{incorrect-scoring-of-the-2016-forecasts}{%
\subsection{Incorrect scoring of the 2016
forecasts}\label{incorrect-scoring-of-the-2016-forecasts}}

BS show a confusiong matrix to score their 2016-H1 forecasts in Table 4.
Although the forecasts are for the probability of civil war onset, in
the replication code they are actually scored using the much more common
incidence of civil war, i.e.~including ongoing civil wars as ``1'''s.

The relevant variables in the data are ``incidence\_civil\_ns'' and
``incidence\_civil\_ns\_plus1'', which appears to be a 1-period lead
version of the DV that is used in the actual prediction models. The
incidence DV contains both 0/1 and missing values. By examining the
pattern of missing values, it seems clear that this was originally an
incidence variable indicating whether a country was at civil war in a
given year or not, and which was converted to an onset version so that
onsets retain the value of 1 but continuing civil war years are coded as
missing. This reflects common practice in how these are coded.

By examining the code used to generated Table 4, we were able to confirm
that the onset forecasts are assessed using incidence, not onset. In the
file \texttt{6mo\_make\_confusion\_matrix.do} on line 52, missing values
in ``incidence\_civil\_ns'' are recoded to 1, thus reverting the onset
coding of this variable back to incidence.

\hypertarget{results-of-the-updated-replication}{%
\section{Results of the updated
replication}\label{results-of-the-updated-replication}}

\hypertarget{basic-interpretation-of-table-1}{%
\subsection{Basic interpretation of Table
1}\label{basic-interpretation-of-table-1}}

Table \ref{table1-smooth} is our replication of BS Table 1 with smoothed
AUC-ROC. The results differ slightly from the original BS Table 1,
typically by no more than 0.01, due to the non-deterministic nature of
the RF models. It is the case that BS set the RNG seed in their
replication code, which should theoretically allow exact reproduction,
but (1) there was a change in more recent versions of R that affected
the RNG seeding process, and (2) we refactored the replication script to
allow one to run the models in parallel. In any case, the interpretation
of results should not be sensitive to random variation, i.e.~it should
not depend on using a specific RNG seed. On the basis of these results,
BS conclude that the Escalation model is generally superior to the
alternatives, and we can replicate that interpretation when using
smoothed ROC curves.

\begin{table}[t]

\caption{\label{tab:table1-smooth}Replication of BS Table 1 with smoothed ROC curves; test set AUC-ROC for various models}
\centering
\begin{tabular}{lrrrrr}
\toprule
Model & Escalation & Quad & Goldstein & CAMEO & Average\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{One-month forecasts}}\\
\hspace{1em}Base specification & 0.85 & 0.80 & 0.79 & 0.83 & 0.82\\
\hspace{1em}Terminal nodes & 0.86 & 0.79 & 0.78 & 0.83 & 0.82\\
\hspace{1em}Sample size & 0.85 & 0.81 & 0.70 & 0.86 & 0.84\\
\hspace{1em}Trees per forest & 0.84 & 0.80 & 0.78 & 0.83 & 0.82\\
\hspace{1em}Training/test sets 1 & 0.86 & 0.78 & 0.75 & 0.81 & 0.80\\
\hspace{1em}Training/test sets 2 & 0.81 & 0.79 & 0.72 & 0.78 & 0.78\\
\hspace{1em}Training/test sets 3 & 0.79 & 0.80 & 0.69 & 0.75 & 0.75\\
\hspace{1em}Coding of DV 1 & 0.86 & 0.81 & 0.80 & 0.84 & 0.83\\
\hspace{1em}Coding of DV 2 & 0.92 & 0.81 & 0.81 & 0.82 & 0.81\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Six-month forecasts}}\\
\hspace{1em}Base specification & 0.82 & 0.78 & 0.82 & 0.77 & 0.79\\
\hspace{1em}Terminal nodes & 0.80 & 0.75 & 0.81 & 0.75 & 0.77\\
\hspace{1em}Sample size & 0.83 & 0.78 & 0.78 & 0.78 & 0.79\\
\hspace{1em}Trees per forest & 0.82 & 0.78 & 0.82 & 0.77 & 0.79\\
\hspace{1em}Training/test sets 1 & 0.79 & 0.77 & 0.81 & 0.75 & 0.78\\
\hspace{1em}Training/test sets 2 & 0.72 & 0.73 & 0.77 & 0.74 & 0.75\\
\hspace{1em}Training/test sets 3 & 0.88 & 0.70 & 0.81 & 0.68 & 0.79\\
\hspace{1em}Coding of DV 1 & 0.83 & 0.77 & 0.82 & 0.79 & 0.80\\
\hspace{1em}Coding of DV 2 & 0.83 & 0.77 & 0.82 & 0.79 & 0.79\\
\bottomrule
\end{tabular}
\end{table}

Table \ref{table1-nosmooth} shows a version of BS Table 1 with the
conventional non-smoothed ROC curves. The Average model outperforms the
Escalation model in 17 out of 18 cases, and the CAMEO model outperforms
in 16 of 18 cases, with one tie. The Goldstein model generally
outperforms the Escalation model in the 6-month version. The Quad model
appears to be roughly on par with the Escalation model. Thus the
original BS conclusion that the Escalation model is superior to the
alternative models is completely conditional on the non-standard use of
smoothed ROC curves, and overturns when using traditional AUC-ROC
calculations.

\begin{table}[t]

\caption{\label{tab:table1-nosmooth}Replication of BS Table 1 \textit{without} smoothed ROC curves; test set AUC-ROC for various models}
\centering
\begin{tabular}{lrrrrr}
\toprule
Model & Escalation & Quad & Goldstein & CAMEO & Average\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{One-month forecasts}}\\
\hspace{1em}Base specification & 0.78 & 0.78 & 0.79 & 0.80 & 0.82\\
\hspace{1em}Terminal nodes & 0.79 & 0.78 & 0.78 & 0.81 & 0.82\\
\hspace{1em}Sample size & 0.79 & 0.80 & 0.74 & 0.82 & 0.84\\
\hspace{1em}Trees per forest & 0.78 & 0.78 & 0.79 & 0.81 & 0.82\\
\hspace{1em}Training/test sets 1 & 0.78 & 0.76 & 0.76 & 0.79 & 0.80\\
\hspace{1em}Training/test sets 2 & 0.75 & 0.77 & 0.73 & 0.76 & 0.78\\
\hspace{1em}Training/test sets 3 & 0.70 & 0.79 & 0.69 & 0.73 & 0.74\\
\hspace{1em}Coding of DV 1 & 0.80 & 0.79 & 0.80 & 0.82 & 0.83\\
\hspace{1em}Coding of DV 2 & 0.80 & 0.82 & 0.78 & 0.83 & 0.81\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Six-month forecasts}}\\
\hspace{1em}Base specification & 0.77 & 0.78 & 0.83 & 0.79 & 0.81\\
\hspace{1em}Terminal nodes & 0.77 & 0.78 & 0.82 & 0.78 & 0.79\\
\hspace{1em}Sample size & 0.78 & 0.77 & 0.80 & 0.80 & 0.82\\
\hspace{1em}Trees per forest & 0.77 & 0.79 & 0.83 & 0.78 & 0.81\\
\hspace{1em}Training/test sets 1 & 0.75 & 0.79 & 0.82 & 0.78 & 0.80\\
\hspace{1em}Training/test sets 2 & 0.70 & 0.75 & 0.78 & 0.75 & 0.77\\
\hspace{1em}Training/test sets 3 & 0.85 & 0.72 & 0.84 & 0.71 & 0.81\\
\hspace{1em}Coding of DV 1 & 0.77 & 0.80 & 0.83 & 0.79 & 0.82\\
\hspace{1em}Coding of DV 2 & 0.80 & 0.78 & 0.84 & 0.80 & 0.81\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{do-structural-variables-add-to-the-escalation-model}{%
\subsection{Do structural variables add to the Escalation
model?}\label{do-structural-variables-add-to-the-escalation-model}}

Table \ref{table2-fixed} shows our replication of BS Table 2 with (1)
regular, not smoothed, AUC-ROC, (2) fixed ``Weighted by PITF'' and
``PITF Split-Population'' models, and (3) AUC-ROC values computed on the
common joint subset of tests cases that all models have non-missing
predictions for. Table \ref{table2-full} further below shows AUC-ROC
values for both smoothed and non-smoothed versions, and both the
original, model-varying test cases sets and our common joint subset.

\begin{table}[t]

\caption{\label{tab:table2-fixed}Replication of BS Table 2: Test set AUC-ROC for Escalation model with and without structural PITF contribution}
\centering
\begin{threeparttable}
\begin{tabular}{r>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{}p{2cm}}
\toprule
Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
\midrule
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{One-month forecasts}}\\
\hspace{1em}0.69 & 0.71 & 0.69 & 0.81 & 0.67\\
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Six-month forecasts}}\\
\hspace{1em}0.79 & 0.87 & 0.80 & 0.69 & 0.71\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item \textit{Note: } 
\item Differences from the original BS Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.
\end{tablenotes}
\end{threeparttable}
\end{table}

BS interpret the results as follows, on page 20:\footnote{We list the
  ``Overall, \ldots{}'' interpretation out of order, last, for clarity.}

\begin{quote}
Of the approaches we test, the split-population analog is most promising
\end{quote}

This is no the case anymore. It outperforms in the 1-month version and
underperforms the Escalation model in the 6-month version.

\begin{quote}
Adding PITF predictors improves the performance of the escalation model
over six-month windows but diminishes it over one-month windows.
\end{quote}

Adding PITF predictors actually improves performance in both cases; the
``With PITF Predictions'' model strictly dominates the ``Escalation
Only'' model.

\begin{quote}
The weighted model performs very poorly regardless.
\end{quote}

It performs roughly on par with the Escalation Only model.

One finding that remains is that the ``PITF Only'' model is outperformed
by the ``Escalation Only'' model. As the former only uses annual inputs,
but the data at hand are the 1-month or 6-months level, this is not
surprising.

\begin{quote}
Overall, our results suggest that while measures of structural risk may
improve predictive performance, the value they add is marginal and
inconsistent. {[}\ldots{}{]} Incorporating PITF thus significantly
reduces or only slightly improves the performance of the escalation
model, regardless of the approach we take.
\end{quote}

The most straightforward method of incorporating the annual, structural
PITF variables--adding them to the predictors of the Escalation RF
model--strictly outperforms the Escalation Only model. Note that the two
other combination models considered are both non-standard, and that the
``PITF Split Population'' model does not in fact actually incorporate
structural information at all. We thus conclude that adding structural
variables actually clearly improves predictive performance.

\hypertarget{the-effect-of-using-smoothed-roc-curves}{%
\subsection{The effect of using smoothed ROC
curves}\label{the-effect-of-using-smoothed-roc-curves}}

\begin{table}[t]

\caption{\label{tab:table1-benefit}Smoothing advantage for BS Table 1: the gain in AUC-ROC when calculated using smoothed ROC curves}
\centering
\begin{tabular}{lrrrrr}
\toprule
Model & Escalation & Quad & Goldstein & CAMEO & Average\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{One-month forecasts}}\\
\hspace{1em}Base specification & 0.07 & 0.02 & 0.00 & 0.02 & 0.00\\
\hspace{1em}Terminal nodes & 0.06 & 0.01 & -0.01 & 0.02 & 0.00\\
\hspace{1em}Sample size & 0.06 & 0.01 & -0.03 & 0.04 & 0.00\\
\hspace{1em}Trees per forest & 0.06 & 0.02 & -0.01 & 0.02 & 0.00\\
\hspace{1em}Training/test sets 1 & 0.09 & 0.02 & -0.01 & 0.02 & 0.00\\
\hspace{1em}Training/test sets 2 & 0.07 & 0.00 & -0.01 & 0.02 & 0.00\\
\hspace{1em}Training/test sets 3 & 0.09 & 0.01 & -0.01 & 0.02 & 0.01\\
\hspace{1em}Coding of DV 1 & 0.06 & 0.02 & 0.00 & 0.03 & 0.00\\
\hspace{1em}Coding of DV 2 & 0.12 & -0.01 & 0.04 & -0.01 & 0.01\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Six-month forecasts}}\\
\hspace{1em}Base specification & 0.05 & -0.01 & -0.01 & -0.02 & -0.02\\
\hspace{1em}Terminal nodes & 0.02 & -0.02 & -0.01 & -0.02 & -0.02\\
\hspace{1em}Sample size & 0.05 & 0.00 & -0.02 & -0.01 & -0.03\\
\hspace{1em}Trees per forest & 0.05 & -0.01 & -0.01 & -0.01 & -0.02\\
\hspace{1em}Training/test sets 1 & 0.04 & -0.01 & -0.01 & -0.02 & -0.02\\
\hspace{1em}Training/test sets 2 & 0.03 & -0.01 & -0.02 & -0.01 & -0.02\\
\hspace{1em}Training/test sets 3 & 0.03 & -0.02 & -0.03 & -0.02 & -0.02\\
\hspace{1em}Coding of DV 1 & 0.05 & -0.03 & -0.01 & -0.01 & -0.02\\
\hspace{1em}Coding of DV 2 & 0.03 & -0.01 & -0.01 & -0.01 & -0.02\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]

\caption{\label{tab:table2-benefit}Smoothing advantage for BS Table 2: the gain in AUC-ROC when calculated using smoothed ROC curves}
\centering
\begin{tabular}{l>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
\toprule
Model & Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{One-month forecasts}}\\
\hspace{1em}Base specification & 0.05 & -0.03 & 0 & -0.02 & 0.02\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Six-month forecasts}}\\
\hspace{1em}Base specification & 0.06 & -0.01 & 0 & 0.01 & 0.00\\
\bottomrule
\end{tabular}
\end{table}

What impact did the ROC smoothing have overall on the results reported
in BS Tables 1 and 2? Tables \ref{table1-benefit} and
\ref{table2-benefit} show the increase in AUC-ROC when using smoothed
ROC curves, compared to the standard non-smoothed AUC-ROC. Positive
values indicate that smoothing increased a model's apparent performance.
The Escalation model is the only model that consistently had a benefit
from smoothing. For all 8 other models, smoothing sometimes gave a
benefit, sometimes not.

A more dramatic difference stands out when we consider the overall
average impact of smoothing across all flavors of a model reported in
the columns of the tables. The Average, Goldstein, Quad, PITF Split
Population, and Weighted by PITF models are slightly hurt by smoothing,
but on the order of less than 0.01 in absolute magnitude. The ``With
PITF Predictors'' model is hurt on the order of 0.02, i.e.~it appears
substantially worse with smoothing. The CAMEO and PITF Only models
benefit slightly, on the order of 0.01 or less. The Escalation Only
model on the other hand has an average boost of 0.058 to it's AUC-ROC
from smoothing. Considering the spread of AUC-ROC values if we compare
across rows in BS Tables 1 and 2, that boost is substantial.

The use of smoothed ROC curves to calculate AUC-ROC values only clearly
benefits the Escalation model, it does so consistently, and by a
considerable margin. All 8 alternative models reported in Tables 1 and 2
do not on average gain when using smoothed ROC curves to calculate AUC.

\hypertarget{how-accurate-were-the-2016-forecasts}{%
\subsection{How accurate were the 2016
forecasts?}\label{how-accurate-were-the-2016-forecasts}}

\hypertarget{additional-concerns}{%
\section{Additional concerns}\label{additional-concerns}}

\hypertarget{random-forest-hyperparameters}{%
\subsection{Random forest
hyperparameters}\label{random-forest-hyperparameters}}

What initially sparked our interest in the paper was the unusual choice
of hyperparameter settings for the random forest models estimated.

\begin{longtable}[]{@{}llll@{}}
\toprule
Hyperparameter & Default heuristic & Default values (Escalation) & B\&S
value\tabularnewline
\midrule
\endhead
type & & classification & regression\tabularnewline
ntree & & 500 & 100,000 or 1e6\tabularnewline
mtry & \texttt{floor(sqrt(ncol(x)))} & 3 & 3\tabularnewline
replace & & true & false\tabularnewline
sampsize & \texttt{nrow(x)} if replace, & 11,869 & 100 or
500\tabularnewline
& else \texttt{ceiling(.632*nrow(x))} & &\tabularnewline
nodesize & 1 for classification & 1 & 1\tabularnewline
maxnodes & & null & 5 or 10\tabularnewline
\bottomrule
\end{longtable}

There is a stark contrast in the default RF settings and the way Blair
and Sambanis use the RF models. The default approach is to train a
relatively small number (ntree; 500) of classification trees, but where
each tree is fairly big in that it is trained on data that has the same
number of rows as the training data, albeit sampled with replacement
(replace is true; sampsize is 11,869), and allowed to grow fairly deep
(this is governed by nodesize (1), which is the minimum size a terminal
node must have). In contrast, Blair and Sambanis grow very extensive
forests with a large number of trees (100,000 compared to 500), but each
tree is very small and shallow; only a 100 rows are sampled from the
training data for each tree, and the trees are constrained to at most 5
terminal nodes (maxnodes).

This approach only works due to the other unorthodox choice, which is to
use regression, not classification, trees. Trying to use classification
trees with the other parameter settings in fact does not work at all
because it is almost guaranteed that a sample of 100 from the 11,869
training data rows with 9 positive cases will only include 0 (negative)
outcomes in the sample. As it is, using regression with a 0 or 1 outcome
produces warnings when estimating the models:

\begin{verbatim}
Warning message:
In randomForest.default(y = as.integer(train_df$incidence_civil_ns_plus1 ==  :
  The response has five or fewer unique values.  Are you sure you want to do regression?
\end{verbatim}

The RF approach BS use is essentially similar to the probability forests
proposed in Malley 2012. It is not clear if this is intentional or not,
as the paper is not cited.

We tried to create versions of the RF models that use the regular
framework for classification forests, but were unable to create models
that approach the BS RFs test set performance.

How were these hyper-parameter values determined? Do they work better
for some models than others?

\hypertarget{auc-roc-sensitivity}{%
\subsection{AUC-ROC sensitivity}\label{auc-roc-sensitivity}}

RNG and variation

\hypertarget{pre-registration}{%
\subsection{Pre-registration}\label{pre-registration}}

\begin{itemize}
\tightlist
\item
  The escalation model specification, prior to looking at data!
\item
  The hyper-parameter tuning strategy.
\end{itemize}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

BS has a clear misunderstanding of the role that out-of-sample
prediction can plan in analysis. On the one hand it can be used for
simple forecasting while on the other it can be used to evaluate the
performance of models, specifically to overfitting and bias. They
further misrepresent cross-validation which is not about fishing for
results, as claimed by BS. The current state-of-the-art uses prediction
and cross-validation most frequently to provide supportive evidence that
is independent of the estimation procedure and the in-sample data. It is
incorrect to paint this procedure as atheoretical since many studies
will have some explanation for how the model was constructed. These
procedures may simply be used to provide evidence for a theoretical
argument. Though that argument may be unconvincing and/or atheoretical.

The BS approach that is advocated is to use theory to guide prediction.
But theory is a quite ambiguous and undefined concept. It is not a
procedure. What they actually do is to create a model with four right
hand side variables that is supposed to capture a complicated
repression-dissent dynamic. The dynamic is probably not linear, but
their model is. There is a wide-ranging literature on this dynamic that
they do not rely on to construct their model. As such their baseline
comparison is hardly a standard bearer for strong theory.

Further, they complete misunderstand the use of ICEWS event data in
current research. They claim that most uses to date have focused on the
quad categories, but this ignores a wide swath of literature
(Steinert-Threlkeld ApsR 2017) and Metternich et al (AJPX 2013) that
uses a sepcific action---such as protest---defined in the CAMEO
ontology. In the AJPS article we hand coded, for example, every actor in
Thailand and focused on an analysis of how those have been interacting.

We encountered several issues in the code underlying the BS analysis.

The issues we encountered are not subjective modeling choices.

When we fix these issue and perform an updated analysis, the conclusions
BS draw all essentially overturn. In other words, BS findings are based
on a faulty analysis, and invalid.

Using the same analysis BS intend to use, we in fact find that:

\begin{itemize}
\tightlist
\item
  the theory-driven Escalation model is outperformed both by the
  low-effort 1,160 predictor all CAMEO model and the Average ensemble
  model
\item
  structural variables substantially improve the Escalation model's
  performance when added to the pool of predictors the underlying random
  forest model draws on
\end{itemize}

\textbf{AB: add a table comparing BS original claims and updated
results.}

\clearpage

\includegraphics{figures/bands-table4.png}

\begin{table}[t]

\caption{\label{tab:unnamed-chunk-2}2016 Confusion Matrices for Six-month Escalation Model.}
\centering
\begin{tabular}{lrrr}
\toprule
header & Observed & 0 & 1\\
\midrule
 & 0 & 134 & 30\\
\cmidrule{2-4}
\multirow{-2}{*}{\raggedright\arraybackslash Assuming Persistence} & 1 & 0 & 0\\
\cmidrule{1-4}
 & 0 & 134 & 28\\
\cmidrule{2-4}
\multirow{-2}{*}{\raggedright\arraybackslash Assuming Change} & 1 & 0 & 2\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]

\caption{\label{tab:table2-N}Number of valid test predictions for each cell in BS Table 2}
\centering
\begin{tabular}{ll>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
\toprule
 & Horizon & Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
\midrule
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Original model-specific cases}}\\
\hspace{1em} & 1 month & 13748 & 13155 & 13461 & 13748 & 13510\\
\cmidrule{2-7}
\hspace{1em} & 6 months & 2366 & 2264 & 2317 & 2366 & 2265\\
\cmidrule{1-7}
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Cases adjusted to common subset}}\\
\hspace{1em} & 1 month & 9811 & 9811 & 9811 & 9811 & 9811\\
\cmidrule{2-7}
\hspace{1em} & 6 months & 1915 & 1915 & 1915 & 1915 & 1915\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]

\caption{\label{tab:table2-full}Replication of BS Table 2 with smoothed/original ROC and with original varying N cases or adjusting for common case set with constant N}
\centering
\begin{tabular}{ll>{\raggedright\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
\toprule
 &  & Smoothed ROC & Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
\midrule
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{Original model-specific cases}}\\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textit{1 month}}\\
\hspace{1em}\hspace{1em} &  & Yes & 0.85 & 0.76 & 0.80 & 0.81 & 0.77\\
\cmidrule{3-8}
\hspace{1em}\hspace{1em} &  & No & 0.78 & 0.77 & 0.80 & 0.80 & 0.75\\
\cmidrule{2-8}
\addlinespace[0.3em]
\multicolumn{8}{l}{\textit{6 months}}\\
\hspace{1em}\hspace{1em} &  & Yes & 0.82 & 0.86 & 0.81 & 0.79 & 0.74\\
\cmidrule{3-8}
\hspace{1em}\hspace{1em} &  & No & 0.77 & 0.86 & 0.81 & 0.82 & 0.74\\
\cmidrule{1-8}
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{Cases adjusted to common subset}}\\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textit{1 month}}\\
\hspace{1em}\hspace{1em} &  & Yes & 0.74 & 0.68 & 0.69 & 0.80 & 0.69\\
\cmidrule{3-8}
\hspace{1em}\hspace{1em} &  & No & 0.69 & 0.71 & 0.69 & 0.81 & 0.67\\
\cmidrule{2-8}
\addlinespace[0.3em]
\multicolumn{8}{l}{\textit{6 months}}\\
\hspace{1em}\hspace{1em} &  & Yes & 0.85 & 0.86 & 0.79 & 0.70 & 0.71\\
\cmidrule{3-8}
\hspace{1em}\hspace{1em} &  & No & 0.79 & 0.87 & 0.80 & 0.69 & 0.71\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-blair2020forecasting}{}%
Blair, Robert A., and Nicholas Sambanis. 2020. ``Forecasting Civil Wars:
Theory and Structure in an Age of `Big Data' and Machine Learning.''
\emph{Journal of Conflict Resolution}.

\end{document}

### 2020-06-03 paper.Rmd
wednesday mdw
some changes

@@
 ```
+# Introduction
+
+@bdm:2011 , @ward:2016 , @montgomery:etal:PS:2012, @chiba:etal:2015, @brandt:2015  @mhw:2012,@ward:etal:2013,@chadefaux:2015,@weidmann:ward:2010,@pilster:bohmelt:2014,@Brandt_etal_2011,@gleditsch:2015,@bennett:stam:2009,@shellman:etal:2013,@hegre:etal:2011,@ward:etal:2010,@freeman:job:1979,@gneiting:raftery:2005
+@choucri:1974,@choucri:robinson:1979,@goldstone:etal:2010,tetlock:etal:2017,
 
@@
 
-- (1) how does the theoretically-driven Escalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories, 
-- (2) does annual, structural information from the PITF instability forecasting model add to the Escalation model's monthly and 6-month predictions,
-- and (3) how accurate were predictions for the first half of 2016 derived from the Escalation model?
+- (1) How does the theoretically-driven rscalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories?
+- (2) Does annual, structural information from the PITF instability forecasting model add to the escalation model's monthly and 6-month predictions?
+- (3) How accurate were predictions for the first half of 2016 derived from the escalation model?
 
@@
 
-To examine the first question, BS compare the test set of the Escalation model to four alternative models.
+To examine the first question, BS compare the test set of the escalation model to four alternative models.
 The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country.
@@
 
-- Escalation: a set of 10 theoretically-informed indicators.
+- Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
 - Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
@@
 These are directed, thus making for 4 total covariates.
-- CAMEO: counts for all CAMEO event codes, thus a total of 1,159 covariates. 
-- Average: unweighted average of the predictions from the 4 models above.
+- CAMEO: counts for all CAMEO event codes, thus a total of 1,159 covariates (CAN THIS BE RIGHT?). 
+- Average: unweighted average of the predictions from the four models briefly described above.
 

### 2020-06-03 paper.tex
wednesday mdw
some changes

@@
 
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+Bueno de Mesquita (2011) , Ward (2016) , Montgomery, Hollenbach, and
+Ward (2012a), Chiba, Metternich, and Ward (2015), Brandt (2015)
+Montgomery, Hollenbach, and Ward (2012b),Ward et al. (2013),Chadefaux
+(2015),Weidmann and Ward (2010),Pilster and Böhmelt (2014),Brandt,
+Freeman, and Schrodt (2011),Gleditsch (2015),Bennett and Stam
+(2009),Shellman, Levey, and Young (2013),Hegre, Hultman, and Nygård
+(2011),Ward, Greenhill, and Bakke (2010),Freeman and Job (1979),Gneiting
+and Raftery (2005) Choucri (1974),Choucri and Robinson (1979),Goldstone
+et al. (2010),tetlock:etal:2017,
+
 \hypertarget{review-of-blair-and-sambanis-2020}{%
@@
   \item
-    how does the theoretically-driven Escalation model compare in
+    How does the theoretically-driven rscalation model compare in
     forecast performance to alternative models not informed specifically
-    by civil war onset theories,
+    by civil war onset theories?
   \end{enumerate}
@@
   \item
-    does annual, structural information from the PITF instability
-    forecasting model add to the Escalation model's monthly and 6-month
-    predictions,
+    Does annual, structural information from the PITF instability
+    forecasting model add to the escalation model's monthly and 6-month
+    predictions?
   \end{enumerate}
 \item
-  and (3) how accurate were predictions for the first half of 2016
-  derived from the Escalation model?
+  \begin{enumerate}
+  \def\labelenumi{(\arabic{enumi})}
+  \setcounter{enumi}{2}
+  \tightlist
+  \item
+    How accurate were predictions for the first half of 2016 derived
+    from the escalation model?
+  \end{enumerate}
 \end{itemize}
@@
 
-To examine the first question, BS compare the test set of the Escalation
+To examine the first question, BS compare the test set of the escalation
 model to four alternative models. The independent variables for the
@@
 \item
-  Escalation: a set of 10 theoretically-informed indicators.
+  Escalation: a set of ten indicators, putatively drawn from a
+  theoretical escalation model.
 \item
@@
   CAMEO: counts for all CAMEO event codes, thus a total of 1,159
-  covariates.
-\item
-  Average: unweighted average of the predictions from the 4 models
-  above.
+  covariates (CAN THIS BE RIGHT?).
+\item
+  Average: unweighted average of the predictions from the four models
+  briefly described above.
 \end{itemize}
@@
 \hypertarget{refs}{}
+\leavevmode\hypertarget{ref-bennett:stam:2009}{}%
+Bennett, D. Scott, and Allan C. Stam. 2009. ``Revisiting Predictions of
+War Duration.'' \emph{Conflict Management and Peace Science} 26 (3):
+256--67.
+
 \leavevmode\hypertarget{ref-blair2020forecasting}{}%
@@
 
+\leavevmode\hypertarget{ref-brandt:2015}{}%
+Brandt, Patrick. 2015. ``Forecasting Conflicts: Long and Short Term
+Predictions Based on Different Training Set Considerations.'' University
+of Texas Dallas; Peace Research Institute Oslo.
+
+\leavevmode\hypertarget{ref-Brandt_etal_2011}{}%
+Brandt, Patrick T., John R. Freeman, and Philip A. Schrodt. 2011. ``Real
+Time, Time Series Forecasting of Inter- and Intra-State Political
+Conflict.'' \emph{Conflict Management and Peace Science} 28 (1): 41--64.
+
+\leavevmode\hypertarget{ref-bdm:2011}{}%
+Bueno de Mesquita, Bruce. 2011. ``A New Model for Predicting Policy
+Choices: Preliminary Tests.'' \emph{Conflict Management and Peace
+Science} 28 (1): 65--85.
+
+\leavevmode\hypertarget{ref-chadefaux:2015}{}%
+Chadefaux, Thomas. 2015. ``Predictably Unpredictable: The Limits of
+Conflict Forecasting.''
+
+\leavevmode\hypertarget{ref-chiba:etal:2015}{}%
+Chiba, Daina, Nils W. Metternich, and Michael D. Ward. 2015. ``Every
+Story Has a Beginning, Middle, and an End (but Not Always in That
+Order): Predicting Duration Dynamics in a Unified Framework.''
+\emph{Political Science Research and Methods} 3 (3): 515--41.
+
+\leavevmode\hypertarget{ref-choucri:1974}{}%
+Choucri, Nazli. 1974. ``Forecasting in International Relations: Problems
+and Prospects.'' \emph{International Interactions} 1 (2): 63--68.
+
+\leavevmode\hypertarget{ref-choucri:robinson:1979}{}%
+Choucri, Nazli, and Thomas W. Robinson, eds. 1979. \emph{Forecasting in
+International Relations: Theory, Methods, Problems, and Prospects}. San
+Francisco, CA: W.H. Freeman.
+
+\leavevmode\hypertarget{ref-freeman:job:1979}{}%
+Freeman, John R., and Brian L. Job. 1979. ``Scientific Forecasts in
+International Relations: Problems of Definition and Epistemology.''
+\emph{International Studies Quarterly} 23 (1). Blackwell Publishing on
+behalf of The International Studies Association: 113--43.
+
+\leavevmode\hypertarget{ref-gleditsch:2015}{}%
+Gleditsch, Kristian Skrede. 2015. ``Predicting Ucdp/Prio Civil Wars:
+Expanding the Inequality and Grievance Model with Event Data.''
+University of Essex; Peace Research Institute Oslo.
+
+\leavevmode\hypertarget{ref-gneiting:raftery:2005}{}%
+Gneiting, Tilman, and Adrian E. Raftery. 2005. ``Weather Forecasting
+with Ensemble Methods.'' \emph{Science} 310: 248--49.
+
+\leavevmode\hypertarget{ref-goldstone:etal:2010}{}%
+Goldstone, Jack A, Robert H Bates, David L Epstein, Ted Robert Gurr,
+Michael B Lustik, Monty G Marshall, Jay Ulfelder, and Mark Woodward.
+2010. ``A Global Model for Forecasting Political Instability.''
+\emph{American Journal of Political Science} 54 (1). Wiley Online
+Library: 190--208.
+
+\leavevmode\hypertarget{ref-hegre:etal:2011}{}%
+Hegre, Håvard, Lisa Hultman, and Håvard Mokleiv Nygård. 2011.
+``Simulating the Effect of Peacekeeping Operations 2010--2035.'' In
+\emph{Social Computing, Behavioral-Cultural Modeling and Prediction},
+edited by John Salerno, Shanchieh Jay Yang, Dana Nau, and Sun-Ki Chai.,
+325--32. Lecture Notes in Computer Science 6589. Springer.
+
+\leavevmode\hypertarget{ref-montgomery:etal:PS:2012}{}%
+Montgomery, Jacob M., Florian M. Hollenbach, and Michael D. Ward. 2012a.
+``Ensemble Predictions of the 2012 Us Presidential Election.'' \emph{PS:
+Political Science \& Politics} 45: 651--54.
+
+\leavevmode\hypertarget{ref-mhw:2012}{}%
+Montgomery, Jacob M., Florian Hollenbach, and Michael D. Ward. 2012b.
+``Improving Predictions Using Ensemble Bayesian Model Averaging.''
+\emph{Political Analysis} 20 (3): 271--91.
+
+\leavevmode\hypertarget{ref-pilster:bohmelt:2014}{}%
+Pilster, Ulrich, and Tobias Böhmelt. 2014. ``Predicting the Duration of
+the Syrian Insurgency.'' \emph{Research \& Politics} 1 (2).
+
+\leavevmode\hypertarget{ref-shellman:etal:2013}{}%
+Shellman, Stephen M., Brian P. Levey, and Joseph K. Young. 2013.
+``Shifting Sands: Explaining and Predicting Phase Shifts by Dissident
+Organizations.'' \emph{Journal of Peace Research} 50 (3). SAGE
+Publications: 319--36.
+
+\leavevmode\hypertarget{ref-ward:2016}{}%
+Ward, Michael D. 2016. ``Can We Predict Politics? Toward What End?''
+\emph{Journal of Global Security Studies} 1 (1): 80--91.
+
+\leavevmode\hypertarget{ref-ward:etal:2010}{}%
+Ward, Michael D., Brian D. Greenhill, and Kristin M. Bakke. 2010. ``The
+Perils of Policy by P-Value: Predicting Civil Conflicts.'' \emph{Journal
+of Peace Research} 47 (4): 363--75.
+
+\leavevmode\hypertarget{ref-ward:etal:2013}{}%
+Ward, Michael D., Nils W. Metternich, Cassy L. Dorff, Max Gallop,
+Florian M. Hollenbach, Anna Schultz, and Simon Weschle. 2013. ``Learning
+from the Past and Stepping into the Future: Toward a New Generation of
+Conflict Prediction.'' \emph{International Studies Review} 16 (4):
+473--644.
+
+\leavevmode\hypertarget{ref-weidmann:ward:2010}{}%
+Weidmann, N.B., and M.D. Ward. 2010. ``Predicting Conflict in Space and
+Time.'' \emph{Journal of Conflict Resolution} 54 (6). SAGE Publications:
+883--901.
+
 \end{document}

### 2020-06-04 paper.Rmd
moar mdw
changes
add local fallback .bib file
so that Rick can knit the file; we will need this eventually anyways for reproducability

@@
     year: 2020
-bibliography: /Users/mdw/Git/whistle/master.bib
+bibliography: 
+  - "`r user <- Sys.info()[['user']]; if (user=='andybega')  '/Users/andybega/Work/whistle/master.bib' else if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else references.bib`"
 ---
@@
 
+@rost:etal:2009,  GLOBUS, 
 # Review of Blair and Sambanis 2020
@@
 
-Lastly, the AUC-ROC values reported in the original BS Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{table2-N}).
+Lastly, the AUC-ROC values reported in the original BS Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}).
 ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available.
@@
 
-Table \ref{table1-smooth} is our replication of BS Table 1 with smoothed AUC-ROC.
+Table \ref{tab:table1-smooth} is our replication of BS Table 1 with smoothed AUC-ROC.
 The results differ slightly from the original BS Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models.
@@
 
-Table \ref{table1-nosmooth} shows a version of BS Table 1 with the conventional non-smoothed ROC curves.
+Table \ref{tab:table1-nosmooth} shows a version of BS Table 1 with the conventional non-smoothed ROC curves.
 The Average model outperforms the Escalation model in 17 out of 18 cases, and the CAMEO model outperforms in 16 of 18 cases, with one tie.
@@
 
-Table \ref{table2-fixed} shows our replication of BS Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common joint subset of tests cases that all models have non-missing predictions for.
-Table \ref{table2-full} further below shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.
+Table \ref{tab:table2-fixed} shows our replication of BS Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common joint subset of tests cases that all models have non-missing predictions for.
+Table \ref{tab:table2-full} further below shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.
 
@@
 What impact did the ROC smoothing have overall on the results reported in BS Tables 1 and 2?
-Tables \ref{table1-benefit} and \ref{table2-benefit} show the increase in AUC-ROC when using smoothed ROC curves, compared to the standard non-smoothed AUC-ROC.
+Tables \ref{tab:table1-benefit} and \ref{tab:table2-benefit} show the increase in AUC-ROC when using smoothed ROC curves, compared to the standard non-smoothed AUC-ROC.
 Positive values indicate that smoothing increased a model's apparent performance.

### 2020-06-04 paper.Rmd
replace "BS" with "B&S"

@@
 @rost:etal:2009,  GLOBUS,
+
 # Review of Blair and Sambanis 2020
 
-@blair2020forecasting (BS for the sake of brevity hereafter) aim to examine whether theory adds value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
+@blair2020forecasting (B&S for the sake of brevity hereafter) aim to examine whether theory adds value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
 They find that it does.
@@
 
-BS specifically examine three questions:
+B&S specifically examine three questions:
 
@@
 
-To assess the first two questions BS use global datasets covering all major countries from 2001 to 2015.
+To assess the first two questions B&S use global datasets covering all major countries from 2001 to 2015.
 Two versions of the dataset are used, one at the country-month level, the other aggregated to 6-month half-years.
@@
 
-The corresponding results for each question are shown in BS Tables 1 and 2, which we will replicate further below.
+The corresponding results for each question are shown in B&S Tables 1 and 2, which we will replicate further below.
 
-To examine the first question, BS compare the test set of the escalation model to four alternative models.
+To examine the first question, B&S compare the test set of the escalation model to four alternative models.
 The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country.
@@
 
-The corresponding results are shown in BS Table 2.
+The corresponding results are shown in B&S Table 2.
 
-Finally, BS used the Escalation model to create forcasts for the first half of 2016, and in their third and final analysis, they score the forecasts accuracy using civil war onset data later observed.
-This is summarized in BS Table 3.
+Finally, B&S used the Escalation model to create forcasts for the first half of 2016, and in their third and final analysis, they score the forecasts accuracy using civil war onset data later observed.
+This is summarized in B&S Table 3.
 
@@
 
-While attempting to replicate BS's results, we found several issues.
+While attempting to replicate B&S's results, we found several issues.
 
@@
 
-The most consequential issue that we found is that all AUC-ROC values reported in BS Tables 1 and 2 are calculated using smoothed ROC curves, not the original, actual ROC curves.
+The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, actual ROC curves.
 A reference to smoothing is made in a single sentence in the paper (p.
@@
 
-Figure 1 shows our replication of both the smoothed ROC curves BS report, and the actual ROC curves on the right.
+Figure 1 shows our replication of both the smoothed ROC curves B&S report, and the actual ROC curves on the right.
 
-![Replication of BS Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)
+![Replication of B&S Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)
 
@@
 In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached.
-For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in BS Table 1, there are only 11 civil war onset cases in the test set.
+For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set.
 Thus the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.
@@
 As we will show in the results further below, the difference between the smoothed and original ROC AUC values is up to 0.12--a huge difference given that row-wise, the models being compared typically differ by only 0.05 or less--and differentially impacts the models that are being compared.
-In fact, BS's original results and interpretation are entirely conditional on the use of smoothed AUC-ROC.
+In fact, B&S's original results and interpretation are entirely conditional on the use of smoothed AUC-ROC.
 
@@
 We are in fact not aware of other work that uses smoothed ROC curves for AUC calculations.
-It might be that there are theoretical reasons justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the BS results, it minimally would have warranted an explicit discussion in the paper.
+It might be that there are theoretical reasons justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the B&S results, it minimally would have warranted an explicit discussion in the paper.
 This is not the case, and it is only mentioned in the sentence we quote above.
@@
 
-The "Weighted by PITF" model is described as follows in BS, page 19:
+The "Weighted by PITF" model is described as follows in B&S, page 19:
 
@@
 
-Similarly, the "PITF Split Population" model appears to be incorrectly implemented. BS describe it as, on page 20:
+Similarly, the "PITF Split Population" model appears to be incorrectly implemented. B&S describe it as, on page 20:
 
@@
 they use the same *x* variables and the same random forest hyper-parameter settings.
-The *only* difference in the models as they are implemented in the BS replication code is due to the non-deterministic nature of the random forest model itself.
+The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself.
 If we ran both with the same RNG seed, they would be identical in every respect, producing identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined.
@@
 
-Lastly, the AUC-ROC values reported in the original BS Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}).
+Lastly, the AUC-ROC values reported in the original B&S Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}).
 ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available.
@@
 Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
-In other words, the results for different models in BS Table 2 are actually not comparable to the other, and any conclusions drawn from such comparison are potentially incorrect.
+In other words, the results for different models in B&S Table 2 are actually not comparable to the other, and any conclusions drawn from such comparison are potentially incorrect.
 
 We fix this issue by only using predictions for common joint subset of cases that all models have non-missing predictions for.
-The original BS Table 2 1-month have *N*=11,806--12,495 versus a common joint subset of 9,811, and for the 6-month row *N*=2,070--2,233 versus *N*=1,915 for the common joint subset.
+The original B&S Table 2 1-month have *N*=11,806--12,495 versus a common joint subset of 9,811, and for the 6-month row *N*=2,070--2,233 versus *N*=1,915 for the common joint subset.
 
@@
 
-BS show a confusiong matrix to score their 2016-H1 forecasts in Table 4.
+B&S show a confusiong matrix to score their 2016-H1 forecasts in Table 4.
 Although the forecasts are for the probability of civil war onset, in the replication code they are actually scored using the much more common incidence of civil war, i.e.
@@
 
-Table \ref{tab:table1-smooth} is our replication of BS Table 1 with smoothed AUC-ROC.
-The results differ slightly from the original BS Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models.
-It is the case that BS set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel.
+Table \ref{tab:table1-smooth} is our replication of B&S Table 1 with smoothed AUC-ROC.
+The results differ slightly from the original B&S Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models.
+It is the case that B&S set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel.
 In any case, the interpretation of results should not be sensitive to random variation, i.e.
 it should not depend on using a specific RNG seed.
-On the basis of these results, BS conclude that the Escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.
+On the basis of these results, B&S conclude that the Escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.
 
@@
     digits = 2, booktabs = TRUE, 
-    caption = "Replication of BS Table 1 with smoothed ROC curves; test set AUC-ROC for various models",
+    caption = "Replication of B\\&S Table 1 with smoothed ROC curves; test set AUC-ROC for various models",
     label = "table1-smooth") %>% 
@@
 
-Table \ref{tab:table1-nosmooth} shows a version of BS Table 1 with the conventional non-smoothed ROC curves.
+Table \ref{tab:table1-nosmooth} shows a version of B&S Table 1 with the conventional non-smoothed ROC curves.
 The Average model outperforms the Escalation model in 17 out of 18 cases, and the CAMEO model outperforms in 16 of 18 cases, with one tie.
@@
 The Quad model appears to be roughly on par with the Escalation model.
-Thus the original BS conclusion that the Escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.
+Thus the original B&S conclusion that the Escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.
 
@@
     digits = 2, booktabs = TRUE, 
-    caption = "Replication of BS Table 1 \\textit{without} smoothed ROC curves; test set AUC-ROC for various models",
+    caption = "Replication of B\\&S Table 1 \\textit{without} smoothed ROC curves; test set AUC-ROC for various models",
     label = "table1-nosmooth") %>% 
@@
 
-Table \ref{tab:table2-fixed} shows our replication of BS Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common joint subset of tests cases that all models have non-missing predictions for.
+Table \ref{tab:table2-fixed} shows our replication of B&S Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common joint subset of tests cases that all models have non-missing predictions for.
 Table \ref{tab:table2-full} further below shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.
@@
     digits = 2, booktabs = TRUE, 
-    caption = "Replication of BS Table 2: Test set AUC-ROC for Escalation model with and without structural PITF contribution",
+    caption = "Replication of B\\&S Table 2: Test set AUC-ROC for Escalation model with and without structural PITF contribution",
     label = "table2-fixed") %>%
@@
   pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1)) %>%
-  footnote("Differences from the original BS Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.", footnote_as_chunk = TRUE, threeparttable = TRUE)
+  footnote("Differences from the original B\\&S Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.", footnote_as_chunk = TRUE, threeparttable = TRUE)
 ```
 
-BS interpret the results as follows, on page 20:^[We list the "Overall, ..." interpretation out of order, last, for clarity.]
+B&S interpret the results as follows, on page 20:^[We list the "Overall, ..." interpretation out of order, last, for clarity.]
 
@@
     digits = 2, booktabs = TRUE, 
-    caption = "Smoothing advantage for BS Table 1: the gain in AUC-ROC when calculated using smoothed ROC curves",
+    caption = "Smoothing advantage for B\\&S Table 1: the gain in AUC-ROC when calculated using smoothed ROC curves",
     label = "table1-benefit") %>% 
@@
     digits = 2, booktabs = TRUE, 
-    caption = "Smoothing advantage for BS Table 2: the gain in AUC-ROC when calculated using smoothed ROC curves",
+    caption = "Smoothing advantage for B\\&S Table 2: the gain in AUC-ROC when calculated using smoothed ROC curves",
     label = "table2-benefit") %>% 
@@
 
-What impact did the ROC smoothing have overall on the results reported in BS Tables 1 and 2?
+What impact did the ROC smoothing have overall on the results reported in B&S Tables 1 and 2?
 Tables \ref{tab:table1-benefit} and \ref{tab:table2-benefit} show the increase in AUC-ROC when using smoothed ROC curves, compared to the standard non-smoothed AUC-ROC.
@@
 The Escalation Only model on the other hand has an average boost of 0.058 to it's AUC-ROC from smoothing.
-Considering the spread of AUC-ROC values if we compare across rows in BS Tables 1 and 2, that boost is substantial.
+Considering the spread of AUC-ROC values if we compare across rows in B&S Tables 1 and 2, that boost is substantial.
 
@@
 What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated.
+Table \ref{tab:hp} shows the default values used by the implementation of random forest that B&S use (from the **randomForest** R package), in contrast to the basic settings used by B&S for most the models reported in the paper.
 
-|Hyperparameter|Default heuristic|Default values (Escalation)|B&S value|
-|--------------|-----------------|--------------------------|---------|
-|type          |                 | classification           | regression|
-|ntree         |                 | 500                      | 100,000 or 1e6 |
-|mtry          |`floor(sqrt(ncol(x)))`|3|3|
-|replace       |                 | true | false |
-|sampsize | `nrow(x)` if replace, | 11,869 | 100 or 500 |
-||else `ceiling(.632*nrow(x))`|||
-|nodesize | 1 for classification | 1 | 1 |
-|maxnodes | | null | 5 or 10 |
+As the outcome is a binary indicator of civil war onset, one would typically use a classification random forest that predicts 0 or 1 labels directly.
+The implementation of random forests that B&S use ([@liaw:wiener:2002]) is based on the original @breiman:2001 implementation and calculates predictived probabilities by averaging over the "0" or "1" votes from all constituent decision trees.
+The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number.
+From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive, and operates on a large bootstrapped sample of the original training data.
+These are of course only heuristics and it is usual to attempt to find better hyper-parameter methods through some form of tuning.
 
-There is a stark contrast in the default RF settings and the way Blair and Sambanis use the RF models.
-The default approach is to train a relatively small number (ntree; 500) of classification trees, but where each tree is fairly big in that it is trained on data that has the same number of rows as the training data, albeit sampled with replacement (replace is true; sampsize is 11,869), and allowed to grow fairly deep (this is governed by nodesize (1), which is the minimum size a terminal node must have).
-In contrast, Blair and Sambanis grow very extensive forests with a large number of trees (100,000 compared to 500), but each tree is very small and shallow; only a 100 rows are sampled from the training data for each tree, and the trees are constrained to at most 5 terminal nodes (maxnodes).
+\begin{table}
+\caption{\label{tab:hp} Random forest (\texttt{randomForest()}) default versus B\&S hyperparameters}
+\begin{tabular}{l>{\raggedright\arraybackslash}p{2in}ll}
+\toprule
+Hyperparameter & Default heuristic & Default values (Escalation) & B\&S value \\
+\midrule
+type & & classification & regression \\
+ntree & & 500 & 100,000 or 1e6 \\
+mtry & \texttt{floor(sqrt(ncol(x)))} & 3 & 3 \\
+replace & & true & false \\
+sampsize & \texttt{nrow(x)} if replace, else \texttt{ceiling(.632*nrow(x))} & 11,869 & 100 or
+500 \\
+nodesize & 1 for classification & 1 & 1 \\
+maxnodes & & null & 5 or 10 \\
+\bottomrule
+\end{tabular}
+\end{table}
 
-This approach only works due to the other unorthodox choice, which is to use regression, not classification, trees.
-Trying to use classification trees with the other parameter settings in fact does not work at all because it is almost guaranteed that a sample of 100 from the 11,869 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample.
+B&S in contrast fit very large forests with 100,000 trees in the basic model form, but where each tree only operates on a very small sub-sample (N=100 or 500), drawn without replacement,  of the available training data.
+This approach only works due to the choice to use regression, not classification, trees.
+Trying to use classification trees with the other parameter settings not work at all because it is almost guaranteed that a sample of 100 from the ~11,000 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample.
 As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:
@@
 
-The RF approach BS use is essentially similar to the probability forests proposed in Malley 2012. It is not clear if this is intentional or not, as the paper is not cited.
+As it turns out, using regression random forests for this kind of binary classification problem in order to obtain probability estimates matches the probability random forest approach suggested and positively evaluated in @malley:etal:2012, and which is used in another prominent R implementation of random forests.^[The **ranger** package.] It is not clear whether this is intentional, as the Malley paper is not cited in B&S.
 
-We tried to create versions of the RF models that use the regular framework for classification forests, but were unable to create models that approach the BS RFs test set performance.
+In any case, B&S's random forest approach appears to work really well.
+We tried to construct classification random forests tuned via cross-validation on the training data set partition, i.e.
+without touching the test data, but were unable to develop models that consistently match the B&S random forest method in both cross-validated out-of-sample training predictions and test set predictions.
 
-How were these hyper-parameter values determined? Do they work better for some models than others?
+Given that they are relatively unorthodox, yet appear to work very well, we wonder how the hyper-parameter values were determined.
+Two specific concerns are that this was not done with an eye towards test set accuracy, which would invalidate the independence of the out-of-sample test set, and whether the specific hyper-parameter values are optimized for only one model, or were optimized and found to work well for all models.
+There is no discussion of the random forest tuning strategy or how the specific hyper-parameter methods were determined in the paper.
 
@@
 
-BS has a clear misunderstanding of the role that out-of-sample prediction can plan in analysis.
+B&S has a clear misunderstanding of the role that out-of-sample prediction can plan in analysis.
 On the one hand it can be used for simple forecasting while on the other it can be used to evaluate the performance of models, specifically to overfitting and bias.
-They further misrepresent cross-validation which is not about fishing for results, as claimed by BS.
+They further misrepresent cross-validation which is not about fishing for results, as claimed by B&S.
 The current state-of-the-art uses prediction and cross-validation most frequently to provide supportive evidence that is independent of the estimation procedure and the in-sample data.
@@
 
-The BS approach that is advocated is to use theory to guide prediction.
+The B&S approach that is advocated is to use theory to guide prediction.
 But theory is a quite ambiguous and undefined concept.
@@
 
-We encountered several issues in the code underlying the BS analysis.
+We encountered several issues in the code underlying the B&S analysis.
 
@@
 
-When we fix these issue and perform an updated analysis, the conclusions BS draw all essentially overturn. In other words, BS findings are based on a faulty analysis, and invalid.
+When we fix these issue and perform an updated analysis, the conclusions B&S draw all essentially overturn. In other words, B&S findings are based on a faulty analysis, and invalid.
 
-Using the same analysis BS intend to use, we in fact find that:
+Using the same analysis B&S intend to use, we in fact find that:
 
@@
 
-**AB: add a table comparing BS original claims and updated results.**
+**AB: add a table comparing B&S original claims and updated results.**
 
@@
     "latex", booktabs = TRUE, digits = 0,
-    caption = "Number of valid test predictions for each cell in BS Table 2",
+    caption = "Number of valid test predictions for each cell in B\\&S Table 2",
     label = "table2-N") %>%
@@
     "latex", booktabs = TRUE, digits = 2,
-               caption = "Replication of BS Table 2 with smoothed/original ROC and with original varying N cases or adjusting for common case set with constant N",
+               caption = "Replication of B\\&S Table 2 with smoothed/original ROC and with original varying N cases or adjusting for common case set with constant N",
     label = "table2-full") %>%

### 2020-06-04 paper.Rmd
update Table 4 replication

@@
 
-\clearpage
+```{r, message=FALSE}
+tab4o <- read_csv("data/table4-original.csv")
+tab4f <- read_csv("data/table4-fixed.csv")
 
-![](figures/bands-table4.png)
+tab4o <- tab4o %>%
+  rename(header = type) %>%
+  mutate(source = "Original, scored with civil war incidence")
+tab4f <- tab4f %>%
+  rename(Predicted0 = `0`, Predicted1 = `1`) %>%
+  mutate(source = "Fixed, scored with civil war onset")
 
-```{r, message=FALSE}
-tab4 <- read_csv("data/tab4-fixed.csv")
+tab4 <- bind_rows(tab4o, tab4f) %>%
+  select(source, header, everything())
 
@@
   knitr::kable(
-    caption = "2016 Confusion Matrices for Six-month Escalation Model.",
+    caption = "Replication of B\\&S Table 4: 2016 Confusion Matrices for Six-month Escalation Model.",
     booktabs = TRUE
   ) %>%
-  kableExtra::collapse_rows(1)
+  kableExtra::collapse_rows(1:2, row_group_label_position = "stack")
 ```

### 2020-06-04 table4/README.Rmd
update Table 4 replication

---
title: "Table 4 investigation"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(readstata13)
library(ggplot2)
library(tidyr)
library(readr)
```

```{r}
data_6month_oos <- read.dta13("data/6mo_data_OOS.dta")

data_6month_oos %>% count(incidence_civil_ns_plus1)
data_6month_oos %>% count(incidence_civil_ns )

pred_escalation_6mo_inc_civil_ns <- read.dta13("data/6mo_predictions_escalation_OOS.dta")

# this object contains OOS forecasts from 2008 to 2016
as_tibble(pred_escalation_6mo_inc_civil_ns)

fcast2016 <- pred_escalation_6mo_inc_civil_ns %>%
  filter(year==2016)

fcast2016 %>% count(incidence_civil_ns_plus1)
fcast2016 %>% count(incidence_civil_ns )

fcast2016 %>%
  filter(is.na(incidence_civil_ns)) %>%
  dplyr::select(year, country)

# Ok, indeed it looks like "incidence_civil_ns" was originally incidence and
# then recoded to onset, with NA for continuing conflicts
data_6month_oos %>%
  ggplot(aes(x = year, y = factor(country_iso3), fill = factor(incidence_civil_ns))) +
  geom_tile()

# How does this look in the full data?
data_6month <- read.dta13("data/6mo_data_OOS.dta")
data_6month %>%
  ggplot(aes(x = year, y = factor(country_iso3), fill = factor(incidence_civil_ns))) +
  geom_tile()

data_6month %>%
  ggplot(aes(x = year, y = factor(country_iso3), fill = factor(incidence_civil_ns_plus1))) +
  geom_tile()

# What's going on with the positive cases here?
dv <- data_6month %>%
  filter(incidence_civil_ns==1) %>%
  dplyr::select(country_name, country_iso3, year, period, incidence_civil_ns)
dv_plus1 <- data_6month %>%
  filter(incidence_civil_ns_plus1==1) %>%
  dplyr::select(country_name, country_iso3, year, period, incidence_civil_ns_plus1)
both_dvs <- full_join(dv_plus1, dv) %>%
  arrange(country_name, year)

ggplot(data_6month, aes(x = period, y = factor(country_iso3),
                        fill = factor(incidence_civil_ns_plus1))) +
  geom_tile()

# Predictions
preds <- read.dta13("data/6mo_predictions_escalation_OOS.dta") %>%
  as_tibble() %>%
  filter(period==max(period)) %>%
  mutate(incidence_civil_ns = as.integer(incidence_civil_ns),
         incidence_civil_ns_plus1 = as.integer(incidence_civil_ns_plus1))

# Hand-code the Table 3 predictions;
# The predictions don't exactly match what is in Table 3, possibley because I
# only ran the 6 month OOS portion of +master.R after setting the seed at the
# top.
pred1 <- c("Nigeria", "India", "Iraq", "Somalia", "Syria", "Pakistan",
           "Philippines", "Turkey", "Afghanistan", "Russia", "Burundi", "Egypt",
           "Yemen", "Colombia", "Mali", "China", "Indonesia", "Ukraine",
           "Sudan", "Lebanon", "Thailand", "Iran", "Myanmar", "Montenegro",
           "Bangladesh", "Niger", "El Salvador", "France", "Ghana", "Tajikistan")
codes <- countrycode::countrycode(pred1, "country.name", "iso3c")
preds$label <- as.integer(preds$country %in% codes)
# should be 30
sum(preds$label)

# What is the truth data in preds?
preds %>% count(incidence_civil_ns)
preds %>% count(incidence_civil_ns_plus1)

# Does this match the OOS data?
data_6month_oos %>% filter(period==31) %>% count(incidence_civil_ns)
data_6month_oos %>% filter(period==31) %>% count(incidence_civil_ns_alt1_plus1)

tab4_top <- preds %>%
  rename(Observed = incidence_civil_ns, Predicted = label) %>%
  replace_na(list(Observed = 0L)) %>%
  mutate(Observed = factor(Observed, levels = c("0", "1")),
         Predicted = factor(Predicted, levels = c("0", "1"))) %>%
  group_by(Observed, Predicted) %>%
  dplyr::summarize(n = n()) %>%
  ungroup() %>%
  tidyr::complete(Observed, Predicted, fill = list(n = 0)) %>%
  mutate(header = "Assuming Persistence")

# From 6mo_make_confusion_matrix.do, for the bottom of Table 3:
# replace incidence_civil_ns_alt=0 if country_name=="Colombia"
# replace incidence_civil_ns_alt=1 if country_name=="Turkey"
# replace incidence_civil_ns_alt=1 if country_name=="Burundi"
preds %>%
  filter(country %in% c("COL", "TUR", "BDI"))

data_6month_oos %>%
  filter(country_iso3 %in% c("COL", "TUR", "BDI")) %>%
  dplyr::select(country_iso3, year, incidence_civil_ns, incidence_civil_ns_plus1)

# Burundi has no ongoing conflict, so this is new onset
# Colombia has ongong conflict, so no difference
# Turkey has no ongoing conflict, so this is new onset
preds_v2 <- preds %>%
  mutate(incidence_civil_ns = case_when(
    country=="TUR" ~ 1L,
    country=="BDI" ~ 1L,
    TRUE ~ incidence_civil_ns
  ))

tab4_bottom <- preds_v2 %>%
  rename(Observed = incidence_civil_ns, Predicted = label) %>%
  replace_na(list(Observed = 0L)) %>%
  mutate(Observed = factor(Observed, levels = c("0", "1")),
         Predicted = factor(Predicted, levels = c("0", "1"))) %>%
  group_by(Observed, Predicted) %>%
  dplyr::summarize(n = n()) %>%
  ungroup() %>%
  tidyr::complete(Observed, Predicted, fill = list(n = 0)) %>%
  mutate(header = "Assuming Change")

tab4 <- bind_rows(tab4_top, tab4_bottom) %>%
  dplyr::select(header, everything()) %>%
  pivot_wider(names_from = "Predicted", values_from = n)

write_csv(tab4, "output/table4-fixed.csv")

tab4
```

### 2020-06-04 table4-investigation.R

@@ -1,132 +1,0 @@
-
-
-data_6month_oos <- read.dta13("data/6mo_data_OOS.dta")
-
-data_6month_oos %>% count(incidence_civil_ns_plus1)
-data_6month_oos %>% count(incidence_civil_ns )
-
-# this object contains OOS forecasts from 2008 to 2016
-as_tibble(pred_escalation_6mo_inc_civil_ns)
-
-fcast2016 <- pred_escalation_6mo_inc_civil_ns %>%
-  filter(year==2016)
-
-fcast2016 %>% count(incidence_civil_ns_plus1)
-fcast2016 %>% count(incidence_civil_ns )
-
-fcast2016 %>%
-  filter(is.na(incidence_civil_ns)) %>%
-  dplyr::select(year, country)
-
-# Ok, indeed it looks like "incidence_civil_ns" was originally incidence and
-# then recoded to onset, with NA for continuing conflicts
-data_6month_oos %>%
-  ggplot(aes(x = year, y = factor(country_iso3), fill = factor(incidence_civil_ns))) +
-  geom_tile()
-
-# How does this look in the full data?
-data_6month <- read.dta13("data/6mo_data_OOS.dta")
-data_6month %>%
-  ggplot(aes(x = year, y = factor(country_iso3), fill = factor(incidence_civil_ns))) +
-  geom_tile()
-
-data_6month %>%
-  ggplot(aes(x = year, y = factor(country_iso3), fill = factor(incidence_civil_ns_plus1))) +
-  geom_tile()
-
-# What's going on with the positive cases here?
-dv <- data_6month %>%
-  filter(incidence_civil_ns==1) %>%
-  dplyr::select(country_name, country_iso3, year, period, incidence_civil_ns)
-dv_plus1 <- data_6month %>%
-  filter(incidence_civil_ns_plus1==1) %>%
-  dplyr::select(country_name, country_iso3, year, period, incidence_civil_ns_plus1)
-both_dvs <- full_join(dv_plus1, dv) %>%
-  arrange(country_name, year)
-
-ggplot(data_6month, aes(x = period, y = factor(country_iso3),
-                        fill = factor(incidence_civil_ns_plus1))) +
-  geom_tile()
-
-
-
-# Predictions
-preds <- read.dta13("data/6mo_predictions_escalation_OOS.dta") %>%
-  as_tibble() %>%
-  filter(period==max(period)) %>%
[…]

### 2020-06-04 table4/table4-investigation.R

@@ -1,0 +1,132 @@
+
+
+data_6month_oos <- read.dta13("data/6mo_data_OOS.dta")
+
+data_6month_oos %>% count(incidence_civil_ns_plus1)
+data_6month_oos %>% count(incidence_civil_ns )
+
+# this object contains OOS forecasts from 2008 to 2016
+as_tibble(pred_escalation_6mo_inc_civil_ns)
+
+fcast2016 <- pred_escalation_6mo_inc_civil_ns %>%
+  filter(year==2016)
+
+fcast2016 %>% count(incidence_civil_ns_plus1)
+fcast2016 %>% count(incidence_civil_ns )
+
+fcast2016 %>%
+  filter(is.na(incidence_civil_ns)) %>%
+  dplyr::select(year, country)
+
+# Ok, indeed it looks like "incidence_civil_ns" was originally incidence and
+# then recoded to onset, with NA for continuing conflicts
+data_6month_oos %>%
+  ggplot(aes(x = year, y = factor(country_iso3), fill = factor(incidence_civil_ns))) +
+  geom_tile()
+
+# How does this look in the full data?
+data_6month <- read.dta13("data/6mo_data_OOS.dta")
+data_6month %>%
+  ggplot(aes(x = year, y = factor(country_iso3), fill = factor(incidence_civil_ns))) +
+  geom_tile()
+
+data_6month %>%
+  ggplot(aes(x = year, y = factor(country_iso3), fill = factor(incidence_civil_ns_plus1))) +
+  geom_tile()
+
+# What's going on with the positive cases here?
+dv <- data_6month %>%
+  filter(incidence_civil_ns==1) %>%
+  dplyr::select(country_name, country_iso3, year, period, incidence_civil_ns)
+dv_plus1 <- data_6month %>%
+  filter(incidence_civil_ns_plus1==1) %>%
+  dplyr::select(country_name, country_iso3, year, period, incidence_civil_ns_plus1)
+both_dvs <- full_join(dv_plus1, dv) %>%
+  arrange(country_name, year)
+
+ggplot(data_6month, aes(x = period, y = factor(country_iso3),
+                        fill = factor(incidence_civil_ns_plus1))) +
+  geom_tile()
+
+
+
+# Predictions
+preds <- read.dta13("data/6mo_predictions_escalation_OOS.dta") %>%
+  as_tibble() %>%
+  filter(period==max(period)) %>%
[…]

### 2020-06-04 update-data.R

@@ -1,0 +1,24 @@
+#
+#  Copy/update the files in data/ that are used in paper.Rmd
+#
+
+library(here)
+
+setwd(here::here())
+
+files_needed <- c("rep_nosmooth/output/tables/table1-nosmooth.csv",
+                  "rep_nosmooth/output/tables/table1-smooth.csv",
+                  "rep_nosmooth/output/tables/table1-smooth-benefit.csv",
+                  "rep_nosmooth/output/tables/table2-for-appendix.csv",
+                  "rep_nosmooth/output/tables/table2-nosmooth.csv",
+                  "rep_nosmooth/output/tables/table2-smooth-benefit.csv",
+                  "rep_nosmooth/output/tables/table2-smooth.csv",
+                  "rep_nosmooth/output/tables/table4-original.csv",
+                  "table4/output/table4-fixed.csv")
+
+success <- file.copy(files_needed, "data", overwrite = TRUE)
+
+if (any(success==FALSE)) {
+  warning(sprintf("Could not find and copy file(s):\n%s",
+                  paste0(files_needed[!success], collapse = "\n")))
+}

### 2020-06-04 paper.Rmd
End of day two
Not as much progress as needed, but were getting there.  Check out the last sentence in the firs paragraph.

@@
 
-@bdm:2011 , @ward:2016 , @montgomery:etal:PS:2012, @chiba:etal:2015, @brandt:2015  @mhw:2012,@ward:etal:2013,@chadefaux:2015,@weidmann:ward:2010,@pilster:bohmelt:2014,@Brandt_etal_2011,@gleditsch:2015,@bennett:stam:2009,@shellman:etal:2013,@hegre:etal:2011,@ward:etal:2010,@freeman:job:1979,@gneiting:raftery:2005
-@choucri:1974,@choucri:robinson:1979,@goldstone:etal:2010,tetlock:etal:2017,
-
-@rost:etal:2009,  GLOBUS,
+Blair and Sambanis (2020; hereafter B&S) argue that theory is important for creating models that have high accuracy in forecasting civil war onset.
+Indeed they argue that with such theory, forecasting is more accurate than has previously been possible.
+Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in their article.
+We find that these claims are false.
+Their theory-based escalation model does not do better than the alternatives that they examine, indeed it does worse.
+The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure: 1) they use smoothed performance curves instead of the unsmoothed curves, 2) they mistakenly encode the dependent variable to be incidents of civil war, rather than as they claim the onset of a civil war, 3) they mistakenly calculate the predicted values for the alternative ``non-theoretical'' models that they compare to.
+In what follows we show the impact of these mistakes on the conclusions, we examine their understanding of prediction it self, and we explore the logic of their assertion about theory as well.
+We believe that this research by B&S is incorrect and should be retracted.
 
@@
 
+@bdm:2011 , @ward:2016 , @montgomery:etal:PS:2012, @chiba:etal:2015, @brandt:2015  @mhw:2012,@ward:etal:2013,@chadefaux:2015,@weidmann:ward:2010,@pilster:bohmelt:2014,@Brandt_etal_2011,@gleditsch:2015,@bennett:stam:2009,@shellman:etal:2013,@hegre:etal:2011,@ward:etal:2010,@freeman:job:1979,@gneiting:raftery:2005
+@choucri:1974,@choucri:robinson:1979,@goldstone:etal:2010,tetlock:etal:2017,
+@rost:etal:2009,  
+@hopple:andriole:1984,
+@andriole:young:1977,
+@brandt:etal:2011,
+@brandt:etal:2014,
+@mellers:etal:2015,
+@tetlock:gardner:2015,
+@hegre:etal:2019,
+@weidmann:ward:2010, 
+@beger:ward:2017,
+@schrodt:1991,
+@obrien:2010,
+@obrien:2002,
+@hegre:etal:2019,
+@hegre:etal:2017,
+@hegre:etal:2016,
+@chiba:gleditsch:2017,
+@buhaug:etal:2014,
+@colaresi:mahmood:2017,
+@ramakrishnan:etal:2014,
+@schleussner:etal2016,
+@kupilik:witmer:2018,
+@witmer:etal:2017
+@oloughlin:etal:2012
+@hegre:etal:2013 
+@buhaug:gleditsch:2008
+
 **AB: add a table comparing B&S original claims and updated results.**
@@
 ```
+\newpage
 

### 2020-06-04 paper.Rmd
end of day two
did a bit more...

@@
 We believe that this research by B&S is incorrect and should be retracted.
+
+B&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars ``remarkable accurately.'' Indeed they argue that this so-called theoretical model outperforms four other ``more mechanical'' alternatives.
+Second, they clam that the integration of structure with process are better over short forecasting windows.
+Third they preregisted the list of thirty countries which have the highest risk of civil war onset.
+They claim that such prospective predictions are rare in the literature when in fact they have been routine for many years with several prominent projects.
+B&S claim to be unique in assessing these forecasts.
+A qualitative analysis of their predictions allows them to conclude that their model is strong and a more precise test is undertaken.
+We will return to their analysis later, after correcting the procedural mistakes we found in their research process.
+
+Before proceeding, we quote B&S (page 24):
+\begin{quote}
+Our theoretically driven model generates accurate forecasts, with base specification
+AUCs of .82 and .85 over one- and six-month windows, respectively, and AUCs
+as high as .92 in other specifications. Our model also consistently and sometimes
+dramatically outperforms the alternatives we test...Cederman and Weidmann (2017, 476) argue that ``the hope that
+big data will somehow yield valid forecasts through theory-free ‘brute force’ is
+misplaced in the area of political violence.'' Our results lend some credence to this
+claim.
+\end{quote}
 

### 2020-06-05 rep_nosmooth/README.Rmd
add note checking RNG seed variance
How much does AUC-ROC vary between model runs when you don't see RNG seeds? Generally much less than 0.01. This is with the base spec escalation model, anecdotally, some of the other models, especially in Table 2, can vary more.

@@
 
-In order to be able to run the models needed to create tables 1 and 2 more quickly than with the original replication code, the code here has been refactored to enable running in parallel.
+In order to be able to run the models needed to create tables 1 and 2 more quickly than with the original replication code, the code here has been re-factored to enable running in parallel.
 Here is a summary of the major changes:
@@
 Adjust the number of WORKERS at the top of the script as needed based on the number of cores on your computer.
-The primary output of this script is `output/model-table-w-results.csv`
-- `recreate-tables.R` recreates the original and non-smooth ROC versions of Tables 1 and 2.
+The primary output of this script is `output/model-table-w-results.csv` and `output/all-predictions.R`.
+- `recreate-tables.R` recreates the original and non-smooth ROC versions of Tables 1 and 2 using `output/model-table-w-results.csv` and `output/all-predictions.R`.
+
+In addition, there are some ancillary materials.
+
+- [variance.md](/variance.md) examines how much the AUC-ROC results vary when you don't set an RNG seed.
+For the base specification escalation model, the variance is significantly below 0.01, and small changes in the Table 1 and 2 replications mainly occur when the distribution straddles a threshold used for rounding, e.g.
+0.825.
+- `make-original-tables.R` is a by-hand coding of the values reported for Tables 1, 2, and 4 in B&S 2020.
 
@@
 These are handled via special treatment in `model-runner.R`, outside the main parallel loop.
-
-## Table 1
-
-### Original, smoothed ROC curves
-
-```{r}
-table1_smooth <- read_csv("output/table1-smooth.csv")
-
-table1_smooth %>%
-  knitr::kable(digits = 2,
-               caption = "Alternative Table 1 with non-smoothed AUC-ROC")
-```
-
-### With conventional ROC curves
-
-```{r}
-table1_nosmooth <- read_csv("output/table1-nosmooth.csv")
-table1_nosmooth %>%
-  knitr::kable(digits = 2,
-               caption = "Alternative Table 1 with non-smoothed AUC-ROC")
-```

### 2020-06-05 rep_nosmooth/variance.Rmd
add note checking RNG seed variance
How much does AUC-ROC vary between model runs when you don't see RNG seeds? Generally much less than 0.01. This is with the base spec escalation model, anecdotally, some of the other models, especially in Table 2, can vary more.

---
title: "How much variance in AUC-ROC is there between model runs?"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Since random forests are non-deterministic models, if you don't set a RNG seed the results for identical input will vary each time due to slightly different initial conditions. This means that the predictions and thus AUC-ROC values will also differ to some extent between model runs. Setting the RNG seed allows exact reproduction, but it doesn't solve this variability due to RNG state. We don't want results to be sensitive to a particular RNG seed. So, how much does AUC-ROC vary between model runs?

This script will run the base specification escalation model several times, in both the 1-month and 6-month versions, and then check the distribution of test AUC-ROC values obtained.

The chunk below will run the models. It's cached by default because this can take more than a few seconds to run, depending on the number of samples.

- `WORKERS`: how many parallel workers to use; number of cores or one less usually is good
- `SAMPLES`: how many times to run the model for each of the two data versions

```{r packages}
# Don't cache this, otherwise the chunks behind run-models will not work
suppressMessages({
  library(dplyr)
  library(readr)
  library(randomForest)
  library(pROC)
  library(foreach)
  library(doFuture)
  library(tidyr)
  library(ggplot2)
})
```

```{r run-models, cache=TRUE}
WORKERS  <- 8
SAMPLES  <- 40

# Function to wrap the AUC-ROC calculation
auc_roc_vec <- function(pred, truth, smooth) {
  roc_obj <- pROC::roc(truth, pred, auc = TRUE, quiet = TRUE, smooth = smooth)
  as.numeric(roc_obj$auc)
}

features <- c("gov_opp_low_level", "gov_reb_low_level", "opp_gov_low_level",
              "reb_gov_low_level", "gov_opp_nonviol_repression", "gov_reb_nonviol_repression",
              "gov_opp_accommodations", "gov_reb_accommodations", "reb_gov_demands",
              "opp_gov_demands")

train_end_year <- 2007

hp_settings <- list(ntree = 100000L, maxnodes = 5L, sampsize = 100L, replace = FALSE)

dv_name <- "incidence_civil_ns_plus1"

data_1mo <- read_rds("trafo-data/1mo_data.rds")
data_6mo <- read_rds("trafo-data/6mo_data.rds")

registerDoFuture()
plan("multisession", workers = WORKERS)

model_table <- crossing(
  horizon = rep(c("1 month", "6 months"), SAMPLES),
  sample  = rep(1:SAMPLES, each = 2)
)

# shuffle the table so the work load on average is more evenly distributed 
# among workers
model_table <- model_table[sample(1:nrow(model_table)), ]

res <- foreach(
  i = 1:nrow(model_table),
  .inorder = FALSE
) %dopar% {

  horizon_i <- model_table[i, ][["horizon"]]
  sample_i  <- model_table[i, ][["sample"]]

  if (horizon_i=="1 month") {
    data_i <- data_1mo
  } else {
    data_i <- data_6mo
  }

  train_data <- data_i %>% filter(year <= train_end_year)
  test_data  <- data_i %>% filter(year >  train_end_year)

  # randomForest with the x, y instead of formulate interface will not drop
  # missing values, so we need to make sure the training data drops those
  # manually
  train_data   <- train_data %>%
    dplyr::select(one_of(c(dv_name, features))) %>%
    filter(complete.cases(.))

  fitted_model <- suppressWarnings(randomForest(
    x = train_data[, features],
    y = train_data[, dv_name],
    ntree    = hp_settings$ntree,
    maxnodes = hp_settings$maxnodes,
    sampsize = hp_settings$sampsize,
    replace = FALSE, do.trace = FALSE, importance = FALSE
  ))

  # Because the original rep code uses the formula interface, the predictions
  # it puts out cover all of the test data, with NA values where needed for
  # incomplete rows. Replicate that here.
  non_missing_test_data <- test_data %>%
    dplyr::select(one_of(c("year", "period", "country_iso3", features))) %>%
    dplyr::filter(complete.cases(.))
  preds <- non_missing_test_data %>%
    dplyr::select(year, period, country_iso3) %>%
    mutate(pred = predict(fitted_model, newdata = non_missing_test_data,
                          type = "response")) %>%
    # add the non-complete rows back in
    right_join(test_data %>%
                 dplyr::select(year, period, country_iso3, one_of(dv_name)),
               by = c("year", "period", "country_iso3")) %>%
    # the original rep scripts do this; doesn't make a difference for ROC
    # calculations
    mutate(pred2 = ifelse(is.na(!!sym(dv_name)), NA_real_, pred)) %>%
    arrange(country_iso3, year, period)

  res_i <- model_table[i, ] %>%
    mutate(
      auc_roc          = auc_roc_vec(pred = preds$pred, truth = preds[, dv_name],
                                     smooth = FALSE),
      auc_roc_smoothed = auc_roc_vec(pred = preds$pred, truth = preds[, dv_name],
                                     smooth = TRUE),
      # The number of non-missing rows in the training data used to fit the mdl
      N_train = nrow(train_data),
      # The number of non-missing predictions
      N_test  = sum(complete.cases(preds))
    )
  res_i
}

res <- bind_rows(res)
write_rds(res, "output/variance.rds")
```

```{r}
res <- read_rds("output/variance.rds")
res_long <- res %>%
  dplyr::select(-N_train, -N_test) %>%
  tidyr::pivot_longer(auc_roc:auc_roc_smoothed, names_to = "smoothed", values_to = "auc_roc") %>%
  mutate(smoothed = ifelse(smoothed=="auc_roc", FALSE, TRUE))

res_tbl <- res_long %>%
  group_by(horizon, smoothed) %>%
  dplyr::summarize(N = n(),
                   mean = mean(auc_roc),
                   min  = min(auc_roc),
                   max  = max(auc_roc),
                   sd   = sqrt(mean((auc_roc-mean(auc_roc))^2)), # pop sd, not sample sd
                   .groups = "drop")
res_tbl %>%
  knitr::kable(digits = 3)
```

Comparing across the two dimensions--horizon and smoothed--actually smoothing has more impact on the results than the forecast horizon.

```{r}
# I want to add horizontal lines at the boundaries for rounding AUC-ROC to 
# two digits. Since the plot below is faceted, need to use a separate data frame
# for this with facet variable
hlines <- bind_rows(
  tibble(smoothed = "Smoothed", y = c(.82, .83, .84, .85) + .005),
  tibble(smoothed = "Original", y = c(.77, .78) + .005)
)

res_long %>%
  mutate(smoothed = ifelse(smoothed, "Smoothed", "Original"),
         horizon  = factor(horizon), 
         smoothed = factor(smoothed, levels = c("Smoothed", "Original"))) %>%
  ggplot() +
  facet_wrap(~smoothed, scales = "free") +
  geom_violin(aes(x = horizon, y = auc_roc)) +
  geom_point(aes(x = horizon, y = auc_roc), alpha = .5) +
  geom_hline(data = hlines, aes(yintercept = y), linetype = 3) +
  theme_bw()
```

The dotted lines in the graph mark the boundaries when rounding AUC-ROC to two digits. So for example, the original 1-month AUC-ROC straddles a boundary, meaning it is likely that the last digit in the tables reported in the paper will flip between model runs. On the other hand, the original 6-months version is between boundaries so the rounded results are likely to not change between runs.

But this straddling of rounding points aside, it seems that the variance of AUC-ROC values is generally pretty low: \(2 \times \textrm{SD} \lesssim 0.01 \). Not enough to affect results interpretation as long as we disregard differences of 0.01 or less. At least based on the escalation model, it seems some of the models in Table 2 might vary a bit more.

### 2020-06-05 paper.Rmd

--- 2020-06-04 paper.Rmd
@@
 Their theory-based escalation model does not do better than the alternatives that they examine, indeed it does worse.
-The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure: 1) they use smoothed performance curves instead of the unsmoothed curves, 2) they mistakenly encode the dependent variable to be incidents of civil war, rather than as they claim the onset of a civil war, 3) they mistakenly calculate the predicted values for the alternative ``non-theoretical'' models that they compare to.
-In what follows we show the impact of these mistakes on the conclusions, we examine their understanding of prediction it self, and we explore the logic of their assertion about theory as well.
+The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure: 1) they use smoothed performance curves instead of the unsmoothed curves, 2) they mistakenly encode the dependent variable to be incidents of civil war, rather than as they claim the onset of a civil war, 3) they mistakenly calculate the predicted values for the alternative "non-theoretical" models that they compare to.
+In what follows we show the impact of these mistakes on the conclusions, we examine their understanding of prediction itself, and we explore the logic of their assertion about theory as well.
 We believe that this research by B&S is incorrect and should be retracted.
+
+B&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkable accurately." Indeed they argue that this so-called theoretical model outperforms four other "more mechanical" alternatives.
+Second, they clam that the integration of structure with process are better over short forecasting windows.
+Third they preregisted the list of thirty countries which have the highest risk of civil war onset.
+They claim that such prospective predictions are rare in the literature when in fact they have been routine for many years with several prominent projects.
+B&S claim to be unique in assessing these forecasts.
+A qualitative analysis of their predictions allows them to conclude that their model is strong and a more precise test is undertaken.
+We will return to their analysis later, after correcting the procedural mistakes we found in their research process.
+
+Before proceeding, we quote B&S (page 24):
+\begin{quote}
+Our theoretically driven model generates accurate forecasts, with base specification
+AUCs of 0.82 and 0.85 over one- and six-month windows, respectively, and AUCs
+as high as 0.92 in other specifications. Our model also consistently and sometimes
+dramatically outperforms the alternatives we test... Cederman and Weidmann (2017, 476) argue that "the hope that
+big data will somehow yield valid forecasts through theory-free 'brute force' is
+misplaced in the area of political violence." Our results lend some credence to this
+claim.
+\end{quote}
 
@@
 
-@blair2020forecasting (B&S for the sake of brevity hereafter) aim to examine whether theory adds value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
+@blair2020forecasting (B&S for the sake of brevity hereafter) aim to examine whether theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
 They find that it does.
@@
 
-- (1) How does the theoretically-driven rscalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories?
+- (1) How does the theoretically-driven escalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories?
 - (2) Does annual, structural information from the PITF instability forecasting model add to the escalation model's monthly and 6-month predictions?
@@
 
-To assess the first two questions B&S use global datasets covering all major countries from 2001 to 2015.
+To assess the first two questions, B&S use global datasets covering all major countries from 2001 to 2015.
 Two versions of the dataset are used, one at the country-month level, the other aggregated to 6-month half-years.
-The main outcome is civil war onset, measured using Sambanis' civil war dataset.
+The main outcome variable is civil war onset, measured using Sambanis' civil war dataset.
 
-Both the first and second questions above rely on comparing the Escalation model to various alternative models. The same procedure is used in both cases:
+Both the first and second questions above rely on comparing their Escalation model to various alternative models. The same procedure is used in both cases:
 
@@
 - Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government one one side and opposition or rebel actors on the other.
-These are directed, thus making for 4 total covariates.
+These are directed, thus making for four total covariates.
 - CAMEO: counts for all CAMEO event codes, thus a total of 1,159 covariates (CAN THIS BE RIGHT?). 
@@
 
-The results in Table 1, aside from the core base specification results, include 8 additional robustness tests for both the 1-month and 6-month versions.
+The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions.
 These robustness checks vary either (1) random forecast hyperparameter values, or (2) the year used to split the train/test data, or (3) alternative codings of the civil war onset dependent variable.
@@
 
-Finally, B&S used the Escalation model to create forcasts for the first half of 2016, and in their third and final analysis, they score the forecasts accuracy using civil war onset data later observed.
+Finally, B&S used their Escalation model to create forecasts for the first half of 2016, and in their third and final analysis, they score the forecasts accuracy using civil war onset data later observed.
 This is summarized in B&S Table 3.
@@
 ROC curves typically appear step-like in response to the distribution of positive and negative cases in the data.
-In this case there are also groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines.
+In this case, there are also groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines as seen in the panels on the right.
 In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached.
 For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set.
-Thus the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.
+Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.
 We can only speculate, but maybe the very "step"-like nature of the ROC curves plus the presence of unusual diagonal lines accounts for the decision to use smoothed ROC curves.
 
-As it turns out, this decision has a dramatic impact on the AUC-ROC calculations.
-As we will show in the results further below, the difference between the smoothed and original ROC AUC values is up to 0.12--a huge difference given that row-wise, the models being compared typically differ by only 0.05 or less--and differentially impacts the models that are being compared.
+As it turns out, this decision has a dramatic impact on the AUC-ROC calculations they use to support their empirical claims.
+As we will show in the results further below, the difference between the smoothed and original ROC AUC values is up to 0.12--a huge difference given that row-wise, the models being compared typically differ by only 0.05 or less--and deferentially impacts the models that are being compared.
 In fact, B&S's original results and interpretation are entirely conditional on the use of smoothed AUC-ROC.
@@
 We infer that the intent is that the Escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
-The implementation actually weighs the test set predictions using the PITF model predictions for the training set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`).
+Their implementation, however, actually weighs the test set predictions using the PITF model predictions for the training set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`).
 The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions.
-Similarly in the 6 month version of this file.] This appears to be a coding error.
+Similarly in the 6-month version of this file.] This appears to be a coding error.
 
@@
 The intention clearly was to run two separate random forest models, one each on the low- and high-risk training data splits.
-The replication code does indeed run two separate random forecasts, but they are both run on the exact same training data, which consists of the full training data all other models are run on.
+The replication code does indeed run two separate random forecasts, but they are both run on the *exact same training data*, which consists of the full training data all other models are run on.
 The models are also identical otherwise, i.e.
@@
 Although the RF models are used for a binary decision problem, the actual implementation uses regression RFs for continuous outcomes.]) trees.
-Thus while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does indeed have 200,000 decision trees.
+Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does indeed have 200,000 decision trees.
 Furthermore, the PITF model predictions do not impact the way the combined RF model predicts at all, not even through a binary low-/high-risk split.
@@
 
-B&S show a confusiong matrix to score their 2016-H1 forecasts in Table 4.
+B&S show a confusion matrix to score their 2016-H1 forecasts in Table 4.
 Although the forecasts are for the probability of civil war onset, in the replication code they are actually scored using the much more common incidence of civil war, i.e.
@@
 The Quad model appears to be roughly on par with the Escalation model.
-Thus the original B&S conclusion that the Escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.
+Thus, the original B&S conclusion that the Escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.
 
@@
 
-This is no the case anymore. It outperforms in the 1-month version and underperforms the Escalation model in the 6-month version.
+This is not the case anymore. It outperforms in the 1-month version and under-performs the Escalation model in the 6-month version.
 
@@
 The Escalation model is the only model that consistently had a benefit from smoothing.
-For all 8 other models, smoothing sometimes gave a benefit, sometimes not.
+For all eight other models, smoothing sometimes gave a benefit, sometimes not.
 
@@
 The CAMEO and PITF Only models benefit slightly, on the order of 0.01 or less.
-The Escalation Only model on the other hand has an average boost of 0.058 to it's AUC-ROC from smoothing.
+The Escalation Only model on the other hand has an average boost of 0.058 to its AUC-ROC from smoothing.
 Considering the spread of AUC-ROC values if we compare across rows in B&S Tables 1 and 2, that boost is substantial.
@@
 The use of smoothed ROC curves to calculate AUC-ROC values only clearly benefits the Escalation model, it does so consistently, and by a considerable margin.
-All 8 alternative models reported in Tables 1 and 2 do not on average gain when using smoothed ROC curves to calculate AUC.
+All eight alternative models reported in Tables 1 and 2 do not on average gain when using smoothed ROC curves to calculate AUC.
 
@@
 As the outcome is a binary indicator of civil war onset, one would typically use a classification random forest that predicts 0 or 1 labels directly.
-The implementation of random forests that B&S use ([@liaw:wiener:2002]) is based on the original @breiman:2001 implementation and calculates predictived probabilities by averaging over the "0" or "1" votes from all constituent decision trees.
+The implementation of random forests that B&S use ([@liaw:wiener:2002]) is based on the original @breiman:2001 implementation and calculates predictive probabilities by averaging over the "0" or "1" votes from all constituent decision trees.
 The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number.
@@
 
-B&S has a clear misunderstanding of the role that out-of-sample prediction can plan in analysis.
+B&S have a clear misunderstanding of the role that out-of-sample prediction can plan in analysis.
 On the one hand it can be used for simple forecasting while on the other it can be used to evaluate the performance of models, specifically to overfitting and bias.
@@
 
-Further, they complete misunderstand the use of ICEWS event data in current research.
-They claim that most uses to date have focused on the quad categories, but this ignores a wide swath of literature (Steinert-Threlkeld ApsR 2017) and Metternich et al (AJPX 2013) that uses a sepcific action---such as protest---defined in the CAMEO ontology.
+Further, they completely misunderstand the use of ICEWS event data in current research.
+They claim that most uses to date have focused on the quad categories, but this ignores a wide swath of literature (Steinert-Threlkeld APSR 2017) and Metternich et al (AJPS 2013) that uses a specific action---such as protest---defined in the CAMEO ontology.
 In the AJPS article we hand coded, for example, every actor in Thailand and focused on an analysis of how those have been interacting.

### 2020-06-05 paper.Rmd
moar editing
I said moar

@@
 ---
-title: "Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
-author: Richard K. Morgan and Andreas Beger and Michael D. Ward
+title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
+author: Andreas Beger and Richard K. Morgan and Michael D. Ward
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
 In what follows we show the impact of these mistakes on the conclusions, we examine their understanding of prediction itself, and we explore the logic of their assertion about theory as well.
-We believe that this research by B&S is incorrect and should be retracted.
 
@@
 
-@blair2020forecasting (B&S for the sake of brevity hereafter) aim to examine whether theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
-They find that it does.
-They arrive at this conclusion by examining the problem of predicting civil war onset, and find that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory and a kitchen sink model with more than 1,000 covariates.
+@blair2020forecasting (B&S hereafter) aim to examine whether theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
+For the moment we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models.
+In short, they uphold their hypothesis that theory guided empirical research produces better conflict predictions than machine learning inspired efforts that are essentially ad hoc combinations of available variables.
+They arrive at this conclusion by examining the problem of predicting civil war onset.
+They report that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory including a \textit{kitchen sink} model with more than 1,000 covariates.
 
@@
 
-- (1) How does the theoretically-driven escalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories?
-- (2) Does annual, structural information from the PITF instability forecasting model add to the escalation model's monthly and 6-month predictions?
-- (3) How accurate were predictions for the first half of 2016 derived from the escalation model?
+1. How does the theoretically-driven escalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories?
+2. Does annual, structural information from the PITF instability forecasting model add to the escalation model's monthly and 6-month predictions?
+3. How accurate were predictions using the escalation model for the first half of 2016?
 
-To assess the first two questions, B&S use global datasets covering all major countries from 2001 to 2015.
+To assess the first two questions, B&S use ICEWS data covering all major countries from 2001 to 2015.
 Two versions of the dataset are used, one at the country-month level, the other aggregated to 6-month half-years.
@@
 
-Both the first and second questions above rely on comparing their Escalation model to various alternative models. The same procedure is used in both cases:
+Both the first and second questions above rely on comparing their escalation model to various alternative models. The same procedure is used in both cases:
 
 1. Split the training data into training (2001 - 2007) and test (2008 - 2015) sets.
-2. Estimate the Escalation and other competing models. 
+2. Estimate the escalation and other competing models. 
 3. Create out-of-sample (OOS) predictions from each model using the test set. 
 4. Calculate AUC-ROC measures for each set of OOS predictions.
-
-The corresponding results for each question are shown in B&S Tables 1 and 2, which we will replicate further below.
 
@@
 The models are:
-
 - Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
@@
 These are directed, thus making for four total covariates.
-- CAMEO: counts for all CAMEO event codes, thus a total of 1,159 covariates (CAN THIS BE RIGHT?). 
+- CAMEO: counts for all CAMEO event codes totalling $1,159$ covariates, which are mostly zero for any country in any month. 
 - Average: unweighted average of the predictions from the four models briefly described above.
 
+The corresponding results for each question are shown in B&S Tables 1 and 2, which we will examine further below. We accurately replicate their Tables 1 and 2, with very tiny differences. 
 The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions.
@@
 
-The second question, whether structural variables add to the Escalation model, is assessed by comparing the original escalation model to four alternatives that incorporate annual, structural variables that are used in the PITF instability forecasting model:
+The second question, whether structural variables add to the escalation model, is assessed by comparing the original escalation model to four alternatives that incorporate annual, structural variables that are used in the PITF instability forecasting model:
 
@@
 
-Finally, B&S used their Escalation model to create forecasts for the first half of 2016, and in their third and final analysis, they score the forecasts accuracy using civil war onset data later observed.
+Finally, B&S used their escalation model to create forecasts for the first half of 2016, and in their third and final analysis, they score the forecasts accuracy using civil war onset data later observed.
 This is summarized in B&S Table 3.
@@
 
-While attempting to replicate B&S's results, we found several issues.
+While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation.
+These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) an incorrect implementation of the weighted by ICEWS data model, c)  an inconsistent test set $ for the models examined, d) incorrect scoring of the 2016 forecasts, e) an inappropriate use of random forrest implementation, with very unusual hyperparameters.
+Our belief is that these research decisions lead B&S to draw conclusion that are incorrect.
+The escalation model is not the best and it actually performs worse that the atheoretical, garbage can model with over $1000$ variables.
+We turn to discussing these five issues below.
 
@@
 This implies that the ROC curves were only smoothed in the references Figure 1, but actually all AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation.
+The reason to use a smoothed curve is that you have a continuous dependent variable (not the case here) which can produce any value of the dependent variable.
+Wherein there is a discrete number of actual outcomes there is no justification for using smoothed ROC plots, nor for caculating statistics based on them.
+Moreover, it is easy to compute the nonsmoothed ROC curve, and associated statistics, so computational ease does not justify this choice.
 
@@
 Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.
-We can only speculate, but maybe the very "step"-like nature of the ROC curves plus the presence of unusual diagonal lines accounts for the decision to use smoothed ROC curves.
+Notice also that the smoothing averages the left most almost straight line with the right most almost straight line in a monotonic way.
+This ignores the fact that the actual ROC
+
+Table \ref{tab:table1-smooth} is our replication of B&S Table 1 with smoothed AUC-ROC.
+The results differ slightly from the original B&S Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models.
+It is the case that B&S set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel.
+In any case, the interpretation of results should not be sensitive to random variation, i.e.
+it should not depend on using a specific RNG seed.
+On the basis of these results, B&S conclude that the escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.
+
+```{r table1-smooth}
+table1_smooth <- read_csv("data/table1-smooth.csv")
+
+table1_smooth %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Replication of B\\&S Table 1 with smoothed ROC curves; test set AUC-ROC for various models",
+    label = "table1-smooth") %>% 
+  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
+```
+
+Table \ref{tab:table1-nosmooth} shows a version of B&S Table 1 with the conventional non-smoothed ROC curves.
+The Average model outperforms the escalation model in 17 out of 18 cases, and the CAMEO model outperforms in 16 of 18 cases, with one tie.
+The Goldstein model generally outperforms the Escalation model in the 6-month version.
+The Quad model appears to be roughly on par with the Escalation model.
+Thus, the original B&S conclusion that the escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.
+
+```{r table1-nosmooth}
+table1_nosmooth <- read_csv("data/table1-nosmooth.csv")
+
+table1_nosmooth %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Replication of B\\&S Table 1 \\textit{without} smoothed ROC curves; test set AUC-ROC for various models",
+    label = "table1-nosmooth") %>% 
+  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
+```
 
 As it turns out, this decision has a dramatic impact on the AUC-ROC calculations they use to support their empirical claims.
-As we will show in the results further below, the difference between the smoothed and original ROC AUC values is up to 0.12--a huge difference given that row-wise, the models being compared typically differ by only 0.05 or less--and deferentially impacts the models that are being compared.
+As we show below, the difference between the smoothed and original ROC AUC values is up to 0.12--a huge difference given that row-wise, the models being compared typically differ by only 0.05 or less--and this differentially impacts the models that are being compared.
 In fact, B&S's original results and interpretation are entirely conditional on the use of smoothed AUC-ROC.
+These results are shown in Tables 1 and 2.
+The Goldstein, CAMEO, and Average model are everywhere superior to the theoretically informed escalation model.
+Even the Quad model is generally as good as the escalation model in all implementations, and is frequently quite a bit better.
 
@@
 
-We infer that the intent is that the Escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
+We infer that the intent is that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
 Their implementation, however, actually weighs the test set predictions using the PITF model predictions for the training set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`).
@@
 Furthermore, the PITF model predictions do not impact the way the combined RF model predicts at all, not even through a binary low-/high-risk split.
-The split-population PITF RF model is practically speaking just another Escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree).
+The split-population PITF RF model is practically speaking just another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree).
 
@@
 
-## Basic interpretation of Table 1
-
-Table \ref{tab:table1-smooth} is our replication of B&S Table 1 with smoothed AUC-ROC.
-The results differ slightly from the original B&S Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models.
-It is the case that B&S set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel.
-In any case, the interpretation of results should not be sensitive to random variation, i.e.
-it should not depend on using a specific RNG seed.
-On the basis of these results, B&S conclude that the Escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.
-
-```{r table1-smooth}
-table1_smooth <- read_csv("data/table1-smooth.csv")
-
-table1_smooth %>%
-  select(-horizon) %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Replication of B\\&S Table 1 with smoothed ROC curves; test set AUC-ROC for various models",
-    label = "table1-smooth") %>% 
-  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
-```
-
-Table \ref{tab:table1-nosmooth} shows a version of B&S Table 1 with the conventional non-smoothed ROC curves.
-The Average model outperforms the Escalation model in 17 out of 18 cases, and the CAMEO model outperforms in 16 of 18 cases, with one tie.
-The Goldstein model generally outperforms the Escalation model in the 6-month version.
-The Quad model appears to be roughly on par with the Escalation model.
-Thus, the original B&S conclusion that the Escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.
-
-```{r table1-nosmooth}
-table1_nosmooth <- read_csv("data/table1-nosmooth.csv")
-
-table1_nosmooth %>%
-  select(-horizon) %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Replication of B\\&S Table 1 \\textit{without} smoothed ROC curves; test set AUC-ROC for various models",
-    label = "table1-nosmooth") %>% 
-  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
-```
-
 ## Do structural variables add to the Escalation model?
@@
     digits = 2, booktabs = TRUE, 
-    caption = "Replication of B\\&S Table 2: Test set AUC-ROC for Escalation model with and without structural PITF contribution",
+    caption = "Replication of B\\&S Table 2: Test set AUC-ROC for escalation model with and without structural PITF contribution",
     label = "table2-fixed") %>%
@@
 
-This is not the case anymore. It outperforms in the 1-month version and under-performs the Escalation model in the 6-month version.
+This is not the case anymore. It outperforms in the 1-month version and under-performs the escalation model in the 6-month version.
 
@@
 Positive values indicate that smoothing increased a model's apparent performance.
-The Escalation model is the only model that consistently had a benefit from smoothing.
+The escalation model is the only model that consistently had a benefit from smoothing.
 For all eight other models, smoothing sometimes gave a benefit, sometimes not.
@@
 The CAMEO and PITF Only models benefit slightly, on the order of 0.01 or less.
-The Escalation Only model on the other hand has an average boost of 0.058 to its AUC-ROC from smoothing.
+The escalation only model on the other hand has an average boost of 0.058 to its AUC-ROC from smoothing.
 Considering the spread of AUC-ROC values if we compare across rows in B&S Tables 1 and 2, that boost is substantial.
 
-The use of smoothed ROC curves to calculate AUC-ROC values only clearly benefits the Escalation model, it does so consistently, and by a considerable margin.
+The use of smoothed ROC curves to calculate AUC-ROC values only clearly benefits the escalation model, it does so consistently, and by a considerable margin.
 All eight alternative models reported in Tables 1 and 2 do not on average gain when using smoothed ROC curves to calculate AUC.
@@
 
-- the theory-driven Escalation model is outperformed both by the low-effort 1,160 predictor all CAMEO model and the Average ensemble model
-- structural variables substantially improve the Escalation model's performance when added to the pool of predictors the underlying random forest model draws on
+- the theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all CAMEO model and the Average ensemble model
+- structural variables substantially improve the escalation model's performance when added to the pool of predictors on which the underlying random forest model is based.
 

### 2020-06-05 paper.Rmd
okay, done for friday in Seattle.

@@
 
-As it turns out, this decision has a dramatic impact on the AUC-ROC calculations they use to support their empirical claims.
+As it turns out, this decision has a dramatic impact on the AUC-ROC calculations B&S use to support their empirical claims.
 As we show below, the difference between the smoothed and original ROC AUC values is up to 0.12--a huge difference given that row-wise, the models being compared typically differ by only 0.05 or less--and this differentially impacts the models that are being compared.
@@
 
-The next three issues we encountered all concern information in Table 2.
+The next three issues we encountered all concern information in B&S Table 2.
 

### 2020-06-06 paper.Rmd
one more before bed
boo

@@
 We infer that the intent is that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
-Their implementation, however, actually weighs the test set predictions using the PITF model predictions for the training set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`).
+B&S, however, actually weights the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`).
 The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions.
@@
 
+NOW SHOW HOW THIS MATTERS?
+
 ## Incorrect "PITF Split Population" implementation
 
-Similarly, the "PITF Split Population" model appears to be incorrectly implemented. B&S describe it as, on page 20:
+Similarly, the "PITF Split Population" model appears to be incorrectly implemented. B&S describe it on page 20:
 
@@
 The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself.
-If we ran both with the same RNG seed, they would be identical in every respect, producing identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined.
+If we ran both with the same random seed, they would be identical in every respect, producing identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined.
 We have documented details at https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5.]
@@
 The split-population PITF RF model is practically speaking just another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree).
+This does not adequately implement their research strategy.
+
+GIVE EXAMPLE OF IMPACT HERE?
 
@@
 
-Lastly, the AUC-ROC values reported in the original B&S Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}).
+Further, the AUC-ROC values reported in the original B&S Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}).
 ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available.
@@
 Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
-In other words, the results for different models in B&S Table 2 are actually not comparable to the other, and any conclusions drawn from such comparison are potentially incorrect.
+In other words, the results for different models in B&S Table 2 are actually not comparable toone another, and any conclusions drawn from such comparison are potentially incorrect.
 
@@
 The original B&S Table 2 1-month have *N*=11,806--12,495 versus a common joint subset of 9,811, and for the 6-month row *N*=2,070--2,233 versus *N*=1,915 for the common joint subset.
+
+Give EXAMPLE OF IMPACT HERE?
 

### 2020-06-06 rep_nosmooth/README.Rmd

@@
 In order to be able to run the models needed to create tables 1 and 2 more quickly than with the original replication code, the code here has been re-factored to enable running in parallel.
-Here is a summary of the major changes:
+The structure of the code has completely changed.
+Instead of a master script that handles file IO and sources other scripts to setup model definition objects (features, training/test data frames, etc.) and run models, the new logic is to:
+
+1.
+Create a central model table that encodes all cells in B&S Tables 1 and 2 and the information needed to define the underlying models, as well as associated model definitions like features, hyper-parameter settings, etc.
+(`output/model-definitions/`).
+2. Run through the model table (mostly in parallel) to estimate all models; save the resulting predictions. 
+3. Re-create output tables and figures.
+
+## (Clean) replication
+
+Below are instructions for how to run the script in order to replicate the updated analysis. To do a clean replication, delete the `output/` folder; this is not required though.
+
+Run through the files in the following order:
 
@@
 - `recreate-tables.R` recreates the original and non-smooth ROC versions of Tables 1 and 2 using `output/model-table-w-results.csv` and `output/all-predictions.R`.
+- `recreate-figure1.R` recreates the ROC curves in B&S Figure 1 using `output/model-table-w-results.csv` and `output/all-predictions.R`.
 
@@
 
-- [variance.md](/variance.md) examines how much the AUC-ROC results vary when you don't set an RNG seed.
+- [variance.md](variance.md) examines how much the AUC-ROC results vary when you don't set an RNG seed.
 For the base specification escalation model, the variance is significantly below 0.01, and small changes in the Table 1 and 2 replications mainly occur when the distribution straddles a threshold used for rounding, e.g.
@@
 
-### Model definitions
+## Model definitions
 

### 2020-06-06 paper.Rmd
sfs
sdfasdf

@@
 # Results of the updated replication
+We now turn to an examination of our replication that fixes the above issues.  The bottom line is that the changes we make turn on their heads the conclusions of B&S.
 
@@
 
-Table \ref{tab:table2-fixed} shows our replication of B&S Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common joint subset of tests cases that all models have non-missing predictions for.
+Table \ref{tab:table2-fixed} shows our replication of B&S Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common, joint subset of tests cases that all models have non-missing predictions for.
 Table \ref{tab:table2-full} further below shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.
@@
 
-B&S interpret the results as follows, on page 20:^[We list the "Overall, ..." interpretation out of order, last, for clarity.]
+B&S interpret their results as follows, on page 20 and we comment on the conclusions seratim:^[We list the "Overall, ..." interpretation out of order, last, for clarity.]
 
@@
 
-It performs roughly on par with the Escalation Only model.
-
+The weighted model performs roughly on par with the Escalation Only model.
 One finding that remains is that the "PITF Only" model is outperformed by the "Escalation Only" model.
-As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is not surprising.
+As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is neither surprising, nor noteworthy.
 
@@
 The most straightforward method of incorporating the annual, structural PITF variables--adding them to the predictors of the Escalation RF model--strictly outperforms the Escalation Only model.
-Note that the two other combination models considered are both non-standard, and that the "PITF Split Population" model does not in fact actually incorporate structural information at all.
-We thus conclude that adding structural variables actually clearly improves predictive performance.
+Note that the two other combination models considered are both non-standard and that the "PITF Split Population" model does not in fact actually incorporate structural information at all.
+We thus conclude that adding structural variables improves predictive performance.
 
@@
 
-The use of smoothed ROC curves to calculate AUC-ROC values only clearly benefits the escalation model, it does so consistently, and by a considerable margin.
-All eight alternative models reported in Tables 1 and 2 do not on average gain when using smoothed ROC curves to calculate AUC.
+The use of smoothed ROC curves to calculate AUC-ROC values clearly benefits \textbf{only} the escalation model.
+It does so consistently and by a considerable margin.
+All eight alternative models reported in Tables 1 and 2 on average do not gain when using smoothed ROC curves to calculate AUC.
 
 ## How accurate were the 2016 forecasts?
+
+THIS SECTION IS EMPTY
 

### 2020-06-06 paper.Rmd
add text about B&S table 4 to paper

@@
 
-THIS SECTION IS EMPTY
+B&S report confusion matrices for their forecasts for the first half of 2016 (2016-H1) in their Table 4.
+To create the confusion matrices, B&S treat the 30 highest ranked predictions as positive predictions ("1s") and the rest as negative predictions ("0s").
+We replicate their Table 4 in the top portion of \ref{tab:table4}.
+There are two confusion matrices for slightly different codings of outcomes in 2016-H1, under the corresponding "Assuming Persistence" and "Assuming Change" headings.
+We should not that the "Assuming Persistence" corresponds to the values in the replication data; the small variations for the "Assuming Change" version are hand-coded in the replication code file that generates the confusion matrices.
+
+We can see that the original table presents 15 or 16 positive cases for 2016-H1, depending on teh DV coding variation.
+This corresponds to a positive rate of around 9.5% for the first half of 2016 data.
+In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a much lower positive rate of around 0.5%.
+Since the positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data, we suspected that the forecasts were erroneously assessed using civil war incidence, not onset.
+As we showed above, by examining the replication code we were able to verify that the forecasts were scored using civil war incidence, i.e.
+including ongoing civil wars, rather than civil war onset years only.
+
+```{r, message=FALSE}
+tab4o <- read_csv("data/table4-original.csv")
+tab4f <- read_csv("data/table4-fixed.csv")
+
+tab4o <- tab4o %>%
+  rename(header = type) %>%
+  mutate(source = "Original, scored with civil war incidence")
+tab4f <- tab4f %>%
+  rename(Predicted0 = `0`, Predicted1 = `1`) %>%
+  mutate(source = "Fixed, scored with civil war onset")
+
+tab4 <- bind_rows(tab4o, tab4f) %>%
+  select(source, header, everything())
+
+tab4 %>%
+  knitr::kable(
+    caption = "Replication of B\\&S Table 4: 2016 Confusion Matrices for Six-month Escalation Model.",
+    label = "table4",
+    booktabs = TRUE
+  ) %>%
+  kableExtra::collapse_rows(1:2, row_group_label_position = "stack")
+```
+
+The correct confusion matrices when using observed onset (or the lack of it) are shown in the second part of Table \ref{tab:table4}.
+In the default "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1.
+Thus the recall values is undefined, while the precision is 0/30 = 0, compared to recall and precision values of 13/15 = 0.87 and 13/30 = 0.43 with the incorrect scoring.
+The alternative coding ("Assuming Change") produces 2 civil war onsets.
+Recall is 1 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.
+
+Another, minor issue or rather coding error, is related to using a lead version of the DV.
+With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 actually indicates the risk of civil war onset in 2016-H2.
+In the Table 4 script referenced above, the 2016-H1 predictions (for 2016-H2) are assessed using the raw DV, "incidence_civil_ns", not the lead version.
+Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1.
+In this case it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.
 
@@
 
-```{r, message=FALSE}
-tab4o <- read_csv("data/table4-original.csv")
-tab4f <- read_csv("data/table4-fixed.csv")
-
-tab4o <- tab4o %>%
-  rename(header = type) %>%
-  mutate(source = "Original, scored with civil war incidence")
-tab4f <- tab4f %>%
-  rename(Predicted0 = `0`, Predicted1 = `1`) %>%
-  mutate(source = "Fixed, scored with civil war onset")
-
-tab4 <- bind_rows(tab4o, tab4f) %>%
-  select(source, header, everything())
-
-tab4 %>%
-  knitr::kable(
-    caption = "Replication of B\\&S Table 4: 2016 Confusion Matrices for Six-month Escalation Model.",
-    booktabs = TRUE
-  ) %>%
-  kableExtra::collapse_rows(1:2, row_group_label_position = "stack")
-```
-
 ```{r}

### 2020-06-06 paper.Rmd
getting closer

@@
 We turn to discussing these five issues below.
+We hold a complete analysis that corrects all these issues until later, as there are many possible permuations of a seriatim unfolding.
 
@@
 
-NOW SHOW HOW THIS MATTERS?
-
 ## Incorrect "PITF Split Population" implementation
@@
 
-GIVE EXAMPLE OF IMPACT HERE?
-
 ## Inconsistent test set N for the models in Table 2
@@
 The original B&S Table 2 1-month have *N*=11,806--12,495 versus a common joint subset of 9,811, and for the 6-month row *N*=2,070--2,233 versus *N*=1,915 for the common joint subset.
-
-Give EXAMPLE OF IMPACT HERE?
 
@@
 There are two confusion matrices for slightly different codings of outcomes in 2016-H1, under the corresponding "Assuming Persistence" and "Assuming Change" headings.
-We should not that the "Assuming Persistence" corresponds to the values in the replication data; the small variations for the "Assuming Change" version are hand-coded in the replication code file that generates the confusion matrices.
+We should note that the "Assuming Persistence" corresponds to the values in the replication data; the small variations for the "Assuming Change" version are hand-coded in the replication code file that generates the confusion matrices and appear to be subjective assessments of B&S.
 
-We can see that the original table presents 15 or 16 positive cases for 2016-H1, depending on teh DV coding variation.
+We can see that the original table presents 15 or 16 positive cases for 2016-H1, depending on the dependent variable coding variation.
 This corresponds to a positive rate of around 9.5% for the first half of 2016 data.
 In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a much lower positive rate of around 0.5%.
-Since the positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data, we suspected that the forecasts were erroneously assessed using civil war incidence, not onset.
+Since the positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data, this suggests that the forecasts were erroneously assessed using civil war incidence, not onset.
 As we showed above, by examining the replication code we were able to verify that the forecasts were scored using civil war incidence, i.e.
@@
 In the default "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1.
-Thus the recall values is undefined, while the precision is 0/30 = 0, compared to recall and precision values of 13/15 = 0.87 and 13/30 = 0.43 with the incorrect scoring.
+Thus, the recall values is undefined, while the precision is 0/30 = 0, compared to reported recall and precision values of 13/15 = 0.87 and 13/30 = 0.43.
 The alternative coding ("Assuming Change") produces 2 civil war onsets.
@@
 
-## AUC-ROC sensitivity
-
-RNG and variation
-
-## Pre-registration
-
-- The escalation model specification, prior to looking at data!
-- The hyper-parameter tuning strategy.
-
 # Conclusion
 
-B&S have a clear misunderstanding of the role that out-of-sample prediction can plan in analysis.
-On the one hand it can be used for simple forecasting while on the other it can be used to evaluate the performance of models, specifically to overfitting and bias.
-They further misrepresent cross-validation which is not about fishing for results, as claimed by B&S.
+B&S have a clear misunderstanding of the role that out-of-sample prediction can play in analysis.
+On the one hand it can be used for simple forecasting, while on the other it can be used to evaluate the performance of models, specifically to overfitting and bias.
+They further misrepresent cross-validation---which is not about fishing for results, as claimed by B&S.
 The current state-of-the-art uses prediction and cross-validation most frequently to provide supportive evidence that is independent of the estimation procedure and the in-sample data.
@@
 The B&S approach that is advocated is to use theory to guide prediction.
-But theory is a quite ambiguous and undefined concept.
+But theory is an ambiguous and undefined concept.
 It is not a procedure.
@@
 There is a wide-ranging literature on this dynamic that they do not rely on to construct their model.
-As such their baseline comparison is hardly a standard bearer for strong theory.
+As such their baseline comparison is a poor standard bearer for strong theory.
 
@@
 We encountered several issues in the code underlying the B&S analysis.
-
 The issues we encountered are not subjective modeling choices.
-
-When we fix these issue and perform an updated analysis, the conclusions B&S draw all essentially overturn. In other words, B&S findings are based on a faulty analysis, and invalid.
+When we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturn.
+In other words, B&S findings are based on a faulty analysis, and invalid.
 
 Using the same analysis B&S intend to use, we in fact find that:
-
 - the theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all CAMEO model and the Average ensemble model
 - structural variables substantially improve the escalation model's performance when added to the pool of predictors on which the underlying random forest model is based.
+
+Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain.
+As social science becomes more adept at predictive analysis this will doubtless be of increasing importance.
+However, these comparisons must be done carefully to ensure that correct inferences are drawn.
+We continue to think that theory is overrated (@ward:2016), and that machine learning and big data will allow us to learn new things, but it will be interesting to see how the evidence for these claims are adjudicated with additional usage and careful evaluation.
 

### 2020-06-06 paper.Rmd
fix fallback bib path

@@
 bibliography: 
-  - "`r user <- Sys.info()[['user']]; if (user=='andybega')  '/Users/andybega/Work/whistle/master.bib' else if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else references.bib`"
+  - "`r user <- Sys.info()[['user']]; if (user=='andybega')  '/Users/andybega/Work/whistle/master.bib' else if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else 'references.bib'`"
 ---

### 2020-06-06 rep_nosmooth/make-original-tables.R

@@ -2,6 +2,9 @@
 #   Encode the original values reported for B&S Tables 1, 2, and 4
 #
 
+library(tibble)
+library(readr)
+
 setwd(here::here("rep_nosmooth"))
 
 dir.create("output/tables", recursive = TRUE, showWarnings = FALSE)

### 2020-06-06 rep_nosmooth/variance.Rmd
add quad model to variance check

@@
   library(ggplot2)
+  library(jsonlite)
 })
@@
 
-features <- c("gov_opp_low_level", "gov_reb_low_level", "opp_gov_low_level",
-              "reb_gov_low_level", "gov_opp_nonviol_repression", "gov_reb_nonviol_repression",
-              "gov_opp_accommodations", "gov_reb_accommodations", "reb_gov_demands",
-              "opp_gov_demands")
+features <- fromJSON("output/model-definitions/feature-specs.json")
 
@@
 model_table <- crossing(
-  horizon = rep(c("1 month", "6 months"), SAMPLES),
-  sample  = rep(1:SAMPLES, each = 2)
+  model   = c("Escalation", "Quad"),
+  horizon = c("1 month", "6 months"),
+  sample  = c(1:SAMPLES)
 )
@@
   sample_i   <- model_table[i, ][["sample"]]
+  features_i <- features[[model_table[i, ][["model"]]]]
 
@@
   train_data   <- train_data %>%
-    dplyr::select(one_of(c(dv_name, features))) %>%
+    dplyr::select(one_of(c(dv_name, features_i))) %>%
     filter(complete.cases(.))
@@
   fitted_model <- suppressWarnings(randomForest(
-    x = train_data[, features],
+    x = train_data[, features_i],
     y = train_data[, dv_name],
@@
   non_missing_test_data <- test_data %>%
-    dplyr::select(one_of(c("year", "period", "country_iso3", features))) %>%
+    dplyr::select(one_of(c("year", "period", "country_iso3", features_i))) %>%
     dplyr::filter(complete.cases(.))
@@
 res_tbl <- res_long %>%
-  group_by(horizon, smoothed) %>%
+  group_by(smoothed, horizon, model) %>%
   dplyr::summarize(N = n(),
@@
 hlines <- bind_rows(
-  tibble(smoothed = "Smoothed", y = c(.82, .83, .84, .85) + .005),
-  tibble(smoothed = "Original", y = c(.77, .78) + .005)
+  tibble(smoothed = "Smoothed", horizon = "1 month", 
+         y = c(.79, .80, .81, .82, .83, .84, .85) + .005),
+  tibble(smoothed = "Original", horizon = "1 month", 
+         y = c(.77, .78, .79, .8) + .005),
+  tibble(smoothed = "Smoothed", horizon = "6 months", 
+         y = c(.77, .78, .79, .80, .81, .82) + .005),
+  tibble(smoothed = "Original", horizon = "6 months", 
+         y = c(.76, .77, .78, .79) + .005)
 )
@@
   ggplot() +
-  facet_wrap(~smoothed, scales = "free") +
-  geom_violin(aes(x = horizon, y = auc_roc)) +
-  geom_point(aes(x = horizon, y = auc_roc), alpha = .5) +
+  facet_wrap(horizon ~ smoothed, scales = "free", nrow = 2) +
+  geom_violin(aes(x = model, y = auc_roc)) +
+  geom_point(aes(x = model, y = auc_roc), alpha = .5) +
   geom_hline(data = hlines, aes(yintercept = y), linetype = 3) +
-  theme_bw()
+  theme_bw() +
+  coord_flip()
 ```

### 2020-06-06 paper.Rmd

@@
 title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
-author: Andreas Beger and Richard K. Morgan and Michael D. Ward
+author: Andreas Beger, Richard K. Morgan, and Michael D. Ward
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
 knitr::opts_chunk$set(echo = FALSE, message=FALSE)
+# options(tinytex.verbose = TRUE)
 
@@
 These are directed, thus making for four total covariates.
-- CAMEO: counts for all CAMEO event codes totalling $1,159$ covariates, which are mostly zero for any country in any month. 
+- CAMEO: counts for all CAMEO event codes totaling $1,159$ covariates, which are mostly zero for any country in any month. 
 - Average: unweighted average of the predictions from the four models briefly described above.
@@
 While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation.
-These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) an incorrect implementation of the weighted by ICEWS data model, c)  an inconsistent test set $ for the models examined, d) incorrect scoring of the 2016 forecasts, e) an inappropriate use of random forrest implementation, with very unusual hyperparameters.
+These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) an incorrect implementation of the weighted by ICEWS data model, c)  an inconsistent test set for the models examined, d) incorrect scoring of the 2016 forecasts, e) an inappropriate use of random forrest implementation, with very unusual hyperparameters.
 Our belief is that these research decisions lead B&S to draw conclusion that are incorrect.

### 2020-06-07 paper.Rmd
I made some more cleanup
Please do anything you want with this over the weekend. I think we should try to get some feedback on it early next week, guys.

@@
 title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
-author: Andreas Beger, Richard K. Morgan, and Michael D. Ward
+author: 
+ - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com. corresponding author]
+ - Richard K. Morgan^[World Bank, rick.morgan2@gmail.com]
+ - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com]
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
 
-@bdm:2011 , @ward:2016 , @montgomery:etal:PS:2012, @chiba:etal:2015, @brandt:2015  @mhw:2012,@ward:etal:2013,@chadefaux:2015,@weidmann:ward:2010,@pilster:bohmelt:2014,@Brandt_etal_2011,@gleditsch:2015,@bennett:stam:2009,@shellman:etal:2013,@hegre:etal:2011,@ward:etal:2010,@freeman:job:1979,@gneiting:raftery:2005
-@choucri:1974,@choucri:robinson:1979,@goldstone:etal:2010,tetlock:etal:2017,
+There is a fairly long tradition of making predictions in academic and policy work on conflict.
+Most of these are not mentioned by B&S.
+We would include work from the last century including @choucri:1974, @andriole:young:1977, @choucri:robinson:1979, @freeman:job:1979,@hopple:andriole:1984, @schrodt:1991, and others.
+The current century has even more work that can't be overlooked or combined into a theory barren classification
+@obrien:2002,
+@buhaug:gleditsch:2008, 
+@bennett:stam:2009,
 @rost:etal:2009, 
-@hopple:andriole:1984,
-@andriole:young:1977,
+@obrien:2010, 
+@weidmann:ward:2010, 
+@weidmann:ward:2010, 
+@ward:etal:2010,
+@goldstone:etal:2010, 
+@bdm:2011, 
 @brandt:etal:2011, 
+@hegre:etal:2011, 
+@Brandt_etal_2011, 
+@oloughlin:etal:2012, 
+@hegre:etal:2013, @shellman:etal:2013, 
+@ward:etal:2013, 
 @brandt:etal:2014, 
+@buhaug:etal:2014, 
+@ramakrishnan:etal:2014, 
+@pilster:bohmelt:2014, 
+@brandt:2015, @chadefaux:2015, 
+@chiba:etal:2015, @gleditsch:2015, 
 @mellers:etal:2015, 
 @tetlock:gardner:2015, 
+@hegre:etal:2016, 
+@schleussner:etal2016, @ward:2016, 
+@beger:ward:2017, 
+@chiba:gleditsch:2017,
+@colaresi:mahmood:2017, 
+@hegre:etal:2017, 
+@tetlock:etal:2017, 
+@witmer:etal:2017, 
+@kupilik:witmer:2018, 
 @hegre:etal:2019, 
-@weidmann:ward:2010, 
-@beger:ward:2017,
-@schrodt:1991,
-@obrien:2010,
-@obrien:2002,
-@hegre:etal:2019,
-@hegre:etal:2017,
-@hegre:etal:2016,
-@chiba:gleditsch:2017,
-@buhaug:etal:2014,
-@colaresi:mahmood:2017,
-@ramakrishnan:etal:2014,
-@schleussner:etal2016,
-@kupilik:witmer:2018,
-@witmer:etal:2017
-@oloughlin:etal:2012
-@hegre:etal:2013 
-@buhaug:gleditsch:2008
-
-**AB: add a table comparing B&S original claims and updated results.**
+@hegre:etal:2019.
+Indeed this scholarship has brought us to the point where predictive analysis of conflict studies is not unknown in the scholarly and policy arenas.
+Most, if not all of these authors, would probably claim some theoretical motivation.
+The a-theoretical case is a counterfactual that either includes everything, or nothing, as aruged in @ward:2016.
+In the B&S article it serves as a strawman.
+Unfortunately for B&S that strawman outperforms there putatively theoretical model.
+Important statistical developments have also added to this research vector, including especially
+@gneiting:raftery:2005, 
+@montgomery:etal:PS:2012, and
+@mhw:2012.
 

### 2020-06-07 paper.Rmd
mike makes moar music
well some tiny changes at least.

@@
     keep_tex: true
-references:
-- id: blair2020forecasting
-  title: "Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning"
-  author:
-  - family: Blair
-    given: Robert A.
-  - family: Sambanis
-    given: Nicholas
-  container-title: Journal of Conflict Resolution
-  type: article-journal
-  issued:
-    year: 2020
 bibliography: 
@@
 
-Blair and Sambanis (2020; hereafter B&S) argue that theory is important for creating models that have high accuracy in forecasting civil war onset.
+Blair and Sambanis (hereafter B&S) argue that theory is important for creating models that have high accuracy in forecasting civil war onset.
 Indeed they argue that with such theory, forecasting is more accurate than has previously been possible.
@@
 Their theory-based escalation model does not do better than the alternatives that they examine, indeed it does worse.
-The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure: 1) they use smoothed performance curves instead of the unsmoothed curves, 2) they mistakenly encode the dependent variable to be incidents of civil war, rather than as they claim the onset of a civil war, 3) they mistakenly calculate the predicted values for the alternative "non-theoretical" models that they compare to.
+The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure: 1) they use smoothed performance curves instead of the unsmoothed curves, 2) they mistakenly encode the dependent variable to be incidents of civil war, rather than as they claim the onset of a civil war, 3) they mistakenly calculate the predicted values for the alternative "non-theoretical" models that they compare to, and they use a random forest model that is appropriate for continuous, not discrete dependent variables..
 In what follows we show the impact of these mistakes on the conclusions, we examine their understanding of prediction itself, and we explore the logic of their assertion about theory as well.
@@
 Second, they clam that the integration of structure with process are better over short forecasting windows.
-Third they preregisted the list of thirty countries which have the highest risk of civil war onset.
+Third they preregistered the list of thirty countries which have the highest risk of civil war onset.
 They claim that such prospective predictions are rare in the literature when in fact they have been routine for many years with several prominent projects.
@@
 
-@blair2020forecasting (B&S hereafter) aim to examine whether theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
+@blair:sambanis:2020 (B&S hereafter) aim to examine whether theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
 For the moment we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models.
@@
 
-> Of the approaches we test, the split-population analog is most promising
+1. B&S: ``Of the approaches we test, the split-population analog is most promising \ldots ''
 
@@
 
-> Adding PITF predictors improves the performance of the escalation model over six-month windows but diminishes it over one-month windows.
+2. B&S: ``Adding PITF predictors improves the performance of the escalation model over six-month windows but diminishes it over one-month windows \ldots ''
 
@@
 
-> The weighted model performs very poorly regardless.
+3. B&S: ``The weighted model performs very poorly regardless \ldots ''
 
@@
 
-> Overall, our results suggest that while measures of structural risk may improve predictive performance, the value they add is marginal and inconsistent.
+4.
+B&S: ``Overall, our results suggest that while measures of structural risk may improve predictive performance, the value they add is marginal and inconsistent.
 [...] Incorporating PITF thus significantly reduces or only slightly improves the performance of the escalation model, regardless of the approach we take.
+\ldots ''
 
@@
 @pilster:bohmelt:2014, 
-@brandt:2015, @chadefaux:2015, 
-@chiba:etal:2015, @gleditsch:2015, 
+@chadefaux:2017, 
+@chiba:etal:2015,
 @mellers:etal:2015, 

### 2020-06-08 table4/README.Rmd
fix up table 4 note

@@
 
+Is B&S Table 4 genered by scoring the civil war onset forecasts using civil war incidence?
+
+The data used to train the forecasting models is `6mo_data_OOS.dta`. This just appears to be an extension of the regular 6 month training data that includes and additional
+
 ```{r}
@@
 
+data_6month_oos %>% filter(year > 2013) %>% count(year, period)
+```
+
+The number of positive events--civil war onsets--matches what's in the training data, namely 20 onsets. There are no onsets recorded for 2016.
+
+```{r}
 data_6month_oos %>% count(incidence_civil_ns_plus1)
 data_6month_oos %>% count(incidence_civil_ns ) 
+data_6month_oos %>% filter(year==2016) %>% count(incidence_civil_ns)
+```
 
-pred_escalation_6mo_inc_civil_ns <- read.dta13("data/6mo_predictions_escalation_OOS.dta")
+The predictions seem to come from `6mo_predictions_escalation_OOS.dta`.
 
-# this object contains OOS forecasts from 2008 to 2016
-as_tibble(pred_escalation_6mo_inc_civil_ns)
+```{r}
+preds <- read.dta13("data/6mo_predictions_escalation_OOS.dta") %>%
+  as_tibble() %>%
+  # for some reason the data vectors as character, not numeric. 
+  mutate(across(-country, as.numeric))
 
-fcast2016 <- pred_escalation_6mo_inc_civil_ns %>%
+preds
+```
+
+This actually has OOS predictions from 2008 to 2016 (this also means the latest forecasts for 2016 use stale 2007 data...):
+
+```{r}
+range(preds$year)
+```
+
+There are no onsets recorded in this either. But there are some missing values in the non-"plus1" version of the outcome.
+
+```{r}
+fcast2016 <- preds %>%
   filter(year==2016)
@@
 fcast2016 %>%
-  filter(is.na(incidence_civil_ns)) %>%
-  dplyr::select(year, country)
+  dplyr::filter(is.na(incidence_civil_ns)) %>%
+  dplyr::select(country, year)
+```
 
-# Ok, indeed it looks like "incidence_civil_ns" was originally incidence and
-# then recoded to onset, with NA for continuing conflicts
+If you start out with an incidence civil war variable, like the name here implies, and you convert it to onset, it might be that this was done by keeping the first year as a "1" and setting the rest to missing.
+Does it look like this was done here?
+
+```{r}
 data_6month_oos %>%
-  ggplot(aes(x = year, y = factor(country_iso3), fill = factor(incidence_civil_ns))) +
-  geom_tile()
+  select(year, period, country_iso3, incidence_civil_ns, incidence_civil_ns_plus1) %>%
+  group_by(country_iso3) %>%
+  mutate(any_non0 = any(is.na(incidence_civil_ns)) | any(incidence_civil_ns==1)) %>%
+  ungroup() %>%
+  filter(any_non0==TRUE) %>%
+  mutate(country_iso3 = factor(country_iso3, levels = rev(unique(country_iso3))),
+         date = sprintf("%s-%02s-01", year, ifelse(period %% 2 == 1, 1, 7)),
+         date = as.Date(date)) %>%
+  ggplot(aes(x = date, y = factor(country_iso3), 
+             fill = factor(incidence_civil_ns))) +
+  geom_tile() +
+  scale_x_date(expand = c(0, 0)) +
+  theme_bw()
+```
 
-# How does this look in the full data?
-data_6month <- read.dta13("data/6mo_data_OOS.dta")
-data_6month %>%
-  ggplot(aes(x = year, y = factor(country_iso3), fill = factor(incidence_civil_ns))) +
-  geom_tile()
+This plot shows country-years--with countries on the *y*-axis and years on the *x*-axis. The fill color corresponds to whether the non-lagged outcome variable was coded as 0, 1, or missing.
 
-data_6month %>%
-  ggplot(aes(x = year, y = factor(country_iso3), fill = factor(incidence_civil_ns_plus1))) +
-  geom_tile()
+It seems that all sequences of NA's that start after 2001 are preceded by a "1" coding. There are some isolated single half-year 1's, and Yemen has 2 "1"'s not separated by a non-civil war year.
 
-# What's going on with the positive cases here?
-dv <- data_6month %>%
-  filter(incidence_civil_ns==1) %>%
-  dplyr::select(country_name, country_iso3, year, period, incidence_civil_ns)
-dv_plus1 <- data_6month %>%
-  filter(incidence_civil_ns_plus1==1) %>%
-  dplyr::select(country_name, country_iso3, year, period, incidence_civil_ns_plus1)
-both_dvs <- full_join(dv_plus1, dv) %>%
-  arrange(country_name, year)
+Looking through the appendix:
 
-ggplot(data_6month, aes(x = period, y = factor(country_iso3),
-                        fill = factor(incidence_civil_ns_plus1))) +
-  geom_tile()
+- Cote d'Ivoire (CIV) indeed has a single-year civil war (p. 30)
+- Libya also looks like it has a short civil war folled by longer one (p. 29)
+- Yemen, ok, there are two long periods of civil war separated by a small gap; that probably explains the onset immediately after NA half year (p. 28)
 
-# Predictions
-preds <- read.dta13("data/6mo_predictions_escalation_OOS.dta") %>%
-  as_tibble() %>%
-  filter(period==max(period)) %>%
-  mutate(incidence_civil_ns = as.integer(incidence_civil_ns),
-         incidence_civil_ns_plus1 = as.integer(incidence_civil_ns_plus1))
+Ok, so this indeed seems to be onset coded from an original incidence variable, as the name suggests anyways.
 
+How do we get from that to B&S Table 4?
+
+Aha, the culprit is in `6mo_make_confusion_matrix.do`. The script loads `predictions/6mo_predictions_escalation_OOS.dta`, drops everything prior to 2016. Then, on lines 50-52:
+
+```stata
+* Recode predictand
+
+replace incidence_civil_ns=1 if incidence_civil_ns==.
+```
+
+So it seems that `incidence_civil_ns` went from some original incidence version to the onset + NA for ongoing coding in the data that is used for the R portion of the work; then prior to creating the confusion matrices for the forecast scoring, the onset + NA for ongoing coding is reversed, giving us back incidence coding.
+So indeed, the forecasts are scored using incidence.
+
+Here's the alternative coding for the bottom portion, btw:
+
+```stata
+		  * Generate alternate coding of predictand
+
+			gen incidence_civil_ns_alt=incidence_civil_ns
+				replace incidence_civil_ns_alt=0 if country_name=="Colombia"
+				replace incidence_civil_ns_alt=1 if country_name=="Turkey"
+				replace incidence_civil_ns_alt=1 if country_name=="Burundi"
+```
+
+Redo a correct scoring against civil war onset.
+
+```{r}
 # Hand-code the Table 3 predictions;
-# The predictions don't exactly match what is in Table 3, possibley because I
+# The predictions don't exactly match what is in Table 3, possibly because I
 # only ran the 6 month OOS portion of +master.R after setting the seed at the
@@
 codes <- countrycode::countrycode(pred1, "country.name", "iso3c")
-preds$label <- as.integer(preds$country %in% codes)
+fcast2016$label <- as.integer(fcast2016$country %in% codes)
 # should be 30
-sum(preds$label)
+sum(fcast2016$label)
 
 # What is the truth data in preds?
-preds %>% count(incidence_civil_ns)
-preds %>% count(incidence_civil_ns_plus1)
+fcast2016 %>% count(incidence_civil_ns)
+fcast2016 %>% count(incidence_civil_ns_plus1)
 
@@
 
-tab4_top <- preds %>%
+tab4_top <- fcast2016 %>%
   rename(Observed = incidence_civil_ns, Predicted = label) %>%
@@
 # replace incidence_civil_ns_alt=1 if country_name=="Burundi"
-preds %>%
+fcast2016 %>%
   filter(country %in% c("COL", "TUR", "BDI"))
@@
 # Turkey has no ongoing conflict, so this is new onset
-preds_v2 <- preds %>%
+fcast2016_v2 <- fcast2016 %>%
   mutate(incidence_civil_ns = case_when(
-    country=="TUR" ~ 1L,
-    country=="BDI" ~ 1L,
+    country=="TUR" ~ 1,
+    country=="BDI" ~ 1,
     TRUE ~ incidence_civil_ns
@@
 
-tab4_bottom <- preds_v2 %>%
+tab4_bottom <- fcast2016_v2 %>%
   rename(Observed = incidence_civil_ns, Predicted = label) %>%

### 2020-06-08 paper.Rmd
replace smooth benefit tables with figure; edits to paper

@@
  - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com. corresponding author]
- - Richard K. Morgan^[World Bank, rick.morgan2@gmail.com]
+ - Richard K. Morgan^[USAID, rick.morgan2@gmail.com. The views expressed are those of the author and do not reflect the official policy or position of USAID or the US Government.]
  - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com]
@@
 
-Blair and Sambanis (hereafter B&S) argue that theory is important for creating models that have high accuracy in forecasting civil war onset.
+Blair and Sambanis (2020; hereafter B&S) argue that theory is important for creating models that have high accuracy in forecasting civil war onset.
 Indeed they argue that with such theory, forecasting is more accurate than has previously been possible.
@@
 Their theory-based escalation model does not do better than the alternatives that they examine, indeed it does worse.
-The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure: 1) they use smoothed performance curves instead of the unsmoothed curves, 2) they mistakenly encode the dependent variable to be incidents of civil war, rather than as they claim the onset of a civil war, 3) they mistakenly calculate the predicted values for the alternative "non-theoretical" models that they compare to, and they use a random forest model that is appropriate for continuous, not discrete dependent variables..
+The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure, the principal of which is that the performance results they report are based on smoothed performance curves, not the original unsmoothed curves.
+In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented, and we also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset.
 In what follows we show the impact of these mistakes on the conclusions, we examine their understanding of prediction itself, and we explore the logic of their assertion about theory as well.
@@
 While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation.
-These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) an incorrect implementation of the weighted by ICEWS data model, c)  an inconsistent test set for the models examined, d) incorrect scoring of the 2016 forecasts, e) an inappropriate use of random forrest implementation, with very unusual hyperparameters.
-Our belief is that these research decisions lead B&S to draw conclusion that are incorrect.
+These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) incorrect implementations of the weighted by PITF and PITF analog split-population models, c) inconsistent test sets for the models examined, and d) incorrect scoring of the 2016 forecasts.
+We also note (e) additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox.
+We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.
+
+Our belief is that these research decisions and issues lead B&S to draw conclusion that are incorrect.
 The escalation model is not the best and it actually performs worse that the atheoretical, garbage can model with over $1000$ variables.
@@
 This implies that the ROC curves were only smoothed in the references Figure 1, but actually all AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation.
-The reason to use a smoothed curve is that you have a continuous dependent variable (not the case here) which can produce any value of the dependent variable.
-Wherein there is a discrete number of actual outcomes there is no justification for using smoothed ROC plots, nor for caculating statistics based on them.
-Moreover, it is easy to compute the nonsmoothed ROC curve, and associated statistics, so computational ease does not justify this choice.
-
+ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result appear step-like.
 Figure 1 shows our replication of both the smoothed ROC curves B&S report, and the actual ROC curves on the right.
+The standard method is to compute the area under the curve AUC statistic on the original, non-smooth ROC curves that are shown on the right.
 
@@
 
-ROC curves typically appear step-like in response to the distribution of positive and negative cases in the data.
-In this case, there are also groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines as seen in the panels on the right.
+ROC curves typically appear step-like in response to the distribution of positive and negative cases in the data, which specifically in rare events problems like this one restricts the number of distinct true positive rate values.
+The specific predictions here also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines as seen in the panels on the right.
 In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached.
@@
 Notice also that the smoothing averages the left most almost straight line with the right most almost straight line in a monotonic way.
-This ignores the fact that the actual ROC
+This ignores the fact that the actual ROC **MIKE**
 
-Table \ref{tab:table1-smooth} is our replication of B&S Table 1 with smoothed AUC-ROC.
-The results differ slightly from the original B&S Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models.
-It is the case that B&S set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel.
-In any case, the interpretation of results should not be sensitive to random variation, i.e.
-it should not depend on using a specific RNG seed.
-On the basis of these results, B&S conclude that the escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.
-
-```{r table1-smooth}
-table1_smooth <- read_csv("data/table1-smooth.csv")
-
-table1_smooth %>%
-  select(-horizon) %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Replication of B\\&S Table 1 with smoothed ROC curves; test set AUC-ROC for various models",
-    label = "table1-smooth") %>% 
-  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
-```
-
-Table \ref{tab:table1-nosmooth} shows a version of B&S Table 1 with the conventional non-smoothed ROC curves.
-The Average model outperforms the escalation model in 17 out of 18 cases, and the CAMEO model outperforms in 16 of 18 cases, with one tie.
-The Goldstein model generally outperforms the Escalation model in the 6-month version.
-The Quad model appears to be roughly on par with the Escalation model.
-Thus, the original B&S conclusion that the escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.
-
-```{r table1-nosmooth}
-table1_nosmooth <- read_csv("data/table1-nosmooth.csv")
-
-table1_nosmooth %>%
-  select(-horizon) %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Replication of B\\&S Table 1 \\textit{without} smoothed ROC curves; test set AUC-ROC for various models",
-    label = "table1-nosmooth") %>% 
-  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
-```
-
-As it turns out, this decision has a dramatic impact on the AUC-ROC calculations B&S use to support their empirical claims.
-As we show below, the difference between the smoothed and original ROC AUC values is up to 0.12--a huge difference given that row-wise, the models being compared typically differ by only 0.05 or less--and this differentially impacts the models that are being compared.
-In fact, B&S's original results and interpretation are entirely conditional on the use of smoothed AUC-ROC.
-These results are shown in Tables 1 and 2.
-The Goldstein, CAMEO, and Average model are everywhere superior to the theoretically informed escalation model.
-Even the Quad model is generally as good as the escalation model in all implementations, and is frequently quite a bit better.
-
+As we show below, the use of smoothed ROC curves to calculuate AUC has a dramatic impact on the results.
 To our knowledge, the norm is to calculate AUC-ROC values using original, not smoothed ROC curves.
@@
 It might be that there are theoretical reasons justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the B&S results, it minimally would have warranted an explicit discussion in the paper.
-This is not the case, and it is only mentioned in the sentence we quote above.
+As it is, B&S offer no justification for using smoothed ROC curves in the AUC-ROC calculations.
 
@@
 Those sets differ across models (columns) for each row in Table 2.
+The top portion of Table \ref{tab:table2-N} shows the number of valid test set predictions that can be scored for each model in the original B&S Table 2.
+For the 1-month data version, the number of predictions differs by up to 500 cases, and in the 6-month data version by around 100 cases.
+These numbers appear small enough to be not important, but crucially the already small number of positive cases also changes: the "With PITF Predictors" and "PITF Only" models lose 2 and 1 of 11 positive cases, respectively, in the 1- and 6-month data versions.
+
 Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
-In other words, the results for different models in B&S Table 2 are actually not comparable toone another, and any conclusions drawn from such comparison are potentially incorrect.
+In other words, the results for different models in B&S Table 2 are actually not comparable to one another, and any conclusions drawn from such comparison are potentially incorrect.
+
+```{r}
+table2_full <- read_csv("data/table2-for-appendix.csv")
+
+table2_full %>%
+  select(horizon, column, N_test, cases) %>%
+  pivot_wider(names_from = column, values_from = N_test) %>%
+  arrange(horizon, desc(cases)) %>%
+  rename(Cases = cases, Horizon = horizon) %>%
+  select(Cases, Horizon, everything()) %>%
+  arrange(desc(Cases), Horizon) %>%
+  knitr::kable(
+    "latex", booktabs = TRUE, digits = 0,
+    caption = "Number of valid test predictions for the escalation and structural alterantives models (B\\&S Table 2)",
+    label = "table2-N") %>%
+  column_spec(3:7, width = "2cm") %>%
+  collapse_rows(1:2, row_group_label_position = "stack") 
+```
 
 We fix this issue by only using predictions for common joint subset of cases that all models have non-missing predictions for.
-The original B&S Table 2 1-month have *N*=11,806--12,495 versus a common joint subset of 9,811, and for the 6-month row *N*=2,070--2,233 versus *N*=1,915 for the common joint subset.
+The resulting number of cases for each model are shown in the bottom portion of the table.
+Since the sets of incomplete cases in the original version above do not completely overlap themselves, the common subset for all models is actually slightly smaller than the smallest N in the top portion of the table.
 
@@
 
-# Results of the updated replication
+## Random forest hyperparameters
+
+What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated.
+Table \ref{tab:hp} shows the default values used by the implementation of random forest that B&S use (from the **randomForest** R package), in contrast to the basic settings used by B&S for most the models reported in the paper.
+
+As the outcome is a binary indicator of civil war onset, one would typically use a classification random forest that predicts 0 or 1 labels directly.
+The implementation of random forests that B&S use ([@liaw:wiener:2002]) is based on the original @breiman:2001 implementation and calculates predictive probabilities by averaging over the "0" or "1" votes from all constituent decision trees.
+The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number.
+From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive, and operates on a large bootstrapped sample of the original training data.
+These are of course only heuristics and it is usual to attempt to find better hyper-parameter methods through some form of tuning.
+
+\begin{table}
+\caption{\label{tab:hp} Random forest (\texttt{randomForest()}) default versus B\&S hyperparameters}
+\begin{tabular}{l>{\raggedright\arraybackslash}p{2in}ll}
+\toprule
+Hyperparameter & Default heuristic & Default values (Escalation) & B\&S value \\
+\midrule
+type & & classification & regression \\
+ntree & & 500 & 100,000 or 1e6 \\
+mtry & \texttt{floor(sqrt(ncol(x)))} & 3 & 3 \\
+replace & & true & false \\
+sampsize & \texttt{nrow(x)} if replace, else \texttt{ceiling(.632*nrow(x))} & 11,869 & 100 or
+500 \\
+nodesize & 1 for classification & 1 & 1 \\
+maxnodes & & null & 5 or 10 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+B&S in contrast fit very large forests with 100,000 trees in the basic model form, but where each tree only operates on a very small sub-sample (N=100 or 500), drawn without replacement, of the available training data.
+This approach only works due to the choice to use regression, not classification, trees.
+Trying to use classification trees with the other parameter settings not work at all because it is almost guaranteed that a sample of 100 from the ~11,000 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample.
+As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:
+
+```
+Warning message:
+In randomForest.default(y = as.integer(train_df$incidence_civil_ns_plus1 ==  :
+  The response has five or fewer unique values.  Are you sure you want to do regression?
+```
+
+As it turns out, using regression random forests for this kind of binary classification problem in order to obtain probability estimates matches the probability random forest approach suggested and positively evaluated in @malley:etal:2012, and which is used in another prominent R implementation of random forests.^[The **ranger** package.] It is not clear whether this is intentional, as the Malley paper is not cited in B&S.
+
+In any case, B&S's random forest approach appears to work really well.
+We tried to construct classification random forests tuned via cross-validation on the training data set partition, i.e.
+without touching the test data, but were unable to develop models that consistently match the B&S random forest method in both cross-validated out-of-sample training predictions and test set predictions.
+
+Given that they are relatively unorthodox, yet appear to work very well, we wonder how the hyper-parameter values were determined.
+Two specific concerns are that this was not done with an eye towards test set accuracy, which would invalidate the independence of the out-of-sample test set, and whether the specific hyper-parameter values are optimized for only one model, or were optimized and found to work well for all models.
+There is no discussion of the random forest tuning strategy or how the specific hyper-parameter methods were determined in the paper.
+
+Given the dramatic changes in results as a result of the preceding issues, we did not further investigate these potential concerns.
+
+# Results of the updated analysis
+
 We now turn to an examination of our replication that fixes the above issues.  The bottom line is that the changes we make turn on their heads the conclusions of B&S.
+
+## Is the Escalation model superior to the alternative ICEWS models?
+
+Table \ref{tab:table1-nosmooth} shows a version of B&S Table 1 with the conventional non-smoothed ROC curves.
+Both the Average and CAMEO models outperform the Escalation model in almost all instances.
+The Goldstein model generally outperforms the Escalation model in the 6-month version.
+The Quad model appears to be roughly on par with the Escalation model.
+Thus, the original B&S conclusion that the escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.
+
+```{r table1-nosmooth}
+table1_nosmooth <- read_csv("data/table1-nosmooth.csv")
+
+table1_nosmooth %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Comparison of the escalation model to alternative ICEWS model using test set AUC-ROC, \\textit{without} smoothed ROC curves (Replication of B\\&S Table 1)",
+    label = "table1-nosmooth") %>% 
+  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
+```
+
+The key difference between our Table \ref{tab:table1-nosmooth} and the original B&S Table 1 is that we calculated AUC-ROC values for the original ROC curves, not smoothed ROC curves.
+B&S's original results and interpretation regarding the superiority of the B&S model over alternative ICEWS models, including the 1,000+ covariate CAMEO model, are entirely conditional on the use of smoothed AUC-ROC.
+Even the Quad model is generally as good as the escalation model in all implementations, and is frequently quite a bit better.
 
@@
     digits = 2, booktabs = TRUE, 
-    caption = "Replication of B\\&S Table 2: Test set AUC-ROC for escalation model with and without structural PITF contribution",
+    caption = "Comparison of escalation model to structural extensions, using test set AUC-ROC (Replication of B\\&S Table 2)",
     label = "table2-fixed") %>%
@@
 
-This is not the case anymore. It outperforms in the 1-month version and under-performs the escalation model in the 6-month version.
+The PITF split-population analog still performs well, but the simple Escalation + PITF predictors model arguable performs better still.
 
@@
 
-Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model.
+Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model, and by quite a maring in the 6-month data version.
 
@@
 
-```{r}
+What impact did the ROC smoothing have overall on the results reported in Table \ref{tab:table1-nosmooth} (B&S Table 1) and Table \ref{tab:table2-fixed} (B&S Table 2)?
+Figure \ref{fig:benefit} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values.
+Each of the gray points corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed}.
+The red crosses mark the average effect of smoothing on AUC-ROC.
+For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible in average.
+The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05.
+This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the result B&S find, namely that the escalation model is generally superior.
+
+```{r benefit, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2)"}
 table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
+table2_benefit <- read_csv("data/table2-smooth-benefit.csv")
+
+benefit <- bind_rows(
   table1_benefit %>%
-  select(-horizon) %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Smoothing advantage for B\\&S Table 1: the gain in AUC-ROC when calculated using smoothed ROC curves",
-    label = "table1-benefit") %>% 
-  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
+    rename(setting = Model) %>%
+    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff"),
+  table2_benefit %>%
+    rename(setting = Model, Escalation = `Escalation Only`) %>%
+    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff")
+)
 
-table2_benefit <- read_csv("data/table2-smooth-benefit.csv") 
-table2_benefit %>%
-  select(-horizon) %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Smoothing advantage for B\\&S Table 2: the gain in AUC-ROC when calculated using smoothed ROC curves",
-    label = "table2-benefit") %>% 
-  column_spec(2:6, width = "2cm") %>%
-  pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1))
+# mean benefit by model
+means <- benefit %>%
+  group_by(model) %>%
+  summarize(avg_benefit = mean(diff))
+
+ann_text <- tibble(horizon = "1 month", diff = 0.07, model = 8.6, label = "Mean")
+
+model_levels <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average", 
+                  "With PITF Predictors", "Weighted by PITF", "PITF Split Population",
+                  "PITF Only")
+benefit %>%
+  mutate(model = factor(model, levels = rev(model_levels))) %>%
+  ggplot(aes(y = model)) + 
+  #facet_wrap(~ horizon, scales = "free_x") +
+  geom_point(aes(x = diff), alpha = 0.5, size = 2, color = "gray50") +
+  geom_point(data = means, aes(x = avg_benefit), shape = 4, color = "red", size = 4) +
+  labs(x = "AUC(smoothed ROC) - AUC(ROC)",
+       y = "") +
+  theme_bw() +
+  geom_text(data = ann_text, aes(x = diff, label = label), col = "red") +
+  geom_vline(xintercept = 0, linetype = 3) +
+  theme(axis.text.y = element_text(size = 12))
 ```
 
-What impact did the ROC smoothing have overall on the results reported in B&S Tables 1 and 2?
-Tables \ref{tab:table1-benefit} and \ref{tab:table2-benefit} show the increase in AUC-ROC when using smoothed ROC curves, compared to the standard non-smoothed AUC-ROC.
-Positive values indicate that smoothing increased a model's apparent performance.
-The escalation model is the only model that consistently had a benefit from smoothing.
-For all eight other models, smoothing sometimes gave a benefit, sometimes not.
-
-A more dramatic difference stands out when we consider the overall average impact of smoothing across all flavors of a model reported in the columns of the tables.
-The Average, Goldstein, Quad, PITF Split Population, and Weighted by PITF models are slightly hurt by smoothing, but on the order of less than 0.01 in absolute magnitude.
-The "With PITF Predictors" model is hurt on the order of 0.02, i.e.
-it appears substantially worse with smoothing.
-The CAMEO and PITF Only models benefit slightly, on the order of 0.01 or less.
-The escalation only model on the other hand has an average boost of 0.058 to its AUC-ROC from smoothing.
-Considering the spread of AUC-ROC values if we compare across rows in B&S Tables 1 and 2, that boost is substantial.
-
-The use of smoothed ROC curves to calculate AUC-ROC values clearly benefits \textbf{only} the escalation model.
+In sum, it is not only the case that using smoothed ROC curves alters the results: the use of smoothed ROC curves to calculate AUC-ROC values clearly benefits \textbf{only} the escalation model.
 It does so consistently and by a considerable margin.
@@
 In this case it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.
-
-# Additional concerns
-
-## Random forest hyperparameters
-
-What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated.
-Table \ref{tab:hp} shows the default values used by the implementation of random forest that B&S use (from the **randomForest** R package), in contrast to the basic settings used by B&S for most the models reported in the paper.
-
-As the outcome is a binary indicator of civil war onset, one would typically use a classification random forest that predicts 0 or 1 labels directly.
-The implementation of random forests that B&S use ([@liaw:wiener:2002]) is based on the original @breiman:2001 implementation and calculates predictive probabilities by averaging over the "0" or "1" votes from all constituent decision trees.
-The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number.
-From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive, and operates on a large bootstrapped sample of the original training data.
-These are of course only heuristics and it is usual to attempt to find better hyper-parameter methods through some form of tuning.
-
-\begin{table}
-\caption{\label{tab:hp} Random forest (\texttt{randomForest()}) default versus B\&S hyperparameters}
-\begin{tabular}{l>{\raggedright\arraybackslash}p{2in}ll}
-\toprule
-Hyperparameter & Default heuristic & Default values (Escalation) & B\&S value \\
-\midrule
-type & & classification & regression \\
-ntree & & 500 & 100,000 or 1e6 \\
-mtry & \texttt{floor(sqrt(ncol(x)))} & 3 & 3 \\
-replace & & true & false \\
-sampsize & \texttt{nrow(x)} if replace, else \texttt{ceiling(.632*nrow(x))} & 11,869 & 100 or
-500 \\
-nodesize & 1 for classification & 1 & 1 \\
-maxnodes & & null & 5 or 10 \\
-\bottomrule
-\end{tabular}
-\end{table}
-
-B&S in contrast fit very large forests with 100,000 trees in the basic model form, but where each tree only operates on a very small sub-sample (N=100 or 500), drawn without replacement, of the available training data.
-This approach only works due to the choice to use regression, not classification, trees.
-Trying to use classification trees with the other parameter settings not work at all because it is almost guaranteed that a sample of 100 from the ~11,000 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample.
-As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:
-
-```
-Warning message:
-In randomForest.default(y = as.integer(train_df$incidence_civil_ns_plus1 ==  :
-  The response has five or fewer unique values.  Are you sure you want to do regression?
-```
-
-As it turns out, using regression random forests for this kind of binary classification problem in order to obtain probability estimates matches the probability random forest approach suggested and positively evaluated in @malley:etal:2012, and which is used in another prominent R implementation of random forests.^[The **ranger** package.] It is not clear whether this is intentional, as the Malley paper is not cited in B&S.
-
-In any case, B&S's random forest approach appears to work really well.
-We tried to construct classification random forests tuned via cross-validation on the training data set partition, i.e.
-without touching the test data, but were unable to develop models that consistently match the B&S random forest method in both cross-validated out-of-sample training predictions and test set predictions.
-
-Given that they are relatively unorthodox, yet appear to work very well, we wonder how the hyper-parameter values were determined.
-Two specific concerns are that this was not done with an eye towards test set accuracy, which would invalidate the independence of the out-of-sample test set, and whether the specific hyper-parameter values are optimized for only one model, or were optimized and found to work well for all models.
-There is no discussion of the random forest tuning strategy or how the specific hyper-parameter methods were determined in the paper.
 
@@
 table2_full %>%
-  select(horizon, column, N_test, cases) %>%
-  pivot_wider(names_from = column, values_from = N_test) %>%
-  arrange(horizon, desc(cases)) %>%
-  rename(Cases = cases, Horizon = horizon) %>%
-  select(Cases, Horizon, everything()) %>%
-  arrange(desc(Cases), Horizon) %>%
-  knitr::kable(
-    "latex", booktabs = TRUE, digits = 0,
-    caption = "Number of valid test predictions for each cell in B\\&S Table 2",
-    label = "table2-N") %>%
-  column_spec(3:7, width = "2cm") %>%
-  collapse_rows(1:2, row_group_label_position = "stack")
-
-table2_full %>%
   pivot_longer(auc_roc:auc_roc_smoothed, names_to = "Smoothed ROC",
@@
 
+# Appendix
+
+Table \ref{tab:table1-smooth} is our replication of B&S Table 1 with smoothed AUC-ROC.
+The results differ slightly from the original B&S Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models.
+It is the case that B&S set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel.
+In any case, the interpretation of results should not be sensitive to random variation, i.e.
+it should not depend on using a specific RNG seed.
+On the basis of these results, B&S conclude that the escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.
+
+```{r table1-smooth}
+table1_smooth <- read_csv("data/table1-smooth.csv")
+
+table1_smooth %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Replication of B\\&S Table 1 with smoothed ROC curves; test set AUC-ROC for various models",
+    label = "table1-smooth") %>% 
+  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
+```
+
+```{r smooth-benefit-tables}
+table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
+table1_benefit %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Smoothing advantage for B\\&S Table 1: the gain in AUC-ROC when calculated using smoothed ROC curves",
+    label = "table1-benefit") %>% 
+  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
+
+table2_benefit <- read_csv("data/table2-smooth-benefit.csv") 
+table2_benefit %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Smoothing advantage for B\\&S Table 2: the gain in AUC-ROC when calculated using smoothed ROC curves",
+    label = "table2-benefit") %>% 
+  column_spec(2:6, width = "2cm") %>%
+  pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1))
+```
+
 # References

### 2020-06-08 rep_nosmooth/recreate-tables.R

@@ -128,6 +128,16 @@ sapply(common_subset, nrow)
 common_subset <- common_subset %>%
   bind_rows(., .id = "horizon")
 
+# Note that the number of positive cases also differs!
+sapply(preds$preds, function(x) {x <- x[complete.cases(x), ]; sum(x$value==1, na.rm=T)})
+preds %>%
+  # reduce each prediction set down to the common case set; after this step
+  # the predictions in each horizon group will have same N.
+  mutate(preds = map(preds, semi_join, y = common_subset,
+                     by = c("year", "period", "country_iso3"))) %>%
+  pull(preds) %>%
+  sapply(., function(x) {x <- x[complete.cases(x), ]; sum(x$value==1, na.rm=T)})
+
 results_table2 <- preds %>%
   # reduce each prediction set down to the common case set; after this step
   # the predictions in each horizon group will have same N.

### 2020-06-08 paper.Rmd
call this version 0.1
Ready to share with close friends.

@@
 
-# Review of Blair and Sambanis 2020
+# Summary of Blair and Sambanis (2020)
 
@@
 
-# Replication problems
+# Implementation issues in B&S
 
 While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation.
-These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) incorrect implementations of the weighted by PITF and PITF analog split-population models, c) inconsistent test sets for the models examined, and d) incorrect scoring of the 2016 forecasts.
-We also note (e) additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox.
-We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.
+These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) incorrect implementations of the weighted by PITF and PITF analog split-population models, c) inconsistent test sets for the models examined, and d) incorrect scoring of the 2016 forecasts.^[We also note (e) additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox.
+We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.]
 
@@
 The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, actual ROC curves.
-A reference to smoothing is made in a single sentence in the paper (p.
-12):
-
-> Figure 1 displays the corresponding ROC curves, smoothed for ease of interpretation.
-
-This implies that the ROC curves were only smoothed in the references Figure 1, but actually all AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation.
-ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result appear step-like.
-Figure 1 shows our replication of both the smoothed ROC curves B&S report, and the actual ROC curves on the right.
-The standard method is to compute the area under the curve AUC statistic on the original, non-smooth ROC curves that are shown on the right.
-
-![Replication of B&S Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)
-
-ROC curves typically appear step-like in response to the distribution of positive and negative cases in the data, which specifically in rare events problems like this one restricts the number of distinct true positive rate values.
-The specific predictions here also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines as seen in the panels on the right.
-In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached.
-For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set.
-Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.
-Notice also that the smoothing averages the left most almost straight line with the right most almost straight line in a monotonic way.
-This ignores the fact that the actual ROC **MIKE**
-
+the data, which specifically in rare events problems like this one restricts the number of distinct true positive rate values.
 As we show below, the use of smoothed ROC curves to calculuate AUC has a dramatic impact on the results.
@@
 The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions.
-Similarly in the 6-month version of this file.] This appears to be a coding error.
+Similarly in the 6-month version of this file.] This appears to be an easily corrected coding error on the part of B&S.
 
@@
 
-Similarly, the "PITF Split Population" model appears to be incorrectly implemented. B&S describe it on page 20:
+Similarly, the "PITF Split Population" model appears to be incorrectly implemented, owing to a coding error. B&S describe it on page 20:
 
@@
 
-The intention clearly was to run two separate random forest models, one each on the low- and high-risk training data splits.
+The intention if B&S appears to have been to run two separate random forest models, one each on the low- and high-risk training data splits.
 The replication code does indeed run two separate random forecasts, but they are both run on the *exact same training data*, which consists of the full training data all other models are run on.
@@
 These numbers appear small enough to be not important, but crucially the already small number of positive cases also changes: the "With PITF Predictors" and "PITF Only" models lose 2 and 1 of 11 positive cases, respectively, in the 1- and 6-month data versions.
+
+## Incorrect scoring of the 2016 forecasts
+
+B&S show a confusion matrix to score their 2016-H1 forecasts in Table 4.
+Although the forecasts are for the probability of civil war onset, in the replication code they are actually scored using the much more common incidence of civil war, i.e.
+including ongoing civil wars as "1"'s.
+
+The relevant variables in the data are "incidence_civil_ns" and "incidence_civil_ns_plus1", which appears to be a 1-period lead version of the DV that is used in the actual prediction models.
+The incidence DV contains both 0/1 and missing values.
+By examining the pattern of missing values, it seems clear that this was originally an incidence variable indicating whether a country was at civil war in a given year or not, and which was converted to an onset version so that onsets retain the value of 1 but continuing civil war years are coded as missing.
+This reflects common practice in how these are coded.
+
+By examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
+In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable back to incidence.
+
+# Results of the updated analysis
+
+We now turn to an examination of our analysis that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/rickmorgan2/Blair-Sambanis-replication.]  The bottom line is that the changes we make turn on their heads the conclusions of B&S.
+
+## Smoothed ROC curves and AUC calculations
+
+A reference to smoothing is made in a single sentence in the paper (p. 12):
+
+> Figure 1 displays the corresponding ROC curves, smoothed for ease of interpretation.
+
+This implies that the ROC curves were only smoothed in the references Figure 1, but actually all AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation.
+ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result appear step-like.
+Figure 1 shows our replication of both the smoothed ROC curves B&S report, and the actual ROC curves on the right.
+The standard method is to compute the area under the curve AUC statistic on the original, non-smooth ROC curves that are shown on the right.
+
+![Replication of B&S Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)
+
+The specific predictions here also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines as seen in the panels on the right.
+In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached.
+For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set.
+Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.
+Notice also that the smoothing averages the left most almost straight line with the right most almost straight line in a monotonic way.
+
+## The effect of using smoothed ROC curves
+
+What impact did the ROC smoothing have overall on the results reported in Table \ref{tab:table1-nosmooth} (B&S Table 1) and Table \ref{tab:table2-fixed} (B&S Table 2)?
+Figure \ref{fig:benefit} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values.
+Each of the gray points corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed}.
+The red crosses mark the average effect of smoothing on AUC-ROC.
+For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible in average.
+The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05.
+This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the result B&S find, namely that the escalation model is generally superior.
+
+```{r benefit, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2)"}
+table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
+table2_benefit <- read_csv("data/table2-smooth-benefit.csv")
+
+benefit <- bind_rows(
+  table1_benefit %>%
+    rename(setting = Model) %>%
+    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff"),
+  table2_benefit %>%
+    rename(setting = Model, Escalation = `Escalation Only`) %>%
+    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff")
+)
+
+# mean benefit by model
+means <- benefit %>%
+  group_by(model) %>%
+  summarize(avg_benefit = mean(diff))
+
+ann_text <- tibble(horizon = "1 month", diff = 0.07, model = 8.6, label = "Mean")
+
+model_levels <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average", 
+                  "With PITF Predictors", "Weighted by PITF", "PITF Split Population",
+                  "PITF Only")
+benefit %>%
+  mutate(model = factor(model, levels = rev(model_levels))) %>%
+  ggplot(aes(y = model)) + 
+  #facet_wrap(~ horizon, scales = "free_x") +
+  geom_point(aes(x = diff), alpha = 0.5, size = 2, color = "gray50") +
+  geom_point(data = means, aes(x = avg_benefit), shape = 4, color = "red", size = 4) +
+  labs(x = "AUC(smoothed ROC) - AUC(ROC)",
+       y = "") +
+  theme_bw() +
+  geom_text(data = ann_text, aes(x = diff, label = label), col = "red") +
+  geom_vline(xintercept = 0, linetype = 3) +
+  theme(axis.text.y = element_text(size = 12))
+```
+
+In sum, it is not only the case that using smoothed ROC curves alters the results: the use of smoothed ROC curves to calculate AUC-ROC values clearly benefits \textbf{only} the escalation model.
+It does so consistently and by a considerable margin.
+All eight alternative models reported in Tables 1 and 2 on average do not gain when using smoothed ROC curves to calculate AUC.
+
+## Using Inconsistent Ns
 
@@
 Since the sets of incomplete cases in the original version above do not completely overlap themselves, the common subset for all models is actually slightly smaller than the smallest N in the top portion of the table.
-
-## Incorrect scoring of the 2016 forecasts
-
-B&S show a confusion matrix to score their 2016-H1 forecasts in Table 4.
-Although the forecasts are for the probability of civil war onset, in the replication code they are actually scored using the much more common incidence of civil war, i.e.
-including ongoing civil wars as "1"'s.
-
-The relevant variables in the data are "incidence_civil_ns" and "incidence_civil_ns_plus1", which appears to be a 1-period lead version of the DV that is used in the actual prediction models.
-The incidence DV contains both 0/1 and missing values.
-By examining the pattern of missing values, it seems clear that this was originally an incidence variable indicating whether a country was at civil war in a given year or not, and which was converted to an onset version so that onsets retain the value of 1 but continuing civil war years are coded as missing.
-This reflects common practice in how these are coded.
-
-By examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
-In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable back to incidence.
-
-## Random forest hyperparameters
-
-What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated.
-Table \ref{tab:hp} shows the default values used by the implementation of random forest that B&S use (from the **randomForest** R package), in contrast to the basic settings used by B&S for most the models reported in the paper.
-
-As the outcome is a binary indicator of civil war onset, one would typically use a classification random forest that predicts 0 or 1 labels directly.
-The implementation of random forests that B&S use ([@liaw:wiener:2002]) is based on the original @breiman:2001 implementation and calculates predictive probabilities by averaging over the "0" or "1" votes from all constituent decision trees.
-The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number.
-From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive, and operates on a large bootstrapped sample of the original training data.
-These are of course only heuristics and it is usual to attempt to find better hyper-parameter methods through some form of tuning.
-
-\begin{table}
-\caption{\label{tab:hp} Random forest (\texttt{randomForest()}) default versus B\&S hyperparameters}
-\begin{tabular}{l>{\raggedright\arraybackslash}p{2in}ll}
-\toprule
-Hyperparameter & Default heuristic & Default values (Escalation) & B\&S value \\
-\midrule
-type & & classification & regression \\
-ntree & & 500 & 100,000 or 1e6 \\
-mtry & \texttt{floor(sqrt(ncol(x)))} & 3 & 3 \\
-replace & & true & false \\
-sampsize & \texttt{nrow(x)} if replace, else \texttt{ceiling(.632*nrow(x))} & 11,869 & 100 or
-500 \\
-nodesize & 1 for classification & 1 & 1 \\
-maxnodes & & null & 5 or 10 \\
-\bottomrule
-\end{tabular}
-\end{table}
-
-B&S in contrast fit very large forests with 100,000 trees in the basic model form, but where each tree only operates on a very small sub-sample (N=100 or 500), drawn without replacement, of the available training data.
-This approach only works due to the choice to use regression, not classification, trees.
-Trying to use classification trees with the other parameter settings not work at all because it is almost guaranteed that a sample of 100 from the ~11,000 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample.
-As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:
-
-```
-Warning message:
-In randomForest.default(y = as.integer(train_df$incidence_civil_ns_plus1 ==  :
-  The response has five or fewer unique values.  Are you sure you want to do regression?
-```
-
-As it turns out, using regression random forests for this kind of binary classification problem in order to obtain probability estimates matches the probability random forest approach suggested and positively evaluated in @malley:etal:2012, and which is used in another prominent R implementation of random forests.^[The **ranger** package.] It is not clear whether this is intentional, as the Malley paper is not cited in B&S.
-
-In any case, B&S's random forest approach appears to work really well.
-We tried to construct classification random forests tuned via cross-validation on the training data set partition, i.e.
-without touching the test data, but were unable to develop models that consistently match the B&S random forest method in both cross-validated out-of-sample training predictions and test set predictions.
-
-Given that they are relatively unorthodox, yet appear to work very well, we wonder how the hyper-parameter values were determined.
-Two specific concerns are that this was not done with an eye towards test set accuracy, which would invalidate the independence of the out-of-sample test set, and whether the specific hyper-parameter values are optimized for only one model, or were optimized and found to work well for all models.
-There is no discussion of the random forest tuning strategy or how the specific hyper-parameter methods were determined in the paper.
-
-Given the dramatic changes in results as a result of the preceding issues, we did not further investigate these potential concerns.
-
-# Results of the updated analysis
-
-We now turn to an examination of our replication that fixes the above issues.  The bottom line is that the changes we make turn on their heads the conclusions of B&S.
 
@@
 
-## The effect of using smoothed ROC curves
-
-What impact did the ROC smoothing have overall on the results reported in Table \ref{tab:table1-nosmooth} (B&S Table 1) and Table \ref{tab:table2-fixed} (B&S Table 2)?
-Figure \ref{fig:benefit} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values.
-Each of the gray points corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed}.
-The red crosses mark the average effect of smoothing on AUC-ROC.
-For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible in average.
-The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05.
-This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the result B&S find, namely that the escalation model is generally superior.
-
-```{r benefit, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2)"}
-table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
-table2_benefit <- read_csv("data/table2-smooth-benefit.csv")
-
-benefit <- bind_rows(
-  table1_benefit %>%
-    rename(setting = Model) %>%
-    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff"),
-  table2_benefit %>%
-    rename(setting = Model, Escalation = `Escalation Only`) %>%
-    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff")
-)
-
-# mean benefit by model
-means <- benefit %>%
-  group_by(model) %>%
-  summarize(avg_benefit = mean(diff))
-
-ann_text <- tibble(horizon = "1 month", diff = 0.07, model = 8.6, label = "Mean")
-
-model_levels <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average", 
-                  "With PITF Predictors", "Weighted by PITF", "PITF Split Population",
-                  "PITF Only")
-benefit %>%
-  mutate(model = factor(model, levels = rev(model_levels))) %>%
-  ggplot(aes(y = model)) + 
-  #facet_wrap(~ horizon, scales = "free_x") +
-  geom_point(aes(x = diff), alpha = 0.5, size = 2, color = "gray50") +
-  geom_point(data = means, aes(x = avg_benefit), shape = 4, color = "red", size = 4) +
-  labs(x = "AUC(smoothed ROC) - AUC(ROC)",
-       y = "") +
-  theme_bw() +
-  geom_text(data = ann_text, aes(x = diff, label = label), col = "red") +
-  geom_vline(xintercept = 0, linetype = 3) +
-  theme(axis.text.y = element_text(size = 12))
-```
-
-In sum, it is not only the case that using smoothed ROC curves alters the results: the use of smoothed ROC curves to calculate AUC-ROC values clearly benefits \textbf{only} the escalation model.
-It does so consistently and by a considerable margin.
-All eight alternative models reported in Tables 1 and 2 on average do not gain when using smoothed ROC curves to calculate AUC.
-
 ## How accurate were the 2016 forecasts?
@@
 
-B&S have a clear misunderstanding of the role that out-of-sample prediction can play in analysis.
-On the one hand it can be used for simple forecasting, while on the other it can be used to evaluate the performance of models, specifically to overfitting and bias.
-They further misrepresent cross-validation---which is not about fishing for results, as claimed by B&S.
-The current state-of-the-art uses prediction and cross-validation most frequently to provide supportive evidence that is independent of the estimation procedure and the in-sample data.
-It is incorrect to paint this procedure as atheoretical since many studies will have some explanation for how the model was constructed.
-These procedures may simply be used to provide evidence for a theoretical argument.
-Though that argument may be unconvincing and/or atheoretical.
-
 The B&S approach that is advocated is to use theory to guide prediction.
@@
 As such their baseline comparison is a poor standard bearer for strong theory.
-
-Further, they completely misunderstand the use of ICEWS event data in current research.
+Moreover, they misunderstand the use of ICEWS event data in current research.
 They claim that most uses to date have focused on the quad categories, but this ignores a wide swath of literature (Steinert-Threlkeld APSR 2017) and Metternich et al (AJPS 2013) that uses a specific action---such as protest---defined in the CAMEO ontology.
-In the AJPS article we hand coded, for example, every actor in Thailand and focused on an analysis of how those have been interacting.
+In the AJPS article on Thailand, Metternich et al hand coded, for example, every actor in Thailand and focused on an analysis of how those interact.
 
@@
 
+\newpage
+
+# (APPENDIX) Appendix {-}
+# Appendix A
+
 ```{r}
@@
 ```
-\newpage
-
-# Appendix
 
@@
 
+# Appendix B
+## Some additional, minor issues
+### Random forest hyperparameters
+
+What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated.
+Table \ref{tab:hp} shows the default values used by the implementation of random forest that B&S use (from the **randomForest** R package), in contrast to the basic settings used by B&S for most the models reported in the paper.
+
+As the outcome is a binary indicator of civil war onset, one would typically use a classification random forest that predicts 0 or 1 labels directly.
+The implementation of random forests that B&S use ([@liaw:wiener:2002]) is based on the original @breiman:2001 implementation and calculates predictive probabilities by averaging over the "0" or "1" votes from all constituent decision trees.
+The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number.
+From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive, and operates on a large bootstrapped sample of the original training data.
+These are of course only heuristics and it is usual to attempt to find better hyper-parameter methods through some form of tuning.
+
+\begin{table}
+\caption{\label{tab:hp} Random forest (\texttt{randomForest()}) default versus B\&S hyperparameters}
+\begin{tabular}{l>{\raggedright\arraybackslash}p{2in}ll}
+\toprule
+Hyperparameter & Default heuristic & Default values (Escalation) & B\&S value \\
+\midrule
+type & & classification & regression \\
+ntree & & 500 & 100,000 or 1e6 \\
+mtry & \texttt{floor(sqrt(ncol(x)))} & 3 & 3 \\
+replace & & true & false \\
+sampsize & \texttt{nrow(x)} if replace, else \texttt{ceiling(.632*nrow(x))} & 11,869 & 100 or
+500 \\
+nodesize & 1 for classification & 1 & 1 \\
+maxnodes & & null & 5 or 10 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+B&S in contrast fit very large forests with 100,000 trees in the basic model form, but where each tree only operates on a very small sub-sample (N=100 or 500), drawn without replacement, of the available training data.
+This approach only works due to the choice to use regression, not classification, trees.
+Trying to use classification trees with the other parameter settings not work at all because it is almost guaranteed that a sample of 100 from the ~11,000 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample.
+As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:
+
+```
+Warning message:
+In randomForest.default(y = as.integer(train_df$incidence_civil_ns_plus1 ==  :
+  The response has five or fewer unique values.  Are you sure you want to do regression?
+```
+
+As it turns out, using regression random forests for this kind of binary classification problem in order to obtain probability estimates matches the probability random forest approach suggested and positively evaluated in @malley:etal:2012, and which is used in another prominent R implementation of random forests.^[The **ranger** package.] It is not clear whether this is intentional, as the Malley paper is not cited in B&S.
+
+In any case, B&S's random forest approach appears to work really well.
+We tried to construct classification random forests tuned via cross-validation on the training data set partition, i.e.
+without touching the test data, but were unable to develop models that consistently match the B&S random forest method in both cross-validated out-of-sample training predictions and test set predictions.
+
+Given that they are relatively unorthodox, yet appear to work very well, we wonder how the hyper-parameter values were determined.
+Two specific concerns are that this was not done with an eye towards test set accuracy, which would invalidate the independence of the out-of-sample test set, and whether the specific hyper-parameter values are optimized for only one model, or were optimized and found to work well for all models.
+There is no discussion of the random forest tuning strategy or how the specific hyper-parameter methods were determined in the paper.
+
+Given the dramatic changes in results as a result of the preceding issues, we did not further investigate these potential concerns.
+
 # References

### 2020-06-08 paper.Rmd
0.11
frakk
let there be appendix

@@
 output: 
-  pdf_document:
+  bookdown::pdf_document2:
     keep_tex: true
+    toc: false
 bibliography: 
@@
 
+# References
+
+<div id="refs"></div>
+
 \newpage
@@
 # (APPENDIX) Appendix {-}
-# Appendix A
+
+# Appendix
+
+\renewcommand{\thefigure}{A\arabic{figure}}
+\setcounter{figure}{0}
+
+\renewcommand{\thetable}{A\arabic{table}} 
+\setcounter{table}{0}
+
+## Additional replication tables
+
+Table \ref{tab:table1-smooth} is our replication of B&S Table 1 with smoothed AUC-ROC.
+The results differ slightly from the original B&S Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models.
+It is the case that B&S set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel.
+In any case, the interpretation of results should not be sensitive to random variation, i.e.
+it should not depend on using a specific RNG seed.
+On the basis of these results, B&S conclude that the escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.
+
+```{r table1-smooth}
+table1_smooth <- read_csv("data/table1-smooth.csv")
+
+table1_smooth %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Replication of B\\&S Table 1 with smoothed ROC curves; test set AUC-ROC for various models",
+    label = "table1-smooth") %>% 
+  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
+```
 
@@
 
-Table \ref{tab:table1-smooth} is our replication of B&S Table 1 with smoothed AUC-ROC.
-The results differ slightly from the original B&S Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models.
-It is the case that B&S set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel.
-In any case, the interpretation of results should not be sensitive to random variation, i.e.
-it should not depend on using a specific RNG seed.
-On the basis of these results, B&S conclude that the escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.
-
-```{r table1-smooth}
-table1_smooth <- read_csv("data/table1-smooth.csv")
-
-table1_smooth %>%
-  select(-horizon) %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Replication of B\\&S Table 1 with smoothed ROC curves; test set AUC-ROC for various models",
-    label = "table1-smooth") %>% 
-  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
-```
-
 ```{r smooth-benefit-tables}
@@
 
-# Appendix B
-## Some additional, minor issues
-### Random forest hyperparameters
+## Random forest hyper-parameter selection
 
@@
 Given the dramatic changes in results as a result of the preceding issues, we did not further investigate these potential concerns.
-
-# References

### 2020-06-08 paper.Rmd
edits
- switch the b&s table 1 substantive interpretation with the test N discussion, so that it is now right below the smoothing effect discussion. The test N issue only impacts table 2 so it makes more sense to have it right before the b&s table 2 substantive discussion.
-

@@
 The implementation error aside, this split-population analog model is actually quite odd and does not actually replicate an analogue of the idea behind split-population modeling.
-Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision (regression^[See further below.
-Although the RF models are used for a binary decision problem, the actual implementation uses regression RFs for continuous outcomes.]) trees.
+Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision trees.
 Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does indeed have 200,000 decision trees.
@@
 ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available.
-Those sets differ across models (columns) for each row in Table 2.
-The top portion of Table \ref{tab:table2-N} shows the number of valid test set predictions that can be scored for each model in the original B&S Table 2.
-For the 1-month data version, the number of predictions differs by up to 500 cases, and in the 6-month data version by around 100 cases.
-These numbers appear small enough to be not important, but crucially the already small number of positive cases also changes: the "With PITF Predictors" and "PITF Only" models lose 2 and 1 of 11 positive cases, respectively, in the 1- and 6-month data versions.
+Those sets differ across models (columns) for each row in B&S Table 2.
+
+Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
+In other words, the results for different models in B&S Table 2 are actually not comparable to one another, and any conclusions drawn from such comparison are potentially incorrect.
 
@@
 
-We now turn to an examination of our analysis that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/rickmorgan2/Blair-Sambanis-replication.]  The bottom line is that the changes we make turn on their heads the conclusions of B&S.
+We now turn to an examination of our analysis that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/rickmorgan2/Blair-Sambanis-replication.] The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B&S Table 2).
+We will review the substantive implications of our updated analysis further below, but the bottom line is that the changes we make turn on their heads the conclusions of B&S.
 
@@
 
-This implies that the ROC curves were only smoothed in the references Figure 1, but actually all AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation.
+This implies that the ROC curves were only smoothed in the referenced Figure 1, but actually all AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation.
 ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result appear step-like.
@@
 
-What impact did the ROC smoothing have overall on the results reported in Table \ref{tab:table1-nosmooth} (B&S Table 1) and Table \ref{tab:table2-fixed} (B&S Table 2)?
+What impact did the ROC smoothing have overall on the performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B&S Table 2)?
 Figure \ref{fig:benefit} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values.
@@
   geom_point(aes(x = diff), alpha = 0.5, size = 2, color = "gray50") +
-  geom_point(data = means, aes(x = avg_benefit), shape = 4, color = "red", size = 4) +
+  geom_point(data = means, aes(x = avg_benefit), shape = "|", color = "black", size = 6) +
   labs(x = "AUC(smoothed ROC) - AUC(ROC)",
@@
   theme_bw() +
-  geom_text(data = ann_text, aes(x = diff, label = label), col = "red") +
+  geom_text(data = ann_text, aes(x = diff, label = label), col = "black") +
   geom_vline(xintercept = 0, linetype = 3) +
@@
 It does so consistently and by a considerable margin.
-All eight alternative models reported in Tables 1 and 2 on average do not gain when using smoothed ROC curves to calculate AUC.
+All eight alternative models reported in B&S Tables 1 and 2 on average do not gain when using smoothed ROC curves to calculate AUC.
+
+## Is the Escalation model superior to the alternative ICEWS models?
+
+Table \ref{tab:table1-nosmooth} (B&S Table 1) shows the comparison of the Escalation model to other alternative models based on ICEWS event data indicators.
+The AUC-ROC values are based on original, non-smoothed ROC curves.
+
+Both the Average and CAMEO models outperform the Escalation model in almost all instances.
+The Goldstein model generally outperforms the Escalation model in the 6-month version.
+The Quad model appears to be roughly on par with the Escalation model.
+Thus, the original B&S conclusion that the escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.
+
+```{r table1-nosmooth}
+table1_nosmooth <- read_csv("data/table1-nosmooth.csv")
+
+table1_nosmooth %>%
+  select(-horizon) %>%
+  knitr::kable(
+    digits = 2, booktabs = TRUE, 
+    caption = "Comparison of the escalation model to alternative ICEWS model using test set AUC-ROC, \\textit{without} smoothed ROC curves (Replication of B\\&S Table 1)",
+    label = "table1-nosmooth") %>% 
+  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
+```
+
+The key difference between Table \ref{tab:table1-nosmooth} and B&S Table 1 is whether the underlying ROC curves were original or smoothed; the test set N and coding issues did not affect this set of results.^[We should note that in our replication we consciously decided to not set RNG seeds, even though the random forest models are non-deterministic and vary slightly from run to run.
+As we changed the replication code to allow running in parallel, we cannot exactly reproduce the B&S results even with the same RNG seed.
+More importantly, the substantive interpretation of results should not depend on a specific RNG seed.
+We found that the escalation model's AUC-ROC values generally fluctuate no more than 0.01 (see https://github.com/rickmorgan2/Blair-Sambanis-replication/blob/master/rep_nosmooth/variance.md), and thus are confident that the patterns we see are robust to the initial RNG state.] B&S's original results and interpretation regarding the superiority of the B&S model over alternative ICEWS models, including the 1,000+ covariate CAMEO model, are thus entirely conditional on the use of smoothed AUC-ROC.
+Even the Quad model is generally as good as the escalation model in all implementations, and is frequently quite a bit better.
 
@@
 
-Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
-In other words, the results for different models in B&S Table 2 are actually not comparable to one another, and any conclusions drawn from such comparison are potentially incorrect.
+The top portion of Table \ref{tab:table2-N} shows the number of valid test set predictions that can be scored for each model in the original B&S Table 2.
+For the 1-month data version, the number of predictions differs by up to 500 cases, and in the 6-month data version by around 100 cases.
+These numbers appear small enough to be not important, but crucially the already small number of positive cases also changes: the "With PITF Predictors" and "PITF Only" models lose 2 and 1 of 11 positive cases, respectively, in the 1- and 6-month data versions.
 
@@
 
-## Is the Escalation model superior to the alternative ICEWS models?
-
-Table \ref{tab:table1-nosmooth} shows a version of B&S Table 1 with the conventional non-smoothed ROC curves.
-Both the Average and CAMEO models outperform the Escalation model in almost all instances.
-The Goldstein model generally outperforms the Escalation model in the 6-month version.
-The Quad model appears to be roughly on par with the Escalation model.
-Thus, the original B&S conclusion that the escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.
-
-```{r table1-nosmooth}
-table1_nosmooth <- read_csv("data/table1-nosmooth.csv")
-
-table1_nosmooth %>%
-  select(-horizon) %>%
-  knitr::kable(
-    digits = 2, booktabs = TRUE, 
-    caption = "Comparison of the escalation model to alternative ICEWS model using test set AUC-ROC, \\textit{without} smoothed ROC curves (Replication of B\\&S Table 1)",
-    label = "table1-nosmooth") %>% 
-  pack_rows(index = c("One-month forecasts" = 9, "Six-month forecasts" = 9))
-```
-
-The key difference between our Table \ref{tab:table1-nosmooth} and the original B&S Table 1 is that we calculated AUC-ROC values for the original ROC curves, not smoothed ROC curves.
-B&S's original results and interpretation regarding the superiority of the B&S model over alternative ICEWS models, including the 1,000+ covariate CAMEO model, are entirely conditional on the use of smoothed AUC-ROC.
-Even the Quad model is generally as good as the escalation model in all implementations, and is frequently quite a bit better.
-
 ## Do structural variables add to the Escalation model?
@@
 Table \ref{tab:table2-fixed} shows our replication of B&S Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common, joint subset of tests cases that all models have non-missing predictions for.
-Table \ref{tab:table2-full} further below shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.
+Table \ref{tab:table2-full} in the appendix shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.
 
@@
 It is not a procedure.
-What they actually do is to create a model with four right hand side variables that is supposed to capture a complicated repression-dissent dynamic.
-The dynamic is probably not linear, but their model is.
+What they actually do is to create a model with ten right hand side variables that is supposed to capture a complicated repression-dissent dynamic.
 There is a wide-ranging literature on this dynamic that they do not rely on to construct their model.

### 2020-06-08 paper.Rmd
version 0.2
yup

@@
 date: "`r format(Sys.time(), '%d %B %Y')`"
+thanks: Cassy L. Dorff provided helpful feedback on this project.
+abstract: We examine the research protocols in Blair and Sambanis (2020).
+We find that there are several important mistakes and research decisions that determine their results.
+Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.
+While their model is not very theoretical, it is outperformed by several of the ad hoc, non-theoretical models they examine.
 output: 
@@
 
+library(bookdown)
 library(readr)
@@
 
-We infer that the intent is that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
+B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
 B&S, however, actually weights the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`).
@@
 We now turn to an examination of our analysis that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/rickmorgan2/Blair-Sambanis-replication.] The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B&S Table 2).
-We will review the substantive implications of our updated analysis further below, but the bottom line is that the changes we make turn on their heads the conclusions of B&S.
+We will review the substantive implications of our updated analysis further below, but the bottom line is that these changes turn B&S's conclusions on their heads.
 
@@
 For the 1-month data version, the number of predictions differs by up to 500 cases, and in the 6-month data version by around 100 cases.
-These numbers appear small enough to be not important, but crucially the already small number of positive cases also changes: the "With PITF Predictors" and "PITF Only" models lose 2 and 1 of 11 positive cases, respectively, in the 1- and 6-month data versions.
+These numbers appear small enough to be not important.
+However, the already small number of positive cases also is affected substantially:  "With PITF Predictors" and "PITF Only" models lose two or one (respectively) of 11 positive cases in the 1- and 6-month data versions.
 
@@
 
-We fix this issue by only using predictions for common joint subset of cases that all models have non-missing predictions for.
-The resulting number of cases for each model are shown in the bottom portion of the table.
+Generally, these comparisions require the same cases.
+Therefore, we use predictions for common joint subset of nonmissing predictions The resulting numbers of cases for each model are shown in the bottom portion of the table.
 Since the sets of incomplete cases in the original version above do not completely overlap themselves, the common subset for all models is actually slightly smaller than the smallest N in the top portion of the table.
@@
 The B&S approach that is advocated is to use theory to guide prediction.
-But theory is an ambiguous and undefined concept.
+But theory as used by B&S is an ambiguous and undefined concept.
 It is not a procedure.

### 2020-06-09 rep_nosmooth/variance.Rmd
add Goldstein to variance note
generally still the case that varation is within 0.01, with some outliers for Quad

@@
 model_table <- crossing(
-  model   = c("Escalation", "Quad"),
+  model   = c("Escalation", "Quad", "Goldstein"),
   horizon = c("1 month", "6 months"),
@@
 hlines <- bind_rows(
-  tibble(smoothed = "Smoothed", horizon = "1 month", 
-         y = c(.79, .80, .81, .82, .83, .84, .85) + .005),
-  tibble(smoothed = "Original", horizon = "1 month", 
-         y = c(.77, .78, .79, .8) + .005),
-  tibble(smoothed = "Smoothed", horizon = "6 months", 
-         y = c(.77, .78, .79, .80, .81, .82) + .005),
-  tibble(smoothed = "Original", horizon = "6 months", 
-         y = c(.76, .77, .78, .79) + .005)
+  tibble(horizon = "1 month", smoothed = "Original", 
+         y = c(.77, .78, .79, .8, .81) + .005),
+  tibble(horizon = "1 month", smoothed = "Smoothed", 
+         y = c(.78, .79, .80, .81, .82, .83, .84, .85) + .005),
+  tibble(horizon = "6 months", smoothed = "Original", 
+         y = seq(.76, .83, by = 0.01) + .005),
+  tibble(horizon = "6 months", smoothed = "Smoothed", 
+         y = seq(.76, .82, by = 0.01) + .005)
 )

### 2020-06-09 paper.Rmd
add model 2 model comparison plots in appendix
also drop in last rep_nosmooth run results

@@
 
-```{r benefit, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2)"}
+```{r benefit-plot, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2)"}
 table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
@@
 
+## Model to model comparison plots
+
+Figures \ref{fig:table1-plot} and \ref{fig:table2-plot} replicate the information in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} in a way makes the comparison of the escalation model to the alternative models easier.
+Each facet shows the escalation AUC-ROC for all model settings (the rows in Table \ref{fig:table1-plot} and base specification for all models in Table \ref{tab:table2-fixed}) on the left, and the AUC-ROC for an alternative model on the right, with a connecting line.
+If the lines slope up to the right, the alternative model is better.
+
+```{r table1-plot, fig.height=5, fig.width=8, out.width = ".9\\linewidth", fig.align = "center", fig.cap = "Escalation to alternative comparisons for the ICEWS models (B\\&S Table 1)"}
+table1_nosmooth <- read_csv("data/table1-nosmooth.csv")
+
+icews_lvls <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average")
+
+df <- table1_nosmooth %>%
+  rename(Spec = Model) %>%
+  pivot_longer(Escalation:Average, names_to = "Model", values_to = "auc_roc") %>%
+  mutate(Model = factor(Model, levels = icews_lvls)) %>%
+  nest(data = c(Spec, auc_roc))
+
+pairs <- full_join(
+  df %>% filter(Model!="Escalation"),
+  df %>% filter(Model=="Escalation") %>% rename(data2 = data) %>% select(horizon, data2),
+  by = c("horizon")
+) %>%
+  pivot_longer(data:data2, names_to = "xname", values_to = "data") %>%
+  mutate(xname = ifelse(xname=="data", as.character(Model), "Escalation"),
+         xname = factor(xname, levels = icews_lvls))
+
+pairs %>%
+  unnest(data) %>%
+  ggplot(aes(x = xname, y = auc_roc)) +
+  facet_grid(horizon ~ Model, scales = "free_x") +
+  geom_point() +
+  geom_line(aes(group = Spec)) +
+  theme_bw() +
+  theme(panel.grid.minor = element_blank()) +
+  labs(x = "", y = "AUC-ROC (empirical)")
+```
+
+```{r table2-plot, fig.height=5, fig.width=8, out.width = ".9\\linewidth", fig.align = "center", fig.cap = "Escalation to alternative comparisons for the structural models (B\\&S Table 2)"}
+table2_nosmooth <- read_csv("data/table2-nosmooth.csv")
+
+model_lvls <- colnames(table2_nosmooth)[-c(1:2)]
+
+df <- table2_nosmooth %>%
+  rename(Spec = Model) %>%
+  pivot_longer(`Escalation Only`:`PITF Only`, names_to = "Model", values_to = "auc_roc") %>%
+  mutate(Model = factor(Model, levels = model_lvls)) %>%
+  nest(data = c(Spec, auc_roc))
+
+pairs <- full_join(
+  df %>% filter(Model!="Escalation Only"),
+  df %>% filter(Model=="Escalation Only") %>% rename(data2 = data) %>% select(horizon, data2),
+  by = c("horizon")
+) %>%
+  pivot_longer(data:data2, names_to = "xname", values_to = "data") %>%
+  mutate(xname = ifelse(xname=="data", as.character(Model), "Escalation Only"),
+         xname = factor(xname, levels = model_lvls))
+
+pairs %>%
+  unnest(data) %>%
+  ggplot(aes(x = xname, y = auc_roc)) +
+  facet_grid(horizon ~ Model, scales = "free_x") +
+  geom_point() +
+  geom_line(aes(group = Spec)) +
+  theme_bw() +
+  theme(panel.grid.minor = element_blank(),
+        axis.text.x = element_text(angle = 30, hjust = 1)) +
+  labs(x = "", y = "AUC-ROC (empirical)")
+```
+
 ## Random forest hyper-parameter selection

### 2020-06-09 paper.Rmd
maybe penultimate
version 0.9

@@
  - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com. corresponding author]
- - Richard K. Morgan^[USAID, rick.morgan2@gmail.com. The views expressed are those of the author and do not reflect the official policy or position of USAID or the US Government.]
+ - Richard K. Morgan^[VDEM, rick.morgan2@gmail.com.]
  - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com]
 date: "`r format(Sys.time(), '%d %B %Y')`"
-thanks: Cassy L. Dorff provided helpful feedback on this project.
+thanks: Cassy L. Dorff and Shahryar Minhas both provided helpful feedback on this project. All the code and several additional analyses can be found at our replication archive.
 abstract: We examine the research protocols in Blair and Sambanis (2020).
@@
 Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.
-While their model is not very theoretical, it is outperformed by several of the ad hoc, non-theoretical models they examine.
+While their model is not very theoretical, it is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.
 output: 
@@
     toc: false
+# spacing: double
 bibliography: 
@@
 # Introduction
-
-Blair and Sambanis (2020; hereafter B&S) argue that theory is important for creating models that have high accuracy in forecasting civil war onset.
-Indeed they argue that with such theory, forecasting is more accurate than has previously been possible.
+ @blair:sambanis:2020 (hereafter B\&S) argue that theory is essential for creating models that have high accuracy in forecasting civil war onset.
+Indeed they assert that with such theory, forecasting is more accurate than has previously been possible.
 Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in their article.
 We find that these claims are false.
-Their theory-based escalation model does not do better than the alternatives that they examine, indeed it does worse.
-The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure, the principal of which is that the performance results they report are based on smoothed performance curves, not the original unsmoothed curves.
-In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented, and we also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset.
-In what follows we show the impact of these mistakes on the conclusions, we examine their understanding of prediction itself, and we explore the logic of their assertion about theory as well.
+Their theory-based escalation model does not do better than the alternatives that they examine.
+Indeed, it does worse.
+The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure.
+The performance results they report are based on smoothed performance curves, not the original unsmoothed curves.
+Two of the structural alternatives to their basic escalation model were incorrectly implemented.
+We also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset.
+In what follows, we show the impact of these mistakes on the conclusions.
 
-B&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkable accurately." Indeed they argue that this so-called theoretical model outperforms four other "more mechanical" alternatives.
-Second, they clam that the integration of structure with process are better over short forecasting windows.
-Third they preregistered the list of thirty countries which have the highest risk of civil war onset.
-They claim that such prospective predictions are rare in the literature when in fact they have been routine for many years with several prominent projects.
-B&S claim to be unique in assessing these forecasts.
-A qualitative analysis of their predictions allows them to conclude that their model is strong and a more precise test is undertaken.
+B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars ``remarkable accurately.'' Indeed they argue that this so-called theoretical model outperforms four other ``more mechanical'' alternatives.
+Second, they claim that the integration of structure with process is better over short forecasting windows.
+Third, they preregistered the list of thirty countries that have the highest risk of civil war onset.
+They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.
+B\&S claim to be unique in assessing these forecasts.
+A qualitative analysis of their predictions allows them to conclude that their model is robust, and a more precise test is undertaken.
 We will return to their analysis later, after correcting the procedural mistakes we found in their research process.
@@
 
-@blair:sambanis:2020 (B&S hereafter) aim to examine whether theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
-For the moment we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models.
-In short, they uphold their hypothesis that theory guided empirical research produces better conflict predictions than machine learning inspired efforts that are essentially ad hoc combinations of available variables.
+Blair and Sambanis (2020) (B\&S hereafter) aim to examine whether 
+theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
+For the moment, we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models.
+In short, they uphold their assumption that theory-guided empirical research produces better conflict predictions than machine learning inspired efforts that are necessarily ad hoc combinations of available variables.
 They arrive at this conclusion by examining the problem of predicting civil war onset.
@@
 - Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
-- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government one one side and opposition or rebel actors on the other.
+- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government on one side and opposition or rebel actors on the other.
 These are directed, thus making for four total covariates.
@@
 
-Our belief is that these research decisions and issues lead B&S to draw conclusion that are incorrect.
-The escalation model is not the best and it actually performs worse that the atheoretical, garbage can model with over $1000$ variables.
-We turn to discussing these five issues below.
-We hold a complete analysis that corrects all these issues until later, as there are many possible permuations of a seriatim unfolding.
+We believe that these research decisions and issues lead B\&S to incorrect conclusions.
+The escalation model is not the best, and it actually performs worse than the atheoretical, garbage can model with over \(1000\) variables.
+We discuss these five issues below.
+We defer a complete analysis that corrects all these issues until later, as there are many possible permutations of a seriatim unfolding.
 
@@
 The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, actual ROC curves.
-the data, which specifically in rare events problems like this one restricts the number of distinct true positive rate values.
-As we show below, the use of smoothed ROC curves to calculuate AUC has a dramatic impact on the results.
+The data in rare events problems like this one restricts the number of distinct true positive rate values.
+As we show below, the use of smoothed ROC curves to calculate AUC has a dramatic impact on the results.
 To our knowledge, the norm is to calculate AUC-ROC values using original, not smoothed ROC curves.
-We are in fact not aware of other work that uses smoothed ROC curves for AUC calculations.
-It might be that there are theoretical reasons justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the B&S results, it minimally would have warranted an explicit discussion in the paper.
+In fact, we are not aware of other work that uses smoothed ROC curves for AUC calculations.
+It might be that theoretical reasons are justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the B&S results, it minimally would have warranted an explicit discussion in the paper.
 As it is, B&S offer no justification for using smoothed ROC curves in the AUC-ROC calculations.
+
+Smoothing of ROC curves is well known.
+Basically, the smoothed curves transform the false positive rate and the false negative rates jointly in a monotone fashion.
+The assumption is that there is a monotone transformation that produces normal distributions for the sensitivity and specificity.
+It is well-known that there is substantial bias that can be introduced by bias.
+Indeed, bootstrapping is often used as a diagnostic to see the amount of bias introduced by smoothing.
 
@@
 The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions.
-Similarly in the 6-month version of this file.] This appears to be an easily corrected coding error on the part of B&S.
+Similarly, in the 6-month version of this file.] This appears to be an easily corrected coding error on the part of B&S.
 
@@
 
-Similarly, the "PITF Split Population" model appears to be incorrectly implemented, owing to a coding error. B&S describe it on page 20:
+The "PITF Split Population" model appears to be incorrectly implemented, owing to a coding error. B&S describe it on page 20:
 
@@
 
-The intention if B&S appears to have been to run two separate random forest models, one each on the low- and high-risk training data splits.
-The replication code does indeed run two separate random forecasts, but they are both run on the *exact same training data*, which consists of the full training data all other models are run on.
+The intention of B&S appears to have been to run two separate random forest models, one each on the low- and high-risk training data splits.
+The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consists of the full training data from all other models.
 The models are also identical otherwise, i.e.
@@
 
-The implementation error aside, this split-population analog model is actually quite odd and does not actually replicate an analogue of the idea behind split-population modeling.
+The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling.
 Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision trees.
-Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does indeed have 200,000 decision trees.
-Furthermore, the PITF model predictions do not impact the way the combined RF model predicts at all, not even through a binary low-/high-risk split.
+Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does have 200,000 decision trees.
+Furthermore, the PITF model predictions do not impact how the combined RF model predicts at all, not even through a binary low-/high-risk split.
 The split-population PITF RF model is practically speaking just another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree).
@@
 
-Further, the AUC-ROC values reported in the original B&S Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}).
+Further, the AUC-ROC values reported in the original B&S Table 2 are calculated based on slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}).
 ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available.
@@
 
-Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
-In other words, the results for different models in B&S Table 2 are actually not comparable to one another, and any conclusions drawn from such comparison are potentially incorrect.
+Thus a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
+In other words, the results for different models in B&S Table 2 are not comparable to one another, and any conclusions drawn from such comparison are potentially incorrect.
 
@@
 
-B&S show a confusion matrix to score their 2016-H1 forecasts in Table 4.
-Although the forecasts are for the probability of civil war onset, in the replication code they are actually scored using the much more common incidence of civil war, i.e.
+B&S present a confusion matrix to score their 2016-H1 forecasts in Table 4.
+Although the forecasts are for the probability of civil war onset, in the replication code, they are actually scored using the much more commonly used incidence of civil war, i.e.
 including ongoing civil wars as "1"'s.
@@
 By examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
-In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable back to incidence.
+In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable to incidence.
 
@@
 ![Replication of B&S Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)
-
-The specific predictions here also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines as seen in the panels on the right.
-In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for a observed positive case is reached.
-For these ROC curves, and for that matter in the basic train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set.
+The specific predictions also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines seen in the panels on the right.
+In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for an observed positive case is reached.
+For these ROC curves, and for that matter in the fundamental train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set.
 Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.
-Notice also that the smoothing averages the left most almost straight line with the right most almost straight line in a monotonic way.
+Notice also that the smoothing averages the left-most almost straight line with the right-most almost straight line in a monotonic way.
 
@@
 
-In sum, it is not only the case that using smoothed ROC curves alters the results: the use of smoothed ROC curves to calculate AUC-ROC values clearly benefits \textbf{only} the escalation model.
+In sum, it is not only the case that using smoothed ROC curves alters the results: the use of smoothed ROC curves to calculate AUC-ROC values benefits \textbf{only} the escalation model.
 It does so consistently and by a considerable margin.
-All eight alternative models reported in B&S Tables 1 and 2 on average do not gain when using smoothed ROC curves to calculate AUC.
+All eight alternative models reported in B&S Tables 1 and 2, on average, do not gain when using smoothed ROC curves to calculate AUC.
 
@@
 The Quad model appears to be roughly on par with the Escalation model.
-Thus, the original B&S conclusion that the escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.
+Thus, the original B&S conclusion that the escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves and overturns when using traditional AUC-ROC calculations.
 
@@
 We found that the escalation model's AUC-ROC values generally fluctuate no more than 0.01 (see https://github.com/rickmorgan2/Blair-Sambanis-replication/blob/master/rep_nosmooth/variance.md), and thus are confident that the patterns we see are robust to the initial RNG state.] B&S's original results and interpretation regarding the superiority of the B&S model over alternative ICEWS models, including the 1,000+ covariate CAMEO model, are thus entirely conditional on the use of smoothed AUC-ROC.
-Even the Quad model is generally as good as the escalation model in all implementations, and is frequently quite a bit better.
+Even the Quad model is generally as good as the escalation model in all implementations and is frequently quite a bit better.
 
@@
 
-Generally, these comparisions require the same cases.
-Therefore, we use predictions for common joint subset of nonmissing predictions The resulting numbers of cases for each model are shown in the bottom portion of the table.
-Since the sets of incomplete cases in the original version above do not completely overlap themselves, the common subset for all models is actually slightly smaller than the smallest N in the top portion of the table.
+Generally, these comparisons require the same cases.
+Therefore, we use predictions for a common joint subset of non-missing predictions.
+The resulting numbers of cases for each model are shown in the bottom portion of the table.
+Since the sets of incomplete cases in the original version above do not entirely overlap themselves, the common subset for all models is slightly smaller than the smallest N in the top portion of the table.
 
@@
 
-Table \ref{tab:table2-fixed} shows our replication of B&S Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common, joint subset of tests cases that all models have non-missing predictions for.
+Table \ref{tab:table2-fixed} shows our replication of B&S Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common, joint subset of tests cases for which all models have non-missing predictions.
 Table \ref{tab:table2-full} in the appendix shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.
@@
 
-The most straightforward method of incorporating the annual, structural PITF variables--adding them to the predictors of the Escalation RF model--strictly outperforms the Escalation Only model.
-Note that the two other combination models considered are both non-standard and that the "PITF Split Population" model does not in fact actually incorporate structural information at all.
+The most straightforward method of incorporating the annual structural PITF variables--adding them to the predictors of the Escalation RF model--strictly outperforms the Escalation Only model.
+Note that the two other combination models considered are both non-standard and that the "PITF Split Population" model does not incorporate structural information at all.
 We thus conclude that adding structural variables improves predictive performance.
@@
 
-B&S report confusion matrices for their forecasts for the first half of 2016 (2016-H1) in their Table 4.
+B&S report confusion matrices for their forecasts for the first half of 2016 (2016-H1) in Table 4.
 To create the confusion matrices, B&S treat the 30 highest ranked predictions as positive predictions ("1s") and the rest as negative predictions ("0s").
@@
 In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a much lower positive rate of around 0.5%.
-Since the positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data, this suggests that the forecasts were erroneously assessed using civil war incidence, not onset.
-As we showed above, by examining the replication code we were able to verify that the forecasts were scored using civil war incidence, i.e.
-including ongoing civil wars, rather than civil war onset years only.
+The positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data.
+This suggests that the forecasts were erroneously assessed using civil war incidence, not onset.
+By examining the replication code we were able to verify that the forecasts were scored using civil war incidence, i.e., including ongoing civil wars, rather than civil war onset years only.
 
@@
 Thus, the recall values is undefined, while the precision is 0/30 = 0, compared to reported recall and precision values of 13/15 = 0.87 and 13/30 = 0.43.
-The alternative coding ("Assuming Change") produces 2 civil war onsets.
-Recall is 1 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.
+The alternative coding ("Assuming Change") produces two civil war onsets.
+Recall is 1.0 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.
 
 Another, minor issue or rather coding error, is related to using a lead version of the DV.
-With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 actually indicates the risk of civil war onset in 2016-H2.
+With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 indicates the risk of civil war onset in 2016-H2.
 In the Table 4 script referenced above, the 2016-H1 predictions (for 2016-H2) are assessed using the raw DV, "incidence_civil_ns", not the lead version.
 Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1.
-In this case it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.
+In this case, it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.
 
@@
 
-The B&S approach that is advocated is to use theory to guide prediction.
+The B&S advocate the use theory to guide prediction.
 But theory as used by B&S is an ambiguous and undefined concept.
 It is not a procedure.
-What they actually do is to create a model with ten right hand side variables that is supposed to capture a complicated repression-dissent dynamic.
-There is a wide-ranging literature on this dynamic that they do not rely on to construct their model.
-As such their baseline comparison is a poor standard bearer for strong theory.
+They actually create a model with ten right hand side variables that are supposed to capture a complicated repression-dissent dynamic.
+There is wide-ranging literature on this dynamic that could justify many specifications.
+As such, their baseline comparison is an unfortunate standard bearer for strong theory.
+
 Moreover, they misunderstand the use of ICEWS event data in current research.
-They claim that most uses to date have focused on the quad categories, but this ignores a wide swath of literature (Steinert-Threlkeld APSR 2017) and Metternich et al (AJPS 2013) that uses a specific action---such as protest---defined in the CAMEO ontology.
-In the AJPS article on Thailand, Metternich et al hand coded, for example, every actor in Thailand and focused on an analysis of how those interact.
+They claim that most applications to date have focused on the quad categories, but this ignores a wide swath of literature (@steinert-threlkeld:2017 and @metternich:etal:2013) that uses a specific action---such as protest---defined in the CAMEO ontology.
+In their article on conflict in Thailand, Metternich et alia hand-coded, for example, every actor in Thailand and focused on an analysis of how those interact.
 
 We encountered several issues in the code underlying the B&S analysis.
-The issues we encountered are not subjective modeling choices.
-When we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturn.
+The problems we encountered are not subjective modeling choices.
+When we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturned.
 In other words, B&S findings are based on a faulty analysis, and invalid.
 
-Using the same analysis B&S intend to use, we in fact find that:
-- the theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all CAMEO model and the Average ensemble model
+Using the same analysis B&S intend to use, we, in fact, find that:
+- the theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model. The Average ensemble model
 - structural variables substantially improve the escalation model's performance when added to the pool of predictors on which the underlying random forest model is based.
@@
 Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain.
-As social science becomes more adept at predictive analysis this will doubtless be of increasing importance.
-However, these comparisons must be done carefully to ensure that correct inferences are drawn.
+As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance.
+However, these comparisons must be made carefully to ensure those correct inferences are drawn.
 We continue to think that theory is overrated (@ward:2016), and that machine learning and big data will allow us to learn new things, but it will be interesting to see how the evidence for these claims are adjudicated with additional usage and careful evaluation.
@@
 In the B&S article it serves as a strawman.
-Unfortunately for B&S that strawman outperforms there putatively theoretical model.
-Important statistical developments have also added to this research vector, including especially
+Unfortunately for B&S that strawman outperforms their putatively theoretical model.^[Important statistical developments have also added to this research vector, including especially
 @gneiting:raftery:2005, 
 @montgomery:etal:PS:2012, and
-@mhw:2012.
+@mhw:2012.]
 

### 2020-06-09 paper.Rmd
0.95
took my crap at the end out

@@
 
-There is a fairly long tradition of making predictions in academic and policy work on conflict.
-Most of these are not mentioned by B&S.
-We would include work from the last century including @choucri:1974, @andriole:young:1977, @choucri:robinson:1979, @freeman:job:1979,@hopple:andriole:1984, @schrodt:1991, and others.
-The current century has even more work that can't be overlooked or combined into a theory barren classification
-@obrien:2002,
-@buhaug:gleditsch:2008, 
-@bennett:stam:2009,
-@rost:etal:2009, 
-@obrien:2010, 
-@weidmann:ward:2010, 
-@weidmann:ward:2010, 
-@ward:etal:2010,
-@goldstone:etal:2010, 
-@bdm:2011, 
-@brandt:etal:2011, 
-@hegre:etal:2011, 
-@Brandt_etal_2011, 
-@oloughlin:etal:2012, 
-@hegre:etal:2013, @shellman:etal:2013, 
-@ward:etal:2013, 
-@brandt:etal:2014, 
-@buhaug:etal:2014, 
-@ramakrishnan:etal:2014, 
-@pilster:bohmelt:2014, 
-@chadefaux:2017, 
-@chiba:etal:2015,
-@mellers:etal:2015, 
-@tetlock:gardner:2015, 
-@hegre:etal:2016, 
-@schleussner:etal2016, @ward:2016, 
-@beger:ward:2017, 
-@chiba:gleditsch:2017,
-@colaresi:mahmood:2017, 
-@hegre:etal:2017, 
-@tetlock:etal:2017, 
-@witmer:etal:2017, 
-@kupilik:witmer:2018, 
-@hegre:etal:2019, 
-@hegre:etal:2019.
-Indeed this scholarship has brought us to the point where predictive analysis of conflict studies is not unknown in the scholarly and policy arenas.
-Most, if not all of these authors, would probably claim some theoretical motivation.
-The a-theoretical case is a counterfactual that either includes everything, or nothing, as aruged in @ward:2016.
-In the B&S article it serves as a strawman.
-Unfortunately for B&S that strawman outperforms their putatively theoretical model.^[Important statistical developments have also added to this research vector, including especially
-@gneiting:raftery:2005, 
-@montgomery:etal:PS:2012, and
-@mhw:2012.]
-
 # References

### 2020-06-10 paper.Rmd
add github repo link to first page

@@
 date: "`r format(Sys.time(), '%d %B %Y')`"
-thanks: Cassy L. Dorff and Shahryar Minhas both provided helpful feedback on this project. All the code and several additional analyses can be found at our replication archive.
+thanks: Cassy L.
+Dorff and Shahryar Minhas both provided helpful feedback on this project.
+All the code and several additional analyses can be found at our replication archive at https://github.com/rickmorgan2/Blair-Sambanis-replication.
 abstract: We examine the research protocols in Blair and Sambanis (2020).
@@
     keep_tex: true
+    keep_md: true
     toc: false
@@
 bibliography: 
-  - "`r user <- Sys.info()[['user']]; if (user=='andybega')  '/Users/andybega/Work/whistle/master.bib' else if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else 'references.bib'`"
+  - "`r user <- Sys.info()[['user']]; if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else 'references.bib'`"
 ---

### 2020-06-10 paper.Rmd
add missing refs

@@
 Moreover, they misunderstand the use of ICEWS event data in current research.
-They claim that most applications to date have focused on the quad categories, but this ignores a wide swath of literature (@steinert-threlkeld:2017 and @metternich:etal:2013) that uses a specific action---such as protest---defined in the CAMEO ontology.
+They claim that most applications to date have focused on the quad categories, but this ignores a wide swath of literature [@steinert-threlkeld:2017; and @metternich:etal:2013] that uses a specific action---such as protest---defined in the CAMEO ontology.
 In their article on conflict in Thailand, Metternich et alia hand-coded, for example, every actor in Thailand and focused on an analysis of how those interact.
@@
 However, these comparisons must be made carefully to ensure those correct inferences are drawn.
-We continue to think that theory is overrated (@ward:2016), and that machine learning and big data will allow us to learn new things, but it will be interesting to see how the evidence for these claims are adjudicated with additional usage and careful evaluation.
+We continue to think that theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things, but it will be interesting to see how the evidence for these claims are adjudicated with additional usage and careful evaluation.
 

### 2020-06-10 make-references-bib.R

@@ -2,14 +2,12 @@
 #   Create the local references.bib file from the private central bib file
 #
 
-library(stringr)
-library(RefManageR)
-
 setwd(here::here("./"))
 
-source("https://raw.githubusercontent.com/andybega/condensebib/master/R/reduce-bib.R")
+remotes::install_github("andybega/condensebib")
+library(condensebib)
 
-reduce_bib(
+res <- reduce_bib(
   file = "paper.Rmd",
   master_bib = "../../whistle/master.bib",
   out_bib    = "references.bib"
@@ -17,5 +15,15 @@ reduce_bib(
 
 # Change \aa to \r{a}
 bib <- readLines("references.bib")
-bib <- str_replace_all(bib, "\\\\aa", "\\\\r\\{a\\}")
+bib <- stringr::str_replace_all(bib, "\\\\aa", "\\\\r\\{a\\}")
 writeLines(bib, "references.bib")
+
+
+# If some references are not showing, check if RefManageR complains about them
+if (FALSE) {
+  msgs <- capture.output(bib <- read_bib("../../whistle/master.bib", quiet = FALSE),
+                         type = "message")
+}
+
+
+

### 2020-06-10 paper.Rmd
work on adding ROC smoothing discussion

@@
  - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com. corresponding author]
- - Richard K. Morgan^[VDEM, rick.morgan2@gmail.com.]
+ - Richard K. Morgan^[V-Dem, rick.morgan2@gmail.com.]
  - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com]
@@
 Before proceeding, we quote B&S (page 24):
-\begin{quote}
-Our theoretically driven model generates accurate forecasts, with base specification
-AUCs of 0.82 and 0.85 over one- and six-month windows, respectively, and AUCs
-as high as 0.92 in other specifications. Our model also consistently and sometimes
-dramatically outperforms the alternatives we test... Cederman and Weidmann (2017, 476) argue that "the hope that
-big data will somehow yield valid forecasts through theory-free 'brute force' is
-misplaced in the area of political violence." Our results lend some credence to this
-claim.
-\end{quote}
+
+> Our theoretically driven model generates accurate forecasts, with base specification AUCs of 0.82 and 0.85 over one- and six-month windows, respectively, and AUCs as high as 0.92 in other specifications.
+Our model also consistently and sometimes dramatically outperforms the alternatives we test...
+Cederman and Weidmann (2017, 476) argue that "the hope that big data will somehow yield valid forecasts through theory-free 'brute force' is misplaced in the area of political violence." Our results lend some credence to this claim.
 
@@
 
-The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, actual ROC curves.
+The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves.
 The data in rare events problems like this one restricts the number of distinct true positive rate values.
 As we show below, the use of  smoothed ROC curves to calculate AUC has a dramatic impact on the results.
-To our knowledge, the norm is to calculate AUC-ROC values using original, not smoothed ROC curves.
-In fact, we are not aware of other work that uses smoothed ROC curves for AUC calculations.
-It might be that theoretical reasons are justifying the use of smoothed ROC curves over the original ROC curves; but given that this decision dramatically impacts the interpretation of the B&S results, it minimally would have warranted an explicit discussion in the paper.
-As it is, B&S offer no justification for using smoothed ROC curves in the AUC-ROC calculations.
 
-Smoothing of ROC curves is well known.
-Basically, the smoothed curves transform the false positive rate and the false negative rates jointly in a monotone fashion.
-The assumption is that there is a monotone transformation that produces normal distributions for the sensitivity and specificity.
-It is well-known that there is substantial bias that can be introduced by bias.
-Indeed, bootstrapping is often used as a diagnostic to see the amount of bias introduced by smoothing.
+Although smoothing might appear to be a simple option to aid in the visual interpretation of ROC curves, and is presented as such by B&S, it actually involves estimating an underlying population ROC curve, using a parametric binormal model in this case.
+There is a literature in statistics and the medical field on such ROC estimators and the conditions under which they might be preferable to ...
+
+The option to smooth the ROC curves is presented by B&S as a trivial choice to aid visual interpretation, but it actually delves into a broader area of parametric or non-parametric population ROC curve estimation.
+A common motivating scenario is in the area of medical testing, where test results might only be binary or ordinal, not continuous, where only a small number of tests with verified accuracy might be available, and where there is a clear larger population one wants to infer to.
+The particular application here deviates in several ways from the conditions under which the properties of these estimators have been studies: rare events, our data is more of a census than independent samples from a much larger population, and the use of non-deterministic random forest models.
+
+Far from being a trivial choice, the question of whether to use ROC estimates rather than the empirical ROC curves is a major decision that dramatically impacts the results.
+B&S do not offer a justification for departing from the common practice of using the empirical ROC curves.
 
@@
 What impact did the ROC smoothing have overall on the performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B&S Table 2)?
-Figure \ref{fig:benefit} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values.
+Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values.
 Each of the gray points corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed}.

### 2020-06-10 paper.Rmd
copy edits
I'm sure I missed some things in this reading. I'll pick it up again tomorrow

@@
 
-B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars ``remarkable accurately.'' Indeed they argue that this so-called theoretical model outperforms four other ``more mechanical'' alternatives.
+B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkable accurately." Indeed they argue that this so-called theoretical model outperforms four other "more mechanical" alternatives.
 Second, they claim that the integration of structure with process is better over short forecasting windows.
@@
 
-Blair and Sambanis (2020) (B\&S hereafter) aim to examine whether 
-theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
+Blair and Sambanis (2020) (B\&S hereafter) aim to examine whether theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
 For the moment, we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models.
@@
 
-The corresponding results for each question are shown in B&S Tables 1 and 2, which we will examine further below. We accurately replicate their Tables 1 and 2, with very tiny differences. 
+The corresponding results for each question are shown in B&S Tables 1 and 2, which we examine further below.
+We accurately replicate their Tables 1 and 2, with very tiny differences.
 The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions.
@@
 Although smoothing might appear to be a simple option to aid in the visual interpretation of ROC curves, and is presented as such by B&S, it actually involves estimating an underlying population ROC curve, using a parametric binormal model in this case.
-There is a literature in statistics and the medical field on such ROC estimators and the conditions under which they might be preferable to ...
+There is a literature in statistics and the medical field on such ROC estimators and the conditions under which they might be preferable to empirical ROC curves.
+For example, when the goal is to extrapolate from a statistical sample to a larger population of potential cases...
+
+(RM: IDK. But I think it can be short and simple. Highlight that smoothing in unconventional in conflict/political forecasting)
 
@@
 
-Far from being a trivial choice, the question of whether to use ROC estimates rather than the empirical ROC curves is a major decision that dramatically impacts the results.
+Far from being a trivial choice, the question of whether to use smoothed ROC estimates rather than the empirical ROC curves is a major decision that dramatically impacts the results.
 B&S do not offer a justification for departing from the common practice of using the empirical ROC curves.
@@
 > The final approach is a random forest analog to split-population modeling.
-first compute the average PITF predicted probability for each country across all years in our training set.
+We first compute the average PITF predicted probability for each country across all years in our training set.
 We define those that fall in the bottom quartile as "low risk" and the rest as "high risk." We then run our escalation model on the high-risk and low-risk subsets separately, combining the results into a single random forest (column 4).
 
-The intention of B&S appears to have been to run two separate random forest models, one each on the low- and high-risk training data splits.
+This description suggests that B&S intended to run two separate random forest models, one each on the low- and high-risk training data splits.
 The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consists of the full training data from all other models.
-The models are also identical otherwise, i.e.
-they use the same *x* variables and the same random forest hyper-parameter settings.
+In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.
+
+The model specifications are also identical; i.e., they use the same *x* variables and the same random forest hyper-parameter settings.
 The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself.
-If we ran both with the same random seed, they would be identical in every respect, producing identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined.
+If we ran both with the same random seed, they would be identical in every respect, producing to forests of identical decision trees and, thus, identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined.
 We have documented details at https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5.]
@@
 The split-population PITF RF model is practically speaking just another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree).
-This does not adequately implement their research strategy.
+This does not adequately implement their split-sample modeling strategy.
 
@@
 
-By examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
+However, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
 In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable to incidence.
@@
 
-This implies that the ROC curves were only smoothed in the referenced Figure 1, but actually all AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation.
-ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result appear step-like.
-Figure 1 shows our replication of both the smoothed ROC curves B&S report, and the actual ROC curves on the right.
-The standard method is to compute the area under the curve AUC statistic on the original, non-smooth ROC curves that are shown on the right.
+This implies that the ROC curves were only smoothed in the referenced Figure 1.
+However, this is not the case.
+All AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation.
+ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like.
+Figure 1 shows our replication of both the estimated smoothed ROC curves from the B&S report (left-hand side) and the actual empirical ROC curves (right-hand side).
+The standard method is to compute the area under the curve (AUC) statistic on the original, empirical ROC curves that are shown on the right.
 
@@
 
-In sum, it is not only the case that using smoothed ROC curves alters the results: the use of smoothed ROC curves to calculate AUC-ROC values benefits \textbf{only} the escalation model.
+In sum, it is not only the case that using smoothed ROC curves alters the results, but also that the use of smoothed ROC curves to calculate AUC-ROC values benefits \textbf{only} the escalation model.
 It does so consistently and by a considerable margin.
@@
 The Quad model appears to be roughly on par with the Escalation model.
-Thus, the original B&S conclusion that the escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves and overturns when using traditional AUC-ROC calculations.
+Thus, the original B&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves and overturns when using traditional AUC-ROC calculations.
 
@@
 
-1. B&S: ``Of the approaches we test, the split-population analog is most promising \ldots ''
+1. B&S: "Of the approaches we test, the split-population analog is most promising \ldots "
 
@@
 
-2. B&S: ``Adding PITF predictors improves the performance of the escalation model over six-month windows but diminishes it over one-month windows \ldots ''
+2. B&S: "Adding PITF predictors improves the performance of the escalation model over six-month windows but diminishes it over one-month windows \ldots "
 
-Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model, and by quite a maring in the 6-month data version.
+Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model, and by quite a margin in the 6-month data version.
 
-3. B&S: ``The weighted model performs very poorly regardless \ldots ''
+3. B&S: "The weighted model performs very poorly regardless \ldots "
 
@@
 4.
-B&S: ``Overall, our results suggest that while measures of structural risk may improve predictive performance, the value they add is marginal and inconsistent.
+B&S: "Overall, our results suggest that while measures of structural risk may improve predictive performance, the value they add is marginal and inconsistent.
 [...] Incorporating PITF thus significantly reduces or only slightly improves the performance of the escalation model, regardless of the approach we take.
-\ldots ''
+\ldots "
 
@@
 tab4f <- tab4f %>%
-  rename(Predicted0 = `0`, Predicted1 = `1`) %>%
+  rename(`Predicted 0` = `0`, `Predicted 1` = `1`) %>%
   mutate(source = "Fixed, scored with civil war onset")
@@
 
-The B&S advocate the use theory to guide prediction.
+B&S advocate the use theory to guide prediction.
 But theory as used by B&S is an ambiguous and undefined concept.

### 2020-06-10 paper.Rmd
fixed table 4
I messed up table 4 in my last commit... my bad

@@
 
-The most straightforward method of incorporating the annual structural PITF variables--adding them to the predictors of the Escalation RF model--strictly outperforms the Escalation Only model.
+The most straightforward method of incorporating the annual structural PITF variables---adding them to the predictors of the Escalation RF model---strictly outperforms the Escalation Only model.
 Note that the two other combination models considered are both non-standard and that the "PITF Split Population" model does not incorporate structural information at all.
@@
 tab4f <- tab4f %>%
-  rename(`Predicted 0` = `0`, `Predicted 1` = `1`) %>%
+  rename(Predicted0 = `0`, Predicted1 = `1`) %>%
   mutate(source = "Fixed, scored with civil war onset")

### 2020-06-10 paper.Rmd
rewrite smoothing section

--- 2020-06-10 paper.Rmd
@@
 The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves.
-The data in rare events problems like this one restricts the number of distinct true positive rate values.
-As we show below, the use of  smoothed ROC curves to calculate AUC has a dramatic impact on the results.
+The data in rare events problems like this one restricts the number of true positive rate values and leads to very non-smooth ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves as easing interpretation.
+But in fact smoothing is used to produce all AUC-ROC values they report.
+Standard practice has been to use empirical, not smoothed ROC curves, both for visualization and when calculating AUCs.
 
-Although smoothing might appear to be a simple option to aid in the visual interpretation of ROC curves, and is presented as such by B&S, it actually involves estimating an underlying population ROC curve, using a parametric binormal model in this case.
-There is a literature in statistics and the medical field on such ROC estimators and the conditions under which they might be preferable to ...
-
-The option to smooth the ROC curves is presented by B&S as a trivial choice to aid visual interpretation, but it actually delves into a broader area of parametric or non-parametric population ROC curve estimation.
-A common motivating scenario is in the area of medical testing, where test results might only be binary or ordinal, not continuous, where only a small number of tests with verified accuracy might be available, and where there is a clear larger population one wants to infer to.
-The particular application here deviates in several ways from the conditions under which the properties of these estimators have been studies: rare events, our data is more of a census than independent samples from a much larger population, and the use of non-deterministic random forest models.
-
-Far from being a trivial choice, the question of whether to use ROC estimates rather than the empirical ROC curves is a major decision that dramatically impacts the results.
-B&S do not offer a justification for departing from the common practice of using the empirical ROC curves.
+Aside from the visual impact of ROC smoothing, the underlying motivation and methods involved in smoothing are much broader and involve parametric estimation of population ROC curves [@henley:1988; @henley:2014].
+The main application is in the evaluation of medical tests, e.g.
+diagnostic imaging, with early methods dating back several decades and ongoing development of both parametric and non-parametric smoothing methods [e.g.
+@zou:etal:1997; @pulit:2016]; it does not seem to be established which methods should be preferable in a given application [@henley:2014].
+It is also not clear whether these methods and rationales are valid in applications like our present one, e.g.
+where our data is closer to a census of the population and with well-known dependencies like spatial and temporal correlation, rather than a random sample that is approximately i.i.d.
 

### 2020-06-10 paper.Rmd
double-spacing, in case we need it in future

@@
     toc: false
-# spacing: double
+#header-includes:
+#  - \usepackage{setspace}\doublespacing
 bibliography: 

### 2020-06-10 paper.Rmd

--- 2020-06-10 paper.Rmd
@@
 The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves.
-The data in rare events problems like this one restricts the number of distinct true positive rate values.
-As we show below, the use of smoothed ROC curves to calculate AUC has a dramatic impact on the results.
+The data in rare events problems like this one restricts the number of true positive rate values and leads to very non-smooth ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves as easing interpretation.
+But in fact smoothing is used to produce all AUC-ROC values they report.
+Standard practice in conflict research and forecasting has been to use empirical, not smoothed ROC curves, both for visualization and when calculating AUCs.
 
-Although smoothing might appear to be a simple option to aid in the visual interpretation of ROC curves, and is presented as such by B&S, it actually involves estimating an underlying population ROC curve, using a parametric binormal model in this case.
-There is a literature in statistics and the medical field on such ROC estimators and the conditions under which they might be preferable to empirical ROC curves.
-For example, when the goal is to extrapolate from a statistical sample to a larger population of potential cases...
-
-(RM: IDK. But I think it can be short and simple. Highlight that smoothing in unconventional in conflict/political forecasting)
-
-The option to smooth the ROC curves is presented by B&S as a trivial choice to aid visual interpretation, but it actually delves into a broader area of parametric or non-parametric population ROC curve estimation.
-A common motivating scenario is in the area of medical testing, where test results might only be binary or ordinal, not continuous, where only a small number of tests with verified accuracy might be available, and where there is a clear larger population one wants to infer to.
-The particular application here deviates in several ways from the conditions under which the properties of these estimators have been studies: rare events, our data is more of a census than independent samples from a much larger population, and the use of non-deterministic random forest models.
-
-Far from being a trivial choice, the question of whether to use smoothed ROC estimates rather than the empirical ROC curves is a major decision that dramatically impacts the results.
-B&S do not offer a justification for departing from the common practice of using the empirical ROC curves.
+Aside from the visual impact of ROC smoothing, the underlying motivation and methods involved in smoothing are broader and involve parametric estimation of population ROC curves [@henley:1988; @henley:2014].
+The main application is in the evaluation of medical tests, e.g.
+diagnostic imaging, with early methods dating back several decades and ongoing development of both parametric and non-parametric smoothing methods [e.g.
+@zou:etal:1997; @pulit:2016]; it is not established which methods should be preferable in a given application [@henley:2014].
+Notably, the context in which ROC smoothing is used differs in important aspects from typical conflict research applications, e.g.
+our data is closer to a census of the population and with well-known dependencies like spatial and temporal correlation, rather than a random sample that is approximately i.i.d.
+It is thus not clear that smoothing is justified nor valid, and we should also note that it does not seem to be widely used outside of the narrow context mentioned above.
 

### 2020-06-10 paper.Rmd
let's call this version 1.0
I may tweak it further. I had to fix the bibliography which was rough.

@@
 bibliography: 
-  - "`r user <- Sys.info()[['user']]; if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else 'references.bib'`"
+  - "`r user <- Sys.info()[['user']]; if (user=='MooseJaw') '/Users/mdw/Git/whistle/master.bib' else 'references.bib'`"
 ---
@@
 We find that these claims are false.
-Their theory-based escalation model does not do better than the alternatives that they examine.
-Indeed, it does worse.
+Their theory-based escalation model does not do better than the alternatives that they examine.Actually, it does worse.
 The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure.
@@
 
-B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkable accurately." Indeed they argue that this so-called theoretical model outperforms four other "more mechanical" alternatives.
+B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars remarkably accurately.
+Indeed they argue that this so-called theoretical model outperforms four other``more mechanical'' alternatives.
 Second, they claim that the integration of structure with process is better over short forecasting windows.
@@
 
-Blair and Sambanis (2020) (B\&S hereafter) aim to examine whether theory adds more value to a forecasting model when compared to non-parametric machine learning models, and specifically random forests, whose specifications are not theory-informed.
+Blair and Sambanis (2020) (B\&S hereafter) machine learning models, and specifically random forests, whose specifications are not theory-informed.
 For the moment, we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models.
@@
 The models are:
-- Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
-- Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
-- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government on one side and opposition or rebel actors on the other.
+\begin{itemize}
+\item Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
+\item Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
+\item Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government on one side and opposition or rebel actors on the other.
 These are directed, thus making for four total covariates.
-- CAMEO: counts for all CAMEO event codes totaling $1,159$ covariates, which are mostly zero for any country in any month. 
-- Average: unweighted average of the predictions from the four models briefly described above.
+\item CAMEO: counts for all CAMEO event codes totaling $1,159$ covariates, which are mostly zero for any country in any month. 
+\item Average: unweighted average of the predictions from the four models briefly described above.
+\end{itemize}
 
@@
 our data is closer to a census of the population and with well-known dependencies like spatial and temporal correlation, rather than a random sample that is approximately i.i.d.
-It is thus not clear that smoothing is justified nor valid, and we should also note that it does not seem to be widely used outside of the narrow context mentioned above.
+It is thus not clear that smoothing is justified nor valid.
+We also note that smoothed ROC plots do
+are not widely used outside of the narrow context mentioned above.
 
@@
 B&S present a confusion matrix to score their 2016-H1 forecasts in Table 4.
-Although the forecasts are for the probability of civil war onset, in the replication code, they are actually scored using the much more commonly used incidence of civil war, i.e.
-including ongoing civil wars as "1"'s.
+Although the forecasts are for the probability of civil war onset, in the replication code, they are actually scored using incidence of civil war.
 
@@
 Each of the gray points corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed}.
-The red crosses mark the average effect of smoothing on AUC-ROC.
+The vertical bar in each plot marks the average effect of smoothing on AUC-ROC.
 For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible in average.
@@
 
-B&S advocate the use theory to guide prediction.
-But theory as used by B&S is an ambiguous and undefined concept.
+B&S advocate the use of theory to guide prediction.
+But ``theory'' as used by B&S is an ambiguous and undefined concept.
 It is not a procedure.
-They actually create a model with ten right hand side variables that are supposed to capture a complicated repression-dissent dynamic.
+They actually create a model with ten right-hand side variables that are supposed to capture a complicated repression-dissent dynamic.
 There is wide-ranging literature on this dynamic that could justify many specifications.
@@
 
-Using the same analysis B&S intend to use, we, in fact, find that:
-- the theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model. The Average ensemble model
-- structural variables substantially improve the escalation model's performance when added to the pool of predictors on which the underlying random forest model is based.
+In contradistinction to the conclusions offered by B&S, we find that:
+\begin{itemize}
+\item The theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model.
+\item The Average ensemble model and the CAMEO models outperform the escalation model in all instances.
+\item Adding structural variables substantially improves the escalation model's performance.
+\end{itemize}
 
@@
 However, these comparisons must be made carefully to ensure those correct inferences are drawn.
-We continue to think that theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things, but it will be interesting to see how the evidence for these claims are adjudicated with additional usage and careful evaluation.
+We continue to think that theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things.
+It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.
+
+\newpage
 

### 2020-06-11 paper.Rmd
fix incorrect linetype in figure 1

@@
 #  - \usepackage{setspace}\doublespacing
-bibliography: 
-  - "`r user <- Sys.info()[['user']]; if (user=='MooseJaw') '/Users/mdw/Git/whistle/master.bib' else 'references.bib'`"
+bibliography: "./references.bib"
 ---
@@
 B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars remarkably accurately.
-Indeed they argue that this so-called theoretical model outperforms four other``more mechanical'' alternatives.
+Indeed they argue that this so-called theoretical model outperforms four other "more mechanical" alternatives.
 Second, they claim that the integration of structure with process is better over short forecasting windows.
@@
 The models are:
+
 \begin{itemize}
@@
 ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like.
-Figure 1 shows our replication of both the estimated smoothed ROC curves from the B&S report (left-hand side) and the actual empirical ROC curves (right-hand side).
+Figure \ref{fig:rocs} shows our replication of both the estimated smoothed ROC curves from the B&S report (left-hand side) and the actual empirical ROC curves (right-hand side).
 The standard method is to compute the area under the curve (AUC) statistic on the original, empirical ROC curves that are shown on the right.
 
-![Replication of B&S Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)
+\begin{figure}
+\caption{Replication of B\&S Figure 1 with both smooth and non-smoothed ROC curves\label{fig:rocs}}
+\centering
+\includegraphics[width=.8\linewidth]{figures/figure1-replicated.png}
+\end{figure}
+
 The specific predictions also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines seen in the panels on the right.
@@
 
-1. B&S: "Of the approaches we test, the split-population analog is most promising \ldots "
+1. B&S: "Of the approaches we test, the split-population analog is most promising ..."
 
@@
 B&S advocate the use of theory to guide prediction.
-But ``theory'' as used by B&S is an ambiguous and undefined concept.
+But "theory" as used by B&S is an ambiguous and undefined concept.
 It is not a procedure.
@@
 In contradistinction to the conclusions offered by B&S, we find that:
+
 \begin{itemize}

### 2020-06-11 rep_nosmooth/recreate-figure1.R

@@ -52,7 +52,7 @@ par(mfrow = c(2, 2))
 plot(roc[["1mo-smooth-escalation"]],
   ylab = "True positive rate",
   xlab = "False positive rate",
-  main = "ROC - Smoothed (1-mon)",             lty = 1)
+  main = "ROC - Smoothed (1-mon)", lty = 1)
 pROC::lines.roc(roc[["1mo-smooth-quad"]],      lty = 2)
 pROC::lines.roc(roc[["1mo-smooth-goldstein"]], lty = 3)
 pROC::lines.roc(roc[["1mo-smooth-cameo"]],     lty = 4)
@@ -73,20 +73,20 @@ plot(roc[["6mo-smooth-escalation"]],
      ylab = "True positive rate",
      xlab = "False positive rate",
      main = "ROC - Smoothed (6-mon)", lty = 1)
-pROC::lines.roc(roc[["6mo-smooth-quad"]], lty = 1)
+pROC::lines.roc(roc[["6mo-smooth-quad"]],      lty = 2)
 pROC::lines.roc(roc[["6mo-smooth-goldstein"]], lty = 3)
-pROC::lines.roc(roc[["6mo-smooth-cameo"]], lty = 4)
-pROC::lines.roc(roc[["6mo-smooth-average"]], lty = 5)
+pROC::lines.roc(roc[["6mo-smooth-cameo"]],     lty = 4)
+pROC::lines.roc(roc[["6mo-smooth-average"]],   lty = 5)
 legend("bottomright", c("Escalation", "Quad", "Goldstein", "CAMEO", "Average"), lty = c(1, 2, 3, 4, 5))
 
 plot(roc[["6mo-orig-escalation"]],
      ylab = "True positive rate",
      xlab = "False positive rate",
      main = "ROC - Not Smoothed (6-mon)", lty = 1)
-pROC::lines.roc(roc[["6mo-orig-quad"]], lty = 1)
+pROC::lines.roc(roc[["6mo-orig-quad"]],      lty = 2)
 pROC::lines.roc(roc[["6mo-orig-goldstein"]], lty = 3)
-pROC::lines.roc(roc[["6mo-orig-cameo"]], lty = 4)
-pROC::lines.roc(roc[["6mo-orig-average"]], lty = 5)
+pROC::lines.roc(roc[["6mo-orig-cameo"]],     lty = 4)
+pROC::lines.roc(roc[["6mo-orig-average"]],   lty = 5)
[…]

### 2020-06-11 paper.Rmd
add color to ICEWS models in figures 1 and 2

@@
 author: 
- - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com. corresponding author]
+ - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
  - Richard K. Morgan^[V-Dem, rick.morgan2@gmail.com.]
- - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com]
+ - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
 # Introduction
-@blair:sambanis:2020 (hereafter B\&S) argue that theory is essential for creating models that have high accuracy in forecasting civil war onset.
+Blair and Sambanis [-@blair:sambanis:2020, hereafter B\&S] argue that theory is essential for creating models that have high accuracy in forecasting civil war onset.
 Indeed they assert that with such theory, forecasting is more accurate than has previously been possible.
@@
 We find that these claims are false.
-Their theory-based escalation model does not do better than the alternatives that they examine.Actually, it does worse.
+Their theory-based escalation model does not do better than the alternatives that they examine.
+Actually, it does worse.
 The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure.
@@
 \centering
-\includegraphics[width=.8\linewidth]{figures/figure1-replicated.png}
+\includegraphics[width=.95\linewidth]{figures/figure1-replicated.png}
 \end{figure}
@@
 table2_benefit <- read_csv("data/table2-smooth-benefit.csv")
+
+cols <- c(
+  Escalation = "#e41a1c",
+  Goldstein = "#ff7f00",
+  Quad = "#984ea3",
+  CAMEO = "#377eb8",
+  Average = "#4daf4a",
+  `With PITF Predictors` = "gray50", 
+  `Weighted by PITF` = "gray50", 
+  `PITF Split Population` = "gray50",
+  `PITF Only` = "gray50"
+)
 
@@
   #facet_wrap(~ horizon, scales = "free_x") +
-  geom_point(aes(x = diff), alpha = 0.5, size = 2, color = "gray50") +
+  geom_point(aes(x = diff, color = model), alpha = 0.5, size = 2) +
+  scale_color_manual(guide = FALSE, values = cols) +
   geom_point(data = means, aes(x = avg_benefit), shape = "|", color = "black", size = 6) +
-  labs(x = "AUC(smoothed ROC) - AUC(ROC)",
+  labs(x = "AUC(smoothed ROC) - AUC(empirical ROC)",
        y = "") +

### 2020-06-11 rep_nosmooth/recreate-figure1.R

@@ -11,7 +11,7 @@ library(readr)
 library(tibble)
 
 setwd(here("rep_nosmooth"))
-dir.create("output/figures")
+dir.create("output/figures", showWarnings = FALSE)
 
 all_predictions <- read_rds("output/all-predictions.rds") %>%
   group_by(cell_id) %>%
@@ -43,8 +43,19 @@ roc <- rocs %>%
   select(list_name, object) %>%
   deframe()
 
+cols <- c(
+  esc = "#e41a1c",
+  gold = "#ff7f00",
+  quad = "#984ea3",
+  cameo = "#377eb8",
+  avg = "#4daf4a"
+)
+
+# linewidth
+ll <- 4
+
 png("output/figures/figure1-replicated.png", width = 1100, height = 1000,
-    units = "px", pointsize = 20)
+    units = "px", pointsize = 24)
 par(mfrow = c(2, 2))
 
 # Plot ROC curves
@@ -52,42 +63,46 @@ par(mfrow = c(2, 2))
 plot(roc[["1mo-smooth-escalation"]],
   ylab = "True positive rate",
   xlab = "False positive rate",
-  main = "ROC - Smoothed (1-mon)", lty = 1)
-pROC::lines.roc(roc[["1mo-smooth-quad"]],      lty = 2)
-pROC::lines.roc(roc[["1mo-smooth-goldstein"]], lty = 3)
-pROC::lines.roc(roc[["1mo-smooth-cameo"]],     lty = 4)
-pROC::lines.roc(roc[["1mo-smooth-average"]],   lty = 5)
-legend("bottomright", c("Escalation", "Quad", "Goldstein", "CAMEO", "Average"), lty = c(1, 2, 3, 4, 5))
+  main = "ROC - Smoothed (1-mon)",             lty = 1, col = cols["esc"], lwd = ll)
+pROC::lines.roc(roc[["1mo-smooth-quad"]],      lty = 2, col = cols["quad"], lwd = ll)
+pROC::lines.roc(roc[["1mo-smooth-goldstein"]], lty = 3, col = cols["gold"], lwd = ll)
+pROC::lines.roc(roc[["1mo-smooth-cameo"]],     lty = 4, col = cols["cameo"], lwd = ll)
+pROC::lines.roc(roc[["1mo-smooth-average"]],   lty = 5, col = cols["avg"], lwd = ll)
+legend("bottomright",
[…]

### 2020-06-11 paper.Rmd
minor edits

@@
 > Our theoretically driven model generates accurate forecasts, with base specification AUCs of 0.82 and 0.85 over one- and six-month windows, respectively, and AUCs as high as 0.92 in other specifications.
-Our model also consistently and sometimes dramatically outperforms the alternatives we test...
-Cederman and Weidmann (2017, 476) argue that "the hope that big data will somehow yield valid forecasts through theory-free 'brute force' is misplaced in the area of political violence." Our results lend some credence to this claim.
+Our model also consistently and sometimes dramatically outperforms the alternatives we test.
+[...] Cederman and Weidmann (2017, 476) argue that "the hope that big data will somehow yield valid forecasts through theory-free 'brute force' is misplaced in the area of political violence." Our results lend some credence to this claim.
 
@@
 
-Blair and Sambanis (2020) (B\&S hereafter) machine learning models, and specifically random forests, whose specifications are not theory-informed.
+B&S' analysis is based on the use of non-linear, non-deterministic machine learning models, and specifically random forests, one of which has a specification they argues is theory-based, and several others with more generic sets of covariates.
+Notably, the analysis is at the country-month or country-(6 month) levels and relies in large part on indicators derived from the ICEWS event data.
 For the moment, we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models.
@@
 They arrive at this conclusion by examining the problem of predicting civil war onset.
-They report that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory including a \textit{kitchen sink} model with more than 1,000 covariates.
+They report that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory, including a \textit{kitchen sink} model with more than 1,000 covariates.
 
@@
 \begin{itemize}
-\item Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
-\item Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
-\item Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government on one side and opposition or rebel actors on the other.
-These are directed, thus making for four total covariates.
-\item CAMEO: counts for all CAMEO event codes totaling $1,159$ covariates, which are mostly zero for any country in any month. 
+\item Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model, for interactions between the government on one side and opposition or rebel actors on the other. 
+\item Quad: ICEWS quad counts, i.e.
+material conflict, material cooperation, verbal conflict, verbal cooperation, for interactions between the government and opposition or rebels.
+These are directed, thus making for four directed dyads, which with four quad categories make for a total of sixteen covariates.
+\item Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for the same four directed dyads, for a toal of four covariates. 
+\item CAMEO: counts for all CAMEO event codes over the four actor dyads, totaling $1,160$ covariates, which are mostly zero for any country in any month. 
 \item Average: unweighted average of the predictions from the four models briefly described above.
@@
 It is thus not clear that smoothing is justified nor valid.
-We also note that smoothed ROC plots do
-are not widely used outside of the narrow context mentioned above.
+We also note that smoothed ROC plots are not widely used outside of the narrow context mentioned above.
 
@@
 B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
-B&S, however, actually weights the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`).
+However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`).
 The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions.
@@
 
-The "PITF Split Population" model appears to be incorrectly implemented, owing to a coding error. B&S describe it on page 20:
+The "PITF Split Population" model also appears to be incorrectly implemented, owing to a coding error. B&S describe it on page 20:
 
@@
 We first compute the average PITF predicted probability for each country across all years in our training set.
-We define those that fall in the bottom quartile as "low risk" and the rest as "high risk." We then run our escalation model on the high-risk and low-risk subsets separately, combining the results into a single random forest (column 4).
+We define those that fall in the bottom quartile as "low risk" and the rest as "high risk." We then run our escalation model on the high-risk and low-risk subsets separately, combining the results into a single random forest [...].
 
@@
 Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values.
-Each of the gray points corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed}.
+Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}).
 The vertical bar in each plot marks the average effect of smoothing on AUC-ROC.
@@
 
-Both the Average and CAMEO models outperform the Escalation model in almost all instances.
+Both the Average and CAMEO models outperform the Escalation model in almost all instances (see also Figure \ref{fig:table1-plot}).
 The Goldstein model generally outperforms the Escalation model in the 6-month version.
 The Quad model appears to be roughly on par with the Escalation model.
-Thus, the original B&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves and overturns when using traditional AUC-ROC calculations.
+Thus, the original B&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves and overturns when using conventional AUC-ROC calculations.
 
@@
 [...] Incorporating PITF thus significantly reduces or only slightly improves the performance of the escalation model, regardless of the approach we take.
-\ldots "
+[...] "
 
 The most straightforward method of incorporating the annual structural PITF variables---adding them to the predictors of the Escalation RF model---strictly outperforms the Escalation Only model.
-Note that the two other combination models considered are both non-standard and that the "PITF Split Population" model does not incorporate structural information at all.
+Note that the two other combination models considered are both non-standard and that the "PITF Split Population" model does not incorporate structural information at all, yet they also both do well.
 We thus conclude that adding structural variables improves predictive performance.
@@
 This approach only works due to the choice to use regression, not classification, trees.
-Trying to use classification trees with the other parameter settings not work at all because it is almost guaranteed that a sample of 100 from the ~11,000 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample.
+Trying to use classification trees with the other parameter settings does not work at all because it is almost guaranteed that a sample of 100 from the ~11,000 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample.
 As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:

### 2020-06-11 paper.Rmd
revert lists to markdown style

@@
 
-B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars remarkably accurately.
+B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately".
 Indeed they argue that this so-called theoretical model outperforms four other "more mechanical" alternatives.
@@
 
-\begin{itemize}
-\item Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model, for interactions between the government on one side and opposition or rebel actors on the other. 
-\item Quad: ICEWS quad counts, i.e.
+- Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model, for interactions between the government on one side and opposition or rebel actors on the other. 
+- Quad: ICEWS quad counts, i.e.
 material conflict, material cooperation, verbal conflict, verbal cooperation, for interactions between the government and opposition or rebels.
 These are directed, thus making for four directed dyads, which with four quad categories make for a total of sixteen covariates.
-\item Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for the same four directed dyads, for a toal of four covariates. 
-\item CAMEO: counts for all CAMEO event codes over the four actor dyads, totaling $1,160$ covariates, which are mostly zero for any country in any month. 
-\item Average: unweighted average of the predictions from the four models briefly described above.
-\end{itemize}
+- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for the same four directed dyads, for a toal of four covariates. 
+- CAMEO: counts for all CAMEO event codes over the four actor dyads, totaling $1,160$ covariates, which are mostly zero for any country in any month. 
+- Average: unweighted average of the predictions from the four models briefly described above.
 
@@
 
-\begin{itemize}
-\item The theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model.
-\item The Average ensemble model and the CAMEO models outperform the escalation model in all instances.
-\item Adding structural variables substantially improves the escalation model's performance.
-\end{itemize}
+- The theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model.
+- The Average ensemble model and the CAMEO models outperform the escalation model in all instances.
+- Adding structural variables substantially improves the escalation model's performance.
 

### 2020-06-11 paper.Rmd
fix typo

@@
   pack_rows(index = c("One-month forecasts" = 1, "Six-month forecasts" = 1)) %>%
-  footnote("Differences from the original B\\&S Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.", footnote_as_chunk = TRUE, threeparttable = TRUE)
+  footnote("Differences from the original B&S Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.", footnote_as_chunk = TRUE, threeparttable = TRUE)
 ```

### 2020-06-12 paper.Rmd
moar final final final changes
By Finn Ally

@@
 date: "`r format(Sys.time(), '%d %B %Y')`"
-thanks: Cassy L.
-Dorff and Shahryar Minhas both provided helpful feedback on this project.
+thanks: John Ahlquist, Cassy L.
+Dorff, and Shahryar Minhas both provided helpful feedback on this project.
 All the code and several additional analyses can be found at our replication archive at https://github.com/rickmorgan2/Blair-Sambanis-replication.
@@
 Blair and Sambanis [-@blair:sambanis:2020, hereafter B\&S] argue that theory is essential for creating models that have high accuracy in forecasting civil war onset.
-Indeed they assert that with such theory, forecasting is more accurate than has previously been possible.
+Indeed they assert that with such theory forecasting is more accurate than has previously been possible.
 Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in their article.
@@
 The performance results they report are based on smoothed performance curves, not the original unsmoothed curves.
-Two of the structural alternatives to their basic escalation model were incorrectly implemented.
+Further, two of the structural alternatives to their basic escalation model were incorrectly implemented.
 We also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset.
@@
 B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately".
-Indeed they argue that this so-called theoretical model outperforms four other "more mechanical" alternatives.
+Indeed, B&S argue that this so-called theoretical model outperforms four other "more mechanical" alternatives.
 Second, they claim that the integration of structure with process is better over short forecasting windows.
@@
 B\&S claim to be unique in assessing these forecasts.
-A qualitative analysis of their predictions allows them to conclude that their model is robust, and a more precise test is undertaken.
+A qualitative analysis of their predictions allows them to conclude that their model is robust.
 We will return to their analysis later, after correcting the procedural mistakes we found in their research process.
+
+nocite: |
+  @cederman:weidmann:2017 
+...
 
@@
 material conflict, material cooperation, verbal conflict, verbal cooperation, for interactions between the government and opposition or rebels.
-These are directed, thus making for four directed dyads, which with four quad categories make for a total of sixteen covariates.
-- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for the same four directed dyads, for a toal of four covariates. 
+These are directed, thus making for four directed dyads, which with four quad categories make sixteen covariates.
+- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for the same four directed dyads, for a total of four covariates. 
 - CAMEO: counts for all CAMEO event codes over the four actor dyads, totaling $1,160$ covariates, which are mostly zero for any country in any month. 
@@
 The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions.
-These robustness checks vary either (1) random forecast hyperparameter values, or (2) the year used to split the train/test data, or (3) alternative codings of the civil war onset dependent variable.
+These robustness checks vary either (1) random forecast hyperparameter values or (2) the year used to split the train/test data, or (3) alternative codings of the civil war onset dependent variable.
 
@@
 
-- Escalation Only: the original basic escalation model with only ICEWS predictors
-- With PITF Predictors: a random forest that as predictors has the escalation model indicators but also the PITF annual, structural variables
-- Weighted by PITF: escalation model predictions weighted using the PITF instability model predictions
-- PITF Split Population: the training data are split into high and low risk portions based on the PITF instability model predictions, two separate escalation random forests are trained on the splits, then re-combined into a single random forest that is used to create the test set predictions
-- PITF Only: a random forest model based only on the annual, structural PITF model predictors
+- Escalation Only: the original basic escalation model with only ICEWS predictors.
+- With PITF Predictors: a random forest that also adds the PITF annual, structural variables.
+- Weighted by PITF: escalation model predictions weighted using the PITF instability model predictions.
+- PITF Split Population: the training data are split into high and low risk portions based on the PITF instability model predictions, two separate escalation random forests are trained on the splits, then re-combined into a single random forest that is used to create the test set predictions.
+- PITF Only: a random forest model based only on the annual, structural PITF model predictors.
 
@@
 
-# Implementation issues in B&S
+# Implementation Issues in B&S
 
 While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation.
-These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) incorrect implementations of the weighted by PITF and PITF analog split-population models, c) inconsistent test sets for the models examined, and d) incorrect scoring of the 2016 forecasts.^[We also note (e) additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox.
+These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) incorrect implementations of the weighted by PITF and PITF analog split-population models, c) inconsistent test sets for the models examined, and d) incorrect scoring of the 2016 forecasts.^[We also note there are additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox.
 We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.]
@@
 
-## Smoothed ROC curves
+## Smoothed ROC Curves
 
 The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves.
-The data in rare events problems like this one restricts the number of true positive rate values and leads to very non-smooth ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves as easing interpretation.
+The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves as easing interpretation.
 But in fact smoothing is used to produce all AUC-ROC values they report.
@@
 Aside from the visual impact of ROC smoothing, the underlying motivation and methods involved in smoothing are broader and involve parametric estimation of population ROC curves [@henley:1988; @henley:2014].
-The main application is in the evaluation of medical tests, e.g.
-diagnostic imaging, with early methods dating back several decades and ongoing development of both parametric and non-parametric smoothing methods [e.g.
-@zou:etal:1997; @pulit:2016]; it is not established which methods should be preferable in a given application [@henley:2014].
-Notably, the context in which ROC smoothing is used differs in important aspects from typical conflict research applications, e.g.
-our data is closer to a census of the population and with well-known dependencies like spatial and temporal correlation, rather than a random sample that is approximately i.i.d.
-It is thus not clear that smoothing is justified nor valid.
+The main application is in the evaluation of medical tests (e.g.
+diagnostic imaging) with early methods dating back several decades and ongoing development of both parametric and non-parametric smoothing methods [e.g.
+@zou:etal:1997; @pulit:2016].
+It is not established which methods are preferable in a given application [@henley:2014].
+Notably, the context in which ROC smoothing is used differs in important aspects from typical conflict research applications.
+Conflict data are closer to a census of the population and with well-known dependencies like spatial and temporal correlation, rather than a random sample that is approximately independently and identically distributed.
+It is thus neither clear that smoothing is justified nor valid.
 We also note that smoothed ROC plots are not widely used outside of the narrow context mentioned above.
@@
 
-## Incorrect "Weighted by PITF" implementation
+## Incorrect "Weighted by PITF" Implementation
 
@@
 
-## Incorrect "PITF Split Population" implementation
+## Incorrect "PITF Split Population" Implementation
 
@@
 
-The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling.
+The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015,beger:etal:2017].
 Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision trees.
@@
 
-The relevant variables in the data are "incidence_civil_ns" and "incidence_civil_ns_plus1", which appears to be a 1-period lead version of the DV that is used in the actual prediction models.
-The incidence DV contains both 0/1 and missing values.
+The relevant variables in the data are "incidence_civil_ns" and "incidence_civil_ns_plus1", which appears to be a 1-period lead (ie., $t_+1$) version of the dependent variable that is used in the actual prediction models.
+The incidence dependent variable  contains both 0/1 and missing values.
 By examining the pattern of missing values, it seems clear that this was originally an incidence variable indicating whether a country was at civil war in a given year or not, and which was converted to an onset version so that onsets retain the value of 1 but continuing civil war years are coded as missing.
-This reflects common practice in how these are coded.
+This reflects common practice.
 
@@
 We now turn to an examination of our analysis that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/rickmorgan2/Blair-Sambanis-replication.] The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B&S Table 2).
-We will review the substantive implications of our updated analysis further below, but the bottom line is that these changes turn B&S's conclusions on their heads.
+We will review the substantive implications of our updated analysis below, but the bottom line is that these changes turn B&S's conclusions on their heads.
 
@@
 
-A reference to smoothing is made in a single sentence in the paper (p. 12):
+A reference to smoothing is made in a single sentence in B&S (p. 12):
 
@@
 
-## The effect of using smoothed ROC curves
+## The Effect of Using Smoothed ROC Curves
 
@@
 All eight alternative models reported in B&S Tables 1 and 2, on average, do not gain when using smoothed ROC curves to calculate AUC.
+Four of the eight show a small positive bias from smoothing, while the other four display a small negative bias.
+The magnitudes of these are not close to the level of positive bias found in the escalation model.
 
@@
 The AUC-ROC values are based on original, non-smoothed ROC curves.
-
 Both the Average and CAMEO models outperform the Escalation model in almost all instances (see also Figure \ref{fig:table1-plot}).
@@
 
-## Do structural variables add to the Escalation model?
+## Do Structural Variables Add to the Escalation Model's Predictive Power?
 
@@
 
-## How accurate were the 2016 forecasts?
+## How Accurate Were the 2016 Forecasts?
 
-B&S report confusion matrices for their forecasts for the first half of 2016 (2016-H1) in Table 4.
+B&S report (Table 4)confusion matrices for their forecasts for the first half of 2016 (2016-H1).
 To create the confusion matrices, B&S treat the 30 highest ranked predictions as positive predictions ("1s") and the rest as negative predictions ("0s").
@@
 
-In contradistinction to the conclusions offered by B&S, we find that:
+In contradistinction to the conclusions offered by B&S, we find that when correctly specified and implemented:
 

### 2020-06-12 rep_nosmooth/rep-check.R
add explicit replication checks
addes tables that compare both the originanl replication and our modified replication results to the values reported in Tables 1 and 2 in the paper

see #7, #8

@@ -1,0 +1,64 @@
+#
+#   How much do our replication results differ from the values reported in the
+#   B&S paper?
+#
+
+library(readr)
+library(dplyr)
+library(tidyr)
+
+setwd(here::here("rep_nosmooth"))
+dir.create("output/tables")
+
+stopifnot(
+  "run model-runner.R first" = file.exists("output/model-table-w-results.rds"),
+  "run make-original-tables.R first" = file.exists(c("output/tables/table1-original.csv", "output/tables/table2-original.csv"))
+)
+
+results <- read_rds("output/model-table-w-results.rds")
+
+table1_orig <- read_csv("output/tables/table1-original.csv")
+table2_orig <- read_csv("output/tables/table2-original.csv")
+
+table1_orig <- table1_orig %>%
+  rename(row = model) %>%
+  pivot_longer(-c(horizon, row), names_to = "column", values_to = "auc_roc_paper")
+
+table2_orig <- table2_orig %>%
+  pivot_longer(-horizon, names_to = "column", values_to = "auc_roc_paper")
+
+table1_smooth <- results %>%
+  filter(table=="Table 1") %>%
+  dplyr::select(horizon, row, column, auc_roc_smoothed) %>%
+  arrange(horizon, row) %>%
+  rename(auc_roc_ourrep = auc_roc_smoothed)
+
+table1 <- full_join(table1_smooth, table1_orig) %>%
+  mutate(diff = auc_roc_ourrep - auc_roc_paper)
+
+table1 <- table1 %>%
+  select(-auc_roc_ourrep, -auc_roc_paper) %>%
+  pivot_wider(names_from = column, values_from = diff)
+
+write_csv(table1, "output/tables/ourrep-check-table1.csv")
+table1 %>%
+  knitr::kable("markdown", digits = 2)
+
+
+
+table2_smooth <- results %>%
+  filter(table=="Table 2") %>%
+  dplyr::select(horizon, row, column, auc_roc_smoothed) %>%
+  arrange(horizon, row) %>%
+  rename(auc_roc_ourrep = auc_roc_smoothed)
+
+table2 <- full_join(table2_smooth, table2_orig) %>%
+  mutate(diff = auc_roc_ourrep - auc_roc_paper)
+
+table2 <- table2 %>%
+  select(-auc_roc_ourrep, -auc_roc_paper) %>%
+  pivot_wider(names_from = column, values_from = diff)
+
[…]

### 2020-06-12 rep_original/_rep-check.R

@@ -1,0 +1,94 @@
+#
+#   How much do the replication results differ from the values reported in the
+#   B&S paper?
+#
+
+library(readr)
+library(dplyr)
+library(tidyr)
+
+setwd(here::here("rep_original"))
+
+table1_orig <- read_csv("extra/table1-original.csv")
+table2_orig <- read_csv("extra/table2-original.csv")
+
+table1_orig <- table1_orig %>%
+  rename(row = model) %>%
+  pivot_longer(-c(horizon, row), names_to = "column", values_to = "auc_roc_paper")
+
+table2_orig <- table2_orig %>%
+  pivot_longer(-horizon, names_to = "column", values_to = "auc_roc_paper")
+
+
+t1_top <- read_csv("tables/table1_top.csv") %>%
+  rename(row = X1) %>%
+  mutate(horizon = "1 month")
+t1_bottom <- read_csv("tables/table1_bottom.csv") %>%
+  rename(row = X1) %>%
+  mutate(horizon = "6 months")
+t1 <- bind_rows(t1_top, t1_bottom) %>%
+  pivot_longer(-c(row, horizon), names_to = "column", values_to = "auc_roc_rep") %>%
+  mutate(
+    row = case_when(
+      row=="base" ~ "Base specification",
+      row=="robust_maxnodes" ~ "Terminal nodes",
+      row=="robust_sampsize" ~ "Sample size",
+      row=="robust_ntree"    ~ "Trees per forest",
+      row=="robust_traintest1" ~ "Training/test sets 1",
+      row=="robust_traintest2" ~ "Training/test sets 2",
+      row=="robust_traintest3" ~ "Training/test sets 3",
+      row=="robust_DV1"        ~ "Coding of DV 1",
+      row=="robust_DV2"        ~ "Coding of DV 2",
+      TRUE ~ row
+    ),
+    column = case_when(
+      column=="escalation" ~ "Escalation",
+      column=="quad"       ~ "Quad",
+      column=="avg"        ~ "Average",
+      TRUE ~ column
+    )
+  )
+
+table1 <- full_join(t1, table1_orig) %>%
+  mutate(diff = auc_roc_rep - auc_roc_paper)
+
+table1 <- table1 %>%
+  select(-auc_roc_rep, -auc_roc_paper) %>%
+  pivot_wider(names_from = column, values_from = diff)
+
+write_csv(table1, "extra/rep-check-table1.csv")
[…]

### 2020-06-16 paper.Rmd
update repo URL after owner move

@@
 Dorff, and Shahryar Minhas both provided helpful feedback on this project.
-All the code and several additional analyses can be found at our replication archive at https://github.com/rickmorgan2/Blair-Sambanis-replication.
+All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
 abstract: We examine the research protocols in Blair and Sambanis (2020).
@@
 If we ran both with the same random seed, they would be identical in every respect, producing to forests of identical decision trees and, thus, identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined.
-We have documented details at https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5.]
+We have documented details at https://github.com/andybega/Blair-Sambanis-replication/issues/5.]
 
@@
 
-We now turn to an examination of our analysis that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/rickmorgan2/Blair-Sambanis-replication.] The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B&S Table 2).
+We now turn to an examination of our analysis that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/andybega/Blair-Sambanis-replication.] The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B&S Table 2).
 We will review the substantive implications of our updated analysis below, but the bottom line is that these changes turn B&S's conclusions on their heads.
@@
 More importantly, the substantive interpretation of results should not depend on a specific RNG seed.
-We found that the escalation model's AUC-ROC values generally fluctuate no more than 0.01 (see https://github.com/rickmorgan2/Blair-Sambanis-replication/blob/master/rep_nosmooth/variance.md), and thus are confident that the patterns we see are robust to the initial RNG state.] B&S's original results and interpretation regarding the superiority of the B&S model over alternative ICEWS models, including the 1,000+ covariate CAMEO model, are thus entirely conditional on the use of smoothed AUC-ROC.
+We found that the escalation model's AUC-ROC values generally fluctuate no more than 0.01 (see https://github.com/andybega/Blair-Sambanis-replication/blob/master/rep_nosmooth/variance.md), and thus are confident that the patterns we see are robust to the initial RNG state.] B&S's original results and interpretation regarding the superiority of the B&S model over alternative ICEWS models, including the 1,000+ covariate CAMEO model, are thus entirely conditional on the use of smoothed AUC-ROC.
 Even the Quad model is generally as good as the escalation model in all implementations and is frequently quite a bit better.

### 2020-06-29 rep_nosmooth/replication-check.Rmd
turn replication check scripts into Rmd/md notes
close #8, see #7

---
title: "Replication verification"
output: github_document
---

How much do our replication results differ from the values reported in the B&S paper? This note compares the B&S Table 1 and 2 AUC-ROC values we obtain from our modified replication, when using smoothed ROC curves (not the standard empirical ROC curves) to calculate AUC, as B&S do.

Our replication relies on completely rewritten code that differs substantially from B&S' original replication code, and thus we want to verify that the results are not spuriously different because our replication code is different. (The reason for rewriting the code was to allow for running the replication in parallel, which reduces the time needed to run it.)

There are a couple of known deviations:

- RNG seed variation: the original B&S code sets a RNG seed value and then proceeds sequentially through the dozens of random forest models reflected in the paper. Even if we were to set the same RNG seed value, our code runs models in parallel and thus the RNG state for any given model will be different than it is for a given model in the sequence of models B&S' code runs. In experiments (see [variance.md](variance.md)) we find that this generally produces on the order of 0.01 differences in AUC-ROC values, but sometimes more.
- We fixed the implementations of the "Weighted by PITF" and "PITF Split Population" models in Table 2, which probably accounts for the more dramatic AUC-ROC differences there.

Overall, it doesn't look to us like there are any glaring deviations from the original B&S results.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
```

```{r}
library(readr)
library(dplyr)
library(tidyr)

dir.create("output/tables")

stopifnot(
  "run model-runner.R first" = file.exists("output/model-table-w-results.rds"),
  "run make-original-tables.R first" = file.exists(c("output/tables/table1-original.csv", "output/tables/table2-original.csv"))
)

results <- read_rds("output/model-table-w-results.rds")

table1_orig <- read_csv("output/tables/table1-original.csv")
table2_orig <- read_csv("output/tables/table2-original.csv")

table1_orig <- table1_orig %>%
  rename(row = model) %>%
  pivot_longer(-c(horizon, row), names_to = "column", values_to = "auc_roc_paper")

table2_orig <- table2_orig %>%
  pivot_longer(-horizon, names_to = "column", values_to = "auc_roc_paper")

```

## B&S Table 1

```{r}
table1_smooth <- results %>%
  filter(table=="Table 1") %>%
  dplyr::select(horizon, row, column, auc_roc_smoothed) %>%
  arrange(horizon, row) %>%
  rename(auc_roc_ourrep = auc_roc_smoothed)

table1 <- full_join(table1_smooth, table1_orig) %>%
  mutate(diff = auc_roc_ourrep - auc_roc_paper)

table1 <- table1 %>%
  select(-auc_roc_ourrep, -auc_roc_paper) %>%
  pivot_wider(names_from = column, values_from = diff)

write_csv(table1, "output/tables/ourrep-check-table1.csv")
table1 %>%
  knitr::kable("markdown", digits = 2)
```

## B&S Table 2

```{r}
table2_smooth <- results %>%
  filter(table=="Table 2") %>%
  dplyr::select(horizon, row, column, auc_roc_smoothed) %>%
  arrange(horizon, row) %>%
  rename(auc_roc_ourrep = auc_roc_smoothed)

table2 <- full_join(table2_smooth, table2_orig) %>%
  mutate(diff = auc_roc_ourrep - auc_roc_paper)

table2 <- table2 %>%
  select(-auc_roc_ourrep, -auc_roc_paper) %>%
  pivot_wider(names_from = column, values_from = diff)

write_csv(table2, "output/tables/ourrep-check-table2.csv")
table2 %>%
  knitr::kable("markdown", digits = 2)
```

### 2020-06-29 rep_nosmooth/README.Rmd
add explicit replication checks
addes tables that compare both the originanl replication and our modified replication results to the values reported in Tables 1 and 2 in the paper

see #7, #8

@@
 
+- [replication-check.md](replication-check.md) compares the AUC-ROC results we obtain when using smoothed ROC curves to the original B&S results (hand-copied from the paper).
+When taking into account the implementation fixes we made for two models, and RNG seed variation, we are able to replicate the original results/patterns.
 - [variance.md](variance.md) examines how much the AUC-ROC results vary when you don't set an RNG seed.
@@
 - `make-original-tables.R` is a by-hand coding of the values reported for Tables 1, 2, and 4 in B&S 2020. 
+- `rep-check.R` compares the results of our modified replication to the original values reported in Tables 1 and 2 of the B&S paper.
 

### 2020-06-29 rep_original/_replication-check.Rmd

---
title: "Replication verification"
output: github_document
---

How much do the replication results differ from the values reported in the B&S paper?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

library(dplyr)
library(readr)
library(tidyr)
```

The values in `table1-original.csv` and `table2-original.csv` are hand-copied values from the paper tables. The paper tables include two decimal digits. In this note, we compare the original Table 1 and 2 AUC-ROC values to the values we obtained from re-running the replication code.

```{r load-data, message=FALSE, warning=FALSE}
table1_orig <- read_csv("extra/table1-original.csv")
table2_orig <- read_csv("extra/table2-original.csv")

t1_top <- read_csv("tables/table1_top.csv") %>%
  rename(row = X1) %>%
  mutate(horizon = "1 month")
t1_bottom <- read_csv("tables/table1_bottom.csv") %>%
  rename(row = X1) %>%
  mutate(horizon = "6 months")

t2_top <- read_csv("tables/table2_top.csv") %>%
  rename(row = X1) %>%
  mutate(horizon = "1 month")
t2_bottom <- read_csv("tables/table2_bottom.csv") %>%
  rename(row = X1) %>%
  mutate(horizon = "6 months")
```

## Table 1

The table below shows the difference between B&S Table 1 and our re-creation. All values are 0, meaning that we match the paper results.

```{r}
table1_orig <- table1_orig %>%
  rename(row = model) %>%
  pivot_longer(-c(horizon, row), names_to = "column", values_to = "auc_roc_paper")

table2_orig <- table2_orig %>%
  pivot_longer(-horizon, names_to = "column", values_to = "auc_roc_paper")

t1 <- bind_rows(t1_top, t1_bottom) %>%
  pivot_longer(-c(row, horizon), names_to = "column", values_to = "auc_roc_rep") %>%
  mutate(
    row = case_when(
      row=="base" ~ "Base specification",
      row=="robust_maxnodes" ~ "Terminal nodes",
      row=="robust_sampsize" ~ "Sample size",
      row=="robust_ntree"    ~ "Trees per forest",
      row=="robust_traintest1" ~ "Training/test sets 1",
      row=="robust_traintest2" ~ "Training/test sets 2",
      row=="robust_traintest3" ~ "Training/test sets 3",
      row=="robust_DV1"        ~ "Coding of DV 1",
      row=="robust_DV2"        ~ "Coding of DV 2",
      TRUE ~ row
    ),
    column = case_when(
      column=="escalation" ~ "Escalation",
      column=="quad"       ~ "Quad",
      column=="avg"        ~ "Average",
      TRUE ~ column
    )
  )

table1 <- full_join(t1, table1_orig) %>%
  mutate(diff = auc_roc_rep - auc_roc_paper)

table1 <- table1 %>%
  select(-auc_roc_rep, -auc_roc_paper) %>%
  pivot_wider(names_from = column, values_from = diff)

write_csv(table1, "extra/rep-check-table1.csv")
table1 %>%
  knitr::kable("markdown", digits = 2, format.args = list(nsmall = 2))
```

## Table 2

Our reproduction values also match the B&S paper values for Table 2.

```{r}
t2 <- bind_rows(t2_top, t2_bottom) %>%
  pivot_longer(-c(row, horizon), names_to = "column", values_to = "auc_roc_rep") %>%
  mutate(
    row = NULL,
    column = case_when(
      column=="escalation"    ~ "Escalation Only",
      column=="with_PITF"     ~ "With PITF Predictors",
      column=="weighted_PITF" ~ "Weighted by PITF",
      column=="split_PITF"    ~ "PITF Split Population",
      column=="PITF"          ~ "PITF Only",
      TRUE ~ column
    )
  )

table2 <- full_join(t2, table2_orig) %>%
  mutate(diff = auc_roc_rep - auc_roc_paper)

table2 <- table2 %>%
  select(-auc_roc_rep, -auc_roc_paper) %>%
  pivot_wider(names_from = column, values_from = diff)

write_csv(table2, "extra/rep-check-table2.csv")
table2 %>%
  knitr::kable("markdown", digits = 2, format.args = list(nsmall = 2))
```

### 2020-06-29 paper.Rmd
add Cederman Weidman 2017 reference

@@
 bibliography: "./references.bib"
+nocite: |
+  @cederman:weidmann:2017
 ---
@@
 # Introduction
+
 Blair and Sambanis [-@blair:sambanis:2020, hereafter B\&S] argue that theory is essential for creating models that have high accuracy in forecasting civil war onset.
@@
 We will return to their analysis later, after correcting the procedural mistakes we found in their research process.
-
-nocite: |
-  @cederman:weidmann:2017 
-...
 

### 2020-07-09 paper.Rmd
add para on model development and test set contamination

@@
  - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
- - Richard K. Morgan^[V-Dem, rick.morgan2@gmail.com.]
+ - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
  - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
@@
 Standard practice in conflict research and forecasting has been to use empirical, not smoothed ROC curves, both for visualization and when calculating AUCs.
-
-Aside from the visual impact of ROC smoothing, the underlying motivation and methods involved in smoothing are broader and involve parametric estimation of population ROC curves [@henley:1988; @henley:2014].
-The main application is in the evaluation of medical tests (e.g.
-diagnostic imaging) with early methods dating back several decades and ongoing development of both parametric and non-parametric smoothing methods [e.g.
-@zou:etal:1997; @pulit:2016].
-It is not established which methods are preferable in a given application [@henley:2014].
-Notably, the context in which ROC smoothing is used differs in important aspects from typical conflict research applications.
-Conflict data are closer to a census of the population and with well-known dependencies like spatial and temporal correlation, rather than a random sample that is approximately independently and identically distributed.
-It is thus neither clear that smoothing is justified nor valid.
-We also note that smoothed ROC plots are not widely used outside of the narrow context mentioned above.
 
@@
 
+Aside from the visual impact of ROC smoothing, the underlying motivation and methods involved in smoothing are broader and involve parametric estimation of population ROC curves [@henley:1988; @henley:2014].
+The main application is in the evaluation of medical tests (e.g.
+diagnostic imaging) with early methods dating back several decades and ongoing development of both parametric and non-parametric smoothing methods [e.g.
+@zou:etal:1997; @pulit:2016].
+It is not established which methods are preferable in a given application [@henley:2014].
+Notably, the context in which ROC smoothing is used differs in important aspects from typical conflict research applications.
+Conflict data are closer to a census of the population and with well-known dependencies like spatial and temporal correlation, rather than a random sample that is approximately independently and identically distributed.
+It is thus neither clear that smoothing is justified nor valid.
+We also note that smoothed ROC plots are not widely used outside of the narrow context mentioned above.
+
 ## Using Inconsistent Ns
@@
 As such, their baseline comparison is an unfortunate standard bearer for strong theory.
+
+Another important consideration when developing a predictive model is to do so without contaminating the out-of-sample test set.
+This occurs if test set performance is used to inform specification---predictive modeling's analogue to p-hacking---and special care is needed [e.g.
+see the approach by @colaresi:mahmood:2017].
+B&S did pre-register their 2016-H1 forecasts, which verifies the integrity of the forecasts, but this does not assure the model development process.
+To be clear: there is no reason to suspect test set contamination in the escalation or other models.
+Rather, it is challenging to provide verifiable assurance thereof in a single analysis iteration that both develops and tests a model on data that was available contemporaneously.
 

### 2020-07-10 submissionVersion.Rmd
sub
mission

--- 2020-07-09 paper.Rmd
@@
 author: 
- - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
- - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
- - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
+# - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
+# - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
+# - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
+ - Anne Author
 date: "`r format(Sys.time(), '%d %B %Y')`"
-thanks: John Ahlquist, Cassy L.
+# thanks: John Ahlquist, Cassy L.
 Dorff, and Shahryar Minhas both provided helpful feedback on this project.
@@
     toc: false
-#header-includes:
-#  - \usepackage{setspace}\doublespacing
+header-includes:
+  - \usepackage{setspace}\doublespacing
 bibliography: "./references.bib"

### 2020-07-10 submissionVersion.Rmd
drac
asdf

@@
 
-The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015,beger:etal:2017].
+The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015;@beger:etal:2017].
 Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision trees.

### 2020-07-10 submissionVersion.Rmd
moar stuffs
sdfdf

@@
 bibliography: "./references.bib"
+biblio-style: chicago.bst
 nocite: |

### 2020-09-04 paper.Rmd
small edit in conclusion
September 1-4 smoothing

@@
 Indeed they assert that with such theory forecasting is more accurate than has previously been possible.
-Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in their article.
-We find that these claims are false.
+Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in support of it.
+We find that these claims to be unsupported and evidence presented for them to be incorrect.
 Their theory-based escalation model does not do better than the alternatives that they examine.
 Actually, it does worse.
-The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure.
-The performance results they report are based on smoothed performance curves, not the original unsmoothed curves.
-Further, two of the structural alternatives to their basic escalation model were incorrectly implemented.
+The reason for this is that they have made several mistakes in their research procedure.
+Further, the performance results they report are based on smoothed performance curves, not the original unsmoothed curves.
+This provides misleading results.
+In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented.
 We also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset.
@@
 B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately".
-Indeed, B&S argue that this so-called theoretical model outperforms four other "more mechanical" alternatives.
-Second, they claim that the integration of structure with process is better over short forecasting windows.
+Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.\footnote{We set aside for here the characteristics of theoretical models, accepting the B\&S denotation.}  They also claim that the integration of structure with process is better than alternaties over short forecasting windows.
 Third, they preregistered the list of thirty countries that have the highest risk of civil war onset.
@@
 
-B&S' analysis is based on the use of non-linear, non-deterministic machine learning models, and specifically random forests, one of which has a specification they argues is theory-based, and several others with more generic sets of covariates.
+B&S' analysis is based on the use of non-linear, non-deterministic machine learning models, and specifically random forests, one of which has a specification they argue is theory-based.
+Several others with more generic sets of covariates are also developed and compared.
 Notably, the analysis is at the country-month or country-(6 month) levels and relies in large part on indicators derived from the ICEWS event data.
-For the moment, we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models.
 In short, they uphold their assumption that theory-guided empirical research produces better conflict predictions than machine learning inspired efforts that are necessarily ad hoc combinations of available variables.
@@
 
-- Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model, for interactions between the government on one side and opposition or rebel actors on the other. 
+- Escalation: a set of ten indicators, drawn from a theoretical escalation model, for interactions between the government on one side and opposition or rebel actors on the other. 
 - Quad: ICEWS quad counts, i.e.
@@
 
-Finally, B&S used their escalation model to create forecasts for the first half of 2016, and in their third and final analysis, they score the forecasts accuracy using civil war onset data later observed.
+Finally, B&S used their escalation model to create forecasts for the first half of 2016.
+In their third analysis, they score the forecasts accuracy using subsequent civil war onset data.
 This is summarized in B&S Table 3.
@@
 The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves.
-The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves as easing interpretation.
-But in fact smoothing is used to produce all AUC-ROC values they report.
-Standard practice in conflict research and forecasting has been to use empirical, not smoothed ROC curves, both for visualization and when calculating AUCs.
+The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves justified as easing interpretation.
+Smoothing is used to produce all AUC-ROC values they report.
+There is no difficulty to be overcome in using non-smoothed plots.
+Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs.
 
@@
 
-B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
+We believe that B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
 However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`).
@@
 This description suggests that B&S intended to run two separate random forest models, one each on the low- and high-risk training data splits.
-The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consists of the full training data from all other models.
+The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consist of the full training data from all other models.
 In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.
@@
 
-Further, the AUC-ROC values reported in the original B&S Table 2 are calculated based on slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}).
+Further, the AUC-ROC values reported in the original B&S Table 2 are calculated  on slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}).
 ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available.
@@
 Thus a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
-In other words, the results for different models in B&S Table 2 are not comparable to one another, and any conclusions drawn from such comparison are potentially incorrect.
+In other words, the results for different models in B&S Table 2 are not comparable to one another, and conclusions drawn from such comparison are potentially incorrect.
 
@@
 
-We now turn to an examination of our analysis that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/andybega/Blair-Sambanis-replication.] The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B&S Table 2).
+We now turn to an examination that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/andybega/Blair-Sambanis-replication.] The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B&S Table 2).
 We will review the substantive implications of our updated analysis below, but the bottom line is that these changes turn B&S's conclusions on their heads.
@@
 
-This implies that the ROC curves were only smoothed in the referenced Figure 1.
+This suggest that using smoothed ROC curves will have no substantive impact on the results; however, in this smoothing determines the comparative results.
+Their statement also implies that the ROC curves were only smoothed in the referenced Figure 1.
 However, this is not the case.
-All AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation.
+All AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves *prior* to AUC calculation.
 ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like.
@@
 Notice also that the smoothing averages the left-most almost straight line with the right-most almost straight line in a monotonic way.
+Thus, smoothing changes the ordering of alternatives.
+It is not a cosmetic change as suggested.
 
@@
 
-What impact did the ROC smoothing have overall on the performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B&S Table 2)?
+What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B&S Table 2)?
 Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values.
@@
 The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05.
-This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the result B&S find, namely that the escalation model is generally superior.
+This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B&S report, namely that the escalation model is generally superior.
 
@@
 The Quad model appears to be roughly on par with the Escalation model.
-Thus, the original B&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves and overturns when using conventional AUC-ROC calculations.
+Thus, the original B&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves.
 
@@
 
-The PITF split-population analog still performs well, but the simple Escalation + PITF predictors model arguable performs better still.
+The PITF split-population analog still performs well, but the simple Escalation + PITF predictors model arguably performs better still.
 
@@
 
-Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model, and by quite a margin in the 6-month data version.
+Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model by quite a margin in the 6-month data version.
 
@@
 One finding that remains is that the "PITF Only" model is outperformed by the "Escalation Only" model.
-As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is neither surprising, nor noteworthy.
+As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is neither surprising, nor especially noteworthy.
 
@@
 
-B&S report (Table 4)confusion matrices for their forecasts for the first half of 2016 (2016-H1).
+B&S report (in their Table 4) confusion matrices for their forecasts for the first half of 2016 (2016-H1).
 To create the confusion matrices, B&S treat the 30 highest ranked predictions as positive predictions ("1s") and the rest as negative predictions ("0s").
@@
 This corresponds to a positive rate of around 9.5% for the first half of 2016 data.
-In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a much lower positive rate of around 0.5%.
+In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a much lower positive rate of around 0.5%, i.e., an order of magnitude smaller.
 The positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data.
@@
 In the default "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1.
-Thus, the recall values is undefined, while the precision is 0/30 = 0, compared to reported recall and precision values of 13/15 = 0.87 and 13/30 = 0.43.
+Thus, the recall value is undefined, while the precision is 0/30 = 0, compared to reported recall and precision values of 13/15 = 0.87 and 13/30 = 0.43.
 The alternative coding ("Assuming Change") produces two civil war onsets.
-Recall is 1.0 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.
-
-Another, minor issue or rather coding error, is related to using a lead version of the DV.
+Recall is 1.0 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.\footnote{Another, minor issue or rather coding error, is related to using a lead version of the DV.
 With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 indicates the risk of civil war onset in 2016-H2.
@@
 Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1.
-In this case, it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.
+In this case, it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.}
 
@@
 When we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturned.
-In other words, B&S findings are based on a faulty analysis, and invalid.
+In other words, B&S findings are based on a faulty analysis, invalid, and should be disregarded.
 
@@
 
-- The theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model.
+- The theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model and the ensemble average model.
 - The Average ensemble model and the CAMEO models outperform the escalation model in all instances.

### 2020-09-04 paper.Rmd
sept 4
fixed on footnote

@@
 
-B&S interpret their results as follows, on page 20 and we comment on the conclusions seratim:^[We list the "Overall, ..." interpretation out of order, last, for clarity.]
+B&S interpret their results as follows, on page 20 and we comment on the conclusions seratim:^[We list the "Overall, ..." interpretation out-of-order for clarity.]
 
@@
 The alternative coding ("Assuming Change") produces two civil war onsets.
-Recall is 1.0 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.\footnote{Another, minor issue or rather coding error, is related to using a lead version of the DV.
+Recall is 1.0 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.^[Another, minor issue is related to using a lead version of the DV.
 With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 indicates the risk of civil war onset in 2016-H2.
@@
 Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1.
-In this case, it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.}
+In this case, it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.]
 

### 2020-09-04 paper.Rmd
add appendix figure with diff't smoothing methods
extends Figure 2 that shows the benefit of (binormal) smoothing

@@
 Given the dramatic changes in results as a result of the preceding issues, we did not further investigate these potential concerns.
+
+## Different smoothing methods
+
+The **pROC** package `smooth()` function includes several different smoothing methods.
+The default, which BS use in their code, is binormal smoothing.
+Here is a replication of the smooth benefits shown in Figure \ref{fig:benefit-plot} with two other smoothing methods.
+(We could not get "logcondens" and "logcondens.smooth" to work, and they are based on another external package that has not been updated since 2016.)
+
+```{r smooth-benefit-plot-extended, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2), with 3 different smoothing methods (binormal is the default method and used in B\\&S)"}
+
+suppressPackageStartupMessages({
+  library(tidyverse)
+  library(yardstick)
+  library(pROC)
+  library(forcats)
+})
+
+preds <- read_rds("rep_nosmooth/output/all-predictions.rds")
+
+auc_roc_vec <- function(pred, truth, smooth, sm = "binormal") {
+  roc_obj <- pROC::roc(truth, pred, auc = TRUE, quiet = TRUE, smooth = smooth, smooth.method = sm)
+  as.numeric(roc_obj$auc)
+}
+
+rocs <- preds %>% 
+  filter(!is.na(value)) %>%
+  group_by(cell_id) %>% 
+  summarize(
+    no = auc_roc_vec(pred = pred, truth = value, smooth = FALSE),
+    binormal = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "binormal"),
+    density = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "density"),
+    fitdistr = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "fitdistr")
+  )
+
+# these two take a long time and error out for some reason
+#logcondens = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "logcondens"),
+#logcondens.smooth = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "logcondens.smooth")
+
+mt <- read_rds("rep_nosmooth/output/model-table-w-results.rds")
+mt <- mt %>% left_join(rocs, by = "cell_id")
+
+cols <- c(
+  Escalation = "#e41a1c",
+  Goldstein = "#ff7f00",
+  Quad = "#984ea3",
+  CAMEO = "#377eb8",
+  Average = "#4daf4a",
+  `With PITF Predictors` = "gray50", 
+  `Weighted by PITF` = "gray50", 
+  `PITF Split Population` = "gray50",
+  `PITF Only` = "gray50"
+)
+
+model_levels <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average", 
+                  "With PITF Predictors", "Weighted by PITF", "PITF Split Population",
+                  "PITF Only")
+
+benefit <- mt %>% 
+  mutate(column = fct_recode(column, Escalation = "Escalation Only"),
+         column = factor(column, levels = rev(model_levels))) %>%
+  mutate(binormal = binormal - no,
+         density  = density - no,
+         fitdistr = fitdistr - no) %>%
+  select(cell_id:non_RF, binormal:fitdistr) %>%
+  pivot_longer(binormal:fitdistr, names_to = "method")
+
+means <- benefit %>%
+  group_by(column, method) %>%
+  summarize(avg_benefit = mean(value))
+
+ggplot(benefit, aes(x = value, y = column, color = column)) +
+  facet_wrap(~method) +
+  geom_point() +
+  scale_color_manual(guide = FALSE, values = cols) +
+  theme(axis.text.y = element_text(size = 12)) +
+  geom_point(data = means, aes(x = avg_benefit), shape = "|", color = "black", size = 6) +
+  labs(x = "AUC(smoothed ROC) - AUC(empirical ROC)",
+       y = "") +
+  theme_bw() +
+  geom_vline(xintercept = 0, linetype = 3) +
+  theme(axis.text.y = element_text(size = 12))
+```

### 2020-09-17 journal-survey/README.Rmd

---
title: "Journal survey"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Survey journals to check use of ROC/AUC and specifically how common it is to use smoothed ROC curves or AUCs derived from them.

"template.csv" is a template for adding a new journal issue (or other group of articles).

## Results

_(This README page is generated from a Rmarkdown file; re-knit that to update the results.)_

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

setwd(here::here("journal-survey"))

articles <- tibble(file = dir("data-raw", full.names = TRUE)) %>%
  mutate(data = suppressMessages(purrr::map(file, read_csv))) %>%
  unnest(data)

# mark articles that were referenced in B&S
articles$bsref <- as.integer(str_detect(articles$file, "bs\\-references"))
articles$file <- NULL

# take out items that are not coded yet (missing binary_outcome column)
articles <- articles[!is.na(articles$binary_outcome), ]

# if one of the payload columns is NA, set it to 0
articles$uses_auc[is.na(articles$uses_auc)] <- 0L
articles$uses_roc[is.na(articles$uses_roc)] <- 0L
articles$uses_smoothed_auc[is.na(articles$uses_smoothed_auc)] <- 0L
articles$uses_smoothed_roc[is.na(articles$uses_smoothed_roc)] <- 0L

cat("How many samples articles had binary outcome?\n")
table(articles$binary_outcome)

# The rest below is only for binary outcome articles
binary <- articles %>% filter(binary_outcome==1L)

smry <- binary %>%
  mutate(any_smoothed = uses_smoothed_auc | uses_smoothed_roc,
         any_empirical = uses_auc | uses_roc,
         uses_any = any_smoothed | any_empirical,
         only_smoothed = any_smoothed & !any_empirical,
         only_empirical = any_empirical & !any_smoothed,
         both = any_smoothed & any_empirical) %>%
  summarize(
    articles_coded = nrow(articles),
    articles_with_binary_outcome = n(),
    use_any_roc_measure = sum(uses_any),
    only_smoothed = sum(only_smoothed),
    only_empirical = sum(only_empirical),
    both = sum(both)
  ) %>%
  pivot_longer(everything())

knitr::kable(smry)

```

### 2020-09-17 paper.Rmd
first edits
mdw

@@
 ---
-title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
+title: "Reassessing the role of theory and machine learning in forecasting civil conflict"
 author: 
@@
 Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs.
+We examined N articles in K journals over the period from 20xx-2020 looking for articles which have a binary dependent variable and use the ROC techonology in the analysis section.
+We found that XX percent of these articles use the empirical ROC calculations and only 100-XX use a smoothed curve.
+This is not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.
+What is surprising is that anyone would choose the smoothed ROC implementation.
+INSERT SMALL STUDY HERE WITH DETAILS.
 
@@
 
-B&S advocate the use of theory to guide prediction.
-But "theory" as used by B&S is an ambiguous and undefined concept.
+B\&S advocate the use of theory to guide prediction.
+But "theory" as used by B\&S is an ambiguous and undefined concept.
+`Theory'' has many interpretations in social science.
+A theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.
+The additional statements can be subjected to empirical scrutiny.
+An example might be William Riker's theory of coalitions which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically.
+For the purposes of discussion we call this strong theory.
+
+However, this is not what many mean by theory in the contemporary social science. For many,
+a theory is a falsifiable story or narrative that makes causal claims about the relationships among variables.
+A theory in this sense could be entirely created by the scholar.
+More typically it is fused together from a meta-analysis of prior work on the topic under consideration.
+We call these types of theory weak theory.
+
+They are weak because they do not include two characteristics of strong theory, though they may include one:
+1. A theory is separate from the case or cases being examined. It is intented to be general.
+2. A theory generates new statements that in principle can be empirically falsified, not proven.
+
+Throwing a bunch of variables that you think are related to a dependent variable and popping then Into a non parametric model is not a theory-driven model.
+
 It is not a procedure.

### 2020-09-18 paper.Rmd
moar pie
mdw

@@
 
-Throwing a bunch of variables that you think are related to a dependent variable and popping then Into a non parametric model is not a theory-driven model.
+There is another type of theory we can define ostensibly.
+It consists of taking a variety of variables that putatively relate to another (dependent) variable and running a non-parametric model such as a logistic regression with this set of variables.
+This even weaker form of theory is the most widespread use of theory in the contemporary social sciences, and it is the type of theory that is employed in B\&S:
 

### 2020-09-18 journal-survey/README.Rmd
finish data entry for B&S references survey

@@
 "template.csv" is a template for adding a new journal issue (or other group of articles).
+
+The sample consists of:
+
+1. All references in the Blair and Sambanis paper, excluding books and non-academic articles. 
+2. Two issues of JCR; 2020 volumes 1 and 9.
 
@@
 
-# take out items that are not coded yet (missing binary_outcome column)
-articles <- articles[!is.na(articles$binary_outcome), ]
+# take out items that are not coded yet (missing "journal" column)
+articles <- articles[!is.na(articles$journal), ]
 
@@
 articles$uses_smoothed_roc[is.na(articles$uses_smoothed_roc)] <- 0L
+articles$unclear_auc[is.na(articles$unclear_auc)] <- 0L
+articles$unclear_roc[is.na(articles$unclear_roc)] <- 0L
 
@@
 
-smry <- binary %>%
+binary <- binary %>%
   mutate(any_smoothed = uses_smoothed_auc | uses_smoothed_roc,
          any_empirical = uses_auc | uses_roc,
-         uses_any = any_smoothed | any_empirical,
-         only_smoothed = any_smoothed & !any_empirical,
-         only_empirical = any_empirical & !any_smoothed,
-         both = any_smoothed & any_empirical) %>%
+         any_unclear = unclear_roc | unclear_auc,
+         uses_any = any_smoothed | any_empirical | any_unclear,
+         only_smoothed = any_smoothed & !any_empirical & !any_unclear,
+         only_empirical = any_empirical & !any_smoothed & !any_unclear,
+         only_unclear = any_unclear & !any_empirical & !any_smoothed)
+
+smry <- binary %>%
   summarize(
-    articles_coded = nrow(articles),
+    items_coded = nrow(articles),
     articles_with_binary_outcome = n(),
@@
     only_empirical = sum(only_empirical),
-    both = sum(both)
+    only_unclear = sum(only_unclear)
   ) %>%
@@
 ```
+
+What items had unclear or ambiguous codings?
+
+```{r}
+binary %>% 
+  filter(any_unclear) %>%
+  select(title, year, journal, notes) %>%
+  knitr::kable()
+```
+
+Note that both articles include Blair as co-author.
+
+## Data
+
+- "uses_roc", "uses_smoothed_roc", "unclear_roc": Does the paper include figures showing (empirical / smoothed / unclear) ROC curves?
+- "uses_auc", "uses_smoothed_auc", "unclear_auc": Does the paper include tables showing (empirical / smoothed / unclear) AUC-ROC values?
+
+If the paper includes visually obvious empirical ROC curves, we assumed that any corresponding AUC-ROC values were also generated from empirical ROC curves.
+
+Some papers had AUC values but no ROC curves. In these cases we consulted replication code. These are listed below.
+
+```{r}
+articles %>% 
+  filter(!uses_roc & uses_auc) %>%
+  select(title, year, journal, notes) %>%
+  knitr::kable()
+```
+
+```{r}
+write_csv(articles, "coded-items.csv")
+```

### 2020-09-18 paper.Rmd
finish data entry for B&S references survey

@@
 Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs.
-We examined N articles in K journals over the period from 20xx-2020 looking for articles which have a binary dependent variable and use the ROC techonology in the analysis section.
-We found that XX percent of these articles use the empirical ROC calculations and only 100-XX use a smoothed curve.
-This is not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.
+We examined all 63 references in B&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), for a total of 78 items, in order to survey the use of ROC curves.
+Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values.
+We found that 90%--17 of 19--clearly used empirical ROC curves, and 2 (10%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code.
+The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B&S.
+The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey.} (Both of the ambiguous cases include Blair as coauthor.) The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.
 What is surprising is that anyone would choose the smoothed ROC implementation.
+
 INSERT SMALL STUDY HERE WITH DETAILS.

### 2020-09-18 paper.Rmd
back to you, andy

@@
 Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs.
-We examined all 63 references in B&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), for a total of 78 items, in order to survey the use of ROC curves.
+We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), for a total of 78 items, in order to survey the use of ROC curves.
 Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values.
 We found that 90%--17 of 19--clearly used empirical ROC curves, and 2 (10%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code.
-The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B&S.
+The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S.
 The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey.} (Both of the ambiguous cases include Blair as coauthor.) The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.
 What is surprising is that anyone would choose the smoothed ROC implementation.
-
-INSERT SMALL STUDY HERE WITH DETAILS.
 
@@
 
-What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B&S Table 2)?
+What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B\&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B\&S Table 2)?
 Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values.
@@
 
-There is another type of theory we can define ostensibly.
-It consists of taking a variety of variables that putatively relate to another (dependent) variable and running a non-parametric model such as a logistic regression with this set of variables.
-This even weaker form of theory is the most widespread use of theory in the contemporary social sciences, and it is the type of theory that is employed in B\&S:
+Most social science research uses a weak form of theory.
+In the case of B\&S they base their theoretical thinking on research that argues that conflict is not the product of structural factors, but rather comes from an interaction of state repression and the changes in costs and benefits to citizens for engaging in anti-government conflict.
+Following a wide swath of scholars, they argue that the escalation and de-escalations of interactions between citizens and government are what typically lead up to the onset of civil wars.
+Operationally, this boils down to four types of concepts: demands, accommodations,
+nonviolent repression, and low-level violence.
+\begin{itemize}
+\item Demands include events in which either opposition or rebels is the source and
+government is the target and typically involve appeals for political or institutional
+reform, policy change, or extension of rights.
+\item Nonviolent repression and accommodation include events in which government
+is the source and either opposition or rebels is the target. Nonviolent
+repression includes threats, sanctions, restrictions on freedom of movement, or
+rejections of group demands.
+\item Low-level violence includes all events in which
+(1) government is the source and opposition is the target, (2) opposition is the
+source and government is the target, (3) government is the source and rebels are
+the target, or (4) rebels are the source and government is the target.
+\end{itemize}
+These variables, taken from ICEWS, are the main components of the escalation model in the B\&S article.
 
-It is not a procedure.
-They actually create a model with ten right-hand side variables that are supposed to capture a complicated repression-dissent dynamic.
-There is wide-ranging literature on this dynamic that could justify many specifications.
-As such, their baseline comparison is an unfortunate standard bearer for strong theory.
+Indeed most empirical social science research follows this approach: 1) delineation of an argument, 2) selection of variables that represent a plausible or compelling representation of the concepts, followed by 3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation.
+It is worth pointing out that this is simply one set of variables that are selected to represent the concepts in the theoretical story.
+A strong theory gives more guidance on compelling choices than does a weak theory.
+Accordingly, while the choice is plausible is not necessarily compelling.
+Another choice might be equally plausible.
+In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups.
+In practical terms there is a wide array of choices that are available to instantiate any particular theory.
+A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated temporal dynamics.
+One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.
+
+MORE HERE ON GENERAL IDEAS.  ANDY ANY BULLET pOINTs TO ADD?
 

### 2020-09-20 paper.Rmd
third shot

@@
 B\&S advocate the use of theory to guide prediction.
-But "theory" as used by B\&S is an ambiguous and undefined concept.
+But "theory" is not defined by B\&S, and may an ambiguous and undefined concept.
 `Theory'' has many interpretations in social science.
-A theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.
+For some---the EITM movement, for example---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.
 The additional statements can be subjected to empirical scrutiny.
 An example might be William Riker's theory of coalitions which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically.
+A more recent example is found in The Political Logic of Survival.
 For the purposes of discussion we call this strong theory.
 
-However, this is not what many mean by theory in the contemporary social science. For many,
+However, for many,
 a theory is a falsifiable story or narrative that makes causal claims about the relationships among variables.
-A theory in this sense could be entirely created by the scholar.
-More typically it is fused together from a meta-analysis of prior work on the topic under consideration.
+Typically it is fused together from a meta-analysis of prior work on the topic under consideration.
 We call these types of theory weak theory.
-
 They are weak because they do not include two characteristics of strong theory, though they may include one:
@@
 These variables, taken from ICEWS, are the main components of the escalation model in the B\&S article.
+The escalation theory is that the forces behind these variables interact such that as each of them increases, they independently increase the probability of civil conflict.
+The actual mechanisms of the dynamism among these variables are not specified.
 
@@
 One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.
+At this level of detail the theoretical model is more about chosing the correct variables.
+And while we don't undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.
 
-MORE HERE ON GENERAL IDEAS.  ANDY ANY BULLET pOINTs TO ADD?
+Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning.
+Even if B\&S were correct that in their study the theory based escalation model performed better, it would simple be a case study of one case where it was preferable.
+It establishes no general tendency, let alone law-like evidence for general theory.
+Nor, against it.
 
-Another important consideration when developing a predictive model is to do so without contaminating the out-of-sample test set.
+#An important consideration when developing a predictive model is to do so without contaminating the out-of-sample test set.
 This occurs if test set performance is used to inform specification---predictive modeling's analogue to p-hacking---and special care is needed [e.g.
@@
 
-Moreover, they misunderstand the use of ICEWS event data in current research.
+#Moreover, they misunderstand the use of ICEWS event data in current research.
 They claim that most applications to date have focused on the quad categories, but this ignores a wide swath of literature [@steinert-threlkeld:2017; and @metternich:etal:2013] that uses a specific action---such as protest---defined in the CAMEO ontology.
@@
 The problems we encountered are not subjective modeling choices.
-When we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturned.
-In other words, B&S findings are based on a faulty analysis, invalid, and should be disregarded.
+They bear no implications on the question of whether theory based model are preferable.
+However,  when we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturned.
+In other words, B&S findings are based on a faulty analysis, invalid, and should be disregarded as evidence in favor of theory-based models.
 
@@
 It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.
+Social science has leaned on theory for (at least) seventy years, but it might be time to find useful ways to incorporate some of the recently developed methods in the arena of machine learning as well.
 

### 2020-09-21 paper.Rmd
add cites
also move some footnotes and text discussion code lines to appendix

@@
 The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves.
-The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves justified as easing interpretation.
-Smoothing is used to produce all AUC-ROC values they report.
+The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation.
+However, smoothing is also used in all AUC-ROC values they report.
 There is no difficulty to be overcome in using non-smoothed plots.
 Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs.
-We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), for a total of 78 items, in order to survey the use of ROC curves.
+We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), in order to survey the use of ROC curves.
 Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values.
@@
 
-The next three issues we encountered all concern information in B&S Table 2.
+The next three issues we encountered all concern information in B&S Table 2, which is the key table in which B&S report AUC-ROC values to demonstrate the "Escalation" models superiority.
 
@@
 We believe that B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
-However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`).
-The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions.
-Similarly, in the 6-month version of this file.] This appears to be an easily corrected coding error on the part of B&S.
+However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.
+This appears to be an easily corrected coding error on the part of B&S.
 
@@
 In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.
-
 The model specifications are also identical; i.e., they use the same *x* variables and the same random forest hyper-parameter settings.
 The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself.
-If we ran both with the same random seed, they would be identical in every respect, producing to forests of identical decision trees and, thus, identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined.
-We have documented details at https://github.com/andybega/Blair-Sambanis-replication/issues/5.]
+If we ran both with the same random seed, they would be identical in every respect, producing to forests of identical decision trees and, thus, identical predictions.
 
@@
 B&S present a confusion matrix to score their 2016-H1 forecasts in Table 4.
-Although the forecasts are for the probability of civil war onset, in the replication code, they are actually scored using incidence of civil war.
-
-The relevant variables in the data are "incidence_civil_ns" and "incidence_civil_ns_plus1", which appears to be a 1-period lead (ie., $t_+1$) version of the dependent variable that is used in the actual prediction models.
-The incidence dependent variable  contains both 0/1 and missing values.
-By examining the pattern of missing values, it seems clear that this was originally an incidence variable indicating whether a country was at civil war in a given year or not, and which was converted to an onset version so that onsets retain the value of 1 but continuing civil war years are coded as missing.
-This reflects common practice.
-
-However, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
-In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable to incidence.
+Although the forecasts are for the probability of civil war _onset_, in the replication code, they are actually scored using _incidence_ of civil war.
+By definition the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict.
+When forecasting rare events, it is common to have many false positives, i.e.
+instances where we predict onset but not onsets occur.
+_Ex ante_, using incidence rather than onset to score onset forecasts should decrease the number of false positives and correspondingly increase the number of true positives (good predictions) and false negatives (missed onsets).
 
@@
 
-```{r benefit-plot, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2)"}
+```{r benefit-plot, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2).
+The Escalation model is the only model which consistently benefits from using smoothed ROC curves, and this advantage accounts for its apparent superiority to the other models."}
 table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
@@
 
+Predicting rare events like civil war onset is exceedingly difficult, and it is not unusual at all that most of the predictions are false positives.
+This will by definition be the case when there are no onsets in a time period, as is true with the default "Assuming Persistence" coding that is in the data.
+Correctly predicting the two onsets in the alternative "Assuming Change" coding is at face value a credit to the model's accuracy.
+
 # Conclusion
@@
 `Theory'' has many interpretations in social science.
-For some---the EITM movement, for example---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.
+For some---the EITM movement, for example [@aldrich:etal:2008]---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.
 The additional statements can be subjected to empirical scrutiny.
-An example might be William Riker's theory of coalitions which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically.
-A more recent example is found in The Political Logic of Survival.
+An example might be William Riker's theory of coalitions [@riker:1962] which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically.
+A more recent example is found in The Political Logic of Survival [@bdm:etal:2005].
 For the purposes of discussion we call this strong theory.
 
-However, for many,
-a theory is a falsifiable story or narrative that makes causal claims about the relationships among variables.
+However, for many, a theory is a falsifiable story or narrative that makes causal claims about the relationships among variables.
 Typically it is fused together from a meta-analysis of prior work on the topic under consideration.
@@
 They are weak because they do not include two characteristics of strong theory, though they may include one:
-1. A theory is separate from the case or cases being examined. It is intented to be general.
+1. A theory is separate from the case or cases being examined. It is intended to be general.
 2. A theory generates new statements that in principle can be empirically falsified, not proven.
@@
 \setcounter{table}{0}
+
+## Code references for coding errors
+
+### Section 3.2: Incorrect “Weighted by PITF” Implementation
+
+The intention with the "Weighted by PITF" model is to combine the PITF test predictions with the Escalation test predictions, i.e.
+weight the latter with the former.
+Due to an error, the PITF _training_ set predictions are used, not the _test_ set predictions.
+
+See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`).
+The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions.
+Similarly, in the 6-month version of this file.
+
+Additional details in issue #3 in the GitHub repo.
+
+### Section 3.3: Incorrect “PITF Split Population” Implementation
+
+The intention is to train seperate random forests on data split into high and low-risk groups based on PITF predictions, and then combine them back into one random forest.
+Actually both random forests are trained on identical data, essentially production a random forest that just has N_trees \* 2.
+
+Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined.
+Details are documented in issue #5 in the GitHub repo.
+
+### Section 3.5: Incorrect scoring of the 2016 forecasts
+
+The relevant variables in the data are "incidence_civil_ns" and "incidence_civil_ns_plus1", which appears to be a 1-period lead (ie., $t_+1$) version of the dependent variable that is used in the actual prediction models.
+The incidence dependent variable  contains both 0/1 and missing values.
+By examining the pattern of missing values, it seems clear that this was originally an incidence variable indicating whether a country was at civil war in a given year or not, and which was converted to an onset version so that onsets retain the value of 1 but continuing civil war years are coded as missing.
+This reflects common practice.
+
+However, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
+In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable to incidence.
 
@@
 
+Only the "binormal" smoothing method produces a pattern of AUC-ROCs that clearly elevates the Escalation model above the alternatives.
+The "density" method does not on average change the AUC-ROC values very much from their empirical ROC curve equivalents.
+The "fitdistr" method tends to increase all AUC-ROC values, although maybe slightly more so for the Escalation model than the main alternatives.
+
 ```{r smooth-benefit-plot-extended, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2), with 3 different smoothing methods (binormal is the default method and used in B\\&S)"}

### 2020-09-21 paper.Rmd
cut paragraphs from conclusion

@@
 
-#An important consideration when developing a predictive model is to do so without contaminating the out-of-sample test set.
-This occurs if test set performance is used to inform specification---predictive modeling's analogue to p-hacking---and special care is needed [e.g.
-see the approach by @colaresi:mahmood:2017].
-B&S did pre-register their 2016-H1 forecasts, which verifies the integrity of the forecasts, but this does not assure the model development process.
-To be clear: there is no reason to suspect test set contamination in the escalation or other models.
-Rather, it is challenging to provide verifiable assurance thereof in a single analysis iteration that both develops and tests a model on data that was available contemporaneously.
-
-#Moreover, they misunderstand the use of ICEWS event data in current research.
-They claim that most applications to date have focused on the quad categories, but this ignores a wide swath of literature [@steinert-threlkeld:2017; and @metternich:etal:2013] that uses a specific action---such as protest---defined in the CAMEO ontology.
-In their article on conflict in Thailand, Metternich et alia hand-coded, for example, every actor in Thailand and focused on an analysis of how those interact.
-
 We encountered several issues in the code underlying the B&S analysis.

### 2020-09-21 paper.Rmd
work on discussion and conclusion
still in progress

@@
 
-# Conclusion
+# Discussion
 
@@
 They are weak because they do not include two characteristics of strong theory, though they may include one:
+
 1. A theory is separate from the case or cases being examined. It is intended to be general.
@@
 nonviolent repression, and low-level violence.
+
 \begin{itemize}
@@
 \end{itemize}
+
 These variables, taken from ICEWS, are the main components of the escalation model in the B\&S article.
@@
 In practical terms there is a wide array of choices that are available to instantiate any particular theory.
-A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated temporal dynamics.
+A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated dynamics.
 One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.
-At this level of detail the theoretical model is more about chosing the correct variables.
+
+At this level of detail the theoretical model is more about choosing the correct variables, and this is the contrast between theoretical prediction models and a-theoretical machine learning that B&S focus on.
+But this dichotomy is misguided.
+The data that are used in a predictive model have to come from somewhere.
+Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful.
+It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well.
+Rather, we posit that most of the cases B&S might characterize as a-theoretical enter the "theoretical" process at step 1.5 above, with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.
+
 And while we don't undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.
+
+Another way in which we could contrast theoretical from a-theoretical prediction models is in terms of their flexibility in regard to functional form.
+Theory should not only tell us what factors matter, but also how they relate to each other.
+We think that few would be strongly committed to the idea that the linear additive equations, maybe with the occasional polynomial or multiplicative interaction terms, that dominate published regression analyses, are *the* correct functional specification.
+Probably it's more complex, and probably we don't really know what a better specification would be.
+Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low?
+Are whatever those relationships are linear in the log-odds?
+Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships.
+B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives.
+
+Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider the ways in which they can mutually reinforce each other.
+Even if the analysis in B&S fails to support the point, we concede that predicting on the basis of a strong theoretical model is preferable to inductive prediction.
+Beyond accurate predictions, such models would allow us to make causal statements that are harder to make on the basis ad hoc models.
+But this is hard to do in practice.
+The problem is that many extant theoretical models are in fact so malleable to interpretation that they essentially are ad hoc.
+They are useful for identifying variables that could be useful for prediction, and this is in fact already taking place.
+
+WHAT CAN ML MODELS TELL US ABOUT THEORY? PRUNING VARIABLES, FUNCTIONAL FORM, INTERACTIONS, PREDICABILITY
 
@@
 Nor, against it.
+
+# Conclusion
 
@@
 - Adding structural variables substantially improves the escalation model's performance.
+
+More importantly, trying to contrast machine learning and theory is misguided.
+Even if a "theory" section is missing, predictive models are informed by previous work and arguments.
+If we had strong general theories of something like civil war onset, we could replace more inductive, ad hoc, methods with cleary specified models.
+But we do not.
+What we instead have are weak theories that could be interpreted to suggest any range of equations for predicting an outcome.
+Instead of trying to find an instantiation of such a "theoretical" model that can outperform a more ad hoc alternative, we could use the flexibility of machine learning methods and insights that predictive modeling brings to improve our theories.
 

### 2020-09-21 paper.Rmd
next to last on Monday

@@
 ---
-title: "Reassessing the role of theory and machine learning in forecasting civil conflict"
+title: "Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict"
 author: 
@@
 Dorff, and Shahryar Minhas both provided helpful feedback on this project.
+We are especially grateful to Paul Huth for his comments and guidance.
 All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
@@
 Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.
-While their model is not very theoretical, it is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.
+Their model is s outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.
 output: 
@@
 Indeed they assert that with such theory forecasting is more accurate than has previously been possible.
-Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in support of it.
+We re-examine the empirical basis for the claims made in support of it.
 We find that these claims to be unsupported and evidence presented for them to be incorrect.
 Their theory-based escalation model does not do better than the alternatives that they examine.
-Actually, it does worse.
+It does worse.
 The reason for this is that they have made several mistakes in their research procedure.
-Further, the performance results they report are based on smoothed performance curves, not the original unsmoothed curves.
+Further, the performance results they report are based on smoothed performance curves, not the empirical, unsmoothed curves.
 This provides misleading results.
 In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented.
-We also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset.
+We also found that the scoring of their forecasts for the first half of 2016 was incorrectly performed using civil war incidence, not onset.
 In what follows, we show the impact of these mistakes on the conclusions.
@@
 B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately".
-Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.\footnote{We set aside for here the characteristics of theoretical models, accepting the B\&S denotation.}  They also claim that the integration of structure with process is better than alternaties over short forecasting windows.
+Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.They also claim that the integration of structure with process is better than alternaties over short forecasting windows.
 Third, they preregistered the list of thirty countries that have the highest risk of civil war onset.
-They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.
-B\&S claim to be unique in assessing these forecasts.
+They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.\footnote{\cite{PITF},\cite{ICEWS},\cite{VIEWS}\cite{others}} B\&S claim to be unique in assessing these forecasts.
 A qualitative analysis of their predictions allows them to conclude that their model is robust.
@@
 
-1. Split the training data into training (2001 - 2007) and test (2008 - 2015) sets.
-2. Estimate the escalation and other competing models. 
+1. Split the training data into training (2001 - 2007) and test (2008 - 2015) data.
+2. Estimate the escalation and other competing models with the training data. 
 3. Create out-of-sample (OOS) predictions from each model using the test set. 
@@
 The corresponding results for each question are shown in B&S Tables 1 and 2, which we examine further below.
-We accurately replicate their Tables 1 and 2, with very tiny differences.
+We accurately replicate their Tables 1 and 2.
 The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions.
@@
 We believe that these research decisions and issues lead B\&S to incorrect conclusions.
-The escalation model is not the best, and it actually performs worse than the atheoretical, garbage can model with over \(1000\) variables.
+The escalation model is not the best; it actually performs worse than the atheoretical, ktchen sink model with over \(1000\) variables.
 We discuss these five issues below.
@@
 The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S.
-The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey.} (Both of the ambiguous cases include Blair as coauthor.) The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.
+The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey.
+Both of the ambiguous cases include Blair as coauthor.} The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.
 What is surprising is that anyone would choose the smoothed ROC implementation.
@@
 
-The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015,beger:etal:2017].
-Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision trees.
+The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015,@beger:etal:2017].
+Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger RF using both component model's underlying decision trees.
 Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does have 200,000 decision trees.
 Furthermore, the PITF model predictions do not impact how the combined RF model predicts at all, not even through a binary low-/high-risk split.
-The split-population PITF RF model is practically speaking just another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree).
+The split-population PITF RF model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree).
 This does not adequately implement their split-sample modeling strategy.
@@
 
-We now turn to an examination that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/andybega/Blair-Sambanis-replication.] The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B&S Table 2).
+We now turn to an examination that addresses and fixes the five issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/andybega/Blair-Sambanis-replication.] The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B&S Table 2).
 We will review the substantive implications of our updated analysis below, but the bottom line is that these changes turn B&S's conclusions on their heads.
@@
 An example might be William Riker's theory of coalitions [@riker:1962] which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically.
-A more recent example is found in The Political Logic of Survival [@bdm:etal:2005].
+A more recent example of strong theory is found in *The Political Logic of Survival* [@bdm:etal:2005].
 For the purposes of discussion we call this strong theory.
@@
 
-At this level of detail the theoretical model is more about choosing the correct variables, and this is the contrast between theoretical prediction models and a-theoretical machine learning that B&S focus on.
+At this level of detail the theoretical model is more about choosing the correct variables and this is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on.
 But this dichotomy is misguided.
@@
 It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well.
-Rather, we posit that most of the cases B&S might characterize as a-theoretical enter the "theoretical" process at step 1.5 above, with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.
+Rather, we posit that most of the cases B&S might characterize as a-theoretical enter the "theoretical" , with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}
 
-And while we don't undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.
-
-Another way in which we could contrast theoretical from a-theoretical prediction models is in terms of their flexibility in regard to functional form.
+Another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form.
 Theory should not only tell us what factors matter, but also how they relate to each other.
 We think that few would be strongly committed to the idea that the linear additive equations, maybe with the occasional polynomial or multiplicative interaction terms, that dominate published regression analyses, are *the* correct functional specification.
-Probably it's more complex, and probably we don't really know what a better specification would be.
 Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low?
@@
 
-WHAT CAN ML MODELS TELL US ABOUT THEORY? PRUNING VARIABLES, FUNCTIONAL FORM, INTERACTIONS, PREDICABILITY
-
 Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning.
-Even if B\&S were correct that in their study the theory based escalation model performed better, it would simple be a case study of one case where it was preferable.
+Even if B\&S were correct that in their study the theory based escalation model performed better, it would simply be a case study of a single case where it was preferable.
 It establishes no general tendency, let alone law-like evidence for general theory.
@@
 
-More importantly, trying to contrast machine learning and theory is misguided.
+More importantly, trying to contrast machine learning and theory may be misguided.
 Even if a "theory" section is missing, predictive models are informed by previous work and arguments.
@@
 But we do not.
-What we instead have are weak theories that could be interpreted to suggest any range of equations for predicting an outcome.
-Instead of trying to find an instantiation of such a "theoretical" model that can outperform a more ad hoc alternative, we could use the flexibility of machine learning methods and insights that predictive modeling brings to improve our theories.
+What we have instead are weak theories that could be interpreted to suggest any range of equations for predicting an outcome.
+Instead of trying to find an instantiation of such a "theoretical" model that can outperform a more ad hoc alternative, we could use the flexibility of machine learning methods and insights that predictive modeling brings to improve our weak theories.
 
@@
 However, these comparisons must be made carefully to ensure those correct inferences are drawn.
-We continue to think that theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things.
+We continue to think that weak theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things.
 It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.

### 2020-09-21 paper.Rmd
final monday change
andy check several places herein where I ask you to clarify. Over to you.

@@
 All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
-abstract: We examine the research protocols in Blair and Sambanis (2020).
+abstract: We examine the research protocols in Blair and Sambanis (nd).
 We find that there are several important mistakes and research decisions that determine their results.
@@
 B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately".
-Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.They also claim that the integration of structure with process is better than alternaties over short forecasting windows.
+Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.They also claim that the integration of structure with process is better than alternatives over short forecasting windows.
 Third, they preregistered the list of thirty countries that have the highest risk of civil war onset.
@@
 This appears to be an easily corrected coding error on the part of B&S.
+ANDY ARE YOU SURE THIS PARAGRAPH IS CORRECT?
+DON'T YOU WANT TO WEIGHT THE TEST BY THE TRAINING?
 
@@
 The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself.
-If we ran both with the same random seed, they would be identical in every respect, producing to forests of identical decision trees and, thus, identical predictions.
+If we ran both with the same random seed, they would be identical in every respect, producing two forests of identical decision trees and, thus, identical predictions.
 
@@
 When forecasting rare events, it is common to have many false positives, i.e.
-instances where we predict onset but not onsets occur.
+instances where we predict onset but no onsets occur.
 _Ex ante_, using incidence rather than onset to score onset forecasts should decrease the number of false positives and correspondingly increase the number of true positives (good predictions) and false negatives (missed onsets).
@@
 The vertical bar in each plot marks the average effect of smoothing on AUC-ROC.
-For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible in average.
+For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible on average.
 The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05.
@@
 
-B&S interpret their results as follows, on page 20 and we comment on the conclusions seratim:^[We list the "Overall, ..." interpretation out-of-order for clarity.]
+B&S interpret their results as follows (page 20) and we comment on the conclusions seratim:^[We list the "Overall, ..." interpretation out-of-order for clarity.]
 
@@
 
+ANDY, THIS SECTION 4.6 AND THE TABLE 4 ARE VERY CONFUSING. CAN YOU TRY TO CLEAN IT UP?
 The correct confusion matrices when using observed onset (or the lack of it) are shown in the second part of Table \ref{tab:table4}.
@@
 B\&S advocate the use of theory to guide prediction.
-But "theory" is not defined by B\&S, and may an ambiguous and undefined concept.
+But "theory" is not defined by B\&S and may be an ambiguous and undefined concept.
 `Theory'' has many interpretations in social science.
@@
 It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well.
-Rather, we posit that most of the cases B&S might characterize as a-theoretical enter the "theoretical" , with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}
+Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical" , with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}
 
@@
 Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low?
-Are whatever those relationships are linear in the log-odds?
+Are those relationships linear in the log-odds?
 Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships.
@@
 It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.
-Social science has leaned on theory for (at least) seventy years, but it might be time to find useful ways to incorporate some of the recently developed methods in the arena of machine learning as well.
+Social science has leaned on theory for (at least) seventy years, but it might be time to find useful ways to incorporate some of the recently developed methods in the arena of machine learning and modern statistics as well.
 

### 2020-09-22 paper.Rmd
drop in references

@@
 Third, they preregistered the list of thirty countries that have the highest risk of civil war onset.
-They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.\footnote{\cite{PITF},\cite{ICEWS},\cite{VIEWS}\cite{others}} B\&S claim to be unique in assessing these forecasts.
+They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.^[PITF [e.g.
+@goldstone:etal:2010; @beger:ward:2017], the W-ICEWS project [@ward:etal:2013], VIEWS [@hegre:etal:2019], and others, e.g.
+@beger:etal:2020.] B\&S claim to be unique in assessing these forecasts.
 A qualitative analysis of their predictions allows them to conclude that their model is robust.
@@
 But "theory" is not defined by B\&S and may be an ambiguous and undefined concept.
-`Theory'' has many interpretations in social science.
+"Theory" has many interpretations in social science.
 For some---the EITM movement, for example [@aldrich:etal:2008]---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.
@@
 It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well.
-Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical" , with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}
+Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical", with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}
 
@@
 Theory should not only tell us what factors matter, but also how they relate to each other.
-We think that few would be strongly committed to the idea that the linear additive equations, maybe with the occasional polynomial or multiplicative interaction terms, that dominate published regression analyses, are *the* correct functional specification.
+We think that few would be strongly committed to the idea that the linear additive equations that dominate published regression analyses, maybe with the occasional polynomial or multiplicative interaction terms, are *the* correct functional specification.
 Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low?
@@
 B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives.
+But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory?
 

### 2020-09-22 paper.Rmd
Redo the confusion matrices in LaTeX

@@
 This appears to be an easily corrected coding error on the part of B&S.
-ANDY ARE YOU SURE THIS PARAGRAPH IS CORRECT?
-DON'T YOU WANT TO WEIGHT THE TEST BY THE TRAINING?
 
@@
 
-B&S report (in their Table 4) confusion matrices for their forecasts for the first half of 2016 (2016-H1).
-To create the confusion matrices, B&S treat the 30 highest ranked predictions as positive predictions ("1s") and the rest as negative predictions ("0s").
-We replicate their Table 4 in the top portion of \ref{tab:table4}.
-There are two confusion matrices for slightly different codings of outcomes in 2016-H1, under the corresponding "Assuming Persistence" and "Assuming Change" headings.
-We should note that the "Assuming Persistence" corresponds to the values in the replication data; the small variations for the "Assuming Change" version are hand-coded in the replication code file that generates the confusion matrices and appear to be subjective assessments of B&S.
+B&S made forecasts for the first half of 2016 (2016-H1), and assess the accuracy of these forecasts.
+The forecasts themselves are probabilistic, i.e.
+range from 0 to 1, but for the purpose of scoring the forecasts they consider the 30 highest predictions to have been positive forecasts ("1") and the remaining forecasts to have been for no onset ("0").
+They then construct confusion matrices for the forecasts, in their Table 4 and which we replicate here in the top half of our Table \ref{tab:table4}.
+They use two alternative codings of the outcomes for 2016-H1, one of which they call "Assuming Persistence" (this is what is in their replication data), and one of which they call "Assuming Change" (hand-coded in one of their replication scripts).
 
-We can see that the original table presents 15 or 16 positive cases for 2016-H1, depending on the dependent variable coding variation.
-This corresponds to a positive rate of around 9.5% for the first half of 2016 data.
-In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a much lower positive rate of around 0.5%, i.e., an order of magnitude smaller.
-The positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data.
-This suggests that the forecasts were erroneously assessed using civil war incidence, not onset.
-By examining the replication code we were able to verify that the forecasts were scored using civil war incidence, i.e., including ongoing civil wars, rather than civil war onset years only.
+With the "Assuming Persistence" version of the outcomes, shown in the first confusion matrix there are 15 civil war onsets, of which the forecasts capture 13 (true positives) and miss 2 (false negatives).
+With the "Assuming Persistence" coding of outcomes, shown in the second confusion matrix, there are 16 onsets and the forecasts capture 14 of those with, still, 2 misses.
+Fifteen to 16 civil war onsets in the first half of 2016 is incredible, especially considering that the entire historical data from 2001 to 2015 only has 20 recorded civil war onsets.
+Indeed, it turns out that B&S scored their forecasts for civil war \textit{onset} using civil war \textit{incidence}, i.e.
+counting ongoing civil wars as if they were continuing onsets.
 
-```{r, message=FALSE}
+```{r, include=FALSE}
+# AB 2020-09-22: I redid the confusion matrices in latex so that the resulting table is clearer than what I could easily do with kable() programatically here in this chunk. 
 tab4o <- read_csv("data/table4-original.csv")
@@
 
-ANDY, THIS SECTION 4.6 AND THE TABLE 4 ARE VERY CONFUSING. CAN YOU TRY TO CLEAN IT UP?
-The correct confusion matrices when using observed onset (or the lack of it) are shown in the second part of Table \ref{tab:table4}.
-In the default "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1.
-Thus, the recall value is undefined, while the precision is 0/30 = 0, compared to reported recall and precision values of 13/15 = 0.87 and 13/30 = 0.43.
-The alternative coding ("Assuming Change") produces two civil war onsets.
-Recall is 1.0 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.^[Another, minor issue is related to using a lead version of the DV.
-With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 indicates the risk of civil war onset in 2016-H2.
-In the Table 4 script referenced above, the 2016-H1 predictions (for 2016-H2) are assessed using the raw DV, "incidence_civil_ns", not the lead version.
-Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1.
-In this case, it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.]
+\begin{table}
+\caption{Replication of B\&S Table 4: Confusion Matrices for Six-month Escalation Model forecasts for the first half of 2016.}\label{tab:table4}
+\centering
+\begin{tabular}{ll|c|c|l}
+\multicolumn{5}{c}{\textbf{Original confusion matrices with civil war \textit{incidence}}} \\
+% BS, assuming persistence
+\multicolumn{1}{l}{\multirow{4}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
+\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
+\cline{3-4}
+& $Y=0$ & 132 & 17 \\
+\cline{3-4}
+& $Y=1$ & 2 & 13 \\
+\cline{3-4}
+% BS, assuming change
+\multicolumn{1}{l}{\multirow{4}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
+\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
+\cline{3-4}
+& $Y=0$ & 132 & 16 \\
+\cline{3-4}
+& $Y=1$ & 2 & 14 \\
+\cline{3-4}
+\\ \midrule
+\multicolumn{5}{c}{\textbf{Fixed confusion matrices with civil war \textit{onset}}} \\
+% fixed, assuming persistence
+\multicolumn{1}{l}{\multirow{4}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
+\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
+\cline{3-4}
+& $Y=0$ & 134 & 30 \\
+\cline{3-4}
+& $Y=1$ & 0 & 0 \\
+\cline{3-4}
+% fixed, assuming change
+\multicolumn{1}{l}{\multirow{4}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
+\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
+\cline{3-4}
+& $Y=0$ & 134 & 28 \\
+\cline{3-4}
+& $Y=1$ & 0 & 2 \\
+\cline{3-4}
+\end{tabular}
+\end{table}
 
-Predicting rare events like civil war onset is exceedingly difficult, and it is not unusual at all that most of the predictions are false positives.
-This will by definition be the case when there are no onsets in a time period, as is true with the default "Assuming Persistence" coding that is in the data.
-Correctly predicting the two onsets in the alternative "Assuming Change" coding is at face value a credit to the model's accuracy.
+The correct confusion matrices that use civil war onsets to score the forecasts are shown in the second part of Table \ref{tab:table4}.
+In the "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1.
+The alternative "Assuming Persistence" coding records 2 civil war onsets in 2016-H1.
+As a result, if we compare the precision and recall of the forecasts with the correct scoring to the incorrect scoring in B&S, almost all of the forecasts are false positives.
+This is the norm when predicting rare events, and thus not unusual at all.
+The forecasts do correctly capture the 2 civil war onsets in the "Assuming Change" version of the outcome coding.
+This is, at face value, a credit to the model's accuracy.
 
@@
 Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical", with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}
+
+Conversely, not all factors that are "theoretically" important are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory.
+For example, both @wgb:2007 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes.
 
@@
 
-The intention with the "Weighted by PITF" model is to combine the PITF test predictions with the Escalation test predictions, i.e.
+We believe the intention with the "Weighted by PITF" model is to combine the PITF test predictions with the Escalation test predictions, i.e.
 weight the latter with the former.
@@
 The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions.
-Similarly, in the 6-month version of this file.
+Similarly in the 6-month version of this file.
 

### 2020-09-22 paper.Rmd
expand discussion; done for now

@@
 
-Conversely, not all factors that are "theoretically" important are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory.
+Furthermore, if what we mean with "theory" is such that it suggests a large range of variables that could be related to an outcome, what do we really gain from having this theory in our predictive model?
+We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off?
+Certainly that's not what B&S are trying to suggest since this would be a problematic way to build a theory.
+But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could post fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and any associated "theory" with any eye towards out-of-sample predictive accuracy [cf.
+mahmood colarsei paper on how to do this properly though].
+Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize, and whose underlying theories are by extension not externally valid [zones farris paper].
+
+Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory.
 For example, both @wgb:2007 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes.
+And it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.
+the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups.^[Andreas Beger and Michael D.
+Ward, 2020, “Coup forecasts for 2020”, https://www.predictiveheuristics.com/forecasts]
 
-Another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form.
+% MIKE: is it actually true that infant mortality has mostly not been studied in the context of explaining conflict risk? Goldstone in their article claim that it has, but for me that's news.
+
+Aside from selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form.
 Theory should not only tell us what factors matter, but also how they relate to each other.

### 2020-09-22 paper.Rmd
finished for the day, this time it's true

@@
 However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.
+
+As a result: (1) in the test set, all countries with names that alphabetically start after "Tongo" are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical.
+For example, the "Weighted by PITF" model predictions for Syria from 2008 to 2015 are the Escalation test predictions for that time period weighted, respectively, by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.
+
 This appears to be an easily corrected coding error on the part of B&S.
@@
 We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off?
-Certainly that's not what B&S are trying to suggest since this would be a problematic way to build a theory.
-But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could post fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and any associated "theory" with any eye towards out-of-sample predictive accuracy [cf.
-mahmood colarsei paper on how to do this properly though].
-Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize, and whose underlying theories are by extension not externally valid [zones farris paper].
+Certainly that's not what B&S are trying to suggest since this would be a problematic way to "evaluate" a theory.
+But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could post fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated "theory" with any eye towards out-of-sample predictive accuracy [cf.
+@colaresi:mahmood:2017 for how to avoid this problem while leveraging out-of-sample predictions].
+Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize, and whose underlying theories may by extension not be externally valid [for a more detailed treatment of this issue, see @fariss:jones:2018].
 
 Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory.
-For example, both @wgb:2007 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes.
+For example, both @ward:etal:2010 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes.
 And it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.
@@
 B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives.
-But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory?
+But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory [see also @fariss:jones:2018]?
 
@@
 
-Additional details in issue #3 in the GitHub repo.
+Uncovering which cases in the PITF training predictions are matched to the cases that the Escalation model test set predictions correspond to requires tracing back several more data objects.
+We have documented details in issue #3 in the GitHub repo.
 

### 2020-09-22 paper.Rmd
ready to call this final for proofing

@@
 
-The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015,@beger:etal:2017].
+The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015;@beger:etal:2017].
 Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger RF using both component model's underlying decision trees.
@@
 It is not a cosmetic change as suggested.
-
-## The Effect of Using Smoothed ROC Curves
 
@@
 
-Furthermore, if what we mean with "theory" is such that it suggests a large range of variables that could be related to an outcome, what do we really gain from having this theory in our predictive model?
+Furthermore, if what we mean by the concept "theory" is such that it suggests a large range of variables that could be related to an outcome, what do we really gain from having this theory in our predictive model?
 We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off?
 Certainly that's not what B&S are trying to suggest since this would be a problematic way to "evaluate" a theory.
-But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could post fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated "theory" with any eye towards out-of-sample predictive accuracy [cf.
+But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could in fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated "theory" with any eye towards out-of-sample predictive accuracy [cf.
 @colaresi:mahmood:2017 for how to avoid this problem while leveraging out-of-sample predictions].
-Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize, and whose underlying theories may by extension not be externally valid [for a more detailed treatment of this issue, see @fariss:jones:2018].
+Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize and whose underlying theories may by extension not be externally valid [for a more detailed treatment of this issue, see @fariss:jones:2018].
 
@@
 Ward, 2020, “Coup forecasts for 2020”, https://www.predictiveheuristics.com/forecasts]
-
-% MIKE: is it actually true that infant mortality has mostly not been studied in the context of explaining conflict risk? Goldstone in their article claim that it has, but for me that's news.
 

### 2020-09-23 paper.Rmd
change abstract and intro

@@
 All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
-abstract: We examine the research protocols in Blair and Sambanis (nd).
-We find that there are several important mistakes and research decisions that determine their results.
+abstract: We examine the research protocols in Blair and Sambanis (nd), who argue that their theory-based model can predict civil war onsets better than several atheoretical alternatives or a model with country-structural factors.
+We find that there are several important mistakes and that their key finding is entirely conditional on the use of parametrically smoothed ROC curves when calculating accuracy, rather than the standard empirical ROC curves that dominate the literature.
 Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.
-Their model is s outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.
+Their model is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.
+More importantly, we argue that rather than trying to contrast the roles of theory and \"atheoretical\" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant theories.
 output: 
@@
     toc: false
-#header-includes:
+header-includes:
+  - \usepackage{floatrow}
+  - \floatsetup[figure]{capposition=bottom}
+  - \usepackage[format=plain,labelfont=bf,up]{caption}
 #  - \usepackage{setspace}\doublespacing
@@
 [...]  Cederman and Weidmann (2017, 476) argue that "the hope that big data will somehow yield valid forecasts through theory-free 'brute force' is misplaced in the area of political violence." Our results lend some credence to this claim.
+
+Although their analysis in the end does not support this point, we do not disagree that theory is important in predictive modeling.
+Of the various ways one could contrast ad hoc machine learning with "theoretical" modeling, B&S focus essentially on theory as a mechanism for variable selection.
+In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting.
+The problem is that the space of extant theories for a phenomenon like civil war is so broad and non-specific in terms of an exact specification of what variables and in which functional form should constitute a prediction model, that it's both not very clear what the "correct" model should be, nor easy to demonstrate that any particular "theoretical" model's ability to predict generalizes beyond the data that was used to create it in the first place.
+Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to "explain" a phenomenon to strengthen and whittle down the broad space of extant theories.
 
@@
 A strong theory gives more guidance on compelling choices than does a weak theory.
-Accordingly, while the choice is plausible is not necessarily compelling.
+Accordingly, while the choice is plausible it is not necessarily compelling.
 Another choice might be equally plausible.
@@
 And it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.
-the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups.^[Andreas Beger and Michael D.
-Ward, 2020, “Coup forecasts for 2020”, https://www.predictiveheuristics.com/forecasts]
+the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups [@beger:ward:2020].
 
@@
 Even if the analysis in B&S fails to support the point, we concede that predicting on the basis of a strong theoretical model is preferable to inductive prediction.
-Beyond accurate predictions, such models would allow us to make causal statements that are harder to make on the basis ad hoc models.
+Beyond accurate predictions, such models would allow us to make causal statements that are harder to make on the basis of ad hoc models.
 But this is hard to do in practice.
@@
 Even if a "theory" section is missing, predictive models are informed by previous work and arguments.
-If we had strong general theories of something like civil war onset, we could replace more inductive, ad hoc, methods with cleary specified models.
+If we had strong general theories of something like civil war onset, we could replace more inductive, ad hoc, methods with clearly specified models.
 But we do not.

### 2020-09-24 paper.Rmd
edits

@@
 Blair and Sambanis [-@blair:sambanis:2020, hereafter B\&S] argue that theory is essential for creating models that have high accuracy in forecasting civil war onset.
-Indeed they assert that with such theory forecasting is more accurate than has previously been possible.
+Indeed, they assert that with such theory forecasting is more accurate than has previously been possible.
 We re-examine the empirical basis for the claims made in support of it.
@@
 B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately".
-Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.They also claim that the integration of structure with process is better than alternatives over short forecasting windows.
+Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.
+They also claim that the integration of structure with process is better than alternatives over short forecasting windows.
 Third, they preregistered the list of thirty countries that have the highest risk of civil war onset.
-They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.^[PITF [e.g.
-@goldstone:etal:2010; @beger:ward:2017], the W-ICEWS project [@ward:etal:2013], VIEWS [@hegre:etal:2019], and others, e.g.
-@beger:etal:2020.] B\&S claim to be unique in assessing these forecasts.
+They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.^[PITF [e.g., @goldstone:etal:2010; @beger:ward:2017], the W-ICEWS project [@ward:etal:2013], VIEWS [@hegre:etal:2019], and others, e.g., @beger:etal:2020.] B\&S claim to be unique in assessing these forecasts.
 A qualitative analysis of their predictions allows them to conclude that their model is robust.
@@
 While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation.
-These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) incorrect implementations of the weighted by PITF and PITF analog split-population models, c) inconsistent test sets for the models examined, and d) incorrect scoring of the 2016 forecasts.^[We also note there are additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox.
+These are (a) the use of smoothed ROC curves to draw conclusions about which model is best, (b) incorrect implementations of the weighted by PITF and PITF analog split-population models, (c) inconsistent test sets for the models examined, and (d) incorrect scoring of the 2016 forecasts.^[We also note there are additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox.
 We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.]
@@
 We believe that these research decisions and issues lead B\&S to incorrect conclusions.
-The escalation model is not the best; it actually performs worse than the atheoretical, ktchen sink model with over \(1000\) variables.
+The escalation model \textit{is not} the best; it actually performs worse than the atheoretical, kitchen sink model with over \(1000\) variables.
 We discuss these five issues below.
@@
 
+<!-- [RM: I'm not sure what this sentence is doing in this paragraph.
+It was the second sentence in the paragraph below] The data in rare events problems, like this one, restrict the number of true positive rate values and lead to non-continuous ROC curves;  -->
+
 The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves.
-The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation.
+B&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation.
 However, smoothing is also used in all AUC-ROC values they report.
@@
 In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.
+
 The model specifications are also identical; i.e., they use the same *x* variables and the same random forest hyper-parameter settings.
@@
 The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015;@beger:etal:2017].
-Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger RF using both component model's underlying decision trees.
-Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does have 200,000 decision trees.
-Furthermore, the PITF model predictions do not impact how the combined RF model predicts at all, not even through a binary low-/high-risk split.
-The split-population PITF RF model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree).
+Although the two random forest models are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger random forest using both component model's underlying decision trees.
+Thus, while all random forest models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new random forest model after combination does have 200,000 decision trees.
+Furthermore, the PITF model predictions do not impact how the combined random forest model predicts at all, not even through a binary low-/high-risk split.
+The split-population PITF random forest model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing random forest randomization facilities (row and column sampling for each decision tree).
 This does not adequately implement their split-sample modeling strategy.
@@
 
-Further, the AUC-ROC values reported in the original B&S Table 2 are calculated  on slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}).
+Further, the AUC-ROC values reported in the original B&S Table 2 are calculated on different numbers of underlying test set cases (see Table \ref{tab:table2-N}).
 ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available.
@@
 
-> Figure 1 displays the corresponding ROC curves, smoothed for ease of interpretation.
+> Figure 1 displays the corresponding ROC curves, _smoothed for ease of interpretation_. (emphasis added)
 
@@
 However, this is not the case.
-All AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves *prior* to AUC calculation.
+All AUC-ROC calculations throughout their replication code use an option to smooth the ROC curves *prior* to AUC calculation.
+
 ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like.
@@
 In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for an observed positive case is reached.
-For these ROC curves, and for that matter in the fundamental train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set.
+For these ROC curves, and in the fundamental train/test split used for 12 of the 18 rows/models in B&S Table 1, for that matter, there are only 11 civil war onset cases in the test set.
 Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates.
@@
 For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible on average.
-The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05.
+The Escalation model, however, _always_ benefits from smoothing, with an average improvement on the order of 0.05.
 This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B&S report, namely that the escalation model is generally superior.
@@
 
-In sum, it is not only the case that using smoothed ROC curves alters the results, but also that the use of smoothed ROC curves to calculate AUC-ROC values benefits \textbf{only} the escalation model.
+In sum, it is not only the case that using smoothed ROC curves alters the results, but also that the use of smoothed ROC curves to calculate AUC-ROC values \textbf{only} benefits the escalation model.
 It does so consistently and by a considerable margin.
@@
 
-The PITF split-population analog still performs well, but the simple Escalation + PITF predictors model arguably performs better still.
+The PITF split-population analog still performs well, but the simple Escalation + PITF predictors model performs better still.
 
@@
 
-With the "Assuming Persistence" version of the outcomes, shown in the first confusion matrix there are 15 civil war onsets, of which the forecasts capture 13 (true positives) and miss 2 (false negatives).
-With the "Assuming Persistence" coding of outcomes, shown in the second confusion matrix, there are 16 onsets and the forecasts capture 14 of those with, still, 2 misses.
-Fifteen to 16 civil war onsets in the first half of 2016 is incredible, especially considering that the entire historical data from 2001 to 2015 only has 20 recorded civil war onsets.
+With the "Assuming Persistence" version of the outcomes, shown in the first confusion matrix, B\&S report that there are 15 civil war onsets, of which the forecasts capture 13 (true positives) and miss 2 (false negatives).
+With the "Assuming Persistence" coding of outcomes, shown in the second confusion matrix, they report that there are 16 onsets and the forecasts capture 14 of those with, still, 2 misses.
+Fifteen to 16 civil war _onsets_ in the first half of 2016 is incredible, especially considering that the entire historical data from 2001 to 2015 only has 20 recorded civil war _onsets_.
 Indeed, it turns out that B&S scored their forecasts for civil war \textit{onset} using civil war \textit{incidence}, i.e.
-counting ongoing civil wars as if they were continuing onsets.
+counting ongoing civil wars as if they were onsets.
 
@@
 \\ \midrule
+\\
 \multicolumn{5}{c}{\textbf{Fixed confusion matrices with civil war \textit{onset}}} \\
@@
 In the "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1.
-The alternative "Assuming Persistence" coding records 2 civil war onsets in 2016-H1.
+The alternative "Assuming Persistence" coding records 2 civil war onsets in 2016-H1, according to the civil conflict dataset they rely on.
 As a result, if we compare the precision and recall of the forecasts with the correct scoring to the incorrect scoring in B&S, almost all of the forecasts are false positives.
@@
 
-However, for many, a theory is a falsifiable story or narrative that makes causal claims about the relationships among variables.
+However, for many, a theory is simply a falsifiable story or narrative that makes causal claims about the relationships among variables.
 Typically it is fused together from a meta-analysis of prior work on the topic under consideration.
@@
 Following a wide swath of scholars, they argue that the escalation and de-escalations of interactions between citizens and government are what typically lead up to the onset of civil wars.
-Operationally, this boils down to four types of concepts: demands, accommodations,
-nonviolent repression, and low-level violence.
+Operationally, this boils down to four types of concepts: demands, accommodations, nonviolent repression, and low-level violence.
 
 \begin{itemize}
-\item Demands include events in which either opposition or rebels is the source and
-government is the target and typically involve appeals for political or institutional
-reform, policy change, or extension of rights.
-\item Nonviolent repression and accommodation include events in which government
-is the source and either opposition or rebels is the target. Nonviolent
-repression includes threats, sanctions, restrictions on freedom of movement, or
-rejections of group demands.
-\item Low-level violence includes all events in which
-(1) government is the source and opposition is the target, (2) opposition is the
-source and government is the target, (3) government is the source and rebels are
-the target, or (4) rebels are the source and government is the target.
+\item Demands include events in which either opposition or rebels is the source and government is the target and typically involve appeals for political or institutional reform, policy change, or extension of rights.
+\item Nonviolent repression and accommodation include events in which the government is the source and either opposition or rebels is the target.
+Nonviolent repression includes threats, sanctions, restrictions on freedom of movement, or rejections of group demands.
+\item Low-level violence includes all events in which (1) the government is the source and opposition is the target, (2) the opposition is the source and government is the target, (3) the government is the source and rebels are the target, or (4) rebels are the source and government is the target.
 \end{itemize}
@@
 
-Indeed most empirical social science research follows this approach: 1) delineation of an argument, 2) selection of variables that represent a plausible or compelling representation of the concepts, followed by 3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation.
+Indeed, most empirical social science research follows this approach: (1) delineation of an argument, (2) selection of variables that represent a plausible or compelling representation of the concepts, followed by (3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation.
 It is worth pointing out that this is simply one set of variables that are selected to represent the concepts in the theoretical story.
+
 A strong theory gives more guidance on compelling choices than does a weak theory.
@@
 
-At this level of detail the theoretical model is more about choosing the correct variables and this is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on.
+At this level of detail, the theoretical model is more about choosing the correct variables and this is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on.
 But this dichotomy is misguided.
@@
 For example, both @ward:etal:2010 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes.
-And it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.
+And, it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.
 the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups [@beger:ward:2020].

### 2020-09-24 paper.Rmd
fix table caption position

@@
   - \floatsetup[figure]{capposition=bottom}
+  - \floatsetup[table]{capposition=top}
   - \usepackage[format=plain,labelfont=bf,up]{caption}

### 2020-09-25 paper.Rmd
forecast scoring: confusion matrix with B&S original

@@
 B&S present a confusion matrix to score their 2016-H1 forecasts in Table 4.
-Although the forecasts are for the probability of civil war _onset_, in the replication code, they are actually scored using _incidence_ of civil war.
+Although the forecasts are for the probability of civil war _onset_, they are actually scored using _incidence_ of civil war.
 By definition the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict.
@@
 
-B&S made forecasts for the first half of 2016 (2016-H1), and assess the accuracy of these forecasts.
+B&S made forecasted the risk of civil war onset for the first half of 2016 (2016-H1, where H2 would be the second half), and assess the accuracy of these forecasts.
 The forecasts themselves are probabilistic, i.e.
 range from 0 to 1, but for the purpose of scoring the forecasts they consider the 30 highest predictions to have been positive forecasts ("1") and the remaining forecasts to have been for no onset ("0").
-They then construct confusion matrices for the forecasts, in their Table 4 and which we replicate here in the top half of our Table \ref{tab:table4}.
-They use two alternative codings of the outcomes for 2016-H1, one of which they call "Assuming Persistence" (this is what is in their replication data), and one of which they call "Assuming Change" (hand-coded in one of their replication scripts).
+They evaluate two alternative codings of observed civil war outcomes, one in which 15 countries, including Colombia, had ongoing civil wars at the end of 2015 (labeled "Assuming Persistence"), and an alternative coding in which the civil war in Colombia ends but Turkey and Benin experience civil war onsets ("Assuming Change").\footnote{\label{fn:leadDV}The former coding is in the main replication data, while the latter is hand-coded in the Stata script that produces their confusion matrices.
+There is also some potential ambiguity about the discussion in this section because the forecasts B\&S claim are for the first half of 2016 are actually for the second half of 2016, but scored against outcomes for the first half of 2016.} However, when they score the forecasts to produce the confusion matrices they report in their Table 4, they score them using civil war incidence, not onset.
+In fact, this a conscious choice B&S seem to acknowledge in the text (if not the table with confusion matrices), but it's not clear to us how this is justifiable.
 
+The way civil war is operationalized by Sambanis is so that a country at any given point in time is either at civil war or not, and it cannot experience another civil war onset when it is already at war.
+In the data, ongoing civil wars are coded as having a missing outcome value, since a new onset is conceptually impossible, and dropped.
+If we place ourselves back to the beginning of 2016, when the forecasts were made (the forecasts are pre-registered in early March), we would presumable already have observed civil war state for the second half of 2015 and seen that 15 countries had ongoing conflicts.
+In all previous analyses in the paper, these kinds of cases drop out of the data.
+However, B&S re-introduce, produce forecasts for them, and include them as positive events ("1"s) in their confusion matrices.
+They write, on page 21:
+
+> Importantly, we make no assumptions about the persistence of civil war between the second half of 2015 and the first half of 2016, meaning that all countries—even those with ongoing conflicts as of December 31, 2015--are included in the 2016 test set.
+As a result, 14 of the 30 countries on the list (and nine of the top 10) are cases of ongoing conflict.
+
+It's not clear to us under what conditions forecasting civil war onset risk for countries with ongoing civil wars make sense.
+Theoretically, a conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, _we would have already known that_.
+Including those cases in 2016-H1 risk set that we are forecasting for would only make sense if all of them had seen civil war termination in 2015-H2, but when making the forecast we would have already known this is not the case.
+
+When creating the confusion matrices, B&S treat these ongoing civil war instances equivalent to civil war onsets, as positive events ("1").
+In the text, B&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also do try to claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases.
+This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g.
+two separatists conflicts in different portions of the country.
+But this is not the case; ongoing conflicts are consistently in all other analyses dropped from the data.
+Thus it is not possible to treat incidence of civil war similar to onset of civil war, which is what the model actually predicts.
+
+We replicate the B&S confusion matrices but distinguish between non-evens that correspond to countries with no civil war, those with an ongoing civil war, and those with an actual civil war onset in Table \ref{tab:table4}.
 With the "Assuming Persistence" version of the outcomes, shown in the first confusion matrix, B\&S report that there are 15 civil war onsets, of which the forecasts capture 13 (true positives) and miss 2 (false negatives).
@@
 % fixed, assuming persistence
-\multicolumn{1}{l}{\multirow{4}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
+\multicolumn{1}{l}{\multirow{5}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
 \multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
 \cline{3-4}
-& $Y=0$ & 134 & 30 \\
+& $Y=0$, no civil war & 134 & 17 \\
 \cline{3-4}
-& $Y=1$ & 0 & 0 \\
+& $Y=0$, ongoing & 2 & 13 \\
+\cline{3-4}
+& $Y=1$, onset & 0 & 0 \\
 \cline{3-4}
 % fixed, assuming change
-\multicolumn{1}{l}{\multirow{4}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
+\multicolumn{1}{l}{\multirow{5}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
 \multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
 \cline{3-4}
-& $Y=0$ & 134 & 28 \\
+& $Y=0$, no civil war & 134 & 16 \\
 \cline{3-4}
-& $Y=1$ & 0 & 2 \\
+& $Y=0$, ongoing & 2 & 12 \\
+\cline{3-4}
+& $Y=1$, onset & 0 & 2 \\
 \cline{3-4}
@@
 
+Regarding the claim in footnote \ref{fn:leadDV} that the 2016-H1 forecasts are actually forecasts for 2016-H2:
+
+1.
+Throughout the R code for the main analyses, the models predict a 1-time period lead version of the civil war onset variables (ending in "_plus1").
+This means that, for example, in the 6-month version of the data, a value of "1" for the "_plus1" outcome variable for 2015-H2 actually refers to a civil war onset in the first half of 2016.
+One can confirm, through some tracing to work back to the original variable assignments, that the ROC curves and AUC are correspondingly, and correctly, also scored using the lead versions of the DV.
+2. The nominally 2016-H1 forecasts are created at the bottom of `+master.R` using the data for 2016-H1 and still the lead DV. Thus the lead DV here refers to outcomes in 2016-H2. 
+3.
+The scoring of the forecasts, whatever time period they are actually for, is done in a separate Stata do file, `6mo_make_confusion_matrix.do`.
+Here, the scoring is done using the non-lead version of the DV.
+Thus there is a mismatch between the time period the forecasts are for and the period who's observed outcomes are used to score them.
+
 The relevant variables in the data are "incidence_civil_ns" and "incidence_civil_ns_plus1", which appears to be a 1-period lead (ie., $t_+1$) version of the dependent variable that is used in the actual prediction models.
@@
 
-However, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
+Furthermoe, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
 In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable to incidence.

### 2020-09-25 paper.Rmd
forecast scoring: take out original confusion matrices

@@
 We replicate the B&S confusion matrices but distinguish between non-evens that correspond to countries with no civil war, those with an ongoing civil war, and those with an actual civil war onset in Table \ref{tab:table4}.
+The original B&S confusion matrices collapse the second two rows in each matrix and treat ongoing and onset as part of the same group.
+It is now more apparent that essentially all of the apparent true positives ($P=1, Y=1$) in the original B&S confusion matrices are actually ongoing civil war cases which by definition _cannot_ experience another civil war onset.
+The alternative outcome coding B&S evaluate, under "Assuming Change", produces the only two civil war onsets we would have observed in 2016-H1.
+
 With the "Assuming Persistence" version of the outcomes, shown in the first confusion matrix, B\&S report that there are 15 civil war onsets, of which the forecasts capture 13 (true positives) and miss 2 (false negatives).
@@
 \begin{tabular}{ll|c|c|l}
-\multicolumn{5}{c}{\textbf{Original confusion matrices with civil war \textit{incidence}}} \\
-% BS, assuming persistence
-\multicolumn{1}{l}{\multirow{4}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
-\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
-\cline{3-4}
-& $Y=0$ & 132 & 17 \\
-\cline{3-4}
-& $Y=1$ & 2 & 13 \\
-\cline{3-4}
-% BS, assuming change
-\multicolumn{1}{l}{\multirow{4}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
-\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
-\cline{3-4}
-& $Y=0$ & 132 & 16 \\
-\cline{3-4}
-& $Y=1$ & 2 & 14 \\
-\cline{3-4}
-\\ \midrule
-\\
-\multicolumn{5}{c}{\textbf{Fixed confusion matrices with civil war \textit{onset}}} \\
-% fixed, assuming persistence
 \multicolumn{1}{l}{\multirow{5}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\

### 2020-09-25 paper.Rmd
note
fix re-writing the forecast scoring section

@@
 
-The way civil war is operationalized by Sambanis is so that a country at any given point in time is either at civil war or not, and it cannot experience another civil war onset when it is already at war.
-In the data, ongoing civil wars are coded as having a missing outcome value, since a new onset is conceptually impossible, and dropped.
-If we place ourselves back to the beginning of 2016, when the forecasts were made (the forecasts are pre-registered in early March), we would presumable already have observed civil war state for the second half of 2015 and seen that 15 countries had ongoing conflicts.
-In all previous analyses in the paper, these kinds of cases drop out of the data.
-However, B&S re-introduce, produce forecasts for them, and include them as positive events ("1"s) in their confusion matrices.
-They write, on page 21:
-
-> Importantly, we make no assumptions about the persistence of civil war between the second half of 2015 and the first half of 2016, meaning that all countries—even those with ongoing conflicts as of December 31, 2015--are included in the 2016 test set.
-As a result, 14 of the 30 countries on the list (and nine of the top 10) are cases of ongoing conflict.
-
-It's not clear to us under what conditions forecasting civil war onset risk for countries with ongoing civil wars make sense.
-Theoretically, a conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, _we would have already known that_.
-Including those cases in 2016-H1 risk set that we are forecasting for would only make sense if all of them had seen civil war termination in 2015-H2, but when making the forecast we would have already known this is not the case.
-
-When creating the confusion matrices, B&S treat these ongoing civil war instances equivalent to civil war onsets, as positive events ("1").
-In the text, B&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also do try to claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases.
-This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g.
-two separatists conflicts in different portions of the country.
-But this is not the case; ongoing conflicts are consistently in all other analyses dropped from the data.
-Thus it is not possible to treat incidence of civil war similar to onset of civil war, which is what the model actually predicts.
-
-We replicate the B&S confusion matrices but distinguish between non-evens that correspond to countries with no civil war, those with an ongoing civil war, and those with an actual civil war onset in Table \ref{tab:table4}.
+For reference for the following discussion, we reproduce confusion matrices for the 2016-H1 forecasts in Table \ref{tab:table4}, but distinguish between civil war onsets ($Y=1$), non-events ($Y=0$) that correspond to countries with an ongoing civil war, and non-events ($Y=0$) that correspond to countries without a civil war, but which could potentially experience a civil war onset.
 The original B&S confusion matrices collapse the second two rows in each matrix and treat ongoing and onset as part of the same group.
-It is now more apparent that essentially all of the apparent true positives ($P=1, Y=1$) in the original B&S confusion matrices are actually ongoing civil war cases which by definition _cannot_ experience another civil war onset.
+Essentially all of the apparent true positives ($P=1, Y=1$) in the original B&S confusion matrices are actually ongoing civil war cases which by definition _cannot_ experience another civil war onset.
 The alternative outcome coding B&S evaluate, under "Assuming Change", produces the only two civil war onsets we would have observed in 2016-H1.
-
-With the "Assuming Persistence" version of the outcomes, shown in the first confusion matrix, B\&S report that there are 15 civil war onsets, of which the forecasts capture 13 (true positives) and miss 2 (false negatives).
-With the "Assuming Persistence" coding of outcomes, shown in the second confusion matrix, they report that there are 16 onsets and the forecasts capture 14 of those with, still, 2 misses.
-Fifteen to 16 civil war _onsets_ in the first half of 2016 is incredible, especially considering that the entire historical data from 2001 to 2015 only has 20 recorded civil war _onsets_.
-Indeed, it turns out that B&S scored their forecasts for civil war \textit{onset} using civil war \textit{incidence}, i.e.
-counting ongoing civil wars as if they were onsets.
 
@@
 
-The correct confusion matrices that use civil war onsets to score the forecasts are shown in the second part of Table \ref{tab:table4}.
-In the "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1.
-The alternative "Assuming Persistence" coding records 2 civil war onsets in 2016-H1, according to the civil conflict dataset they rely on.
-As a result, if we compare the precision and recall of the forecasts with the correct scoring to the incorrect scoring in B&S, almost all of the forecasts are false positives.
-This is the norm when predicting rare events, and thus not unusual at all.
+The fact that countries with an ongoing civil war cannot experience another civil war onset follows from the way the Sambanis civil war data conceptualize and operationalize civil war.
+Civil war appears to be a binary property of a country at any given moment in time, i.e.
+a country at any given point is either at civil war or not, and onset is the first time period corresponding to this state.
+In the data, ongoing civil wars are coded as having a missing outcome value, and dropped.
+
+If we place ourselves back to the beginning of 2016, when the forecasts were made (the forecasts are pre-registered in early March), this means that we would have observed that there are 15 countries that had ongoing conflicts in the second half of 2015 which had not terminated prior to the forecasting period.
+In all previous analyses in the paper, these kinds of cases drop out of the data.
+However, B&S re-introduce, produce forecasts for them, and include them as positive events ("1"s) in their confusion matrices.
+They write, on page 21:
+
+> Importantly, we make no assumptions about the persistence of civil war between the second half of 2015 and the first half of 2016, meaning that all countries—even those with ongoing conflicts as of December 31, 2015--are included in the 2016 test set.
+As a result, 14 of the 30 countries on the list (and nine of the top 10) are cases of ongoing conflict.
+
+It's not clear to us under what conditions forecasting civil war onset risk for countries with ongoing civil wars make sense.
+Theoretically, a conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, _we would have already known that_.
+Rather than not making any "assumptions about the persistence of civil war", this would only make sense if we in fact assume that all ongoing civil wars ended in 2015-H2.
+Since Sambanis is the author of the civil war data that is used, it would  already have been known in early 2016 that these 15 civil wars had in fact not terminated, and were still ongoing.
+It is thus both by definition impossible for countries with ongoing civil wars to experience another onset, and at the time the forecasts were made it would have been known that all conflicts were still ongoing up through the beginning of the forecasting period.
+
+Does it make sense to count a high probability for the risk of onset for a case with an ongoing civil war as a partial or complete "win" for the model?
+When creating the confusion matrices, B&S treat these ongoing civil war instances equivalent to civil war onsets, as positive events ("1").
+In the text, B&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also do try to claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases.
+This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g.
+two separatists conflicts in different portions of the country.
+But this is not the case; ongoing conflicts are consistently in all other analyses dropped from the data as if a new civil war onset for a country with an ongoing conflict is conceptually impossible.
+
+If we turn back to the confusion matrices to re-assess the accuracy of the 2016-H1 forecasts, we can thus conclude that almost all of the predictions are false positives.
+There are no onsets in the "Assuming Persistence" coding, and two onsets in the "Assuming Change" coding.
+This is normal.
+Civil war onsets are infrequent, and it is to be expected that forecasts of rare events will mostly consist of false positives.
 The forecasts do correctly capture the 2 civil war onsets in the "Assuming Change" version of the outcome coding.

### 2020-09-25 paper.Rmd
what

@@
 Their model is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.
-More importantly, we argue that rather than trying to contrast the roles of theory and \"atheoretical\" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant theories.
+More importantly, we argue that rather than trying to contrast the roles of theory and
+\"atheoretical\" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant predictive work.
+Instead, we argue that predictive modeling and machine learning are effective tools for theory testing.
+
 output: 
@@
 
-Although their analysis in the end does not support this point, we do not disagree that theory is important in predictive modeling.
+Although their analysis in the end does not support this point, we do not disagree that theory can important in predictive modeling.
 Of the various ways one could contrast ad hoc machine learning with "theoretical" modeling, B&S focus essentially on theory as a mechanism for variable selection.
 In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting.
-The problem is that the space of extant theories for a phenomenon like civil war is so broad and non-specific in terms of an exact specification of what variables and in which functional form should constitute a prediction model, that it's both not very clear what the "correct" model should be, nor easy to demonstrate that any particular "theoretical" model's ability to predict generalizes beyond the data that was used to create it in the first place.
+The problem is that the extent of extant theories for a phenomenon like civil war is so broad and non-specific in terms.
+They do not provide much guidance on an exact specification of what variables and which functional form is required.
+As a result it is both not very clear what the "correct" model should be, nor easy to demonstrate that any particular "theoretical" model's ability to predict generalizes beyond the data that was used to create it in the first place.
 Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to "explain" a phenomenon to strengthen and whittle down the broad space of extant theories.
@@
 
-The next three issues we encountered all concern information in B&S Table 2, which is the key table in which B&S report AUC-ROC values to demonstrate the "Escalation" models superiority.
+The next three issues we encountered all concern information in B&S Table 2, which is the key table in which B&S report AUC-ROC values to demonstrate the "Escalation" model's superiority.
 
@@
 
-At this level of detail, the theoretical model is more about choosing the correct variables and this is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on.
+Instantiating the theoretical model is more about choosing the correct variables.
+This is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on.
 But this dichotomy is misguided.
@@
 It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.
-Social science has leaned on theory for (at least) seventy years, but it might be time to find useful ways to incorporate some of the recently developed methods in the arena of machine learning and modern statistics as well.
+Social science has leaned on theory for (at least) seventy years, but it is time to find useful ways to incorporate recently developed methods in the arena of machine learning and modern statistics.
 

### 2020-09-25 paper.Rmd
take out line break in YAML abstract

@@
 Their model is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.
-More importantly, we argue that rather than trying to contrast the roles of theory and
-\"atheoretical\" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant predictive work.
+More importantly, we argue that rather than trying to contrast the roles of theory and \"atheoretical\" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant predictive work.
 Instead, we argue that predictive modeling and machine learning are effective tools for theory testing.

### 2020-09-25 paper.Rmd
submission revision (sent)

@@
   - \usepackage[format=plain,labelfont=bf,up]{caption}
-#  - \usepackage{setspace}\doublespacing
+  - \usepackage{setspace}\doublespacing
 bibliography: "./references.bib"
@@
 
-B&S made forecasted the risk of civil war onset for the first half of 2016 (2016-H1, where H2 would be the second half), and assess the accuracy of these forecasts.
+B&S forecast the risk of civil war onset for the first half of 2016 (2016-H1, where H2 would be the second half), and assessed the accuracy of these forecasts.
 The forecasts themselves are probabilistic, i.e.
@@
 
-It's not clear to us under what conditions forecasting civil war onset risk for countries with ongoing civil wars make sense.
-Theoretically, a conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, _we would have already known that_.
+It's not clear to us under what conditions forecasting civil war onset risk for countries with ongoing civil wars make sense. A
+conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, _we would have already known that_.
 Rather than not making any "assumptions about the persistence of civil war", this would only make sense if we in fact assume that all ongoing civil wars ended in 2015-H2.
@@
 When creating the confusion matrices, B&S treat these ongoing civil war instances equivalent to civil war onsets, as positive events ("1").
-In the text, B&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also do try to claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases.
+In the text, B&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases.
 This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g.
 two separatists conflicts in different portions of the country.
-But this is not the case; ongoing conflicts are consistently in all other analyses dropped from the data as if a new civil war onset for a country with an ongoing conflict is conceptually impossible.
+But this is not the case; ongoing conflicts in all other analyses  are consistently dropped from the data as if a new civil war onset for a country with an ongoing conflict is conceptually impossible.
 

### 2020-09-26 paper.Rmd
single space

@@
   - \usepackage[format=plain,labelfont=bf,up]{caption}
-  - \usepackage{setspace}\doublespacing
+#  - \usepackage{setspace}\doublespacing
 bibliography: "./references.bib"

### 2020-11-09 paper/paper.Rmd
move figure generation to separate files
Rmarkdown does less lifting

--- 2020-09-25 paper.Rmd
@@
 \centering
-\includegraphics[width=.95\linewidth]{figures/figure1-replicated.png}
+\includegraphics[width=.95\linewidth]{figures/fig-1-figure1-replicated.png}
 \end{figure}
@@
 
-```{r benefit-plot, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2).
-The Escalation model is the only model which consistently benefits from using smoothed ROC curves, and this advantage accounts for its apparent superiority to the other models."}
-table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
-table2_benefit <- read_csv("data/table2-smooth-benefit.csv")
-
-cols <- c(
-  Escalation = "#e41a1c",
-  Goldstein = "#ff7f00",
-  Quad = "#984ea3",
-  CAMEO = "#377eb8",
-  Average = "#4daf4a",
-  `With PITF Predictors` = "gray50", 
-  `Weighted by PITF` = "gray50", 
-  `PITF Split Population` = "gray50",
-  `PITF Only` = "gray50"
-)
-
-benefit <- bind_rows(
-  table1_benefit %>%
-    rename(setting = Model) %>%
-    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff"),
-  table2_benefit %>%
-    rename(setting = Model, Escalation = `Escalation Only`) %>%
-    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff")
-)
-
-# mean benefit by model
-means <- benefit %>%
-  group_by(model) %>%
-  summarize(avg_benefit = mean(diff))
-
-ann_text <- tibble(horizon = "1 month", diff = 0.07, model = 8.6, label = "Mean")
-
-model_levels <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average", 
-                  "With PITF Predictors", "Weighted by PITF", "PITF Split Population",
-                  "PITF Only")
-benefit %>%
-  mutate(model = factor(model, levels = rev(model_levels))) %>%
-  ggplot(aes(y = model)) + 
-  #facet_wrap(~ horizon, scales = "free_x") +
-  geom_point(aes(x = diff, color = model), alpha = 0.5, size = 2) +
-  scale_color_manual(guide = FALSE, values = cols) +
-  geom_point(data = means, aes(x = avg_benefit), shape = "|", color = "black", size = 6) +
-  labs(x = "AUC(smoothed ROC) - AUC(empirical ROC)",
-       y = "") +
-  theme_bw() +
-  geom_text(data = ann_text, aes(x = diff, label = label), col = "black") +
-  geom_vline(xintercept = 0, linetype = 3) +
-  theme(axis.text.y = element_text(size = 12))
-```
+\begin{figure}
+\caption{Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\&S Tables 1 and 2).
+The Escalation model is the only model which consistently benefits from using smoothed ROC curves, and this advantage accounts for its apparent superiority to the other models.\label{fig:benefit-plot}}
+\centering
+\includegraphics[width=.8\linewidth]{figures/fig-2-benefit-plot.png}
+\end{figure}
 
@@
 
-The intention is to train seperate random forests on data split into high and low-risk groups based on PITF predictions, and then combine them back into one random forest.
+The intention is to train separate random forests on data split into high and low-risk groups based on PITF predictions, and then combine them back into one random forest.
 Actually both random forests are trained on identical data, essentially production a random forest that just has N_trees \* 2.
@@
 
+\clearpage
+
 ## Model to model comparison plots
@@
 
-```{r table1-plot, fig.height=5, fig.width=8, out.width = ".9\\linewidth", fig.align = "center", fig.cap = "Escalation to alternative comparisons for the ICEWS models (B\\&S Table 1)"}
-table1_nosmooth <- read_csv("data/table1-nosmooth.csv")
+\begin{figure}
+{\centering
+\includegraphics[width=.9\linewidth]{figures/fig-A1-table1-pairs.png}
+}
+\caption{Escalation to alternative comparisons for the ICEWS models (B\&S Table 1)\label{fig:table1-plot}}
+\end{figure}
 
-icews_lvls <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average")
+\begin{figure}
+{\centering
+\includegraphics[width=.9\linewidth]{figures/fig-A2-table2-pairs.png}
+}
+\caption{Escalation to alternative comparisons for the structural models (B\&S Table 2)\label{fig:table2-plot}}
+\end{figure}
 
-df <- table1_nosmooth %>%
-  rename(Spec = Model) %>%
-  pivot_longer(Escalation:Average, names_to = "Model", values_to = "auc_roc") %>%
-  mutate(Model = factor(Model, levels = icews_lvls)) %>%
-  nest(data = c(Spec, auc_roc))
-
-pairs <- full_join(
-  df %>% filter(Model!="Escalation"),
-  df %>% filter(Model=="Escalation") %>% rename(data2 = data) %>% select(horizon, data2),
-  by = c("horizon")
-) %>%
-  pivot_longer(data:data2, names_to = "xname", values_to = "data") %>%
-  mutate(xname = ifelse(xname=="data", as.character(Model), "Escalation"),
-         xname = factor(xname, levels = icews_lvls))
-
-pairs %>%
-  unnest(data) %>%
-  ggplot(aes(x = xname, y = auc_roc)) +
-  facet_grid(horizon ~ Model, scales = "free_x") +
-  geom_point() +
-  geom_line(aes(group = Spec)) +
-  theme_bw() +
-  theme(panel.grid.minor = element_blank()) +
-  labs(x = "", y = "AUC-ROC (empirical)")
-```
-
-```{r table2-plot, fig.height=5, fig.width=8, out.width = ".9\\linewidth", fig.align = "center", fig.cap = "Escalation to alternative comparisons for the structural models (B\\&S Table 2)"}
-table2_nosmooth <- read_csv("data/table2-nosmooth.csv")
-
-model_lvls <- colnames(table2_nosmooth)[-c(1:2)]
-
-df <- table2_nosmooth %>%
-  rename(Spec = Model) %>%
-  pivot_longer(`Escalation Only`:`PITF Only`, names_to = "Model", values_to = "auc_roc") %>%
-  mutate(Model = factor(Model, levels = model_lvls)) %>%
-  nest(data = c(Spec, auc_roc))
-
-pairs <- full_join(
-  df %>% filter(Model!="Escalation Only"),
-  df %>% filter(Model=="Escalation Only") %>% rename(data2 = data) %>% select(horizon, data2),
-  by = c("horizon")
-) %>%
-  pivot_longer(data:data2, names_to = "xname", values_to = "data") %>%
-  mutate(xname = ifelse(xname=="data", as.character(Model), "Escalation Only"),
-         xname = factor(xname, levels = model_lvls))
-
-pairs %>%
-  unnest(data) %>%
-  ggplot(aes(x = xname, y = auc_roc)) +
-  facet_grid(horizon ~ Model, scales = "free_x") +
-  geom_point() +
-  geom_line(aes(group = Spec)) +
-  theme_bw() +
-  theme(panel.grid.minor = element_blank(),
-        axis.text.x = element_text(angle = 30, hjust = 1)) +
-  labs(x = "", y = "AUC-ROC (empirical)")
-```
+\newpage
 
@@
 
+\newpage
+
 ## Different smoothing methods
@@
 
-```{r smooth-benefit-plot-extended, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2), with 3 different smoothing methods (binormal is the default method and used in B\\&S)"}
-
-suppressPackageStartupMessages({
-  library(tidyverse)
-  library(yardstick)
-  library(pROC)
-  library(forcats)
-})
-
-preds <- read_rds("rep_nosmooth/output/all-predictions.rds")
-
-auc_roc_vec <- function(pred, truth, smooth, sm = "binormal") {
-  roc_obj <- pROC::roc(truth, pred, auc = TRUE, quiet = TRUE, smooth = smooth, smooth.method = sm)
-  as.numeric(roc_obj$auc)
-}
-
-rocs <- preds %>% 
-  filter(!is.na(value)) %>%
-  group_by(cell_id) %>% 
-  summarize(
-    no = auc_roc_vec(pred = pred, truth = value, smooth = FALSE),
-    binormal = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "binormal"),
-    density = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "density"),
-    fitdistr = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "fitdistr")
-  )
-
-# these two take a long time and error out for some reason
-#logcondens = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "logcondens"),
-#logcondens.smooth = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "logcondens.smooth")
-
-mt <- read_rds("rep_nosmooth/output/model-table-w-results.rds")
-mt <- mt %>% left_join(rocs, by = "cell_id")
-
-cols <- c(
-  Escalation = "#e41a1c",
-  Goldstein = "#ff7f00",
-  Quad = "#984ea3",
-  CAMEO = "#377eb8",
-  Average = "#4daf4a",
-  `With PITF Predictors` = "gray50", 
-  `Weighted by PITF` = "gray50", 
-  `PITF Split Population` = "gray50",
-  `PITF Only` = "gray50"
-)
-
-model_levels <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average", 
-                  "With PITF Predictors", "Weighted by PITF", "PITF Split Population",
-                  "PITF Only")
-
-benefit <- mt %>% 
-  mutate(column = fct_recode(column, Escalation = "Escalation Only"),
-         column = factor(column, levels = rev(model_levels))) %>%
-  mutate(binormal = binormal - no,
-         density  = density - no,
-         fitdistr = fitdistr - no) %>%
-  select(cell_id:non_RF, binormal:fitdistr) %>%
-  pivot_longer(binormal:fitdistr, names_to = "method")
-
-means <- benefit %>%
-  group_by(column, method) %>%
-  summarize(avg_benefit = mean(value))
-
-ggplot(benefit, aes(x = value, y = column, color = column)) +
-  facet_wrap(~method) +
-  geom_point() +
-  scale_color_manual(guide = FALSE, values = cols) +
-  theme(axis.text.y = element_text(size = 12)) +
-  geom_point(data = means, aes(x = avg_benefit), shape = "|", color = "black", size = 6) +
-  labs(x = "AUC(smoothed ROC) - AUC(empirical ROC)",
-       y = "") +
-  theme_bw() +
-  geom_vline(xintercept = 0, linetype = 3) +
-  theme(axis.text.y = element_text(size = 12))
-```
+\begin{figure}
+{\centering \includegraphics[width=.95\linewidth]{figures/fig-a3-benefit-plot-extended.png}}
+\caption{Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\&S Tables 1 and 2), with 3 different smoothing methods (binormal is the default method and used in B\&S)\label{fig:smooth-benefit-plot-extended}}
+\end{figure}

### 2020-11-09 rep_nosmooth/7-make-figure2.R

@@ -1,0 +1,63 @@
+#
+#   Create Figure 2: Gain from using smoothed ROC to calculate AUC
+#
+
+library(ggplot2)
+library(readr)
+library(tibble)
+library(dplyr)
+library(tidyr)
+library(here)
+
+setwd(here::here("rep_nosmooth"))
+
+
+table1_benefit <- read_csv("output/tables/table1-smooth-benefit.csv")
+table2_benefit <- read_csv("output/tables/table2-smooth-benefit.csv")
+
+cols <- c(
+  Escalation = "#e41a1c",
+  Goldstein = "#ff7f00",
+  Quad = "#984ea3",
+  CAMEO = "#377eb8",
+  Average = "#4daf4a",
+  `With PITF Predictors` = "gray50",
+  `Weighted by PITF` = "gray50",
+  `PITF Split Population` = "gray50",
+  `PITF Only` = "gray50"
+)
+
+benefit <- bind_rows(
+  table1_benefit %>%
+    rename(setting = Model) %>%
+    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff"),
+  table2_benefit %>%
+    rename(setting = Model, Escalation = `Escalation Only`) %>%
+    pivot_longer(-c(horizon, setting), names_to = "model", values_to = "diff")
+)
+
+# mean benefit by model
+means <- benefit %>%
+  group_by(model) %>%
+  summarize(avg_benefit = mean(diff))
+
+ann_text <- tibble(horizon = "1 month", diff = 0.07, model = 8.6, label = "Mean")
+
+model_levels <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average",
+                  "With PITF Predictors", "Weighted by PITF", "PITF Split Population",
+                  "PITF Only")
+p <- benefit %>%
+  mutate(model = factor(model, levels = rev(model_levels))) %>%
+  ggplot(aes(y = model)) +
+  #facet_wrap(~ horizon, scales = "free_x") +
+  geom_point(aes(x = diff, color = model), alpha = 0.5, size = 2) +
+  scale_color_manual(guide = FALSE, values = cols) +
+  geom_point(data = means, aes(x = avg_benefit), shape = "|", color = "black", size = 6) +
+  labs(x = "AUC(smoothed ROC) - AUC(empirical ROC)",
+       y = "") +
+  theme_bw() +
+  geom_text(data = ann_text, aes(x = diff, label = label), col = "black") +
[…]

### 2020-11-09 rep_nosmooth/8-make-figureA1.R

@@ -1,0 +1,45 @@
+#
+#   Generate Figure A.1., Comparing Escalation to alternative models in B&S
+#   Table 1.
+#
+
+library(ggplot2)
+library(readr)
+library(dplyr)
+library(tidyr)
+library(here)
+
+setwd(here::here("rep_nosmooth"))
+
+table1_nosmooth <- read_csv("output/tables/table1-nosmooth.csv")
+
+icews_lvls <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average")
+
+df <- table1_nosmooth %>%
+  rename(Spec = Model) %>%
+  pivot_longer(Escalation:Average, names_to = "Model", values_to = "auc_roc") %>%
+  mutate(Model = factor(Model, levels = icews_lvls)) %>%
+  nest(data = c(Spec, auc_roc))
+
+pairs <- full_join(
+  df %>% filter(Model!="Escalation"),
+  df %>% filter(Model=="Escalation") %>% rename(data2 = data) %>% select(horizon, data2),
+  by = c("horizon")
+) %>%
+  pivot_longer(data:data2, names_to = "xname", values_to = "data") %>%
+  mutate(xname = ifelse(xname=="data", as.character(Model), "Escalation"),
+         xname = factor(xname, levels = icews_lvls))
+
+p <- pairs %>%
+  unnest(data) %>%
+  ggplot(aes(x = xname, y = auc_roc)) +
+  facet_grid(horizon ~ Model, scales = "free_x") +
+  geom_point() +
+  geom_line(aes(group = Spec)) +
+  theme_bw() +
+  theme(panel.grid.minor = element_blank()) +
+  labs(x = "", y = "AUC-ROC (empirical)")
+
+ggsave(plot=p, filename = "output/figures/fig-A1-table1-pairs.png",
+       height = 5, width = 8)
+

### 2020-11-09 rep_nosmooth/8-make-figureA3.R

@@ -1,0 +1,83 @@
+#
+#   Create Figure A.3: Extended smooth AUC-ROC benefit plot
+#
+
+
+library(tidyverse)
+library(yardstick)
+library(pROC)
+library(forcats)
+library(here)
+
+setwd(here::here("rep_nosmooth"))
+
+
+preds <- read_rds("output/all-predictions.rds")
+
+auc_roc_vec <- function(pred, truth, smooth, sm = "binormal") {
+  roc_obj <- pROC::roc(truth, pred, auc = TRUE, quiet = TRUE, smooth = smooth, smooth.method = sm)
+  as.numeric(roc_obj$auc)
+}
+
+rocs <- preds %>%
+  filter(!is.na(value)) %>%
+  group_by(cell_id) %>%
+  summarize(
+    no = auc_roc_vec(pred = pred, truth = value, smooth = FALSE),
+    binormal = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "binormal"),
+    density = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "density"),
+    fitdistr = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "fitdistr")
+  )
+
+# these two take a long time and error out for some reason
+#logcondens = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "logcondens"),
+#logcondens.smooth = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "logcondens.smooth")
+
+mt <- read_rds("../rep_nosmooth/output/model-table-w-results.rds")
+mt <- mt %>% left_join(rocs, by = "cell_id")
+
+cols <- c(
+  Escalation = "#e41a1c",
+  Goldstein = "#ff7f00",
+  Quad = "#984ea3",
+  CAMEO = "#377eb8",
+  Average = "#4daf4a",
+  `With PITF Predictors` = "gray50",
+  `Weighted by PITF` = "gray50",
+  `PITF Split Population` = "gray50",
+  `PITF Only` = "gray50"
+)
+
+model_levels <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average",
+                  "With PITF Predictors", "Weighted by PITF", "PITF Split Population",
+                  "PITF Only")
+
+benefit <- mt %>%
+  mutate(column = fct_recode(column, Escalation = "Escalation Only"),
+         column = factor(column, levels = rev(model_levels))) %>%
[…]

### 2020-11-09 rep_nosmooth/README.Rmd
renumber replication files and update README

@@
 
-## (Clean) replication
+## Reproducing our replication
 
@@
 
-Run through the files in the following order:
+The file numbering indicates the order in which files should be run, and correspond to the steps above.
+Files with the same number can be run in any order, as long as lower number files have already been run.
 
-- `re-save-data.R` changes the format of the saved 1-month and 6-month data versions from Stata 13 .dta files to R .rds format. This can be read much quicker than the Stata format. 
-- `setup-model-table.R` encodes all the various pieces of information needed to actually run the models reflected in tables 1 and 2.
+Setup:
+
+- `0-packages.R` makes sure all required R packages are installed.
+
+Preparing data for model estimation:
+
+- `1-re-save-data.R` changes the format of the saved 1-month and 6-month data versions from Stata 13 .dta files to R .rds format. This can be read much quicker than the Stata format. 
+- `1-make-original-tables.R` is a by-hand coding of the values reported for Tables 1, 2, and 4 in B&S 2020. 
+- `1-setup-model-table.R` encodes all the various pieces of information needed to actually run the models reflected in tables 1 and 2.
 In the original replication files this is done in bits and pieces in `+master.R`, `code/[1|6]mo_define_models.R`, and each model runner file, e.g.
@@
 More details on this below.
-- `model-runner.R` runs the models.
+
+Estimate all models (this is where the heavy lifting takes places):
+
+- `2-model-runner.R` runs the models.
 Adjust the number of WORKERS at the top of the script as needed based on the number of cores on your computer.
 The primary output of this script is `output/model-table-w-results.csv` and `output/all-predictions.R`.
-- `recreate-tables.R` recreates the original and non-smooth ROC versions of Tables 1 and 2 using `output/model-table-w-results.csv` and `output/all-predictions.R`.
-- `recreate-figure1.R` recreates the ROC curves in B&S Figure 1 using `output/model-table-w-results.csv` and `output/all-predictions.R`.
 
-In addition, there are some ancillary materials.
+Process results; figures, tables, etc.:
+
+- `3-...`: the various scripts starting with "3-" recreate the various tables and figures that end up in the paper.
+These are written to the `rep_nosmooth/output` folder here; the `paper/` folder includes a `update-inputs.R` script that will copy needed pieces over to that folder.
+
+In addition, there are some ancillary materials that are not required for replicating the results in the paper, but add additional information. These are easiest to read in GitHub.
 
@@
 0.825.
-- `make-original-tables.R` is a by-hand coding of the values reported for Tables 1, 2, and 4 in B&S 2020. 
 - `rep-check.R` compares the results of our modified replication to the original values reported in Tables 1 and 2 of the B&S paper.

### 2020-11-09 rep_nosmooth/2-model-runner.R

@@ -340,22 +340,32 @@ preds <- cell_ids %>%
   select(-pred_file) %>%
   tidyr::unnest(preds)
 
+
+# Calculate a summary of just the AUC-ROCs for git
+short_results <- model_table_w_results %>%
+  dplyr::select(cell_id, table, horizon, row, column, auc_roc, auc_roc_smoothed)
+
 if (!is.null(RNG_SEED)) {
   out_file <- sprintf("output/model-table-w-results-%s.rds", RNG_SEED)
   write_rds(model_table_w_results, out_file)
 
   out_file <- sprintf("output/all-predictions-%s.rds", RNG_SEED)
   write_rds(preds, out_file)
+
+  out_file <- sprintf("output/model-table-w-results-%s.md", RNG_SEED)
+  short_results %>%
+    knitr::kable("markdown", digits = 2) %>%
+    writeLines(out_file)
+
 } else {
   write_rds(model_table_w_results, "output/model-table-w-results.rds")
   write_rds(preds, "output/all-predictions.rds")
+  short_results %>%
+    knitr::kable("markdown", digits = 2) %>%
+    writeLines("output/model-table-w-results.md")
 }
 
 
-# Write a summary of just the AUC-ROC for git
-model_table_w_results %>%
-  dplyr::select(cell_id, table, horizon, row, column, auc_roc, auc_roc_smoothed) %>%
-  knitr::kable("markdown", digits = 2) %>%
-  writeLines("output/model-table-w-results.md")
+
 
 lgr$info("Script finished")

### 2020-11-09 rep_nosmooth/README.Rmd
update numbering
recreate-tables has to be run before figure scripts

@@
 
-The file numbering indicates the order in which files should be run, and correspond to the steps above.
+The file numbering indicates the order in which files should be run, and roughly correspond to the steps above.
 Files with the same number can be run in any order, as long as lower number files have already been run.
@@
 
-Process results; figures, tables, etc.:
+Recreate AUC-ROC tables:
 
-- `3-...`: the various scripts starting with "3-" recreate the various tables and figures that end up in the paper.
-These are written to the `rep_nosmooth/output` folder here; the `paper/` folder includes a `update-inputs.R` script that will copy needed pieces over to that folder.
+- `3-recreate-tables.R` produces various tables summarizing the performance of the core event data models (B&S Table 1) and the structural models (B&S Table 2); some of these are used as inputs to generate figures below.
+
+Generate figures:
+
+- `4-...`: the various scripts starting with "4-" recreate the various figures that end up in the paper.
+
+The processed results created in the above two sets of scripts ("3-..." and "4-...") are written to the `rep_nosmooth/output` folder here.
+The `paper/` folder includes a `update-inputs.R` script that will copy needed pieces over to that folder.
 
@@
 - `rep-check.R` compares the results of our modified replication to the original values reported in Tables 1 and 2 of the B&S paper.
+
+## Variation due to random number generator seed
+
+We consciously chose not do set a random number generator (RNG) seed in the model estimation script (`2-model-runner.R`), because the substantive results should not change as a result of the small amount of variation related to RNG state.
+The [variance.md](variance.md) note explores this to some extent and leads us to believe that this is the case, with variation for the Table 1 models generally less than 0.01 AUC-ROC.
+
+Anyways, the model runner script does include a provision for setting a RNG seed, in which case the key output files will be suffixed with the RNG seed value.
+Note that the results processing scripts are not setup to work this these output files, but it should be fairly obvious what lines need modifications to make them work.
 

### 2020-11-09 rep_nosmooth/2-model-runner.R

@@ -1,6 +1,8 @@
 #
 #   Run all the models behind Tables 1 and 2
 #
+#   2020-11-09: It took me ~1h15m to run this script. (AB)
+#
 
 WORKERS  <- 8
 # if RNG_SEED is NULL, no seed will be set and the output file name will be as

### 2020-11-09 rep_nosmooth/4-make-figure2.R

@@ -10,7 +10,7 @@ library(tidyr)
 library(here)
 
 setwd(here::here("rep_nosmooth"))
-
+dir.create("output/figures", showWarnings = FALSE)
 
 table1_benefit <- read_csv("output/tables/table1-smooth-benefit.csv")
 table2_benefit <- read_csv("output/tables/table2-smooth-benefit.csv")

### 2020-11-09 rep_nosmooth/4-make-figureA1.R

@@ -10,6 +10,7 @@ library(tidyr)
 library(here)
 
 setwd(here::here("rep_nosmooth"))
+dir.create("output/figures", showWarnings = FALSE)
 
 table1_nosmooth <- read_csv("output/tables/table1-nosmooth.csv")
 

### 2020-11-09 rep_nosmooth/4-make-figureA2.R

@@ -10,6 +10,7 @@ library(tidyr)
 library(here)
 
 setwd(here::here("rep_nosmooth"))
+dir.create("output/figures", showWarnings = FALSE)
 
 table2_nosmooth <- read_csv("output/tables/table2-nosmooth.csv")
 

### 2020-11-09 rep_nosmooth/4-make-figureA3.R

@@ -10,7 +10,7 @@ library(forcats)
 library(here)
 
 setwd(here::here("rep_nosmooth"))
-
+dir.create("output/figures", showWarnings = FALSE)
 
 preds <- read_rds("output/all-predictions.rds")
 

### 2020-11-10 paper/paper.tex
track .tex file for a second here

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict},
  pdfauthor={Andreas Beger; Richard K. Morgan; Michael D. Ward},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{floatrow}
\floatsetup[figure]{capposition=bottom}
\floatsetup[table]{capposition=top}
\usepackage[format=plain,labelfont=bf,up]{caption}
\usepackage{setspace}\doublespacing
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict\thanks{John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. We are especially grateful to Paul Huth for his comments and guidance. All the code and several additional analyses can be found at our replication archive at \url{https://github.com/andybega/Blair-Sambanis-replication}.}}
\author{Andreas Beger\footnote{Predictive Heuristics, \href{mailto:adbeger@gmail.com}{\nolinkurl{adbeger@gmail.com}}.} \and Richard K. Morgan\footnote{Varieties of Democracy Institute, University of Gothenburg, \href{mailto:rick.morgan2@gmail.com}{\nolinkurl{rick.morgan2@gmail.com}}.} \and Michael D. Ward\footnote{Predictive Heuristics, Duke University, University of Washington, \href{mailto:michael.don.ward@gmail.com}{\nolinkurl{michael.don.ward@gmail.com}}. Corresponding author.}}
\date{10 November 2020}

\begin{document}
\maketitle
\begin{abstract}
We examine the research protocols in Blair and Sambanis (nd), who argue that their theory-based model can predict civil war onsets better than several atheoretical alternatives or a model with country-structural factors. We find that there are several important mistakes and that their key finding is entirely conditional on the use of parametrically smoothed ROC curves when calculating accuracy, rather than the standard empirical ROC curves that dominate the literature. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models. Their model is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine. More importantly, we argue that rather than trying to contrast the roles of theory and "atheoretical" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant predictive work. Instead, we argue that predictive modeling and machine learning are effective tools for theory testing.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Blair and Sambanis (n.d., hereafter B\&S) argue that theory is essential for creating models that have high accuracy in forecasting civil war onset. Indeed, they assert that with such theory forecasting is more accurate than has previously been possible. We re-examine the empirical basis for the claims made in support of it. We find that these claims to be unsupported and evidence presented for them to be incorrect. Their theory-based escalation model does not do better than the alternatives that they examine. It does worse. The reason for this is that they have made several mistakes in their research procedure. Further, the performance results they report are based on smoothed performance curves, not the empirical, unsmoothed curves. This provides misleading results. In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented. We also found that the scoring of their forecasts for the first half of 2016 was incorrectly performed using civil war incidence, not onset. In what follows, we show the impact of these mistakes on the conclusions.

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars ``remarkably accurately''. Indeed, B\&S argue that this theoretical model outperforms four other ``more mechanical'' alternatives. They also claim that the integration of structure with process is better than alternatives over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.\footnote{PITF (e.g., Goldstone et al. 2010; Beger and Ward 2017), the W-ICEWS project (Ward et al. 2013), VIEWS (Hegre et al. 2019), and others, e.g., Beger, Morgan, and Maxwell (2020).} B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

Before proceeding, we quote B\&S (page 24):

\begin{quote}
Our theoretically driven model generates accurate forecasts, with base specification AUCs of 0.82 and 0.85 over one- and six-month windows, respectively, and AUCs as high as 0.92 in other specifications. Our model also consistently and sometimes dramatically outperforms the alternatives we test. {[}\ldots{]} Cederman and Weidmann (2017, 476) argue that ``the hope that big data will somehow yield valid forecasts through theory-free `brute force' is misplaced in the area of political violence.'' Our results lend some credence to this claim.
\end{quote}

Although their analysis in the end does not support this point, we do not disagree that theory can important in predictive modeling. Of the various ways one could contrast ad hoc machine learning with ``theoretical'' modeling, B\&S focus essentially on theory as a mechanism for variable selection. In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting. The problem is that the extent of extant theories for a phenomenon like civil war is so broad and non-specific in terms. They do not provide much guidance on an exact specification of what variables and which functional form is required. As a result it is both not very clear what the ``correct'' model should be, nor easy to demonstrate that any particular ``theoretical'' model's ability to predict generalizes beyond the data that was used to create it in the first place. Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to ``explain'' a phenomenon to strengthen and whittle down the broad space of extant theories.

\hypertarget{summary-of-blair-and-sambanis-2020}{%
\section{Summary of Blair and Sambanis (2020)}\label{summary-of-blair-and-sambanis-2020}}

B\&S' analysis is based on the use of non-linear, non-deterministic machine learning models, and specifically random forests, one of which has a specification they argue is theory-based. Several others with more generic sets of covariates are also developed and compared. Notably, the analysis is at the country-month or country-(6 month) levels and relies in large part on indicators derived from the ICEWS event data. In short, they uphold their assumption that theory-guided empirical research produces better conflict predictions than machine learning inspired efforts that are necessarily ad hoc combinations of available variables. They arrive at this conclusion by examining the problem of predicting civil war onset. They report that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory, including a \textit{kitchen sink} model with more than 1,000 covariates.

B\&S specifically examine three questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  How does the theoretically-driven escalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories?
\item
  Does annual, structural information from the PITF instability forecasting model add to the escalation model's monthly and 6-month predictions?
\item
  How accurate were predictions using the escalation model for the first half of 2016?
\end{enumerate}

To assess the first two questions, B\&S use ICEWS data covering all major countries from 2001 to 2015. Two versions of the dataset are used, one at the country-month level, the other aggregated to 6-month half-years. The main outcome variable is civil war onset, measured using Sambanis' civil war dataset.

Both the first and second questions above rely on comparing their escalation model to various alternative models. The same procedure is used in both cases:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Split the training data into training (2001 - 2007) and test (2008 - 2015) data.
\item
  Estimate the escalation and other competing models with the training data.
\item
  Create out-of-sample (OOS) predictions from each model using the test set.
\item
  Calculate AUC-ROC measures for each set of OOS predictions.
\end{enumerate}

To examine the first question, B\&S compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:

\begin{itemize}
\tightlist
\item
  Escalation: a set of ten indicators, drawn from a theoretical escalation model, for interactions between the government on one side and opposition or rebel actors on the other.
\item
  Quad: ICEWS quad counts, i.e.~material conflict, material cooperation, verbal conflict, verbal cooperation, for interactions between the government and opposition or rebels. These are directed, thus making for four directed dyads, which with four quad categories make sixteen covariates.
\item
  Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for the same four directed dyads, for a total of four covariates.
\item
  CAMEO: counts for all CAMEO event codes over the four actor dyads, totaling \(1,160\) covariates, which are mostly zero for any country in any month.
\item
  Average: unweighted average of the predictions from the four models briefly described above.
\end{itemize}

The corresponding results for each question are shown in B\&S Tables 1 and 2, which we examine further below. We accurately replicate their Tables 1 and 2. The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions. These robustness checks vary either (1) random forecast hyperparameter values or (2) the year used to split the train/test data, or (3) alternative codings of the civil war onset dependent variable.

The second question, whether structural variables add to the escalation model, is assessed by comparing the original escalation model to four alternatives that incorporate annual, structural variables that are used in the PITF instability forecasting model:

\begin{itemize}
\tightlist
\item
  Escalation Only: the original basic escalation model with only ICEWS predictors.
\item
  With PITF Predictors: a random forest that also adds the PITF annual, structural variables.
\item
  Weighted by PITF: escalation model predictions weighted using the PITF instability model predictions.
\item
  PITF Split Population: the training data are split into high and low risk portions based on the PITF instability model predictions, two separate escalation random forests are trained on the splits, then re-combined into a single random forest that is used to create the test set predictions.
\item
  PITF Only: a random forest model based only on the annual, structural PITF model predictors.
\end{itemize}

The corresponding results are shown in B\&S Table 2.

Finally, B\&S used their escalation model to create forecasts for the first half of 2016. In their third analysis, they score the forecasts accuracy using subsequent civil war onset data. This is summarized in B\&S Table 3.

\hypertarget{implementation-issues-in-bs}{%
\section{Implementation Issues in B\&S}\label{implementation-issues-in-bs}}

While replicating and analyzing B\&S's results, we found several issues worthy of further discussion and investigation. These are (a) the use of smoothed ROC curves to draw conclusions about which model is best, (b) incorrect implementations of the weighted by PITF and PITF analog split-population models, (c) inconsistent test sets for the models examined, and (d) incorrect scoring of the 2016 forecasts.\footnote{We also note there are additional concerns arising from the question of how the random forest models were tuned by B\&S, especially given the way they are used is unorthodox. We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.}

We believe that these research decisions and issues lead B\&S to incorrect conclusions. The escalation model \textit{is not} the best; it actually performs worse than the atheoretical, kitchen sink model with over \(1000\) variables. We discuss these five issues below. We defer a complete analysis that corrects all these issues until later, as there are many possible permutations of a seriatim unfolding.

\hypertarget{smoothed-roc-curves}{%
\subsection{Smoothed ROC Curves}\label{smoothed-roc-curves}}

The most consequential issue that we found is that all AUC-ROC values reported in B\&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. B\&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation. However, smoothing is also used in all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs. We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), in order to survey the use of ROC curves. Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values. We found that 90\%--17 of 19--clearly used empirical ROC curves, and 2 (10\%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code. The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S. The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey. Both of the ambiguous cases include Blair as coauthor.} The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard. What is surprising is that anyone would choose the smoothed ROC implementation.

The next three issues we encountered all concern information in B\&S Table 2, which is the key table in which B\&S report AUC-ROC values to demonstrate the ``Escalation'' model's superiority.

\hypertarget{incorrect-weighted-by-pitf-implementation}{%
\subsection{Incorrect ``Weighted by PITF'' Implementation}\label{incorrect-weighted-by-pitf-implementation}}

The ``Weighted by PITF'' model is described as follows in B\&S, page 19:

\begin{quote}
The {[}Weighted by PITF model{]} uses PITF predicted probabilities to weight the results of the escalation model, ensuring that high-risk and low-risk countries that happen to take similar values on ICEWS-based predictors are nonetheless assigned different predicted probabilities in most months.
\end{quote}

We believe that B\&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.

As a result: (1) in the test set, all countries with names that alphabetically start after ``Tongo'' are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical. For example, the ``Weighted by PITF'' model predictions for Syria from 2008 to 2015 are the Escalation test predictions for that time period weighted, respectively, by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.

This appears to be an easily corrected coding error on the part of B\&S.

\hypertarget{incorrect-pitf-split-population-implementation}{%
\subsection{Incorrect ``PITF Split Population'' Implementation}\label{incorrect-pitf-split-population-implementation}}

The ``PITF Split Population'' model also appears to be incorrectly implemented, owing to a coding error. B\&S describe it on page 20:

\begin{quote}
The final approach is a random forest analog to split-population modeling. We first compute the average PITF predicted probability for each country across all years in our training set. We define those that fall in the bottom quartile as ``low risk'' and the rest as ``high risk.'' We then run our escalation model on the high-risk and low-risk subsets separately, combining the results into a single random forest {[}\ldots{]}.
\end{quote}

This description suggests that B\&S intended to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the \emph{exact same training data}, which consist of the full training data from all other models. In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.

The model specifications are also identical; i.e., they use the same \emph{x} variables and the same random forest hyper-parameter settings. The \emph{only} difference in the models as they are implemented in the B\&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing two forests of identical decision trees and, thus, identical predictions.

The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling (Chiba, Metternich, and Ward 2015; Beger et al. 2017). Although the two random forest models are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger random forest using both component model's underlying decision trees. Thus, while all random forest models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (\texttt{ntree}), the new random forest model after combination does have 200,000 decision trees. Furthermore, the PITF model predictions do not impact how the combined random forest model predicts at all, not even through a binary low-/high-risk split. The split-population PITF random forest model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing random forest randomization facilities (row and column sampling for each decision tree). This does not adequately implement their split-sample modeling strategy.

\hypertarget{inconsistent-test-set-n-for-the-models-in-table-2}{%
\subsection{Inconsistent test set N for the models in Table 2}\label{inconsistent-test-set-n-for-the-models-in-table-2}}

Further, the AUC-ROC values reported in the original B\&S Table 2 are calculated on different numbers of underlying test set cases (see Table \ref{tab:table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in B\&S Table 2.

Thus a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in B\&S Table 2 are not comparable to one another, and conclusions drawn from such comparison are potentially incorrect.

\hypertarget{incorrect-scoring-of-the-2016-forecasts}{%
\subsection{Incorrect scoring of the 2016 forecasts}\label{incorrect-scoring-of-the-2016-forecasts}}

B\&S present a confusion matrix to score their 2016-H1 forecasts in Table 4. Although the forecasts are for the probability of civil war \emph{onset}, they are actually scored using \emph{incidence} of civil war. By definition the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict. When forecasting rare events, it is common to have many false positives, i.e.~instances where we predict onset but no onsets occur. \emph{Ex ante}, using incidence rather than onset to score onset forecasts should decrease the number of false positives and correspondingly increase the number of true positives (good predictions) and false negatives (missed onsets).

\hypertarget{results-of-the-updated-analysis}{%
\section{Results of the updated analysis}\label{results-of-the-updated-analysis}}

We now turn to an examination that addresses and fixes the five issues discussed above.\footnote{The code for all of our analysis undertaken for this effort may be found at \url{https://github.com/andybega/Blair-Sambanis-replication}.} The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B\&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B\&S Table 2). We will review the substantive implications of our updated analysis below, but the bottom line is that these changes turn B\&S's conclusions on their heads.

\hypertarget{smoothed-roc-curves-and-auc-calculations}{%
\subsection{Smoothed ROC curves and AUC calculations}\label{smoothed-roc-curves-and-auc-calculations}}

A reference to smoothing is made in a single sentence in B\&S (p.~12):

\begin{quote}
Figure 1 displays the corresponding ROC curves, \emph{smoothed for ease of interpretation}. (emphasis added)
\end{quote}

This suggest that using smoothed ROC curves will have no substantive impact on the results; however, in this smoothing determines the comparative results. Their statement also implies that the ROC curves were only smoothed in the referenced Figure 1. However, this is not the case. All AUC-ROC calculations throughout their replication code use an option to smooth the ROC curves \emph{prior} to AUC calculation.

ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like. Figure \ref{fig:rocs} shows our replication of both the estimated smoothed ROC curves from the B\&S report (left-hand side) and the actual empirical ROC curves (right-hand side). The standard method is to compute the area under the curve (AUC) statistic on the original, empirical ROC curves that are shown on the right.

\begin{figure}
\caption{Replication of B\&S Figure 1 with both smooth and non-smoothed ROC curves\label{fig:rocs}}
\centering
\includegraphics[width=.95\linewidth]{figures/fig-1-figure1-replicated.png}
\end{figure}

The specific predictions also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the \emph{y}-axis only changes when the prediction for an observed positive case is reached. For these ROC curves, and in the fundamental train/test split used for 12 of the 18 rows/models in B\&S Table 1, for that matter, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct \emph{y} coordinates. Notice also that the smoothing averages the left-most almost straight line with the right-most almost straight line in a monotonic way. Thus, smoothing changes the ordering of alternatives. It is not a cosmetic change as suggested.

What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B\&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B\&S Table 2)? Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}). The vertical bar in each plot marks the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to ``PITF Only'', smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible on average. The Escalation model, however, \emph{always} benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B\&S report, namely that the escalation model is generally superior.

\begin{figure}
\caption{Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\&S Tables 1 and 2). The Escalation model is the only model which consistently benefits from using smoothed ROC curves, and this advantage accounts for its apparent superiority to the other models.\label{fig:benefit-plot}}
\centering
\includegraphics[width=.8\linewidth]{figures/fig-2-benefit-plot.png}
\end{figure}

In sum, it is not only the case that using smoothed ROC curves alters the results, but also that the use of smoothed ROC curves to calculate AUC-ROC values \textbf{only} benefits the escalation model. It does so consistently and by a considerable margin. All eight alternative models reported in B\&S Tables 1 and 2, on average, do not gain when using smoothed ROC curves to calculate AUC. Four of the eight show a small positive bias from smoothing, while the other four display a small negative bias. The magnitudes of these are not close to the level of positive bias found in the escalation model.

\hypertarget{is-the-escalation-model-superior-to-the-alternative-icews-models}{%
\subsection{Is the Escalation model superior to the alternative ICEWS models?}\label{is-the-escalation-model-superior-to-the-alternative-icews-models}}

Table \ref{tab:table1-nosmooth} (B\&S Table 1) shows the comparison of the Escalation model to other alternative models based on ICEWS event data indicators. The AUC-ROC values are based on original, non-smoothed ROC curves. Both the Average and CAMEO models outperform the Escalation model in almost all instances (see also Figure \ref{fig:table1-plot}). The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B\&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves.

\begin{table}

\caption{\label{tab:table1-nosmooth}Comparison of the escalation model to alternative ICEWS model using test set AUC-ROC, \textit{without} smoothed ROC curves (Replication of B\&S Table 1)}
\centering
\begin{tabular}[t]{lrrrrr}
\toprule
Model & Escalation & Quad & Goldstein & CAMEO & Average\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{One-month forecasts}}\\
\hspace{1em}Base specification & 0.78 & 0.78 & 0.80 & 0.81 & 0.82\\
\hspace{1em}Terminal nodes & 0.79 & 0.78 & 0.79 & 0.81 & 0.82\\
\hspace{1em}Sample size & 0.79 & 0.80 & 0.74 & 0.82 & 0.84\\
\hspace{1em}Trees per forest & 0.78 & 0.78 & 0.79 & 0.81 & 0.82\\
\hspace{1em}Training/test sets 1 & 0.77 & 0.76 & 0.77 & 0.79 & 0.80\\
\hspace{1em}Training/test sets 2 & 0.75 & 0.77 & 0.74 & 0.76 & 0.78\\
\hspace{1em}Training/test sets 3 & 0.71 & 0.79 & 0.69 & 0.72 & 0.74\\
\hspace{1em}Coding of DV 1 & 0.80 & 0.80 & 0.80 & 0.82 & 0.83\\
\hspace{1em}Coding of DV 2 & 0.80 & 0.82 & 0.77 & 0.83 & 0.81\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Six-month forecasts}}\\
\hspace{1em}Base specification & 0.77 & 0.79 & 0.83 & 0.78 & 0.81\\
\hspace{1em}Terminal nodes & 0.77 & 0.78 & 0.82 & 0.78 & 0.79\\
\hspace{1em}Sample size & 0.77 & 0.77 & 0.80 & 0.80 & 0.82\\
\hspace{1em}Trees per forest & 0.77 & 0.79 & 0.83 & 0.79 & 0.81\\
\hspace{1em}Training/test sets 1 & 0.75 & 0.79 & 0.82 & 0.77 & 0.80\\
\hspace{1em}Training/test sets 2 & 0.70 & 0.75 & 0.78 & 0.75 & 0.77\\
\hspace{1em}Training/test sets 3 & 0.85 & 0.72 & 0.84 & 0.72 & 0.81\\
\hspace{1em}Coding of DV 1 & 0.78 & 0.80 & 0.83 & 0.79 & 0.82\\
\hspace{1em}Coding of DV 2 & 0.80 & 0.78 & 0.84 & 0.80 & 0.81\\
\bottomrule
\end{tabular}
\end{table}

The key difference between Table \ref{tab:table1-nosmooth} and B\&S Table 1 is whether the underlying ROC curves were original or smoothed; the test set N and coding issues did not affect this set of results.\footnote{We should note that in our replication we consciously decided to not set RNG seeds, even though the random forest models are non-deterministic and vary slightly from run to run. As we changed the replication code to allow running in parallel, we cannot exactly reproduce the B\&S results even with the same RNG seed. More importantly, the substantive interpretation of results should not depend on a specific RNG seed. We found that the escalation model's AUC-ROC values generally fluctuate no more than 0.01 (see \url{https://github.com/andybega/Blair-Sambanis-replication/blob/master/rep_nosmooth/variance.md}), and thus are confident that the patterns we see are robust to the initial RNG state.} B\&S's original results and interpretation regarding the superiority of the B\&S model over alternative ICEWS models, including the 1,000+ covariate CAMEO model, are thus entirely conditional on the use of smoothed AUC-ROC. Even the Quad model is generally as good as the escalation model in all implementations and is frequently quite a bit better.

Aside from the visual impact of ROC smoothing, the underlying motivation and methods involved in smoothing are broader and involve parametric estimation of population ROC curves (Hanley 1988, 2014). The main application is in the evaluation of medical tests (e.g.~diagnostic imaging) with early methods dating back several decades and ongoing development of both parametric and non-parametric smoothing methods (e.g.~Zou, Hall, and Shapiro 1997; Pulit 2016). It is not established which methods are preferable in a given application (Hanley 2014). Notably, the context in which ROC smoothing is used differs in important aspects from typical conflict research applications. Conflict data are closer to a census of the population and with well-known dependencies like spatial and temporal correlation, rather than a random sample that is approximately independently and identically distributed. It is thus neither clear that smoothing is justified nor valid. We also note that smoothed ROC plots are not widely used outside of the narrow context mentioned above.

\hypertarget{using-inconsistent-ns}{%
\subsection{Using Inconsistent Ns}\label{using-inconsistent-ns}}

The top portion of Table \ref{tab:table2-N} shows the number of valid test set predictions that can be scored for each model in the original B\&S Table 2. For the 1-month data version, the number of predictions differs by up to 500 cases, and in the 6-month data version by around 100 cases. These numbers appear small enough to be not important. However, the already small number of positive cases also is affected substantially: ``With PITF Predictors'' and ``PITF Only'' models lose two or one (respectively) of 11 positive cases in the 1- and 6-month data versions.

\begin{table}

\caption{\label{tab:table2-N}Number of valid test predictions for the escalation and structural alterantives models (B\&S Table 2)}
\centering
\begin{tabular}[t]{ll>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
\toprule
 & Horizon & Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
\midrule
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Original model-specific cases}}\\
\hspace{1em} & 1 month & 13748 & 13155 & 13461 & 13748 & 13510\\
\cmidrule{2-7}
\hspace{1em} & 6 months & 2366 & 2264 & 2317 & 2366 & 2265\\
\cmidrule{1-7}
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Cases adjusted to common subset}}\\
\hspace{1em} & 1 month & 13062 & 13062 & 13062 & 13062 & 13062\\
\cmidrule{2-7}
\hspace{1em} & 6 months & 2250 & 2250 & 2250 & 2250 & 2250\\
\bottomrule
\end{tabular}
\end{table}

Generally, these comparisons require the same cases. Therefore, we use predictions for a common joint subset of non-missing predictions. The resulting numbers of cases for each model are shown in the bottom portion of the table. Since the sets of incomplete cases in the original version above do not entirely overlap themselves, the common subset for all models is slightly smaller than the smallest N in the top portion of the table.

\hypertarget{do-structural-variables-add-to-the-escalation-models-predictive-power}{%
\subsection{Do Structural Variables Add to the Escalation Model's Predictive Power?}\label{do-structural-variables-add-to-the-escalation-models-predictive-power}}

Table \ref{tab:table2-fixed} shows our replication of B\&S Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed ``Weighted by PITF'' and ``PITF Split-Population'' models, and (3) AUC-ROC values computed on the common, joint subset of tests cases for which all models have non-missing predictions. Table \ref{tab:table2-full} in the appendix shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.

\begin{table}

\caption{\label{tab:table2-fixed}Comparison of escalation model to structural extensions, using test set AUC-ROC (Replication of B\&S Table 2)}
\centering
\begin{threeparttable}
\begin{tabular}[t]{r>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{}p{2cm}}
\toprule
Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
\midrule
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{One-month forecasts}}\\
\hspace{1em}0.75 & 0.78 & 0.78 & 0.85 & 0.75\\
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Six-month forecasts}}\\
\hspace{1em}0.76 & 0.86 & 0.80 & 0.77 & 0.74\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item \textit{Note: } 
\item Differences from the original B\&S Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.
\end{tablenotes}
\end{threeparttable}
\end{table}

B\&S interpret their results as follows (page 20) and we comment on the conclusions seratim:\footnote{We list the ``Overall, \ldots{}'' interpretation out-of-order for clarity.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  B\&S: ``Of the approaches we test, the split-population analog is most promising \ldots{}''
\end{enumerate}

The PITF split-population analog still performs well, but the simple Escalation + PITF predictors model performs better still.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  B\&S: ``Adding PITF predictors improves the performance of the escalation model over six-month windows but diminishes it over one-month windows \ldots ''
\end{enumerate}

Adding PITF predictors actually improves performance in both cases; the ``With PITF Predictions'' model strictly dominates the ``Escalation Only'' model by quite a margin in the 6-month data version.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  B\&S: ``The weighted model performs very poorly regardless \ldots ''
\end{enumerate}

The weighted model performs roughly on par with the Escalation Only model. One finding that remains is that the ``PITF Only'' model is outperformed by the ``Escalation Only'' model. As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is neither surprising, nor especially noteworthy.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  B\&S: ``Overall, our results suggest that while measures of structural risk may improve predictive performance, the value they add is marginal and inconsistent. {[}\ldots{]} Incorporating PITF thus significantly reduces or only slightly improves the performance of the escalation model, regardless of the approach we take. {[}\ldots{]}''
\end{enumerate}

The most straightforward method of incorporating the annual structural PITF variables---adding them to the predictors of the Escalation RF model---strictly outperforms the Escalation Only model. Note that the two other combination models considered are both non-standard and that the ``PITF Split Population'' model does not incorporate structural information at all, yet they also both do well. We thus conclude that adding structural variables improves predictive performance.

\hypertarget{how-accurate-were-the-2016-forecasts}{%
\subsection{How Accurate Were the 2016 Forecasts?}\label{how-accurate-were-the-2016-forecasts}}

B\&S forecast the risk of civil war onset for the first half of 2016 (2016-H1, where H2 would be the second half), and assessed the accuracy of these forecasts. The forecasts themselves are probabilistic, i.e.~range from 0 to 1, but for the purpose of scoring the forecasts they consider the 30 highest predictions to have been positive forecasts (``1'') and the remaining forecasts to have been for no onset (``0''). They evaluate two alternative codings of observed civil war outcomes, one in which 15 countries, including Colombia, had ongoing civil wars at the end of 2015 (labeled ``Assuming Persistence''), and an alternative coding in which the civil war in Colombia ends but Turkey and Benin experience civil war onsets (``Assuming Change'').\footnote{\label{fn:leadDV}The former coding is in the main replication data, while the latter is hand-coded in the Stata script that produces their confusion matrices. There is also some potential ambiguity about the discussion in this section because the forecasts B\&S claim are for the first half of 2016 are actually for the second half of 2016, but scored against outcomes for the first half of 2016.} However, when they score the forecasts to produce the confusion matrices they report in their Table 4, they score them using civil war incidence, not onset. In fact, this a conscious choice B\&S seem to acknowledge in the text (if not the table with confusion matrices), but it's not clear to us how this is justifiable.

For reference for the following discussion, we reproduce confusion matrices for the 2016-H1 forecasts in Table \ref{tab:table4}, but distinguish between civil war onsets (\(Y=1\)), non-events (\(Y=0\)) that correspond to countries with an ongoing civil war, and non-events (\(Y=0\)) that correspond to countries without a civil war, but which could potentially experience a civil war onset. The original B\&S confusion matrices collapse the second two rows in each matrix and treat ongoing and onset as part of the same group. Essentially all of the apparent true positives (\(P=1, Y=1\)) in the original B\&S confusion matrices are actually ongoing civil war cases which by definition \emph{cannot} experience another civil war onset. The alternative outcome coding B\&S evaluate, under ``Assuming Change'', produces the only two civil war onsets we would have observed in 2016-H1.

\begin{table}
\caption{Replication of B\&S Table 4: Confusion Matrices for Six-month Escalation Model forecasts for the first half of 2016.}\label{tab:table4}
\centering
\begin{tabular}{ll|c|c|l}
\multicolumn{1}{l}{\multirow{5}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$, no civil war & 134 & 17 \\
\cline{3-4}
& $Y=0$, ongoing & 2 & 13 \\
\cline{3-4}
& $Y=1$, onset & 0 & 0 \\
\cline{3-4}
% fixed, assuming change
\multicolumn{1}{l}{\multirow{5}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$, no civil war & 134 & 16 \\
\cline{3-4}
& $Y=0$, ongoing & 2 & 12 \\
\cline{3-4}
& $Y=1$, onset & 0 & 2 \\
\cline{3-4}
\end{tabular}
\end{table}

The fact that countries with an ongoing civil war cannot experience another civil war onset follows from the way the Sambanis civil war data conceptualize and operationalize civil war. Civil war appears to be a binary property of a country at any given moment in time, i.e.~a country at any given point is either at civil war or not, and onset is the first time period corresponding to this state. In the data, ongoing civil wars are coded as having a missing outcome value, and dropped.

If we place ourselves back to the beginning of 2016, when the forecasts were made (the forecasts are pre-registered in early March), this means that we would have observed that there are 15 countries that had ongoing conflicts in the second half of 2015 which had not terminated prior to the forecasting period. In all previous analyses in the paper, these kinds of cases drop out of the data. However, B\&S re-introduce, produce forecasts for them, and include them as positive events (``1''s) in their confusion matrices. They write, on page 21:

\begin{quote}
Importantly, we make no assumptions about the persistence of civil war between the second half of 2015 and the first half of 2016, meaning that all countries---even those with ongoing conflicts as of December 31, 2015--are included in the 2016 test set. As a result, 14 of the 30 countries on the list (and nine of the top 10) are cases of ongoing conflict.
\end{quote}

It's not clear to us under what conditions forecasting civil war onset risk for countries with ongoing civil wars make sense. A
conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, \emph{we would have already known that}. Rather than not making any ``assumptions about the persistence of civil war'', this would only make sense if we in fact assume that all ongoing civil wars ended in 2015-H2. Since Sambanis is the author of the civil war data that is used, it would already have been known in early 2016 that these 15 civil wars had in fact not terminated, and were still ongoing. It is thus both by definition impossible for countries with ongoing civil wars to experience another onset, and at the time the forecasts were made it would have been known that all conflicts were still ongoing up through the beginning of the forecasting period.

Does it make sense to count a high probability for the risk of onset for a case with an ongoing civil war as a partial or complete ``win'' for the model? When creating the confusion matrices, B\&S treat these ongoing civil war instances equivalent to civil war onsets, as positive events (``1''). In the text, B\&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases. This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g.~two separatists conflicts in different portions of the country. But this is not the case; ongoing conflicts in all other analyses are consistently dropped from the data as if a new civil war onset for a country with an ongoing conflict is conceptually impossible.

If we turn back to the confusion matrices to re-assess the accuracy of the 2016-H1 forecasts, we can thus conclude that almost all of the predictions are false positives. There are no onsets in the ``Assuming Persistence'' coding, and two onsets in the ``Assuming Change'' coding. This is normal. Civil war onsets are infrequent, and it is to be expected that forecasts of rare events will mostly consist of false positives. The forecasts do correctly capture the 2 civil war onsets in the ``Assuming Change'' version of the outcome coding. This is, at face value, a credit to the model's accuracy.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

B\&S advocate the use of theory to guide prediction. But ``theory'' is not defined by B\&S and may be an ambiguous and undefined concept. ``Theory'' has many interpretations in social science. For some---the EITM movement, for example (Aldrich, Alt, and Lupia 2008)---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced. The additional statements can be subjected to empirical scrutiny. An example might be William Riker's theory of coalitions (Riker 1962) which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically. A more recent example of strong theory is found in \emph{The Political Logic of Survival} (Bueno de Mesquita et al. 2005). For the purposes of discussion we call this strong theory.

However, for many, a theory is simply a falsifiable story or narrative that makes causal claims about the relationships among variables. Typically it is fused together from a meta-analysis of prior work on the topic under consideration. We call these types of theory weak theory. They are weak because they do not include two characteristics of strong theory, though they may include one:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A theory is separate from the case or cases being examined. It is intended to be general.
\item
  A theory generates new statements that in principle can be empirically falsified, not proven.
\end{enumerate}

Most social science research uses a weak form of theory. In the case of B\&S they base their theoretical thinking on research that argues that conflict is not the product of structural factors, but rather comes from an interaction of state repression and the changes in costs and benefits to citizens for engaging in anti-government conflict. Following a wide swath of scholars, they argue that the escalation and de-escalations of interactions between citizens and government are what typically lead up to the onset of civil wars. Operationally, this boils down to four types of concepts: demands, accommodations, nonviolent repression, and low-level violence.

\begin{itemize}
\item Demands include events in which either opposition or rebels is the source and government is the target and typically involve appeals for political or institutional reform, policy change, or extension of rights.
\item Nonviolent repression and accommodation include events in which the government is the source and either opposition or rebels is the target. Nonviolent repression includes threats, sanctions, restrictions on freedom of movement, or rejections of group demands.
\item Low-level violence includes all events in which (1) the government is the source and opposition is the target, (2) the opposition is the source and government is the target, (3) the government is the source and rebels are the target, or (4) rebels are the source and government is the target.
\end{itemize}

These variables, taken from ICEWS, are the main components of the escalation model in the B\&S article. The escalation theory is that the forces behind these variables interact such that as each of them increases, they independently increase the probability of civil conflict. The actual mechanisms of the dynamism among these variables are not specified.

Indeed, most empirical social science research follows this approach: (1) delineation of an argument, (2) selection of variables that represent a plausible or compelling representation of the concepts, followed by (3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation. It is worth pointing out that this is simply one set of variables that are selected to represent the concepts in the theoretical story.

A strong theory gives more guidance on compelling choices than does a weak theory. Accordingly, while the choice is plausible it is not necessarily compelling. Another choice might be equally plausible. In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups. In practical terms there is a wide array of choices that are available to instantiate any particular theory. A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated dynamics. One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.

Instantiating the theoretical model is more about choosing the correct variables. This is the contrast between theoretical prediction models and atheoretical machine learning that B\&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B\&S might characterize as atheoretical enter the ``theoretical'', with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

Furthermore, if what we mean by the concept ``theory'' is such that it suggests a large range of variables that could be related to an outcome, what do we really gain from having this theory in our predictive model? We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off? Certainly that's not what B\&S are trying to suggest since this would be a problematic way to ``evaluate'' a theory. But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could in fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated ``theory'' with any eye towards out-of-sample predictive accuracy (cf.~Colaresi and Mahmood 2017 for how to avoid this problem while leveraging out-of-sample predictions). Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize and whose underlying theories may by extension not be externally valid (for a more detailed treatment of this issue, see Fariss and Jones 2018).

Conversely, not all factors that are ``theoretically'' important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both Ward, Greenhill, and Bakke (2010) and Hill and Jones (2014) examine a variety of variables that are considered ``theoretically'' important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes. And, it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.~the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al. 2010), irregular leadership changes (Beger et al. 2017), and coups (Beger and Ward 2020).

Aside from selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations that dominate published regression analyses, maybe with the occasional polynomial or multiplicative interaction terms, are \emph{the} correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are those relationships linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B\&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives. But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory (see also Fariss and Jones 2018)?

Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider the ways in which they can mutually reinforce each other. Even if the analysis in B\&S fails to support the point, we concede that predicting on the basis of a strong theoretical model is preferable to inductive prediction. Beyond accurate predictions, such models would allow us to make causal statements that are harder to make on the basis of ad hoc models. But this is hard to do in practice. The problem is that many extant theoretical models are in fact so malleable to interpretation that they essentially are ad hoc. They are useful for identifying variables that could be useful for prediction, and this is in fact already taking place.

Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning. Even if B\&S were correct that in their study the theory based escalation model performed better, it would simply be a case study of a single case where it was preferable. It establishes no general tendency, let alone law-like evidence for general theory. Nor, against it.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

We encountered several issues in the code underlying the B\&S analysis. The problems we encountered are not subjective modeling choices. They bear no implications on the question of whether theory based model are preferable. However, when we fix these issues and perform an updated analysis, the B\&S conclusions are all essentially overturned. In other words, B\&S findings are based on a faulty analysis, invalid, and should be disregarded as evidence in favor of theory-based models.

In contradistinction to the conclusions offered by B\&S, we find that when correctly specified and implemented:

\begin{itemize}
\tightlist
\item
  The theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model and the ensemble average model.
\item
  The Average ensemble model and the CAMEO models outperform the escalation model in all instances.
\item
  Adding structural variables substantially improves the escalation model's performance.
\end{itemize}

More importantly, trying to contrast machine learning and theory may be misguided. Even if a ``theory'' section is missing, predictive models are informed by previous work and arguments. If we had strong general theories of something like civil war onset, we could replace more inductive, ad hoc, methods with clearly specified models. But we do not. What we have instead are weak theories that could be interpreted to suggest any range of equations for predicting an outcome. Instead of trying to find an instantiation of such a ``theoretical'' model that can outperform a more ad hoc alternative, we could use the flexibility of machine learning methods and insights that predictive modeling brings to improve our weak theories.

Blair \& Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain. As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance. However, these comparisons must be made carefully to ensure those correct inferences are drawn. We continue to think that weak theory is overrated (Ward 2016), and that machine learning and big data will allow us to learn new things. It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation. Social science has leaned on theory for (at least) seventy years, but it is time to find useful ways to incorporate recently developed methods in the arena of machine learning and modern statistics.

\newpage

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-aldrich:etal:2008}{}%
Aldrich, John H., James E. Alt, and Arthur Lupia. 2008. ``The Eitm Approach: Origins and Interpretations.'' In \emph{The Oxford Handbook of Political Methodology}.

\leavevmode\hypertarget{ref-beger:etal:2017}{}%
Beger, Andreas, Daniel W. Hill, Jr., Nils. W. Metternich, Shahryar Minhas, and Michael D. Ward. 2017. ``Splitting It Up: The \texttt{spduration} Split-Population Duration Regression Package for Time-Varying Covariates.'' \emph{The \(\mathcal{R}\) Journal} 9 (2): 474--86. \url{https://journal.r-project.org/archive/2017/RJ-2017-056/index.html}.

\leavevmode\hypertarget{ref-beger:etal:2020}{}%
Beger, Andreas, Richard K. Morgan, and Laura Maxwell. 2020. ``The Democratic Spaces Barometer: Global Forecasts of Autocratization and Democratization.'' \url{https://www.v-dem.net/en/analysis/DemSpace/}.

\leavevmode\hypertarget{ref-beger:ward:2017}{}%
Beger, Andreas, and Michael D. Ward. 2017. ``Lessons from Near Real-Time Forecasting of Irregular Leadership Changes.'' \emph{Journal of Peace Research} 54 (2): 141--56.

\leavevmode\hypertarget{ref-beger:ward:2020}{}%
---------. 2020. ``Coup Forecasts for 2020.'' \url{https://www.predictiveheuristics.com/forecasts}.

\leavevmode\hypertarget{ref-blair:sambanis:2020}{}%
Blair, Robert A., and Nicholas Sambanis. n.d. ``Forecasting Civil Wars: Theory and Structure in an Age of `Big Data' and Machine Learning.'' \emph{Journal of Conflict Resolution} TBD (firstview): 0022002720918923.

\leavevmode\hypertarget{ref-breiman:2001}{}%
Breiman, Leo. 2001. ``Random Forests.'' \emph{Machine Learning} 45 (1): 5--32.

\leavevmode\hypertarget{ref-bdm:etal:2005}{}%
Bueno de Mesquita, Bruce, Alastair Smith, Randolph M. Siverson, and James D. Morrow. 2005. \emph{The Logic of Political Survival}. Cambridge, Mass: MIT Press.

\leavevmode\hypertarget{ref-cederman:weidmann:2017}{}%
Cederman, Lars-Erik, and Nils B. Weidmann. 2017. ``Predicting Armed Conflict: Time to Adjust Our Expectations?'' \emph{Science} 355 (6324): 474--76.

\leavevmode\hypertarget{ref-chiba:etal:2015}{}%
Chiba, Daina, Nils W. Metternich, and Michael D. Ward. 2015. ``Every Story Has a Beginning, Middle, and an End (but Not Always in That Order): Predicting Duration Dynamics in a Unified Framework.'' \emph{Political Science Research and Methods} 3 (3): 515--41.

\leavevmode\hypertarget{ref-colaresi:mahmood:2017}{}%
Colaresi, Michael, and Zuhaib Mahmood. 2017. ``Do the Robot: Lessons from Machine Learning to Improve Conflict Forecasting.'' \emph{Journal of Peace Research} 54 (2): 193--214.

\leavevmode\hypertarget{ref-fariss:jones:2018}{}%
Fariss, Christopher J., and Zachary M. Jones. 2018. ``Enhancing Validity in Observational Settings When Replication Is Not Possible.'' \emph{Political Science Research and Methods} 6 (2): 365--80.

\leavevmode\hypertarget{ref-goldstone:etal:2010}{}%
Goldstone, Jack A., Robert H. Bates, David L. Epstein, Ted Robert Gurr, Michael B. Lustik, Monty G. Marshall, Jay Ulfelder, and Mark Woodward. 2010. ``A Global Model for Forecasting Political Instability.'' \emph{American Journal of Political Science} 54 (1): 190--208.

\leavevmode\hypertarget{ref-henley:1988}{}%
Hanley, James A. 1988. ``The Robustness of the "Binormal" Assumptions Used in Fitting ROC Curves.'' \emph{Medical Decision Making} 8 (3): 197--203. \url{https://doi.org/10.1177/0272989X8800800308}.

\leavevmode\hypertarget{ref-henley:2014}{}%
---------. 2014. ``Receiver Operating Characteristic (ROC) Curves.'' In \emph{Wiley Statsref: Statistics Reference Online}. American Cancer Society. \url{https://doi.org/10.1002/9781118445112.stat05255}.

\leavevmode\hypertarget{ref-hegre:etal:2019}{}%
Hegre, Håvard, Marie Allansson, Matthias Basedau, Michael Colaresi, Mihai Croicu, Hanne Fjelde, Frederick Hoyles, et al. 2019. ``ViEWS: A Political Violence Early-Warning System.'' \emph{Journal of Peace Research} 56 (2): 155--74.

\leavevmode\hypertarget{ref-hill:jones:2014}{}%
Hill, Daniel W., and Zachary M. Jones. 2014. ``An Empirical Evaluation of Explanations for State Repression.'' \emph{American Political Science Review} 108 (03): 661--87.

\leavevmode\hypertarget{ref-liaw:wiener:2002}{}%
Liaw, Andy, and Matthew Wiener. 2002. ``Classification and Regression by randomForest.'' \emph{R News} 2 (3): 18--22. \url{https://CRAN.R-project.org/doc/Rnews/}.

\leavevmode\hypertarget{ref-malley:etal:2012}{}%
Malley, James D., Jochen Kruppa, Anirban DasGupta, Karen G. Malley, and Andreas Ziegler. 2012. ``Probability Machines: Consistent Probability Estimation Using Nonparametric Learning Machines.'' \emph{Methods of Information in Medicine} 51 (1): 74--81. \url{https://www.thieme-connect.de/products/ejournals/abstract/10.3414/ME00-01-0052}.

\leavevmode\hypertarget{ref-pulit:2016}{}%
Pulit, Michał. 2016. ``A New Method of Kernel-Smoothing Estimation of the Roc Curve.'' \emph{Metrika} 79 (5): 603--34.

\leavevmode\hypertarget{ref-riker:1962}{}%
Riker, William H. 1962. \emph{The Theory of Political Coalitions}. Yale University Press.

\leavevmode\hypertarget{ref-ward:2016}{}%
Ward, Michael D. 2016. ``Can We Predict Politics? Toward What End?'' \emph{Journal of Global Security Studies} 1 (1): 80--91.

\leavevmode\hypertarget{ref-ward:etal:2010}{}%
Ward, Michael D., Brian D. Greenhill, and Kristin M. Bakke. 2010. ``The Perils of Policy by P-Value: Predicting Civil Conflicts.'' \emph{Journal of Peace Research} 47 (4): 363--75.

\leavevmode\hypertarget{ref-ward:etal:2013}{}%
Ward, Michael D., Nils W. Metternich, Cassy L. Dorff, Max Gallop, Florian M. Hollenbach, Anna Schultz, and Simon Weschle. 2013. ``Learning from the Past and Stepping into the Future: Toward a New Generation of Conflict Prediction.'' \emph{International Studies Review} 16 (4): 473--644.

\leavevmode\hypertarget{ref-zou:etal:1997}{}%
Zou, Kelly H., William J. Hall, and David E. Shapiro. 1997. ``Smooth Non-Parametric Receiver Operating Characteristic (ROC) Curves for Continuous Diagnostic Tests.'' \emph{Statistics in Medicine} 16 (19): 2143--56.
\end{cslreferences}

\newpage

\hypertarget{appendix-appendix}{%
\appendix}

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}

\renewcommand{\thetable}{A\arabic{table}} 
\setcounter{table}{0}

\hypertarget{code-references-for-coding-errors}{%
\subsection{Code references for coding errors}\label{code-references-for-coding-errors}}

\hypertarget{section-3.2-incorrect-weighted-by-pitf-implementation}{%
\subsubsection{Section 3.2: Incorrect ``Weighted by PITF'' Implementation}\label{section-3.2-incorrect-weighted-by-pitf-implementation}}

We believe the intention with the ``Weighted by PITF'' model is to combine the PITF test predictions with the Escalation test predictions, i.e.~weight the latter with the former. Due to an error, the PITF \emph{training} set predictions are used, not the \emph{test} set predictions.

See \texttt{1mo\_run\_escalation\_weighted\_PITF.R} line 4, where the PITF predictions are taken from the training data set (\texttt{train\$pred\_prob\_plus1}). The next line is a hack extending the shorter \texttt{weight} vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly in the 6-month version of this file.

Uncovering which cases in the PITF training predictions are matched to the cases that the Escalation model test set predictions correspond to requires tracing back several more data objects. We have documented details in issue \#3 in the GitHub repo.

\hypertarget{section-3.3-incorrect-pitf-split-population-implementation}{%
\subsubsection{Section 3.3: Incorrect ``PITF Split Population'' Implementation}\label{section-3.3-incorrect-pitf-split-population-implementation}}

The intention is to train separate random forests on data split into high and low-risk groups based on PITF predictions, and then combine them back into one random forest. Actually both random forests are trained on identical data, essentially production a random forest that just has N\_trees * 2.

Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined. Details are documented in issue \#5 in the GitHub repo.

\hypertarget{section-3.5-incorrect-scoring-of-the-2016-forecasts}{%
\subsubsection{Section 3.5: Incorrect scoring of the 2016 forecasts}\label{section-3.5-incorrect-scoring-of-the-2016-forecasts}}

Regarding the claim in footnote \ref{fn:leadDV} that the 2016-H1 forecasts are actually forecasts for 2016-H2:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Throughout the R code for the main analyses, the models predict a 1-time period lead version of the civil war onset variables (ending in "\_plus1``). This means that, for example, in the 6-month version of the data, a value of''1" for the "\_plus1" outcome variable for 2015-H2 actually refers to a civil war onset in the first half of 2016. One can confirm, through some tracing to work back to the original variable assignments, that the ROC curves and AUC are correspondingly, and correctly, also scored using the lead versions of the DV.
\item
  The nominally 2016-H1 forecasts are created at the bottom of \texttt{+master.R} using the data for 2016-H1 and still the lead DV. Thus the lead DV here refers to outcomes in 2016-H2.
\item
  The scoring of the forecasts, whatever time period they are actually for, is done in a separate Stata do file, \texttt{6mo\_make\_confusion\_matrix.do}. Here, the scoring is done using the non-lead version of the DV. Thus there is a mismatch between the time period the forecasts are for and the period who's observed outcomes are used to score them.
\end{enumerate}

The relevant variables in the data are ``incidence\_civil\_ns'' and ``incidence\_civil\_ns\_plus1'', which appears to be a 1-period lead (ie., \(t_+1\)) version of the dependent variable that is used in the actual prediction models. The incidence dependent variable contains both 0/1 and missing values. By examining the pattern of missing values, it seems clear that this was originally an incidence variable indicating whether a country was at civil war in a given year or not, and which was converted to an onset version so that onsets retain the value of 1 but continuing civil war years are coded as missing. This reflects common practice.

Furthermoe, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset. In the file \texttt{6mo\_make\_confusion\_matrix.do} on line 52, missing values in ``incidence\_civil\_ns'' are recoded to 1, thus reverting the onset coding of this variable to incidence.

\hypertarget{additional-replication-tables}{%
\subsection{Additional replication tables}\label{additional-replication-tables}}

Table \ref{tab:table1-smooth} is our replication of B\&S Table 1 with smoothed AUC-ROC. The results differ slightly from the original B\&S Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models. It is the case that B\&S set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel. In any case, the interpretation of results should not be sensitive to random variation, i.e.~it should not depend on using a specific RNG seed. On the basis of these results, B\&S conclude that the escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.

\begin{table}

\caption{\label{tab:table1-smooth}Replication of B\&S Table 1 with smoothed ROC curves; test set AUC-ROC for various models}
\centering
\begin{tabular}[t]{lrrrrr}
\toprule
Model & Escalation & Quad & Goldstein & CAMEO & Average\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{One-month forecasts}}\\
\hspace{1em}Base specification & 0.85 & 0.80 & 0.79 & 0.84 & 0.83\\
\hspace{1em}Terminal nodes & 0.86 & 0.79 & 0.77 & 0.83 & 0.82\\
\hspace{1em}Sample size & 0.85 & 0.81 & 0.71 & 0.86 & 0.84\\
\hspace{1em}Trees per forest & 0.85 & 0.80 & 0.78 & 0.83 & 0.82\\
\hspace{1em}Training/test sets 1 & 0.86 & 0.78 & 0.76 & 0.81 & 0.80\\
\hspace{1em}Training/test sets 2 & 0.82 & 0.79 & 0.72 & 0.77 & 0.78\\
\hspace{1em}Training/test sets 3 & 0.80 & 0.80 & 0.69 & 0.74 & 0.75\\
\hspace{1em}Coding of DV 1 & 0.86 & 0.81 & 0.80 & 0.84 & 0.83\\
\hspace{1em}Coding of DV 2 & 0.92 & 0.80 & 0.81 & 0.81 & 0.81\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Six-month forecasts}}\\
\hspace{1em}Base specification & 0.82 & 0.78 & 0.82 & 0.77 & 0.79\\
\hspace{1em}Terminal nodes & 0.79 & 0.76 & 0.81 & 0.77 & 0.77\\
\hspace{1em}Sample size & 0.83 & 0.78 & 0.78 & 0.79 & 0.79\\
\hspace{1em}Trees per forest & 0.82 & 0.77 & 0.82 & 0.77 & 0.79\\
\hspace{1em}Training/test sets 1 & 0.80 & 0.78 & 0.81 & 0.76 & 0.78\\
\hspace{1em}Training/test sets 2 & 0.72 & 0.74 & 0.77 & 0.73 & 0.75\\
\hspace{1em}Training/test sets 3 & 0.88 & 0.71 & 0.81 & 0.68 & 0.79\\
\hspace{1em}Coding of DV 1 & 0.83 & 0.77 & 0.82 & 0.79 & 0.80\\
\hspace{1em}Coding of DV 2 & 0.83 & 0.77 & 0.83 & 0.78 & 0.79\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}

\caption{\label{tab:table2-full}Replication of B\&S Table 2 with smoothed/original ROC and with original varying N cases or adjusting for common case set with constant N}
\centering
\begin{tabular}[t]{ll>{\raggedright\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
\toprule
 &  & Smoothed ROC & Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
\midrule
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{Original model-specific cases}}\\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textit{1 month}}\\
\hspace{1em}\hspace{1em} &  & Yes & 0.85 & 0.77 & 0.80 & 0.81 & 0.76\\
\cmidrule{3-8}
\hspace{1em}\hspace{1em} &  & No & 0.79 & 0.78 & 0.80 & 0.79 & 0.75\\
\cmidrule{2-8}
\addlinespace[0.3em]
\multicolumn{8}{l}{\textit{6 months}}\\
\hspace{1em}\hspace{1em} &  & Yes & 0.82 & 0.85 & 0.81 & 0.78 & 0.74\\
\cmidrule{3-8}
\hspace{1em}\hspace{1em} &  & No & 0.77 & 0.86 & 0.81 & 0.78 & 0.74\\
\cmidrule{1-8}
\addlinespace[0.3em]
\multicolumn{8}{l}{\textbf{Cases adjusted to common subset}}\\
\addlinespace[0.3em]
\multicolumn{8}{l}{\textit{1 month}}\\
\hspace{1em}\hspace{1em} &  & Yes & 0.81 & 0.77 & 0.79 & 0.86 & 0.76\\
\cmidrule{3-8}
\hspace{1em}\hspace{1em} &  & No & 0.75 & 0.78 & 0.78 & 0.85 & 0.75\\
\cmidrule{2-8}
\addlinespace[0.3em]
\multicolumn{8}{l}{\textit{6 months}}\\
\hspace{1em}\hspace{1em} &  & Yes & 0.82 & 0.85 & 0.80 & 0.77 & 0.74\\
\cmidrule{3-8}
\hspace{1em}\hspace{1em} &  & No & 0.76 & 0.86 & 0.80 & 0.77 & 0.74\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}

\caption{\label{tab:table1-benefit}Smoothing advantage for B\&S Table 1: the gain in AUC-ROC when calculated using smoothed ROC curves}
\centering
\begin{tabular}[t]{lrrrrr}
\toprule
Model & Escalation & Quad & Goldstein & CAMEO & Average\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{One-month forecasts}}\\
\hspace{1em}Base specification & 0.06 & 0.02 & -0.01 & 0.03 & 0.00\\
\hspace{1em}Terminal nodes & 0.07 & 0.01 & -0.01 & 0.02 & 0.00\\
\hspace{1em}Sample size & 0.06 & 0.01 & -0.04 & 0.04 & 0.00\\
\hspace{1em}Trees per forest & 0.06 & 0.02 & -0.01 & 0.02 & 0.00\\
\hspace{1em}Training/test sets 1 & 0.09 & 0.02 & -0.01 & 0.02 & 0.00\\
\hspace{1em}Training/test sets 2 & 0.07 & 0.02 & -0.01 & 0.02 & 0.00\\
\hspace{1em}Training/test sets 3 & 0.09 & 0.01 & 0.00 & 0.02 & 0.01\\
\hspace{1em}Coding of DV 1 & 0.07 & 0.02 & 0.00 & 0.02 & 0.00\\
\hspace{1em}Coding of DV 2 & 0.12 & -0.02 & 0.04 & -0.01 & 0.01\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Six-month forecasts}}\\
\hspace{1em}Base specification & 0.05 & -0.01 & -0.01 & -0.01 & -0.02\\
\hspace{1em}Terminal nodes & 0.02 & -0.02 & -0.01 & -0.01 & -0.02\\
\hspace{1em}Sample size & 0.05 & 0.00 & -0.02 & -0.01 & -0.03\\
\hspace{1em}Trees per forest & 0.05 & -0.01 & -0.01 & -0.02 & -0.02\\
\hspace{1em}Training/test sets 1 & 0.04 & -0.01 & -0.01 & -0.02 & -0.02\\
\hspace{1em}Training/test sets 2 & 0.02 & -0.01 & -0.02 & -0.01 & -0.02\\
\hspace{1em}Training/test sets 3 & 0.03 & -0.02 & -0.03 & -0.04 & -0.02\\
\hspace{1em}Coding of DV 1 & 0.05 & -0.03 & -0.01 & 0.00 & -0.02\\
\hspace{1em}Coding of DV 2 & 0.03 & -0.01 & -0.01 & -0.02 & -0.02\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}

\caption{\label{tab:table2-benefit}Smoothing advantage for B\&S Table 2: the gain in AUC-ROC when calculated using smoothed ROC curves}
\centering
\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
\toprule
Model & Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{One-month forecasts}}\\
\hspace{1em}Base specification & 0.06 & -0.01 & 0.01 & 0.01 & 0.01\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Six-month forecasts}}\\
\hspace{1em}Base specification & 0.06 & -0.01 & 0.00 & 0.00 & 0.00\\
\bottomrule
\end{tabular}
\end{table}

\clearpage

\hypertarget{model-to-model-comparison-plots}{%
\subsection{Model to model comparison plots}\label{model-to-model-comparison-plots}}

Figures \ref{fig:table1-plot} and \ref{fig:table2-plot} replicate the information in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} in a way makes the comparison of the escalation model to the alternative models easier. Each facet shows the escalation AUC-ROC for all model settings (the rows in Table \ref{fig:table1-plot} and base specification for all models in Table \ref{tab:table2-fixed}) on the left, and the AUC-ROC for an alternative model on the right, with a connecting line. If the lines slope up to the right, the alternative model is better.

\begin{figure}
{\centering
\includegraphics[width=.9\linewidth]{figures/fig-A1-table1-pairs.png}
}
\caption{Escalation to alternative comparisons for the ICEWS models (B\&S Table 1)\label{fig:table1-plot}}
\end{figure}

\begin{figure}
{\centering
\includegraphics[width=.9\linewidth]{figures/fig-A2-table2-pairs.png}
}
\caption{Escalation to alternative comparisons for the structural models (B\&S Table 2)\label{fig:table2-plot}}
\end{figure}

\newpage

\hypertarget{random-forest-hyper-parameter-selection}{%
\subsection{Random forest hyper-parameter selection}\label{random-forest-hyper-parameter-selection}}

What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated. Table \ref{tab:hp} shows the default values used by the implementation of random forest that B\&S use (from the \textbf{randomForest} R package), in contrast to the basic settings used by B\&S for most the models reported in the paper.

As the outcome is a binary indicator of civil war onset, one would typically use a classification random forest that predicts 0 or 1 labels directly. The implementation of random forests that B\&S use ((Liaw and Wiener 2002)) is based on the original Breiman (2001) implementation and calculates predictive probabilities by averaging over the ``0'' or ``1'' votes from all constituent decision trees. The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number. From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive, and operates on a large bootstrapped sample of the original training data. These are of course only heuristics and it is usual to attempt to find better hyper-parameter methods through some form of tuning.

\begin{table}
\caption{\label{tab:hp} Random forest (\texttt{randomForest()}) default versus B\&S hyperparameters}
\begin{tabular}{l>{\raggedright\arraybackslash}p{2in}ll}
\toprule
Hyperparameter & Default heuristic & Default values (Escalation) & B\&S value \\
\midrule
type & & classification & regression \\
ntree & & 500 & 100,000 or 1e6 \\
mtry & \texttt{floor(sqrt(ncol(x)))} & 3 & 3 \\
replace & & true & false \\
sampsize & \texttt{nrow(x)} if replace, else \texttt{ceiling(.632*nrow(x))} & 11,869 & 100 or
500 \\
nodesize & 1 for classification & 1 & 1 \\
maxnodes & & null & 5 or 10 \\
\bottomrule
\end{tabular}
\end{table}

B\&S in contrast fit very large forests with 100,000 trees in the basic model form, but where each tree only operates on a very small sub-sample (N=100 or 500), drawn without replacement, of the available training data. This approach only works due to the choice to use regression, not classification, trees. Trying to use classification trees with the other parameter settings does not work at all because it is almost guaranteed that a sample of 100 from the \textasciitilde11,000 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample. As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:

\begin{verbatim}
Warning message:
In randomForest.default(y = as.integer(train_df$incidence_civil_ns_plus1 ==  :
  The response has five or fewer unique values.  Are you sure you want to do regression?
\end{verbatim}

As it turns out, using regression random forests for this kind of binary classification problem in order to obtain probability estimates matches the probability random forest approach suggested and positively evaluated in Malley et al. (2012), and which is used in another prominent R implementation of random forests.\footnote{The \textbf{ranger} package.} It is not clear whether this is intentional, as the Malley paper is not cited in B\&S.

In any case, B\&S's random forest approach appears to work really well. We tried to construct classification random forests tuned via cross-validation on the training data set partition, i.e.~without touching the test data, but were unable to develop models that consistently match the B\&S random forest method in both cross-validated out-of-sample training predictions and test set predictions.

Given that they are relatively unorthodox, yet appear to work very well, we wonder how the hyper-parameter values were determined. Two specific concerns are that this was not done with an eye towards test set accuracy, which would invalidate the independence of the out-of-sample test set, and whether the specific hyper-parameter values are optimized for only one model, or were optimized and found to work well for all models. There is no discussion of the random forest tuning strategy or how the specific hyper-parameter methods were determined in the paper.

Given the dramatic changes in results as a result of the preceding issues, we did not further investigate these potential concerns.

\newpage

\hypertarget{different-smoothing-methods}{%
\subsection{Different smoothing methods}\label{different-smoothing-methods}}

The \textbf{pROC} package \texttt{smooth()} function includes several different smoothing methods. The default, which BS use in their code, is binormal smoothing. Here is a replication of the smooth benefits shown in Figure \ref{fig:benefit-plot} with two other smoothing methods. (We could not get ``logcondens'' and ``logcondens.smooth'' to work, and they are based on another external package that has not been updated since 2016.)

Only the ``binormal'' smoothing method produces a pattern of AUC-ROCs that clearly elevates the Escalation model above the alternatives. The ``density'' method does not on average change the AUC-ROC values very much from their empirical ROC curve equivalents. The ``fitdistr'' method tends to increase all AUC-ROC values, although maybe slightly more so for the Escalation model than the main alternatives.

\begin{figure}
{\centering \includegraphics[width=.95\linewidth]{figures/fig-a3-benefit-plot-extended.png}}
\caption{Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\&S Tables 1 and 2), with 3 different smoothing methods (binormal is the default method and used in B\&S)\label{fig:smooth-benefit-plot-extended}}
\end{figure}

\end{document}

### 2020-11-10 paper/paper.Rmd
drop small edits into .Rmd

@@
 
-Although their analysis in the end does not support this point, we do not disagree that theory can important in predictive modeling.
+Although their analysis in the end does not support this point, we do not disagree that theory can be important in predictive modeling.
 Of the various ways one could contrast ad hoc machine learning with "theoretical" modeling, B&S focus essentially on theory as a mechanism for variable selection.
@@
 It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well.
-Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical", with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}
+Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical" with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}
 
@@
 For example, both @ward:etal:2010 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes.
-And, it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.
+It is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.
 the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups [@beger:ward:2020].
@@
 However, these comparisons must be made carefully to ensure those correct inferences are drawn.
-We continue to think that weak theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things.
+We continue to think that weak theory is overrated [@ward:2016] and that machine learning and big data will allow us to learn new things.
 It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.

### 2020-11-10 paper/paper.tex
drop small edits into .Rmd

@@
 
-Although their analysis in the end does not support this point, we do not disagree that theory can important in predictive modeling.
+Although their analysis in the end does not support this point, we do not disagree that theory can be important in predictive modeling.
 Of the various ways one could contrast ad hoc machine learning with ``theoretical'' modeling, B\&S focus essentially on theory as a mechanism for variable selection.
@@
 It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well.
-Rather, we posit that most of the cases B\&S might characterize as atheoretical enter the ``theoretical'', with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}
+Rather, we posit that most of the cases B\&S might characterize as atheoretical enter the ``theoretical'' with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}
 
@@
 For example, both Ward, Greenhill, and Bakke (2010) and Hill and Jones (2014) examine a variety of variables that are considered ``theoretically'' important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes.
-And, it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.~the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al.
+It is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.~the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al.
 2010), irregular leadership changes (Beger et al.
@@
 However, these comparisons must be made carefully to ensure those correct inferences are drawn.
-We continue to think that weak theory is overrated (Ward 2016), and that machine learning and big data will allow us to learn new things.
+We continue to think that weak theory is overrated (Ward 2016) and that machine learning and big data will allow us to learn new things.
 It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.

### 2020-11-10 paper/paper.Rmd
some edits suggested by Grammarly

@@
 We find that there are several important mistakes and that their key finding is entirely conditional on the use of parametrically smoothed ROC curves when calculating accuracy, rather than the standard empirical ROC curves that dominate the literature.
-Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.
+Fixing these mistakes results in a reversal of their claim that theory-based models of escalation are better at predicting onsets of civil war than other kinds of models.
 Their model is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.
-More importantly, we argue that rather than trying to contrast the roles of theory and \"atheoretical\" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant predictive work.
+More importantly, we argue that rather than trying to contrast the roles of theory and \"atheoretical\" machine learning in predictive modeling, it would be more productive to focus on ways in which predictive modeling and machine learning could be used to strengthen extant predictive work.
 Instead, we argue that predictive modeling and machine learning are effective tools for theory testing.
-
 output: 
@@
 They do not provide much guidance on an exact specification of what variables and which functional form is required.
-As a result it is both not very clear what the "correct" model should be, nor easy to demonstrate that any particular "theoretical" model's ability to predict generalizes beyond the data that was used to create it in the first place.
+As a result, it is both not very clear what the "correct" model should be, nor easy to demonstrate that any particular "theoretical" model's ability to predict generalizes beyond the data that was used to create it in the first place.
 Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to "explain" a phenomenon to strengthen and whittle down the broad space of extant theories.
@@
 To examine the first question, B&S compare the test set of the escalation model to four alternative models.
-The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country.
+The independent variables for the first set of analyses reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country.
 The models are:
@@
 - CAMEO: counts for all CAMEO event codes over the four actor dyads, totaling $1,160$ covariates, which are mostly zero for any country in any month. 
-- Average: unweighted average of the predictions from the four models briefly described above.
+- Average: an unweighted average of the predictions from the four models briefly described above.
 
@@
 - Weighted by PITF: escalation model predictions weighted using the PITF instability model predictions.
-- PITF Split Population: the training data are split into high and low risk portions based on the PITF instability model predictions, two separate escalation random forests are trained on the splits, then re-combined into a single random forest that is used to create the test set predictions.
+- PITF Split Population: the training data are split into high- and low-risk portions based on the PITF instability model predictions, two separate escalation random forests are trained on the splits, then re-combined into a single random forest that is used to create the test set predictions.
 - PITF Only: a random forest model based only on the annual, structural PITF model predictors.
@@
 Finally, B&S used their escalation model to create forecasts for the first half of 2016.
-In their third analysis, they score the forecasts accuracy using subsequent civil war onset data.
+In their third analysis, they score the forecasts' accuracy using subsequent civil war onset data.
 This is summarized in B&S Table 3.
@@
 While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation.
-These are (a) the use of smoothed ROC curves to draw conclusions about which model is best, (b) incorrect implementations of the weighted by PITF and PITF analog split-population models, (c) inconsistent test sets for the models examined, and (d) incorrect scoring of the 2016 forecasts.^[We also note there are additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox.
+These are (a) the use of smoothed ROC curves to draw conclusions about which model is best, (b) incorrect implementations of the weighted by PITF and PITF analog split-population models, (c) inconsistent test sets for the models examined, and (d) incorrect scoring of the 2016 forecasts.^[We also note additional concerns arising from the question of how the random forest models were tuned by B&S, especially given the way they are used is unorthodox.
 We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.]
@@
 We believe that these research decisions and issues lead B\&S to incorrect conclusions.
-The escalation model \textit{is not} the best; it actually performs worse than the atheoretical, kitchen sink model with over \(1000\) variables.
+The escalation model _is not_ the best; it actually performs worse than the atheoretical, kitchen sink model with over \(1000\) variables.
 We discuss these five issues below.
@@
 The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves.
-B&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation.
+B&S refer to smoothing only in the context of the ROC curves in Figure 1, justified as easing interpretation.
 However, smoothing is also used in all AUC-ROC values they report.
@@
 Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs.
-We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), in order to survey the use of ROC curves.
+We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), to survey the use of ROC curves.
 Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values.
-We found that 90%--17 of 19--clearly used empirical ROC curves, and 2 (10%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code.
+We found that 90%--17 of 19--used empirical ROC curves, and 2 (10%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code.
 The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S.
@@
 We believe that B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
-However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.
+However, they actually weigh the **test** set predictions using the PITF model predictions for the **training** set.
 
 As a result: (1) in the test set, all countries with names that alphabetically start after "Tongo" are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical.
-For example, the "Weighted by PITF" model predictions for Syria from 2008 to 2015 are the Escalation test predictions for that time period weighted, respectively, by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.
+For example, the "Weighted by PITF" model predictions for Syria from 2008 to 2015 are the Escalation test predictions weighted by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.
 
@@
 This description suggests that B&S intended to run two separate random forest models, one each on the low- and high-risk training data splits.
-The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consist of the full training data from all other models.
+The replication code does indeed run two separate random forecasts, but they both utilize the *same training data*, which consists of the full training data from all other models.
 In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.
@@
 B&S present a confusion matrix to score their 2016-H1 forecasts in Table 4.
-Although the forecasts are for the probability of civil war _onset_, they are actually scored using _incidence_ of civil war.
-By definition the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict.
+Although the forecasts are for the probability of civil war _onset_, they are scored using _incidence_ of civil war.
+By definition, the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict.
 When forecasting rare events, it is common to have many false positives, i.e.
@@
 
-This suggest that using smoothed ROC curves will have no substantive impact on the results; however, in this smoothing determines the comparative results.
-Their statement also implies that the ROC curves were only smoothed in the referenced Figure 1.
+This suggests that using smoothed ROC curves will have no substantive impact on the results; however, we find that smoothing does determine the comparative results.
+Their statement also implies that the ROC curves were only smoothed in referenced Figure 1.
 However, this is not the case.
-All AUC-ROC calculations throughout their replication code use an option to smooth the ROC curves *prior* to AUC calculation.
+All AUC-ROC calculations throughout their replication code use an option to smooth the ROC curves before AUC calculation.
 
-ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like.
+ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result, they appear step-like.
 Figure \ref{fig:rocs} shows our replication of both the estimated smoothed ROC curves from the B&S report (left-hand side) and the actual empirical ROC curves (right-hand side).
@@
 The Escalation model, however, _always_ benefits from smoothing, with an average improvement on the order of 0.05.
-This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B&S report, namely that the escalation model is generally superior.
+This is sufficient to push the Escalation model ahead of the alternative models and accounts for the incorrect result B&S report, namely that the escalation model is generally superior.
 
@@
 
-In sum, it is not only the case that using smoothed ROC curves alters the results, but also that the use of smoothed ROC curves to calculate AUC-ROC values \textbf{only} benefits the escalation model.
+In sum, it is not only the case that using smoothed ROC curves alters the results, but also that the use of smoothed ROC curves to calculate AUC-ROC values **only** benefits the escalation model.
 It does so consistently and by a considerable margin.
@@
 
-Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model by quite a margin in the 6-month data version.
+Adding PITF predictors improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model by quite a margin in the 6-month data version.
 
@@
 One finding that remains is that the "PITF Only" model is outperformed by the "Escalation Only" model.
-As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is neither surprising, nor especially noteworthy.
+As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is neither surprising nor especially noteworthy.
 
@@
 The forecasts themselves are probabilistic, i.e.
-range from 0 to 1, but for the purpose of scoring the forecasts they consider the 30 highest predictions to have been positive forecasts ("1") and the remaining forecasts to have been for no onset ("0").
+range from 0 to 1, but to score the forecasts they consider the 30 highest predictions to have been positive forecasts ("1") and the remaining forecasts to have been for no onset ("0").
 They evaluate two alternative codings of observed civil war outcomes, one in which 15 countries, including Colombia, had ongoing civil wars at the end of 2015 (labeled "Assuming Persistence"), and an alternative coding in which the civil war in Colombia ends but Turkey and Benin experience civil war onsets ("Assuming Change").\footnote{\label{fn:leadDV}The former coding is in the main replication data, while the latter is hand-coded in the Stata script that produces their confusion matrices.
 There is also some potential ambiguity about the discussion in this section because the forecasts B\&S claim are for the first half of 2016 are actually for the second half of 2016, but scored against outcomes for the first half of 2016.} However, when they score the forecasts to produce the confusion matrices they report in their Table 4, they score them using civil war incidence, not onset.
-In fact, this a conscious choice B&S seem to acknowledge in the text (if not the table with confusion matrices), but it's not clear to us how this is justifiable.
+This a conscious choice B&S acknowledge in the text (if not the table with confusion matrices), but it's not clear to us how this is justifiable.
 
@@
 The original B&S confusion matrices collapse the second two rows in each matrix and treat ongoing and onset as part of the same group.
-Essentially all of the apparent true positives ($P=1, Y=1$) in the original B&S confusion matrices are actually ongoing civil war cases which by definition _cannot_ experience another civil war onset.
+Essentially all of the apparent true positives ($P=1, Y=1$) in the original B&S confusion matrices are ongoing civil war cases which by definition _cannot_ experience another civil war onset.
 The alternative outcome coding B&S evaluate, under "Assuming Change", produces the only two civil war onsets we would have observed in 2016-H1.
@@
 a country at any given point is either at civil war or not, and onset is the first time period corresponding to this state.
-In the data, ongoing civil wars are coded as having a missing outcome value, and dropped.
+In the data, ongoing civil wars are coded as having a missing outcome value and dropped.
 
-If we place ourselves back to the beginning of 2016, when the forecasts were made (the forecasts are pre-registered in early March), this means that we would have observed that there are 15 countries that had ongoing conflicts in the second half of 2015 which had not terminated prior to the forecasting period.
+If we place ourselves back to the beginning of 2016, when the forecasts were made (the forecasts are pre-registered in early March), this means that we would have observed that 15 countries had ongoing conflicts in the second half of 2015 which had not terminated prior to the forecasting period.
 In all previous analyses in the paper, these kinds of cases drop out of the data.
@@
 conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, _we would have already known that_.
-Rather than not making any "assumptions about the persistence of civil war", this would only make sense if we in fact assume that all ongoing civil wars ended in 2015-H2.
-Since Sambanis is the author of the civil war data that is used, it would  already have been known in early 2016 that these 15 civil wars had in fact not terminated, and were still ongoing.
+Rather than not making any "assumptions about the persistence of civil war", this would only make sense if we assume that all ongoing civil wars ended in 2015-H2.
+Since Sambanis is the author of the civil war data that is used, it would  already have been known in early 2016 that these 15 civil wars had not terminated, and were still ongoing.
 It is thus both by definition impossible for countries with ongoing civil wars to experience another onset, and at the time the forecasts were made it would have been known that all conflicts were still ongoing up through the beginning of the forecasting period.
@@
 The additional statements can be subjected to empirical scrutiny.
-An example might be William Riker's theory of coalitions [@riker:1962] which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically.
+An example might be William Riker's theory of coalitions [@riker:1962] which posits propositions in game-theoretic form from which the size principle of minimal winning coalitions may be deduced and further examined empirically.
 A more recent example of strong theory is found in *The Political Logic of Survival* [@bdm:etal:2005].
-For the purposes of discussion we call this strong theory.
+For the purposes of discussion, we call this strong theory.
 
@@
 In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups.
-In practical terms there is a wide array of choices that are available to instantiate any particular theory.
+In practical terms, there is a wide array of choices that are available to instantiate any particular theory.
 A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated dynamics.
-One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.
+One general principle that is on point in the B\&S article and is more widely relevant is that dynamical phenomena like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.
 
@@
 We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off?
-Certainly that's not what B&S are trying to suggest since this would be a problematic way to "evaluate" a theory.
-But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could in fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated "theory" with any eye towards out-of-sample predictive accuracy [cf.
+Certainly, that's not what B&S are trying to suggest since this would be a problematic way to "evaluate" a theory.
+But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could, in fact, be tied to a theory, it is hard to demonstrate that this theory is not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated "theory" with an eye towards out-of-sample predictive accuracy [cf.
 @colaresi:mahmood:2017 for how to avoid this problem while leveraging out-of-sample predictions].
@@
 Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory.
-For example, both @ward:etal:2010 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes.
-It is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.
+For example, both @ward:etal:2010 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for actually predicting those outcomes.
+It is also possible that building accurate predictive models could uncover factors overlooked in the literature, e.g.
 the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups [@beger:ward:2020].
 
-Aside from selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form.
+Aside from the selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility regarding functional form.
 Theory should not only tell us what factors matter, but also how they relate to each other.
@@
 Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning.
-Even if B\&S were correct that in their study the theory based escalation model performed better, it would simply be a case study of a single case where it was preferable.
+Even if B\&S were correct that in their study the theory-based escalation model performed better, it would simply be a case study of a single case where it was preferable.
 It establishes no general tendency, let alone law-like evidence for general theory.
@@
 The problems we encountered are not subjective modeling choices.
-They bear no implications on the question of whether theory based model are preferable.
+They bear no implications on the question of whether theory-based model are preferable.
 However, when we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturned.

### 2020-11-10 paper/paper.tex
some edits suggested by Grammarly

@@
 We find that there are several important mistakes and that their key finding is entirely conditional on the use of parametrically smoothed ROC curves when calculating accuracy, rather than the standard empirical ROC curves that dominate the literature.
-Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.
+Fixing these mistakes results in a reversal of their claim that theory-based models of escalation are better at predicting onsets of civil war than other kinds of models.
 Their model is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.
-More importantly, we argue that rather than trying to contrast the roles of theory and "atheoretical" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant predictive work.
+More importantly, we argue that rather than trying to contrast the roles of theory and "atheoretical" machine learning in predictive modeling, it would be more productive to focus on ways in which predictive modeling and machine learning could be used to strengthen extant predictive work.
 Instead, we argue that predictive modeling and machine learning are effective tools for theory testing.
@@
 They do not provide much guidance on an exact specification of what variables and which functional form is required.
-As a result it is both not very clear what the ``correct'' model should be, nor easy to demonstrate that any particular ``theoretical'' model's ability to predict generalizes beyond the data that was used to create it in the first place.
+As a result, it is both not very clear what the ``correct'' model should be, nor easy to demonstrate that any particular ``theoretical'' model's ability to predict generalizes beyond the data that was used to create it in the first place.
 Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to ``explain'' a phenomenon to strengthen and whittle down the broad space of extant theories.
@@
 To examine the first question, B\&S compare the test set of the escalation model to four alternative models.
-The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country.
+The independent variables for the first set of analyses reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country.
 The models are:
@@
 \item
-  Average: unweighted average of the predictions from the four models briefly described above.
+  Average: an unweighted average of the predictions from the four models briefly described above.
 \end{itemize}
@@
 \item
-  PITF Split Population: the training data are split into high and low risk portions based on the PITF instability model predictions, two separate escalation random forests are trained on the splits, then re-combined into a single random forest that is used to create the test set predictions.
+  PITF Split Population: the training data are split into high- and low-risk portions based on the PITF instability model predictions, two separate escalation random forests are trained on the splits, then re-combined into a single random forest that is used to create the test set predictions.
 \item
@@
 Finally, B\&S used their escalation model to create forecasts for the first half of 2016.
-In their third analysis, they score the forecasts accuracy using subsequent civil war onset data.
+In their third analysis, they score the forecasts' accuracy using subsequent civil war onset data.
 This is summarized in B\&S Table 3.
@@
 While replicating and analyzing B\&S's results, we found several issues worthy of further discussion and investigation.
-These are (a) the use of smoothed ROC curves to draw conclusions about which model is best, (b) incorrect implementations of the weighted by PITF and PITF analog split-population models, (c) inconsistent test sets for the models examined, and (d) incorrect scoring of the 2016 forecasts.\footnote{We also note there are additional concerns arising from the question of how the random forest models were tuned by B\&S, especially given the way they are used is unorthodox.
+These are (a) the use of smoothed ROC curves to draw conclusions about which model is best, (b) incorrect implementations of the weighted by PITF and PITF analog split-population models, (c) inconsistent test sets for the models examined, and (d) incorrect scoring of the 2016 forecasts.\footnote{We also note additional concerns arising from the question of how the random forest models were tuned by B\&S, especially given the way they are used is unorthodox.
 We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.}
@@
 We believe that these research decisions and issues lead B\&S to incorrect conclusions.
-The escalation model \textit{is not} the best; it actually performs worse than the atheoretical, kitchen sink model with over \(1000\) variables.
+The escalation model \emph{is not} the best; it actually performs worse than the atheoretical, kitchen sink model with over \(1000\) variables.
 We discuss these five issues below.
@@
 The most consequential issue that we found is that all AUC-ROC values reported in B\&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves.
-B\&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation.
+B\&S refer to smoothing only in the context of the ROC curves in Figure 1, justified as easing interpretation.
 However, smoothing is also used in all AUC-ROC values they report.
@@
 Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs.
-We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), in order to survey the use of ROC curves.
+We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), to survey the use of ROC curves.
 Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values.
-We found that 90\%--17 of 19--clearly used empirical ROC curves, and 2 (10\%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code.
+We found that 90\%--17 of 19--used empirical ROC curves, and 2 (10\%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code.
 The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S.
@@
 We believe that B\&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
-However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.
+However, they actually weigh the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.
 
 As a result: (1) in the test set, all countries with names that alphabetically start after ``Tongo'' are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical.
-For example, the ``Weighted by PITF'' model predictions for Syria from 2008 to 2015 are the Escalation test predictions for that time period weighted, respectively, by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.
+For example, the ``Weighted by PITF'' model predictions for Syria from 2008 to 2015 are the Escalation test predictions weighted by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.
 
@@
 This description suggests that B\&S intended to run two separate random forest models, one each on the low- and high-risk training data splits.
-The replication code does indeed run two separate random forecasts, but they both utilize the \emph{exact same training data}, which consist of the full training data from all other models.
+The replication code does indeed run two separate random forecasts, but they both utilize the \emph{same training data}, which consists of the full training data from all other models.
 In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.
@@
 B\&S present a confusion matrix to score their 2016-H1 forecasts in Table 4.
-Although the forecasts are for the probability of civil war \emph{onset}, they are actually scored using \emph{incidence} of civil war.
-By definition the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict.
+Although the forecasts are for the probability of civil war \emph{onset}, they are scored using \emph{incidence} of civil war.
+By definition, the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict.
 When forecasting rare events, it is common to have many false positives, i.e.~instances where we predict onset but no onsets occur.
@@
 
-This suggest that using smoothed ROC curves will have no substantive impact on the results; however, in this smoothing determines the comparative results.
-Their statement also implies that the ROC curves were only smoothed in the referenced Figure 1.
+This suggests that using smoothed ROC curves will have no substantive impact on the results; however, we find that smoothing does determine the comparative results.
+Their statement also implies that the ROC curves were only smoothed in referenced Figure 1.
 However, this is not the case.
-All AUC-ROC calculations throughout their replication code use an option to smooth the ROC curves \emph{prior} to AUC calculation.
-
-ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like.
+All AUC-ROC calculations throughout their replication code use an option to smooth the ROC curves before AUC calculation.
+
+ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result, they appear step-like.
 Figure \ref{fig:rocs} shows our replication of both the estimated smoothed ROC curves from the B\&S report (left-hand side) and the actual empirical ROC curves (right-hand side).
@@
 The Escalation model, however, \emph{always} benefits from smoothing, with an average improvement on the order of 0.05.
-This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B\&S report, namely that the escalation model is generally superior.
+This is sufficient to push the Escalation model ahead of the alternative models and accounts for the incorrect result B\&S report, namely that the escalation model is generally superior.
 
@@
 
-Adding PITF predictors actually improves performance in both cases; the ``With PITF Predictions'' model strictly dominates the ``Escalation Only'' model by quite a margin in the 6-month data version.
+Adding PITF predictors improves performance in both cases; the ``With PITF Predictions'' model strictly dominates the ``Escalation Only'' model by quite a margin in the 6-month data version.
 
@@
 One finding that remains is that the ``PITF Only'' model is outperformed by the ``Escalation Only'' model.
-As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is neither surprising, nor especially noteworthy.
+As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is neither surprising nor especially noteworthy.
 
@@
 B\&S forecast the risk of civil war onset for the first half of 2016 (2016-H1, where H2 would be the second half), and assessed the accuracy of these forecasts.
-The forecasts themselves are probabilistic, i.e.~range from 0 to 1, but for the purpose of scoring the forecasts they consider the 30 highest predictions to have been positive forecasts (``1'') and the remaining forecasts to have been for no onset (``0'').
+The forecasts themselves are probabilistic, i.e.~range from 0 to 1, but to score the forecasts they consider the 30 highest predictions to have been positive forecasts (``1'') and the remaining forecasts to have been for no onset (``0'').
 They evaluate two alternative codings of observed civil war outcomes, one in which 15 countries, including Colombia, had ongoing civil wars at the end of 2015 (labeled ``Assuming Persistence''), and an alternative coding in which the civil war in Colombia ends but Turkey and Benin experience civil war onsets (``Assuming Change'').\footnote{\label{fn:leadDV}The former coding is in the main replication data, while the latter is hand-coded in the Stata script that produces their confusion matrices.
 There is also some potential ambiguity about the discussion in this section because the forecasts B\&S claim are for the first half of 2016 are actually for the second half of 2016, but scored against outcomes for the first half of 2016.} However, when they score the forecasts to produce the confusion matrices they report in their Table 4, they score them using civil war incidence, not onset.
-In fact, this a conscious choice B\&S seem to acknowledge in the text (if not the table with confusion matrices), but it's not clear to us how this is justifiable.
+This a conscious choice B\&S acknowledge in the text (if not the table with confusion matrices), but it's not clear to us how this is justifiable.
 
@@
 The original B\&S confusion matrices collapse the second two rows in each matrix and treat ongoing and onset as part of the same group.
-Essentially all of the apparent true positives (\(P=1, Y=1\)) in the original B\&S confusion matrices are actually ongoing civil war cases which by definition \emph{cannot} experience another civil war onset.
+Essentially all of the apparent true positives (\(P=1, Y=1\)) in the original B\&S confusion matrices are ongoing civil war cases which by definition \emph{cannot} experience another civil war onset.
 The alternative outcome coding B\&S evaluate, under ``Assuming Change'', produces the only two civil war onsets we would have observed in 2016-H1.
@@
 Civil war appears to be a binary property of a country at any given moment in time, i.e.~a country at any given point is either at civil war or not, and onset is the first time period corresponding to this state.
-In the data, ongoing civil wars are coded as having a missing outcome value, and dropped.
-
-If we place ourselves back to the beginning of 2016, when the forecasts were made (the forecasts are pre-registered in early March), this means that we would have observed that there are 15 countries that had ongoing conflicts in the second half of 2015 which had not terminated prior to the forecasting period.
+In the data, ongoing civil wars are coded as having a missing outcome value and dropped.
+
+If we place ourselves back to the beginning of 2016, when the forecasts were made (the forecasts are pre-registered in early March), this means that we would have observed that 15 countries had ongoing conflicts in the second half of 2015 which had not terminated prior to the forecasting period.
 In all previous analyses in the paper, these kinds of cases drop out of the data.
@@
 conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, \emph{we would have already known that}.
-Rather than not making any ``assumptions about the persistence of civil war'', this would only make sense if we in fact assume that all ongoing civil wars ended in 2015-H2.
-Since Sambanis is the author of the civil war data that is used, it would already have been known in early 2016 that these 15 civil wars had in fact not terminated, and were still ongoing.
+Rather than not making any ``assumptions about the persistence of civil war'', this would only make sense if we assume that all ongoing civil wars ended in 2015-H2.
+Since Sambanis is the author of the civil war data that is used, it would already have been known in early 2016 that these 15 civil wars had not terminated, and were still ongoing.
 It is thus both by definition impossible for countries with ongoing civil wars to experience another onset, and at the time the forecasts were made it would have been known that all conflicts were still ongoing up through the beginning of the forecasting period.
@@
 The additional statements can be subjected to empirical scrutiny.
-An example might be William Riker's theory of coalitions (Riker 1962) which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically.
+An example might be William Riker's theory of coalitions (Riker 1962) which posits propositions in game-theoretic form from which the size principle of minimal winning coalitions may be deduced and further examined empirically.
 A more recent example of strong theory is found in \emph{The Political Logic of Survival} (Bueno de Mesquita et al.
 2005).
-For the purposes of discussion we call this strong theory.
+For the purposes of discussion, we call this strong theory.
 
@@
 We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off?
-Certainly that's not what B\&S are trying to suggest since this would be a problematic way to ``evaluate'' a theory.
-But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could in fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated ``theory'' with any eye towards out-of-sample predictive accuracy (cf.~Colaresi and Mahmood 2017 for how to avoid this problem while leveraging out-of-sample predictions).
+Certainly, that's not what B\&S are trying to suggest since this would be a problematic way to ``evaluate'' a theory.
+But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could, in fact, be tied to a theory, it is hard to demonstrate that this theory is not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated ``theory'' with an eye towards out-of-sample predictive accuracy (cf.~Colaresi and Mahmood 2017 for how to avoid this problem while leveraging out-of-sample predictions).
 Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize and whose underlying theories may by extension not be externally valid (for a more detailed treatment of this issue, see Fariss and Jones 2018).
@@
 Conversely, not all factors that are ``theoretically'' important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory.
-For example, both Ward, Greenhill, and Bakke (2010) and Hill and Jones (2014) examine a variety of variables that are considered ``theoretically'' important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes.
-It is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.~the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al.
+For example, both Ward, Greenhill, and Bakke (2010) and Hill and Jones (2014) examine a variety of variables that are considered ``theoretically'' important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for actually predicting those outcomes.
+It is also possible that building accurate predictive models could uncover factors overlooked in the literature, e.g.~the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al.
 2010), irregular leadership changes (Beger et al.
@@
 
-Aside from selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form.
+Aside from the selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility regarding functional form.
 Theory should not only tell us what factors matter, but also how they relate to each other.
@@
 The problems we encountered are not subjective modeling choices.
-They bear no implications on the question of whether theory based model are preferable.
+They bear no implications on the question of whether theory-based model are preferable.
 However, when we fix these issues and perform an updated analysis, the B\&S conclusions are all essentially overturned.

### 2020-11-10 paper/paper.Rmd
a few more Grammarly suggested edits

@@
 ```
+
 # Introduction
@@
 
-Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider the ways in which they can mutually reinforce each other.
+Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider how they can mutually reinforce each other.
 Even if the analysis in B&S fails to support the point, we concede that predicting on the basis of a strong theoretical model is preferable to inductive prediction.
@@
 The problem is that many extant theoretical models are in fact so malleable to interpretation that they essentially are ad hoc.
-They are useful for identifying variables that could be useful for prediction, and this is in fact already taking place.
+They are useful for identifying variables that could be useful for prediction, and this is already taking place.
 
-Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning.
+Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based on machine learning.
 Even if B\&S were correct that in their study the theory-based escalation model performed better, it would simply be a case study of a single case where it was preferable.
@@
 Here, the scoring is done using the non-lead version of the DV.
-Thus there is a mismatch between the time period the forecasts are for and the period who's observed outcomes are used to score them.
+Thus there is a mismatch between the time period the forecasts are for and the period whose observed outcomes are used to score them.
 
@@
 
-Furthermoe, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
+Furthermore, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
 In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable to incidence.
@@
 What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated.
-Table \ref{tab:hp} shows the default values used by the implementation of random forest that B&S use (from the **randomForest** R package), in contrast to the basic settings used by B&S for most the models reported in the paper.
+Table \ref{tab:hp} shows the default values used by the implementation of random forest that B&S use (from the **randomForest** R package), in contrast to the basic settings used by B&S for most of the models reported in the paper.
 
@@
 The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number.
-From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive, and operates on a large bootstrapped sample of the original training data.
+From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive and operates on a large bootstrapped sample of the original training data.
 These are of course only heuristics and it is usual to attempt to find better hyper-parameter methods through some form of tuning.

### 2020-11-10 paper/paper.tex
a few more Grammarly suggested edits

@@
 
-Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider the ways in which they can mutually reinforce each other.
+Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider how they can mutually reinforce each other.
 Even if the analysis in B\&S fails to support the point, we concede that predicting on the basis of a strong theoretical model is preferable to inductive prediction.
@@
 The problem is that many extant theoretical models are in fact so malleable to interpretation that they essentially are ad hoc.
-They are useful for identifying variables that could be useful for prediction, and this is in fact already taking place.
-
-Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning.
+They are useful for identifying variables that could be useful for prediction, and this is already taking place.
+
+Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based on machine learning.
 Even if B\&S were correct that in their study the theory-based escalation model performed better, it would simply be a case study of a single case where it was preferable.
@@
 Here, the scoring is done using the non-lead version of the DV.
-Thus there is a mismatch between the time period the forecasts are for and the period who's observed outcomes are used to score them.
+Thus there is a mismatch between the time period the forecasts are for and the period whose observed outcomes are used to score them.
 \end{enumerate}
@@
 
-Furthermoe, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
+Furthermore, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
 In the file \texttt{6mo\_make\_confusion\_matrix.do} on line 52, missing values in ``incidence\_civil\_ns'' are recoded to 1, thus reverting the onset coding of this variable to incidence.
@@
 What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated.
-Table \ref{tab:hp} shows the default values used by the implementation of random forest that B\&S use (from the \textbf{randomForest} R package), in contrast to the basic settings used by B\&S for most the models reported in the paper.
+Table \ref{tab:hp} shows the default values used by the implementation of random forest that B\&S use (from the \textbf{randomForest} R package), in contrast to the basic settings used by B\&S for most of the models reported in the paper.
 

### 2020-11-10 rep_nosmooth/README.Rmd
add a master.R script for replication

@@
 ---
-title: "Modified replication"
+title: "Replication materials for Beger, Morgan, Ward, nd"
 output: 
@@
 ```
+
+**tl;dr:** run through `master.R`
 
@@
 
-One of the principal outputs of `setup-model-definitions.R` is this encoding, in `output/model-definitions/model-table.rds`:
+One of the principal outputs of `1-setup-model-definitions.R` is this encoding, in `output/model-definitions/model-table.rds`:
 

### 2020-11-10 rep_nosmooth/master.R

@@ -1,0 +1,30 @@
+#
+#   Replication master script
+#
+
+# Make sure required packages are installed
+source("0-packages.R")
+
+# Encode original B&S Tables 1 and 2
+source("1-make-original-tables.R")
+
+# Re-save data from Stata to R formats for faster loading
+source("1-re-save-data.R")
+
+# Create model table that we will use for running models in parallel
+source("1-setup-model-table.R")
+
+# Run models (this should take about 1 hour)
+source("2-model-runner.R")
+
+# Recreate Tables 1 and 2 from new output
+source("3-recreate-tables.R")
+
+# Figures
+source("4-make-figure1.R")
+source("4-make-figure2.R")
+source("4-make-figureA1.R")
+source("4-make-figureA2.R")
+source("4-make-figureA3.R")
+
+

### 2020-11-11 paper/paper.Rmd
copy edits
I went through the doc twice. I caught a few things, but someone else might want to read through it again to make sure...

@@
  - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
- - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
+ - Richard K. Morgan^[Independent Researcher, rick.morgan2@gmail.com.]
  - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
@@
 We find that these claims to be unsupported and evidence presented for them to be incorrect.
-Their theory-based escalation model does not do better than the alternatives that they examine.
+Their theory-based escalation model does not do better than the alternatives they examine.
 It does worse.
-The reason for this is that they have made several mistakes in their research procedure.
+The reason for this is that they made several mistakes in their research procedure.
 Further, the performance results they report are based on smoothed performance curves, not the empirical, unsmoothed curves.
@@
 
-B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately".
+B\&S claim (page 3) to argue that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately".
 Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.
@@
 Third, they preregistered the list of thirty countries that have the highest risk of civil war onset.
-They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.^[PITF [e.g., @goldstone:etal:2010; @beger:ward:2017], the W-ICEWS project [@ward:etal:2013], VIEWS [@hegre:etal:2019], and others, e.g., @beger:etal:2020.] B\&S claim to be unique in assessing these forecasts.
+They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.^[PITF [e.g., @goldstone:etal:2010; @beger:ward:2017], the W-ICEWS project [@ward:etal:2013], ViEWS [@hegre:etal:2019], and others, e.g., @beger:etal:2020.] B\&S claim to be unique in assessing these forecasts.
 A qualitative analysis of their predictions allows them to conclude that their model is robust.
@@
 
-As a result: (1) in the test set, all countries with names that alphabetically start after "Tongo" are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical.
+As a result: (1) in the test set, all countries with names that alphabetically start after "Tonga" are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical.
 For example, the "Weighted by PITF" model predictions for Syria from 2008 to 2015 are the Escalation test predictions weighted by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.
@@
 Although the two random forest models are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger random forest using both component model's underlying decision trees.
-Thus, while all random forest models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new random forest model after combination does have 200,000 decision trees.
+Thus, while all random forest models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new random forest model after combination has 200,000 decision trees.
 Furthermore, the PITF model predictions do not impact how the combined random forest model predicts at all, not even through a binary low-/high-risk split.
-The split-population PITF random forest model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing random forest randomization facilities (row and column sampling for each decision tree).
+The split-population PITF random forest model is, practically speaking, simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra (odd) randomization step added to the already existing random forest randomization facilities (row and column sampling for each decision tree).
 This does not adequately implement their split-sample modeling strategy.
@@
 
-Thus a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
+Thus, a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
 In other words, the results for different models in B&S Table 2 are not comparable to one another, and conclusions drawn from such comparison are potentially incorrect.
@@
 The AUC-ROC values are based on original, non-smoothed ROC curves.
-Both the Average and CAMEO models outperform the Escalation model in almost all instances (see also Figure \ref{fig:table1-plot}).
+Both the Average and the CAMEO models outperform the Escalation model in almost all instances (see also Figure \ref{fig:table1-plot}).
 The Goldstein model generally outperforms the Escalation model in the 6-month version.
@@
     "latex", booktabs = TRUE, digits = 0,
-    caption = "Number of valid test predictions for the escalation and structural alterantives models (B\\&S Table 2)",
+    caption = "Number of valid test predictions for the escalation and structural alternatives models (B\\&S Table 2)",
     label = "table2-N") %>%
@@
 
-B&S interpret their results as follows (page 20) and we comment on the conclusions seratim:^[We list the "Overall, ..." interpretation out-of-order for clarity.]
+B&S interpret their results as follows (page 20) and we comment on the conclusions seriatim:^[We list the "Overall, ..." interpretation out-of-order for clarity.]
 
@@
 The original B&S confusion matrices collapse the second two rows in each matrix and treat ongoing and onset as part of the same group.
-Essentially all of the apparent true positives ($P=1, Y=1$) in the original B&S confusion matrices are ongoing civil war cases which by definition _cannot_ experience another civil war onset.
+Essentially, all of the apparent true positives ($P=1, Y=1$) in the original B&S confusion matrices are ongoing civil war cases which by definition _cannot_ experience another civil war onset.
 The alternative outcome coding B&S evaluate, under "Assuming Change", produces the only two civil war onsets we would have observed in 2016-H1.
@@
 \cline{3-4}
+\\
 \end{tabular}
@@
 
-> Importantly, we make no assumptions about the persistence of civil war between the second half of 2015 and the first half of 2016, meaning that all countries—even those with ongoing conflicts as of December 31, 2015--are included in the 2016 test set.
+> Importantly, we make no assumptions about the persistence of civil war between the second half of 2015 and the first half of 2016, meaning that all countries---even those with ongoing conflicts as of December 31, 2015---are included in the 2016 test set.
 As a result, 14 of the 30 countries on the list (and nine of the top 10) are cases of ongoing conflict.
@@
 In the text, B&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases.
-This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g.
-two separatists conflicts in different portions of the country.
+This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g., two separatists' conflicts in different portions of the country.
 But this is not the case; ongoing conflicts in all other analyses  are consistently dropped from the data as if a new civil war onset for a country with an ongoing conflict is conceptually impossible.
@@
 However, for many, a theory is simply a falsifiable story or narrative that makes causal claims about the relationships among variables.
-Typically it is fused together from a meta-analysis of prior work on the topic under consideration.
+Typically, it is fused together from a meta-analysis of prior work on the topic under consideration.
 We call these types of theory weak theory.
@@
 A strong theory gives more guidance on compelling choices than does a weak theory.
-Accordingly, while the choice is plausible it is not necessarily compelling.
+Accordingly, while the choice is plausible, it is not necessarily compelling.
 Another choice might be equally plausible.
-In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups.
+In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look at the dynamics among opposition groups.
 In practical terms, there is a wide array of choices that are available to instantiate any particular theory.
 A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated dynamics.
-One general principle that is on point in the B\&S article and is more widely relevant is that dynamical phenomena like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.
+One general principle in the B\&S article that is on point, and is more widely relevant, is that dynamical phenomena like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.
 
@@
 
-Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory.
+Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that are good predictors do not originate from prior theory.
 For example, both @ward:etal:2010 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for actually predicting those outcomes.
-It is also possible that building accurate predictive models could uncover factors overlooked in the literature, e.g.
-the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups [@beger:ward:2020].
+It is also possible that building accurate predictive models could uncover factors overlooked in the literature, e.g., the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups [@beger:ward:2020].
 
@@
 Are those relationships linear in the log-odds?
-Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships.
+Some machine learning models, random forests included, are flexible in this regard and, to some extent, are capable of learning non-linear and interactive relationships.
 B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives.
@@
 The problems we encountered are not subjective modeling choices.
-They bear no implications on the question of whether theory-based model are preferable.
+They bear no implications on the question of whether theory-based models are preferable.
 However, when we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturned.
@@
 See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`).
-The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions.
-Similarly in the 6-month version of this file.
+The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions -- similarly in the 6-month version of this file.
 
@@
 The intention is to train separate random forests on data split into high and low-risk groups based on PITF predictions, and then combine them back into one random forest.
-Actually both random forests are trained on identical data, essentially production a random forest that just has N_trees \* 2.
+Actually, both random forests are trained on identical data, essentially production a random forest that just has N_trees \* 2.
 
@@
 One can confirm, through some tracing to work back to the original variable assignments, that the ROC curves and AUC are correspondingly, and correctly, also scored using the lead versions of the DV.
-2. The nominally 2016-H1 forecasts are created at the bottom of `+master.R` using the data for 2016-H1 and still the lead DV. Thus the lead DV here refers to outcomes in 2016-H2. 
+2. The nominally 2016-H1 forecasts are created at the bottom of `+master.R` using the data for 2016-H1 and still the lead DV. Thus, the lead DV here refers to outcomes in 2016-H2. 
 3.
@@
 Here, the scoring is done using the non-lead version of the DV.
-Thus there is a mismatch between the time period the forecasts are for and the period whose observed outcomes are used to score them.
+Thus, there is a mismatch between the time period the forecasts are for and the period whose observed outcomes are used to score them.
 
-The relevant variables in the data are "incidence_civil_ns" and "incidence_civil_ns_plus1", which appears to be a 1-period lead (ie., $t_+1$) version of the dependent variable that is used in the actual prediction models.
+The relevant variables in the data are "incidence_civil_ns" and "incidence_civil_ns_plus1", which appears to be a 1-period lead (i.e., $t_{+1}$) version of the dependent variable that is used in the actual prediction models.
 The incidence dependent variable  contains both 0/1 and missing values.

### 2020-11-12 paper.tex

--- 2020-06-03 paper.tex
@@
-\documentclass[]{article}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+]{article}
 \usepackage{lmodern}
@@
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
 \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
   \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
 }{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  pdftitle={Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict},
+  pdfauthor={Andreas Beger; Richard K. Morgan; Michael D. Ward},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdftitle={Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020, ``Forecasting Civil Wars: Theory and Structure in an Age of `Big Data' and Machine Learning'', JCR},
-            pdfauthor={Richard K. Morgan and Andreas Beger and Michael D. Ward},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
 \usepackage{longtable,booktabs}
-\usepackage{graphicx,grffile}
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\providecommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020,
-``Forecasting Civil Wars: Theory and Structure in an Age of `Big Data'
-and Machine Learning'', \emph{JCR}}
-    \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
-    \author{Richard K. Morgan and Andreas Beger and Michael D. Ward}
-    \preauthor{\centering\large\emph}
-  \postauthor{\par}
-      \predate{\centering\large\emph}
-  \postdate{\par}
-    \date{03 June 2020}
-
+\setcounter{secnumdepth}{5}
+\usepackage{floatrow}
+\floatsetup[figure]{capposition=bottom}
+\usepackage[format=plain,labelfont=bf,up]{caption}
 \usepackage{booktabs}
@@
 \usepackage{xcolor}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newenvironment{cslreferences}%
+  {\setlength{\parindent}{0pt}%
+  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
+  {\par}
+
+\title{Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict\thanks{John Ahlquist, Cassy L.
+Dorff, and Shahryar Minhas both provided helpful feedback on this project.
+We are especially grateful to Paul Huth for his comments and guidance.
+All the code and several additional analyses can be found at our replication archive at \url{https://github.com/andybega/Blair-Sambanis-replication}.}}
+\author{Andreas Beger\footnote{Predictive Heuristics, \href{mailto:adbeger@gmail.com}{\nolinkurl{adbeger@gmail.com}}.} \and Richard K.
+Morgan\footnote{Varieties of Democracy Institute, University of Gothenburg, \href{mailto:rick.morgan2@gmail.com}{\nolinkurl{rick.morgan2@gmail.com}}.} \and Michael D.
+Ward\footnote{Predictive Heuristics, Duke University, University of Washington, \href{mailto:michael.don.ward@gmail.com}{\nolinkurl{michael.don.ward@gmail.com}}.
+Corresponding author.}}
+\date{24 September 2020}
 
@@
 \maketitle
-
-\hypertarget{review-of-blair-and-sambanis-2020}{%
-\section{Review of Blair and Sambanis
-2020}\label{review-of-blair-and-sambanis-2020}}
-
-Blair and Sambanis (2020) (BS for the sake of brevity hereafter) aim to
-examine whether theory adds value to a forecasting model when compared
-to non-parametric machine learning models, and specifically random
-forests, whose specifications are not theory-informed. They find that it
-does. They arrive at this conclusion by examining the problem of
-predicting civil war onset, and find that a parsimonious model using a
-small number of covariates derived from escalation theories of conflict
-can forecast civil war onset better than alternative specifications
-based on generic covariates not specifically informed by theory and a
-kitchen sink model with more than 1,000 covariates.
-
-BS specifically examine three questions:
-
-\begin{itemize}
-\item
-  \begin{enumerate}
-  \def\labelenumi{(\arabic{enumi})}
-  \tightlist
-  \item
-    how does the theoretically-driven Escalation model compare in
-    forecast performance to alternative models not informed specifically
-    by civil war onset theories,
-  \end{enumerate}
-\item
-  \begin{enumerate}
-  \def\labelenumi{(\arabic{enumi})}
-  \setcounter{enumi}{1}
-  \tightlist
-  \item
-    does annual, structural information from the PITF instability
-    forecasting model add to the Escalation model's monthly and 6-month
-    predictions,
-  \end{enumerate}
-\item
-  and (3) how accurate were predictions for the first half of 2016
-  derived from the Escalation model?
-\end{itemize}
-
-To assess the first two questions BS use global datasets covering all
-major countries from 2001 to 2015. Two versions of the dataset are used,
-one at the country-month level, the other aggregated to 6-month
-half-years. The main outcome is civil war onset, measured using
-Sambanis' civil war dataset.
-
-Both the first and second questions above rely on comparing the
-Escalation model to various alternative models. The same procedure is
-used in both cases:
+\begin{abstract}
+We examine the research protocols in Blair and Sambanis (nd), who argue that their theory-based model can predict civil war onsets better than several atheoretical alternatives or a model with country-structural factors.
+We find that there are several important mistakes and that their key finding is entirely conditional on the use of parametrically smoothed ROC curves when calculating accuracy, rather than the standard empirical ROC curves that dominate the literature.
+Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.
+Their model is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.
+More importantly, we argue that rather than trying to contrast the roles of theory and "atheoretical" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant theories.
+\end{abstract}
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+Blair and Sambanis (n.d., hereafter B\&S) argue that theory is essential for creating models that have high accuracy in forecasting civil war onset.
+Indeed, they assert that with such theory forecasting is more accurate than has previously been possible.
+We re-examine the empirical basis for the claims made in support of it.
+We find that these claims to be unsupported and evidence presented for them to be incorrect.
+Their theory-based escalation model does not do better than the alternatives that they examine.
+It does worse.
+The reason for this is that they have made several mistakes in their research procedure.
+Further, the performance results they report are based on smoothed performance curves, not the empirical, unsmoothed curves.
+This provides misleading results.
+In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented.
+We also found that the scoring of their forecasts for the first half of 2016 was incorrectly performed using civil war incidence, not onset.
+In what follows, we show the impact of these mistakes on the conclusions.
+
+B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars ``remarkably accurately''.
+Indeed, B\&S argue that this theoretical model outperforms four other ``more mechanical'' alternatives.
+They also claim that the integration of structure with process is better than alternatives over short forecasting windows.
+Third, they preregistered the list of thirty countries that have the highest risk of civil war onset.
+They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.\footnote{PITF (e.g., Goldstone et al.
+2010; Beger and Ward 2017), the W-ICEWS project (Ward et al.
+2013), VIEWS (Hegre et al.
+2019), and others, e.g., Beger, Morgan, and Maxwell (2020).} B\&S claim to be unique in assessing these forecasts.
+A qualitative analysis of their predictions allows them to conclude that their model is robust.
+We will return to their analysis later, after correcting the procedural mistakes we found in their research process.
+
+Before proceeding, we quote B\&S (page 24):
+
+\begin{quote}
+Our theoretically driven model generates accurate forecasts, with base specification AUCs of 0.82 and 0.85 over one- and six-month windows, respectively, and AUCs as high as 0.92 in other specifications.
+Our model also consistently and sometimes dramatically outperforms the alternatives we test.
+{[}\ldots{]} Cederman and Weidmann (2017, 476) argue that ``the hope that big data will somehow yield valid forecasts through theory-free `brute force' is misplaced in the area of political violence.'' Our results lend some credence to this claim.
+\end{quote}
+
+Although their analysis in the end does not support this point, we do not disagree that theory is important in predictive modeling.
+Of the various ways one could contrast ad hoc machine learning with ``theoretical'' modeling, B\&S focus essentially on theory as a mechanism for variable selection.
+In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting.
+The problem is that the space of extant theories for a phenomenon like civil war is so broad and non-specific in terms of an exact specification of what variables and in which functional form should constitute a prediction model, that it's both not very clear what the ``correct'' model should be, nor easy to demonstrate that any particular ``theoretical'' model's ability to predict generalizes beyond the data that was used to create it in the first place.
+Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to ``explain'' a phenomenon to strengthen and whittle down the broad space of extant theories.
+
+\hypertarget{summary-of-blair-and-sambanis-2020}{%
+\section{Summary of Blair and Sambanis (2020)}\label{summary-of-blair-and-sambanis-2020}}
+
+B\&S' analysis is based on the use of non-linear, non-deterministic machine learning models, and specifically random forests, one of which has a specification they argue is theory-based.
+Several others with more generic sets of covariates are also developed and compared.
+Notably, the analysis is at the country-month or country-(6 month) levels and relies in large part on indicators derived from the ICEWS event data.
+In short, they uphold their assumption that theory-guided empirical research produces better conflict predictions than machine learning inspired efforts that are necessarily ad hoc combinations of available variables.
+They arrive at this conclusion by examining the problem of predicting civil war onset.
+They report that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory, including a \textit{kitchen sink} model with more than 1,000 covariates.
+
+B\&S specifically examine three questions:
 
@@
 \item
-  Split the training data into training (2001 - 2007) and test (2008 -
-  2015) sets.
-\item
-  Estimate the Escalation and other competing models.
-\item
-  Create out-of-sample (OOS) predictions from each model using the test
-  set.
+  How does the theoretically-driven escalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories?
+\item
+  Does annual, structural information from the PITF instability forecasting model add to the escalation model's monthly and 6-month predictions?
+\item
+  How accurate were predictions using the escalation model for the first half of 2016?
+\end{enumerate}
+
+To assess the first two questions, B\&S use ICEWS data covering all major countries from 2001 to 2015.
+Two versions of the dataset are used, one at the country-month level, the other aggregated to 6-month half-years.
+The main outcome variable is civil war onset, measured using Sambanis' civil war dataset.
+
+Both the first and second questions above rely on comparing their escalation model to various alternative models. The same procedure is used in both cases:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Split the training data into training (2001 - 2007) and test (2008 - 2015) data.
+\item
+  Estimate the escalation and other competing models with the training data.
+\item
+  Create out-of-sample (OOS) predictions from each model using the test set.
 \item
@@
 
-The corresponding results for each question are shown in BS Tables 1 and
-2, which we will replicate further below.
-
-To examine the first question, BS compare the test set of the Escalation
-model to four alternative models. The independent variables for the
-first set of analysis reported in Table 1 in the paper are all derived
-from the ICEWS event data, using domestic events between actors within a
-country. The models are:
+To examine the first question, B\&S compare the test set of the escalation model to four alternative models.
+The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country.
+The models are:
 
@@
 \item
-  Escalation: a set of 10 theoretically-informed indicators.
-\item
-  Quad: ICEWS quad counts, i.e.~material conflict, material cooperation,
-  verbal conflict, verbal cooperation.
-\item
-  Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from
-  the ICEWS data for interactions between the government one one side
-  and opposition or rebel actors on the other. These are directed, thus
-  making for 4 total covariates.
-\item
-  CAMEO: counts for all CAMEO event codes, thus a total of 1,159
-  covariates.
-\item
-  Average: unweighted average of the predictions from the 4 models
-  above.
+  Escalation: a set of ten indicators, drawn from a theoretical escalation model, for interactions between the government on one side and opposition or rebel actors on the other.
+\item
+  Quad: ICEWS quad counts, i.e.~material conflict, material cooperation, verbal conflict, verbal cooperation, for interactions between the government and opposition or rebels.
+These are directed, thus making for four directed dyads, which with four quad categories make sixteen covariates.
+\item
+  Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for the same four directed dyads, for a total of four covariates.
+\item
+  CAMEO: counts for all CAMEO event codes over the four actor dyads, totaling \(1,160\) covariates, which are mostly zero for any country in any month.
+\item
+  Average: unweighted average of the predictions from the four models briefly described above.
 \end{itemize}
 
-The results in Table 1, aside from the core base specification results,
-include 8 additional robustness tests for both the 1-month and 6-month
-versions. These robustness checks vary either (1) random forecast
-hyperparameter values, or (2) the year used to split the train/test
-data, or (3) alternative codings of the civil war onset dependent
-variable.
-
-The second question, whether structural variables add to the Escalation
-model, is assessed by comparing the original escalation model to four
-alternatives that incorporate annual, structural variables that are used
-in the PITF instability forecasting model:
+The corresponding results for each question are shown in B\&S Tables 1 and 2, which we examine further below.
+We accurately replicate their Tables 1 and 2.
+The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions.
+These robustness checks vary either (1) random forecast hyperparameter values or (2) the year used to split the train/test data, or (3) alternative codings of the civil war onset dependent variable.
+
+The second question, whether structural variables add to the escalation model, is assessed by comparing the original escalation model to four alternatives that incorporate annual, structural variables that are used in the PITF instability forecasting model:
 
@@
 \item
-  Escalation Only: the original basic escalation model with only ICEWS
-  predictors
-\item
-  With PITF Predictors: a random forest that as predictors has the
-  escalation model indicators but also the PITF annual, structural
-  variables
-\item
-  Weighted by PITF: escalation model predictions weighted using the PITF
-  instability model predictions
-\item
-  PITF Split Population: the training data are split into high and low
-  risk portions based on the PITF instability model predictions, two
-  separate escalation random forests are trained on the splits, then
-  re-combined into a single random forest that is used to create the
-  test set predictions
-\item
-  PITF Only: a random forest model based only on the annual, structural
-  PITF model predictors
+  Escalation Only: the original basic escalation model with only ICEWS predictors.
+\item
+  With PITF Predictors: a random forest that also adds the PITF annual, structural variables.
+\item
+  Weighted by PITF: escalation model predictions weighted using the PITF instability model predictions.
+\item
+  PITF Split Population: the training data are split into high and low risk portions based on the PITF instability model predictions, two separate escalation random forests are trained on the splits, then re-combined into a single random forest that is used to create the test set predictions.
+\item
+  PITF Only: a random forest model based only on the annual, structural PITF model predictors.
 \end{itemize}
 
-The corresponding results are shown in BS Table 2.
-
-Finally, BS used the Escalation model to create forcasts for the first
-half of 2016, and in their third and final analysis, they score the
-forecasts accuracy using civil war onset data later observed. This is
-summarized in BS Table 3.
-
-\hypertarget{replication-problems}{%
-\section{Replication problems}\label{replication-problems}}
-
-While attempting to replicate BS's results, we found several issues.
+The corresponding results are shown in B\&S Table 2.
+
+Finally, B\&S used their escalation model to create forecasts for the first half of 2016.
+In their third analysis, they score the forecasts accuracy using subsequent civil war onset data.
+This is summarized in B\&S Table 3.
+
+\hypertarget{implementation-issues-in-bs}{%
+\section{Implementation Issues in B\&S}\label{implementation-issues-in-bs}}
+
+While replicating and analyzing B\&S's results, we found several issues worthy of further discussion and investigation.
+These are (a) the use of smoothed ROC curves to draw conclusions about which model is best, (b) incorrect implementations of the weighted by PITF and PITF analog split-population models, (c) inconsistent test sets for the models examined, and (d) incorrect scoring of the 2016 forecasts.\footnote{We also note there are additional concerns arising from the question of how the random forest models were tuned by B\&S, especially given the way they are used is unorthodox.
+We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.}
+
+We believe that these research decisions and issues lead B\&S to incorrect conclusions.
+The escalation model \textit{is not} the best; it actually performs worse than the atheoretical, kitchen sink model with over \(1000\) variables.
+We discuss these five issues below.
+We defer a complete analysis that corrects all these issues until later, as there are many possible permutations of a seriatim unfolding.
 
 \hypertarget{smoothed-roc-curves}{%
-\subsection{Smoothed ROC curves}\label{smoothed-roc-curves}}
-
-The most consequential issue that we found is that all AUC-ROC values
-reported in BS Tables 1 and 2 are calculated using smoothed ROC curves,
-not the original, actual ROC curves. A reference to smoothing is made in
-a single sentence in the paper (p.~12):
+\subsection{Smoothed ROC Curves}\label{smoothed-roc-curves}}
+
+The most consequential issue that we found is that all AUC-ROC values reported in B\&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves.
+B\&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation.
+However, smoothing is also used in all AUC-ROC values they report.
+There is no difficulty to be overcome in using non-smoothed plots.
+Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs.
+We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), in order to survey the use of ROC curves.
+Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values.
+We found that 90\%--17 of 19--clearly used empirical ROC curves, and 2 (10\%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code.
+The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S.
+The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey.
+Both of the ambiguous cases include Blair as coauthor.} The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.
+What is surprising is that anyone would choose the smoothed ROC implementation.
+
+The next three issues we encountered all concern information in B\&S Table 2, which is the key table in which B\&S report AUC-ROC values to demonstrate the ``Escalation'' models superiority.
+
+\hypertarget{incorrect-weighted-by-pitf-implementation}{%
+\subsection{Incorrect ``Weighted by PITF'' Implementation}\label{incorrect-weighted-by-pitf-implementation}}
+
+The ``Weighted by PITF'' model is described as follows in B\&S, page 19:
 
 \begin{quote}
-Figure 1 displays the corresponding ROC curves, smoothed for ease of
-interpretation.
+The {[}Weighted by PITF model{]} uses PITF predicted probabilities to weight the results of the escalation model, ensuring that high-risk and low-risk countries that happen to take similar values on ICEWS-based predictors are nonetheless assigned different predicted probabilities in most months.
 \end{quote}
 
-This implies that the ROC curves were only smoothed in the references
-Figure 1, but actually all AUC-ROC calculations throughout the
-replication code use an option to smooth the ROC curves prior to AUC
-calculation.
-
-Figure 1 shows our replication of both the smoothed ROC curves BS
-report, and the actual ROC curves on the right.
+We believe that B\&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set.
+However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.
+
+As a result: (1) in the test set, all countries with names that alphabetically start after ``Tongo'' are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical.
+For example, the ``Weighted by PITF'' model predictions for Syria from 2008 to 2015 are the Escalation test predictions for that time period weighted, respectively, by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.
+
+This appears to be an easily corrected coding error on the part of B\&S.
+
+\hypertarget{incorrect-pitf-split-population-implementation}{%
+\subsection{Incorrect ``PITF Split Population'' Implementation}\label{incorrect-pitf-split-population-implementation}}
+
+The ``PITF Split Population'' model also appears to be incorrectly implemented, owing to a coding error. B\&S describe it on page 20:
+
+\begin{quote}
+The final approach is a random forest analog to split-population modeling.
+We first compute the average PITF predicted probability for each country across all years in our training set.
+We define those that fall in the bottom quartile as ``low risk'' and the rest as ``high risk.'' We then run our escalation model on the high-risk and low-risk subsets separately, combining the results into a single random forest {[}\ldots{]}.
+\end{quote}
+
+This description suggests that B\&S intended to run two separate random forest models, one each on the low- and high-risk training data splits.
+The replication code does indeed run two separate random forecasts, but they both utilize the \emph{exact same training data}, which consist of the full training data from all other models.
+In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.
+
+The model specifications are also identical; i.e., they use the same \emph{x} variables and the same random forest hyper-parameter settings.
+The \emph{only} difference in the models as they are implemented in the B\&S replication code is due to the non-deterministic nature of the random forest model itself.
+If we ran both with the same random seed, they would be identical in every respect, producing two forests of identical decision trees and, thus, identical predictions.
+
+The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling (Chiba, Metternich, and Ward 2015; Beger et al.
+2017).
+Although the two random forest models are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger random forest using both component model's underlying decision trees.
+Thus, while all random forest models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (\texttt{ntree}), the new random forest model after combination does have 200,000 decision trees.
+Furthermore, the PITF model predictions do not impact how the combined random forest model predicts at all, not even through a binary low-/high-risk split.
+The split-population PITF random forest model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing random forest randomization facilities (row and column sampling for each decision tree).
+This does not adequately implement their split-sample modeling strategy.
+
+\hypertarget{inconsistent-test-set-n-for-the-models-in-table-2}{%
+\subsection{Inconsistent test set N for the models in Table 2}\label{inconsistent-test-set-n-for-the-models-in-table-2}}
+
+Further, the AUC-ROC values reported in the original B\&S Table 2 are calculated on different numbers of underlying test set cases (see Table \ref{tab:table2-N}).
+ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available.
+Those sets differ across models (columns) for each row in B\&S Table 2.
+
+Thus a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
+In other words, the results for different models in B\&S Table 2 are not comparable to one another, and conclusions drawn from such comparison are potentially incorrect.
+
+\hypertarget{incorrect-scoring-of-the-2016-forecasts}{%
+\subsection{Incorrect scoring of the 2016 forecasts}\label{incorrect-scoring-of-the-2016-forecasts}}
+
+B\&S present a confusion matrix to score their 2016-H1 forecasts in Table 4.
+Although the forecasts are for the probability of civil war \emph{onset}, in the replication code, they are actually scored using \emph{incidence} of civil war.
+By definition the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict.
+When forecasting rare events, it is common to have many false positives, i.e.~instances where we predict onset but no onsets occur.
+\emph{Ex ante}, using incidence rather than onset to score onset forecasts should decrease the number of false positives and correspondingly increase the number of true positives (good predictions) and false negatives (missed onsets).
+
+\hypertarget{results-of-the-updated-analysis}{%
+\section{Results of the updated analysis}\label{results-of-the-updated-analysis}}
+
+We now turn to an examination that addresses and fixes the five issues discussed above.\footnote{The code for all of our analysis undertaken for this effort may be found at \url{https://github.com/andybega/Blair-Sambanis-replication}.} The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B\&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B\&S Table 2).
+We will review the substantive implications of our updated analysis below, but the bottom line is that these changes turn B\&S's conclusions on their heads.
+
+\hypertarget{smoothed-roc-curves-and-auc-calculations}{%
+\subsection{Smoothed ROC curves and AUC calculations}\label{smoothed-roc-curves-and-auc-calculations}}
+
+A reference to smoothing is made in a single sentence in B\&S (p.~12):
+
+\begin{quote}
+Figure 1 displays the corresponding ROC curves, \emph{smoothed for ease of interpretation}. (emphasis added)
+\end{quote}
+
+This suggest that using smoothed ROC curves will have no substantive impact on the results; however, in this smoothing determines the comparative results.
+Their statement also implies that the ROC curves were only smoothed in the referenced Figure 1.
+However, this is not the case.
+All AUC-ROC calculations throughout their replication code use an option to smooth the ROC curves \emph{prior} to AUC calculation.
+
+ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like.
+Figure \ref{fig:rocs} shows our replication of both the estimated smoothed ROC curves from the B\&S report (left-hand side) and the actual empirical ROC curves (right-hand side).
+The standard method is to compute the area under the curve (AUC) statistic on the original, empirical ROC curves that are shown on the right.
 
 \begin{figure}
+\caption{Replication of B\&S Figure 1 with both smooth and non-smoothed ROC curves\label{fig:rocs}}
 \centering
-\includegraphics{figures/figure1-replicated.png}
-\caption{Replication of BS Figure 1 with both smooth and non-smoothed
-ROC curves.}
+\includegraphics[width=.95\linewidth]{figures/figure1-replicated.png}
 \end{figure}
 
-ROC curves typically appear step-like in response to the distribution of
-positive and negative cases in the data. In this case there are also
-groups of cases with identical predicted probabilities, which accounts
-for the unusual diagonal lines. In any case, with a sparse outcome like
-civil war onsets, the true positive rate on the \emph{y}-axis only
-changes when the prediction for a observed positive case is reached. For
-these ROC curves, and for that matter in the basic train/test split used
-for 12 of the 18 rows/models in BS Table 1, there are only 11 civil war
-onset cases in the test set. Thus the ROC curves here are very
-step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct
-\emph{y} coordinates. We can only speculate, but maybe the very
-``step''-like nature of the ROC curves plus the presence of unusual
-diagonal lines accounts for the decision to use smoothed ROC curves.
-
-As it turns out, this decision has a dramatic impact on the AUC-ROC
-calculations. As we will show in the results further below, the
-difference between the smoothed and original ROC AUC values is up to
-0.12--a huge difference given that row-wise, the models being compared
-typically differ by only 0.05 or less--and differentially impacts the
-models that are being compared. In fact, BS's original results and
-interpretation are entirely conditional on the use of smoothed AUC-ROC.
-
-To our knowledge, the norm is to calculate AUC-ROC values using
-original, not smoothed ROC curves. We are in fact not aware of other
-work that uses smoothed ROC curves for AUC calculations. It might be
-that there are theoretical reasons justifying the use of smoothed ROC
-curves over the original ROC curves; but given that this decision
-dramatically impacts the interpretation of the BS results, it minimally
-would have warranted an explicit discussion in the paper. This is not
-the case, and it is only mentioned in the sentence we quote above.
-
-The next three issues we encountered all concern information in Table 2.
-
-\hypertarget{incorrect-weighted-by-pitf-implementation}{%
-\subsection{Incorrect ``Weighted by PITF''
-implementation}\label{incorrect-weighted-by-pitf-implementation}}
-
-The ``Weighted by PITF'' model is described as follows in BS, page 19:
-
-\begin{quote}
-The {[}Weighted by PITF model{]} uses PITF predicted probabilities to
-weight the results of the escalation model, ensuring that high-risk and
-low-risk countries that happen to take similar values on ICEWS-based
-predictors are nonetheless assigned different predicted probabilities in
-most months.
-\end{quote}
-
-We infer that the intent is that the Escalation model's predictions for
-the test set are weighted by the PITF model predictions for the test
-set. The implementation actually weighs the test set predictions using
-the PITF model predictions for the training set.\footnote{See
-  \texttt{1mo\_run\_escalation\_weighted\_PITF.R} line 4, where the PITF
-  predictions are taken from the training data set
-  (\texttt{train\$pred\_prob\_plus1}). The next line is a hack extending
-  the shorter \texttt{weight} vector with missing values to avoid a R
-  warning when it is multiplied with the longer vector of escalation
-  model test set predictions. Similarly in the 6 month version of this
-  file.} This appears to be a coding error.
-
-\hypertarget{incorrect-pitf-split-population-implementation}{%
-\subsection{Incorrect ``PITF Split Population''
-implementation}\label{incorrect-pitf-split-population-implementation}}
-
-Similarly, the ``PITF Split Population'' model appears to be incorrectly
-implemented. BS describe it as, on page 20:
-
-\begin{quote}
-The final approach is a random forest analog to split-population
-modeling. first compute the average PITF predicted probability for each
-country across all years in our training set. We define those that fall
-in the bottom quartile as ``low risk'' and the rest as ``high risk.'' We
-then run our escalation model on the high-risk and low-risk subsets
-separately, combining the results into a single random forest (column
-4).
-\end{quote}
-
-The intention clearly was to run two separate random forest models, one
-each on the low- and high-risk training data splits. The replication
-code does indeed run two separate random forecasts, but they are both
-run on the exact same training data, which consists of the full training
-data all other models are run on. The models are also identical
-otherwise, i.e.~they use the same \emph{x} variables and the same random
-forest hyper-parameter settings. The \emph{only} difference in the
-models as they are implemented in the BS replication code is due to the
-non-deterministic nature of the random forest model itself. If we ran
-both with the same RNG seed, they would be identical in every respect,
-producing identical predictions.\footnote{Disentangling this coding
-  error is not straightforward as it occurs over several R scripts and
-  requires (or at least is easier to verify by) running partway through
-  the actual replication until the objects holding the training data for
-  the models are instantiated and can be examined. We have documented
-  details at
-  \url{https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5}.}
-
-The implementation error aside, this split-population analog model is
-actually quite odd and does not actually replicate an analogue of the
-idea behind split-population modeling. Although the two RFs are trained
-on separate data (in our updated, fixed replication), the process of
-combining them actually just creates a new, larger RF using both
-component model's underlying decision (regression\footnote{See further
-  below. Although the RF models are used for a binary decision problem,
-  the actual implementation uses regression RFs for continuous outcomes.})
-trees. Thus while all RF models throughout (except for one of the
-robustness checks) are trained with 100,000 decision trees
-(\texttt{ntree}), the new RF model after combination does indeed have
-200,000 decision trees. Furthermore, the PITF model predictions do not
-impact the way the combined RF model predicts at all, not even through a
-binary low-/high-risk split. The split-population PITF RF model is
-practically speaking just another Escalation model trained with
-N=200,000 instead of N=100,000 trees and an extra odd randomization step
-added to the already existing RF randomization facilities (row and
-column sampling for each decision tree).
-
-\hypertarget{inconsistent-test-set-n-for-the-models-in-table-2}{%
-\subsection{Inconsistent test set N for the models in Table
-2}\label{inconsistent-test-set-n-for-the-models-in-table-2}}
-
-Lastly, the AUC-ROC values reported in the original BS Table 2 are
-calculated on the basis of slightly different numbers of underlying test
-set cases (see Table \ref{table2-N}). ROC calculations for a set of
-predictions can only be done on the set of cases for which both
-non-missing predictions and non-missing outcomes are available. Those
-sets differ across models (columns) for each row in Table 2. Thus a
-difference in AUC-ROC values for two models could be due to the fact
-that they were calculated on different sets of underlying cases, not
-because the models are systemically performing at a different level. In
-other words, the results for different models in BS Table 2 are actually
-not comparable to the other, and any conclusions drawn from such
-comparison are potentially incorrect.
-
-We fix this issue by only using predictions for common joint subset of
-cases that all models have non-missing predictions for. The original BS
-Table 2 1-month have \emph{N}=11,806--12,495 versus a common joint
-subset of 9,811, and for the 6-month row \emph{N}=2,070--2,233 versus
-\emph{N}=1,915 for the common joint subset.
-
-\hypertarget{incorrect-scoring-of-the-2016-forecasts}{%
-\subsection{Incorrect scoring of the 2016
-forecasts}\label{incorrect-scoring-of-the-2016-forecasts}}
-
-BS show a confusiong matrix to score their 2016-H1 forecasts in Table 4.
-Although the forecasts are for the probability of civil war onset, in
-the replication code they are actually scored using the much more common
-incidence of civil war, i.e.~including ongoing civil wars as ``1'''s.
-
-The relevant variables in the data are ``incidence\_civil\_ns'' and
-``incidence\_civil\_ns\_plus1'', which appears to be a 1-period lead
-version of the DV that is used in the actual prediction models. The
-incidence DV contains both 0/1 and missing values. By examining the
-pattern of missing values, it seems clear that this was originally an
-incidence variable indicating whether a country was at civil war in a
-given year or not, and which was converted to an onset version so that
-onsets retain the value of 1 but continuing civil war years are coded as
-missing. This reflects common practice in how these are coded.
-
-By examining the code used to generated Table 4, we were able to confirm
-that the onset forecasts are assessed using incidence, not onset. In the
-file \texttt{6mo\_make\_confusion\_matrix.do} on line 52, missing values
-in ``incidence\_civil\_ns'' are recoded to 1, thus reverting the onset
-coding of this variable back to incidence.
-
-\hypertarget{results-of-the-updated-replication}{%
-\section{Results of the updated
-replication}\label{results-of-the-updated-replication}}
-
-\hypertarget{basic-interpretation-of-table-1}{%
-\subsection{Basic interpretation of Table
-1}\label{basic-interpretation-of-table-1}}
-
-Table \ref{table1-smooth} is our replication of BS Table 1 with smoothed
-AUC-ROC. The results differ slightly from the original BS Table 1,
-typically by no more than 0.01, due to the non-deterministic nature of
-the RF models. It is the case that BS set the RNG seed in their
-replication code, which should theoretically allow exact reproduction,
-but (1) there was a change in more recent versions of R that affected
-the RNG seeding process, and (2) we refactored the replication script to
-allow one to run the models in parallel. In any case, the interpretation
-of results should not be sensitive to random variation, i.e.~it should
-not depend on using a specific RNG seed. On the basis of these results,
-BS conclude that the Escalation model is generally superior to the
-alternatives, and we can replicate that interpretation when using
-smoothed ROC curves.
-
-\begin{table}[t]
-
-\caption{\label{tab:table1-smooth}Replication of BS Table 1 with smoothed ROC curves; test set AUC-ROC for various models}
+The specific predictions also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines seen in the panels on the right.
+In any case, with a sparse outcome like civil war onsets, the true positive rate on the \emph{y}-axis only changes when the prediction for an observed positive case is reached.
+For these ROC curves, and in the fundamental train/test split used for 12 of the 18 rows/models in B\&S Table 1, for that matter, there are only 11 civil war onset cases in the test set.
+Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct \emph{y} coordinates.
+Notice also that the smoothing averages the left-most almost straight line with the right-most almost straight line in a monotonic way.
+Thus, smoothing changes the ordering of alternatives.
+It is not a cosmetic change as suggested.
+
+What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B\&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B\&S Table 2)?
+Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values.
+Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}).
+The vertical bar in each plot marks the average effect of smoothing on AUC-ROC.
+For all alternative models, from Quad to ``PITF Only'', smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible on average.
+The Escalation model, however, \emph{always} benefits from smoothing, with an average improvement on the order of 0.05.
+This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B\&S report, namely that the escalation model is generally superior.
+
+\begin{figure}
+
+{\centering \includegraphics[width=.8\linewidth]{paper_files/figure-latex/benefit-plot-1}
+
+}
+
+\caption{Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\&S Tables 1 and 2).
+The Escalation model is the only model which consistently benefits from using smoothed ROC curves, and this advantage accounts for its apparent superiority to the other models.}\label{fig:benefit-plot}
+\end{figure}
+
+In sum, it is not only the case that using smoothed ROC curves alters the results, but also that the use of smoothed ROC curves to calculate AUC-ROC values \textbf{only} benefits the escalation model.
+It does so consistently and by a considerable margin.
+All eight alternative models reported in B\&S Tables 1 and 2, on average, do not gain when using smoothed ROC curves to calculate AUC.
+Four of the eight show a small positive bias from smoothing, while the other four display a small negative bias.
+The magnitudes of these are not close to the level of positive bias found in the escalation model.
+
+\hypertarget{is-the-escalation-model-superior-to-the-alternative-icews-models}{%
+\subsection{Is the Escalation model superior to the alternative ICEWS models?}\label{is-the-escalation-model-superior-to-the-alternative-icews-models}}
+
+Table \ref{tab:table1-nosmooth} (B\&S Table 1) shows the comparison of the Escalation model to other alternative models based on ICEWS event data indicators.
+The AUC-ROC values are based on original, non-smoothed ROC curves.
+Both the Average and CAMEO models outperform the Escalation model in almost all instances (see also Figure \ref{fig:table1-plot}).
+The Goldstein model generally outperforms the Escalation model in the 6-month version.
+The Quad model appears to be roughly on par with the Escalation model.
+Thus, the original B\&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves.
+
+\begin{table}
+
+\caption{\label{tab:table1-nosmooth}Comparison of the escalation model to alternative ICEWS model using test set AUC-ROC, \textit{without} smoothed ROC curves (Replication of B\&S Table 1)}
 \centering
-\begin{tabular}{lrrrrr}
+\begin{tabular}[t]{lrrrrr}
 \toprule
@@
 \multicolumn{6}{l}{\textbf{One-month forecasts}}\\
-\hspace{1em}Base specification & 0.85 & 0.80 & 0.79 & 0.83 & 0.82\\
-\hspace{1em}Terminal nodes & 0.86 & 0.79 & 0.78 & 0.83 & 0.82\\
-\hspace{1em}Sample size & 0.85 & 0.81 & 0.70 & 0.86 & 0.84\\
-\hspace{1em}Trees per forest & 0.84 & 0.80 & 0.78 & 0.83 & 0.82\\
-\hspace{1em}Training/test sets 1 & 0.86 & 0.78 & 0.75 & 0.81 & 0.80\\
-\hspace{1em}Training/test sets 2 & 0.81 & 0.79 & 0.72 & 0.78 & 0.78\\
-\hspace{1em}Training/test sets 3 & 0.79 & 0.80 & 0.69 & 0.75 & 0.75\\
-\hspace{1em}Coding of DV 1 & 0.86 & 0.81 & 0.80 & 0.84 & 0.83\\
-\hspace{1em}Coding of DV 2 & 0.92 & 0.81 & 0.81 & 0.82 & 0.81\\
-\addlinespace[0.3em]
-\multicolumn{6}{l}{\textbf{Six-month forecasts}}\\
-\hspace{1em}Base specification & 0.82 & 0.78 & 0.82 & 0.77 & 0.79\\
-\hspace{1em}Terminal nodes & 0.80 & 0.75 & 0.81 & 0.75 & 0.77\\
-\hspace{1em}Sample size & 0.83 & 0.78 & 0.78 & 0.78 & 0.79\\
-\hspace{1em}Trees per forest & 0.82 & 0.78 & 0.82 & 0.77 & 0.79\\
-\hspace{1em}Training/test sets 1 & 0.79 & 0.77 & 0.81 & 0.75 & 0.78\\
-\hspace{1em}Training/test sets 2 & 0.72 & 0.73 & 0.77 & 0.74 & 0.75\\
-\hspace{1em}Training/test sets 3 & 0.88 & 0.70 & 0.81 & 0.68 & 0.79\\
-\hspace{1em}Coding of DV 1 & 0.83 & 0.77 & 0.82 & 0.79 & 0.80\\
-\hspace{1em}Coding of DV 2 & 0.83 & 0.77 & 0.82 & 0.79 & 0.79\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Table \ref{table1-nosmooth} shows a version of BS Table 1 with the
-conventional non-smoothed ROC curves. The Average model outperforms the
-Escalation model in 17 out of 18 cases, and the CAMEO model outperforms
-in 16 of 18 cases, with one tie. The Goldstein model generally
-outperforms the Escalation model in the 6-month version. The Quad model
-appears to be roughly on par with the Escalation model. Thus the
-original BS conclusion that the Escalation model is superior to the
-alternative models is completely conditional on the non-standard use of
-smoothed ROC curves, and overturns when using traditional AUC-ROC
-calculations.
-
-\begin{table}[t]
-
-\caption{\label{tab:table1-nosmooth}Replication of BS Table 1 \textit{without} smoothed ROC curves; test set AUC-ROC for various models}
-\centering
-\begin{tabular}{lrrrrr}
-\toprule
-Model & Escalation & Quad & Goldstein & CAMEO & Average\\
-\midrule
-\addlinespace[0.3em]
-\multicolumn{6}{l}{\textbf{One-month forecasts}}\\
-\hspace{1em}Base specification & 0.78 & 0.78 & 0.79 & 0.80 & 0.82\\
-\hspace{1em}Terminal nodes & 0.79 & 0.78 & 0.78 & 0.81 & 0.82\\
+\hspace{1em}Base specification & 0.78 & 0.78 & 0.80 & 0.81 & 0.82\\
+\hspace{1em}Terminal nodes & 0.79 & 0.78 & 0.79 & 0.81 & 0.82\\
 \hspace{1em}Sample size & 0.79 & 0.80 & 0.74 & 0.82 & 0.84\\
 \hspace{1em}Trees per forest & 0.78 & 0.78 & 0.79 & 0.81 & 0.82\\
-\hspace{1em}Training/test sets 1 & 0.78 & 0.76 & 0.76 & 0.79 & 0.80\\
-\hspace{1em}Training/test sets 2 & 0.75 & 0.77 & 0.73 & 0.76 & 0.78\\
-\hspace{1em}Training/test sets 3 & 0.70 & 0.79 & 0.69 & 0.73 & 0.74\\
-\hspace{1em}Coding of DV 1 & 0.80 & 0.79 & 0.80 & 0.82 & 0.83\\
-\hspace{1em}Coding of DV 2 & 0.80 & 0.82 & 0.78 & 0.83 & 0.81\\
+\hspace{1em}Training/test sets 1 & 0.77 & 0.76 & 0.77 & 0.79 & 0.80\\
+\hspace{1em}Training/test sets 2 & 0.75 & 0.77 & 0.74 & 0.76 & 0.78\\
+\hspace{1em}Training/test sets 3 & 0.71 & 0.79 & 0.69 & 0.72 & 0.74\\
+\hspace{1em}Coding of DV 1 & 0.80 & 0.80 & 0.80 & 0.82 & 0.83\\
+\hspace{1em}Coding of DV 2 & 0.80 & 0.82 & 0.77 & 0.83 & 0.81\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Six-month forecasts}}\\
-\hspace{1em}Base specification & 0.77 & 0.78 & 0.83 & 0.79 & 0.81\\
+\hspace{1em}Base specification & 0.77 & 0.79 & 0.83 & 0.78 & 0.81\\
 \hspace{1em}Terminal nodes & 0.77 & 0.78 & 0.82 & 0.78 & 0.79\\
-\hspace{1em}Sample size & 0.78 & 0.77 & 0.80 & 0.80 & 0.82\\
-\hspace{1em}Trees per forest & 0.77 & 0.79 & 0.83 & 0.78 & 0.81\\
-\hspace{1em}Training/test sets 1 & 0.75 & 0.79 & 0.82 & 0.78 & 0.80\\
+\hspace{1em}Sample size & 0.77 & 0.77 & 0.80 & 0.80 & 0.82\\
+\hspace{1em}Trees per forest & 0.77 & 0.79 & 0.83 & 0.79 & 0.81\\
+\hspace{1em}Training/test sets 1 & 0.75 & 0.79 & 0.82 & 0.77 & 0.80\\
 \hspace{1em}Training/test sets 2 & 0.70 & 0.75 & 0.78 & 0.75 & 0.77\\
-\hspace{1em}Training/test sets 3 & 0.85 & 0.72 & 0.84 & 0.71 & 0.81\\
-\hspace{1em}Coding of DV 1 & 0.77 & 0.80 & 0.83 & 0.79 & 0.82\\
+\hspace{1em}Training/test sets 3 & 0.85 & 0.72 & 0.84 & 0.72 & 0.81\\
+\hspace{1em}Coding of DV 1 & 0.78 & 0.80 & 0.83 & 0.79 & 0.82\\
 \hspace{1em}Coding of DV 2 & 0.80 & 0.78 & 0.84 & 0.80 & 0.81\\
@@
 
-\hypertarget{do-structural-variables-add-to-the-escalation-model}{%
-\subsection{Do structural variables add to the Escalation
-model?}\label{do-structural-variables-add-to-the-escalation-model}}
-
-Table \ref{table2-fixed} shows our replication of BS Table 2 with (1)
-regular, not smoothed, AUC-ROC, (2) fixed ``Weighted by PITF'' and
-``PITF Split-Population'' models, and (3) AUC-ROC values computed on the
-common joint subset of tests cases that all models have non-missing
-predictions for. Table \ref{table2-full} further below shows AUC-ROC
-values for both smoothed and non-smoothed versions, and both the
-original, model-varying test cases sets and our common joint subset.
-
-\begin{table}[t]
-
-\caption{\label{tab:table2-fixed}Replication of BS Table 2: Test set AUC-ROC for Escalation model with and without structural PITF contribution}
+The key difference between Table \ref{tab:table1-nosmooth} and B\&S Table 1 is whether the underlying ROC curves were original or smoothed; the test set N and coding issues did not affect this set of results.\footnote{We should note that in our replication we consciously decided to not set RNG seeds, even though the random forest models are non-deterministic and vary slightly from run to run.
+As we changed the replication code to allow running in parallel, we cannot exactly reproduce the B\&S results even with the same RNG seed.
+More importantly, the substantive interpretation of results should not depend on a specific RNG seed.
+We found that the escalation model's AUC-ROC values generally fluctuate no more than 0.01 (see \url{https://github.com/andybega/Blair-Sambanis-replication/blob/master/rep_nosmooth/variance.md}), and thus are confident that the patterns we see are robust to the initial RNG state.} B\&S's original results and interpretation regarding the superiority of the B\&S model over alternative ICEWS models, including the 1,000+ covariate CAMEO model, are thus entirely conditional on the use of smoothed AUC-ROC.
+Even the Quad model is generally as good as the escalation model in all implementations and is frequently quite a bit better.
+
+Aside from the visual impact of ROC smoothing, the underlying motivation and methods involved in smoothing are broader and involve parametric estimation of population ROC curves (Hanley 1988, 2014).
+The main application is in the evaluation of medical tests (e.g.~diagnostic imaging) with early methods dating back several decades and ongoing development of both parametric and non-parametric smoothing methods (e.g.~Zou, Hall, and Shapiro 1997; Pulit 2016).
+It is not established which methods are preferable in a given application (Hanley 2014).
+Notably, the context in which ROC smoothing is used differs in important aspects from typical conflict research applications.
+Conflict data are closer to a census of the population and with well-known dependencies like spatial and temporal correlation, rather than a random sample that is approximately independently and identically distributed.
+It is thus neither clear that smoothing is justified nor valid.
+We also note that smoothed ROC plots are not widely used outside of the narrow context mentioned above.
+
+\hypertarget{using-inconsistent-ns}{%
+\subsection{Using Inconsistent Ns}\label{using-inconsistent-ns}}
+
+The top portion of Table \ref{tab:table2-N} shows the number of valid test set predictions that can be scored for each model in the original B\&S Table 2.
+For the 1-month data version, the number of predictions differs by up to 500 cases, and in the 6-month data version by around 100 cases.
+These numbers appear small enough to be not important.
+However, the already small number of positive cases also is affected substantially: ``With PITF Predictors'' and ``PITF Only'' models lose two or one (respectively) of 11 positive cases in the 1- and 6-month data versions.
+
+\begin{table}
+
+\caption{\label{tab:table2-N}Number of valid test predictions for the escalation and structural alterantives models (B\&S Table 2)}
+\centering
+\begin{tabular}[t]{ll>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
+\toprule
+ & Horizon & Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{7}{l}{\textbf{Original model-specific cases}}\\
+\hspace{1em} & 1 month & 13748 & 13155 & 13461 & 13748 & 13510\\
+\cmidrule{2-7}
+\hspace{1em} & 6 months & 2366 & 2264 & 2317 & 2366 & 2265\\
+\cmidrule{1-7}
+\addlinespace[0.3em]
+\multicolumn{7}{l}{\textbf{Cases adjusted to common subset}}\\
+\hspace{1em} & 1 month & 13062 & 13062 & 13062 & 13062 & 13062\\
+\cmidrule{2-7}
+\hspace{1em} & 6 months & 2250 & 2250 & 2250 & 2250 & 2250\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Generally, these comparisons require the same cases.
+Therefore, we use predictions for a common joint subset of non-missing predictions.
+The resulting numbers of cases for each model are shown in the bottom portion of the table.
+Since the sets of incomplete cases in the original version above do not entirely overlap themselves, the common subset for all models is slightly smaller than the smallest N in the top portion of the table.
+
+\hypertarget{do-structural-variables-add-to-the-escalation-models-predictive-power}{%
+\subsection{Do Structural Variables Add to the Escalation Model's Predictive Power?}\label{do-structural-variables-add-to-the-escalation-models-predictive-power}}
+
+Table \ref{tab:table2-fixed} shows our replication of B\&S Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed ``Weighted by PITF'' and ``PITF Split-Population'' models, and (3) AUC-ROC values computed on the common, joint subset of tests cases for which all models have non-missing predictions.
+Table \ref{tab:table2-full} in the appendix shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.
+
+\begin{table}
+
+\caption{\label{tab:table2-fixed}Comparison of escalation model to structural extensions, using test set AUC-ROC (Replication of B\&S Table 2)}
 \centering
 \begin{threeparttable}
-\begin{tabular}{r>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{}p{2cm}}
+\begin{tabular}[t]{r>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{}p{2cm}}
 \toprule
@@
 \multicolumn{5}{l}{\textbf{One-month forecasts}}\\
-\hspace{1em}0.69 & 0.71 & 0.69 & 0.81 & 0.67\\
+\hspace{1em}0.75 & 0.78 & 0.78 & 0.85 & 0.75\\
 \addlinespace[0.3em]
 \multicolumn{5}{l}{\textbf{Six-month forecasts}}\\
-\hspace{1em}0.79 & 0.87 & 0.80 & 0.69 & 0.71\\
+\hspace{1em}0.76 & 0.86 & 0.80 & 0.77 & 0.74\\
 \bottomrule
@@
 \item \textit{Note: } 
-\item Differences from the original BS Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.
+\item Differences from the original B\&S Table 2: (1) AUC-ROC values are computed on the common subset of cases, meaning that N is equal in each row; (2) AUC-ROC values are computed using original, non-smoothed ROC curves.
 \end{tablenotes}
@@
 
-BS interpret the results as follows, on page 20:\footnote{We list the
-  ``Overall, \ldots{}'' interpretation out of order, last, for clarity.}
-
-\begin{quote}
-Of the approaches we test, the split-population analog is most promising
-\end{quote}
-
-This is no the case anymore. It outperforms in the 1-month version and
-underperforms the Escalation model in the 6-month version.
-
-\begin{quote}
-Adding PITF predictors improves the performance of the escalation model
-over six-month windows but diminishes it over one-month windows.
-\end{quote}
-
-Adding PITF predictors actually improves performance in both cases; the
-``With PITF Predictions'' model strictly dominates the ``Escalation
-Only'' model.
-
-\begin{quote}
-The weighted model performs very poorly regardless.
-\end{quote}
-
-It performs roughly on par with the Escalation Only model.
-
-One finding that remains is that the ``PITF Only'' model is outperformed
-by the ``Escalation Only'' model. As the former only uses annual inputs,
-but the data at hand are the 1-month or 6-months level, this is not
-surprising.
-
-\begin{quote}
-Overall, our results suggest that while measures of structural risk may
-improve predictive performance, the value they add is marginal and
-inconsistent. {[}\ldots{}{]} Incorporating PITF thus significantly
-reduces or only slightly improves the performance of the escalation
-model, regardless of the approach we take.
-\end{quote}
-
-The most straightforward method of incorporating the annual, structural
-PITF variables--adding them to the predictors of the Escalation RF
-model--strictly outperforms the Escalation Only model. Note that the two
-other combination models considered are both non-standard, and that the
-``PITF Split Population'' model does not in fact actually incorporate
-structural information at all. We thus conclude that adding structural
-variables actually clearly improves predictive performance.
-
-\hypertarget{the-effect-of-using-smoothed-roc-curves}{%
-\subsection{The effect of using smoothed ROC
-curves}\label{the-effect-of-using-smoothed-roc-curves}}
-
-\begin{table}[t]
-
-\caption{\label{tab:table1-benefit}Smoothing advantage for BS Table 1: the gain in AUC-ROC when calculated using smoothed ROC curves}
+B\&S interpret their results as follows (page 20) and we comment on the conclusions seratim:\footnote{We list the ``Overall, \ldots{}'' interpretation out-of-order for clarity.}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  B\&S: ``Of the approaches we test, the split-population analog is most promising \ldots{}''
+\end{enumerate}
+
+The PITF split-population analog still performs well, but the simple Escalation + PITF predictors model performs better still.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  B\&S: ``Adding PITF predictors improves the performance of the escalation model over six-month windows but diminishes it over one-month windows \ldots ''
+\end{enumerate}
+
+Adding PITF predictors actually improves performance in both cases; the ``With PITF Predictions'' model strictly dominates the ``Escalation Only'' model by quite a margin in the 6-month data version.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  B\&S: ``The weighted model performs very poorly regardless \ldots ''
+\end{enumerate}
+
+The weighted model performs roughly on par with the Escalation Only model.
+One finding that remains is that the ``PITF Only'' model is outperformed by the ``Escalation Only'' model.
+As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is neither surprising, nor especially noteworthy.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{3}
+\tightlist
+\item
+  B\&S: ``Overall, our results suggest that while measures of structural risk may improve predictive performance, the value they add is marginal and inconsistent.
+{[}\ldots{]} Incorporating PITF thus significantly reduces or only slightly improves the performance of the escalation model, regardless of the approach we take.
+{[}\ldots{]}''
+\end{enumerate}
+
+The most straightforward method of incorporating the annual structural PITF variables---adding them to the predictors of the Escalation RF model---strictly outperforms the Escalation Only model.
+Note that the two other combination models considered are both non-standard and that the ``PITF Split Population'' model does not incorporate structural information at all, yet they also both do well.
+We thus conclude that adding structural variables improves predictive performance.
+
+\hypertarget{how-accurate-were-the-2016-forecasts}{%
+\subsection{How Accurate Were the 2016 Forecasts?}\label{how-accurate-were-the-2016-forecasts}}
+
+B\&S made forecasts for the first half of 2016 (2016-H1), and assess the accuracy of these forecasts.
+The forecasts themselves are probabilistic, i.e.~range from 0 to 1, but for the purpose of scoring the forecasts they consider the 30 highest predictions to have been positive forecasts (``1'') and the remaining forecasts to have been for no onset (``0'').
+They then construct confusion matrices for the forecasts, in their Table 4 and which we replicate here in the top half of our Table \ref{tab:table4}.
+They use two alternative codings of the outcomes for 2016-H1, one of which they call ``Assuming Persistence'' (this is what is in their replication data), and one of which they call ``Assuming Change'' (hand-coded in one of their replication scripts).
+
+With the ``Assuming Persistence'' version of the outcomes, shown in the first confusion matrix, B\&S report that there are 15 civil war onsets, of which the forecasts capture 13 (true positives) and miss 2 (false negatives).
+With the ``Assuming Persistence'' coding of outcomes, shown in the second confusion matrix, they report that there are 16 onsets and the forecasts capture 14 of those with, still, 2 misses.
+Fifteen to 16 civil war \emph{onsets} in the first half of 2016 is incredible, especially considering that the entire historical data from 2001 to 2015 only has 20 recorded civil war \emph{onsets}.
+Indeed, it turns out that B\&S scored their forecasts for civil war \textit{onset} using civil war \textit{incidence}, i.e.~counting ongoing civil wars as if they were onsets.
+
+\begin{table}
+\caption{Replication of B\&S Table 4: Confusion Matrices for Six-month Escalation Model forecasts for the first half of 2016.}\label{tab:table4}
 \centering
-\begin{tabular}{lrrrrr}
+\begin{tabular}{ll|c|c|l}
+\multicolumn{5}{c}{\textbf{Original confusion matrices with civil war \textit{incidence}}} \\
+% BS, assuming persistence
+\multicolumn{1}{l}{\multirow{4}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
+\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
+\cline{3-4}
+& $Y=0$ & 132 & 17 \\
+\cline{3-4}
+& $Y=1$ & 2 & 13 \\
+\cline{3-4}
+% BS, assuming change
+\multicolumn{1}{l}{\multirow{4}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
+\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
+\cline{3-4}
+& $Y=0$ & 132 & 16 \\
+\cline{3-4}
+& $Y=1$ & 2 & 14 \\
+\cline{3-4}
+\\ \midrule
+\\
+\multicolumn{5}{c}{\textbf{Fixed confusion matrices with civil war \textit{onset}}} \\
+% fixed, assuming persistence
+\multicolumn{1}{l}{\multirow{4}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
+\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
+\cline{3-4}
+& $Y=0$ & 134 & 30 \\
+\cline{3-4}
+& $Y=1$ & 0 & 0 \\
+\cline{3-4}
+% fixed, assuming change
+\multicolumn{1}{l}{\multirow{4}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
+\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
+\cline{3-4}
+& $Y=0$ & 134 & 28 \\
+\cline{3-4}
+& $Y=1$ & 0 & 2 \\
+\cline{3-4}
+\end{tabular}
+\end{table}
+
+The correct confusion matrices that use civil war onsets to score the forecasts are shown in the second part of Table \ref{tab:table4}.
+In the ``Assuming Persistence'' coding, there are no civil war onsets in the data for 2016-H1.
+The alternative ``Assuming Persistence'' coding records 2 civil war onsets in 2016-H1, according to the civil conflict dataset they rely on.
+As a result, if we compare the precision and recall of the forecasts with the correct scoring to the incorrect scoring in B\&S, almost all of the forecasts are false positives.
+This is the norm when predicting rare events, and thus not unusual at all.
+The forecasts do correctly capture the 2 civil war onsets in the ``Assuming Change'' version of the outcome coding.
+This is, at face value, a credit to the model's accuracy.
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+B\&S advocate the use of theory to guide prediction.
+But ``theory'' is not defined by B\&S and may be an ambiguous and undefined concept.
+``Theory'' has many interpretations in social science.
+For some---the EITM movement, for example (Aldrich, Alt, and Lupia 2008)---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.
+The additional statements can be subjected to empirical scrutiny.
+An example might be William Riker's theory of coalitions (Riker 1962) which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically.
+A more recent example of strong theory is found in \emph{The Political Logic of Survival} (Bueno de Mesquita et al.
+2005).
+For the purposes of discussion we call this strong theory.
+
+However, for many, a theory is simply a falsifiable story or narrative that makes causal claims about the relationships among variables.
+Typically it is fused together from a meta-analysis of prior work on the topic under consideration.
+We call these types of theory weak theory.
+They are weak because they do not include two characteristics of strong theory, though they may include one:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  A theory is separate from the case or cases being examined. It is intended to be general.
+\item
+  A theory generates new statements that in principle can be empirically falsified, not proven.
+\end{enumerate}
+
+Most social science research uses a weak form of theory.
+In the case of B\&S they base their theoretical thinking on research that argues that conflict is not the product of structural factors, but rather comes from an interaction of state repression and the changes in costs and benefits to citizens for engaging in anti-government conflict.
+Following a wide swath of scholars, they argue that the escalation and de-escalations of interactions between citizens and government are what typically lead up to the onset of civil wars.
+Operationally, this boils down to four types of concepts: demands, accommodations, nonviolent repression, and low-level violence.
+
+\begin{itemize}
+\item Demands include events in which either opposition or rebels is the source and government is the target and typically involve appeals for political or institutional reform, policy change, or extension of rights.
+\item Nonviolent repression and accommodation include events in which the government is the source and either opposition or rebels is the target.
+Nonviolent repression includes threats, sanctions, restrictions on freedom of movement, or rejections of group demands.
+\item Low-level violence includes all events in which (1) the government is the source and opposition is the target, (2) the opposition is the source and government is the target, (3) the government is the source and rebels are the target, or (4) rebels are the source and government is the target.
+\end{itemize}
+
+These variables, taken from ICEWS, are the main components of the escalation model in the B\&S article.
+The escalation theory is that the forces behind these variables interact such that as each of them increases, they independently increase the probability of civil conflict.
+The actual mechanisms of the dynamism among these variables are not specified.
+
+Indeed, most empirical social science research follows this approach: (1) delineation of an argument, (2) selection of variables that represent a plausible or compelling representation of the concepts, followed by (3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation.
+It is worth pointing out that this is simply one set of variables that are selected to represent the concepts in the theoretical story.
+
+A strong theory gives more guidance on compelling choices than does a weak theory.
+Accordingly, while the choice is plausible it is not necessarily compelling.
+Another choice might be equally plausible.
+In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups.
+In practical terms there is a wide array of choices that are available to instantiate any particular theory.
+A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated dynamics.
+One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.
+
+At this level of detail, the theoretical model is more about choosing the correct variables and this is the contrast between theoretical prediction models and atheoretical machine learning that B\&S focus on.
+But this dichotomy is misguided.
+The data that are used in a predictive model have to come from somewhere.
+Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful.
+It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well.
+Rather, we posit that most of the cases B\&S might characterize as atheoretical enter the ``theoretical'', with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}
+
+Furthermore, if what we mean by the concept ``theory'' is such that it suggests a large range of variables that could be related to an outcome, what do we really gain from having this theory in our predictive model?
+We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off?
+Certainly that's not what B\&S are trying to suggest since this would be a problematic way to ``evaluate'' a theory.
+But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could in fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and associated ``theory'' with any eye towards out-of-sample predictive accuracy (cf.~Colaresi and Mahmood 2017 for how to avoid this problem while leveraging out-of-sample predictions).
+Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize and whose underlying theories may by extension not be externally valid (for a more detailed treatment of this issue, see Fariss and Jones 2018).
+
+Conversely, not all factors that are ``theoretically'' important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory.
+For example, both Ward, Greenhill, and Bakke (2010) and Hill and Jones (2014) examine a variety of variables that are considered ``theoretically'' important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes.
+And, it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.~the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al.
+2010), irregular leadership changes (Beger et al.
+2017), and coups (Beger and Ward 2020).
+
+Aside from selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form.
+Theory should not only tell us what factors matter, but also how they relate to each other.
+We think that few would be strongly committed to the idea that the linear additive equations that dominate published regression analyses, maybe with the occasional polynomial or multiplicative interaction terms, are \emph{the} correct functional specification.
+Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low?
+Are those relationships linear in the log-odds?
+Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships.
+B\&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives.
+But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory (see also Fariss and Jones 2018)?
+
+Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider the ways in which they can mutually reinforce each other.
+Even if the analysis in B\&S fails to support the point, we concede that predicting on the basis of a strong theoretical model is preferable to inductive prediction.
+Beyond accurate predictions, such models would allow us to make causal statements that are harder to make on the basis of ad hoc models.
+But this is hard to do in practice.
+The problem is that many extant theoretical models are in fact so malleable to interpretation that they essentially are ad hoc.
+They are useful for identifying variables that could be useful for prediction, and this is in fact already taking place.
+
+Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning.
+Even if B\&S were correct that in their study the theory based escalation model performed better, it would simply be a case study of a single case where it was preferable.
+It establishes no general tendency, let alone law-like evidence for general theory.
+Nor, against it.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+We encountered several issues in the code underlying the B\&S analysis.
+The problems we encountered are not subjective modeling choices.
+They bear no implications on the question of whether theory based model are preferable.
+However, when we fix these issues and perform an updated analysis, the B\&S conclusions are all essentially overturned.
+In other words, B\&S findings are based on a faulty analysis, invalid, and should be disregarded as evidence in favor of theory-based models.
+
+In contradistinction to the conclusions offered by B\&S, we find that when correctly specified and implemented:
+
+\begin{itemize}
+\tightlist
+\item
+  The theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model and the ensemble average model.
+\item
+  The Average ensemble model and the CAMEO models outperform the escalation model in all instances.
+\item
+  Adding structural variables substantially improves the escalation model's performance.
+\end{itemize}
+
+More importantly, trying to contrast machine learning and theory may be misguided.
+Even if a ``theory'' section is missing, predictive models are informed by previous work and arguments.
+If we had strong general theories of something like civil war onset, we could replace more inductive, ad hoc, methods with clearly specified models.
+But we do not.
+What we have instead are weak theories that could be interpreted to suggest any range of equations for predicting an outcome.
+Instead of trying to find an instantiation of such a ``theoretical'' model that can outperform a more ad hoc alternative, we could use the flexibility of machine learning methods and insights that predictive modeling brings to improve our weak theories.
+
+Blair \& Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain.
+As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance.
+However, these comparisons must be made carefully to ensure those correct inferences are drawn.
+We continue to think that weak theory is overrated (Ward 2016), and that machine learning and big data will allow us to learn new things.
+It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.
+Social science has leaned on theory for (at least) seventy years, but it might be time to find useful ways to incorporate some of the recently developed methods in the arena of machine learning and modern statistics as well.
+
+\newpage
+
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\hypertarget{refs}{}
+\begin{cslreferences}
+\leavevmode\hypertarget{ref-aldrich:etal:2008}{}%
+Aldrich, John H., James E. Alt, and Arthur Lupia. 2008. ``The Eitm Approach: Origins and Interpretations.'' In \emph{The Oxford Handbook of Political Methodology}.
+
+\leavevmode\hypertarget{ref-beger:etal:2017}{}%
+Beger, Andreas, Daniel W.
+Hill, Jr., Nils.
+W.
+Metternich, Shahryar Minhas, and Michael D.
+Ward.
+2017.
+``Splitting It Up: The \texttt{spduration} Split-Population Duration Regression Package for Time-Varying Covariates.'' \emph{The \(\mathcal{R}\) Journal} 9 (2): 474--86.
+\url{https://journal.r-project.org/archive/2017/RJ-2017-056/index.html}.
+
+\leavevmode\hypertarget{ref-beger:etal:2020}{}%
+Beger, Andreas, Richard K.
+Morgan, and Laura Maxwell.
+2020.
+``The Democratic Spaces Barometer: Global Forecasts of Autocratization and Democratization.'' \url{https://www.v-dem.net/en/analysis/DemSpace/}.
+
+\leavevmode\hypertarget{ref-beger:ward:2017}{}%
+Beger, Andreas, and Michael D. Ward. 2017. ``Lessons from Near Real-Time Forecasting of Irregular Leadership Changes.'' \emph{Journal of Peace Research} 54 (2): 141--56.
+
+\leavevmode\hypertarget{ref-beger:ward:2020}{}%
+---------. 2020. ``Coup Forecasts for 2020.'' \url{https://www.predictiveheuristics.com/forecasts}.
+
+\leavevmode\hypertarget{ref-blair:sambanis:2020}{}%
+Blair, Robert A., and Nicholas Sambanis.
+n.d.
+``Forecasting Civil Wars: Theory and Structure in an Age of `Big Data' and Machine Learning.'' \emph{Journal of Conflict Resolution} TBD (firstview): 0022002720918923.
+
+\leavevmode\hypertarget{ref-breiman:2001}{}%
+Breiman, Leo. 2001. ``Random Forests.'' \emph{Machine Learning} 45 (1): 5--32.
+
+\leavevmode\hypertarget{ref-bdm:etal:2005}{}%
+Bueno de Mesquita, Bruce, Alastair Smith, Randolph M. Siverson, and James D. Morrow. 2005. \emph{The Logic of Political Survival}. Cambridge, Mass: MIT Press.
+
+\leavevmode\hypertarget{ref-cederman:weidmann:2017}{}%
+Cederman, Lars-Erik, and Nils B. Weidmann. 2017. ``Predicting Armed Conflict: Time to Adjust Our Expectations?'' \emph{Science} 355 (6324): 474--76.
+
+\leavevmode\hypertarget{ref-chiba:etal:2015}{}%
+Chiba, Daina, Nils W.
+Metternich, and Michael D.
+Ward.
+2015.
+``Every Story Has a Beginning, Middle, and an End (but Not Always in That Order): Predicting Duration Dynamics in a Unified Framework.'' \emph{Political Science Research and Methods} 3 (3): 515--41.
+
+\leavevmode\hypertarget{ref-colaresi:mahmood:2017}{}%
+Colaresi, Michael, and Zuhaib Mahmood. 2017. ``Do the Robot: Lessons from Machine Learning to Improve Conflict Forecasting.'' \emph{Journal of Peace Research} 54 (2): 193--214.
+
+\leavevmode\hypertarget{ref-fariss:jones:2018}{}%
+Fariss, Christopher J., and Zachary M. Jones. 2018. ``Enhancing Validity in Observational Settings When Replication Is Not Possible.'' \emph{Political Science Research and Methods} 6 (2): 365--80.
+
+\leavevmode\hypertarget{ref-goldstone:etal:2010}{}%
+Goldstone, Jack A., Robert H.
+Bates, David L.
+Epstein, Ted Robert Gurr, Michael B.
+Lustik, Monty G.
+Marshall, Jay Ulfelder, and Mark Woodward.
+2010.
+``A Global Model for Forecasting Political Instability.'' \emph{American Journal of Political Science} 54 (1): 190--208.
+
+\leavevmode\hypertarget{ref-henley:1988}{}%
+Hanley, James A. 1988. ``The Robustness of the "Binormal" Assumptions Used in Fitting ROC Curves.'' \emph{Medical Decision Making} 8 (3): 197--203. \url{https://doi.org/10.1177/0272989X8800800308}.
+
+\leavevmode\hypertarget{ref-henley:2014}{}%
+---------.
+2014.
+``Receiver Operating Characteristic (ROC) Curves.'' In \emph{Wiley Statsref: Statistics Reference Online}.
+American Cancer Society.
+\url{https://doi.org/10.1002/9781118445112.stat05255}.
+
+\leavevmode\hypertarget{ref-hegre:etal:2019}{}%
+Hegre, Håvard, Marie Allansson, Matthias Basedau, Michael Colaresi, Mihai Croicu, Hanne Fjelde, Frederick Hoyles, et al.
+2019.
+``ViEWS: A Political Violence Early-Warning System.'' \emph{Journal of Peace Research} 56 (2): 155--74.
+
+\leavevmode\hypertarget{ref-hill:jones:2014}{}%
+Hill, Daniel W., and Zachary M. Jones. 2014. ``An Empirical Evaluation of Explanations for State Repression.'' \emph{American Political Science Review} 108 (03): 661--87.
+
+\leavevmode\hypertarget{ref-liaw:wiener:2002}{}%
+Liaw, Andy, and Matthew Wiener. 2002. ``Classification and Regression by randomForest.'' \emph{R News} 2 (3): 18--22. \url{https://CRAN.R-project.org/doc/Rnews/}.
+
+\leavevmode\hypertarget{ref-malley:etal:2012}{}%
+Malley, James D., Jochen Kruppa, Anirban DasGupta, Karen G.
+Malley, and Andreas Ziegler.
+2012.
+``Probability Machines: Consistent Probability Estimation Using Nonparametric Learning Machines.'' \emph{Methods of Information in Medicine} 51 (1): 74--81.
+\url{https://www.thieme-connect.de/products/ejournals/abstract/10.3414/ME00-01-0052}.
+
+\leavevmode\hypertarget{ref-pulit:2016}{}%
+Pulit, Michał. 2016. ``A New Method of Kernel-Smoothing Estimation of the Roc Curve.'' \emph{Metrika} 79 (5): 603--34.
+
+\leavevmode\hypertarget{ref-riker:1962}{}%
+Riker, William H. 1962. \emph{The Theory of Political Coalitions}. Yale University Press.
+
+\leavevmode\hypertarget{ref-ward:2016}{}%
+Ward, Michael D. 2016. ``Can We Predict Politics? Toward What End?'' \emph{Journal of Global Security Studies} 1 (1): 80--91.
+
+\leavevmode\hypertarget{ref-ward:etal:2010}{}%
+Ward, Michael D., Brian D. Greenhill, and Kristin M. Bakke. 2010. ``The Perils of Policy by P-Value: Predicting Civil Conflicts.'' \emph{Journal of Peace Research} 47 (4): 363--75.
+
+\leavevmode\hypertarget{ref-ward:etal:2013}{}%
+Ward, Michael D., Nils W.
+Metternich, Cassy L.
+Dorff, Max Gallop, Florian M.
+Hollenbach, Anna Schultz, and Simon Weschle.
+2013.
+``Learning from the Past and Stepping into the Future: Toward a New Generation of Conflict Prediction.'' \emph{International Studies Review} 16 (4): 473--644.
+
+\leavevmode\hypertarget{ref-zou:etal:1997}{}%
+Zou, Kelly H., William J.
+Hall, and David E.
+Shapiro.
+1997.
+``Smooth Non-Parametric Receiver Operating Characteristic (ROC) Curves for Continuous Diagnostic Tests.'' \emph{Statistics in Medicine} 16 (19): 2143--56.
+\end{cslreferences}
+
+\newpage
+
+\hypertarget{appendix-appendix}{%
+\appendix}
+
+\hypertarget{appendix}{%
+\section{Appendix}\label{appendix}}
+
+\renewcommand{\thefigure}{A\arabic{figure}}
+\setcounter{figure}{0}
+
+\renewcommand{\thetable}{A\arabic{table}} 
+\setcounter{table}{0}
+
+\hypertarget{code-references-for-coding-errors}{%
+\subsection{Code references for coding errors}\label{code-references-for-coding-errors}}
+
+\hypertarget{section-3.2-incorrect-weighted-by-pitf-implementation}{%
+\subsubsection{Section 3.2: Incorrect ``Weighted by PITF'' Implementation}\label{section-3.2-incorrect-weighted-by-pitf-implementation}}
+
+We believe the intention with the ``Weighted by PITF'' model is to combine the PITF test predictions with the Escalation test predictions, i.e.~weight the latter with the former.
+Due to an error, the PITF \emph{training} set predictions are used, not the \emph{test} set predictions.
+
+See \texttt{1mo\_run\_escalation\_weighted\_PITF.R} line 4, where the PITF predictions are taken from the training data set (\texttt{train\$pred\_prob\_plus1}).
+The next line is a hack extending the shorter \texttt{weight} vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions.
+Similarly in the 6-month version of this file.
+
+Uncovering which cases in the PITF training predictions are matched to the cases that the Escalation model test set predictions correspond to requires tracing back several more data objects.
+We have documented details in issue \#3 in the GitHub repo.
+
+\hypertarget{section-3.3-incorrect-pitf-split-population-implementation}{%
+\subsubsection{Section 3.3: Incorrect ``PITF Split Population'' Implementation}\label{section-3.3-incorrect-pitf-split-population-implementation}}
+
+The intention is to train seperate random forests on data split into high and low-risk groups based on PITF predictions, and then combine them back into one random forest.
+Actually both random forests are trained on identical data, essentially production a random forest that just has N\_trees * 2.
+
+Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined.
+Details are documented in issue \#5 in the GitHub repo.
+
+\hypertarget{section-3.5-incorrect-scoring-of-the-2016-forecasts}{%
+\subsubsection{Section 3.5: Incorrect scoring of the 2016 forecasts}\label{section-3.5-incorrect-scoring-of-the-2016-forecasts}}
+
+The relevant variables in the data are ``incidence\_civil\_ns'' and ``incidence\_civil\_ns\_plus1'', which appears to be a 1-period lead (ie., \(t_+1\)) version of the dependent variable that is used in the actual prediction models.
+The incidence dependent variable contains both 0/1 and missing values.
+By examining the pattern of missing values, it seems clear that this was originally an incidence variable indicating whether a country was at civil war in a given year or not, and which was converted to an onset version so that onsets retain the value of 1 but continuing civil war years are coded as missing.
+This reflects common practice.
+
+However, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset.
+In the file \texttt{6mo\_make\_confusion\_matrix.do} on line 52, missing values in ``incidence\_civil\_ns'' are recoded to 1, thus reverting the onset coding of this variable to incidence.
+
+\hypertarget{additional-replication-tables}{%
+\subsection{Additional replication tables}\label{additional-replication-tables}}
+
+Table \ref{tab:table1-smooth} is our replication of B\&S Table 1 with smoothed AUC-ROC.
+The results differ slightly from the original B\&S Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models.
+It is the case that B\&S set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel.
+In any case, the interpretation of results should not be sensitive to random variation, i.e.~it should not depend on using a specific RNG seed.
+On the basis of these results, B\&S conclude that the escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.
+
+\begin{table}
+
+\caption{\label{tab:table1-smooth}Replication of B\&S Table 1 with smoothed ROC curves; test set AUC-ROC for various models}
+\centering
+\begin{tabular}[t]{lrrrrr}
 \toprule
@@
 \multicolumn{6}{l}{\textbf{One-month forecasts}}\\
-\hspace{1em}Base specification & 0.07 & 0.02 & 0.00 & 0.02 & 0.00\\
-\hspace{1em}Terminal nodes & 0.06 & 0.01 & -0.01 & 0.02 & 0.00\\
-\hspace{1em}Sample size & 0.06 & 0.01 & -0.03 & 0.04 & 0.00\\
-\hspace{1em}Trees per forest & 0.06 & 0.02 & -0.01 & 0.02 & 0.00\\
-\hspace{1em}Training/test sets 1 & 0.09 & 0.02 & -0.01 & 0.02 & 0.00\\
-\hspace{1em}Training/test sets 2 & 0.07 & 0.00 & -0.01 & 0.02 & 0.00\\
-\hspace{1em}Training/test sets 3 & 0.09 & 0.01 & -0.01 & 0.02 & 0.01\\
-\hspace{1em}Coding of DV 1 & 0.06 & 0.02 & 0.00 & 0.03 & 0.00\\
-\hspace{1em}Coding of DV 2 & 0.12 & -0.01 & 0.04 & -0.01 & 0.01\\
+\hspace{1em}Base specification & 0.85 & 0.80 & 0.79 & 0.84 & 0.83\\
+\hspace{1em}Terminal nodes & 0.86 & 0.79 & 0.77 & 0.83 & 0.82\\
+\hspace{1em}Sample size & 0.85 & 0.81 & 0.71 & 0.86 & 0.84\\
+\hspace{1em}Trees per forest & 0.85 & 0.80 & 0.78 & 0.83 & 0.82\\
+\hspace{1em}Training/test sets 1 & 0.86 & 0.78 & 0.76 & 0.81 & 0.80\\
+\hspace{1em}Training/test sets 2 & 0.82 & 0.79 & 0.72 & 0.77 & 0.78\\
+\hspace{1em}Training/test sets 3 & 0.80 & 0.80 & 0.69 & 0.74 & 0.75\\
+\hspace{1em}Coding of DV 1 & 0.86 & 0.81 & 0.80 & 0.84 & 0.83\\
+\hspace{1em}Coding of DV 2 & 0.92 & 0.80 & 0.81 & 0.81 & 0.81\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Six-month forecasts}}\\
-\hspace{1em}Base specification & 0.05 & -0.01 & -0.01 & -0.02 & -0.02\\
-\hspace{1em}Terminal nodes & 0.02 & -0.02 & -0.01 & -0.02 & -0.02\\
-\hspace{1em}Sample size & 0.05 & 0.00 & -0.02 & -0.01 & -0.03\\
-\hspace{1em}Trees per forest & 0.05 & -0.01 & -0.01 & -0.01 & -0.02\\
-\hspace{1em}Training/test sets 1 & 0.04 & -0.01 & -0.01 & -0.02 & -0.02\\
-\hspace{1em}Training/test sets 2 & 0.03 & -0.01 & -0.02 & -0.01 & -0.02\\
-\hspace{1em}Training/test sets 3 & 0.03 & -0.02 & -0.03 & -0.02 & -0.02\\
-\hspace{1em}Coding of DV 1 & 0.05 & -0.03 & -0.01 & -0.01 & -0.02\\
-\hspace{1em}Coding of DV 2 & 0.03 & -0.01 & -0.01 & -0.01 & -0.02\\
+\hspace{1em}Base specification & 0.82 & 0.78 & 0.82 & 0.77 & 0.79\\
+\hspace{1em}Terminal nodes & 0.79 & 0.76 & 0.81 & 0.77 & 0.77\\
+\hspace{1em}Sample size & 0.83 & 0.78 & 0.78 & 0.79 & 0.79\\
+\hspace{1em}Trees per forest & 0.82 & 0.77 & 0.82 & 0.77 & 0.79\\
+\hspace{1em}Training/test sets 1 & 0.80 & 0.78 & 0.81 & 0.76 & 0.78\\
+\hspace{1em}Training/test sets 2 & 0.72 & 0.74 & 0.77 & 0.73 & 0.75\\
+\hspace{1em}Training/test sets 3 & 0.88 & 0.71 & 0.81 & 0.68 & 0.79\\
+\hspace{1em}Coding of DV 1 & 0.83 & 0.77 & 0.82 & 0.79 & 0.80\\
+\hspace{1em}Coding of DV 2 & 0.83 & 0.77 & 0.83 & 0.78 & 0.79\\
 \bottomrule
@@
 
-\begin{table}[t]
-
-\caption{\label{tab:table2-benefit}Smoothing advantage for BS Table 2: the gain in AUC-ROC when calculated using smoothed ROC curves}
+\begin{table}
+
+\caption{\label{tab:table2-full}Replication of B\&S Table 2 with smoothed/original ROC and with original varying N cases or adjusting for common case set with constant N}
 \centering
-\begin{tabular}{l>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
+\begin{tabular}[t]{ll>{\raggedright\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
+\toprule
+ &  & Smoothed ROC & Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{8}{l}{\textbf{Original model-specific cases}}\\
+\addlinespace[0.3em]
+\multicolumn{8}{l}{\textit{1 month}}\\
+\hspace{1em}\hspace{1em} &  & Yes & 0.85 & 0.77 & 0.80 & 0.81 & 0.76\\
+\cmidrule{3-8}
+\hspace{1em}\hspace{1em} &  & No & 0.79 & 0.78 & 0.80 & 0.79 & 0.75\\
+\cmidrule{2-8}
+\addlinespace[0.3em]
+\multicolumn{8}{l}{\textit{6 months}}\\
+\hspace{1em}\hspace{1em} &  & Yes & 0.82 & 0.85 & 0.81 & 0.78 & 0.74\\
+\cmidrule{3-8}
+\hspace{1em}\hspace{1em} &  & No & 0.77 & 0.86 & 0.81 & 0.78 & 0.74\\
+\cmidrule{1-8}
+\addlinespace[0.3em]
+\multicolumn{8}{l}{\textbf{Cases adjusted to common subset}}\\
+\addlinespace[0.3em]
+\multicolumn{8}{l}{\textit{1 month}}\\
+\hspace{1em}\hspace{1em} &  & Yes & 0.81 & 0.77 & 0.79 & 0.86 & 0.76\\
+\cmidrule{3-8}
+\hspace{1em}\hspace{1em} &  & No & 0.75 & 0.78 & 0.78 & 0.85 & 0.75\\
+\cmidrule{2-8}
+\addlinespace[0.3em]
+\multicolumn{8}{l}{\textit{6 months}}\\
+\hspace{1em}\hspace{1em} &  & Yes & 0.82 & 0.85 & 0.80 & 0.77 & 0.74\\
+\cmidrule{3-8}
+\hspace{1em}\hspace{1em} &  & No & 0.76 & 0.86 & 0.80 & 0.77 & 0.74\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}
+
+\caption{\label{tab:table1-benefit}Smoothing advantage for B\&S Table 1: the gain in AUC-ROC when calculated using smoothed ROC curves}
+\centering
+\begin{tabular}[t]{lrrrrr}
+\toprule
+Model & Escalation & Quad & Goldstein & CAMEO & Average\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{6}{l}{\textbf{One-month forecasts}}\\
+\hspace{1em}Base specification & 0.06 & 0.02 & -0.01 & 0.03 & 0.00\\
+\hspace{1em}Terminal nodes & 0.07 & 0.01 & -0.01 & 0.02 & 0.00\\
+\hspace{1em}Sample size & 0.06 & 0.01 & -0.04 & 0.04 & 0.00\\
+\hspace{1em}Trees per forest & 0.06 & 0.02 & -0.01 & 0.02 & 0.00\\
+\hspace{1em}Training/test sets 1 & 0.09 & 0.02 & -0.01 & 0.02 & 0.00\\
+\hspace{1em}Training/test sets 2 & 0.07 & 0.02 & -0.01 & 0.02 & 0.00\\
+\hspace{1em}Training/test sets 3 & 0.09 & 0.01 & 0.00 & 0.02 & 0.01\\
+\hspace{1em}Coding of DV 1 & 0.07 & 0.02 & 0.00 & 0.02 & 0.00\\
+\hspace{1em}Coding of DV 2 & 0.12 & -0.02 & 0.04 & -0.01 & 0.01\\
+\addlinespace[0.3em]
+\multicolumn{6}{l}{\textbf{Six-month forecasts}}\\
+\hspace{1em}Base specification & 0.05 & -0.01 & -0.01 & -0.01 & -0.02\\
+\hspace{1em}Terminal nodes & 0.02 & -0.02 & -0.01 & -0.01 & -0.02\\
+\hspace{1em}Sample size & 0.05 & 0.00 & -0.02 & -0.01 & -0.03\\
+\hspace{1em}Trees per forest & 0.05 & -0.01 & -0.01 & -0.02 & -0.02\\
+\hspace{1em}Training/test sets 1 & 0.04 & -0.01 & -0.01 & -0.02 & -0.02\\
+\hspace{1em}Training/test sets 2 & 0.02 & -0.01 & -0.02 & -0.01 & -0.02\\
+\hspace{1em}Training/test sets 3 & 0.03 & -0.02 & -0.03 & -0.04 & -0.02\\
+\hspace{1em}Coding of DV 1 & 0.05 & -0.03 & -0.01 & 0.00 & -0.02\\
+\hspace{1em}Coding of DV 2 & 0.03 & -0.01 & -0.01 & -0.02 & -0.02\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}
+
+\caption{\label{tab:table2-benefit}Smoothing advantage for B\&S Table 2: the gain in AUC-ROC when calculated using smoothed ROC curves}
+\centering
+\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
 \toprule
@@
 \multicolumn{6}{l}{\textbf{One-month forecasts}}\\
-\hspace{1em}Base specification & 0.05 & -0.03 & 0 & -0.02 & 0.02\\
+\hspace{1em}Base specification & 0.06 & -0.01 & 0.01 & 0.01 & 0.01\\
 \addlinespace[0.3em]
 \multicolumn{6}{l}{\textbf{Six-month forecasts}}\\
-\hspace{1em}Base specification & 0.06 & -0.01 & 0 & 0.01 & 0.00\\
+\hspace{1em}Base specification & 0.06 & -0.01 & 0.00 & 0.00 & 0.00\\
 \bottomrule
@@
 
-What impact did the ROC smoothing have overall on the results reported
-in BS Tables 1 and 2? Tables \ref{table1-benefit} and
-\ref{table2-benefit} show the increase in AUC-ROC when using smoothed
-ROC curves, compared to the standard non-smoothed AUC-ROC. Positive
-values indicate that smoothing increased a model's apparent performance.
-The Escalation model is the only model that consistently had a benefit
-from smoothing. For all 8 other models, smoothing sometimes gave a
-benefit, sometimes not.
-
-A more dramatic difference stands out when we consider the overall
-average impact of smoothing across all flavors of a model reported in
-the columns of the tables. The Average, Goldstein, Quad, PITF Split
-Population, and Weighted by PITF models are slightly hurt by smoothing,
-but on the order of less than 0.01 in absolute magnitude. The ``With
-PITF Predictors'' model is hurt on the order of 0.02, i.e.~it appears
-substantially worse with smoothing. The CAMEO and PITF Only models
-benefit slightly, on the order of 0.01 or less. The Escalation Only
-model on the other hand has an average boost of 0.058 to it's AUC-ROC
-from smoothing. Considering the spread of AUC-ROC values if we compare
-across rows in BS Tables 1 and 2, that boost is substantial.
-
-The use of smoothed ROC curves to calculate AUC-ROC values only clearly
-benefits the Escalation model, it does so consistently, and by a
-considerable margin. All 8 alternative models reported in Tables 1 and 2
-do not on average gain when using smoothed ROC curves to calculate AUC.
-
-\hypertarget{how-accurate-were-the-2016-forecasts}{%
-\subsection{How accurate were the 2016
-forecasts?}\label{how-accurate-were-the-2016-forecasts}}
-
-\hypertarget{additional-concerns}{%
-\section{Additional concerns}\label{additional-concerns}}
-
-\hypertarget{random-forest-hyperparameters}{%
-\subsection{Random forest
-hyperparameters}\label{random-forest-hyperparameters}}
-
-What initially sparked our interest in the paper was the unusual choice
-of hyperparameter settings for the random forest models estimated.
-
-\begin{longtable}[]{@{}llll@{}}
+\hypertarget{model-to-model-comparison-plots}{%
+\subsection{Model to model comparison plots}\label{model-to-model-comparison-plots}}
+
+Figures \ref{fig:table1-plot} and \ref{fig:table2-plot} replicate the information in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} in a way makes the comparison of the escalation model to the alternative models easier.
+Each facet shows the escalation AUC-ROC for all model settings (the rows in Table \ref{fig:table1-plot} and base specification for all models in Table \ref{tab:table2-fixed}) on the left, and the AUC-ROC for an alternative model on the right, with a connecting line.
+If the lines slope up to the right, the alternative model is better.
+
+\begin{figure}
+
+{\centering \includegraphics[width=.9\linewidth]{paper_files/figure-latex/table1-plot-1}
+
+}
+
+\caption{Escalation to alternative comparisons for the ICEWS models (B\&S Table 1)}\label{fig:table1-plot}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=.9\linewidth]{paper_files/figure-latex/table2-plot-1}
+
+}
+
+\caption{Escalation to alternative comparisons for the structural models (B\&S Table 2)}\label{fig:table2-plot}
+\end{figure}
+
+\hypertarget{random-forest-hyper-parameter-selection}{%
+\subsection{Random forest hyper-parameter selection}\label{random-forest-hyper-parameter-selection}}
+
+What initially sparked our interest in the paper was the unusual choice of hyperparameter settings for the random forest models estimated.
+Table \ref{tab:hp} shows the default values used by the implementation of random forest that B\&S use (from the \textbf{randomForest} R package), in contrast to the basic settings used by B\&S for most the models reported in the paper.
+
+As the outcome is a binary indicator of civil war onset, one would typically use a classification random forest that predicts 0 or 1 labels directly.
+The implementation of random forests that B\&S use ((Liaw and Wiener 2002)) is based on the original Breiman (2001) implementation and calculates predictive probabilities by averaging over the ``0'' or ``1'' votes from all constituent decision trees.
+The conventional wisdom regarding the number of trees in a random forest is that it needs to be large enough to stabilize performance, but without any additional gain or harm in accuracy beyond a certain number.
+From the other default settings, which are generally not uninformed choices, one can see that the basic logic is to grow a forest with a relatively small number of trees, but where each tree is fairly extensive, and operates on a large bootstrapped sample of the original training data.
+These are of course only heuristics and it is usual to attempt to find better hyper-parameter methods through some form of tuning.
+
+\begin{table}
+\caption{\label{tab:hp} Random forest (\texttt{randomForest()}) default versus B\&S hyperparameters}
+\begin{tabular}{l>{\raggedright\arraybackslash}p{2in}ll}
 \toprule
-Hyperparameter & Default heuristic & Default values (Escalation) & B\&S
-value\tabularnewline
+Hyperparameter & Default heuristic & Default values (Escalation) & B\&S value \\
 \midrule
-\endhead
-type & & classification & regression\tabularnewline
-ntree & & 500 & 100,000 or 1e6\tabularnewline
-mtry & \texttt{floor(sqrt(ncol(x)))} & 3 & 3\tabularnewline
-replace & & true & false\tabularnewline
-sampsize & \texttt{nrow(x)} if replace, & 11,869 & 100 or
-500\tabularnewline
-& else \texttt{ceiling(.632*nrow(x))} & &\tabularnewline
-nodesize & 1 for classification & 1 & 1\tabularnewline
-maxnodes & & null & 5 or 10\tabularnewline
+type & & classification & regression \\
+ntree & & 500 & 100,000 or 1e6 \\
+mtry & \texttt{floor(sqrt(ncol(x)))} & 3 & 3 \\
+replace & & true & false \\
+sampsize & \texttt{nrow(x)} if replace, else \texttt{ceiling(.632*nrow(x))} & 11,869 & 100 or
+500 \\
+nodesize & 1 for classification & 1 & 1 \\
+maxnodes & & null & 5 or 10 \\
 \bottomrule
-\end{longtable}
-
-There is a stark contrast in the default RF settings and the way Blair
-and Sambanis use the RF models. The default approach is to train a
-relatively small number (ntree; 500) of classification trees, but where
-each tree is fairly big in that it is trained on data that has the same
-number of rows as the training data, albeit sampled with replacement
-(replace is true; sampsize is 11,869), and allowed to grow fairly deep
-(this is governed by nodesize (1), which is the minimum size a terminal
-node must have). In contrast, Blair and Sambanis grow very extensive
-forests with a large number of trees (100,000 compared to 500), but each
-tree is very small and shallow; only a 100 rows are sampled from the
-training data for each tree, and the trees are constrained to at most 5
-terminal nodes (maxnodes).
-
-This approach only works due to the other unorthodox choice, which is to
-use regression, not classification, trees. Trying to use classification
-trees with the other parameter settings in fact does not work at all
-because it is almost guaranteed that a sample of 100 from the 11,869
-training data rows with 9 positive cases will only include 0 (negative)
-outcomes in the sample. As it is, using regression with a 0 or 1 outcome
-produces warnings when estimating the models:
+\end{tabular}
+\end{table}
+
+B\&S in contrast fit very large forests with 100,000 trees in the basic model form, but where each tree only operates on a very small sub-sample (N=100 or 500), drawn without replacement, of the available training data.
+This approach only works due to the choice to use regression, not classification, trees.
+Trying to use classification trees with the other parameter settings does not work at all because it is almost guaranteed that a sample of 100 from the \textasciitilde11,000 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample.
+As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:
 
@@
 
-The RF approach BS use is essentially similar to the probability forests
-proposed in Malley 2012. It is not clear if this is intentional or not,
-as the paper is not cited.
-
-We tried to create versions of the RF models that use the regular
-framework for classification forests, but were unable to create models
-that approach the BS RFs test set performance.
-
-How were these hyper-parameter values determined? Do they work better
-for some models than others?
-
-\hypertarget{auc-roc-sensitivity}{%
-\subsection{AUC-ROC sensitivity}\label{auc-roc-sensitivity}}
-
-RNG and variation
-
-\hypertarget{pre-registration}{%
-\subsection{Pre-registration}\label{pre-registration}}
-
-\begin{itemize}
-\tightlist
-\item
-  The escalation model specification, prior to looking at data!
-\item
-  The hyper-parameter tuning strategy.
-\end{itemize}
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-BS has a clear misunderstanding of the role that out-of-sample
-prediction can plan in analysis. On the one hand it can be used for
-simple forecasting while on the other it can be used to evaluate the
-performance of models, specifically to overfitting and bias. They
-further misrepresent cross-validation which is not about fishing for
-results, as claimed by BS. The current state-of-the-art uses prediction
-and cross-validation most frequently to provide supportive evidence that
-is independent of the estimation procedure and the in-sample data. It is
-incorrect to paint this procedure as atheoretical since many studies
-will have some explanation for how the model was constructed. These
-procedures may simply be used to provide evidence for a theoretical
-argument. Though that argument may be unconvincing and/or atheoretical.
-
-The BS approach that is advocated is to use theory to guide prediction.
-But theory is a quite ambiguous and undefined concept. It is not a
-procedure. What they actually do is to create a model with four right
-hand side variables that is supposed to capture a complicated
-repression-dissent dynamic. The dynamic is probably not linear, but
-their model is. There is a wide-ranging literature on this dynamic that
-they do not rely on to construct their model. As such their baseline
-comparison is hardly a standard bearer for strong theory.
-
-Further, they complete misunderstand the use of ICEWS event data in
-current research. They claim that most uses to date have focused on the
-quad categories, but this ignores a wide swath of literature
-(Steinert-Threlkeld ApsR 2017) and Metternich et al (AJPX 2013) that
-uses a sepcific action---such as protest---defined in the CAMEO
-ontology. In the AJPS article we hand coded, for example, every actor in
-Thailand and focused on an analysis of how those have been interacting.
-
-We encountered several issues in the code underlying the BS analysis.
-
-The issues we encountered are not subjective modeling choices.
-
-When we fix these issue and perform an updated analysis, the conclusions
-BS draw all essentially overturn. In other words, BS findings are based
-on a faulty analysis, and invalid.
-
-Using the same analysis BS intend to use, we in fact find that:
-
-\begin{itemize}
-\tightlist
-\item
-  the theory-driven Escalation model is outperformed both by the
-  low-effort 1,160 predictor all CAMEO model and the Average ensemble
-  model
-\item
-  structural variables substantially improve the Escalation model's
-  performance when added to the pool of predictors the underlying random
-  forest model draws on
-\end{itemize}
-
-\textbf{AB: add a table comparing BS original claims and updated
-results.}
-
-\clearpage
-
-\includegraphics{figures/bands-table4.png}
-
-\begin{table}[t]
-
-\caption{\label{tab:unnamed-chunk-2}2016 Confusion Matrices for Six-month Escalation Model.}
-\centering
-\begin{tabular}{lrrr}
-\toprule
-header & Observed & 0 & 1\\
-\midrule
- & 0 & 134 & 30\\
-\cmidrule{2-4}
-\multirow{-2}{*}{\raggedright\arraybackslash Assuming Persistence} & 1 & 0 & 0\\
-\cmidrule{1-4}
- & 0 & 134 & 28\\
-\cmidrule{2-4}
-\multirow{-2}{*}{\raggedright\arraybackslash Assuming Change} & 1 & 0 & 2\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}[t]
-
-\caption{\label{tab:table2-N}Number of valid test predictions for each cell in BS Table 2}
-\centering
-\begin{tabular}{ll>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
-\toprule
- & Horizon & Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
-\midrule
-\addlinespace[0.3em]
-\multicolumn{7}{l}{\textbf{Original model-specific cases}}\\
-\hspace{1em} & 1 month & 13748 & 13155 & 13461 & 13748 & 13510\\
-\cmidrule{2-7}
-\hspace{1em} & 6 months & 2366 & 2264 & 2317 & 2366 & 2265\\
-\cmidrule{1-7}
-\addlinespace[0.3em]
-\multicolumn{7}{l}{\textbf{Cases adjusted to common subset}}\\
-\hspace{1em} & 1 month & 9811 & 9811 & 9811 & 9811 & 9811\\
-\cmidrule{2-7}
-\hspace{1em} & 6 months & 1915 & 1915 & 1915 & 1915 & 1915\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}[t]
-
-\caption{\label{tab:table2-full}Replication of BS Table 2 with smoothed/original ROC and with original varying N cases or adjusting for common case set with constant N}
-\centering
-\begin{tabular}{ll>{\raggedright\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
-\toprule
- &  & Smoothed ROC & Escalation Only & With PITF Predictors & Weighted by PITF & PITF Split Population & PITF Only\\
-\midrule
-\addlinespace[0.3em]
-\multicolumn{8}{l}{\textbf{Original model-specific cases}}\\
-\addlinespace[0.3em]
-\multicolumn{8}{l}{\textit{1 month}}\\
-\hspace{1em}\hspace{1em} &  & Yes & 0.85 & 0.76 & 0.80 & 0.81 & 0.77\\
-\cmidrule{3-8}
-\hspace{1em}\hspace{1em} &  & No & 0.78 & 0.77 & 0.80 & 0.80 & 0.75\\
-\cmidrule{2-8}
-\addlinespace[0.3em]
-\multicolumn{8}{l}{\textit{6 months}}\\
-\hspace{1em}\hspace{1em} &  & Yes & 0.82 & 0.86 & 0.81 & 0.79 & 0.74\\
-\cmidrule{3-8}
-\hspace{1em}\hspace{1em} &  & No & 0.77 & 0.86 & 0.81 & 0.82 & 0.74\\
-\cmidrule{1-8}
-\addlinespace[0.3em]
-\multicolumn{8}{l}{\textbf{Cases adjusted to common subset}}\\
-\addlinespace[0.3em]
-\multicolumn{8}{l}{\textit{1 month}}\\
-\hspace{1em}\hspace{1em} &  & Yes & 0.74 & 0.68 & 0.69 & 0.80 & 0.69\\
-\cmidrule{3-8}
-\hspace{1em}\hspace{1em} &  & No & 0.69 & 0.71 & 0.69 & 0.81 & 0.67\\
-\cmidrule{2-8}
-\addlinespace[0.3em]
-\multicolumn{8}{l}{\textit{6 months}}\\
-\hspace{1em}\hspace{1em} &  & Yes & 0.85 & 0.86 & 0.79 & 0.70 & 0.71\\
-\cmidrule{3-8}
-\hspace{1em}\hspace{1em} &  & No & 0.79 & 0.87 & 0.80 & 0.69 & 0.71\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-blair2020forecasting}{}%
-Blair, Robert A., and Nicholas Sambanis. 2020. ``Forecasting Civil Wars:
-Theory and Structure in an Age of `Big Data' and Machine Learning.''
-\emph{Journal of Conflict Resolution}.
+As it turns out, using regression random forests for this kind of binary classification problem in order to obtain probability estimates matches the probability random forest approach suggested and positively evaluated in Malley et al.
+(2012), and which is used in another prominent R implementation of random forests.\footnote{The \textbf{ranger} package.} It is not clear whether this is intentional, as the Malley paper is not cited in B\&S.
+
+In any case, B\&S's random forest approach appears to work really well.
+We tried to construct classification random forests tuned via cross-validation on the training data set partition, i.e.~without touching the test data, but were unable to develop models that consistently match the B\&S random forest method in both cross-validated out-of-sample training predictions and test set predictions.
+
+Given that they are relatively unorthodox, yet appear to work very well, we wonder how the hyper-parameter values were determined.
+Two specific concerns are that this was not done with an eye towards test set accuracy, which would invalidate the independence of the out-of-sample test set, and whether the specific hyper-parameter values are optimized for only one model, or were optimized and found to work well for all models.
+There is no discussion of the random forest tuning strategy or how the specific hyper-parameter methods were determined in the paper.
+
+Given the dramatic changes in results as a result of the preceding issues, we did not further investigate these potential concerns.
+
+\hypertarget{different-smoothing-methods}{%
+\subsection{Different smoothing methods}\label{different-smoothing-methods}}
+
+The \textbf{pROC} package \texttt{smooth()} function includes several different smoothing methods.
+The default, which BS use in their code, is binormal smoothing.
+Here is a replication of the smooth benefits shown in Figure \ref{fig:benefit-plot} with two other smoothing methods.
+(We could not get ``logcondens'' and ``logcondens.smooth'' to work, and they are based on another external package that has not been updated since 2016.)
+
+Only the ``binormal'' smoothing method produces a pattern of AUC-ROCs that clearly elevates the Escalation model above the alternatives.
+The ``density'' method does not on average change the AUC-ROC values very much from their empirical ROC curve equivalents.
+The ``fitdistr'' method tends to increase all AUC-ROC values, although maybe slightly more so for the Escalation model than the main alternatives.
+
+\begin{figure}
+
+{\centering \includegraphics[width=.8\linewidth]{paper_files/figure-latex/smooth-benefit-plot-extended-1}
+
+}
+
+\caption{Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\&S Tables 1 and 2), with 3 different smoothing methods (binormal is the default method and used in B\&S)}\label{fig:smooth-benefit-plot-extended}
+\end{figure}
 

### 2020-11-12 paper/paper.tex
recompile paper to make sure its current

@@
 \author{Andreas Beger\footnote{Predictive Heuristics, \href{mailto:adbeger@gmail.com}{\nolinkurl{adbeger@gmail.com}}.} \and Richard K.
-Morgan\footnote{Varieties of Democracy Institute, University of Gothenburg, \href{mailto:rick.morgan2@gmail.com}{\nolinkurl{rick.morgan2@gmail.com}}.} \and Michael D.
+Morgan\footnote{Independent Researcher, \href{mailto:rick.morgan2@gmail.com}{\nolinkurl{rick.morgan2@gmail.com}}.} \and Michael D.
 Ward\footnote{Predictive Heuristics, Duke University, University of Washington, \href{mailto:michael.don.ward@gmail.com}{\nolinkurl{michael.don.ward@gmail.com}}.
 Corresponding author.}}
-\date{10 November 2020}
+\date{12 November 2020}
 
@@
 
-B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars ``remarkably accurately''.
+B\&S claim (page 3) to argue that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars ``remarkably accurately''.
 Indeed, B\&S argue that this theoretical model outperforms four other ``more mechanical'' alternatives.
@@
 2010; Beger and Ward 2017), the W-ICEWS project (Ward et al.
-2013), VIEWS (Hegre et al.
+2013), ViEWS (Hegre et al.
 2019), and others, e.g., Beger, Morgan, and Maxwell (2020).} B\&S claim to be unique in assessing these forecasts.
@@
 
-As a result: (1) in the test set, all countries with names that alphabetically start after ``Tongo'' are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical.
+As a result: (1) in the test set, all countries with names that alphabetically start after ``Tonga'' are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical.
 For example, the ``Weighted by PITF'' model predictions for Syria from 2008 to 2015 are the Escalation test predictions weighted by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.
@@
 Although the two random forest models are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger random forest using both component model's underlying decision trees.
-Thus, while all random forest models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (\texttt{ntree}), the new random forest model after combination does have 200,000 decision trees.
+Thus, while all random forest models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (\texttt{ntree}), the new random forest model after combination has 200,000 decision trees.
 Furthermore, the PITF model predictions do not impact how the combined random forest model predicts at all, not even through a binary low-/high-risk split.
-The split-population PITF random forest model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing random forest randomization facilities (row and column sampling for each decision tree).
+The split-population PITF random forest model is, practically speaking, simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra (odd) randomization step added to the already existing random forest randomization facilities (row and column sampling for each decision tree).
 This does not adequately implement their split-sample modeling strategy.
@@
 
-Thus a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
+Thus, a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level.
 In other words, the results for different models in B\&S Table 2 are not comparable to one another, and conclusions drawn from such comparison are potentially incorrect.
@@
 
-\caption{\label{tab:table2-N}Number of valid test predictions for the escalation and structural alterantives models (B\&S Table 2)}
+\caption{\label{tab:table2-N}Number of valid test predictions for the escalation and structural alternatives models (B\&S Table 2)}
 \centering
@@
 
-B\&S interpret their results as follows (page 20) and we comment on the conclusions seratim:\footnote{We list the ``Overall, \ldots{}'' interpretation out-of-order for clarity.}
+B\&S interpret their results as follows (page 20) and we comment on the conclusions seriatim:\footnote{We list the ``Overall, \ldots{}'' interpretation out-of-order for clarity.}
 
@@
 The original B\&S confusion matrices collapse the second two rows in each matrix and treat ongoing and onset as part of the same group.
-Essentially all of the apparent true positives (\(P=1, Y=1\)) in the original B\&S confusion matrices are ongoing civil war cases which by definition \emph{cannot} experience another civil war onset.
+Essentially, all of the apparent true positives (\(P=1, Y=1\)) in the original B\&S confusion matrices are ongoing civil war cases which by definition \emph{cannot} experience another civil war onset.
 The alternative outcome coding B\&S evaluate, under ``Assuming Change'', produces the only two civil war onsets we would have observed in 2016-H1.
@@
 \begin{quote}
-Importantly, we make no assumptions about the persistence of civil war between the second half of 2015 and the first half of 2016, meaning that all countries---even those with ongoing conflicts as of December 31, 2015--are included in the 2016 test set.
+Importantly, we make no assumptions about the persistence of civil war between the second half of 2015 and the first half of 2016, meaning that all countries---even those with ongoing conflicts as of December 31, 2015---are included in the 2016 test set.
 As a result, 14 of the 30 countries on the list (and nine of the top 10) are cases of ongoing conflict.
@@
 In the text, B\&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases.
-This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g.~two separatists conflicts in different portions of the country.
+This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g., two separatists' conflicts in different portions of the country.
 But this is not the case; ongoing conflicts in all other analyses are consistently dropped from the data as if a new civil war onset for a country with an ongoing conflict is conceptually impossible.
@@
 
-Conversely, not all factors that are ``theoretically'' important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory.
+Conversely, not all factors that are ``theoretically'' important or suggested by the literature are also important for prediction, and some factors that are good predictors do not originate from prior theory.
 For example, both Ward, Greenhill, and Bakke (2010) and Hill and Jones (2014) examine a variety of variables that are considered ``theoretically'' important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for actually predicting those outcomes.
-It is also possible that building accurate predictive models could uncover factors overlooked in the literature, e.g.~the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al.
+It is also possible that building accurate predictive models could uncover factors overlooked in the literature, e.g., the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al.
 2010), irregular leadership changes (Beger et al.
@@
 Are those relationships linear in the log-odds?
-Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships.
+Some machine learning models, random forests included, are flexible in this regard and, to some extent, are capable of learning non-linear and interactive relationships.
 B\&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives.
@@
 The problems we encountered are not subjective modeling choices.
-They bear no implications on the question of whether theory-based model are preferable.
+They bear no implications on the question of whether theory-based models are preferable.
 However, when we fix these issues and perform an updated analysis, the B\&S conclusions are all essentially overturned.
@@
 See \texttt{1mo\_run\_escalation\_weighted\_PITF.R} line 4, where the PITF predictions are taken from the training data set (\texttt{train\$pred\_prob\_plus1}).
-The next line is a hack extending the shorter \texttt{weight} vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions.
-Similarly in the 6-month version of this file.
+The next line is a hack extending the shorter \texttt{weight} vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions -- similarly in the 6-month version of this file.
 
@@
 The intention is to train separate random forests on data split into high and low-risk groups based on PITF predictions, and then combine them back into one random forest.
-Actually both random forests are trained on identical data, essentially production a random forest that just has N\_trees * 2.
+Actually, both random forests are trained on identical data, essentially production a random forest that just has N\_trees * 2.
 
@@
 \item
-  The nominally 2016-H1 forecasts are created at the bottom of \texttt{+master.R} using the data for 2016-H1 and still the lead DV. Thus the lead DV here refers to outcomes in 2016-H2.
+  The nominally 2016-H1 forecasts are created at the bottom of \texttt{+master.R} using the data for 2016-H1 and still the lead DV. Thus, the lead DV here refers to outcomes in 2016-H2.
 \item
@@
 Here, the scoring is done using the non-lead version of the DV.
-Thus there is a mismatch between the time period the forecasts are for and the period whose observed outcomes are used to score them.
+Thus, there is a mismatch between the time period the forecasts are for and the period whose observed outcomes are used to score them.
 \end{enumerate}
 
-The relevant variables in the data are ``incidence\_civil\_ns'' and ``incidence\_civil\_ns\_plus1'', which appears to be a 1-period lead (ie., \(t_+1\)) version of the dependent variable that is used in the actual prediction models.
+The relevant variables in the data are ``incidence\_civil\_ns'' and ``incidence\_civil\_ns\_plus1'', which appears to be a 1-period lead (i.e., \(t_{+1}\)) version of the dependent variable that is used in the actual prediction models.
 The incidence dependent variable contains both 0/1 and missing values.

### 2020-11-13 rep_nosmooth/README.Rmd
don't change working directories in analysis scripts
Use here() directly for the input/output files instead of changing working directories at the top of each script.

@@
 ---
-title: "Replication materials for Beger, Morgan, Ward, nd"
+title: "Re-analysis of B&S with both smoothed and empirical AUC-ROC"
 output: 
@@
 
-**tl;dr:** run through `master.R`
+The easiest way to reproduce our re-analysis is to run `master.R`, which will 
+run the other scripts in correct order.
+
+- Adjust the number of parallel workers in `2-model-runner.R`
+- Go through `master.R`.
+
+If you want to run without `master.R`, the numbers in the other R scripts
+indicate dependence. For example, the script starting with "3-" for requires 
+all "1-" and "2-" scripts to have been run.
+
+Our re-analysis is based on the original B&S replication materials. The main
+substantive changes are:
+
+- We evaluate AUC values from both the original smoothed ROC curves but also
+  standard empirical ROC curves.
+- Two of the structural model implementations have been changed.
+
+While the substantive changes are minor, we have changed the structure of the
+code substantially so that the random forest models can be run in parallel.
+
+We intentionally do not set a RNG seed, since the substantive conclusions
+of our re-analysis should not depend on particular RNG seed values. There is
+some slight variation between re-runs of the models as a result. There is 
+a provision in the model runner for setting RNG seeds, but the downstream
+scripts won't work with the resulting file names without some manual 
+adjustment.
+
+*****
+
+## Overview
 
@@
 
-Below are instructions for how to run the script in order to replicate the updated analysis. To do a clean replication, delete the `output/` folder; this is not required though.
+Below are instructions for how to manually run the scripts in order to replicate the updated analysis. To do a clean replication, delete the `output/` folder; this is not required though.
 
@@
 Anyways, the model runner script does include a provision for setting a RNG seed, in which case the key output files will be suffixed with the RNG seed value.
-Note that the results processing scripts are not setup to work this these output files, but it should be fairly obvious what lines need modifications to make them work.
+Note that the results processing scripts are not setup to work this these output files.
+Instead of adjusting file names in the downstream scripts, the easiest way to get this working again is to copy the output + RNG seed files and rename them by dropping the RNG seed suffix.
 

### 2020-11-13 rep_nosmooth/1-make-original-tables.R

@@ -4,10 +4,10 @@
 
 library(tibble)
 library(readr)
+library(here)
 
-setwd(here::here("rep_nosmooth"))
-
-dir.create("output/tables", recursive = TRUE, showWarnings = FALSE)
+dir.create(here("rep_nosmooth/output/tables"),
+           recursive = TRUE, showWarnings = FALSE)
 
 table1 <- tribble(
   ~horizon, ~model, ~Escalation, ~Quad, ~Goldstein, ~CAMEO, ~Average,
@@ -46,7 +46,7 @@ table4 <- tribble(
   "Assuming Change", 1L, 2L, 14L
 )
 
-write_csv(table1, "output/tables/table1-original.csv")
-write_csv(table2, "output/tables/table2-original.csv")
-write_csv(table4, "output/tables/table4-original.csv")
+write_csv(table1, here("rep_nosmooth/output/tables/table1-original.csv"))
+write_csv(table2, here("rep_nosmooth/output/tables/table2-original.csv"))
+write_csv(table4, here("rep_nosmooth/output/tables/table4-original.csv"))
 

### 2020-11-13 rep_nosmooth/1-re-save-data.R

@@ -7,12 +7,10 @@ library(readstata13)
 library(readr)
 library(here)
 
-setwd(here::here("rep_nosmooth"))
+dir.create(here("rep_nosmooth/trafo-data"), showWarnings = FALSE)
 
-dir.create("trafo-data", showWarnings = FALSE)
+data_1mo <- read.dta13(here("rep_nosmooth/data/1mo_data.dta"))
+write_rds(data_1mo, here("rep_nosmooth/trafo-data/1mo_data.rds"))
 
-data_1mo <- read.dta13("data/1mo_data.dta")
-write_rds(data_1mo, "trafo-data/1mo_data.rds")
-
-data_6mo <- read.dta13("data/6mo_data.dta")
-write_rds(data_6mo, "trafo-data/6mo_data.rds")
+data_6mo <- read.dta13(here("rep_nosmooth/data/6mo_data.dta"))
+write_rds(data_6mo, here("rep_nosmooth/trafo-data/6mo_data.rds"))

### 2020-11-13 rep_nosmooth/1-setup-model-table.R

@@ -13,7 +13,6 @@ library(purrr)
 library(yaml)
 library(jsonlite)
 
-setwd(here::here("rep_nosmooth"))
 
 
 # Table 1 cell definitions ------------------------------------------------
@@ -596,13 +595,19 @@ stopifnot(
 #   JSON.
 #
 
-dir.create("output/model-definitions", recursive = TRUE, showWarnings = FALSE)
+dir.create(here("rep_nosmooth/output/model-definitions"),
+           recursive = TRUE, showWarnings = FALSE)
 
-write_rds(model_table,    "output/model-definitions/model-table.rds")
-write_json(feature_specs, "output/model-definitions/feature-specs.json",
+write_rds(model_table,
+          here("rep_nosmooth/output/model-definitions/model-table.rds"))
+write_json(feature_specs,
+           here("rep_nosmooth/output/model-definitions/feature-specs.json"),
            pretty = TRUE, auto_unbox = TRUE)
-write_yaml(hp_settings,   "output/model-definitions/hp-settings.yaml")
-write_yaml(dv_specs,      "output/model-definitions/dv-specs.yaml")
-write_yaml(train_end_year, "output/model-definitions/train-end-year.yaml")
+write_yaml(hp_settings,
+           here("rep_nosmooth/output/model-definitions/hp-settings.yaml"))
+write_yaml(dv_specs,
+           here("rep_nosmooth/output/model-definitions/dv-specs.yaml"))
+write_yaml(train_end_year,
+           here("rep_nosmooth/output/model-definitions/train-end-year.yaml"))
 
 

### 2020-11-13 rep_nosmooth/2-model-runner.R

@@ -10,7 +10,14 @@ WORKERS  <- 8
 # last element of its file name.
 RNG_SEED <- NULL
 
-setwd(here::here("rep_nosmooth"))
+
+oldwd <- getwd()
+newwd <- here::here("rep_nosmooth")
+if (oldwd != newwd) {
+  lgr::lgr$info("Temporarily setting wd to '%s'", newwd)
+  setwd(newwd)
+}
+
 
 library(dplyr)
 library(readr)
@@ -367,7 +374,8 @@ if (!is.null(RNG_SEED)) {
     writeLines("output/model-table-w-results.md")
 }
 
-
-
-
+if (oldwd != newwd) {
+  lgr$info("Reseting working directory to '%s'", oldwd)
+  setwd(oldwd)
+}
 lgr$info("Script finished")

### 2020-11-13 rep_nosmooth/4-make-figure1.R

@@ -10,13 +10,12 @@ library(dplyr)
 library(readr)
 library(tibble)
 
-setwd(here("rep_nosmooth"))
-dir.create("output/figures", showWarnings = FALSE)
+dir.create(here("rep_nosmooth/output/figures"), showWarnings = FALSE)
 
-all_predictions <- read_rds("output/all-predictions.rds") %>%
+all_predictions <- read_rds(here("rep_nosmooth/output/all-predictions.rds")) %>%
   group_by(cell_id) %>%
   nest()
-model_table <- read_rds("output/model-table-w-results.rds")
+model_table <- read_rds(here("rep_nosmooth/output/model-table-w-results.rds"))
 
 # We need predictions for Table 1, base specification
 cell_ids <- model_table %>%
@@ -54,7 +53,8 @@ cols <- c(
 # linewidth
 ll <- 4
 
-png("output/figures/fig-1-figure1-replicated.png", width = 1100, height = 1000,
+png(here("rep_nosmooth/output/figures/fig-1-figure1-replicated.png"),
+    width = 1100, height = 1000,
     units = "px", pointsize = 24)
 par(mfrow = c(2, 2))
 

### 2020-11-13 rep_nosmooth/4-make-figure2.R

@@ -9,11 +9,10 @@ library(dplyr)
 library(tidyr)
 library(here)
 
-setwd(here::here("rep_nosmooth"))
-dir.create("output/figures", showWarnings = FALSE)
+dir.create(here("rep_nosmooth/output/figures"), showWarnings = FALSE)
 
-table1_benefit <- read_csv("output/tables/table1-smooth-benefit.csv")
-table2_benefit <- read_csv("output/tables/table2-smooth-benefit.csv")
+table1_benefit <- read_csv(here("rep_nosmooth/output/tables/table1-smooth-benefit.csv"))
+table2_benefit <- read_csv(here("rep_nosmooth/output/tables/table2-smooth-benefit.csv"))
 
 cols <- c(
   Escalation = "#e41a1c",
@@ -60,4 +59,6 @@ p <- benefit %>%
   geom_vline(xintercept = 0, linetype = 3) +
   theme(axis.text.y = element_text(size = 12))
 
-ggsave(plot = p, filename = "output/figures/fig-2-benefit-plot.png", height = 3.5, width = 5.5)
+ggsave(plot = p,
+       filename = here("rep_nosmooth/output/figures/fig-2-benefit-plot.png"),
+       height = 3.5, width = 5.5)

### 2020-11-13 rep_nosmooth/4-make-figureA1.R

@@ -9,10 +9,9 @@ library(dplyr)
 library(tidyr)
 library(here)
 
-setwd(here::here("rep_nosmooth"))
-dir.create("output/figures", showWarnings = FALSE)
+dir.create(here("rep_nosmooth/output/figures"), showWarnings = FALSE)
 
-table1_nosmooth <- read_csv("output/tables/table1-nosmooth.csv")
+table1_nosmooth <- read_csv(here("rep_nosmooth/output/tables/table1-nosmooth.csv"))
 
 icews_lvls <- c("Escalation", "Quad", "Goldstein", "CAMEO", "Average")
 
@@ -41,6 +40,7 @@ p <- pairs %>%
   theme(panel.grid.minor = element_blank()) +
   labs(x = "", y = "AUC-ROC (empirical)")
 
-ggsave(plot=p, filename = "output/figures/fig-A1-table1-pairs.png",
+ggsave(plot=p,
+       filename = here("rep_nosmooth/output/figures/fig-A1-table1-pairs.png"),
        height = 5, width = 8)
 

### 2020-11-13 rep_nosmooth/4-make-figureA2.R

@@ -9,10 +9,9 @@ library(dplyr)
 library(tidyr)
 library(here)
 
-setwd(here::here("rep_nosmooth"))
-dir.create("output/figures", showWarnings = FALSE)
+dir.create(here("rep_nosmooth/output/figures"), showWarnings = FALSE)
 
-table2_nosmooth <- read_csv("output/tables/table2-nosmooth.csv")
+table2_nosmooth <- read_csv(here("rep_nosmooth/output/tables/table2-nosmooth.csv"))
 
 model_lvls <- colnames(table2_nosmooth)[-c(1:2)]
 
@@ -42,6 +41,7 @@ p <- pairs %>%
         axis.text.x = element_text(angle = 30, hjust = 1)) +
   labs(x = "", y = "AUC-ROC (empirical)")
 
-ggsave(plot=p, filename = "output/figures/fig-A2-table2-pairs.png",
+ggsave(plot=p,
+       filename = here("rep_nosmooth/output/figures/fig-A2-table2-pairs.png"),
        height = 5, width = 8)
 

### 2020-11-13 rep_nosmooth/4-make-figureA3.R

@@ -9,10 +9,9 @@ library(pROC)
 library(forcats)
 library(here)
 
-setwd(here::here("rep_nosmooth"))
-dir.create("output/figures", showWarnings = FALSE)
+dir.create(here("rep_nosmooth/output/figures"), showWarnings = FALSE)
 
-preds <- read_rds("output/all-predictions.rds")
+preds <- read_rds(here("rep_nosmooth/output/all-predictions.rds"))
 
 auc_roc_vec <- function(pred, truth, smooth, sm = "binormal") {
   roc_obj <- pROC::roc(truth, pred, auc = TRUE, quiet = TRUE, smooth = smooth, smooth.method = sm)
@@ -33,7 +32,7 @@ rocs <- preds %>%
 #logcondens = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "logcondens"),
 #logcondens.smooth = auc_roc_vec(pred = pred, truth = value, smooth = TRUE, sm = "logcondens.smooth")
 
-mt <- read_rds("output/model-table-w-results.rds")
+mt <- read_rds(here("rep_nosmooth/output/model-table-w-results.rds"))
 mt <- mt %>% left_join(rocs, by = "cell_id")
 
 cols <- c(
@@ -77,7 +76,8 @@ p <- ggplot(benefit, aes(x = value, y = column, color = column)) +
   geom_vline(xintercept = 0, linetype = 3) +
   theme(axis.text.y = element_text(size = 12))
 
-ggsave(plot = p, filename = "output/figures/fig-A3-benefit-plot-extended.png",
+ggsave(plot = p,
+       filename = here("rep_nosmooth/output/figures/fig-A3-benefit-plot-extended.png"),
        height = 3.5, width = 7.5)
 
 

### 2020-11-13 rep_nosmooth/master.R

@@ -2,6 +2,15 @@
 #   Replication master script
 #
 
+# The scripts use the here package to set working directories. The root
+# directory should be the Blair-Sambanis-replication folder, and this should
+# work out of the box if you cloned the whole repo.
+# Otherwise, read ?here for how to get here to recognize the correct root
+# directory.
+#
+# NB: in 2-model-runner.R, check and adjust WORKERS on line 7
+#
+
 # Make sure required packages are installed
 source("0-packages.R")
 
@@ -15,6 +24,7 @@ source("1-re-save-data.R")
 source("1-setup-model-table.R")
 
 # Run models (this should take about 1 hour)
+# NB: check and adjust WORKERS on line 7 before running this
 source("2-model-runner.R")
 
 # Recreate Tables 1 and 2 from new output

### 2022-07-25 rep_nosmooth/README.Rmd
require version 4.6-14 of randomForest (see #12)

@@
 adjustment.
+
+_Note (July 2022): The latest version of randomForest on CRAN leads to a memory issue when trying to the part of the replication code that trains all of the models.
+Memory usage increases until the process errors out or system crashes.
+Thus the old randomForest 4.6-14 is required for replication (this is also the version we used when originally conducting the replication).
+The package setup script will check for this, and issue guidance as needed.
+Thanks to Christian Oswald for raising this issue.
+(#12)_
 

### 2022-07-25 rep_nosmooth/0-packages.R

@@ -12,6 +12,23 @@ check_packages <- function() {
 
   not_installed <- need[!need %in% installed]
 
+  # special treatment for randomForest (#12)
+  if ("randomForest" %in% not_installed) {
+    message("Please install version 4.6-14 of randomForest from source:\n\n",
+            "devtools::install_version(\"randomForest\", version = \"4.6-14\")",
+            "\n\nSee https://github.com/andybega/Blair-Sambanis-replication/issues/12 for more details.\n")
+  } else {
+    # make sure correct version of randomForest is at hand
+    ver <- installed.packages()["randomForest", "Version"]
+    if (ver > "4.6-14") {
+      message("Due to a memory issue with the latest version of {randomForest} on CRAN, ",
+              "an older version of the package is needed.\n",
+              "See https://github.com/andybega/Blair-Sambanis-replication/issues/12 for more details.\n\n",
+              "devtools::install_version(\"randomForest\", version = \"4.6-14\")")
+      return(invisible(FALSE))
+    }
+  }
+
   if ("dplyr" %in% installed) {
     # need v 1.0.0 or greater in order to be able to group_by a list column in
     # 3-recreate-tables.R
